### 2018-02-21 tex/paper.tex
Add skeleton for paper

\documentclass[
      acmsmall
    , screen
%    , authordraft
%    , anonymous
    , review=true
  ]{acmart}

% Metadata Information
\acmConference[ICFP]{International Conference on Functional Programming}{23--29 September 2018}{St.~Louis, Missouri, United States}
%\acmJournal{PACMPL}
%\acmVolume{9}
%\acmNumber{4}
%\acmArticle{39}
%\acmYear{2010}
%\acmMonth{3}
%\copyrightyear{2009}
%\acmArticleSeq{9}

% Copyright
%\setcopyright{acmcopyright}
\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
%\acmDOI{0000001.0000001}

% Paper history
%\received{February 2007}
%\received[revised]{March 2009}
%\received[accepted]{June 2009}

\citestyle{acmauthoryear}

%%% PACKAGES %%%

\usepackage[inline]{enumitem}
\usepackage[nomargin,inline,draft]{fixme}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{pbox}
\usepackage{tikz}
\usepackage{url}

% Must come after hyperref
\usepackage[capitalise]{cleveref}

%%% FIXME PACKAGE SETUP %%%

\fxusetheme{color}
\fxuseenvlayout{color}

\FXRegisterAuthor{pls}{apls}{\color{teal}[Pablo]}
\FXRegisterAuthor{sjt}{asjt}{\color{magenta}[Simon]}

%%% MACROS %%%

% Document starts
\begin{document}

% Title portion. Note the short title for running heads
\title
  [Untitled]
  {Untitled}

\author{Pablo Lamela Seijas}
\affiliation{%
  \institution{University of Kent}
%  \streetaddress{}
  \city{Canterbury}
  \state{Kent}
  \postcode{CT2 7NZ}
  \country{UK}}
\email{pl292@kent.ac.uk}

\author{Simon J. Thompson}
\orcid{0000-0002-2350-301X}
\affiliation{%
  \institution{University of Kent}
%  \streetaddress{}
  \city{Canterbury}
  \state{Kent}
  \postcode{CT2 7NZ}
  \country{UK}}
\email{sjt@kent.ac.uk}

\begin{abstract}
  No abstract
\end{abstract}

% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below.

%
% End generated code
%

\keywords{
  No keywords
}

\maketitle

% If the default list of authors is too long for headers, it can be redefined
%\renewcommand{\shortauthors}{}

\section{Introduction}

No introduction \cite{garfield1996cite}.

\clearpage
\bibliographystyle{ACM-Reference-Format}
\bibliography{paper}

\end{document}

### 2018-02-26 tex/paper.tex
Minor changes, suggested title.
incorporated and reformatted the google docs paper (except for RPS)

@@
 \usepackage{url}
+\usepackage{alltt}
 
@@
 \title
-  [Untitled]
-  {Untitled}
+  [Marlowe]
+  {Marlowe: financial contracts on blockchain}
 
@@
 
-\author{Simon J. Thompson}
+\author{Simon Thompson}
 \orcid{0000-0002-2350-301X}
@@
   \country{UK}}
-\email{sjt@kent.ac.uk}
+\email{s.j.thompson@kent.ac.uk}
 
@@
 
+\section{Contract DSL model discussion, Version 1.1, September 2017}
+The aim of this document is to set out the assumptions behind the Haskell model of a contract DSL blockchain. Code for this model is in the repository \url{https://github.com/input-output-hk/scdsl}
+
+The document is structured like this:
+
+\begin{enumerate}
+\item An overview of the model, including the assumptions made in designing it, the types of the principal functions and a description of the DSL as an algebraic type, constructor by constructor.
+The section aims to give a high level overview, and some of the concrete details are to be found in the repo.
+\item A discussion of the kinds of formal analysis which might be made for the DSL.
+\item A set of questions raised by the process of modelling.
+\item A response to some particular points relevant to this work and raised in the email discussion on ``Some Thoughts on Smart Contracts'' initiated by Charles, and taken up by Duncan and Phil.
+\item Background: an overview of the Rock-Paper-Scissors example from the repo and the discussion.
+\item A response to Charles’ comments as posted to slack.
+\end{enumerate}
+
+\subsection{The model}
+The DSL is modelled as an algebraic type in Haskell, together with an executable small-step semantics. 
+A running contract will interact with its environment in two ways.
+
+First, it will need to observe different kinds of varying quantities including, for example, the current time, the current block number and, random numbers, as well as ``real world'' quantities like ``the price of oil'' or ``the exchange rate between currencies A and B''.
+Each instance of such an observable will be recorded on the blockchain to allow the computation to be repeated for verification purposes.
+It is assumed that at each step of the execution of the contract, the values of some observables will be needed.
+
+Second, it will be necessary to ensure that commitments of various kinds have been made by the participants in a contract. We have included two kinds of commitments here:
+\begin{itemize}
+\item A commitment to a value: this would include a choice of a play in a game, which would be chosen and concealed from the other players until a later point in play, when it is revealed.
+
+\item A commitment to a certain amount of cash: while a commitment to a value is indefinite, a cash commitment needs to have a timeout so that progress can be forced in a game.
+After the timeout period the cash is automatically refunded; in practice this might require the user explicitly to create a transaction to reclaim the cash.
+\end{itemize}
+
+At each step, a number of commitments may be made.
+This information constitutes an input to the execution of each step.
+Moreover, information about the commitments currently in force forms the state, which is (potentially) modified at each execution step.
+
+Finally, a running contract will initiate a number of payment actions on the blockchain at each step, represented by a list of actions.
+
+This analysis justifies the type of the step function:
+
+\begin{verbatim}
+step :: Commits -> State -> Contract -> OS -> (State,Contract,AS)
+\end{verbatim}
+
+Where \verb+OS+ contains the current values for observables, and \verb+AS+ is a list of actions (currently payments) that are carried out as a result of executing the step.
+
+The model is based on a number of assumptions.
+
+\begin{itemize}
+\item It is assumed that cryptographic ``services'' or operations are provided by a wrapper layer external to the system, and are not modelled explicitly here.
+\item Making a commitment is not something that a contract can perform; rather, it can require that a commitment is made, but that then has to be established externally: hence the input of (a set of) commitments at each step.
+The same applies to revealing a value commitment, and to manually redeeming a cash commitment.
+\item Observables are observed at a particular time and in a particular context.
+We assume that the recording and identification of each (instance of an) observation is provided by the infrastructure of the system.
+We provide a ``little language'' for describing observations: Observation\footnote{Why doesn't Observation appear in the type of step?
+I would expect a contract would need to yield a request for observations to be made, and accept the value of those observations.}\footnote{The current values of *observables* required at each step are the input OS to the step function.
+We assume that the infrastructure is designed to record these, but that is not part of the DSL itself.
+The *observations* over OS are part of the contract (e.g.
+the Choice constructor).}
+\item The fact of making commitments also needs to be recorded.
+\item It is important to be aware that there may be different instances of contracts executed, and for each instance there will typically be different instances of observables (and commitments) observed and recorded.
+\item The model manages the release of funds back to the committer when a cash commitment expires (see discussion of the full\_step function below).
+\end{itemize}
+
+The type of contracts is given by
+
+\begin{alltt}
+data Contract =
+   Null |
+   RedeemCC IdentCC Contract |
+   RevealCV IdentCV Contract |
+   Pay Person Person Int Contract |
+   Both Contract Contract |
+   Choice Observation\footnote{Doesn't the step function need to record what observations are made, so they can be archived in the blockchain?
+I don't see how your design will support doing so.}\footnote{Yes, we assume that it is part of the infrastructure of this language to record the values of *observables* that are used.
+However, we have chosen no to make it part of the language itself.} Contract Contract |
+   CommitValue IdentCV Person [Value] Contract |
+   CommitCash IdentCC Person Cash Time Contract |
+   When Observation Time Contract Contract
+\end{alltt}
+
+The step function is total, so that for every contract a result of stepping is defined.
+However, for some kinds of contracts – commits, redeems or time-shifted contracts – it is possible that the result of a step is to produce the same contract as the result; such a step is called quiescent whereas all other steps make progress.
+We use this distinction in the explanation that follows.
+
+At each computational step, the step function is wrapped by a call to full\_step which in addition will cause expired cash commitments to be refunded.
+
+The constructors should be interpreted thus:
+
+\begin{description} 
+\item[\texttt Null] is the null contract; it will always be quiescent.
+\item[\texttt RedeemCC idCC k] (CC stands for cash commitment.) For this contract to make progress, the creator of the cash commitment with identifier idCC is required to redeem the unspent funds in that commitment; the contract then continues as k.
+If the cash commitment with identifier idCC has expired, the remaining funds will be redeemed automatically by the full\_step function and a ExpiredRedeem action will be produced\footnote{How does the contract determine what actions resulted from a given step?
+Or is this not required?}\footnote{
+If it is required, restructuring the DSL so that the result action appears as the 'value' of executing a step (using a monadic interface as described for id generation below) would be a good idea.}\footnote{We specify the actions generated by each kind of constructor; at each step the list of such actions, of type AS, is a component of the result.}.
+Using RedeemCC to redeem the same identifier idCC twice will result in the issue of a DuplicateRedeem action.
+Otherwise the contract is quiescent, waiting for the commitment to be redeemed or to expire.
+\item[\texttt RevealCV idCV k] (CV stands for value commitment.) For this contract to make progress, the creator of the value commitment with identifier idCV is required to reveal the value; the contract then continues as k.
+Otherwise the contract is quiescent, waiting for the value to be revealed.
+For the reveal to be valid, the value revealed must be in the vals list provided in the corresponding CommitValue constructor (the one with the same idCV). 
+\item[\texttt Pay p1 p2 int k] requires that int is paid by p1 to p2, and continues as k.
+It uses the currency allocated by the cash commitments made by p1 that expire the earliest.
+This contract will result in a FailedPay action if the funds are not available; otherwise a SuccessfulPay action is generated.
+\item[\texttt Both k1 k2] enforces the behaviour of both contracts k1 and k2.
+Because the model is stateful and produces output actions, to make a step it is necessary to execute a single step of each of  the contracts k1 and k2 in sequence: first k1 then k2.
+\item[\texttt Choice obs k1 k2] behaves as either k1 or k2 depending on the (Boolean) result of obs at the time that the observation is made, k1 if it is True and k2 if False.
+\item[\texttt CommitValue idCV p vals k]  For this contract to make progress, the user p is required to make a value commitment with identifier idCV; the contract then continues as k.
+In order for the commitment to be valid, the value must be one of the specified in the vals list.
+Otherwise the contract is quiescent, waiting for the value to be committed.
+\item[\texttt CommitCash idCC p n timeout k]  For this contract to make progress, the user p is required to make a cash commitment of value n and timeout timeout (specified as a block number) with identifier idCC; the contract then continues as k.
+Otherwise the contract is quiescent, waiting for the cash to be committed.
+When the timeout is reached, the remaining funds will be redeemed automatically and a SuccessfulPay action will be issued that indicates a self-payment from the person p with the remaining funds.
+This is performed by the full\_step function, which wraps the step function.
+\item[\texttt When obs timeout k def\_k] This contract will not progress until obs is True or until the current block number is greater than or equal to the one specified by timeout.
+In case the timeout applies, the contract will continue as def\_k, if the timeout does not apply and obs is True, then the contract continues as k.
+Otherwise the contract is quiescent.
+\end{description}
+
+\subsection{Analysis}
+What questions can we ask? What answers can we give?
+
+\subsubsection{What are the valid contracts?}
+
+One possible definition is ``those contracts that do not lead to a FailedPay action''.
+Does this make sense independently of the inputs to a contract (i.e.
+observables, commitments, redeems, time, and randomness)?
+Some relevant points (NB discussion incomplete):
+\begin{itemize}
+\item For Pay require a sufficient balance not to produce a FailedPay action. 
+\item For Both require a sufficient balance for both contracts combined.
+Need to think here whether the two conjuncts should be independent, or whether a commitment in one half can be used in the other.
+\item For Choice require a sufficient balance for the maximum of the requirements of the individual contracts. 
+\item For When, need to ``age'' the existing commitments relative to the new starting point of the continuation contract. 
+\item Are there any constraints on  Commit*, RevealCV and RedeemCC statements?
+Not directly, but there need to be sufficient funds available to pay if needed, or enough time for the commitments to time out, if not.
+\item Originally we had included that we should include all the ``progress'' requirements from point 2 below; but that’s probably not the right thing here.
+\end{itemize}
+
+\subsubsection{Do all contracts potentially evaluate to Null?}
+
+That is, for each contract do there exist sequences of values of Commit and OS such that stepping the contract with these inputs leads to Null? Might we have to rule out unsatisfiable observations?
+Some relevant points (NB discussion incomplete):
+\begin{itemize}
+\item For When, need to be able to make interpretObs st obs os true. Since there is a timeout to When then there’s no need to impose this condition, as we’ll progress at timeout if not before.
+\item For Commit*, need to have the right commits in the Commit argument.
+\item Similarly for RevealCV and RedeemCC statements.
+\item It would also be possible to put timeouts on to Commit, RevealCV and RedeemCC operations, and if this were to be done there would be no remaining requirements for contract execution to make progress.
+\item There are no constraints on other constructs.
+\end{itemize}
+
+\subsection{Questions}
+
+The model outlined here raises a number of questions, enumerated here.
+
+\begin{enumerate}
+\item Identifiers for commitments are part of the model, and supplied as part of each contract.
+It might well be that these should be generated by the system; this could guarantee uniqueness, freshness etc, and could also lead to a monadic DSL, with bind used for binding names to commitments.
+\item Does the Both constructor potentially produce invalid/misleading contracts, combining a commitment in one half with its use in the other?
+The problem is that the ``use'' conjunct could potentially be executed, leading to a failed payment.
+\item How do we ensure that contracts can be enforced? Is there a notion of validity for contracts that can be enforced? To which extent is this notion enforced by current constructors?
+\item Should there be a timeout on the process of making or revealing a commitment, or is it sufficient to have timeouts on the commitments themselves?
+\item (Related to 3) Is this also an argument to have timeouts on value commitments too?
+\item The model takes time as an observable.
+The type Time represents the index of evaluation, and BlockNumber is the number of the current block of the blockchain.
+We’ve not included a ``real'' time, but should this be included?
+\item (Related to 5) What should denominate the timeouts on commitments: Time or BlockNumber or another notion of time? Currently we are using the latter.
+\item Is it possible for block number to depend on progress of computation?
+We conjecture not, because otherwise it would be possible to produce timelocks, under which it becomes impossible for time to progress.
+\item Does a contract expire if a surrounding commitment does? We think not.
+\item The model is deterministic, with all non-determinism being external, and confined to observables.
+This is surely necessary in order to support computation replay, but does it restrict the expressibility of the model?
+\item Is there a notion of causality for these contracts? Would we be able to characterise it semantically? Syntactically?
+\item We have modelled cash as fungible: we make payments using the cash commitments in order of expiry time, with those expiring earliest being used first.
+Would there be any reason for using a different approach?
+\item Timeouts are expressed in absolute time: presumably they could be relative (with recording on the blockchain of the times at which they are triggered).\footnote{This contradicts what you've said elsewhere.
+I think timeouts should be times, not blockchain numbers.
+I suspect it would be better not to use blockchain numbers.
+I don't want to have to react more quickly if lots of people are making transactions that day.
+Also, it avoids question 8.}\footnote{See addition about slot time from Ouroboros paper (as referenced in the body of the paper).}
+\end{enumerate}
+
+Added: we should probably use ``slot time'' https://eprint.iacr.org/2016/889.pdf ``We consider a setting where time is divided into discrete units called slots.
+A ledger, described in more detail below, associates with each time slot (at most) one ledger block.''
+\begin{enumerate}
+\item Some observables have a ``monotonicity'' property: once something is revealed, it is revealed forever; on the other hand, something unrevealed is only unrevealed so far; do we need to reflect this in the observations that we allow?
+\item The seesaw construction is key to implementing e.g.
+poker, and it can be described in terms of the existing constructs (modulo the answer to point 4 above).
+However, should seesaw be a primitive in its own right?\footnote{No.
+You will have a successful language when you can define seesaw as an abstraction in terms of the other primitives *within the language*}\footnote{And indeed we *did* define it, at least for the RPS example.}
+
+\item Constructs that can time out could have a single continuation (do this next) or two continuations: do this if progressed normally, and do the other continuation if timed out.
+A candidate for this is a timed out When, where the second option is what we have implemented.
+\item RedeemCC is intended for premature repayment of a cash commitment, i.e. before the commitment has timed out.
+\footnote{Why is RedeemCC a separate construct? Can't it be implemented with Pay?}\footnote{It is for someone to claim their cash back, rather than making a payment to another person.}
+ This is potentially dangerous – leading to a FailedPay, for example – but useful to make a repayment at the end of a contract. 
+\item Alternatively, it would be possible for the underlying model to repay all committed but unspent cash automatically at the end of a contract, but we might want to do the same thing at the end of each ``round'' of  a game, for instance.
+\end{enumerate}
+
+\subsection{Response to ``Some Thoughts on Smart Contracts''}
+
+The contracts implemented in the DSL language have an implicit interface defined for what would correspond to the client side of a Dapp, as shown by the interactive interface.
+
+We have designed the contract DSL assuming that each contract would be compiled into
+\begin{itemize} 
+\item one or more smart contracts (potentially written in Plutus) and 
+\item a client side program or interface that has a limited number of choices to be made at each point in time (including an interface for the observables). 
+\end{itemize}
+
+Of course, we probably do not want to give a user a programmatic interface but an actual application with user interface.
+So there is still some work to be done to create interfaces on top of the client side of the contracts.
+
+About cost, there are two types of cost (at least):
+
+\begin{itemize} 
+\item The cost of a single transaction execution (a single Plutus validator).
+\item The cost of the whole contract, which may consists of many transactions.
+\end{itemize}
+
+Even in Ethereum this distinction is important because single contract execution is limited by gas (and is thus finite), whereas the life of the whole contract may be unbounded and consist of an indefinite number of transactions.
+
+So far we are assuming contracts defined in the DSL to be finite in both aspects. They could be infinite if we were to define them – co-inductively – as recursive members of the DSL type.
+
+\subsection{Rock-paper-scissors explanation (not re-formatted)}
+In this section, we explain how the DSL can be used for designing a contract that enforces a round of rock-paper-scissors where the winner will receive from the loser 10 units of currency and not revealing your hand also has a penalty of 10 units.
+
+We will start top to bottom. First we define identities for the players and the commitments.
+The player 1 will be identified by integer 1:
+player1 = 1
+the cash commitment of player 1 will be identified by:
+p1 = IdentCC 1
+and his value commitment will be identified by:
+v1 = IdentCV 1
+In the same way, player 2 will be identified by integer 2:
+player2 = 2
+the cash commitment player 2 will be identified by:
+p2 = IdentCC 2
+And his value commitment will be identified by:
+v2 = IdentCV 2
+
+And because choices must be represented as Int we encode them as follows:
+
+data RPS = Rock | Paper | Scissors
+
+all\_rps = [Rock, Paper, Scissors]
+
+encoding\_rps Rock = 1      -- rock is encoded as 1
+encoding\_rps Paper = 2     -- paper is encoded as 2
+encoding\_rps Scissors = 3  -- scissors is encoded as 3
+
+all\_rps\_encoded = map (encoding\_rps) all\_rps
+
+We first ask the players to commit money.
+We will need 20 units from each player since we want 10 units to enforce that the hand is revealed, and 10 units for the prize.
+Theoretically we could reuse just 10 units, but in practice recommitting payments may limit the applicability of the DSL to certain systems like Bitcoin, so it does not support it at the moment.
+
+The commitments are chained as follows:
+* Cash commitment p1 is done by player 1, consists of 20 units and expires on block 42 same for p2 by player 2.
+
+CommitCash p1 player1 20 42
+  (CommitCash p2 player2 20 42
+
+* Value commitments are done for two choices of one of the three values that encode rock, paper, or scissors.
+
+     (CommitValue v1 (all\_rps\_encoded) player1
+        (CommitValue v2 (all\_rps\_encoded) player2
+
+* The see-saw mechanism ensures that both players reveal or none does.
+This is fundamentally achieved by forcing a player to reveal the other when he reveals.
+But we need to do this steadily to give players enough time to do the actions required by the rest of the contract, otherwise one player may push the other one to timeout.
+
+    (RevealCV v1
+            (Choice (BelowTimeout 10)
+                    (Pay player2 player1 10
+                       (RevealCV v2
+                              (Choice (BelowTimeout 20)
+                                  (Pay 1 2 10
+                                       reward\_winner)
+                                  Null)))
+                    Null)))))
+
+First RevealCV waits for the owner of value commitment v1 (player 1) to reveal his choice.
+If the reveal happens before block number 10 (after the start of the contract instance), player 2 pays 10 currency units to player 1.
+Otherwise we consider the contract timed out and make the contract Null, the existing cash commitments will be automatically refunded on block 42.
+The same is done for player 2.
+This time he is incentivised to reveal his hand before block number 20, since that is the only way he will recover the 10 currency unit.
+He might already know that he has lost and decide not to reveal, that will have the same effect than revealing.
+If desired we could increase the see-saw penalty to make not revealing worse than losing.
+
+If the contract remaining becomes reward\_winner, we already have enough information to know who won and we will have exactly 10 currency units committed by each player.
+
+So we define reward\_winner as follows. We just use Choice twice, one to check if Player 1 won, one to check if Player 2 won, and the default option is just refunding everything:
+
+-- If player 1 won
+Choice (OrObs (AndObs (CVRevealedAs v1 (encoding\_rps Paper))
+                      (CVRevealedAs v2 (encoding\_rps Rock)))
+              (OrObs (AndObs (CVRevealedAs v1 (encoding\_rps Scissors))
+                             (CVRevealedAs v2 (encoding\_rps Paper)))
+                     (AndObs (CVRevealedAs v1 (encoding\_rps Rock))
+                             (CVRevealedAs v2 (encoding\_rps Scissors)))))
+-- player2 pays 10 currency units to player 1
+       (Pay player2 player1 10
+-- and everything else is redeemed
+          (Both (RedeemCC p1 Null)
+                (RedeemCC p2 Null)))
+-- if player 2 won
+       (Choice (OrObs (AndObs (CVRevealedAs v2 (encoding\_rps Paper))
+                              (CVRevealedAs v1 (encoding\_rps Rock)))
+                      (OrObs (AndObs (CVRevealedAs v2 (encoding\_rps Scissors))
+                                     (CVRevealedAs v1 (encoding\_rps Paper)))
+                             (AndObs (CVRevealedAs v2 (encoding\_rps Rock))
+                                     (CVRevealedAs v1 (encoding\_rps Scissors)))))
+-- player1 pays 10 currency units to player 2
+               (Pay player1 player2 10
+-- and everything else is redeemd
+                  (Both (RedeemCC p1 Null)
+                        (RedeemCC p2 Null)))
+-- otherwise everything else is redeemed
+               (Both (RedeemCC p1 Null)
+                     (RedeemCC p2 Null)))
+
+It is also worth noting that, because the possible choices are specified, RedeemCV requires that the choice revealed is valid, otherwise it behaves as nothing was revealed.
+Thus, we know that if no player won it must be a draw.
+
+\subsection{Response to comments on the original document made by Charles on Slack}
+
+Legend: {\it a quote from this document}, Charles' response, {\bf our reply}:
+
+{\it
+First, it will need to observe different kinds of varying quantities including, for example, the current time, the current block number and, random numbers, as well as ``real world'' quantities like ``the price of oil'' or ``the exchange rate between currencies A and B''.
+Each instance of such an observable will be recorded on the blockchain to allow the computation to be repeated for verification purposes.
+It is assumed that at each step of the execution of the contract, the values of some observables will be needed.
+}
+
+We need to get a better notion of observation.
+
+There seem to be two broad categories of events worth observing. One that is derived from the network itself and one that is from an external source.
+
+For example, we could consider a game where the contract executes based upon the value of a block header, this is something known after it occurs and cannot be changed.
+Whereas the price of oil is not.
+
+{\bf
+Indeed, we agree.
+The former kind of observations are already artefacts on the blockchain, whereas the latter need to be stored on the blockchain.
+We have assumed that the infrastructure for the language supports this recording: we would not expect language users to have to do this for themselves.
+}
+
+{\bf
+Apart from this difference in the infrastructure, we see no difference between the two different sorts of observations: is this correct?
+}
+
+The question of commitment to blockchain is interesting as who should have write access and also how do we prevent an enormous amount of bad data from being posted? And is it necessary at all?
+
+We could specify the format of an answer and have them commit it to the blockchain as a signed hash and then it could be externally shared for example and everyone would be able to determine it is correct.
+
+{\bf
+We saw that commitments were an integral part of the game playing process, and so we wanted to include them as primitives in the language.
+Yes, they are usually implemented as hashes, but we wanted to give a more abstract presentation here.
+
+Concerning bad data, a commitment will usually just store the hash, but the problem seems to come at the reveal, when all the data has to be shown, somehow.
+We don’t have to store the whole data (just the hash) but it does have to be revealed in some way that can be observed, and the naive interpretation of this is that it is made available somehow, thereby using storage on the blockchain.
+
+There may be other ways of doing this. Is there some sort of compact proof that can be used here?
+
+Or can we devolve to a trusted third party, off chain?
+}
+
+I'm also curious about expectations. For example, say we have an oil contract.
+
+We have an expectation of exchange rates or something like that so we'd expect some int to float and then a float to float. Examples:
+
+   * 1 barrel for 120.55 dollars.
+   * 22 dollars for 1 ada.
+
+Gives us everything we need to act on the contract accordingly in a deterministic way.
+
+First, we guess these may be fixed or rational rather than float :-) We would assume that exchange rates are observables, with values recorded at the point of use, so that re-computation is possible (and deterministic).
+
+{\bf
+In general, we need a more powerful expression language (e.g. through embedding the DSL). Maybe we could use a subset of Plutus.
+}
+
+{\it
+Second, it will be necessary to ensure that commitments of various kinds have been made by the participants in a contract.
+}
+
+Ok, Bernardo knows a lot more about this than I do. He's been doing this type of stuff with his MPC work. I'd recommend having a discussion with Aggelos and Bernardo about
+commitments in the cryptocurrency world.
+
+{\bf
+Yes, certainly, we will do this.
+}
+
+{\it
+What are the valid contracts?
+One possible definition is ``those contracts that do not lead to a FailedPay action''.
+Does this make sense independently of the inputs to a contract (i.e.
+observables, commitments, redeems, time, and randomness)?
+}
+
+I tend to think of compositions.
+So we have things like multisig where the players setup a special contract tailored to a set of business and trust assumptions. This game ends when funds are spent.
+Complexity could be fairly arbitrary as the conditions to spend could be quite complex and involve lots of business logic. The end of the game could initiate a new game.
+
+{\bf
+A game that starts a new game is essentially a bigger game.
+There seems to be a limitation about commitments.
+Commitments need to be voluntary by nature, so if there are no commitments inherited between the two games, then it should be no different to model both the games separately.
+
+There should not be a problem to model infinite games either by using co-induction.
+}
+
+For example creating a new multisig account with the same logic for the change address.
+
+Then we have event based games, where actions occur when something happens, a triggering event.
+
+For example a deadman switch.
+
+If bob doesn't sign a special transaction each month, then send all the money to alice.
+
+These are at the mercy of the oracle connected to the contract.
+
+Then we have games of interaction, such as poker for example.
+
+Bob and Alice have earned the right to interact with each other with special rules and pending how those interactions end up, we have a transfer of value between them and others.
+
+Abstractly, I suppose there is a notion of entry condition, allowed actions and exit conditions.
+
+{\bf
+Not sure what entry and exit conditions are, but commitments may well be entry conditions, or at least a crucial component of them.
+We think that all the examples described can be implemented in the DSL (assuming we add the proper observables and oracles).
+Maybe we need some more cryptographic primitives for poker, that is probably something we should discuss with Aggelos and Bernardo.
+}
+
+{\it
+Null is the null contract; it will always be quiescent.
+}
+
+Just to clarify, null means termination or just dormancy?
+
+{\bf
+Both. A contract has terminated if it has evolved to be exactly null. But there exist contracts that contain the null contract but have not terminated.
+}
+
+{\it
+Do all contracts potentially evaluate to Null?
+That is, for each contract do there exist sequences of values of Commit and OS such that stepping the contract with these inputs leads to Null?
+Might we have to rule out unsatisfiable observations?
+}
+
+If I understand null to mean the contract has finished all possible execution, then this is a reasonable assumption given a certain viewpoint completion of a multisig contract for example could be interpreted as a null case.
+
+But in a strict sense the answer is no, there can be cases where something is in a pending state indefinitely.
+For example, for a 2 of 3 contract, two keyholders could die then you'd have an active contract waiting for completion, but it will never happen.
+
+{\bf
+I understand this as: all contracts should reduce to null, but not independently of the observables.
+We can then think about analysing which observables stall the contract (which sounds like a boolean satisfiability problem).
+Maybe we just want to guarantee that there is a set of external actions that will cause the contract to reduce to null no matter its state.
+
+In the particular example of death of participants, we would assume that the contract would have been formulated with timeouts on inactivity of participants (when activity requested).
+}
+
+What we need to do is determine what functionality can be black-boxed in this model with some degree of reliability.
+
+Bernardo's MPC library is going to provide a lot of functionality for things like games for example and provides strong cryptographic guarantees for outcomes given some initial assumptions.
+
+{\bf
+We will definitely incorporate these.
+}
+
+Second, we need to have a set of canonical actions that we could test against the proposed model. And see how they would be constructed.
+
+{\bf
+Is this like a test suite? Not sure how it would work, but would like to know more.
+}
+
+Third, we need to clearly understand what is an oracle in this model and how we intend on using them.
+
+{\bf
+Good point, we need to define an interface for oracles.
+}
+
 \clearpage

### 2018-02-27 tex/paper.tex
formatting changes.

@@
 \usepackage{alltt}
+%\usepackage[dvipsnames]{xcolor}
 
@@
 %%% MACROS %%%
+
+\definecolor{myblue}{rgb}{0.2, 0.3, 1.0}
+\newcommand{\bluett}[1]{\textcolor{myblue}{\texttt{#1}}}
+\newenvironment{allbluett}{\begin{alltt}\color{myblue}}{\end{alltt}}
 
@@
 
-\begin{verbatim}
+\begin{allbluett}
 step :: Commits -> State -> Contract -> OS -> (State,Contract,AS)
-\end{verbatim}
+\end{allbluett}
 
-Where \verb+OS+ contains the current values for observables, and \verb+AS+ is a list of actions (currently payments) that are carried out as a result of executing the step.
+Where \bluett{OS} contains the current values for observables, and \bluett{AS} is a list of actions (currently payments) that are carried out as a result of executing the step.
 
@@
 
-\begin{alltt}
+\begin{allbluett}
 data Contract =
@@
    When Observation Time Contract Contract
-\end{alltt}
+\end{allbluett}
 
@@
 
-\begin{description} 
-\item[\texttt Null] is the null contract; it will always be quiescent.
-\item[\texttt RedeemCC idCC k] (CC stands for cash commitment.) For this contract to make progress, the creator of the cash commitment with identifier idCC is required to redeem the unspent funds in that commitment; the contract then continues as k.
-If the cash commitment with identifier idCC has expired, the remaining funds will be redeemed automatically by the full\_step function and a ExpiredRedeem action will be produced\footnote{How does the contract determine what actions resulted from a given step?
+\begin{itemize} 
+\item \bluett{Null} is the null contract; it will always be quiescent.
+\item \bluett{RedeemCC idCC k} (\bluett{CC} stands for cash commitment.) For this contract to make progress, the creator of the cash commitment with identifier \bluett{idCC} is required to redeem the unspent funds in that commitment; the contract then continues as \bluett{k}.
+If the cash commitment with identifier \bluett{idCC} has expired, the remaining funds will be redeemed automatically by the \bluett{full\_step} function and a \bluett{ExpiredRedeem }action will be produced\footnote{How does the contract determine what actions resulted from a given step?
 Or is this not required?}\footnote{
-If it is required, restructuring the DSL so that the result action appears as the 'value' of executing a step (using a monadic interface as described for id generation below) would be a good idea.}\footnote{We specify the actions generated by each kind of constructor; at each step the list of such actions, of type AS, is a component of the result.}.
-Using RedeemCC to redeem the same identifier idCC twice will result in the issue of a DuplicateRedeem action.
+If it is required, restructuring the DSL so that the result action appears as the 'value' of executing a step (using a monadic interface as described for id generation below) would be a good idea.}\footnote{We specify the actions generated by each kind of constructor; at each step the list of such actions, of type \bluett{AS}, is a component of the result.}.
+Using \bluett{RedeemCC} to redeem the same identifier \bluett{idCC} twice will result in the issue of a DuplicateRedeem action.
 Otherwise the contract is quiescent, waiting for the commitment to be redeemed or to expire.
-\item[\texttt RevealCV idCV k] (CV stands for value commitment.) For this contract to make progress, the creator of the value commitment with identifier idCV is required to reveal the value; the contract then continues as k.
+\item  \bluett{RevealCV idCV k} (\bluett{CV} stands for value commitment.) For this contract to make progress, the creator of the value commitment with identifier \bluett{idCV} is required to reveal the value; the contract then continues as \bluett{k}.
 Otherwise the contract is quiescent, waiting for the value to be revealed.
 For the reveal to be valid, the value revealed must be in the vals list provided in the corresponding CommitValue constructor (the one with the same idCV). 
-\item[\texttt Pay p1 p2 int k] requires that int is paid by p1 to p2, and continues as k.
+\item \bluett{Pay p1 p2 int k} requires that int is paid by p1 to p2, and continues as k.
 It uses the currency allocated by the cash commitments made by p1 that expire the earliest.
 This contract will result in a FailedPay action if the funds are not available; otherwise a SuccessfulPay action is generated.
-\item[\texttt Both k1 k2] enforces the behaviour of both contracts k1 and k2.
+\item \bluett{Both k1 k2} enforces the behaviour of both contracts k1 and k2.
 Because the model is stateful and produces output actions, to make a step it is necessary to execute a single step of each of  the contracts k1 and k2 in sequence: first k1 then k2.
-\item[\texttt Choice obs k1 k2] behaves as either k1 or k2 depending on the (Boolean) result of obs at the time that the observation is made, k1 if it is True and k2 if False.
-\item[\texttt CommitValue idCV p vals k]  For this contract to make progress, the user p is required to make a value commitment with identifier idCV; the contract then continues as k.
+\item \bluett{Choice obs k1 k2} behaves as either k1 or k2 depending on the (Boolean) result of obs at the time that the observation is made, k1 if it is True and k2 if False.
+\item \bluett{CommitValue idCV p vals k}  For this contract to make progress, the user p is required to make a value commitment with identifier idCV; the contract then continues as k.
 In order for the commitment to be valid, the value must be one of the specified in the vals list.
 Otherwise the contract is quiescent, waiting for the value to be committed.
-\item[\texttt CommitCash idCC p n timeout k]  For this contract to make progress, the user p is required to make a cash commitment of value n and timeout timeout (specified as a block number) with identifier idCC; the contract then continues as k.
+\item \bluett{CommitCash idCC p n timeout k}  For this contract to make progress, the user p is required to make a cash commitment of value n and timeout timeout (specified as a block number) with identifier idCC; the contract then continues as k.
 Otherwise the contract is quiescent, waiting for the cash to be committed.
@@
 This is performed by the full\_step function, which wraps the step function.
-\item[\texttt When obs timeout k def\_k] This contract will not progress until obs is True or until the current block number is greater than or equal to the one specified by timeout.
+\item \bluett{When obs timeout k def\_k} This contract will not progress until obs is True or until the current block number is greater than or equal to the one specified by timeout.
 In case the timeout applies, the contract will continue as def\_k, if the timeout does not apply and obs is True, then the contract continues as k.
 Otherwise the contract is quiescent.
-\end{description}
+\end{itemize}
 

### 2018-02-28 tex/paper.tex
more on intro; beginning of section 2

@@
 \usepackage{alltt}
-%\usepackage[dvipsnames]{xcolor}
 
@@
   \institution{University of Kent}
-%  \streetaddress{}
+%  \streetaddress{School of Computing}
   \city{Canterbury}
   \state{Kent}
-  \postcode{CT2 7NZ}
+  \postcode{CT2 7NF}
   \country{UK}}
@@
 
-No introduction \cite{garfield1996cite}.
+\noindent
+\sjtnote{This is a preliminary version, return to this at the end of the writing process.}
 
-\section{Contract DSL model discussion, Version 1.1, September 2017}
-The aim of this document is to set out the assumptions behind the Haskell model of a contract DSL blockchain. Code for this model is in the repository \url{https://github.com/input-output-hk/scdsl}
+This paper explores the design of a domain specific language, Marlowe,\footnote{Marlowe is named after Christopher Marlowe, the Elizabethan poet, dramatist and spy, who was born and educated in Canterbury: \url{https://en.wikipedia.org/wiki/Christopher_Marlowe}}\footnote{The code for Marlowe is available from \url{https://github.com/input-output-hk/scdsl}
+} targetted at the execution of financial contracts in the style of Peyton Jones, Eber and Seward~\cite{PeytonJones:2000} on blockchains.
+This work is a part of the Cardano project, and is supported by IOHK.\footnote{\url{https://iohk.io}}
 
-The document is structured like this:
+In doing this we are required to refine the model of contracts in a number of ways in order to fit with a radically different context. 
+\begin{itemize}
+\item In the traditional model, enforcement of the contract is the responsibility of the legal system; in the decentralised blockchain model, in which there is no central authority, the contract needs to be enforced by design.
+\item As noted by  the authors of Findel~\cite{findel} this means in particular that payments need to be bounded; this is made concrete in our account by our notion of commitments.
+\item There is not a single blockchain model, and we give a brief overview of the different mechanisms for scripting on blockchain (although this is not the primary purpose of this article). 
+\begin{itemize}
+\item
+A particularly important technical distinction here is that between an account-style model as used by Ethereum~\cite{EthereumRationale}, and the unspent transaction output (UTxO) model of bitcoin~\cite{sok}.
+\item There is the question of how precisely the execution of the contract interacts with the blockchain, and how information is to be recorded on the chain.
+\item Does the execution ``push'' for changes  to happen, or do external actors have to ``pull'' to effect changes enabled by execution of the contract?
+\end{itemize}
+\item \sjtnote{What else do we want to say here?}
+\end{itemize}
 
-\begin{enumerate}
-\item An overview of the model, including the assumptions made in designing it, the types of the principal functions and a description of the DSL as an algebraic type, constructor by constructor.
+\noindent
+\sjtnote{Preliminary overview; need to revisit.}
+The paper is structured like this: we begin with an overview of the model, including the assumptions made in designing it, the types of the principal functions and a description of the DSL as an algebraic type, constructor by constructor.
 The section aims to give a high level overview, and some of the concrete details are to be found in the repo.
-\item A discussion of the kinds of formal analysis which might be made for the DSL.
-\item A set of questions raised by the process of modelling.
-\item A response to some particular points relevant to this work and raised in the email discussion on ``Some Thoughts on Smart Contracts'' initiated by Charles, and taken up by Duncan and Phil.
-\item Background: an overview of the Rock-Paper-Scissors example from the repo and the discussion.
-\item A response to Charles’ comments as posted to slack.
-\end{enumerate}
+We follow this with discussions of the kinds of formal analysis which might be made for the DSL, as well as discussing a  set of questions raised by the process of modelling.
+We then look at the example of BLAH, as an illustration, and then look at how Marlowe relates to other work in the area.
+We conclude by pointing out
 
-\subsection{The model}
+\section{The model}
 The DSL is modelled as an algebraic type in Haskell, together with an executable small-step semantics. 
@@
 
-Second, it will be necessary to ensure that commitments of various kinds have been made by the participants in a contract. We have included two kinds of commitments here:
-\begin{itemize}
-\item A commitment to a value: this would include a choice of a play in a game, which would be chosen and concealed from the other players until a later point in play, when it is revealed.
-
-\item A commitment to a certain amount of cash: while a commitment to a value is indefinite, a cash commitment needs to have a timeout so that progress can be forced in a game.
-After the timeout period the cash is automatically refunded; in practice this might require the user explicitly to create a transaction to reclaim the cash.
-\end{itemize}
+Second, it will be necessary to ensure that commitments of currency (or ``cash'') can be  made by the participants in a contract.
+While informally we might see a commitment to something as being indefinite, it is important to realise that on blockchain a commitment needs to have a timeout so that progress can be forced in a game.
+After the timeout period the cash can be refunded through the user creating a transaction to reclaim the cash.
 
@@
 
-Finally, a running contract will initiate a number of payment actions on the blockchain at each step, represented by a list of actions.
+\sjtnote{Still true? CHECK} Finally, a running contract will initiate a number of payment actions on the blockchain at each step, represented by a list of actions.
 
-This analysis justifies the type of the step function:
+This analysis justifies the type of the \bluett{step} function:
 
@@
 
-Where \bluett{OS} contains the current values for observables, and \bluett{AS} is a list of actions (currently payments) that are carried out as a result of executing the step.
+Where \bluett{OS} contains the values for observables available at the step in question, and \bluett{AS} is a list of actions (currently payments) that are carried out as a result of executing the step.
 
@@
 \begin{itemize}
-\item It is assumed that cryptographic ``services'' or operations are provided by a wrapper layer external to the system, and are not modelled explicitly here.
+\item It is assumed that cryptographic ``services'' or operations are provided by a layer external to the system, and so they need not be modelled explicitly here.
 \item Making a commitment is not something that a contract can perform; rather, it can require that a commitment is made, but that then has to be established externally: hence the input of (a set of) commitments at each step.
-The same applies to revealing a value commitment, and to manually redeeming a cash commitment.
+The same observation also applies to redeeming a commitment of cash.
 \item Observables are observed at a particular time and in a particular context.
 We assume that the recording and identification of each (instance of an) observation is provided by the infrastructure of the system.
-We provide a ``little language'' for describing observations: Observation\footnote{Why doesn't Observation appear in the type of step?
-I would expect a contract would need to yield a request for observations to be made, and accept the value of those observations.}\footnote{The current values of *observables* required at each step are the input OS to the step function.
-We assume that the infrastructure is designed to record these, but that is not part of the DSL itself.
-The *observations* over OS are part of the contract (e.g.
-the Choice constructor).}
+We provide a ``little language'' for describing observations: \bluett{Observation}.
 \item The fact of making commitments also needs to be recorded.
 \item It is important to be aware that there may be different instances of contracts executed, and for each instance there will typically be different instances of observables (and commitments) observed and recorded.
-\item The model manages the release of funds back to the committer when a cash commitment expires (see discussion of the full\_step function below).
+\item The model manages the release of funds back to the committer when a cash commitment expires (see discussion of the \bluett{full\_step} function below).
 \end{itemize}
 
+\noindent
+\sjtnote{Revised to here, 14:53 Wednesday 28 February.}
+
+\medskip
+\noindent
 The type of contracts is given by
@@
    Both Contract Contract |
-   Choice Observation\footnote{Doesn't the step function need to record what observations are made, so they can be archived in the blockchain?
-I don't see how your design will support doing so.}\footnote{Yes, we assume that it is part of the infrastructure of this language to record the values of *observables* that are used.
-However, we have chosen no to make it part of the language itself.} Contract Contract |
+   Choice Observation Contract Contract |
    CommitValue IdentCV Person [Value] Contract |
@@
 \end{allbluett}
+%Originally on "Choice": \footnote{Doesn't the step function need to record what observations are made, so they can be archived in the blockchain?
+I don't see how your design will support doing so.}\footnote{Yes, we assume that it is part of the infrastructure of this language to record the values of *observables* that are used.
+However, we have chosen no to make it part of the language itself.}
 
-The step function is total, so that for every contract a result of stepping is defined.
-However, for some kinds of contracts – commits, redeems or time-shifted contracts – it is possible that the result of a step is to produce the same contract as the result; such a step is called quiescent whereas all other steps make progress.
+The \bluett{step} function is total, so that for every contract a result of stepping is defined.
+However, for some kinds of contracts -- commits, redeems or time-shifted contracts -- it is possible that the result of a step is to produce the same contract as the result; such a step is called \emph{quiescent} whereas all other steps \emph{make progress}.
 We use this distinction in the explanation that follows.
 
-At each computational step, the step function is wrapped by a call to full\_step which in addition will cause expired cash commitments to be refunded.
+At each computational step, the \bluett{step} function is wrapped by a call to the \bluett{full\_step} function, which in addition to performing \bluett{step} will cause expired cash commitments to be refunded.
 
@@
 
-\subsection{Analysis}
+\subsection{Potential additions to the model}
+
+The model that we have discussed so far is deliberately kept as compact as possible. However there are a number of potential extensions to the model which we introduce here.
+
+We have examined a single kind of commitment, namely a commitment of a certain amount of currency.
+In different applications, such as on-chain game playing of games like poker, a commitment to a value will be required.
+A value commitment would include a choice of a play in a game, which would be chosen and concealed from the other players until a later point in play, when it can be revealed.
+\sjtnote{Add citation to the poker paper?}
+
+\section{Analysis}
 What questions can we ask? What answers can we give?
@@
 Not directly, but there need to be sufficient funds available to pay if needed, or enough time for the commitments to time out, if not.
-\item Originally we had included that we should include all the ``progress'' requirements from point 2 below; but that’s probably not the right thing here.
+\item Originally we had included that we should include all the ``progress'' requirements from point 2 below; but that's probably not the right thing here.
 \end{itemize}
@@
 \begin{itemize}
-\item For When, need to be able to make interpretObs st obs os true. Since there is a timeout to When then there’s no need to impose this condition, as we’ll progress at timeout if not before.
+\item For When, need to be able to make interpretObs st obs os true. Since there is a timeout to When then there's no need to impose this condition, as we'll progress at timeout if not before.
 \item For Commit*, need to have the right commits in the Commit argument.
@@
 
-\subsection{Questions}
+\section{Questions}
 
@@
 The type Time represents the index of evaluation, and BlockNumber is the number of the current block of the blockchain.
-We’ve not included a ``real'' time, but should this be included?
+We've not included a ``real'' time, but should this be included?
 \item (Related to 5) What should denominate the timeouts on commitments: Time or BlockNumber or another notion of time? Currently we are using the latter.
@@
 \footnote{Why is RedeemCC a separate construct? Can't it be implemented with Pay?}\footnote{It is for someone to claim their cash back, rather than making a payment to another person.}
- This is potentially dangerous – leading to a FailedPay, for example – but useful to make a repayment at the end of a contract. 
+ This is potentially dangerous Ð leading to a FailedPay, for example Ð but useful to make a repayment at the end of a contract. 
 \item Alternatively, it would be possible for the underlying model to repay all committed but unspent cash automatically at the end of a contract, but we might want to do the same thing at the end of each ``round'' of  a game, for instance.
@@
 
-\subsection{Response to ``Some Thoughts on Smart Contracts''}
+\section{Response to ``Some Thoughts on Smart Contracts''}
 
@@
 
-So far we are assuming contracts defined in the DSL to be finite in both aspects. They could be infinite if we were to define them – co-inductively – as recursive members of the DSL type.
+So far we are assuming contracts defined in the DSL to be finite in both aspects. They could be infinite if we were to define them -- co-inductively -- as recursive members of the DSL type.
 
-\subsection{Rock-paper-scissors explanation (not re-formatted)}
+\section{Related work}
+
+\begin{itemize}
+\item Original SPJ paper~\cite{PeytonJones:2000}, and multi-party successor. Original Szabo paper?
+\item Findel~\cite{findel}
+\end{itemize}
+
+\section{Rock-paper-scissors explanation (not re-formatted)}
+
 In this section, we explain how the DSL can be used for designing a contract that enforces a round of rock-paper-scissors where the winner will receive from the loser 10 units of currency and not revealing your hand also has a penalty of 10 units.
@@
 Concerning bad data, a commitment will usually just store the hash, but the problem seems to come at the reveal, when all the data has to be shown, somehow.
-We don’t have to store the whole data (just the hash) but it does have to be revealed in some way that can be observed, and the naive interpretation of this is that it is made available somehow, thereby using storage on the blockchain.
+We don't have to store the whole data (just the hash) but it does have to be revealed in some way that can be observed, and the naive interpretation of this is that it is made available somehow, thereby using storage on the blockchain.
 

### 2018-02-28 src/Semantics.hs
Needs further comments / restructuring according to QUESTIONs

@@ -26,9 +26,14 @@ import qualified Data.List as List
  -- Basic data types --
  ----------------------} 
 
-type Key = Int -- Public key
+ -- People are represented by their public keys,
+ -- which in turn are given by integers.
 
+type Key         = Int   -- Public key
 type Person      = Key
+
+-- Block numbers and random numbers are both integers.
+ 
 type Random      = Int
 type BlockNumber = Int
 
@@ -39,8 +44,16 @@ type BlockNumber = Int
 --  - We assume that some mechanism exists which will ensure that the value is looked up and recorded, or …
 --  - … we actually provide that mechanism explicitly, e.g. with inter-contract comms or transaction generation or something.
 
+-- Other observables are possible, e.g. the value of the oil price at this time.
+-- It is assumed that these would be provided in some agreed way by an oracle of some sort.
+
+-- The Observable data type represents the different sorts of observables, …
+
 data Observable = Random | BlockNumber
                     deriving (Eq,Ord,Show,Read)
+
+-- … while the type OS gives a particular random value and block number. 
+-- Can think of these as the values available at a particular step.
 
 data OS =  OS { random       :: Random,
                 blockNumber  :: BlockNumber }
@@ -48,14 +61,22 @@ data OS =  OS { random       :: Random,
 
 -- Inputs
 -- Types for cash commits, money redeems, and choices.
+--
+-- A cash commitment is an integer (should be positive integer?)
+-- Concrete values are sometimes chosen too: these are integers for the sake of this model.
 
-type Value    = Int
- 
 type Cash     = Int
-
 type ConcreteChoice = Int
 
+-- We need to put timeouts on various operations. These could be some abstract time
+-- domain, but it only really makes sense for these to be block numbers.
+
 type Timeout = BlockNumber
+
+-- Commitments, choices and payments are all identified by identifiers.
+-- Their types are given here. In a more sophisticated model these would
+-- be generated automatically (and so uniquely); here we simply assume that 
+-- they are unique.
 
 newtype IdentCC = IdentCC Int
                deriving (Eq,Ord,Show,Read)
@@ -66,10 +87,23 @@ newtype IdentChoice = IdentChoice Int
 newtype IdentPay = IdentPay Int
                deriving (Eq,Ord,Show,Read)
 
+-- A cash commitment is made by a person, for a particular amount and timeout.               
+
[…]

### 2018-02-28 tex/paper.tex
progress in writing, but needs clarification in Semantics.hs before do more

@@
 \begin{allbluett}
-step :: Commits -> State -> Contract -> OS -> (State,Contract,AS)
+step :: Input -> State -> Contract -> OS -> (State,Contract,AS)
 \end{allbluett}
@@
    RedeemCC IdentCC Contract |
-   RevealCV IdentCV Contract |
-   Pay Person Person Int Contract |
+   Pay IdentPay Person Person Cash Timeout Contract |   -- CHANGED
    Both Contract Contract |
    Choice Observation Contract Contract |
-   CommitValue IdentCV Person [Value] Contract |
-   CommitCash IdentCC Person Cash Time Contract |
-   When Observation Time Contract Contract
+   CommitCash IdentCC Person Cash Timeout Timeout Contract |   -- CHANGED
+   When Observation Timeout Contract Contract    -- CHANGED
 \end{allbluett}
@@
 Otherwise the contract is quiescent, waiting for the commitment to be redeemed or to expire.
-\item  \bluett{RevealCV idCV k} (\bluett{CV} stands for value commitment.) For this contract to make progress, the creator of the value commitment with identifier \bluett{idCV} is required to reveal the value; the contract then continues as \bluett{k}.
-Otherwise the contract is quiescent, waiting for the value to be revealed.
-For the reveal to be valid, the value revealed must be in the vals list provided in the corresponding CommitValue constructor (the one with the same idCV). 
 \item \bluett{Pay p1 p2 int k} requires that int is paid by p1 to p2, and continues as k.
@@
 \item \bluett{Choice obs k1 k2} behaves as either k1 or k2 depending on the (Boolean) result of obs at the time that the observation is made, k1 if it is True and k2 if False.
-\item \bluett{CommitValue idCV p vals k}  For this contract to make progress, the user p is required to make a value commitment with identifier idCV; the contract then continues as k.
-In order for the commitment to be valid, the value must be one of the specified in the vals list.
-Otherwise the contract is quiescent, waiting for the value to be committed.
 \item \bluett{CommitCash idCC p n timeout k}  For this contract to make progress, the user p is required to make a cash commitment of value n and timeout timeout (specified as a block number) with identifier idCC; the contract then continues as k.

### 2018-03-01 tex/paper.tex
Answer some questions and add citation

@@
 
-\sjtnote{Still true? CHECK} Finally, a running contract will initiate a number of payment actions on the blockchain at each step, represented by a list of actions.
+Finally, payments can be granted by using committed money, but they must be manually redeemed by the recipient, in the 
+same way that cash commitments are redeemed when they expire. The effects of the contract in the blockchain are 
+represented by a list of actions that is derived from the execution of each step of the semantics.
 
@@
 
-We have examined a single kind of commitment, namely a commitment of a certain amount of currency.
-In different applications, such as on-chain game playing of games like poker, a commitment to a value will be required.
-A value commitment would include a choice of a play in a game, which would be chosen and concealed from the other players until a later point in play, when it can be revealed.
-\sjtnote{Add citation to the poker paper?}
+We have examined a single kind of commitment, namely a commitment of a certain amount of currency. In different 
+applications, such as on-chain game playing of games like poker, a commitment to a value will be required. A value 
+commitment would include a choice of a play in a game, which would be chosen and concealed from the other players until 
+a later point in play, when it can be revealed. The paper \cite{kumaresan2015use} makes use of this and other 
+cryptographic primitives to enforce the rules of poker while maintaining the cards of each player secret.
 

### 2018-03-01 src/Semantics.hs
Answer some quesitons and reorganise

@@ -16,7 +16,7 @@ import qualified Data.List as List
  --     - type of State (of the internal state of a DSL contract evaluation)                    --
  --     - type of Observations (of values of Observables)                                       --
  --     - type of Contracts                                                                     --
- --     - single step evaluation (the step function) which is wrapped by fullStep …            --
+ --     - single step evaluation (the step function) which is wrapped by fullStep …             --
  --     - … which expires and refunds cash commitments                                          --
  --                                                                                             --
  -- Further discussion in accompanying document.                                                --
@@ -100,10 +100,19 @@ data RC = RC IdentCC Person Cash
 -- The input to a step has four components
 --      - a set of cash commitments made at that step
 --      - a set of cash redemptions made at that step
---      - a collection of payments: each payment may pay to multiple people, 
---        so represented by a map (IdentPay, Person) |-> Cash
---      - a collection of choices QUESTION: WHAT IS THE RATIONALE HERE???
---        WHY ISN'T IT A MAP FROM THE IDENT TO A PERSON,CHOICE PAIR??
+--      - a map of payment requests made at that step
+--      - a map of choices made at that step
+--
+-- In practice, we could use sets for all of them,
+-- but using a map ensures that there is only one
+-- entry per identifier and person pair and would
+-- make access more efficient if we needed to find
+-- an entry without knowing the concrete choice
+-- or ammount of cash being claimed.
+
+-- If we want to be able to accept commitments that are
+-- more generous than established we would need to make
+-- cc a map too.
 
 data Input = Input {
                 cc  :: Set.Set CC,
@@ -116,11 +125,18 @@ data Input = Input {
 -- Actions are things that have an effect on the blockchain, and a set
 -- of actions is generated at each step. We are not responsible for
 -- making these happen: this is passed to the blockchain.
--- QUESTION: the rest of this comment should be deleted?
--- The current implementation doesn't properly reflect this.
+--
+-- There are some actions that would not have an effect on
+-- the blockchain but serve the purpouse of logging what happened
[…]

### 2018-03-05 src/Semantics.hs
renamed reduceMoney

@@ -185,13 +185,13 @@ data Money = AvailableMoney IdentCC |
 -- Function that evaluates a Money construct given
 -- the current state. 
 
-reduceMoney :: State -> Money -> Cash
-reduceMoney (State {sc = scv}) (AvailableMoney id) =
+evalMoney :: State -> Money -> Cash
+evalMoney (State {sc = scv}) (AvailableMoney id) =
   case Map.lookup id scv of
     Just (_, NotRedeemed c _) -> c
     _ -> 0
-reduceMoney s (AddMoney a b) = (reduceMoney s a) + (reduceMoney s b)
-reduceMoney _ (ConstMoney c) = c
+evalMoney s (AddMoney a b) = (evalMoney s a) + (evalMoney s b)
+evalMoney _ (ConstMoney c) = c
 
 
 -- Representation of observations over observables and the state.
@@ -229,7 +229,7 @@ interpretObs st (PersonChoseThis choice_id person reference_choice) _
 interpretObs st (PersonChoseSomething choice_id person) _
     = Map.member (choice_id, person) (sch st)
 interpretObs st (ValueGE a b) _
-    = (reduceMoney st a) >= (reduceMoney st b)
+    = (evalMoney st a) >= (evalMoney st b)
 interpretObs _ TrueObs _
     = True
 interpretObs _ FalseObs _

### 2018-03-05 src/Semantics.hs
Remove actions when CommitCash expires before committing

@@ -300,8 +300,7 @@ step _ st (When obs expi con con2) os
 -- May want to relax this
 
 step commits st c@(CommitCash ident person val start_timeout end_timeout con) os
-  | cexe || cexs = (st {sc = ust}, con, [SuccessfulCommit ident person 0,
-                                         CommitRedeemed ident person 0])
+  | cexe || cexs = (st {sc = ust}, con, [])
   | Set.member (CC ident person val end_timeout) (cc commits)
         = (st {sc = ust}, con, [SuccessfulCommit ident person val])
   | otherwise = (st, c, [])

### 2018-03-05 tex/paper.tex
Changed DSL explanation etc.

@@
 \usepackage{alltt}
+%\usepackage[cache=false]{minted}
+\usepackage{minted}
+\usepackage{syntax}
 
@@
 \FXRegisterAuthor{sjt}{asjt}{\color{magenta}[Simon]}
+
+%%% MINTED COMMANDS FOR HASKELL %%%
+
+\newmint{haskell}{}
+\newminted{haskell}{}
+\newmintinline{haskell}{}
 
@@
 Each instance of such an observable will be recorded on the blockchain to allow the computation to be repeated for verification purposes.
-It is assumed that at each step of the execution of the contract, the values of some observables will be needed.
+It is assumed that at each step of the execution of the contract, the values of some observables will be needed, and these values are (together) given by a value of type \haskellinline{OS}.
+Note that these values are not determined by the participants in the contract, but rather by the external environment.
 
-Second, it will be necessary to ensure that commitments of currency (or ``cash'') can be  made by the participants in a contract.
+On the other hand, at each step there are -- potentially, at least -- a variety of inputs available from the participants.
+These include commitments of currency (or ``cash''), redemption of commitments, and claims of payments by a participant; moreover it is also possible for a participant to input an arbitrary value (which we term a ``choice'').
+The particular inputs at a given step are described by a value of type \haskellinline{Input}.
+
 While informally we might see a commitment to something as being indefinite, it is important to realise that on blockchain a commitment needs to have a timeout so that progress can be forced in a game.
@@
 
-At each step, a number of commitments may be made.
-This information constitutes an input to the execution of each step.
-Moreover, information about the commitments currently in force forms the state, which is (potentially) modified at each execution step.
+Moreover, information about the commitments currently in force forms the \haskellinline{State}, which is (potentially) modified at each execution step.
 
@@
 same way that cash commitments are redeemed when they expire. The effects of the contract in the blockchain are 
-represented by a list of actions that is derived from the execution of each step of the semantics.
+represented by a list \haskellinline{AS} of \haskellinline{Action}s that is derived from the execution of each step of the semantics.
 
-This analysis justifies the type of the \bluett{step} function:
+This analysis justifies the type of the \haskellinline{step} function:
 
-\begin{allbluett}
+\begin{haskellcode}
 step :: Input -> State -> Contract -> OS -> (State,Contract,AS)
-\end{allbluett}
-
-Where \bluett{OS} contains the values for observables available at the step in question, and \bluett{AS} is a list of actions (currently payments) that are carried out as a result of executing the step.
+\end{haskellcode}
 
@@
 We assume that the recording and identification of each (instance of an) observation is provided by the infrastructure of the system.
-We provide a ``little language'' for describing observations: \bluett{Observation}.
+We provide a ``little language'' for describing observations: \haskellinline{Observation}.
 \item The fact of making commitments also needs to be recorded.
 \item It is important to be aware that there may be different instances of contracts executed, and for each instance there will typically be different instances of observables (and commitments) observed and recorded.
-\item The model manages the release of funds back to the committer when a cash commitment expires (see discussion of the \bluett{full\_step} function below).
+\item The model manages the release of funds back to the committer when a cash commitment expires (see discussion of the \haskellinline{fullStep} function below).
+\item We assume that time is ``coarse grained'' and measured simply by block number, so that timeouts etc.\ are delimited using nothing finer than block number. 
 \end{itemize}
-
-\noindent
-\sjtnote{Revised to here, 14:53 Wednesday 28 February.}
 
@@
 
-\begin{allbluett}
+\begin{haskellcode}
 data Contract =
    Null |
+   CommitCash IdentCC Person Cash Timeout Timeout Contract |  
    RedeemCC IdentCC Contract |
-   Pay IdentPay Person Person Cash Timeout Contract |   -- CHANGED
+   Pay IdentPay Person Person Cash Timeout Contract |  
    Both Contract Contract |
    Choice Observation Contract Contract |
-   CommitCash IdentCC Person Cash Timeout Timeout Contract |   -- CHANGED
-   When Observation Timeout Contract Contract    -- CHANGED
-\end{allbluett}
+   When Observation Timeout Contract Contract   
+   \end{haskellcode}
 %Originally on "Choice": \footnote{Doesn't the step function need to record what observations are made, so they can be archived in the blockchain?
@@
 
-The \bluett{step} function is total, so that for every contract a result of stepping is defined.
+The \haskellinline{step} function is total, so that for every contract a result of stepping is defined.
 However, for some kinds of contracts -- commits, redeems or time-shifted contracts -- it is possible that the result of a step is to produce the same contract as the result; such a step is called \emph{quiescent} whereas all other steps \emph{make progress}.
@@
 
-At each computational step, the \bluett{step} function is wrapped by a call to the \bluett{full\_step} function, which in addition to performing \bluett{step} will cause expired cash commitments to be refunded.
+At each computational step, the \haskellinline{step} function is wrapped by a call to the \haskellinline{fullStep} function, which in addition to performing \haskellinline{step} will enable expired cash commitments to be refunded and record any choices made at that step.
 
@@
 \begin{itemize}
-\item \bluett{Null} is the null contract; it will always be quiescent.
-\item \bluett{RedeemCC idCC k} (\bluett{CC} stands for cash commitment.) For this contract to make progress, the creator of the cash commitment with identifier \bluett{idCC} is required to redeem the unspent funds in that commitment; the contract then continues as \bluett{k}.
-If the cash commitment with identifier \bluett{idCC} has expired, the remaining funds will be redeemed automatically by the \bluett{full\_step} function and a \bluett{ExpiredRedeem }action will be produced\footnote{How does the contract determine what actions resulted from a given step?
-Or is this not required?}\footnote{
-If it is required, restructuring the DSL so that the result action appears as the 'value' of executing a step (using a monadic interface as described for id generation below) would be a good idea.}\footnote{We specify the actions generated by each kind of constructor; at each step the list of such actions, of type \bluett{AS}, is a component of the result.}.
-Using \bluett{RedeemCC} to redeem the same identifier \bluett{idCC} twice will result in the issue of a DuplicateRedeem action.
-Otherwise the contract is quiescent, waiting for the commitment to be redeemed or to expire.
-\item \bluett{Pay p1 p2 int k} requires that int is paid by p1 to p2, and continues as k.
-It uses the currency allocated by the cash commitments made by p1 that expire the earliest.
-This contract will result in a FailedPay action if the funds are not available; otherwise a SuccessfulPay action is generated.
-\item \bluett{Both k1 k2} enforces the behaviour of both contracts k1 and k2.
-Because the model is stateful and produces output actions, to make a step it is necessary to execute a single step of each of  the contracts k1 and k2 in sequence: first k1 then k2.
-\item \bluett{Choice obs k1 k2} behaves as either k1 or k2 depending on the (Boolean) result of obs at the time that the observation is made, k1 if it is True and k2 if False.
-\item \bluett{CommitCash idCC p n timeout k}  For this contract to make progress, the user p is required to make a cash commitment of value n and timeout timeout (specified as a block number) with identifier idCC; the contract then continues as k.
-Otherwise the contract is quiescent, waiting for the cash to be committed.
-When the timeout is reached, the remaining funds will be redeemed automatically and a SuccessfulPay action will be issued that indicates a self-payment from the person p with the remaining funds.
-This is performed by the full\_step function, which wraps the step function.
-\item \bluett{When obs timeout k def\_k} This contract will not progress until obs is True or until the current block number is greater than or equal to the one specified by timeout.
-In case the timeout applies, the contract will continue as def\_k, if the timeout does not apply and obs is True, then the contract continues as k.
+
+\item \haskellinline{Null} is the null contract; it will always be quiescent.
+
+\item \haskellinline{CommitCash idCC p n t1 t2 k}  For this contract to make progress,\begin{itemize}
+\item either before the timeout \haskellinline{t1} the user \haskellinline{p} makes a cash commitment of value \haskellinline{n} and timeout \haskellinline{t2}  with the identifier \haskellinline{idCC} ,
+\item
+or the timeout \haskellinline{t1} is exceed.
+\end{itemize} 
+In the first case, a \haskellinline{SuccessfulCommit} action is generated; in both cases, the contract then continues as \haskellinline{k}.
+While neither case holds the contract is quiescent, waiting for the cash to be committed.
+
+If the cash is committed successfully and the timeout \haskellinline{t2} is reached, then it is impossible to further spend the committed cash, and any unspent funds can be reclaimed by participant \haskellinline{p}.
+ This is performed by the \haskellinline{fullStep} function, as noted above.
+
+\item \haskellinline{RedeemCC idCC k} (\haskellinline{CC} stands for cash commitment.) For this contract to make progress, the creator of the cash commitment with identifier \haskellinline{idCC} is required to redeem the unspent funds in that commitment; the contract then continues as \haskellinline{k}, and the action \haskellinline{CommitRedeemed} is produced.
+Committed cash can only be redeemed once, and an attempt to redeem it a second time will produce a  \haskellinline{DuplicateRedeem} action, continuing as \haskellinline{k}.
+
+If the cash commitment with identifier \haskellinline{idCC} has expired, it becomes possible for the remaining funds to be redeemed by the committer; this can be done by the \haskellinline{fullStep} function processing the appropriate \haskellinline{Input}, and an \haskellinline{ExpiredCommitRedeemed} action will be produced.
+
+Once the commitment \haskellinline{idCC} has expired and is redeemed, a \haskellinline{RedeemCC idCC k} contract will immediately evolve to \haskellinline{k}.
+%\footnote{How does the contract determine what actions resulted from a given step? Or is this not required?}
+%\footnote{If it is required, restructuring the DSL so that the result action appears as the 'value' of executing a step (using a monadic interface as described for id generation below) would be a good idea.}
+%\footnote{We specify the actions generated by each kind of constructor; at each step the list of such actions, of type \haskellinline{AS}, is a component of the result.}.
+
+\item \haskellinline{Pay p1 p2 int t k} makes it possible, assuming that sufficient funds are available, for \haskellinline{p1} to claim  a payment of \haskellinline{int} from \haskellinline{p2} before the timeout \haskellinline{t}.
+The contract continues as \haskellinline{k}.
+By `available' we mean that sufficient commitments have been made and  not yet expired to cover the payment; in this case the payment  uses the currency allocated by the cash commitments made by \haskellinline{p1} that expire the earliest.
+
+ This contract will result in a \haskellinline{FailedPay} action if the funds are not available; otherwise a \haskellinline{SuccessfulPay} action is generated.
+
+\item \haskellinline{Both k1 k2} enforces the behaviour of both contracts \haskellinline{k1} and \haskellinline{k2}.
+Because the model is stateful and produces output actions, to make a step it is necessary to execute a single step of each of  the contracts \haskellinline{k1} and \haskellinline{k2} in sequence: first \haskellinline{k1} then \haskellinline{k2}.
+
+\item \haskellinline{Choice obs k1 k2} behaves as either \haskellinline{k1} or \haskellinline{k2} depending on the (Boolean) result of \haskellinline{obs} at the time that the observation is made, \haskellinline{k1} if it is \haskellinline{True} and \haskellinline{k2} if \haskellinline{False}.
+
+\item \haskellinline{When obs timeout k def_k} This contract will not progress until \haskellinline{obs} is \haskellinline{True} or until the current block number is greater than or equal to the one specified by timeout.
+In case the timeout applies, the contract will continue as \haskellinline{def_k}, if the timeout does not apply and \haskellinline{obs} is \haskellinline{True}, then the contract continues as \haskellinline{k}.
 Otherwise the contract is quiescent.
@@
 
+We should monadify, so that identifiers are automatically generated.
+This gives two nice properties.
+First, they are guaranteed to be unique; secondly they can be generated only in cases where a commit has been made successfully (within the required time period).
+It might well be that these should be generated by the system; this could guarantee uniqueness, freshness etc, and could also lead to a monadic DSL, with bind used for binding names to commitments.
+
 \section{Analysis}
@@
 
-One possible definition is ``those contracts that do not lead to a FailedPay action''.
+One possible definition is ``those contracts that do not lead to a \haskellinline{FailedPay} action''.
 Does this make sense independently of the inputs to a contract (i.e.
-observables, commitments, redeems, time, and randomness)?
+observables, commitments, redeems, timeouts, and randomness)?
 Some relevant points (NB discussion incomplete):
 \begin{itemize}
-\item For Pay require a sufficient balance not to produce a FailedPay action. 
-\item For Both require a sufficient balance for both contracts combined.
+\item For \haskellinline{Pay} require a sufficient balance not to produce a \haskellinline{FailedPay} or a \haskellinline{DuplicateRedeem} action. 
+\item For \haskellinline{Both} require a sufficient balance for both contracts combined.
 Need to think here whether the two conjuncts should be independent, or whether a commitment in one half can be used in the other.
-\item For Choice require a sufficient balance for the maximum of the requirements of the individual contracts. 
-\item For When, need to ``age'' the existing commitments relative to the new starting point of the continuation contract. 
-\item Are there any constraints on  Commit*, RevealCV and RedeemCC statements?
+\item For \haskellinline{Choice} require a sufficient balance for the maximum of the requirements of the individual contracts. 
+\item For \haskellinline{When}, need to ``age'' the existing commitments relative to the new starting point of the continuation contract. 
+\item Are there any constraints on  \haskellinline{CommitCash} and \haskellinline{RedeemCC} statements?
 Not directly, but there need to be sufficient funds available to pay if needed, or enough time for the commitments to time out, if not.
 \item Originally we had included that we should include all the ``progress'' requirements from point 2 below; but that's probably not the right thing here.
+
 \end{itemize}
 
-\subsubsection{Do all contracts potentially evaluate to Null?}
+\subsubsection{Do all contracts potentially evaluate to \haskellinline{Null}?}
 
@@
 \begin{itemize}
-\item For When, need to be able to make interpretObs st obs os true. Since there is a timeout to When then there's no need to impose this condition, as we'll progress at timeout if not before.
-\item For Commit*, need to have the right commits in the Commit argument.
-\item Similarly for RevealCV and RedeemCC statements.
-\item It would also be possible to put timeouts on to Commit, RevealCV and RedeemCC operations, and if this were to be done there would be no remaining requirements for contract execution to make progress.
+\item For \haskellinline{When}, need to be able to make \haskellinline{interpretObs st obs os} true.
+Since there is a timeout to \haskellinline{When} then there's no need to impose this condition, as we'll progress at timeout if not before.
+\item For \haskellinline{CommitCash}, need to have the right commits in the \haskellinline{Commit} argument.
+\item Similarly for \haskellinline{RedeemCC} statements.
+\item It would also be possible to put timeouts on to \haskellinline{CommitCash},  and \haskellinline{RedeemCC} operations, and if this were to be done there would be no remaining requirements for contract execution to make progress.
 \item There are no constraints on other constructs.
@@
 \begin{enumerate}
-\item Identifiers for commitments are part of the model, and supplied as part of each contract.
-It might well be that these should be generated by the system; this could guarantee uniqueness, freshness etc, and could also lead to a monadic DSL, with bind used for binding names to commitments.
-\item Does the Both constructor potentially produce invalid/misleading contracts, combining a commitment in one half with its use in the other?
+\item Does the \haskellinline{Both} constructor potentially produce invalid/misleading contracts, combining a commitment in one half with its use in the other?
 The problem is that the ``use'' conjunct could potentially be executed, leading to a failed payment.
 \item How do we ensure that contracts can be enforced? Is there a notion of validity for contracts that can be enforced? To which extent is this notion enforced by current constructors?
-\item Should there be a timeout on the process of making or revealing a commitment, or is it sufficient to have timeouts on the commitments themselves?
-\item (Related to 3) Is this also an argument to have timeouts on value commitments too?
-\item The model takes time as an observable.
-The type Time represents the index of evaluation, and BlockNumber is the number of the current block of the blockchain.
-We've not included a ``real'' time, but should this be included?
-\item (Related to 5) What should denominate the timeouts on commitments: Time or BlockNumber or another notion of time? Currently we are using the latter.
+\item The model does not take time as an observable, but rather it is defined in terms of \haskellinline{BlockNumber} only. 
+\item We could use ``slot time'' https://eprint.iacr.org/2016/889.pdf ``We consider a setting where time is divided into discrete units called slots.
+A ledger, described in more detail below, associates with each time slot (at most) one ledger block.''
 \item Is it possible for block number to depend on progress of computation?
@@
 Would there be any reason for using a different approach?
-\item Timeouts are expressed in absolute time: presumably they could be relative (with recording on the blockchain of the times at which they are triggered).\footnote{This contradicts what you've said elsewhere.
+Yes!
+UTxO.
+\item Timeouts are expressed in absolute time: presumably they could be relative (with recording on the blockchain of the times at which they are triggered).
+%\footnote{This contradicts what you've said elsewhere.
 I think timeouts should be times, not blockchain numbers.
@@
 Also, it avoids question 8.}\footnote{See addition about slot time from Ouroboros paper (as referenced in the body of the paper).}
-\end{enumerate}
-
-Added: we should probably use ``slot time'' https://eprint.iacr.org/2016/889.pdf ``We consider a setting where time is divided into discrete units called slots.
-A ledger, described in more detail below, associates with each time slot (at most) one ledger block.''
-\begin{enumerate}
-\item Some observables have a ``monotonicity'' property: once something is revealed, it is revealed forever; on the other hand, something unrevealed is only unrevealed so far; do we need to reflect this in the observations that we allow?
-\item The seesaw construction is key to implementing e.g.
+\item Some observables have a ``monotonicity'' property: once a choice is made, it is fixed forever; on the other hand, something not yet chosen is only so thus far; do we need to reflect this in the observations that we allow?
+%\item The seesaw construction is key to implementing e.g.
 poker, and it can be described in terms of the existing constructs (modulo the answer to point 4 above).
-However, should seesaw be a primitive in its own right?\footnote{No.
+However, should seesaw be a primitive in its own right?
+%\footnote{No.
 You will have a successful language when you can define seesaw as an abstraction in terms of the other primitives *within the language*}\footnote{And indeed we *did* define it, at least for the RPS example.}
@@
 \item Constructs that can time out could have a single continuation (do this next) or two continuations: do this if progressed normally, and do the other continuation if timed out.
-A candidate for this is a timed out When, where the second option is what we have implemented.
-\item RedeemCC is intended for premature repayment of a cash commitment, i.e. before the commitment has timed out.
-\footnote{Why is RedeemCC a separate construct? Can't it be implemented with Pay?}\footnote{It is for someone to claim their cash back, rather than making a payment to another person.}
- This is potentially dangerous Ð leading to a FailedPay, for example Ð but useful to make a repayment at the end of a contract. 
+A candidate for this is a timed out \haskellinline{When}, where the second option is what we have implemented.
+\item \haskellinline{RedeemCC} is intended for premature repayment of a cash commitment, i.e.
+before the commitment has timed out.
+Why is \haskellinline{RedeemCC} a separate construct?
+Can't it be implemented with \haskellinline{Pay}?%\footnote{It is for someone to claim their cash back, rather than making a payment to another person.}
+ This is potentially dangerous -- leading to a \haskellinline{FailedPay}, for example -- but useful to make a repayment at the end of a contract. 
 \item Alternatively, it would be possible for the underlying model to repay all committed but unspent cash automatically at the end of a contract, but we might want to do the same thing at the end of each ``round'' of  a game, for instance.

### 2018-03-05 src/Semantics.hs
changes

@@ -239,11 +239,11 @@ interpretObs _ FalseObs _
 
 data Contract =
     Null |
+    CommitCash IdentCC Person Cash Timeout Timeout Contract |
     RedeemCC IdentCC Contract |
     Pay IdentPay Person Person Cash Timeout Contract |
     Both Contract Contract |
     Choice Observation Contract Contract |
-    CommitCash IdentCC Person Cash Timeout Timeout Contract |
     When Observation Timeout Contract Contract
                deriving (Eq,Ord,Show,Read)
 

### 2018-03-06 tex/paper.tex
Add timeout alternative continuation to CommitCash

@@
    Null |
-   CommitCash IdentCC Person Cash Timeout Timeout Contract |  
+   CommitCash IdentCC Person Cash Timeout Timeout Contract Contract |  
    RedeemCC IdentCC Contract |
@@
 
-\item \haskellinline{CommitCash idCC p n t1 t2 k}  For this contract to make progress,\begin{itemize}
+\item \haskellinline{CommitCash idCC p n t1 t2 k1 k2}  For this contract to make progress,\begin{itemize}
 \item either before the timeout \haskellinline{t1} the user \haskellinline{p} makes a cash commitment of value \haskellinline{n} and timeout \haskellinline{t2}  with the identifier \haskellinline{idCC} ,
@@
 \end{itemize} 
-In the first case, a \haskellinline{SuccessfulCommit} action is generated; in both cases, the contract then continues as \haskellinline{k}.
-While neither case holds the contract is quiescent, waiting for the cash to be committed.
+In the first case, a \haskellinline{SuccessfulCommit} action is generated and the contract continues as 
+\haskellinline{k1}; in the second case no action is generated and the contract continues as 
+\haskellinline{k2}. While neither case holds the contract is quiescent, waiting for the cash to be committed.
 

### 2018-03-06 blockly/blocks/stubs.js

@@ -58,6 +58,14 @@ Blockly.JavaScript['observation_personchosethis'] = function(block) {
   return [code, Blockly.JavaScript.ORDER_NONE];
 };
 
+Blockly.JavaScript['observation_timeabove'] = function(block) {
+  var text_block_number = block.getFieldValue('block_number');
+  // TODO: Assemble JavaScript into code variable.
+  var code = '...';
+  // TODO: Change ORDER_NONE to the correct strength.
+  return [code, Blockly.JavaScript.ORDER_NONE];
+};
+
 Blockly.JavaScript['observation_trueobs'] = function(block) {
   // TODO: Assemble JavaScript into code variable.
   var code = '...';
@@ -66,7 +74,7 @@ Blockly.JavaScript['observation_trueobs'] = function(block) {
 };
 
 Blockly.JavaScript['base_contract'] = function(block) {
-  var statements_name = Blockly.JavaScript.statementToCode(block, 'NAME');
+  var statements_base_contract = Blockly.JavaScript.statementToCode(block, 'base_contract');
   // TODO: Assemble JavaScript into code variable.
   var code = '...;\n';
   return code;
@@ -74,9 +82,8 @@ Blockly.JavaScript['base_contract'] = function(block) {
 
 Blockly.JavaScript['contract_null'] = function(block) {
   // TODO: Assemble JavaScript into code variable.
-  var code = '...';
-  // TODO: Change ORDER_NONE to the correct strength.
-  return [code, Blockly.JavaScript.ORDER_NONE];
+  var code = '...;\n';
+  return code;
 };
 
 Blockly.JavaScript['contract_redeemcc'] = function(block) {
@@ -120,9 +127,10 @@ Blockly.JavaScript['contract_commitcash'] = function(block) {
   var number_ccommit_id = block.getFieldValue('ccommit_id');
   var number_person_id = block.getFieldValue('person_id');
   var number_ammount = block.getFieldValue('ammount');
-  var number_start_expiration = block.getFieldValue('start_expiration');
   var number_end_expiration = block.getFieldValue('end_expiration');
-  var statements_contract = Blockly.JavaScript.statementToCode(block, 'contract');
+  var number_start_expiration = block.getFieldValue('start_expiration');
+  var statements_contract1 = Blockly.JavaScript.statementToCode(block, 'contract1');
[…]

### 2018-03-06 blockly/src/Main.hs

@@ -242,15 +242,17 @@ block_to_blockly (Choice obs c1 c2) =
      connect_as_input "contract1" tb cb1
      connect_as_input "contract2" tb cb2
      return tb
-block_to_blockly (CommitCash (IdentCC id) p cash b1 b2 c) =
-  do cb <- block_to_blockly c
+block_to_blockly (CommitCash (IdentCC id) p cash b1 b2 c1 c2) =
+  do cb1 <- block_to_blockly c1
+     cb2 <- block_to_blockly c2
      tb <- create_block "contract_commitcash"
      set_field_value tb "ccommit_id" id
      set_field_value tb "person_id" p
      set_field_value tb "ammount" cash
      set_field_value tb "start_expiration" b1
      set_field_value tb "end_expiration" b2
-     connect_as_input "contract" tb cb
+     connect_as_input "contract1" tb cb1
+     connect_as_input "contract2" tb cb2
      return tb
 block_to_blockly (When obs b c1 c2) =
   do obb <- obs_to_blockly obs

### 2018-03-06 src/Analysis.hs

@@ -258,7 +258,7 @@ findChoice (RedeemCC _ con) = findChoice con
 findChoice (Pay _ _ _ _ _ con) = findChoice con
 findChoice (Both con1 con2) = minimum (map (findChoice) [con1, con2])
 findChoice (Choice obs con1 con2) = minimum ((findChoiceObs obs):(map (findChoice) [con1, con2]))
-findChoice (CommitCash _ _ _ _ _ con) = findChoice con
+findChoice (CommitCash _ _ _ _ _ con1 con2) = minimum (map (findChoice) [con1, con2])
 findChoice (When obs _ con1 con2) = minimum ((findChoiceObs obs):(map (findChoice) [con1, con2]))
 
 
@@ -330,7 +330,7 @@ analyse_one_step_commits as = (analyse_one_step_commits_aux as)
 -- Remove top actions from contract (including those in Both trees)
 remove_actions :: Contract -> Contract
 remove_actions (Both contr1 contr2) = (Both (remove_actions contr1) (remove_actions contr2))
-remove_actions (CommitCash _ _ _ _ _ contr) = contr
+remove_actions (CommitCash _ _ _ _ _ contr1 contr2) = (Both (remove_actions contr1) (remove_actions contr2))
 remove_actions (RedeemCC _ contr) = contr
 remove_actions contr = contr
 
@@ -414,7 +414,7 @@ analyse_one_step_observables as =
 
 -- Analyses possible combinations of actions regarding the root node
 analyse_one_step_action_aux :: AnalysisState -> [AnalysisState]
-analyse_one_step_action_aux (as@ AS {rem_contract = (CommitCash ident per cash tout _ _),
+analyse_one_step_action_aux (as@ AS {rem_contract = (CommitCash ident per cash tout _ _ _),
                                      commits = comm@Input {cc = cc_old}, -- CommitCash
                                      curr_event = ce,
                                      possible_block = blo

### 2018-03-06 src/DepositIncentive.hs

@@ -16,14 +16,12 @@ import Semantics
 
 depositIncentive :: Contract
 depositIncentive = CommitCash (IdentCC 1) 1 100 10 200
-                     (Choice (BelowTimeout 10)
-                             (CommitCash (IdentCC 2) 2 20 20 200
-                               (Choice (BelowTimeout 20)
-                                       (When (PersonChoseSomething (IdentChoice 1) 1) 100
-                                             (Both (RedeemCC (IdentCC 1) Null)
-                                                   (RedeemCC (IdentCC 2) Null))
-                                             (Pay (IdentPay 1) 2 1 20 200
-                                                  (Both (RedeemCC (IdentCC 1) Null)
-                                                        (RedeemCC (IdentCC 2) Null))))
-                                       (RedeemCC (IdentCC 1) Null)))
-                             Null)
+                              (CommitCash (IdentCC 2) 2 20 20 200
+                                          (When (PersonChoseSomething (IdentChoice 1) 1) 100
+                                                (Both (RedeemCC (IdentCC 1) Null)
+                                                      (RedeemCC (IdentCC 2) Null))
+                                                (Pay (IdentPay 1) 2 1 20 200
+                                                     (Both (RedeemCC (IdentCC 1) Null)
+                                                           (RedeemCC (IdentCC 2) Null))))
+                                          (RedeemCC (IdentCC 1) Null))
+                              Null

### 2018-03-06 src/Semantics.hs

@@ -239,7 +239,7 @@ interpretObs _ FalseObs _
 
 data Contract =
     Null |
-    CommitCash IdentCC Person Cash Timeout Timeout Contract |
+    CommitCash IdentCC Person Cash Timeout Timeout Contract Contract |
     RedeemCC IdentCC Contract |
     Pay IdentPay Person Person Cash Timeout Contract |
     Both Contract Contract |
@@ -299,10 +299,10 @@ step _ st (When obs expi con con2) os
 -- Note that conformance of the commitment here is exact
 -- May want to relax this
 
-step commits st c@(CommitCash ident person val start_timeout end_timeout con) os
-  | cexe || cexs = (st {sc = ust}, con, [])
+step commits st c@(CommitCash ident person val start_timeout end_timeout con1 con2) os
+  | cexe || cexs = (st {sc = ust}, con2, [])
   | Set.member (CC ident person val end_timeout) (cc commits)
-        = (st {sc = ust}, con, [SuccessfulCommit ident person val])
+        = (st {sc = ust}, con1, [SuccessfulCommit ident person val])
   | otherwise = (st, c, [])
   where ccs = sc st
         cexs = expired (blockNumber os) start_timeout

### 2018-03-21 src/CrowdFunding.hs
Implement escrow and crowd-funding examples

@@ -1,0 +1,82 @@
+module CrowdFunding where
+
+import Semantics    
+
+-------------------------------------------------
+-- Implementation of a "limited crowd"-funding --
+-------------------------------------------------
+
+-- The contract allows a person 5 to collect 1000 ADA
+-- by inviting 4 people to donate.
+-- People 1 to 4 can first choose how much to donate
+-- and then commit that ammount of money until
+-- block 20. At block 10, if the ammount collected
+-- is equal or higher than 1000 ADA, person 5 will
+-- have until block 20 to claim all of it.
+-- In any case, people 1 to 4 will be able to redeem
+-- unclaimed money after block 20.
+
+crowdFunding :: Contract
+crowdFunding = Both (Both (Both (When (AndObs (PersonChoseSomething (IdentChoice 1) 1)
+                                               (ValueGE (MoneyFromChoice (IdentChoice 1) 1
+                                                                         (ConstMoney 0))
+                                                        (ConstMoney 1)))
+                                       10
+                                       (CommitCash (IdentCC 1) 1
+                                                   (MoneyFromChoice (IdentChoice 1) 1
+                                                                    (ConstMoney 0))
+                                                   10 20 Null Null)
+                                       Null)
+                                 (When (AndObs (PersonChoseSomething (IdentChoice 2) 2)
+                                               (ValueGE (MoneyFromChoice (IdentChoice 2) 2
+                                                                         (ConstMoney 0))
+                                                        (ConstMoney 1)))
+                                       10
+                                       (CommitCash (IdentCC 2) 2
+                                                   (MoneyFromChoice (IdentChoice 2) 2
+                                                                    (ConstMoney 0))
+                                                   10 20 Null Null)
+                                       Null))
+                           (Both (When (AndObs (PersonChoseSomething (IdentChoice 3) 3)
+                                               (ValueGE (MoneyFromChoice (IdentChoice 3) 3
+                                                                         (ConstMoney 0))
+                                                        (ConstMoney 1)))
+                                       10
+                                       (CommitCash (IdentCC 3) 3
+                                                   (MoneyFromChoice (IdentChoice 3) 3
+                                                                    (ConstMoney 0))
+                                                   10 20 Null Null)
+                                       Null)
+                                 (When (AndObs (PersonChoseSomething (IdentChoice 4) 4)
+                                               (ValueGE (MoneyFromChoice (IdentChoice 4) 4
+                                                                         (ConstMoney 0))
[…]

### 2018-03-21 src/Escrow.hs

@@ -1,0 +1,46 @@
+module CrowdFunding where
+
+import Semantics    
+
+------------------------------------------
+-- Implementation of an escrow contract --
+------------------------------------------
+
+-- The contract allows person 1 to pay 450 ADA
+-- to person 2 by using person 3 as an escrow.
+--
+-- Person 1 can commit 450 ADA until block 100.
+-- The payment will go through only if 2 out of the 3
+-- participants choose the number "1".
+--
+-- If 2 out of the 3 participants chooses the number "0"
+-- or if payment did not go through after block 100,
+-- the money will be refunded to person 1.
+
+escrow :: Contract
+escrow = CommitCash (IdentCC 1) 1
+                    (ConstMoney 450)
+                    10 100
+                    (When (OrObs (OrObs (AndObs (PersonChoseThis (IdentChoice 1) 1 0)
+                                                (OrObs (PersonChoseThis (IdentChoice 2) 2 0)
+                                                       (PersonChoseThis (IdentChoice 3) 3 0)))
+                                        (AndObs (PersonChoseThis (IdentChoice 2) 2 0)
+                                                (PersonChoseThis (IdentChoice 3) 3 0)))
+                                 (OrObs (AndObs (PersonChoseThis (IdentChoice 1) 1 1)
+                                                (OrObs (PersonChoseThis (IdentChoice 2) 2 1)
+                                                       (PersonChoseThis (IdentChoice 3) 3 1)))
+                                        (AndObs (PersonChoseThis (IdentChoice 2) 2 1)
+                                                (PersonChoseThis (IdentChoice 3) 3 1))))
+                          90
+                          (Choice (OrObs (AndObs (PersonChoseThis (IdentChoice 1) 1 1)
+                                                 (OrObs (PersonChoseThis (IdentChoice 2) 2 1)
+                                                        (PersonChoseThis (IdentChoice 3) 3 1)))
+                                         (AndObs (PersonChoseThis (IdentChoice 2) 2 1)
+                                                 (PersonChoseThis (IdentChoice 3) 3 1)))
+                                  (Pay (IdentPay 1) 1 2
+                                       (AvailableMoney (IdentCC 1))
+                                       100
+                                       (RedeemCC (IdentCC 1) Null))
+                                  (RedeemCC (IdentCC 1) Null))
+                          (RedeemCC (IdentCC 1) Null))
+                    Null

### 2018-03-21 src/Escrow.hs
Fix module name

@@ -1,6 +1,6 @@
-module CrowdFunding where
+module Escrow where
 
-import Semantics    
+import Semantics
 
 ------------------------------------------
 -- Implementation of an escrow contract --

### 2018-03-21 blockly/src/Main.hs
Add examples to web demo

@@ -5,6 +5,8 @@ import qualified Haste.JSString as J
 import qualified Haste.Prim as P
 import qualified Haste.DOM as D
 import qualified DepositIncentive as DI
+import qualified CrowFunding as CF
+import qualified Escrow as E
 import Semantics
 import ContractFormatter
 import SmartInputs
@@ -367,6 +369,19 @@ clear_everything = do clear_workspace
                       set_input_text "([], [], [], [])"
                       set_output_text ""
 
+-- examples
+depositIncentive = do clear_everything
+                      code_to_contract DI.depositIncentive
+                      c2b 
+
+crowdFunding = do clear_everything
+                  code_to_contract CF.crowdFunding
+                  c2b
+
+escrow = do clear_everything
+            code_to_contract E.escrow
+            c2b
+
 -- workspace to code
 workspace_to_code_aux :: IO (String)
 workspace_to_code_aux = F.ffi (J.pack "(function () {return Blockly.Marlowe.workspaceToCode(demoWorkspace);})")
@@ -525,6 +540,9 @@ main = do c2b
           F.export (J.pack "refreshActions") refreshActions
           F.export (J.pack "addAction") addAction
           F.export (J.pack "addActionWithNum") addActionWithNum
+          F.export (J.pack "depositIncentive") depositIncentive
+          F.export (J.pack "crowdFunding") crowdFunding
+          F.export (J.pack "escrow") escrow
           refreshActions
           return ()
 

### 2018-03-21 tex/paper.tex
Add example and note about Money to paper

@@
 \sjtnote{Preliminary overview; need to revisit.}
-The paper is structured like this: we begin with an overview of the model, including the assumptions made in designing it, the types of the principal functions and a description of the DSL as an algebraic type, constructor by constructor.
-The section aims to give a high level overview, and some of the concrete details are to be found in the repo.
-We follow this with discussions of the kinds of formal analysis which might be made for the DSL, as well as discussing a  set of questions raised by the process of modelling.
-We then look at the example of BLAH, as an illustration, and then look at how Marlowe relates to other work in the area.
-We conclude by pointing out
+The paper is structured like this: we begin with an overview of the model, including the assumptions made in designing 
+it, the types of the principal functions and a description of the DSL as an algebraic type, constructor by constructor. 
+The section aims to give a high level overview, and some of the concrete details are to be found in the repo. We follow 
+this with discussions of the kinds of formal analysis which might be made for the DSL, as well as discussing a  set of 
+questions raised by the process of modelling. We then look at the example of ``incentivised saving'', as an 
+illustration, and then look at how Marlowe relates to other work in the area. We conclude by pointing out
 
@@
 The same observation also applies to redeeming a commitment of cash.
-\item Observables are observed at a particular time and in a particular context.
-We assume that the recording and identification of each (instance of an) observation is provided by the infrastructure of the system.
-We provide a ``little language'' for describing observations: \haskellinline{Observation}.
+\item Observables are observed at a particular time and in a particular context. We assume that the recording and 
+identification of each (instance of an) observation is provided by the infrastructure of the system. We provide a 
+``little language'' for describing observations: \haskellinline{Observation}. Similar properties are true for
+\haskellinline{Money} amounts: their values are calculated depending on the inputs and state of execution of the 
+contract, they can be deterministically established given the context, and they are described through a small 
+``expression language''.
 \item The fact of making commitments also needs to be recorded.
@@
    Null |
-   CommitCash IdentCC Person Cash Timeout Timeout Contract Contract |  
+   CommitCash IdentCC Person Money Timeout Timeout Contract Contract |  
    RedeemCC IdentCC Contract |
-   Pay IdentPay Person Person Cash Timeout Contract |  
+   Pay IdentPay Person Person Money Timeout Contract |  
    Both Contract Contract |
@@
 
-\item \haskellinline{CommitCash idCC p n t1 t2 k1 k2}  For this contract to make progress,\begin{itemize}
-\item either before the timeout \haskellinline{t1} the user \haskellinline{p} makes a cash commitment of value \haskellinline{n} and timeout \haskellinline{t2}  with the identifier \haskellinline{idCC} ,
+\item \haskellinline{CommitCash idCC p v t1 t2 k1 k2}  For this contract to make progress,\begin{itemize}
+\item either before the timeout \haskellinline{t1} the user \haskellinline{p} makes a cash commitment of value 
+\haskellinline{v} and timeout \haskellinline{t2}  with the identifier \haskellinline{idCC}, 
 \item
@@
 
-\item \haskellinline{Pay p1 p2 int t k} makes it possible, assuming that sufficient funds are available, for \haskellinline{p1} to claim  a payment of \haskellinline{int} from \haskellinline{p2} before the timeout \haskellinline{t}.
-The contract continues as \haskellinline{k}.
-By `available' we mean that sufficient commitments have been made and  not yet expired to cover the payment; in this case the payment  uses the currency allocated by the cash commitments made by \haskellinline{p1} that expire the earliest.
+\item \haskellinline{Pay p1 p2 v t k} makes it possible, assuming that sufficient funds are available, for 
+\haskellinline{p1} to claim  a payment of \haskellinline{v} from \haskellinline{p2} before the timeout 
+\haskellinline{t}. The contract continues as \haskellinline{k}. By `available' we mean that sufficient commitments have 
+been made and  not yet expired to cover the payment; in this case the payment  uses the currency allocated by the cash 
+commitments made by \haskellinline{p1} that expire the earliest.
 
@@
 
-\section{Rock-paper-scissors explanation (not re-formatted)}
+\section{Incentivised saving}
 
-In this section, we explain how the DSL can be used for designing a contract that enforces a round of rock-paper-scissors where the winner will receive from the loser 10 units of currency and not revealing your hand also has a penalty of 10 units.
+In this section, we explain how the DSL can be used for designing a contract that incentivises saving. We use the 
+following example: ``participant 2'' offers 20 ADA to ``participant 1'' in exchange for saving 100 ADA for 100 blocks. 
+``Participant 1'' will be able to withdraw the savings after block 20, but if that option is chosen, then ``participant 
+2'' will not pay the extra 20 ADA to ``participant 1''.
 
-We will start top to bottom. First we define identities for the players and the commitments.
-The player 1 will be identified by integer 1:
-player1 = 1
-the cash commitment of player 1 will be identified by:
-p1 = IdentCC 1
-and his value commitment will be identified by:
-v1 = IdentCV 1
-In the same way, player 2 will be identified by integer 2:
-player2 = 2
-the cash commitment player 2 will be identified by:
-p2 = IdentCC 2
-And his value commitment will be identified by:
-v2 = IdentCV 2
+We will start from outer to inner primitives. The first thing is to allow ``participant 1'' to commit the 100 ADA. The 
+primitive says that with identity \haskellinline{IdentCC 1} (this is the identity of the commitment), ``participant 
+\haskellinline{1}'' can commit \haskellinline{100} ADA before block \haskellinline{10}, and the commitment must 
+expire on block \haskellinline{200}. But if the money is not committed before block \haskellinline{10}, then the 
+contract will continue as \haskellinline{Null}.
 
-And because choices must be represented as Int we encode them as follows:
+\begin{haskellcode}
+CommitCash (IdentCC 1) 1 (ConstMoney 100) 10 200
+           step2
+           Null
+\end{haskellcode}
 
-data RPS = Rock | Paper | Scissors
+If the commitment is done correctly and on time, we continue as in \haskellinline{step2}. \haskellinline{step2} is 
+another commitment, this time by ``participant 2''. The primitive says that with identity \haskellinline{IdentCC 2},
+``participant \haskellinline{2}'' can commit \haskellinline{20} ADA before block \haskellinline{20}, and the commitment
+must expire on block \haskellinline{200}. But if the money is not committed before block \haskellinline{20}, then the
+contract will continue as \haskellinline{RedeemCC (IdentCC 1) Null}, which in turn gives ``participant 1'' the 
+opportunity to withdraw all the money committed immediately.
 
-all\_rps = [Rock, Paper, Scissors]
+\begin{haskellcode}
+CommitCash (IdentCC 2) 2 (ConstMoney 20) 20 200
+           step3
+           (RedeemCC (IdentCC 1) Null)
+\end{haskellcode}
 
-encoding\_rps Rock = 1      -- rock is encoded as 1
-encoding\_rps Paper = 2     -- paper is encoded as 2
-encoding\_rps Scissors = 3  -- scissors is encoded as 3
+If the commitment is done correctly and on time, we continue as in \haskellinline{step3}. \haskellinline{step3}
+waits until choice with identity \haskellinline{IdentChoice 1} is made by ``participant \haskellinline{1}'' or
+until block \haskellinline{100} (whatever happens first). If the choice is made before block \haskellinline{100}, 
+it continues as \haskellinline{stepRedeem}, if block \haskellinline{100} comes first, then it continues as 
+\haskellinline{step4}.
 
-all\_rps\_encoded = map (encoding\_rps) all\_rps
+\begin{haskellcode}
+When (PersonChoseSomething (IdentChoice 1) 1) 100
+     stepRedeem
+     step4
+\end{haskellcode}
 
-We first ask the players to commit money.
-We will need 20 units from each player since we want 10 units to enforce that the hand is revealed, and 10 units for the prize.
-Theoretically we could reuse just 10 units, but in practice recommitting payments may limit the applicability of the DSL to certain systems like Bitcoin, so it does not support it at the moment.
+\haskellinline{step4} pays the \haskellinline{20} ADA committed by ``participant \haskellinline{2}'' in the commit with 
+identity \haskellinline{IdentPay 1} to ``participant \haskellinline{1}'', if the payment is claimed before block 
+\haskellinline{200}. Once the payment is claimed or after block 200, it continues as \haskellinline{stepRedeem}.
 
-The commitments are chained as follows:
-* Cash commitment p1 is done by player 1, consists of 20 units and expires on block 42 same for p2 by player 2.
+\begin{haskellcode}
+Pay (IdentPay 1) 2 1 (ConstMoney 20) 200 stepRedeem
+\end{haskellcode}
 
-CommitCash p1 player1 20 42
-  (CommitCash p2 player2 20 42
+Finally, \haskellinline{stepRedeem} allows both participants to redeem all the remaining committed funds.
 
-* Value commitments are done for two choices of one of the three values that encode rock, paper, or scissors.
+\begin{haskellcode}
+Both (RedeemCC (IdentCC 1) Null)
+     (RedeemCC (IdentCC 2) Null)
+\end{haskellcode}
 
-     (CommitValue v1 (all\_rps\_encoded) player1
-        (CommitValue v2 (all\_rps\_encoded) player2
-
-* The see-saw mechanism ensures that both players reveal or none does.
-This is fundamentally achieved by forcing a player to reveal the other when he reveals.
-But we need to do this steadily to give players enough time to do the actions required by the rest of the contract, otherwise one player may push the other one to timeout.
-
-    (RevealCV v1
-            (Choice (BelowTimeout 10)
-                    (Pay player2 player1 10
-                       (RevealCV v2
-                              (Choice (BelowTimeout 20)
-                                  (Pay 1 2 10
-                                       reward\_winner)
-                                  Null)))
-                    Null)))))
-
-First RevealCV waits for the owner of value commitment v1 (player 1) to reveal his choice.
-If the reveal happens before block number 10 (after the start of the contract instance), player 2 pays 10 currency units to player 1.
-Otherwise we consider the contract timed out and make the contract Null, the existing cash commitments will be automatically refunded on block 42.
-The same is done for player 2.
-This time he is incentivised to reveal his hand before block number 20, since that is the only way he will recover the 10 currency unit.
-He might already know that he has lost and decide not to reveal, that will have the same effect than revealing.
-If desired we could increase the see-saw penalty to make not revealing worse than losing.
-
-If the contract remaining becomes reward\_winner, we already have enough information to know who won and we will have exactly 10 currency units committed by each player.
-
-So we define reward\_winner as follows. We just use Choice twice, one to check if Player 1 won, one to check if Player 2 won, and the default option is just refunding everything:
-
--- If player 1 won
-Choice (OrObs (AndObs (CVRevealedAs v1 (encoding\_rps Paper))
-                      (CVRevealedAs v2 (encoding\_rps Rock)))
-              (OrObs (AndObs (CVRevealedAs v1 (encoding\_rps Scissors))
-                             (CVRevealedAs v2 (encoding\_rps Paper)))
-                     (AndObs (CVRevealedAs v1 (encoding\_rps Rock))
-                             (CVRevealedAs v2 (encoding\_rps Scissors)))))
--- player2 pays 10 currency units to player 1
-       (Pay player2 player1 10
--- and everything else is redeemed
-          (Both (RedeemCC p1 Null)
-                (RedeemCC p2 Null)))
--- if player 2 won
-       (Choice (OrObs (AndObs (CVRevealedAs v2 (encoding\_rps Paper))
-                              (CVRevealedAs v1 (encoding\_rps Rock)))
-                      (OrObs (AndObs (CVRevealedAs v2 (encoding\_rps Scissors))
-                                     (CVRevealedAs v1 (encoding\_rps Paper)))
-                             (AndObs (CVRevealedAs v2 (encoding\_rps Rock))
-                                     (CVRevealedAs v1 (encoding\_rps Scissors)))))
--- player1 pays 10 currency units to player 2
-               (Pay player1 player2 10
--- and everything else is redeemd
-                  (Both (RedeemCC p1 Null)
-                        (RedeemCC p2 Null)))
--- otherwise everything else is redeemed
-               (Both (RedeemCC p1 Null)
-                     (RedeemCC p2 Null)))
-
-It is also worth noting that, because the possible choices are specified, RedeemCV requires that the choice revealed is valid, otherwise it behaves as nothing was revealed.
-Thus, we know that if no player won it must be a draw.
+The whole contract remains as follows:
+\begin{haskellcode}
+CommitCash (IdentCC 1) 1
+           (ConstMoney 100)
+           10 200
+           (CommitCash (IdentCC 2) 2
+                       (ConstMoney 20)
+                       20 200
+                       (When (PersonChoseSomething (IdentChoice 1) 1)
+                             100
+                             (Both (RedeemCC (IdentCC 1) Null)
+                                   (RedeemCC (IdentCC 2) Null))
+                             (Pay (IdentPay 1) 2 1
+                                  (ConstMoney 20)
+                                  200
+                                  (Both (RedeemCC (IdentCC 1) Null)
+                                        (RedeemCC (IdentCC 2) Null))))
+                       (RedeemCC (IdentCC 1) Null))
+           Null
+\end{haskellcode}
 

### 2018-03-24 tex/paper.tex
initial work on design rationale, plus references; more to do

@@
 illustration, and then look at how Marlowe relates to other work in the area. We conclude by pointing out
+
+\sjtnote{To talk about, in no particular order \ldots}
+\begin{itemize}
+\item
+Marlowe as the ``canary in the coalmine'': test out how it will work, how it will interface with
+other aspects of the Cardano platform
+\item
+Be clear that this paper describes version 1.2 of Marlowe.
+\end{itemize}
+
+\section{Design rationale}
+
+\sjtnote{References to background papers still to be added.}
+
+\medskip
+\noindent
+In this section we explore the rationale for building the Marlowe language, and put it into the context of other work in the field.
+
+\subsection{Smart contracts on blockchain}
+
+As more than one wag has observed, many of the programs that run on blockchains are neither smart nor indeed form contracts in any real sense.
+Nonetheless, this terminology has stuck, and has become shorthand for ``programmed interactions on the blockchain''.
+In an earlier paper we presented a survey of the approaches available~\cite{cryptoeprint:2016:1156}; we give an overview here.
+
+Blockchains are executed in a replicated form by parties who cannot be guaranteed not to be hostile, either by directly trying to change the contents of the chain, or through trying to affect other properties of the chain by indirect means (such as swamping honest parties with work).
+Programming on the blockchain therefore needs to be constrained in some ways, since it has to be amenable to replication or verification within a reasonable time if the security and integrity of the chain are to be preserved.
+We look at representatives of the main approaches now.
+
+\paragraph{Bitcoin script}
+
+One approach to this is to choose mechanisms such as bitcoin script which are manifestly non-Turing complete.
+A bitcoin script, written in a Forth-like language, is essentially linear: it can branch, but the language contains neither looping constructs nor recursion.
+It is therefore straightforward not only to see that scripts will terminate, but also to give an accurate estimate of the time taken to execute a script.
+
+\paragraph{Ethereum}
+
+On the other hand, the Ethereum system provides a Turing-complete language for the EVM virtual machine, and a higher-level programming language, Solidity, that compiles into EVM code.
+However, EVM and Solidity programs are constrained \emph{post hoc} by two mechanisms: program execution must be paid for using `gas' proportional to the effort expended, and a set of \emph{ad hoc} limits on program execution, e.g.\ on stack size.
+
+\paragraph{Nxt}
+
+Programmability of the system is provided through a ``fat'' high-level API, which is accessible from Nxt clients through a REST interface.
+The API provides functionality supporting various kinds of transactions.
+The core software itself doesn't support any form of scripting language; rather, users are expected to work with the built in transaction types and transactions that support some 250 primitive operations; these can be ``scripted'' in a client (only) using a binding to the API, which is available, for instance, in JavaScript.
+
+\paragraph{Other approaches}
+
+Other approaches to scripting include using business process notations, such as the OMG's BPMN, process calculi such as Rholang, logic rules (in Formal Contract Logic) and finite state machines.
+More details of these are provided in the survey~\cite{cryptoeprint:2016:1156}.
+
+\subsection{Blockchain platforms: Cardano and others}
+
+There is not a single model of how a blockchain operates, and in this section we look at some of the tradeoffs possible in this space.
+This allows us to give a context to some of the design decisions taken in designing Marlowe.
+
+The simplest blockchain is Bitcoin.
+Transactions on Bitcoin are made by spending the (as yet) unspent outputs of previous transactions (`unspent transaction outputs' or UTxOs).The chain need not maintain any state, such as the current value owned by any particular participant: such \emph{account} information is implicit and external to the Bitcoin blockchain.
+On the other hand, the Ethereum model explicitly keeps track of account values, and so this information needs to be kept as state information on chain.
+While it is possible to translate between UTxO-based and account-based chains \sjtnote{cite Bruno paper} the UTxO model is more fundamental, and requires less support from the implementation of the chain itself.
+
+Some chains have a layered design, and Cardano~\cite{cardano} does this.
+The settlement layer (SL) is to support settlement of transactions, whereas the computation layer (or CL) supports more powerful general computations over the chain.
+Moreover, the Cardano roadmap \cite{cardano-rationale} envisages the possibility of computation taking place, in part at least, on a sidechain \sjtnote{citation} rather than on the main chain itself, which may support only the SL.
+
+\medskip
+\noindent
+\sjtnote{To talk about, in no particular order \ldots}
+\begin{itemize}
+\item
+push vs pull
+\item
+DSLs and Embedded DSLs
+\item
+Multiple languages: can reference aeternity here (see presentation), and tezos (Michelson: stack-based and Liquidity, strongly typed functional)
+\end{itemize}
 

### 2018-03-24 tex/paper.tex
completed work on background to design rationale (new sections 2.1 and 2.2), plus references; more to do

@@
 
-\sjtnote{References to background papers still to be added.}
+\sjtnote{References to some of the background work still to be added.}
 
@@
 \noindent
-In this section we explore the rationale for building the Marlowe language, and put it into the context of other work in the field.
+In this section we look at existing work on smart contracts and blockchains in general, before exploring the rationale for building the Marlowe language.
 
-\subsection{Smart contracts on blockchain}
+\subsection{Smart contracts on blockchains}
 
@@
 
+\paragraph{Multiple languages}
+
+A common feature of many blockchain platforms is that they provide multiple scripting languages.
+As we noted earlier, Ethereum can be programmed at the EVM level as well as by using the high-level Solidity language.
+Tezos~\cite{tezos-white-paper} supports the stack-based, strongly-typed, functional language Michelson, but also provides a high-level language, Liquidity, that compiles into Michelson.
+Liquidity is also functional and strongly typed, but provides more constructs in a more familiar syntax, namely that of a subset of OCaml.
+The \ae{}ternity system~\cite{aeternity} provides multiple languages and VMs:\footnote{As presented by Erik Stenman at CODE BEAM SF, March 2018, \url{https://www.youtube.com/watch?v=VXsqvfPIdWg}} the functional language  Sophia (akin to Reason) and the Functional Typed Warded Virtual Machine (FTWVM) for safe ``system level programming'', the language Varna and the HLM for simple contracts, and a (port of) Solidity and the EVM for compatibility with Ethereum.
+In a similar way, Cardano provides support for IELE~\sjtnote{citation}, a rational reconstruction of the EVM, and thus for Solidity contracts too.
+
+\paragraph{Domain-specific languages}
+
+A domain-specific language or DSL is a high-level language designed to work in a specific field or domain.
+The intention is that because the users will know about the field, the constructs of the language can be designed to be meaningful to them, and also that because of its nature the DSL need not include all the features of a general purpose language.
+Removing this clutter and having the remaining operations directly reflect the application area is intended to make the language more accessible to domain experts who do not necessarily see themselves as programmers.
+
+Stand-alone DSLs have the advantage of providing appropriate, domain-level error messages when things go wrong, but suffer the disadvantage of having to be implemented from scratch.
+An alternative to this are \emph{embedded} DSLs (EDSLs), which provide a ``little language'' for the particular domain embedded within a general-purpose host language.
+This means that parts of the host -- such as arithmetical expressions, or list idioms --  can be used to extend the expressibility of the DSL.
+A notable example of this is the financial contracts language described by Peyton Jones and his collaborators~\cite{PeytonJones:2000}, which is embedded in Haskell.
+
 \paragraph{Other approaches}
 
-Other approaches to scripting include using business process notations, such as the OMG's BPMN, process calculi such as Rholang, logic rules (in Formal Contract Logic) and finite state machines.
+There are a variety of other approaches to scripting, including business process notations, such as the OMG's BPMN; process calculi such as Rholang; logic rules (in Formal Contract Logic), and finite state machines.
 More details of these are provided in the survey~\cite{cryptoeprint:2016:1156}.
@@
 
+\paragraph{UTxO and accounts}
+
 The simplest blockchain is Bitcoin.
@@
 
-Some chains have a layered design, and Cardano~\cite{cardano} does this.
-The settlement layer (SL) is to support settlement of transactions, whereas the computation layer (or CL) supports more powerful general computations over the chain.
-Moreover, the Cardano roadmap \cite{cardano-rationale} envisages the possibility of computation taking place, in part at least, on a sidechain \sjtnote{citation} rather than on the main chain itself, which may support only the SL.
+\paragraph{Interaction modality}
 
-\medskip
-\noindent
-\sjtnote{To talk about, in no particular order \ldots}
-\begin{itemize}
-\item
-push vs pull
-\item
-DSLs and Embedded DSLs
-\item
-Multiple languages: can reference aeternity here (see presentation), and tezos (Michelson: stack-based and Liquidity, strongly typed functional)
-\end{itemize}
+Contracts can be conceived of as acting in two ways.
+In a \emph{push} model, contracts are seen to be pro-active, and to make things happen.
+In the case of blockchain, this would include making payments or transactions happen.
+Alternatively, in the \emph{pull} model contracts enable certain things to happen, such as transactions, but the transactions need to be effected by an external player.
+The pull model makes fewer demands on the blockchain implementation than the push model, which requires the contract to have direct access to modify the chain.
+
+\paragraph{Layered design and sidechains}
+
+Some chains have a layered design, and in particular Cardano~\cite{cardano} has.
+The settlement layer (SL) is to support settlement of transactions but nothing more complicated, whereas the computation layer (or CL) supports more powerful general computations over the chain.
+Moreover, the Cardano roadmap \cite{cardano-rationale} envisages the possibility of computation taking place, in part at least, on a sidechain rather than on the main chain itself, which may support only the SL.
+Other off-chain approaches include the Lightning network, overlaid on blockchains (usually Bitcoin) and the channels of \ae{}ternity, introduced below.
+
+\subsection{Designing Marlowe}
+
+\sjtnote{To be completed.}
 

### 2018-03-24 tex/paper.tex
initial work on commentary; more to do

@@
 
+\setlength{\itemsep}{2pt}
+
 % Document starts
@@
 \item In the traditional model, enforcement of the contract is the responsibility of the legal system; in the decentralised blockchain model, in which there is no central authority, the contract needs to be enforced by design.
-\item As noted by  the authors of Findel~\cite{findel} this means in particular that payments need to be bounded; this is made concrete in our account by our notion of commitments.
 \item There is not a single blockchain model, and we give a brief overview of the different mechanisms for scripting on blockchain (although this is not the primary purpose of this article). 
@@
 Be clear that this paper describes version 1.2 of Marlowe.
+\item 
+Enforcement and promptness.
 \end{itemize}
@@
 Liquidity is also functional and strongly typed, but provides more constructs in a more familiar syntax, namely that of a subset of OCaml.
+
 The \ae{}ternity system~\cite{aeternity} provides multiple languages and VMs:\footnote{As presented by Erik Stenman at CODE BEAM SF, March 2018, \url{https://www.youtube.com/watch?v=VXsqvfPIdWg}} the functional language  Sophia (akin to Reason) and the Functional Typed Warded Virtual Machine (FTWVM) for safe ``system level programming'', the language Varna and the HLM for simple contracts, and a (port of) Solidity and the EVM for compatibility with Ethereum.
@@
 There are a variety of other approaches to scripting, including business process notations, such as the OMG's BPMN; process calculi such as Rholang; logic rules (in Formal Contract Logic), and finite state machines.
-More details of these are provided in the survey~\cite{cryptoeprint:2016:1156}.
+More details of these are provided in~\cite{cryptoeprint:2016:1156}.
+
+\paragraph{Related work}
+
+The Findel~\cite{findel} project examines financial contracts on the Ethereum platform, and the authors note that  payments need to be bounded; this is made concrete in our account by our notion of commitments.
+They take no account of commitments or timeouts as we do below, but it should be noted that the Ethereum platform is more powerful than the one that we target in this paper.
 
@@
 
-Some chains have a layered design, and in particular Cardano~\cite{cardano} has.
-The settlement layer (SL) is to support settlement of transactions but nothing more complicated, whereas the computation layer (or CL) supports more powerful general computations over the chain.
+Some chains have a layered design, in particular Cardano~\cite{cardano}.
+The Cardano Settlement Layer (SL) is to support settlement of transactions but nothing more complicated, whereas the Computation Layer (or CL) supports more powerful general computations over the chain.
 Moreover, the Cardano roadmap \cite{cardano-rationale} envisages the possibility of computation taking place, in part at least, on a sidechain rather than on the main chain itself, which may support only the SL.
@@
 
-\sjtnote{To be completed.}
+In designing Marlowe we have chosen a number of options; this description applies to Marlowe version 1.2 \sjtnote{Is there an obvious way of making that clear in the repo?
+I really don't understand how git deals with ``releases'' or versions.}
+
+\paragraph{Pull}
+
+While earlier versions used a ``push'' model of operations, we have revised the language to give a ``pull'' semantics.
+This means, for example, that the language can \emph{enable} a payment to be made, but cannot make the payment happen.
+It is up to the interested party (or parties) to make the change happen.
+It is therefore possible for the payment not to happen, despite the fact that this is against the interest of the receiving party.
+The same
+
+\paragraph{UTxO vs.\ accounts} \sjtnote{Pablo: please check this section carefully}
+
+The language semantics is described in terms of \emph{commitments} of sums of cryptocurrency -- which we choose to abbreviate to ``cash'' -- and so is agnostic of the underlying implementation.
+However, the current redemption mechanism in Plutus Core is not strong enough to support Marlowe, and we are actively investigating the best way to extend this mechanism to implement Marlowe in full on Plutus Core.
+
+\paragraph{DSL or EDSL?}
+
+Marlowe is described as an algebraic \haskellinline{data} type in Haskell, and as such can be seen as a simple EDSL.
+Note, however, that the implementation presented here is for the purposes of presenting the semantics and behaviour of the language in the abstract.
+The implementation on the Cardano platform could be stand-alone, but more likely it will be a DSL embedded in Plutus or Plutus Core.
+The details of this remain to be decided, as noted in the introduction.
+
+\begin{figure}[t]
+\begin{center}
+\includegraphics[width=0.6\textwidth]{pix/context.png}
+\caption{The context for the contract}
+\label{fig:context}
+\end{center}
+\end{figure}
 
 \section{The model}
+
 The DSL is modelled as an algebraic type in Haskell, together with an executable small-step semantics. 
-A running contract will interact with its environment in two ways.
+We start by looking at the different types used by the model, and then look at the \haskellinline{Contract} DSL itself. \sjtnote{To complete this introduction}
+
+\subsection{The model types}
+
+A running contract will interact with its environment in two ways, as shown in Figure \ref{fig:context}.
+
+\paragraph{Observables}
 
@@
 
+\paragraph{Inputs}
+
 On the other hand, at each step there are -- potentially, at least -- a variety of inputs available from the participants.
@@
 The particular inputs at a given step are described by a value of type \haskellinline{Input}.
+
+\paragraph{Commitments}
 
@@
 
+\begin{figure}[t]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{pix/step-type.png}
+\caption{The \haskellinline{step} function}
+\label{fig:step-function}
+\end{center}
+\end{figure}
+
+\paragraph{The \haskellinline{step} function}
+
 This analysis justifies the type of the \haskellinline{step} function:
@@
 
-The model is based on a number of assumptions.
+which is also shown in Figure \ref{fig:step-function}.
+
+\paragraph{Infrastructure}
+
+The model makes a number of assumptions about the blockchain infrastructure in which it is run.
 
 \begin{itemize}
-\item It is assumed that cryptographic ``services'' or operations are provided by a layer external to the system, and so they need not be modelled explicitly here.
-\item Making a commitment is not something that a contract can perform; rather, it can require that a commitment is made, but that then has to be established externally: hence the input of (a set of) commitments at each step.
+\item It is assumed that cryptographic functions and operations are provided by a layer external to the system, and so they need not be modelled explicitly here.
+\item We assume that time is ``coarse grained'' and measured simply by block number, so that timeouts etc.\ are delimited using nothing finer than block number. 
+\item Because we assume a ``pull'' model, making a commitment is not something that a contract can perform; rather, it can enable that a commitment is made, but that then has to be established externally: hence the input of (a set of) commitments at each step.
 The same observation also applies to redeeming a commitment of cash.
+Commitments that have been made or redeemed need to be recorded somehow.
+
 \item Observables are observed at a particular time and in a particular context. We assume that the recording and 
 identification of each (instance of an) observation is provided by the infrastructure of the system. We provide a 
-``little language'' for describing observations: \haskellinline{Observation}. Similar properties are true for
+``little language'' for describing observations: \haskellinline{Observation}. 
+\item
+Similar properties are true for
 \haskellinline{Money} amounts: their values are calculated depending on the inputs and state of execution of the 
@@
 ``expression language''.
-\item The fact of making commitments also needs to be recorded.
 \item It is important to be aware that there may be different instances of contracts executed, and for each instance there will typically be different instances of observables (and commitments) observed and recorded.
 \item The model manages the release of funds back to the committer when a cash commitment expires (see discussion of the \haskellinline{fullStep} function below).
-\item We assume that time is ``coarse grained'' and measured simply by block number, so that timeouts etc.\ are delimited using nothing finer than block number. 
 \end{itemize}
 
-\medskip
-\noindent
+\subsection{The \haskellinline{Contract} type}
+
 The type of contracts is given by
@@
 
+\subsection{The \haskellinline{step} function}
+
 The \haskellinline{step} function is total, so that for every contract a result of stepping is defined.
@@
 
-The constructors should be interpreted thus:
+The constructors are interpreted like this:
 
@@
 \end{itemize}
-
-\subsection{Potential additions to the model}
-
-The model that we have discussed so far is deliberately kept as compact as possible. However there are a number of potential extensions to the model which we introduce here.
-
-We have examined a single kind of commitment, namely a commitment of a certain amount of currency. In different 
-applications, such as on-chain game playing of games like poker, a commitment to a value will be required. A value 
-commitment would include a choice of a play in a game, which would be chosen and concealed from the other players until 
-a later point in play, when it can be revealed. The paper \cite{kumaresan2015use} makes use of this and other 
-cryptographic primitives to enforce the rules of poker while maintaining the cards of each player secret.
-
-We should monadify, so that identifiers are automatically generated.
-This gives two nice properties.
-First, they are guaranteed to be unique; secondly they can be generated only in cases where a commit has been made successfully (within the required time period).
-It might well be that these should be generated by the system; this could guarantee uniqueness, freshness etc, and could also lead to a monadic DSL, with bind used for binding names to commitments.
-
-\section{Analysis}
-What questions can we ask? What answers can we give?
-
-\subsubsection{What are the valid contracts?}
-
-One possible definition is ``those contracts that do not lead to a \haskellinline{FailedPay} action''.
-Does this make sense independently of the inputs to a contract (i.e.
-observables, commitments, redeems, timeouts, and randomness)?
-Some relevant points (NB discussion incomplete):
-\begin{itemize}
-\item For \haskellinline{Pay} require a sufficient balance not to produce a \haskellinline{FailedPay} or a \haskellinline{DuplicateRedeem} action. 
-\item For \haskellinline{Both} require a sufficient balance for both contracts combined.
-Need to think here whether the two conjuncts should be independent, or whether a commitment in one half can be used in the other.
-\item For \haskellinline{Choice} require a sufficient balance for the maximum of the requirements of the individual contracts. 
-\item For \haskellinline{When}, need to ``age'' the existing commitments relative to the new starting point of the continuation contract. 
-\item Are there any constraints on  \haskellinline{CommitCash} and \haskellinline{RedeemCC} statements?
-Not directly, but there need to be sufficient funds available to pay if needed, or enough time for the commitments to time out, if not.
-\item Originally we had included that we should include all the ``progress'' requirements from point 2 below; but that's probably not the right thing here.
-
-\end{itemize}
-
-\subsubsection{Do all contracts potentially evaluate to \haskellinline{Null}?}
-
-That is, for each contract do there exist sequences of values of Commit and OS such that stepping the contract with these inputs leads to Null? Might we have to rule out unsatisfiable observations?
-Some relevant points (NB discussion incomplete):
-\begin{itemize}
-\item For \haskellinline{When}, need to be able to make \haskellinline{interpretObs st obs os} true.
-Since there is a timeout to \haskellinline{When} then there's no need to impose this condition, as we'll progress at timeout if not before.
-\item For \haskellinline{CommitCash}, need to have the right commits in the \haskellinline{Commit} argument.
-\item Similarly for \haskellinline{RedeemCC} statements.
-\item It would also be possible to put timeouts on to \haskellinline{CommitCash},  and \haskellinline{RedeemCC} operations, and if this were to be done there would be no remaining requirements for contract execution to make progress.
-\item There are no constraints on other constructs.
-\end{itemize}
-
-\section{Questions}
-
-The model outlined here raises a number of questions, enumerated here.
-
-\begin{enumerate}
-\item Does the \haskellinline{Both} constructor potentially produce invalid/misleading contracts, combining a commitment in one half with its use in the other?
-The problem is that the ``use'' conjunct could potentially be executed, leading to a failed payment.
-\item How do we ensure that contracts can be enforced? Is there a notion of validity for contracts that can be enforced? To which extent is this notion enforced by current constructors?
-\item The model does not take time as an observable, but rather it is defined in terms of \haskellinline{BlockNumber} only. 
-\item We could use ``slot time'' https://eprint.iacr.org/2016/889.pdf ``We consider a setting where time is divided into discrete units called slots.
-A ledger, described in more detail below, associates with each time slot (at most) one ledger block.''
-\item Is it possible for block number to depend on progress of computation?
-We conjecture not, because otherwise it would be possible to produce timelocks, under which it becomes impossible for time to progress.
-\item Does a contract expire if a surrounding commitment does? We think not.
-\item The model is deterministic, with all non-determinism being external, and confined to observables.
-This is surely necessary in order to support computation replay, but does it restrict the expressibility of the model?
-\item Is there a notion of causality for these contracts? Would we be able to characterise it semantically? Syntactically?
-\item We have modelled cash as fungible: we make payments using the cash commitments in order of expiry time, with those expiring earliest being used first.
-Would there be any reason for using a different approach?
-Yes!
-UTxO.
-\item Timeouts are expressed in absolute time: presumably they could be relative (with recording on the blockchain of the times at which they are triggered).
-%\footnote{This contradicts what you've said elsewhere.
-I think timeouts should be times, not blockchain numbers.
-I suspect it would be better not to use blockchain numbers.
-I don't want to have to react more quickly if lots of people are making transactions that day.
-Also, it avoids question 8.}\footnote{See addition about slot time from Ouroboros paper (as referenced in the body of the paper).}
-\item Some observables have a ``monotonicity'' property: once a choice is made, it is fixed forever; on the other hand, something not yet chosen is only so thus far; do we need to reflect this in the observations that we allow?
-%\item The seesaw construction is key to implementing e.g.
-poker, and it can be described in terms of the existing constructs (modulo the answer to point 4 above).
-However, should seesaw be a primitive in its own right?
-%\footnote{No.
-You will have a successful language when you can define seesaw as an abstraction in terms of the other primitives *within the language*}\footnote{And indeed we *did* define it, at least for the RPS example.}
-
-\item Constructs that can time out could have a single continuation (do this next) or two continuations: do this if progressed normally, and do the other continuation if timed out.
-A candidate for this is a timed out \haskellinline{When}, where the second option is what we have implemented.
-\item \haskellinline{RedeemCC} is intended for premature repayment of a cash commitment, i.e.
-before the commitment has timed out.
-Why is \haskellinline{RedeemCC} a separate construct?
-Can't it be implemented with \haskellinline{Pay}?%\footnote{It is for someone to claim their cash back, rather than making a payment to another person.}
- This is potentially dangerous -- leading to a \haskellinline{FailedPay}, for example -- but useful to make a repayment at the end of a contract. 
-\item Alternatively, it would be possible for the underlying model to repay all committed but unspent cash automatically at the end of a contract, but we might want to do the same thing at the end of each ``round'' of  a game, for instance.
-\end{enumerate}
-
-\section{Response to ``Some Thoughts on Smart Contracts''}
-
-The contracts implemented in the DSL language have an implicit interface defined for what would correspond to the client side of a Dapp, as shown by the interactive interface.
-
-We have designed the contract DSL assuming that each contract would be compiled into
-\begin{itemize} 
-\item one or more smart contracts (potentially written in Plutus) and 
-\item a client side program or interface that has a limited number of choices to be made at each point in time (including an interface for the observables). 
-\end{itemize}
-
-Of course, we probably do not want to give a user a programmatic interface but an actual application with user interface.
-So there is still some work to be done to create interfaces on top of the client side of the contracts.
-
-About cost, there are two types of cost (at least):
-
-\begin{itemize} 
-\item The cost of a single transaction execution (a single Plutus validator).
-\item The cost of the whole contract, which may consists of many transactions.
-\end{itemize}
-
-Even in Ethereum this distinction is important because single contract execution is limited by gas (and is thus finite), whereas the life of the whole contract may be unbounded and consist of an indefinite number of transactions.
-
-So far we are assuming contracts defined in the DSL to be finite in both aspects. They could be infinite if we were to define them -- co-inductively -- as recursive members of the DSL type.
-
-\section{Related work}
-
-\begin{itemize}
-\item Original SPJ paper~\cite{PeytonJones:2000}, and multi-party successor. Original Szabo paper?
-\item Findel~\cite{findel}
-\end{itemize}
+We look next at an example of Marlowe in action.
 
@@
 
-In this section, we explain how the DSL can be used for designing a contract that incentivises saving. We use the 
+In this section, we explain how Marlowe can be used for designing a contract that incentivises saving. We use the 
 following example: ``participant 2'' offers 20 ADA to ``participant 1'' in exchange for saving 100 ADA for 100 blocks. 
@@
 \end{haskellcode}
+In the next section we look again at the model and the Marlowe language.
 
-\subsection{Response to comments on the original document made by Charles on Slack}
+\section{Reflection}
+We have given a concrete model and DSL, namely Marlowe version 1.2. In this section we examine how the model can be taken further, and reflect on different possible choices that might be made.
+
+\subsection{Potential additions to the model}
+
+The model that we have discussed so far is deliberately kept as compact as possible. However there are a number of potential extensions to the model which we introduce here.
+
+\paragraph{Value commitments}
+
+We have examined a single kind of commitment, namely a commitment of a certain amount of currency. In different 
+applications, such as on-chain game playing of games like poker, a commitment to a value will be required. A value 
+commitment would include a choice of a play in a game, which would be chosen and concealed from the other players until 
+a later point in play, when it can be revealed. The paper \cite{kumaresan2015use} makes use of this and other 
+cryptographic primitives to enforce the rules of poker while maintaining the cards of each player secret.
+
+\paragraph{A monadic DSL}
+
+Haskell is a pure functional programming language, and uses the \haskellinline{Monad} type class to encapsulate operations that can potentially have a side effect.
+Many introductions to monads exist \sjtnote{cite some} but for the sake of this discussion it is enough to think of a value of type \haskellinline{m a} where \haskellinline{m} is a monad as a ``computation that results in something of type \haskellinline{a}''.
+The \haskellinline{Monad} API allows such computations to be assembled, and then once assembled they may be run, in general causing some side-effects as a part of the computation.
+
+We can make the \haskellinline{Contract} type monadic, so that new, unique identifiers are automatically generated.
+This gives two nice properties.
+First, they are guaranteed to be unique; secondly they can be generated only in cases where a commit has been made successfully (within the required time period).
+The
+\haskellinline{do} notation supported by monads allows the generated names to be captured by the \haskellinline{bind} operation and referred to later in the computation.
+
+\subsection{Potential model analyses}
+
+Given that we have a language and a semantics for it, we can examine its properties.
+These might be properties of individual runs of a particular contract, of all runs of a given contract, or indeed properties of all the contracts of the language.
+In this section we look at two particularly obvious questions, but these are not the only ones that we might ask.
+
+\paragraph{What are the valid contracts?}
+
+One possible definition is ``those contracts that do not lead to a \haskellinline{FailedPay} action''.
+Given the mechanisms for making and using commitments are limited by timeouts, it should be possible to write contracts that have this property, perhaps with some assumptions about the value of observables.
+Looking at this in a little more detail, we would expect the following points to hold.
+\begin{itemize}
+\item For a \haskellinline{Pay} contract a sufficient balance not to produce a \haskellinline{FailedPay} or a \haskellinline{DuplicateRedeem} action is required.. 
+\item For \haskellinline{Both} we require a sufficient balance for both contracts combined.
+Need to think here whether the two conjuncts should be independent, or whether a commitment in one half can be used in the other.
+\item For \haskellinline{Choice} we require a sufficient balance for the maximum of the requirements of the individual contracts. 
+\item For \haskellinline{When}, need to ``age'' the existing commitments relative to the new starting point of the continuation contract. 
+\item Are there any constraints on  \haskellinline{CommitCash} and \haskellinline{RedeemCC} statements?
+Not directly, but there need to be sufficient funds available to pay if needed, or enough time for the commitments to time out, if not.
+\end{itemize}
+
+\paragraph{Do all contracts potentially evaluate to \haskellinline{Null}?}
+
+That is, for each contract do there exist sequences of values of \haskellinline{Commit} and \haskellinline{OS} such that stepping the contract with these inputs leads to \haskellinline{Null}?
+Some relevant points include:
+\begin{itemize}
+\item For \haskellinline{When}, need to be able to make \haskellinline{interpretObs st obs os} true.
+Since there is a timeout to \haskellinline{When} then there's no need to impose this condition, as we'll progress at timeout if not before.
+\item As we have put timeouts on to \haskellinline{CommitCash} and \haskellinline{RedeemCC} operations, there are no remaining requirements on these constructs for contract execution to make progress, albeit eventually.
+\item We need to check the status of \haskellinline{Observations}: what if an observation is unsatisfiable?
+\item There are no constraints on other constructs.
+\end{itemize}
+
+\subsection{Alternatives}
+
+The model outlined here raises a number of questions, enumerated here.
+
+\begin{enumerate}
+\item Does the \haskellinline{Both} constructor potentially produce invalid/misleading contracts, combining a commitment in one half with its use in the other?
+The problem is that the ``use'' conjunct could potentially be executed, leading to a failed payment.
+\item How do we ensure that contracts can be enforced? Is there a notion of validity for contracts that can be enforced? To which extent is this notion enforced by current constructors?
+\item The model does not take time as an observable, but rather it is defined in terms of \haskellinline{BlockNumber} only. 
+\item We could use ``slot time'' https://eprint.iacr.org/2016/889.pdf ``We consider a setting where time is divided into discrete units called slots.
+A ledger, described in more detail below, associates with each time slot (at most) one ledger block.''
+\item Is it possible for block number to depend on progress of computation?
+We conjecture not, because otherwise it would be possible to produce timelocks, under which it becomes impossible for time to progress.
+\item Does a contract expire if a surrounding commitment does? We think not.
+\item The model is deterministic, with all non-determinism being external, and confined to observables.
+This is surely necessary in order to support computation replay, but does it restrict the expressibility of the model?
+\item Is there a notion of causality for these contracts? Would we be able to characterise it semantically? Syntactically?
+\item We have modelled cash as fungible: we make payments using the cash commitments in order of expiry time, with those expiring earliest being used first.
+Would there be any reason for using a different approach?
+Yes!
+UTxO.
+\item Timeouts are expressed in absolute time: presumably they could be relative (with recording on the blockchain of the times at which they are triggered).
+%\footnote{This contradicts what you've said elsewhere.
+I think timeouts should be times, not blockchain numbers.
+I suspect it would be better not to use blockchain numbers.
+I don't want to have to react more quickly if lots of people are making transactions that day.
+Also, it avoids question 8.}\footnote{See addition about slot time from Ouroboros paper (as referenced in the body of the paper).}
+\item Some observables have a ``monotonicity'' property: once a choice is made, it is fixed forever; on the other hand, something not yet chosen is only so thus far; do we need to reflect this in the observations that we allow?
+%\item The seesaw construction is key to implementing e.g.
+poker, and it can be described in terms of the existing constructs (modulo the answer to point 4 above).
+However, should seesaw be a primitive in its own right?
+%\footnote{No.
+You will have a successful language when you can define seesaw as an abstraction in terms of the other primitives *within the language*}\footnote{And indeed we *did* define it, at least for the RPS example.}
+
+\item Constructs that can time out could have a single continuation (do this next) or two continuations: do this if progressed normally, and do the other continuation if timed out.
+A candidate for this is a timed out \haskellinline{When}, where the second option is what we have implemented.
+\item \haskellinline{RedeemCC} is intended for premature repayment of a cash commitment, i.e.
+before the commitment has timed out.
+Why is \haskellinline{RedeemCC} a separate construct?
+Can't it be implemented with \haskellinline{Pay}?%\footnote{It is for someone to claim their cash back, rather than making a payment to another person.}
+ This is potentially dangerous -- leading to a \haskellinline{FailedPay}, for example -- but useful to make a repayment at the end of a contract. 
+\item Alternatively, it would be possible for the underlying model to repay all committed but unspent cash automatically at the end of a contract, but we might want to do the same thing at the end of each ``round'' of  a game, for instance.
+\end{enumerate}
+
+\section{Response to ``Some Thoughts on Smart Contracts''}
+
+The contracts implemented in the DSL language have an implicit interface defined for what would correspond to the client side of a Dapp, as shown by the interactive interface.
+
+We have designed the contract DSL assuming that each contract would be compiled into
+\begin{itemize} 
+\item one or more smart contracts (potentially written in Plutus) and 
+\item a client side program or interface that has a limited number of choices to be made at each point in time (including an interface for the observables). 
+\end{itemize}
+
+Of course, we probably do not want to give a user a programmatic interface but an actual application with user interface.
+So there is still some work to be done to create interfaces on top of the client side of the contracts.
+
+About cost, there are two types of cost (at least):
+
+\begin{itemize} 
+\item The cost of a single transaction execution (a single Plutus validator).
+\item The cost of the whole contract, which may consists of many transactions.
+\end{itemize}
+
+Even in Ethereum this distinction is important because single contract execution is limited by gas (and is thus finite), whereas the life of the whole contract may be unbounded and consist of an indefinite number of transactions.
+
+So far we are assuming contracts defined in the DSL to be finite in both aspects. They could be infinite if we were to define them -- co-inductively -- as recursive members of the DSL type.
+
+\section{Response to comments on the original document made by Charles on Slack}
 

### 2018-03-24 tex/paper.tex
completed first draft to the end of the current section 6

@@
 \end{haskellcode}
+In the next section we cover the online contract visualisation and interaction tool for Marlowe.
+
+\section{Visualising the contract}
+
+\sjtnote{Pablo: can you add something here about the online visualisation, including one or more suitable screenshots?
+It might be worth giving a condensed screenshot of the whole screen, and then more detailed ones of parts of the display.}
+
+\sjtnote{Should it have a name?}
+
 In the next section we look again at the model and the Marlowe language.
@@
 \section{Reflection}
-We have given a concrete model and DSL, namely Marlowe version 1.2. In this section we examine how the model can be taken further, and reflect on different possible choices that might be made.
+We have given a concrete model and DSL, namely Marlowe version 1.2.
+In this section we examine how the model can be extended, how contacts can be analysed, reflect on different design decisions that might be made and discuss some aspects of the implementation on Marlowe within a blockchain environment.
 
@@
 
-\subsection{Alternatives}
+\subsection{Questions and alternatives}
 
-The model outlined here raises a number of questions, enumerated here.
+The model outlined here raises a number of questions, or suggests a number of alternative choices.
 
-\begin{enumerate}
-\item Does the \haskellinline{Both} constructor potentially produce invalid/misleading contracts, combining a commitment in one half with its use in the other?
+\paragraph{Time}
+
+The model does not take time as an observable, but rather it is defined in terms of \haskellinline{BlockNumber} only.
+We could  use ``slot time'' instead: \emph{``We consider a setting where time is divided into discrete units called slots.
+A ledger, described in more detail below, associates with each time slot (at most) one ledger block.''}\footnote{\url{https://eprint.iacr.org/2016/889.pdf}}
+
+\paragraph{Timelocks}
+
+In systems where time is significant, it is possible to build ``timelocks'' under which 
+it becomes impossible for time to progress.
+Is it possible for block number to depend on progress of computation?
+We conjecture not, because otherwise it would be possible to produce timelocks, under which it would become impossible for time to progress.
+
+\paragraph{Relative time}
+
+Timeouts are expressed in absolute time: presumably they could be relative (with recording on the blockchain of the times at which they are triggered).
+
+\paragraph{Determinism}
+
+The model is deterministic, with all non-determinism being external, and confined to observables.
+This is surely necessary in order to support computation replay, but does it restrict the expressibility of the model?
+
+\paragraph{Causality}
+
+Is there a notion of causality for these contracts? Would we be able to characterise it semantically? Syntactically?
+
+\paragraph{Expiry}
+
+Does a contract expire if a surrounding commitment does? We have not made this a feature of the model.
+
+\paragraph{Premature repayment}
+
+The \haskellinline{RedeemCC} construct is intended for premature repayment of a cash commitment, i.e.
+before the commitment has timed out.
+Alternatively, it would be possible for the underlying model to repay all committed but unspent cash automatically at the end of a contract (but before all commitments have expired), but we might want to do the same thing at the end of each ``round'' of  a game, for instance.
+
+\paragraph{Fungibility}
+
+We have modelled cash as fungible: we make payments using the cash commitments in order of expiry time, with those expiring earliest being used first.
+Building a model on UTxO might be a reason for using a different approach.
+
+\paragraph{The \haskellinline{Both} construct}
+
+Does the \haskellinline{Both} constructor potentially produce invalid/misleading contracts, combining a commitment in one half with its use in the other?
 The problem is that the ``use'' conjunct could potentially be executed, leading to a failed payment.
-\item How do we ensure that contracts can be enforced? Is there a notion of validity for contracts that can be enforced? To which extent is this notion enforced by current constructors?
-\item The model does not take time as an observable, but rather it is defined in terms of \haskellinline{BlockNumber} only. 
-\item We could use ``slot time'' https://eprint.iacr.org/2016/889.pdf ``We consider a setting where time is divided into discrete units called slots.
-A ledger, described in more detail below, associates with each time slot (at most) one ledger block.''
-\item Is it possible for block number to depend on progress of computation?
-We conjecture not, because otherwise it would be possible to produce timelocks, under which it becomes impossible for time to progress.
-\item Does a contract expire if a surrounding commitment does? We think not.
-\item The model is deterministic, with all non-determinism being external, and confined to observables.
-This is surely necessary in order to support computation replay, but does it restrict the expressibility of the model?
-\item Is there a notion of causality for these contracts? Would we be able to characterise it semantically? Syntactically?
-\item We have modelled cash as fungible: we make payments using the cash commitments in order of expiry time, with those expiring earliest being used first.
-Would there be any reason for using a different approach?
-Yes!
-UTxO.
-\item Timeouts are expressed in absolute time: presumably they could be relative (with recording on the blockchain of the times at which they are triggered).
-%\footnote{This contradicts what you've said elsewhere.
-I think timeouts should be times, not blockchain numbers.
-I suspect it would be better not to use blockchain numbers.
-I don't want to have to react more quickly if lots of people are making transactions that day.
-Also, it avoids question 8.}\footnote{See addition about slot time from Ouroboros paper (as referenced in the body of the paper).}
-\item Some observables have a ``monotonicity'' property: once a choice is made, it is fixed forever; on the other hand, something not yet chosen is only so thus far; do we need to reflect this in the observations that we allow?
-%\item The seesaw construction is key to implementing e.g.
-poker, and it can be described in terms of the existing constructs (modulo the answer to point 4 above).
-However, should seesaw be a primitive in its own right?
-%\footnote{No.
-You will have a successful language when you can define seesaw as an abstraction in terms of the other primitives *within the language*}\footnote{And indeed we *did* define it, at least for the RPS example.}
 
-\item Constructs that can time out could have a single continuation (do this next) or two continuations: do this if progressed normally, and do the other continuation if timed out.
-A candidate for this is a timed out \haskellinline{When}, where the second option is what we have implemented.
-\item \haskellinline{RedeemCC} is intended for premature repayment of a cash commitment, i.e.
-before the commitment has timed out.
-Why is \haskellinline{RedeemCC} a separate construct?
-Can't it be implemented with \haskellinline{Pay}?%\footnote{It is for someone to claim their cash back, rather than making a payment to another person.}
- This is potentially dangerous -- leading to a \haskellinline{FailedPay}, for example -- but useful to make a repayment at the end of a contract. 
-\item Alternatively, it would be possible for the underlying model to repay all committed but unspent cash automatically at the end of a contract, but we might want to do the same thing at the end of each ``round'' of  a game, for instance.
-\end{enumerate}
+\paragraph{Monotonicity}
 
-\section{Response to ``Some Thoughts on Smart Contracts''}
+Some observables have a ``monotonicity'' property: once a choice is made, it is fixed forever; on the other hand, something not yet chosen is only so thus far; do we need to reflect this in the observations that we allow?
 
-The contracts implemented in the DSL language have an implicit interface defined for what would correspond to the client side of a Dapp, as shown by the interactive interface.
+\subsection{Towards an implementation}
 
-We have designed the contract DSL assuming that each contract would be compiled into
+The contracts implemented in the Marlowe language have an implicit interface defined for what would correspond to the client side of a distributed application, as shown by the interactive interface. 
+We have designed Marlowe assuming that each contract would be compiled into
 \begin{itemize} 
@@
 \end{itemize}
-
 Of course, we probably do not want to give a user a programmatic interface but an actual application with user interface.
@@
 
-About cost, there are two types of cost (at least):
-
+The Cardano platform has a notion of cost, and Marlowe contracts will potentially incur costs of two kinds:
 \begin{itemize} 
 \item The cost of a single transaction execution (a single Plutus validator).
-\item The cost of the whole contract, which may consists of many transactions.
+\item The cost of the whole contract, which may consist of many transactions.
 \end{itemize} 
-
 Even in Ethereum this distinction is important because single contract execution is limited by gas (and is thus finite), whereas the life of the whole contract may be unbounded and consist of an indefinite number of transactions.
-
-So far we are assuming contracts defined in the DSL to be finite in both aspects. They could be infinite if we were to define them -- co-inductively -- as recursive members of the DSL type.
+So far we are assuming contracts defined in Marlowe to be finite in both aspects.
+They could be infinite if we were to define them co-inductively as recursive members of the DSL type.
 

### 2018-03-25 tex/paper.tex
completed introduction and reflection section; added next steps

@@
 
-This paper explores the design of a domain specific language, Marlowe,\footnote{Marlowe is named after Christopher Marlowe, the Elizabethan poet, dramatist and spy, who was born and educated in Canterbury: \url{https://en.wikipedia.org/wiki/Christopher_Marlowe}}\footnote{The code for Marlowe is available from \url{https://github.com/input-output-hk/scdsl}
+This paper explores the design of a domain specific language, Marlowe,\footnote{Marlowe is named after Christopher Marlowe, the Elizabethan poet, dramatist and spy, who was born and educated in Canterbury, which is also where our university is based.
+\url{https://en.wikipedia.org/wiki/Christopher_Marlowe}}\footnote{The code for Marlowe is available from \url{https://github.com/input-output-hk/scdsl}
 } targetted at the execution of financial contracts in the style of Peyton Jones, Eber and Seward~\cite{PeytonJones:2000} on blockchains.
 This work is a part of the Cardano project, and is supported by IOHK.\footnote{\url{https://iohk.io}}
+In doing this we are required to refine the model of contracts in a number of ways in order to fit with a radically different context.
 
-In doing this we are required to refine the model of contracts in a number of ways in order to fit with a radically different context. 
-\begin{itemize}
-\item In the traditional model, enforcement of the contract is the responsibility of the legal system; in the decentralised blockchain model, in which there is no central authority, the contract needs to be enforced by design.
-\item There is not a single blockchain model, and we give a brief overview of the different mechanisms for scripting on blockchain (although this is not the primary purpose of this article). 
-\begin{itemize}
-\item
+In the traditional model, enforcement of the contract is the responsibility of the legal system; in the decentralised blockchain model, in which there is no central authority, the contract needs to be enforced by design.
+This means that we require participants to commit currency (or ``cash'' as we term it) to cover all possible expenditure in advance of the contract executing; we also have to use timeouts to ensure that the contract does not become stuck even if one or more of the participants stop interacting with it.
+
+Moreover, there is not a single blockchain model, and we give a brief overview of the different mechanisms for scripting on blockchain (although this is not the primary purpose of this article). 
 A particularly important technical distinction here is that between an account-style model as used by Ethereum~\cite{EthereumRationale}, and the unspent transaction output (UTxO) model of bitcoin~\cite{sok}.
-\item There is the question of how precisely the execution of the contract interacts with the blockchain, and how information is to be recorded on the chain.
-\item Does the execution ``push'' for changes  to happen, or do external actors have to ``pull'' to effect changes enabled by execution of the contract?
-\end{itemize}
-\item \sjtnote{What else do we want to say here?}
-\end{itemize}
+There is also the question of how precisely the execution of the contract interacts with the blockchain, and how information is to be recorded on the chain.
+Finally, should the execution ``push'' for changes  to happen, or do external actors have to ``pull'' to effect changes enabled by execution of the contract?
 
-\noindent
-\sjtnote{Preliminary overview; need to revisit.}
-The paper is structured like this: we begin with an overview of the model, including the assumptions made in designing 
+The paper describes version 1.2 of Marlowe, and it should be seen as a work in progress. We have designed it to provide financial contracts in Cardano, but also in order to start to answer the twin
+ questions of what smart contracts will look like in Cardano, and how those contracts will be implemented in the overall system.
+
+The paper is structured like this: we begin in Section \ref{section:design} with a discussion of the design rationale for Marlowe, including an overview of the various approaches to scripting in blockchain as well as the different designs of the blockchains themselves.
+We conclude Section \ref{section:design} with a set of decisions about the blockchain environment for Marlowe.
+The central part of the paper is Section \ref{section:model} where we introduce the Marlowe model, including the assumptions made in designing
 it, the types of the principal functions and a description of the DSL as an algebraic type, constructor by constructor.
-The section aims to give a high level overview, and some of the concrete details are to be found in the repo. We follow 
-this with discussions of the kinds of formal analysis which might be made for the DSL, as well as discussing a  set of 
-questions raised by the process of modelling. We then look at the example of ``incentivised saving'', as an 
-illustration, and then look at how Marlowe relates to other work in the area. We conclude by pointing out
-
-\sjtnote{To talk about, in no particular order \ldots}
-\begin{itemize}
-\item
-Marlowe as the ``canary in the coalmine'': test out how it will work, how it will interface with
-other aspects of the Cardano platform
-\item
-Be clear that this paper describes version 1.2 of Marlowe.
-\item 
-Enforcement and promptness.
-\end{itemize}
+Section \ref{section:example} presents an example of a larger Marlowe contract, for incentivised saving, and Section \ref{section:tool} introduces our tool for visualising and interacting with Marlowe contracts.
+Section \ref{section:reflection} reflects on the system design, and observes some alternative designs and Section \ref{section:next-steps} enumerates the next steps for the project.
 
 \section{Design rationale}
+\label{section:design}
 
@@
 
-The Findel~\cite{findel} project examines financial contracts on the Ethereum platform, and the authors note that  payments need to be bounded; this is made concrete in our account by our notion of commitments.
+The Findel project~\cite{findel} examines financial contracts on the Ethereum platform, based on the seminal~\cite{PeytonJones:2000}, and the authors note that payments need to be bounded; this is made concrete in our account by our notion of commitments.
 They take no account of commitments or timeouts as we do below, but it should be noted that the Ethereum platform is more powerful than the one that we target in this paper.
@@
 \section{The model}
+\label{section:model}
 
 The DSL is modelled as an algebraic type in Haskell, together with an executable small-step semantics. 
-We start by looking at the different types used by the model, and then look at the \haskellinline{Contract} DSL itself. \sjtnote{To complete this introduction}
+We start by looking at the different types used by the model, then we look at the \haskellinline{Contract} DSL itself, and finally we give its semantics in Haskell.
+Section \ref{section:example} gives an example of Marlowe in action.
 
@@
 
+As the examples illustrate, observables come both from aspects of the blockchain (e.g.\ current block number) and externally.
+In the latter case it will be necessary to agree a trusted oracle or beacon giving the value, which can then be stored (or a hash stored) on chain.
+
 \paragraph{Inputs}
@@
 After the timeout period the cash can be refunded through the user creating a transaction to reclaim the cash.
+Information about the commitments currently in force forms the \haskellinline{State}, which is (potentially) modified at each execution step.
 
-Moreover, information about the commitments currently in force forms the \haskellinline{State}, which is (potentially) modified at each execution step.
-
-Finally, payments can be granted by using committed money, but they must be manually redeemed by the recipient, in the 
+Payments can be granted by using committed money, but they must be manually redeemed by the recipient, in the 
 same way that cash commitments are redeemed when they expire. The effects of the contract in the blockchain are 
-represented by a list \haskellinline{AS} of \haskellinline{Action}s that is derived from the execution of each step of the semantics.
+represented by a list \haskellinline{AS} of \haskellinline{Actions} that is derived from the execution of each step of the semantics.
 
@@
 \section{Incentivised saving}
+\label{section:example}
 
-In this section, we explain how Marlowe can be used for designing a contract that incentivises saving. We use the 
+In this section, we show how Marlowe can be used for designing a contract that incentivises saving. We use the 
 following example: ``participant 2'' offers 20 ADA to ``participant 1'' in exchange for saving 100 ADA for 100 blocks. 
@@
 
-\section{Visualising the contract}
+\section{Visualising and interacting with contracts}
+\label{section:tool}
 
@@
 
-In the next section we look again at the model and the Marlowe language.
+In the next section we re-examine the model and the Marlowe language, looking at possible extensions, alternative design decisions and ways in which we can formally analyse Marlowe contracts.
 
 \section{Reflection}
+\label{section:reflection}
+
 We have given a concrete model and DSL, namely Marlowe version 1.2.
@@
 \subsection{Potential model analyses}
+\label{section:analysis}
 
@@
 
-\subsection{Towards an implementation}
+\subsection{Towards an implementation in Cardano}
+\label{section:towards-implementation}
+
+\sjtnote{I have expanded this to include a number of responses to Charles' comments on an earlier document. Other responses appear in the previous sub-section.}
 
 The contracts implemented in the Marlowe language have an implicit interface defined for what would correspond to the client side of a distributed application, as shown by the interactive interface.
+
+\paragraph{Compilation}
+
 We have designed Marlowe assuming that each contract would be compiled into
@@
 So there is still some work to be done to create interfaces on top of the client side of the contracts.
+
+\paragraph{Cost}
 
@@
 
-\section{Response to comments on the original document made by Charles on Slack}
+\paragraph{Recording information on chain}
 
-Legend: {\it a quote from this document}, Charles' response, {\bf our reply}:
+Commitments and values of observables need to be recorded somehow, and this raises the potential issue about bloating the chain with data.
+In general it will only be necessary to store a signed hash value of the observable, and this will keep data usage bounded.
 
-{\it
-First, it will need to observe different kinds of varying quantities including, for example, the current time, the current block number and, random numbers, as well as ``real world'' quantities like ``the price of oil'' or ``the exchange rate between currencies A and B''.
-Each instance of such an observable will be recorded on the blockchain to allow the computation to be repeated for verification purposes.
-It is assumed that at each step of the execution of the contract, the values of some observables will be needed.
-}
+\paragraph{General computations}
 
-We need to get a better notion of observation.
+If we have contracts that involve, for example, the oil price then we may need to convert a published price in USD into ADA, using a value for the prevailing USD/ADA exchange rate.
+This will require a computation: in this case, a multiplication.
+If Marlowe is to be stand-alone then we need to be extend it with arithmetic and other operations, but we expect Marlowe to be embedded in Plutus (Core) and so to be able to use the facilities of the host language here.
 
-There seem to be two broad categories of events worth observing. One that is derived from the network itself and one that is from an external source.
+\paragraph{Sidechains and wallets}
 
-For example, we could consider a game where the contract executes based upon the value of a block header, this is something known after it occurs and cannot be changed.
-Whereas the price of oil is not.
+The implementation may split contract execution between the main chain and a sidechain (or chains). This will allow links from Marlowe to work on securing multiparty sidechain games/contracts (MPCs).
 
-{\bf
-Indeed, we agree.
-The former kind of observations are already artefacts on the blockchain, whereas the latter need to be stored on the blockchain.
-We have assumed that the infrastructure for the language supports this recording: we would not expect language users to have to do this for themselves.
-}
+\section{Next steps}
+\label{section:next-steps}
 
-{\bf
-Apart from this difference in the infrastructure, we see no difference between the two different sorts of observations: is this correct?
-}
+We plan to continue the work with Marlowe in a number of directions. First, we need to sign off a definition of the language, as the current status of Marlowe 1.2 is that of a proposal.
 
-The question of commitment to blockchain is interesting as who should have write access and also how do we prevent an enormous amount of bad data from being posted? And is it necessary at all?
+\paragraph{Context}
 
-We could specify the format of an answer and have them commit it to the blockchain as a signed hash and then it could be externally shared for example and everyone would be able to determine it is correct.
+We should decide on the final form of the language -- for example, should it be monadic or not -- and the relation of Marlowe to Plutus (Core).
+In particular, answering the latter question will require work on the current redeemer/validator model of transactions in Plutus Core.
 
-{\bf
-We saw that commitments were an integral part of the game playing process, and so we wanted to include them as primitives in the language.
-Yes, they are usually implemented as hashes, but we wanted to give a more abstract presentation here.
+\paragraph{Implement in Cardano}
 
-Concerning bad data, a commitment will usually just store the hash, but the problem seems to come at the reveal, when all the data has to be shown, somehow.
-We don't have to store the whole data (just the hash) but it does have to be revealed in some way that can be observed, and the naive interpretation of this is that it is made available somehow, thereby using storage on the blockchain.
+An immediate goal is to implement Marlowe in Cardano, noting the observations of Section \ref{section:towards-implementation} and to deploy it in a test network to observe and measure its behaviour in practice.
 
-There may be other ways of doing this. Is there some sort of compact proof that can be used here?
+\paragraph{Compilation}
 
-Or can we devolve to a trusted third party, off chain?
-}
+Marlowe contracts could be generated by a compilation process from `vanilla' financial contracts.
+This also suggests that models with off-chain enforcement could be investigated too, with an MPC foundation.
 
-I'm also curious about expectations. For example, say we have an oil contract.
+\paragraph{Validation and Verification}
 
-We have an expectation of exchange rates or something like that so we'd expect some int to float and then a float to float. Examples:
+Some degree of validation is provided by the visualisation tool presented in Section \ref{section:tool}, albeit for smaller-scale contracts, but there is a general question of how properties of contracts can be inferred.
+One relatively low-cost is to use QuickCheck-style property-based testing~\cite{quickCheck}, and properties developed here become candidates for fully fledged verification in a formalisation of the (semantics of) Marlowe.
 
-   * 1 barrel for 120.55 dollars.
-   * 22 dollars for 1 ada.
+\paragraph{Analysis}
 
-Gives us everything we need to act on the contract accordingly in a deterministic way.
+As we noted in Section \ref{section:analysis} we can develop analyses of various aspects of Marlowe contracts.
+While these might initially be ``pencil and paper'' analyses, we could expect these to be based on a formalisation of the language and its semantics.
 
-First, we guess these may be fixed or rational rather than float :-) We would assume that exchange rates are observables, with values recorded at the point of use, so that re-computation is possible (and deterministic).
+\paragraph{Formalisation}
 
-{\bf
-In general, we need a more powerful expression language (e.g. through embedding the DSL). Maybe we could use a subset of Plutus.
-}
+Marlowe and its semantics could be formalised in a theorem prover such as Isabelle~\cite{IsabelleHOL}, which has been used in the formalisation of language semantics.
+A formalisation could provide a sanity check on the language definition, as well as formalise proofs of important properties, based on analysis of the contract language.
 
-{\it
-Second, it will be necessary to ensure that commitments of various kinds have been made by the participants in a contract.
-}
+\section{Acknowledgements}
 
-Ok, Bernardo knows a lot more about this than I do. He's been doing this type of stuff with his MPC work. I'd recommend having a discussion with Aggelos and Bernardo about
-commitments in the cryptocurrency world.
+\sjtnote{Have I forgotten anyone?}
 
-{\bf
-Yes, certainly, we will do this.
-}
+Thanks to IOHK staff Manuel Chakravarty, Duncan Coutts, Bernardo David, Charles Hoskinson, Aggelos Kiyakis, Darry McAdams, Bruno Woltzenlogel Paleo, and Phil Wadler for their collaboration, and to colleagues from the Universities of Kent and Leicester for their comments on previous presentations of this work.
 
-{\it
-What are the valid contracts?
-One possible definition is ``those contracts that do not lead to a FailedPay action''.
-Does this make sense independently of the inputs to a contract (i.e.
-observables, commitments, redeems, time, and randomness)?
-}
-
-I tend to think of compositions.
-So we have things like multisig where the players setup a special contract tailored to a set of business and trust assumptions. This game ends when funds are spent.
-Complexity could be fairly arbitrary as the conditions to spend could be quite complex and involve lots of business logic. The end of the game could initiate a new game.
-
-{\bf
-A game that starts a new game is essentially a bigger game.
-There seems to be a limitation about commitments.
-Commitments need to be voluntary by nature, so if there are no commitments inherited between the two games, then it should be no different to model both the games separately.
-
-There should not be a problem to model infinite games either by using co-induction.
-}
-
-For example creating a new multisig account with the same logic for the change address.
-
-Then we have event based games, where actions occur when something happens, a triggering event.
-
-For example a deadman switch.
-
-If bob doesn't sign a special transaction each month, then send all the money to alice.
-
-These are at the mercy of the oracle connected to the contract.
-
-Then we have games of interaction, such as poker for example.
-
-Bob and Alice have earned the right to interact with each other with special rules and pending how those interactions end up, we have a transfer of value between them and others.
-
-Abstractly, I suppose there is a notion of entry condition, allowed actions and exit conditions.
-
-{\bf
-Not sure what entry and exit conditions are, but commitments may well be entry conditions, or at least a crucial component of them.
-We think that all the examples described can be implemented in the DSL (assuming we add the proper observables and oracles).
-Maybe we need some more cryptographic primitives for poker, that is probably something we should discuss with Aggelos and Bernardo.
-}
-
-{\it
-Null is the null contract; it will always be quiescent.
-}
-
-Just to clarify, null means termination or just dormancy?
-
-{\bf
-Both. A contract has terminated if it has evolved to be exactly null. But there exist contracts that contain the null contract but have not terminated.
-}
-
-{\it
-Do all contracts potentially evaluate to Null?
-That is, for each contract do there exist sequences of values of Commit and OS such that stepping the contract with these inputs leads to Null?
-Might we have to rule out unsatisfiable observations?
-}
-
-If I understand null to mean the contract has finished all possible execution, then this is a reasonable assumption given a certain viewpoint completion of a multisig contract for example could be interpreted as a null case.
-
-But in a strict sense the answer is no, there can be cases where something is in a pending state indefinitely.
-For example, for a 2 of 3 contract, two keyholders could die then you'd have an active contract waiting for completion, but it will never happen.
-
-{\bf
-I understand this as: all contracts should reduce to null, but not independently of the observables.
-We can then think about analysing which observables stall the contract (which sounds like a boolean satisfiability problem).
-Maybe we just want to guarantee that there is a set of external actions that will cause the contract to reduce to null no matter its state.
-
-In the particular example of death of participants, we would assume that the contract would have been formulated with timeouts on inactivity of participants (when activity requested).
-}
-
-What we need to do is determine what functionality can be black-boxed in this model with some degree of reliability.
-
-Bernardo's MPC library is going to provide a lot of functionality for things like games for example and provides strong cryptographic guarantees for outcomes given some initial assumptions.
-
-{\bf
-We will definitely incorporate these.
-}
-
-Second, we need to have a set of canonical actions that we could test against the proposed model. And see how they would be constructed.
-
-{\bf
-Is this like a test suite? Not sure how it would work, but would like to know more.
-}
-
-Third, we need to clearly understand what is an oracle in this model and how we intend on using them.
-
-{\bf
-Good point, we need to define an interface for oracles.
-}
-
-\clearpage
 \bibliographystyle{ACM-Reference-Format}

### 2018-03-26 tex/paper.tex
Fix some typos and add notes

@@
 
-As more than one wag has observed, many of the programs that run on blockchains are neither smart nor indeed form contracts in any real sense.
-Nonetheless, this terminology has stuck, and has become shorthand for ``programmed interactions on the blockchain''.
-In an earlier paper we presented a survey of the approaches available~\cite{cryptoeprint:2016:1156}; we give an overview here.
+As more than one wag has observed, many of the programs that run on blockchains are neither smart nor indeed form 
+contracts in any real sense. Nonetheless, this terminology has stuck, and has become shorthand for ``programmed 
+interactions on the blockchain''. In an earlier paper, we presented a survey of the approaches 
+available~\cite{cryptoeprint:2016:1156}; we give an overview here.
 
@@
 
-A domain-specific language or DSL is a high-level language designed to work in a specific field or domain.
-The intention is that because the users will know about the field, the constructs of the language can be designed to be meaningful to them, and also that because of its nature the DSL need not include all the features of a general purpose language.
-Removing this clutter and having the remaining operations directly reflect the application area is intended to make the language more accessible to domain experts who do not necessarily see themselves as programmers.
+A domain-specific language or DSL is a high-level language designed to work in a specific field or domain. The intention 
+is that because the users will know about the field, the constructs of the language can be designed to be meaningful to 
+them, and also that, because of its nature, the DSL need not include all the features of a general purpose language. 
+Removing this clutter and having the remaining operations directly reflect the application area is intended to make the 
+language more accessible to domain experts who do not necessarily see themselves as programmers.
 
@@
 
-The simplest blockchain is Bitcoin.
-Transactions on Bitcoin are made by spending the (as yet) unspent outputs of previous transactions (`unspent transaction outputs' or UTxOs).The chain need not maintain any state, such as the current value owned by any particular participant: such \emph{account} information is implicit and external to the Bitcoin blockchain.
-On the other hand, the Ethereum model explicitly keeps track of account values, and so this information needs to be kept as state information on chain.
-While it is possible to translate between UTxO-based and account-based chains \sjtnote{cite Bruno paper} the UTxO model is more fundamental, and requires less support from the implementation of the chain itself.
+The simplest blockchain is Bitcoin. Transactions on Bitcoin are made by spending the (as yet) unspent outputs of 
+previous transactions (`unspent transaction outputs' or UTxOs). The chain need not maintain any state, such as the 
+current value owned by any particular participant: such \emph{account} information is implicit and external to the 
+Bitcoin blockchain. On the other hand, the Ethereum model explicitly keeps track of account values, and so this 
+information needs to be kept as state information on chain. While it is possible to translate between UTxO-based and 
+account-based chains \sjtnote{cite Bruno paper} the UTxO model is more fundamental, and requires less support from the 
+implementation of the chain itself.
 
@@
 
-Contracts can be conceived of as acting in two ways.
-In a \emph{push} model, contracts are seen to be pro-active, and to make things happen.
-In the case of blockchain, this would include making payments or transactions happen.
-Alternatively, in the \emph{pull} model contracts enable certain things to happen, such as transactions, but the transactions need to be effected by an external player.
-The pull model makes fewer demands on the blockchain implementation than the push model, which requires the contract to have direct access to modify the chain.
+Contracts can be conceived of as acting in two ways. In a \emph{push} model, contracts are seen to be pro-active, and to 
+make things happen. In the case of blockchain, this would include making payments or transactions happen. 
+Alternatively, in the \emph{pull} model, contracts enable certain things to happen, such as transactions, but the 
+transactions need to be effected by an external player. The pull model makes fewer demands on the blockchain 
+implementation than the push model, which requires the contract to be triggered automatically.
 
@@
 
-In designing Marlowe we have chosen a number of options; this description applies to Marlowe version 1.2 \sjtnote{Is there an obvious way of making that clear in the repo?
-I really don't understand how git deals with ``releases'' or versions.}
+In designing, Marlowe we have chosen a number of options; this description applies to Marlowe version 1.2 \sjtnote{Is 
+there an obvious way of making that clear in the repo? I really don't understand how git deals with ``releases'' or 
+versions.}\plsnote{We can create a branch and point to that.}
 
@@
 This means, for example, that the language can \emph{enable} a payment to be made, but cannot make the payment happen. 
-It is up to the interested party (or parties) to make the change happen.
-It is therefore possible for the payment not to happen, despite the fact that this is against the interest of the receiving party.
-The same
+It is up to the interested party (or parties) to make the change happen. It is therefore possible for the payment not to 
+happen, despite the fact that this is against the interest of the receiving party.\plsnote{There was a trailing ``the 
+same'' here. Did you think of another example?}
 
@@
 
-The language semantics is described in terms of \emph{commitments} of sums of cryptocurrency -- which we choose to abbreviate to ``cash'' -- and so is agnostic of the underlying implementation.
-However, the current redemption mechanism in Plutus Core is not strong enough to support Marlowe, and we are actively investigating the best way to extend this mechanism to implement Marlowe in full on Plutus Core.
+The language semantics is described in terms of \emph{commitments} of sums of cryptocurrency -- which we choose to 
+abbreviate to ``cash'' or ``money'' -- and so is agnostic of the underlying implementation.  However, the current 
+redemption mechanism in Plutus Core is not strong enough to support Marlowe, and we are actively investigating the best 
+way to extend this mechanism to implement Marlowe in full on Plutus Core.
 
@@
 \end{haskellcode}
-In the next section we cover the online contract visualisation and interaction tool for Marlowe.
+In the next section, we cover the online contract visualisation and interaction tool for Marlowe.
 
@@
 
-We have given a concrete model and DSL, namely Marlowe version 1.2.
-In this section we examine how the model can be extended, how contacts can be analysed, reflect on different design decisions that might be made and discuss some aspects of the implementation on Marlowe within a blockchain environment.
+We have given a concrete model and DSL, namely Marlowe version 1.2. In this section we examine how the model can be 
+extended, how contacts can be analysed, reflect on different design decisions that might be made and discuss some 
+aspects of the implementation of Marlowe within a blockchain environment.
 
@@
 
-Haskell is a pure functional programming language, and uses the \haskellinline{Monad} type class to encapsulate operations that can potentially have a side effect.
-Many introductions to monads exist \sjtnote{cite some} but for the sake of this discussion it is enough to think of a value of type \haskellinline{m a} where \haskellinline{m} is a monad as a ``computation that results in something of type \haskellinline{a}''.
-The \haskellinline{Monad} API allows such computations to be assembled, and then once assembled they may be run, in general causing some side-effects as a part of the computation.
+Haskell is a pure functional programming language, and uses the \haskellinline{Monad} type class to encapsulate
+operations that can potentially have a side effect. Many introductions to monads exist \sjtnote{cite some} but for the 
+sake of this discussion it is enough to think of a value of type \haskellinline{m a} where \haskellinline{m} is a monad 
+as a ``computation that results in something of type \haskellinline{a}''. The \haskellinline{Monad} API allows 
+suchcomputations to be assembled, and then, once assembled, they may be run, in general causing some side-effects as a 
+part of the computation.
 
@@
 
-Given that we have a language and a semantics for it, we can examine its properties.
-These might be properties of individual runs of a particular contract, of all runs of a given contract, or indeed properties of all the contracts of the language.
-In this section we look at two particularly obvious questions, but these are not the only ones that we might ask.
+Given that we have a language and a semantics for it, we can examine its properties. These might be properties of 
+individual runs of a particular contract, of all runs of a given contract, or indeed properties of all the contracts of 
+the language. In this section, we look at two particularly obvious questions, but these are not the only ones that we 
+might ask.
 
@@
 \begin{itemize}
-\item For a \haskellinline{Pay} contract a sufficient balance not to produce a \haskellinline{FailedPay} or a \haskellinline{DuplicateRedeem} action is required.. 
+\item For a \haskellinline{Pay} contract a sufficient balance not to produce a \haskellinline{FailedPay}.
 \item For \haskellinline{Both} we require a sufficient balance for both contracts combined.
@@
 \item For \haskellinline{Choice} we require a sufficient balance for the maximum of the requirements of the individual contracts. 
-\item For \haskellinline{When}, need to ``age'' the existing commitments relative to the new starting point of the continuation contract. 
-\item Are there any constraints on  \haskellinline{CommitCash} and \haskellinline{RedeemCC} statements?
-Not directly, but there need to be sufficient funds available to pay if needed, or enough time for the commitments to time out, if not.
+\item For \haskellinline{When}, need to ``age'' the existing commitments relative to the new starting point of the 
+continuation contract.
+\item \haskellinline{RedeemCC} should not happen twice in any execution, so as to avoid producing a
+\haskellinline{DuplicateRedeem} action.
+\item Are there any constraints on \haskellinline{CommitCash} statements? Not directly, but there need to be sufficient 
+funds available to pay if needed, and enough time for the commitments to time out.
 \end{itemize}
@@
 \begin{itemize}
-\item For \haskellinline{When}, need to be able to make \haskellinline{interpretObs st obs os} true.
-Since there is a timeout to \haskellinline{When} then there's no need to impose this condition, as we'll progress at timeout if not before.
+\item For \haskellinline{When}, we need to be able to make \haskellinline{interpretObs st obs os} true. Since there is 
+a timeout to \haskellinline{When} then there is no need to impose this condition, as we will progress at timeout if not 
+before.
 \item As we have put timeouts on to \haskellinline{CommitCash} and \haskellinline{RedeemCC} operations, there are no remaining requirements on these constructs for contract execution to make progress, albeit eventually.
@@
 
-Does the \haskellinline{Both} constructor potentially produce invalid/misleading contracts, combining a commitment in one half with its use in the other?
-The problem is that the ``use'' conjunct could potentially be executed, leading to a failed payment.
+Does the \haskellinline{Both} constructor potentially produce invalid/misleading contracts, combining a commitment in 
+one half with its use in the other? The problem is that the ``use'' conjunct could potentially be executed, leading to 
+a failed payment. \plsnote{What do you mean with the ``use'' conjunct?}
 
@@
 
-Commitments and values of observables need to be recorded somehow, and this raises the potential issue about bloating the chain with data.
-In general it will only be necessary to store a signed hash value of the observable, and this will keep data usage bounded.
+Commitments and values of observables need to be recorded somehow, and this raises the potential issue about bloating 
+the chain with data. In general, it will only be necessary to store a signed hash value of the observable, and this will 
+keep data usage bounded.
 
@@
 
-Thanks to IOHK staff Manuel Chakravarty, Duncan Coutts, Bernardo David, Charles Hoskinson, Aggelos Kiyakis, Darry McAdams, Bruno Woltzenlogel Paleo, and Phil Wadler for their collaboration, and to colleagues from the Universities of Kent and Leicester for their comments on previous presentations of this work.
+Thanks to IOHK staff Manuel Chakravarty, Duncan Coutts, Bernardo David, Charles Hoskinson, Aggelos Kiyakis, Darryl 
+McAdams, Bruno Woltzenlogel Paleo, and Phil Wadler for their collaboration, and to colleagues from the Universities of 
+Kent and Leicester for their comments on previous presentations of this work.
 

### 2018-03-26 tex/paper.tex
small mods.

@@
 
-\noindent
-\sjtnote{This is a preliminary version, return to this at the end of the writing process.}
-
 This paper explores the design of a domain specific language, Marlowe,\footnote{Marlowe is named after Christopher Marlowe, the Elizabethan poet, dramatist and spy, who was born and educated in Canterbury, which is also where our university is based.
@@
 
-In designing, Marlowe we have chosen a number of options; this description applies to Marlowe version 1.2 \sjtnote{Is 
-there an obvious way of making that clear in the repo? I really don't understand how git deals with ``releases'' or 
-versions.}\plsnote{We can create a branch and point to that.}
+In designing, Marlowe we have chosen a number of options; this description applies to Marlowe version 1.2
 
@@
 It is up to the interested party (or parties) to make the change happen. It is therefore possible for the payment not to 
-happen, despite the fact that this is against the interest of the receiving party.\plsnote{There was a trailing ``the 
-same'' here. Did you think of another example?}
+happen, despite the fact that this is against the interest of the receiving party.
 
-\paragraph{UTxO vs.\ accounts} \sjtnote{Pablo: please check this section carefully}
+\paragraph{UTxO vs.\ accounts}
 
@@
 Does the \haskellinline{Both} constructor potentially produce invalid/misleading contracts, combining a commitment in 
-one half with its use in the other? The problem is that the ``use'' conjunct could potentially be executed, leading to 
-a failed payment. \plsnote{What do you mean with the ``use'' conjunct?}
+one half with its use in the other? The problem is that this might potentially lead to 
+a failed payment.
 
@@
 \label{section:towards-implementation}
-
-\sjtnote{I have expanded this to include a number of responses to Charles' comments on an earlier document. Other responses appear in the previous sub-section.}
 
@@
 
-\sjtnote{Have I forgotten anyone?}
-
-Thanks to IOHK staff Manuel Chakravarty, Duncan Coutts, Bernardo David, Charles Hoskinson, Aggelos Kiyakis, Darryl 
+Thanks to IOHK staff Manuel Chakravarty, Duncan Coutts, Bernardo David, Charles Hoskinson, Aggelos Kiayias, Darryl 
 McAdams, Bruno Woltzenlogel Paleo, and Phil Wadler for their collaboration, and to colleagues from the Universities of 

### 2018-03-26 tex/paper.tex
Add explanation of web demo

@@
 \section{Visualising and interacting with contracts}
+
 \label{section:tool}
 
-\sjtnote{Pablo: can you add something here about the online visualisation, including one or more suitable screenshots?
-It might be worth giving a condensed screenshot of the whole screen, and then more detailed ones of parts of the display.}
+One of the objectives of Marlowe is to be accessible to a wider audience
+than a general purpose language would. With this aim, and to aid Marlowe's
+usage by people that does not know is not familiar with its syntax,
+we have developed a web tool that supports the interactive construction,
+edition, and simulation of smart-contracts written in Marlowe. The
+tool is publicly available in the url: \url{https://input-output-hk.github.io/scdsl/}.
+In Figure~\ref{fig:full-screenshot-demo}, we provide a screenshot
+of the tool in the middle of simulating the execution of the ``incentivised
+saving'' contract described in Section~\ref{section:example}.
 
-\sjtnote{Should it have a name?}
+The tool has been mainly written in Haskell and compiled to JavaScript
+by the Haste compiler \plsnote{Citation needed}, it also relies on
+the Blockly library \plsnote{Citation needed} for providing a visual
+editor for smart contracts written in Marlowe. Blockly's editor allows
+the user to visualise and edit smart-contracts as interlocking blocks
+that can be dragged and dropped like pieces of a jigsaw puzzle (see
+Figure~\ref{fig:detail-of-block}). The tool also provides functionality
+to generate syntactically correct and formatted code (that is displayed
+in the upper right corner of the application), and to convert the
+code back to its block representation.
 
-In the next section we re-examine the model and the Marlowe language, looking at possible extensions, alternative design decisions and ways in which we can formally analyse Marlowe contracts.
+\begin{figure}
+\begin{centering}
+\includegraphics[width=1\textwidth]{pix/screenshot1}
+\par\end{centering}
+\caption{\label{fig:full-screenshot-demo}Screenshot of the visualisation tool
+running the incentivised deposit contract.}
+\end{figure}
+
+\begin{figure}
+\centering{}\includegraphics[scale=0.7]{pix/detail1}\caption{\label{fig:detail-of-block}Detail of five interlocked 
+blocks.}
+\end{figure}
+
+The execution of complete contracts can be simulated block by block
+by using the panel on the right (see Figure~\ref{fig:detail-of-interface}),
+which includes text fields to view and edit the current block number,
+the state of the contract, the current inputs, and the outputs from
+last block.
+
+\begin{figure}
+\centering{}\includegraphics[scale=0.7]{pix/detail2}\caption{\label{fig:detail-of-interface}Detail of the interface for 
+contract
+execution simulation.}
+\end{figure}
+
+Additionally, to facilitate the introduction of inputs, the tool provides
+two different interfaces:
+\begin{itemize}
+\item The manual interface (see Figure~\ref{fig:detail-of-manual-interface}),
+provides a template for each of the four possible types of input:
+commits, redeems, payment claims, and choices.
+\item The smart interface (see Figure~\ref{fig:detail-of-smart-interface}),
+calculates the possible operations that would make sense given the
+current inputs, state of the contract, block number, and remaining
+contract; and it provides them in a table with most of the parameters
+already filled in.
+\end{itemize}
+\begin{figure}
+\begin{centering}
+\includegraphics[scale=0.7]{pix/detail3}
+\par\end{centering}
+\caption{\label{fig:detail-of-manual-interface}Detail of the manual interface.}
+\end{figure}
+
+\begin{figure}
+\centering{}\includegraphics[scale=0.7]{pix/detail4}\caption{\label{fig:detail-of-smart-interface}Detail of the smart 
+interface.}
+\end{figure}
+
+The smart interface is usually more convenient to use than the manual
+interface since the later provides between 3 and 4 fields for each
+operation whereas the former ``guesses'' the possible intentions
+of the user and usually can input new operations with a single click
+(except for choices, which still may require the user to input a number).
 

### 2018-03-26 tex/paper.tex
Restore lost paragraph
Retake snapshots and rewrite progress section

@@
 \caption{\label{fig:full-screenshot-demo}Screenshot of the visualisation tool
-running the incentivised deposit contract.}
+simulating the incentivised deposit contract.}
 \end{figure}
@@
 \begin{figure}
-\centering{}\includegraphics[scale=0.7]{pix/detail1}\caption{\label{fig:detail-of-block}Detail of five interlocked 
+\centering{}\includegraphics[scale=0.5]{pix/detail1}\caption{\label{fig:detail-of-block}Detail of five interlocked 
 blocks.}
@@
 \begin{figure}
-\centering{}\includegraphics[scale=0.7]{pix/detail2}\caption{\label{fig:detail-of-interface}Detail of the interface for 
+\centering{}\includegraphics[scale=0.5]{pix/detail2}\caption{\label{fig:detail-of-interface}Detail of the interface for 
 contract
@@
 \begin{centering}
-\includegraphics[scale=0.7]{pix/detail3}
+\includegraphics[scale=0.5]{pix/detail3}
 \par\end{centering}
@@
 \begin{figure}
-\centering{}\includegraphics[scale=0.7]{pix/detail4}\caption{\label{fig:detail-of-smart-interface}Detail of the smart 
+\centering{}\includegraphics[scale=0.5]{pix/detail4}\caption{\label{fig:detail-of-smart-interface}Detail of the smart 
 interface.}
@@
 (except for choices, which still may require the user to input a number).
+
+In the next section we re-examine the model and the Marlowe language, looking at possible extensions, alternative design decisions and ways in which we can formally analyse Marlowe contracts.
 
@@
 
-That is, for each contract do there exist sequences of values of \haskellinline{Commit} and \haskellinline{OS} such that stepping the contract with these inputs leads to \haskellinline{Null}?
-Some relevant points include:
+That is, for each contract do there exist sequences of inputs and observables such that stepping the contract with them 
+leads to \haskellinline{Null}? Furthermore, do all sequences of inputs and observables eventually evolve to 
+\haskellinline{Null}? Some relevant points include:
 \begin{itemize}
-\item For \haskellinline{When}, we need to be able to make \haskellinline{interpretObs st obs os} true. Since there is 
-a timeout to \haskellinline{When} then there is no need to impose this condition, as we will progress at timeout if not 
-before.
-\item As we have put timeouts on to \haskellinline{CommitCash} and \haskellinline{RedeemCC} operations, there are no remaining requirements on these constructs for contract execution to make progress, albeit eventually.
-\item We need to check the status of \haskellinline{Observations}: what if an observation is unsatisfiable?
-\item There are no constraints on other constructs.
+\item For \haskellinline{When}, in principle, we would need to be able to make \haskellinline{interpretObs st obs os} 
+true. However, since there is  a timeout to \haskellinline{When} then there is no need to impose this condition, as we 
+will progress at timeout if not before.
+\item As we have put timeouts on to \haskellinline{CommitCash} and \haskellinline{Pay}, the only construct that may not 
+reduce to \haskellinline{Null} eventually is \haskellinline{RedeemCC}. \haskellinline{RedeemCC} will still reduce to 
+\haskellinline{Null} if the commitment that it refers to has expired (and it will produce a 
+\haskellinline{DuplicateRedeem} action), but it will be quiescent forever if the money is never comitted in the first 
+place.
+\item Other constructs will reduce to \haskellinline{Null} eventually as long as their continuations also reduce to 
+\haskellinline{Null} (as is the case of \haskellinline{Both}).
 \end{itemize}

### 2018-03-26 tex/paper.tex
Add references to Blockly, Haste, and version 1.2

@@
 The tool has been mainly written in Haskell and compiled to JavaScript
-by the Haste compiler \plsnote{Citation needed}, it also relies on
-the Blockly library \plsnote{Citation needed} for providing a visual
+by the Haste compiler \cite{Haste}, it also relies on
+the Blockly library \cite{Blockly} for providing a visual
 editor for smart contracts written in Marlowe. Blockly's editor allows
@@
 
-We have given a concrete model and DSL, namely Marlowe version 1.2. In this section we examine how the model can be 
-extended, how contacts can be analysed, reflect on different design decisions that might be made and discuss some 
-aspects of the implementation of Marlowe within a blockchain environment.
+We have given a concrete model and DSL, namely Marlowe version 
+1.2\footnote{\url{https://github.com/input-output-hk/scdsl/releases/tag/v1.2}}.
+In this section we examine how the model can be extended, how contacts can be analysed, reflect on different design 
+decisions that might be made and discuss some aspects of the implementation of Marlowe within a blockchain environment.
 

### 2018-03-26 tex/paper.tex
Fix layout problem
Add draft abstract

@@
 \begin{abstract}
-  No abstract
+
+Blockchains allow the specification of contracts in the form of programs that guarantee their fulfilment. 
+Nevertheless, errors in those programs can cause important, and often irreversible, economical loses. General purpose 
+languages tend to be unsuitable for this end, since their behaviour in exceptional situations tends to be hard to 
+predict, and they are difficult to analyse due to their generality (they are usually turing-complete).
+
+This paper explores the design of a domain specific language, Marlowe, targeted at the execution of financial 
+contracts in the style of Peyton Jones on blockchains. We have focused on predictability, enforceability, simplicity, 
+and ease of analysis and use.
+
 \end{abstract}
@@
 continuation contract.
-\item \haskellinline{RedeemCC} should not happen twice in any execution, so as to avoid producing a
+\item \haskellinline{RedeemCC} should not happen twice in any execution, so as to avoid a
 \haskellinline{DuplicateRedeem} action.

### 2018-03-26 tex/paper.tex
Fix typo

@@
 as a ``computation that results in something of type \haskellinline{a}''. The \haskellinline{Monad} API allows 
-suchcomputations to be assembled, and then, once assembled, they may be run, in general causing some side-effects as a 
+such computations to be assembled, and then, once assembled, they may be run, in general causing some side-effects as a 
 part of the computation.

### 2018-03-26 tex/paper.tex
revised abstract

@@
 Blockchains allow the specification of contracts in the form of programs that guarantee their fulfilment. 
-Nevertheless, errors in those programs can cause important, and often irreversible, economical loses. General purpose 
-languages tend to be unsuitable for this end, since their behaviour in exceptional situations tends to be hard to 
-predict, and they are difficult to analyse due to their generality (they are usually turing-complete).
+Nevertheless, errors in those programs can cause important, and often irretrievable, monetary loss.
+General-purpose languages provide a platform on which contracts can be built, but by their very generality they have the potential to exhibit behaviours of an unpredictable kind, and are also not easy to read or comprehend for general users.
 
-This paper explores the design of a domain specific language, Marlowe, targeted at the execution of financial 
-contracts in the style of Peyton Jones on blockchains. We have focused on predictability, enforceability, simplicity, 
+An alternative solution is provided by domain-specific languages (DSLs), which are designed to express programs in a particular field.
+This paper explores the design of one DSL, Marlowe, targeted at the execution of financial
+contracts in the style of Peyton Jones \emph{et al} on blockchains. We have focused on predictability, enforceability, simplicity, 
 and ease of analysis and use.

### 2018-03-26 tex/paper.tex
Add some references

@@
 
-\sjtnote{References to some of the background work still to be added.}
-
 \medskip
@@
 
-One approach to this is to choose mechanisms such as bitcoin script which are manifestly non-Turing complete.
-A bitcoin script, written in a Forth-like language, is essentially linear: it can branch, but the language contains neither looping constructs nor recursion.
-It is therefore straightforward not only to see that scripts will terminate, but also to give an accurate estimate of the time taken to execute a script.
+One approach to this is to choose mechanisms such as bitcoin script~\cite{BitcoinWikiScript} which are manifestly 
+non-Turing complete. A bitcoin script, written in a Forth-like language, is essentially linear: it can branch, but the 
+language contains neither looping constructs nor recursion. It is therefore straightforward not only to see that scripts 
+will terminate, but also to give an accurate estimate of the time taken to execute a script.
 
@@
 
-On the other hand, the Ethereum system provides a Turing-complete language for the EVM virtual machine, and a higher-level programming language, Solidity, that compiles into EVM code.
-However, EVM and Solidity programs are constrained \emph{post hoc} by two mechanisms: program execution must be paid for using `gas' proportional to the effort expended, and a set of \emph{ad hoc} limits on program execution, e.g.\ on stack size.
+On the other hand, the Ethereum system~\cite{wood2014ethereum} provides a Turing-complete language for the EVM virtual 
+machine, and a higher-level programming language, Solidity, that compiles into EVM code. However, EVM and Solidity 
+programs are constrained \emph{post hoc} by two mechanisms: program execution must be paid for using `gas' proportional 
+to the effort expended, and a set of \emph{ad hoc} limits on program execution, e.g.\ on stack size.
 
@@
 
-Programmability of the system is provided through a ``fat'' high-level API, which is accessible from Nxt clients through a REST interface.
-The API provides functionality supporting various kinds of transactions.
-The core software itself doesn't support any form of scripting language; rather, users are expected to work with the built in transaction types and transactions that support some 250 primitive operations; these can be ``scripted'' in a client (only) using a binding to the API, which is available, for instance, in JavaScript.
+In Nxt~\cite{Nxt}, programmability of the system is provided through a ``fat'' high-level 
+API, which is accessible from Nxt clients through a REST interface. The API provides functionality supporting various 
+kinds of transactions. The core software itself doesn't support any form of scripting language; rather, users are 
+expected to work with the built in transaction types and transactions that support some 250 primitive operations; these 
+can be ``scripted'' in a client (only) using a binding to the API, which is available, for instance, in JavaScript.
 
@@
 The \ae{}ternity system~\cite{aeternity} provides multiple languages and VMs:\footnote{As presented by Erik Stenman at CODE BEAM SF, March 2018, \url{https://www.youtube.com/watch?v=VXsqvfPIdWg}} the functional language  Sophia (akin to Reason) and the Functional Typed Warded Virtual Machine (FTWVM) for safe ``system level programming'', the language Varna and the HLM for simple contracts, and a (port of) Solidity and the EVM for compatibility with Ethereum.
-In a similar way, Cardano provides support for IELE~\sjtnote{citation}, a rational reconstruction of the EVM, and thus for Solidity contracts too.
+In a similar way, Cardano provides support for IELE~\cite{IELE}, a rational reconstruction of the EVM, and thus for Solidity contracts too.
 
@@
 
-There are a variety of other approaches to scripting, including business process notations, such as the OMG's BPMN; process calculi such as Rholang; logic rules (in Formal Contract Logic), and finite state machines.
-More details of these are provided in~\cite{cryptoeprint:2016:1156}.
+There are a variety of other approaches to scripting, including business process notations, such as the OMG's BPMN; 
+process calculi such as Rholang; logic rules (in Formal Contract Logic), and finite state machines. More details of 
+these are provided in~\cite{cryptoeprint:2016:1156}.
 
@@
 The Cardano Settlement Layer (SL) is to support settlement of transactions but nothing more complicated, whereas the Computation Layer (or CL) supports more powerful general computations over the chain.
-Moreover, the Cardano roadmap \cite{cardano-rationale} envisages the possibility of computation taking place, in part at least, on a sidechain rather than on the main chain itself, which may support only the SL.
+Moreover, the Cardano roadmap~\cite{cardano-rationale} envisages the possibility of computation taking place, in part at least, on a sidechain rather than on the main chain itself, which may support only the SL.
 Other off-chain approaches include the Lightning network, overlaid on blockchains (usually Bitcoin) and the channels of \ae{}ternity, introduced below.
@@
 The tool has been mainly written in Haskell and compiled to JavaScript
-by the Haste compiler \cite{Haste}, it also relies on
-the Blockly library \cite{Blockly} for providing a visual
+by the Haste compiler~\cite{Haste}, it also relies on
+the Blockly library~\cite{Blockly} for providing a visual
 editor for smart contracts written in Marlowe. Blockly's editor allows
@@
 commitment would include a choice of a play in a game, which would be chosen and concealed from the other players until 
-a later point in play, when it can be revealed. The paper \cite{kumaresan2015use} makes use of this and other 
+a later point in play, when it can be revealed. The paper~\cite{kumaresan2015use} makes use of this and other 
 cryptographic primitives to enforce the rules of poker while maintaining the cards of each player secret.
@@
 Haskell is a pure functional programming language, and uses the \haskellinline{Monad} type class to encapsulate
-operations that can potentially have a side effect. Many introductions to monads exist \sjtnote{cite some} but for the 
-sake of this discussion it is enough to think of a value of type \haskellinline{m a} where \haskellinline{m} is a monad 
-as a ``computation that results in something of type \haskellinline{a}''. The \haskellinline{Monad} API allows 
-such computations to be assembled, and then, once assembled, they may be run, in general causing some side-effects as a 
-part of the computation.
+operations that can potentially have a side effect. Many introductions to monads exist \cite{wadler1990comprehending} 
+but for the sake of this discussion it is enough to think of a value of type \haskellinline{m a} where \haskellinline{m} 
+is a monad as a ``computation that results in something of type \haskellinline{a}''. The \haskellinline{Monad} API 
+allows such computations to be assembled, and then, once assembled, they may be run, in general causing some 
+side-effects as a part of the computation.
 

### 2018-03-26 tex/paper.tex
Give a name to the web demo

@@
 it, the types of the principal functions and a description of the DSL as an algebraic type, constructor by constructor. 
-Section \ref{section:example} presents an example of a larger Marlowe contract, for incentivised saving, and Section \ref{section:tool} introduces our tool for visualising and interacting with Marlowe contracts.
+Section \ref{section:example} presents an example of a larger Marlowe contract, for incentivised saving, and Section 
+\ref{section:tool} introduces Meadow, our tool for visualising and interacting with Marlowe contracts.
 Section \ref{section:reflection} reflects on the system design, and observes some alternative designs and Section \ref{section:next-steps} enumerates the next steps for the project.
@@
 usage by people that does not know is not familiar with its syntax,
-we have developed a web tool that supports the interactive construction,
+we have developed Meadow, a web tool that supports the interactive construction,
 edition, and simulation of smart-contracts written in Marlowe. The
@@
 In Figure~\ref{fig:full-screenshot-demo}, we provide a screenshot
-of the tool in the middle of simulating the execution of the ``incentivised
+of Meadow in the middle of simulating the execution of the ``incentivised
 saving'' contract described in Section~\ref{section:example}.
 
-The tool has been mainly written in Haskell and compiled to JavaScript
+Meadow has been mainly written in Haskell and compiled to JavaScript
 by the Haste compiler~\cite{Haste}, it also relies on
@@
 that can be dragged and dropped like pieces of a jigsaw puzzle (see
-Figure~\ref{fig:detail-of-block}). The tool also provides functionality
+Figure~\ref{fig:detail-of-block}). Meadow also provides functionality
 to generate syntactically correct and formatted code (that is displayed
@@
 \par\end{centering}
-\caption{\label{fig:full-screenshot-demo}Screenshot of the visualisation tool
+\caption{\label{fig:full-screenshot-demo}Screenshot of Meadow tool
 simulating the incentivised deposit contract.}
@@
 
-Additionally, to facilitate the introduction of inputs, the tool provides
+Additionally, to facilitate the introduction of inputs, Meadow provides
 two different interfaces:
@@
 
-Some degree of validation is provided by the visualisation tool presented in Section \ref{section:tool}, albeit for smaller-scale contracts, but there is a general question of how properties of contracts can be inferred.
-One relatively low-cost is to use QuickCheck-style property-based testing~\cite{quickCheck}, and properties developed here become candidates for fully fledged verification in a formalisation of the (semantics of) Marlowe.
+Some degree of validation is provided by Meadow presented in Section \ref{section:tool}, albeit for smaller-scale 
+contracts, but there is a general question of how properties of contracts can be inferred. One relatively low-cost is to 
+use QuickCheck-style property-based testing~\cite{quickCheck}, and properties developed here become candidates for 
+fully fledged verification in a formalisation of the (semantics of) Marlowe.
 

### 2018-04-04 tex/paper.tex
small changes
Explain implementation of operations as transactions

@@
 %    , anonymous
-    , review=true
+%    , review=true
   ]{acmart}
@@
 % Metadata Information
-\acmConference[ICFP]{International Conference on Functional Programming}{23--29 September 2018}{St.~Louis, Missouri, United States}
+%\acmConference[DRAFT]{Draft paper}
 %\acmJournal{PACMPL}
@@
 
-The contracts implemented in the Marlowe language have an implicit interface defined for what would correspond to the client side of a distributed application, as shown by the interactive interface.
+The contracts implemented in the Marlowe language have an implicit interface defined for what would correspond to the 
+client side of a distributed application, as shown by the interactive interface.
 
@@
 \item one or more smart contracts (potentially written in Plutus) and 
-\item a client side program or interface that has a limited number of choices to be made at each point in time (including an interface for the observables). 
+\item a client side program or interface that has a limited number of choices to be made at each point in time 
+(including an interface for the observables). 
 \end{itemize}
-Of course, we probably do not want to give a user a programmatic interface but an actual application with user interface.
-So there is still some work to be done to create interfaces on top of the client side of the contracts.
+Of course, we probably do not want to give a user a programmatic interface but an actual application with user 
+interface. So there is still some work to be done to create interfaces on top of the client side of the contracts.
+
+Since the current version of Marlowe follows the `pull' model, a contract written in Marlowe could also be potentially
+compiled to a contract written in Ethereum. This could be done in two ways, it could be compiled so that an Ethereum
+contract represents a single instance of the contract (with a fixed set of participants), or alternatively it would be
+possible to create a generic contract with an extra ``instantiate'' operation that sets-up' a new contract instance
+with a given set of participants (represented by a set of public keys).
+
+\paragraph{Transactions}
+
+In principle, we can directly translate each operation in the model (i.e: cash commitment, payment claim, choice, or
+redemption) into a separate transaction in the blockchain. In Ethereum model this would work in the same 
+way that is simulated by Meadow: several operations can be issued in parallel and the blockchain subsystem will 
+automatically apply them in some order and integrate them into the next block.
+
+Another approach would be to use a UTXO model with continuations in which a UTXO represents a contract. In this model,
+an operation would correspond to a transaction that spends a UTXO and creates (at least) a new UTXO that represents the 
+contract after the operation.
+
+Usually, in the UTXO model, it is necessary that a UTXO is in the blockchain in order for another transaction to be able 
+to spend it. This could potentially limit the number of operations that can be applied in a single block to one, which 
+would make infeasible the approach of using the same UTXO to represent several instances of the same contract (since it 
+would not scale).
+
+On the other hand, it would prevent non-determinism and, with it, it would remove the possibility of race-conditions. A 
+change to the state of a contract before a transaction (and its corresponding operation) is accepted in the blockchain 
+would invalidate the transaction and require the user to issue it again.
+
+Even if we only allow one transaction per block, it would still be possible to combine several transactions off-chain 
+and to issue a transaction that contains several operations. However, that would require an offline protocol 
+independent from the blockchain and it would require participants to collaborate.
 

### 2018-04-04 tex/paper.tex
Fix typo and clarify

@@
 Since the current version of Marlowe follows the `pull' model, a contract written in Marlowe could also be potentially
-compiled to a contract written in Ethereum. This could be done in two ways, it could be compiled so that an Ethereum
-contract represents a single instance of the contract (with a fixed set of participants), or alternatively it would be
-possible to create a generic contract with an extra ``instantiate'' operation that sets-up' a new contract instance
-with a given set of participants (represented by a set of public keys).
+compiled to a smart-contract written in Ethereum. This could be done in two ways, it could be compiled so that an 
+smart-contract represents a single instance of the Marlowe contract (with a fixed set of participants), or 
+alternatively it would be possible to create a generic smart-contract with an extra ``instantiate'' operation that 
+sets-up a new Marlowe contract instance with a given set of participants (represented by a set of public keys).
 

### 2018-04-04 meadow/src/Main.hs
Add choice analysis for CommitCash and Pay

@@ -5,7 +5,7 @@ import qualified Haste.JSString as J
 import qualified Haste.Prim as P
 import qualified Haste.DOM as D
 import qualified DepositIncentive as DI
-import qualified CrowFunding as CF
+import qualified CrowdFunding as CF
 import qualified Escrow as E
 import Semantics
 import ContractFormatter

### 2018-04-04 src/SmartInputs.hs

@@ -42,14 +42,23 @@ getUsedChoiceNumbersObs st (AndObs obs1 obs2) =
 getUsedChoiceNumbersObs st (OrObs obs1 obs2) =
   foldl1 combineInputs $ map (getUsedChoiceNumbersObs st) [obs1, obs2]
 getUsedChoiceNumbersObs st (NotObs obs) = getUsedChoiceNumbersObs st obs
+getUsedChoiceNumbersObs st (ValueGE m1 m2) = foldl1 combineInputs $ map (getUsedChoiceNumbersMoney st) [m1, m2]
 getUsedChoiceNumbersObs _ _ = emptyInput
+
+-- Obtains all the choices that are potentially checked by a money expression (taken as "0")
+getUsedChoiceNumbersMoney :: State -> Money -> Input
+getUsedChoiceNumbersMoney st (AddMoney m1 m2) = foldl1 combineInputs $ map (getUsedChoiceNumbersMoney st) [m1, m2]
+getUsedChoiceNumbersMoney st (MoneyFromChoice identch per m)
+ = foldl1 combineInputs $ [emptyInput {ic = Map.singleton (identch, per) 0}, getUsedChoiceNumbersMoney st m]
+getUsedChoiceNumbersMoney st _ = emptyInput
 
 -- Obtains all the choices that are potentially checked by a contract (taken as "0")
 getUsedChoiceNumbers :: State -> Contract -> Input
-getUsedChoiceNumbers st (CommitCash _ _ _ _ _ c1 c2) =
-  foldl1 combineInputs $ map (getUsedChoiceNumbers st) [c1, c2]
+getUsedChoiceNumbers st (CommitCash _ _ m _ _ c1 c2)
+ = foldl1 combineInputs $ ((getUsedChoiceNumbersMoney st m):(map (getUsedChoiceNumbers st) [c1, c2]))
 getUsedChoiceNumbers st (RedeemCC _ c) = getUsedChoiceNumbers st c
-getUsedChoiceNumbers st (Pay _ _ _ _ _ c) = getUsedChoiceNumbers st c
+getUsedChoiceNumbers st (Pay _ _ _ m _ c)
+ = foldl1 combineInputs $ [getUsedChoiceNumbersMoney st m, getUsedChoiceNumbers st c]
 getUsedChoiceNumbers st (Both c1 c2) =
   foldl1 combineInputs $ map (getUsedChoiceNumbers st) [c1, c2]
 getUsedChoiceNumbers _ Null = emptyInput

### 2018-04-13 tex/paper.tex
Add escrow example to paper

@@
 \end{haskellcode}
+
+In the next section, we look at how to take advantage of Haskell embedding when implementing Marlowe contracts.
+
+\section{Haskell embedding example}
+\label{section:example}
+
+In this section, we give an example of how we can have Marlowe contracts that are easier to write, read, and 
+understand, by embedding them into Haskel code (taking advantage that Marlowe contracts are implemented as Haskell 
+terms).
+
+The example contract we use through this section implements an escrow contract. Let us imagine that ``participant 1'' 
+is buying some item from ``participant 2''. On one hand, if ``participant 1'' pays before receiving the item, there 
+is a risk that ``participant 2'' will never give ``participant 1'' the item. On the other hand, if ``participant 2'' 
+gives ``participant 1'' the item before ``participant 1'' pays, then there is a risk that ``participant 1'' will never 
+pay.
+
+The escrow mechanism allows ``participant 1'' to deposit the money into a contract, in a way that the money will only 
+be released when two out of three participants agree on whether ``participant 2'' has indeed given ``participant 1'' 
+the item. The third participant (``participant 3'') is supposed to be a neutral third party that will decide in case of 
+dispute.
+
+This way, if ``participant 1'' and ``participant 2'' are honest, they will just agree on the result of the transaction 
+and ``participant 3'' will not need to do anything. If ``participant 1'' and ``participant 2'' disagree, ``participant 
+3'' will be able to choose whether the money must go to ``participant 1'' or to ``participant 2''.
+
+In addition to this, in our implementation, the money paid for the item is 450 ADA, it must be committed by 
+``participant 1'' before block 10, and it will be refunded to ``participant 1'' if there is no consensus before block 
+90.
+
+We start by defining some Haskell functions. We can encapsulate identifiers in functions to make the contract more 
+readable. That way we can generate an identifier for the cash commitment:
+
+\begin{haskellcode}
+iCC1 :: IdentCC
+iCC1 = IdentCC 1
+\end{haskellcode}
+
+And an identifier for the payment:
+
+\begin{haskellcode}
+iP1 :: IdentPay
+iP1 = IdentPay 1
+\end{haskellcode}
+
+We can also create a sub-contract that allows the money to be redeemed by ``participant 1'':
+
+\begin{haskellcode}
+redeem_original :: Contract
+redeem_original = RedeemCC iCC1 Null
+\end{haskellcode}
+
+Each participant has a say on who deserves the money: either ``participant 1'' deserves a refund, which we represent 
+with the number $0$; or ``participant 2'' deserves a payment, which we represent with the number $1$. Because there is 
+only one choice to make per participant, we use the same \haskellinline{IdentChoice} as their participant id. We can 
+define a function that returns an observation that is true if and only if the participant \haskellinline{per} has 
+chosen the number \haskellinline{c} as follows:
+
+\begin{haskellcode}
+chose :: Int -> ConcreteChoice -> Observation
+chose per c = PersonChoseThis (IdentChoice per) per c
+\end{haskellcode}
+
+Then we can easily write a function that returns an observation that is true if and only if at least one of the 
+participants \haskellinline{per} and \haskellinline{per'} has chosen the number \haskellinline{c} as follows:
+
+\begin{haskellcode}
+one_chose :: Person -> Person -> ConcreteChoice -> Observation
+one_chose per per' val = (OrObs (chose per val) (chose per' val)) 
+\end{haskellcode}
+
+After that, we can write a function that returns an observation that is true if and only if at least 
+two of the three participants \haskellinline{p1}, \haskellinline{p2}, and \haskellinline{p3} have chosen the number 
+\haskellinline{c} as follows:
+
+\begin{haskellcode}
+two_chose :: Person -> Person -> Person -> ConcreteChoice -> Observation
+two_chose p1 p2 p3 c = OrObs (AndObs (chose p1 c) (one_chose p2 p3 c))
+                             (AndObs (chose p2 c) (chose p3 c))
+\end{haskellcode}
+
+Finally, we can write the escrow contract, which would remain as follows:
+
+\begin{haskellcode}
+escrow :: Contract
+escrow = CommitCash iCC1 1 (ConstMoney 450) 10 100
+                    (When (OrObs (two_chose 1 2 3 0)
+                                 (two_chose 1 2 3 1))
+                          90
+                          (Choice (two_chose 1 2 3 1)
+                                  (Pay iP1 1 2 (AvailableMoney iCC1) 100
+                                       redeem_original)
+                                  redeem_original)
+                          redeem_original)
+                    Null
+\end{haskellcode}
+
+The outermost primitive \haskellinline{CommitCash} allows the ``participant 1'' to commit 450 ADA before block 10, 
+with the promise that money will be released on block 100 if they are not claimed before that.
+
+The next primitive, \haskellinline{When}, waits for one of three things to happen:
+\begin{enumerate}
+ \item Two out of three choose $0$.
+ \item Two out of three choose $1$.
+ \item The block 90 is added to the blockchain.
+\end{enumerate}
+
+If the third option happens first, the money is refunded. Otherwise, there is a \haskellinline{Choice} that will 
+immediately refund the money to ``participant 1'' unless two out of three chose $1$; in the later case, 
+\haskellinline{Pay} will give ``participant 2'' the opportunity to claim the funds available in \haskellinline{iCC1} 
+before block 100. If the funds are not claimed before block 100 they will also be refunded to ``participant 1''.
+
 In the next section, we cover the online contract visualisation and interaction tool for Marlowe.

### 2018-04-13 src/EscrowV2.hs

@@ -1,4 +1,4 @@
-module Escrow where
+module EscrowV2 where
 
 import Semantics
 

### 2018-04-15 tex/paper.tex
added Haskell code for the step function to the paper

@@
 
-\item \haskellinline{Null} is the null contract; it will always be quiescent.
+\item \haskellinline{Null} is the null contract; it will always be quiescent:
+
+\smallskip
+\begin{haskellcode}
+step _ st Null _ = (st, Null, [])
+\end{haskellcode}
+\smallskip
 
@@
 \item
-or the timeout \haskellinline{t1} is exceed.
+or the timeout \haskellinline{t1} is exceed:
 \end{itemize}
+
+\smallskip
+\begin{haskellcode}
+step 
+  commits 
+  st 
+  c@(CommitCash ident person val start_timeout end_timeout con1 con2) 
+  os
+  | cexe || cexs = (st {sc = ust}, con2, [])
+  | Set.member (CC ident person cval end_timeout) (cc commits)
+        = (st {sc = ust}, con1, [SuccessfulCommit ident person cval])
+  | otherwise = (st, c, [])
+  where ccs = sc st
+        cexs = expired (blockNumber os) start_timeout
+        cexe = expired (blockNumber os) end_timeout
+        cns = (person, if cexe || cexs 
+                          then ManuallyRedeemed 
+                          else NotRedeemed cval end_timeout)
+        ust = Map.insert ident cns ccs
+        cval = evalMoney st val
+\end{haskellcode}
+\smallskip
 In the first case, a \haskellinline{SuccessfulCommit} action is generated and the contract continues as 
@@
 \item \haskellinline{RedeemCC idCC k} (\haskellinline{CC} stands for cash commitment.) For this contract to make progress, the creator of the cash commitment with identifier \haskellinline{idCC} is required to redeem the unspent funds in that commitment; the contract then continues as \haskellinline{k}, and the action \haskellinline{CommitRedeemed} is produced.
+
+\smallskip
+\begin{haskellcode}
+step commits st c@(RedeemCC ident con) _ =
+    case Map.lookup ident ccs of
+      Just (person, NotRedeemed val _) ->
+        let newstate = 
+                st {sc = Map.insert ident (person, ManuallyRedeemed) ccs} in
+        if Set.member (RC ident person val) (rc commits)
+        then (newstate, con, [CommitRedeemed ident person val])
+        else (st, c, [])
+      Just (person, ManuallyRedeemed) ->
+        (st, con, [DuplicateRedeem ident person])
+      Nothing -> (st,c,[])
+    where
+        ccs = sc st
+\end{haskellcode}
+\smallskip
+
 Committed cash can only be redeemed once, and an attempt to redeem it a second time will produce a  \haskellinline{DuplicateRedeem} action, continuing as \haskellinline{k}.
@@
 \haskellinline{p1} to claim  a payment of \haskellinline{v} from \haskellinline{p2} before the timeout 
-\haskellinline{t}. The contract continues as \haskellinline{k}. By `available' we mean that sufficient commitments have 
+\haskellinline{t}. The contract continues as \haskellinline{k}.
+
+\smallskip
+\begin{haskellcode}
+step inp st c@(Pay idpay from to val expi con) os
+  | expired (blockNumber os) expi = (st, con, [ExpiredPay idpay from to cval])
+  | right_claim =
+    if committed st from bn >= cval
+      then (newstate, con, [SuccessfulPay idpay from to cval])
+      else (st, con, [FailedPay idpay from to cval])
+  | otherwise = (st, c, [])
+  where
+    cval = evalMoney st val
+    newstate = stateUpdate st from to bn cval
+    bn = blockNumber os
+    right_claim =
+      case Map.lookup (idpay, to) (rp inp) of
+        Just claimed_val -> claimed_val == cval
+        Nothing -> False
+\end{haskellcode}
+\smallskip
+
+By `available' we mean that sufficient commitments have 
 been made and  not yet expired to cover the payment; in this case the payment  uses the currency allocated by the cash 
@@
 \item \haskellinline{Both k1 k2} enforces the behaviour of both contracts \haskellinline{k1} and \haskellinline{k2}.
+
+\smallskip
+\begin{haskellcode}
+step comms st (Both con1 con2) os =
+    (st2, result, ac1 ++ ac2)
+    where
+        result | res1 == Null = res2
+               | res2 == Null = res1
+               | otherwise = Both res1 res2
+        (st1,res1,ac1) = step comms st con1 os
+        (st2,res2,ac2) = step comms st1 con2 os
+\end{haskellcode}
+\smallskip
+
 Because the model is stateful and produces output actions, to make a step it is necessary to execute a single step of each of  the contracts \haskellinline{k1} and \haskellinline{k2} in sequence: first \haskellinline{k1} then \haskellinline{k2}.
@@
 \item \haskellinline{Choice obs k1 k2} behaves as either \haskellinline{k1} or \haskellinline{k2} depending on the (Boolean) result of \haskellinline{obs} at the time that the observation is made, \haskellinline{k1} if it is \haskellinline{True} and \haskellinline{k2} if \haskellinline{False}.
+
+\smallskip
+\begin{haskellcode}
+step _ st (Choice obs conT conF) os =
+    if interpretObs st obs os
+        then (st,conT,[])
+        else (st,conF,[])
+\end{haskellcode}
+\smallskip
 
@@
 Otherwise the contract is quiescent.
+
+\smallskip
+\begin{haskellcode}
+step _ st (When obs expi con con2) os
+  | expired (blockNumber os) expi = (st,con2,[])
+  | interpretObs st obs os = (st,con,[])
+  | otherwise = (st, When obs expi con con2, [])
+  \end{haskellcode}
+\smallskip
+
 \end{itemize}

### 2018-04-15 tex/paper.tex
Added simplified escrow example to introduction; fixed multiple label problem.

@@
 This work is a part of the Cardano project, and is supported by IOHK.\footnote{\url{https://iohk.io}}
+
 In doing this we are required to refine the model of contracts in a number of ways in order to fit with a radically different context.
-
-In the traditional model, enforcement of the contract is the responsibility of the legal system; in the decentralised blockchain model, in which there is no central authority, the contract needs to be enforced by design.
-This means that we require participants to commit currency (or ``cash'' as we term it) to cover all possible expenditure in advance of the contract executing; we also have to use timeouts to ensure that the contract does not become stuck even if one or more of the participants stop interacting with it.
+Taking the following example of an ``escrow'' contract we can explain the differences more concretely.
+The aim of this contract, written in functional pseudocode in the style of~\cite{PeytonJones:2000} involves players 1, 2 and 3.
+Player 1 is to pay an amount of money to player 2 on receipt of goods.
+She pays the sum into escrow controlled by player 3.
+There are two options for the money: if two players agree to \haskellinline{pay} it goes ahead; if, on the other hand, two of the players opt to \haskellinline{refund} the money, that is done.
+The \haskellinline{When} construct ensures fires when its first argument -- a condition -- becomes true; the second argument, a \haskellinline{Contract} itself, is then performed, and this makes the payment according to the players' \haskellinline{Choice}.
+
+\begin{haskellcode}
+(When (Or (two_chose 1 2 3 refund)
+          (two_chose 1 2 3 pay))
+      (Choice (two_chose 1 2 3 pay)
+              (Pay 1 2 AvailableMoney)
+              redeem_original))
+\end{haskellcode}
+
+In the traditional model, enforcement of the contract is the responsibility of the legal system.
+If player 1 doesn't pay the money into escrow, or play 3 chooses to keep it for herself, then we can assume that they can be sued for the money and probably damages on top, thus providing both legal and financial incentives for compliance.
+On the other hand, in the decentralised blockchain model, where there is no central authority, the contract needs to be enforced by design.
+
+This means that we require participants to commit money to cover all possible expenditure \emph{in advance of the contract executing}.
+In order to make sure that player continue to engage with a contract we ensure urgency by imposing timeouts: money is committed for a finite period, for example, and when waiting for a player to make a commitment we also impose a timeout to ensure that the contract does not become stuck even if one or more of the participants stops interacting with it.
 
@@
 it, the types of the principal functions and a description of the DSL as an algebraic type, constructor by constructor. 
-Section \ref{section:example} presents an example of a larger Marlowe contract, for incentivised saving, and Section 
+Section \ref{section:example-saving} presents an example of a larger Marlowe contract, for incentivised saving.
+Section \ref{section:example-escrow} revisits the escrow example, and shows how it is described using a combination of Marlowe and Haskell constructs: that is, we use Marlowe as an embedded DSL.
+Section
 \ref{section:tool} introduces Meadow, our tool for visualising and interacting with Marlowe contracts.
@@
 We start by looking at the different types used by the model, then we look at the \haskellinline{Contract} DSL itself, and finally we give its semantics in Haskell.
-Section \ref{section:example} gives an example of Marlowe in action.
+Section \ref{section:example-saving} gives an example of Marlowe in action, and Section
+\ref{section:example-escrow} shows the advantage of seeing Marlowe as a DSL embedded in Haskell.
 
@@
 \section{Incentivised saving}
-\label{section:example}
+\label{section:example-saving}
 
@@
 
-\section{Haskell embedding example}
-\label{section:example}
+\section{Marlowe as a DSL embedded in Haskell}
+\label{section:example-escrow}
 
@@
 of Meadow in the middle of simulating the execution of the ``incentivised
-saving'' contract described in Section~\ref{section:example}.
+saving'' contract described in Section~\ref{section:example-saving}.
 

### 2018-04-15 tex/paper.tex
Corrections to introduction

@@
 In doing this we are required to refine the model of contracts in a number of ways in order to fit with a radically different context.
-Taking the following example of an ``escrow'' contract we can explain the differences more concretely.
-The aim of this contract, written in functional pseudocode in the style of~\cite{PeytonJones:2000} involves players 1, 2 and 3.
-Player 1 is to pay an amount of money to player 2 on receipt of goods.
-She pays the sum into escrow controlled by player 3.
-There are two options for the money: if two players agree to \haskellinline{pay} it goes ahead; if, on the other hand, two of the players opt to \haskellinline{refund} the money, that is done.
+Let us the following example of an ``escrow'' contract we can explain the motivation more concretely.
+The aim of this contract, written in functional pseudocode in the style of~\cite{PeytonJones:2000} involves three participants \haskellinline{1}, \haskellinline{2} and \haskellinline{3}.
+Participant \haskellinline{1} is to pay an amount of money to participant \haskellinline{2} on receipt of goods from her.
+Participant \haskellinline{1} pays the money into escrow controlled by participant \haskellinline{3}.
+There are two options for the money: if two participants agree to \haskellinline{pay} it to \haskellinline{2}, that goes ahead; if, on the other hand, two of the players opt to \haskellinline{refund} the money to \haskellinline{1}, that is done instead.
 The \haskellinline{When} construct ensures fires when its first argument -- a condition -- becomes true; the second argument, a \haskellinline{Contract} itself, is then performed, and this makes the payment according to the players' \haskellinline{Choice}.
@@
 In the traditional model, enforcement of the contract is the responsibility of the legal system.
-If player 1 doesn't pay the money into escrow, or play 3 chooses to keep it for herself, then we can assume that they can be sued for the money and probably damages on top, thus providing both legal and financial incentives for compliance.
+If participant \haskellinline{1} doesn't pay the money into escrow, or participant \haskellinline{3} chooses to keep it for herself, then we can assume that they can be sued for the money and probably damages on top, thus providing both legal and financial incentives for compliance.
 On the other hand, in the decentralised blockchain model, where there is no central authority, the contract needs to be enforced by design.

### 2018-04-15 tex/paper.tex
Some updates to section 5

@@
 In order to make sure that player continue to engage with a contract we ensure urgency by imposing timeouts: money is committed for a finite period, for example, and when waiting for a player to make a commitment we also impose a timeout to ensure that the contract does not become stuck even if one or more of the participants stops interacting with it.
+We revisit this example in Section~\ref{section:example-escrow} where we show how commitments, timeouts and other features of Marlowe are used to describe this as a self-enforcing contract ``on the blockchain''.
 
@@
 In this section, we show how Marlowe can be used for designing a contract that incentivises saving. We use the 
-following example: ``participant 2'' offers 20 ADA to ``participant 1'' in exchange for saving 100 ADA for 100 blocks. 
-``Participant 1'' will be able to withdraw the savings after block 20, but if that option is chosen, then ``participant 
-2'' will not pay the extra 20 ADA to ``participant 1''.
-
-We will start from outer to inner primitives. The first thing is to allow ``participant 1'' to commit the 100 ADA. The 
-primitive says that with identity \haskellinline{IdentCC 1} (this is the identity of the commitment), ``participant 
-\haskellinline{1}'' can commit \haskellinline{100} ADA before block \haskellinline{10}, and the commitment must 
+following example: participant \haskellinline{2} offers 20 ADA to participant \haskellinline{1} in exchange for saving 100 ADA for 100 blocks. 
+participant \haskellinline{1} will be able to withdraw the savings after block 20, but if that option is chosen, then participant 
+ \haskellinline{2} will not pay the extra 20 ADA to participant \haskellinline{1}.
+
+We will start from outer to inner primitives. The first thing is to allow participant \haskellinline{1} to commit the 100 ADA. The 
+primitive says that with identity \haskellinline{IdentCC 1} (this is the identity of the commitment), participant 
+\haskellinline{1} can commit \haskellinline{100} ADA before block \haskellinline{10}, and the commitment must 
 expire on block \haskellinline{200}. But if the money is not committed before block \haskellinline{10}, then the 
@@
 If the commitment is done correctly and on time, we continue as in \haskellinline{step2}. \haskellinline{step2} is 
-another commitment, this time by ``participant 2''. The primitive says that with identity \haskellinline{IdentCC 2},
-``participant \haskellinline{2}'' can commit \haskellinline{20} ADA before block \haskellinline{20}, and the commitment
+another commitment, this time by participant \haskellinline{2}. The primitive says that with identity \haskellinline{IdentCC 2},
+participant \haskellinline{2} can commit \haskellinline{20} ADA before block \haskellinline{20}, and the commitment
 must expire on block \haskellinline{200}. But if the money is not committed before block \haskellinline{20}, then the
-contract will continue as \haskellinline{RedeemCC (IdentCC 1) Null}, which in turn gives ``participant 1'' the 
+contract will continue as \haskellinline{RedeemCC (IdentCC 1) Null}, which in turn gives participant \haskellinline{1} the 
 opportunity to withdraw all the money committed immediately.
@@
 If the commitment is done correctly and on time, we continue as in \haskellinline{step3}. \haskellinline{step3}
-waits until choice with identity \haskellinline{IdentChoice 1} is made by ``participant \haskellinline{1}'' or
+waits until choice with identity \haskellinline{IdentChoice 1} is made by participant \haskellinline{1} or
 until block \haskellinline{100} (whatever happens first). If the choice is made before block \haskellinline{100}, 
@@
 
-\haskellinline{step4} pays the \haskellinline{20} ADA committed by ``participant \haskellinline{2}'' in the commit with 
-identity \haskellinline{IdentPay 1} to ``participant \haskellinline{1}'', if the payment is claimed before block 
+\haskellinline{step4} pays the \haskellinline{20} ADA committed by participant \haskellinline{2} in the commit with 
+identity \haskellinline{IdentPay 1} to participant \haskellinline{1}, if the payment is claimed before block 
 \haskellinline{200}. Once the payment is claimed or after block 200, it continues as \haskellinline{stepRedeem}.
@@
 
-In this section, we give an example of how we can have Marlowe contracts that are easier to write, read, and 
-understand, by embedding them into Haskel code (taking advantage that Marlowe contracts are implemented as Haskell 
-terms).
-
-The example contract we use through this section implements an escrow contract. Let us imagine that ``participant 1'' 
-is buying some item from ``participant 2''. On one hand, if ``participant 1'' pays before receiving the item, there 
-is a risk that ``participant 2'' will never give ``participant 1'' the item. On the other hand, if ``participant 2'' 
-gives ``participant 1'' the item before ``participant 1'' pays, then there is a risk that ``participant 1'' will never 
+In this section, we revisit the escrow example that we discussed briefly in the introduction, and show  how we can make Marlowe contracts that are easier to write, read, and 
+understand, by embedding them into Haskell code, taking advantage of the fact that Marlowe contracts are implemented as Haskell 
+terms.
+
+The example contract we use through this section implements an escrow contract. Let us imagine that participant \haskellinline{1} 
+is buying some item from participant \haskellinline{2}. On one hand, if participant \haskellinline{1} pays before receiving the item, there 
+is a risk that participant \haskellinline{2} will never give \haskellinline{1} the item. On the other hand, if participant \haskellinline{2} 
+gives participant \haskellinline{1} the item before \haskellinline{1} pays, then there is a risk that \haskellinline{1} will never 
 pay.
 
-The escrow mechanism allows ``participant 1'' to deposit the money into a contract, in a way that the money will only 
-be released when two out of three participants agree on whether ``participant 2'' has indeed given ``participant 1'' 
-the item. The third participant (``participant 3'') is supposed to be a neutral third party that will decide in case of 
+The escrow mechanism allows participant \haskellinline{1} to deposit the money into a contract, in a way that the money will only 
+be released when two out of three participants agree on whether participant \haskellinline{2} has indeed given participant \haskellinline{1} 
+the item. The third participant (participant \haskellinline{3}) is supposed to be a neutral third party that will decide in case of 
 dispute.
 
-This way, if ``participant 1'' and ``participant 2'' are honest, they will just agree on the result of the transaction 
-and ``participant 3'' will not need to do anything. If ``participant 1'' and ``participant 2'' disagree, ``participant 
-3'' will be able to choose whether the money must go to ``participant 1'' or to ``participant 2''.
+This way, if participants \haskellinline{1} and \haskellinline{2} are honest, they will just agree on the result of the transaction 
+and \haskellinline{3} will not need to do anything.
+If \haskellinline{1} and \haskellinline{2} disagree, participant \haskellinline{3} will be able to choose whether the money must go to \haskellinline{1} or to \haskellinline{2}.
 
 In addition to this, in our implementation, the money paid for the item is 450 ADA, it must be committed by 
-``participant 1'' before block 10, and it will be refunded to ``participant 1'' if there is no consensus before block 
-90.
+participant \haskellinline{1} before block \haskellinline{10}, and it will be refunded to participant \haskellinline{1} if there is no consensus before block 
+\haskellinline{90}.
 
@@
 \end{haskellcode}
-
 And an identifier for the payment:
-
 \begin{haskellcode}
@@
 \end{haskellcode}
-
-We can also create a sub-contract that allows the money to be redeemed by ``participant 1'':
-
+We can also create a sub-contract that allows the money to be redeemed by participant \haskellinline{1}:
 \begin{haskellcode}
@@
 \end{haskellcode}
-
-Each participant has a say on who deserves the money: either ``participant 1'' deserves a refund, which we represent 
-with the number $0$; or ``participant 2'' deserves a payment, which we represent with the number $1$. Because there is 
+Each participant has a say on who deserves the money: either participant \haskellinline{1} deserves a refund, which we represent 
+with the number $0$; or participant \haskellinline{2} deserves a payment, which we represent with the number $1$. Because there is 
 only one choice to make per participant, we use the same \haskellinline{IdentChoice} as their participant id. We can 
@@
 chosen the number \haskellinline{c} as follows:
-
 \begin{haskellcode}
@@
 \end{haskellcode}
-
 Then we can easily write a function that returns an observation that is true if and only if at least one of the 
 participants \haskellinline{per} and \haskellinline{per'} has chosen the number \haskellinline{c} as follows:
-
 \begin{haskellcode}
@@
 \end{haskellcode}
-
 After that, we can write a function that returns an observation that is true if and only if at least 
@@
 \haskellinline{c} as follows:
-
 \begin{haskellcode}
@@
 \end{haskellcode}
-
-Finally, we can write the escrow contract, which would remain as follows:
-
+Finally, we can write the escrow contract, thus:
 \begin{haskellcode}
@@
 
-The outermost primitive \haskellinline{CommitCash} allows the ``participant 1'' to commit 450 ADA before block 10, 
+The outermost primitive \haskellinline{CommitCash} allows the participant \haskellinline{1} to commit 450 ADA before block 10, 
 with the promise that money will be released on block 100 if they are not claimed before that.
@@
 \begin{enumerate}
- \item Two out of three choose $0$.
- \item Two out of three choose $1$.
+ \item Two out of three choose $0$, for \haskellinline{pay}. 
+ \item Two out of three choose $1$, for \haskellinline{refund}.
  \item The block 90 is added to the blockchain.
@@
 If the third option happens first, the money is refunded. Otherwise, there is a \haskellinline{Choice} that will 
-immediately refund the money to ``participant 1'' unless two out of three chose $1$; in the later case, 
-\haskellinline{Pay} will give ``participant 2'' the opportunity to claim the funds available in \haskellinline{iCC1} 
-before block 100. If the funds are not claimed before block 100 they will also be refunded to ``participant 1''.
+immediately refund the money to participant \haskellinline{1} unless two out of three chose option $1$; in the later case, 
+\haskellinline{Pay} will give participant \haskellinline{2} the opportunity to claim the funds available in \haskellinline{iCC1} 
+before block 100. If the funds are not claimed before block 100 they will also be refunded to participant \haskellinline{1}.
 

### 2018-04-15 tex/paper.tex
Added discussion of where to end of section 5

@@
 \begin{enumerate}
- \item Two out of three choose $0$, for \haskellinline{pay}. 
- \item Two out of three choose $1$, for \haskellinline{refund}.
+ \item Two out of three choose $0$, for `\haskellinline{pay}'. 
+ \item Two out of three choose $1$, for `\haskellinline{refund}'.
  \item The block 90 is added to the blockchain.
@@
 before block 100. If the funds are not claimed before block 100 they will also be refunded to participant \haskellinline{1}.
+
+Reflecting on the example we can see that using Haskell definitions has made the contract substantially more comprehensible; it will also support more efficient operation, if we were to replace the repeated  expression \haskellinline{two_chose 1 2 3 1} above by a \haskellinline{where} clause,
+\begin{haskellcode}
+escrow = ...
+                    (When (OrObs (two_chose 1 2 3 0)
+                                 chose_refund)
+                          90
+                          (Choice chose_refund
+                           ...
+         ...
+         where 
+         chose_refund = two_chose 1 2 3 1                 
+\end{haskellcode}
+then the repeated computation of the expression will be avoided.
 

### 2018-04-24 src/Semantics.hs
Tentative refactoring of semantics

@@ -389,31 +389,32 @@ expireCommits inp scf os = (Map.union uexp nsc, pas)
         pas = [ExpiredCommitRedeemed ident p val | (ident, (p, NotRedeemed val _)) <- Map.toList expi]
         et = blockNumber os
         uexp = Map.map markRedeemed expi
--- Wraps step function to refund expired cash commitments
-
-fullStep :: Input -> State -> Contract -> OS -> (State, Contract, AS)
-
-fullStep inp st con os = (rs, rcon, nas)
-  where (nsch, chas) = recordChoices inp (sch st)
-        (nsc, pas) = expireCommits inp (sc st) os
-        nst = st { sc = nsc, sch = nsch }
-        nas = chas ++ pas ++ ras
-        (rs, rcon, ras) = step inp nst con os
 
 -- Repeatedly calls the step function (fullStep function actually) until
 -- it does not change anything or produces any actions
 
+stepAll :: Input -> State -> Contract -> OS -> (State, Contract, AS)
+
+stepAll com st con os = stepAllAux com st con os []
+
+stepAllAux :: Input -> State -> Contract -> OS -> AS -> (State, Contract, AS)
+
+stepAllAux com st con os ac
+  | (nst == st) && (ncon == con) && null nac = (st, con, ac)
+  | otherwise = stepAllAux com nst ncon os (nac ++ ac)
+  where
+    (nst, ncon, nac) = step com st con os
+
+-- Wraps step function to refund expired cash commitments
+
 computeAll :: Input -> State -> Contract -> OS -> (State, Contract, AS)
 
-computeAll com st con os = computeAllAux com st con os []
-
-computeAllAux :: Input -> State -> Contract -> OS -> AS -> (State, Contract, AS)
-
-computeAllAux com st con os ac
-  | (nst == st) && (ncon == con) && null nac = (st, con, ac)
-  | otherwise = computeAllAux com nst ncon os (nac ++ ac)
-  where
-    (nst, ncon, nac) = fullStep com st con os
+computeAll inp st con os = (rs, rcon, nas)
+  where (nsch, chas) = recordChoices inp (sch st)
+        (nsc, pas) = expireCommits inp (sc st) os
+        nst = st { sc = nsc, sch = nsch }
+        nas = chas ++ pas ++ ras
+        (rs, rcon, ras) = stepAll inp nst con os
 
 -------------------------
 -- Auxiliary functions --
@@ -510,7 +511,7 @@ driver start_state contract input =
  case input of
[…]

### 2018-04-24 src/Semantics.hs
Fix comments

@@ -333,9 +333,9 @@ step commits st c@(RedeemCC ident con) _ =
     where
         ccs = sc st
 
----------------------------
--- fullStep & computeAll --
----------------------------
+--------------------------
+-- stepAll & computeAll --
+--------------------------
 
 -- Given a choice, if no previous choice for its id has been recorded,
 -- it records it in the map, and adds an action ChoiceMade to the list.
@@ -390,7 +390,7 @@ expireCommits inp scf os = (Map.union uexp nsc, pas)
         et = blockNumber os
         uexp = Map.map markRedeemed expi
 
--- Repeatedly calls the step function (fullStep function actually) until
+-- Repeatedly calls the step function until
 -- it does not change anything or produces any actions
 
 stepAll :: Input -> State -> Contract -> OS -> (State, Contract, AS)
@@ -405,7 +405,8 @@ stepAllAux com st con os ac
   where
     (nst, ncon, nac) = step com st con os
 
--- Wraps step function to refund expired cash commitments
+-- Wraps stepAll function to carry out actions that need to be
+-- done once per block (refund expired cash commitments, and record choices)
 
 computeAll :: Input -> State -> Contract -> OS -> (State, Contract, AS)
 

### 2018-04-24 tex/paper.tex
Fix some typos

@@
 
-In doing this we are required to refine the model of contracts in a number of ways in order to fit with a radically different context.
-Let us the following example of an ``escrow'' contract we can explain the motivation more concretely.
+In doing this, we are required to refine the model of contracts in a number of ways in order to fit with a radically different context.
+Let us consider the following example of an ``escrow'' contract so that we can explain the motivation more concretely.
 The aim of this contract, written in functional pseudocode in the style of~\cite{PeytonJones:2000} involves three participants \haskellinline{1}, \haskellinline{2} and \haskellinline{3}.
@@
 Participant \haskellinline{1} pays the money into escrow controlled by participant \haskellinline{3}.
-There are two options for the money: if two participants agree to \haskellinline{pay} it to \haskellinline{2}, that goes ahead; if, on the other hand, two of the players opt to \haskellinline{refund} the money to \haskellinline{1}, that is done instead.
-The \haskellinline{When} construct ensures fires when its first argument -- a condition -- becomes true; the second argument, a \haskellinline{Contract} itself, is then performed, and this makes the payment according to the players' \haskellinline{Choice}.
+There are two options for the money: if two participants agree to \haskellinline{pay} it to \haskellinline{2}, that goes ahead; if, on the other hand, two of the participants opt to \haskellinline{refund} the money to \haskellinline{1}, that is done instead.
+The \haskellinline{When} construct fires when its first argument -- a condition -- becomes true; the second argument, a \haskellinline{Contract} itself, is then performed, and this makes the payment according to the participants' \haskellinline{Choice}.
 
@@
 In the traditional model, enforcement of the contract is the responsibility of the legal system.
-If participant \haskellinline{1} doesn't pay the money into escrow, or participant \haskellinline{3} chooses to keep it for herself, then we can assume that they can be sued for the money and probably damages on top, thus providing both legal and financial incentives for compliance.
+If participant \haskellinline{1} does not pay the money into escrow, or participant \haskellinline{3} chooses to keep it for herself, then we can assume that they can be sued for the money and probably damages on top, thus providing both legal and financial incentives for compliance.
 On the other hand, in the decentralised blockchain model, where there is no central authority, the contract needs to be enforced by design.
@@
 This means that we require participants to commit money to cover all possible expenditure \emph{in advance of the contract executing}.
-In order to make sure that player continue to engage with a contract we ensure urgency by imposing timeouts: money is committed for a finite period, for example, and when waiting for a player to make a commitment we also impose a timeout to ensure that the contract does not become stuck even if one or more of the participants stops interacting with it.
+In order to make sure that participants continue to engage with a contract, we ensure urgency by imposing timeouts: money is committed for a finite period, for example, and when waiting for a participant to make a commitment we also impose a timeout to ensure that the contract does not become stuck even if one or more of the participants stop interacting with it.
 We revisit this example in Section~\ref{section:example-escrow} where we show how commitments, timeouts and other features of Marlowe are used to describe this as a self-enforcing contract ``on the blockchain''.
@@
 API, which is accessible from Nxt clients through a REST interface. The API provides functionality supporting various 
-kinds of transactions. The core software itself doesn't support any form of scripting language; rather, users are 
+kinds of transactions. The core software itself does not support any form of scripting language; rather, users are 
 expected to work with the built in transaction types and transactions that support some 250 primitive operations; these 
@@
 Alternatively, in the \emph{pull} model, contracts enable certain things to happen, such as transactions, but the 
-transactions need to be effected by an external player. The pull model makes fewer demands on the blockchain 
+transactions need to be effected by an external actor. The pull model makes fewer demands on the blockchain 
 implementation than the push model, which requires the contract to be triggered automatically.

### 2018-04-24 tex/paper.tex
Fix explanation of wrapper functions

@@
 \item It is important to be aware that there may be different instances of contracts executed, and for each instance there will typically be different instances of observables (and commitments) observed and recorded.
-\item The model manages the release of funds back to the committer when a cash commitment expires (see discussion of the \haskellinline{fullStep} function below).
+\item The model manages the release of funds back to the committer when a cash commitment expires (see discussion of the \haskellinline{computeAll} function below).
 \end{itemize}
@@
 
-At each computational step, the \haskellinline{step} function is wrapped by a call to the \haskellinline{fullStep} function, which in addition to performing \haskellinline{step} will enable expired cash commitments to be refunded and record any choices made at that step.
+The function \haskellinline{step} is wrapped by the \haskellinline{stepAll} function, which calls \haskellinline{step} until it stops having any effects.
+For every block, the \haskellinline{computeAll} function is called.
+In addition to calling \haskellinline{stepAll}, \haskellinline{computeAll} will first enable expired cash commitments to be refunded and record any choices made at that step.
 
@@
 If the cash is committed successfully and the timeout \haskellinline{t2} is reached, then it is impossible to further spend the committed cash, and any unspent funds can be reclaimed by participant \haskellinline{p}.
- This is performed by the \haskellinline{fullStep} function, as noted above.
+ This is performed by the \haskellinline{computeAll} function, as noted above.
 
@@
 
-If the cash commitment with identifier \haskellinline{idCC} has expired, it becomes possible for the remaining funds to be redeemed by the committer; this can be done by the \haskellinline{fullStep} function processing the appropriate \haskellinline{Input}, and an \haskellinline{ExpiredCommitRedeemed} action will be produced.
+If the cash commitment with identifier \haskellinline{idCC} has expired, it becomes possible for the remaining funds to be redeemed by the committer; this can be done by the \haskellinline{computeAll} function processing the appropriate \haskellinline{Input}, and an \haskellinline{ExpiredCommitRedeemed} action will be produced.
 

### 2018-04-24 tex/paper.tex
Rename computeAll to stepBlock and other fixes

@@
 \item It is important to be aware that there may be different instances of contracts executed, and for each instance there will typically be different instances of observables (and commitments) observed and recorded.
-\item The model manages the release of funds back to the committer when a cash commitment expires (see discussion of the \haskellinline{computeAll} function below).
+\item The model manages the release of funds back to the committer when a cash commitment expires (see discussion of the \haskellinline{stepBlock} function below).
 \end{itemize}
@@
 The function \haskellinline{step} is wrapped by the \haskellinline{stepAll} function, which calls \haskellinline{step} until it stops having any effects.
-For every block, the \haskellinline{computeAll} function is called.
-In addition to calling \haskellinline{stepAll}, \haskellinline{computeAll} will first enable expired cash commitments to be refunded and record any choices made at that step.
+For every block, the \haskellinline{stepBlock} function is called.
+In addition to calling \haskellinline{stepAll}, \haskellinline{stepBlock} will first enable expired cash commitments to be refunded and record any choices made at that step.
 
@@
 If the cash is committed successfully and the timeout \haskellinline{t2} is reached, then it is impossible to further spend the committed cash, and any unspent funds can be reclaimed by participant \haskellinline{p}.
- This is performed by the \haskellinline{computeAll} function, as noted above.
+ This is performed by the \haskellinline{stepBlock} function, as noted above.
 
@@
 
-If the cash commitment with identifier \haskellinline{idCC} has expired, it becomes possible for the remaining funds to be redeemed by the committer; this can be done by the \haskellinline{computeAll} function processing the appropriate \haskellinline{Input}, and an \haskellinline{ExpiredCommitRedeemed} action will be produced.
+If the cash commitment with identifier \haskellinline{idCC} has expired, it becomes possible for the remaining funds to be redeemed by the committer; this can be done by the \haskellinline{stepBlock} function processing the appropriate \haskellinline{Input}, and an \haskellinline{ExpiredCommitRedeemed} action will be produced.
 

### 2018-04-24 meadow/src/Main.hs

@@ -448,7 +448,7 @@ execute =
      st <- get_state
      contr <- workspace_to_contract
      blk <- get_blocknum
-     let (nst, ncontr, outp) = computeAll inp st contr (OS {random = 42, blockNumber = blk}) in do {set_output outp;
+     let (nst, ncontr, outp) = stepBlock inp st contr (OS {random = 42, blockNumber = blk}) in do {set_output outp;
                          code_to_contract ncontr;
                          set_state nst;
                          set_input_text "([], [], [], [])";

### 2018-04-24 src/Analysis.hs

@@ -1,541 +1,0 @@
-module Analysis where
-
-import Semantics
-
-import qualified Data.Maybe as Maybe
-import qualified Data.Set as Set
-import qualified Data.Map as Map
-import qualified Data.List as List
-
-----------------------
--- Interval algebra --
-----------------------
-
-type Interval = (Maybe Int, Maybe Int)
-type Domain = [Interval]
-
-
--- Comparison of Maybe Ints representing potential infinites
--- a double var in a side means a Nothing on that side
--- represents -inf otherwise it represents +inf
-
-(|<=|) :: Maybe Int -> Maybe Int -> Bool
-(|<=|) (Just a) (Just b) = a <= b
-(|<=|) Nothing (Just _) = False
-(|<=|) _ _ = True
-
-(||<=|) :: Maybe Int -> Maybe Int -> Bool
-(||<=|) (Just a) (Just b) = a <= b
-(||<=|) _ _ = True
-
-(|<=||) :: Maybe Int -> Maybe Int -> Bool
-(|<=||) (Just a) (Just b) = a <= b
-(|<=||) _ _ = False
-
-(||<=||) :: Maybe Int -> Maybe Int -> Bool
-(||<=||) (Just a) (Just b) = a <= b
-(||<=||) Nothing _ = True
-(||<=||) _ _ = False
-
-(|>|) :: Maybe Int -> Maybe Int -> Bool
-(|>|) a b = not $ (|<=|) a b
-
-(||>|) :: Maybe Int -> Maybe Int -> Bool
-(||>|) a b = not $ (||<=|) a b
-
-(|>||) :: Maybe Int -> Maybe Int -> Bool
-(|>||) a b = not $ (|<=||) a b
-
-(||>||) :: Maybe Int -> Maybe Int -> Bool
-(||>||) a b  = not $ (||<=||) a b
-
--- Absurd domain detection
-
-possible :: Interval -> Bool
-possible (low, high) = low ||<=| high
-
-maybePossible :: Interval -> Maybe Interval
-maybePossible i = if (possible i) then Just i else Nothing
-
-
--- Example extraction
-
-get_one :: Domain -> Int
-get_one [] = error "Empty domain on get_one!"
-get_one (h:_) = case h of
-                  (Nothing, Nothing) -> 0
-                  (Just x, _) -> x
-                  (Nothing, Just y) -> y
-
--- Get first element in Domain
-
-
--- Get next element in Domain
[…]

### 2018-04-24 src/Semantics.hs

@@ -17,7 +17,7 @@ import qualified Data.Maybe as Maybe
  --     - type of State (of the internal state of a DSL contract evaluation)                    --
  --     - type of Observations (of values of Observables)                                       --
  --     - type of Contracts                                                                     --
- --     - single step evaluation (the step function) which is wrapped by fullStep …             --
+ --     - single step evaluation (the step function) which is wrapped by stepBlock …            --
  --     - … which expires and refunds cash commitments                                          --
  --                                                                                             --
  -- Further discussion in accompanying document.                                                --
@@ -333,9 +333,9 @@ step commits st c@(RedeemCC ident con) _ =
     where
         ccs = sc st
 
---------------------------
--- stepAll & computeAll --
---------------------------
+-------------------------
+-- stepAll & stepBlock --
+-------------------------
 
 -- Given a choice, if no previous choice for its id has been recorded,
 -- it records it in the map, and adds an action ChoiceMade to the list.
@@ -408,9 +408,9 @@ stepAllAux com st con os ac
 -- Wraps stepAll function to carry out actions that need to be
 -- done once per block (refund expired cash commitments, and record choices)
 
-computeAll :: Input -> State -> Contract -> OS -> (State, Contract, AS)
+stepBlock :: Input -> State -> Contract -> OS -> (State, Contract, AS)
 
-computeAll inp st con os = (rs, rcon, nas)
+stepBlock inp st con os = (rs, rcon, nas)
   where (nsch, chas) = recordChoices inp (sch st)
         (nsc, pas) = expireCommits inp (sc st) os
         nst = st { sc = nsc, sch = nsch }
@@ -499,7 +499,7 @@ lowerExpirationDateButNotExpired _ _ = EQ
  - Driver -
  ----------}
 
--- Driver for a single step of execution, using the fullStep function
+-- Driver for a single step of execution, using the stepBlock function
 -- This first performs any repayments due because of an expired commit,
 -- then calls the step function.
 
@@ -512,7 +512,7 @@ driver start_state contract input =
  case input of
   [] -> error "Input should be infinite in driver"
   (com1,os1):rest_input ->
-    let (next_st,next_con,aset) = computeAll com1 start_state contract os1 in
+    let (next_st,next_con,aset) = stepBlock com1 start_state contract os1 in
     let rest                    = driver next_st next_con rest_input in aset:rest
 
 

### 2018-04-24 src/SmartInputs.hs

@@ -78,7 +78,7 @@ getPotentialInputsFromState os State {sc = commits} =
 -- It takes all choices as "0", independently of what would be useful
 getPossibleInputs :: OS -> Input -> Contract -> State -> Input
 getPossibleInputs os inp contr stat = foldl1 combineInputs [inputsFromContract, inputsFromState, choices]
-  where (statGivenInputs, contractGivenInputs, _) = computeAll inp stat contr os
+  where (statGivenInputs, contractGivenInputs, _) = stepBlock inp stat contr os
         inputsFromContract = getPotentialInputsFromContract os contractGivenInputs statGivenInputs
         inputsFromState = getPotentialInputsFromState os statGivenInputs
         choices = getUsedChoiceNumbers statGivenInputs contractGivenInputs

### 2018-04-24 src/TreeAnalysis.hs

@@ -1,100 +1,0 @@
-module TreeAnalysis where
-
-import Analysis
-import Semantics
-import TreeVisualise
-import qualified Data.Map as Map
-import qualified Data.List as List
-
------------------------
--- Tree construction --
------------------------
-
-type ATran = [Event]
-type ANode = (Domain, [Action])
-
-data Acc = Acc {
-                 node_info :: Map.Map Int AnalysisState,
-                 node_info_rev :: Map.Map AnalysisState Int,
-                 transitions :: [(Int, Int, ATran)],
-                 nodes :: [(Int, ANode)],
-                 last_node_id :: Int
-               }
-           deriving (Eq,Ord,Show,Read)
-
-only_element_or_nothing :: [[Event]] -> [Event]
-only_element_or_nothing [] = []
-only_element_or_nothing [x] = x 
-only_element_or_nothing _ = error ("Several elements in only_element_or_nothing!")
-
-get_tree_aux :: [(Int, AnalysisState)] -> Acc -> Acc
-get_tree_aux [] acc = acc
-get_tree_aux ((parent, curr_as):t) (acc@(Acc {transitions = tra,
-                                              nodes = nods,
-                                              node_info = ni,
-                                              node_info_rev = nirev,
-                                              last_node_id = lnid}))
-  | is_rep curr_as = get_tree_aux t acc
-  | List.null diff && List.null diff_act =
-                        get_tree_aux ([(parent, new_as)
-                                       | new_as <- (analysis_step [curr_as]),
-                                         not $ is_rep new_as] ++ t) new_acc
-  | otherwise = get_tree_aux ((new_node_id, curr_as):t) new_node_acc
-  where curr_eve = get_events curr_as
-        past_eve = get_events (Map.findWithDefault err parent $ node_info acc)
-        diff = curr_eve List.\\ past_eve
-        curr_act = get_actions curr_as
-        past_act = get_actions (Map.findWithDefault err parent $ node_info acc)
-        diff_act = curr_act List.\\ past_act
-        err = (error "Parent does not exist in get_tree_aux!")
-        is_rep x = Map.member x nirev 
-        new_node_acc = acc {last_node_id = new_node_id,
-                            transitions = ((parent, new_node_id, (only_element_or_nothing diff)):tra),
-                            nodes = ((new_node_id, (possible_block curr_as,
-                                                    diff_act)):nods),
[…]

### 2018-04-24 src/TreeVisualise.hs

@@ -1,54 +1,0 @@
-module TreeVisualise(GraphData, visualise, renderToFile) where
-
-import qualified Data.Graph.Inductive.PatriciaTree as PT
-import Data.Graph.Inductive.Graph(mkGraph)
-import Data.GraphViz.Attributes.Complete(Attribute(..),Label(StrLabel))
-import Data.GraphViz(Labellable(..),Attributes,DotGraph,GraphvizParams(..),
-                     GraphvizCommand(..),runGraphvizCommand,GraphvizCanvas(..),
-                     GraphvizOutput(..),nonClusteredParams,graphToDot,runGraphvizCanvas')
-import Data.Text.Lazy (pack)
-
-type GraphData = ([(Int, String)], [(Int, Int, String)])
-
-getStates :: GraphData -> [(Int, DNode)]
-getStates (a, _) = map (\(x, y) -> (x, DNode y)) a
-
-getTransitions :: GraphData -> [(Int, Int, DTran)]
-getTransitions (_, b) = map (\(x, y, z) -> (x, y, DTran z)) b
-
-toGr :: GraphData -> PT.Gr DNode DTran
-toGr x = mkGraph (getStates x) (getTransitions x)
-
-data DNode = DNode String
-  deriving (Eq, Show, Ord)
-
-instance Labellable DNode where
-  toLabelValue (DNode s) = StrLabel (pack s)
-
-data DTran = DTran String
-  deriving (Eq, Show, Ord)
-
-instance Labellable DTran where
-  toLabelValue (DTran s) = StrLabel (pack s)
-
-
-ncp :: GraphvizParams Int DNode DTran () DNode
-ncp = nonClusteredParams { fmtNode = \(_, n) -> decorateNode n
-                         , fmtEdge = \(_, _, e) -> decorateEdge e }
-
-toDot :: PT.Gr DNode DTran -> DotGraph Int 
-toDot = graphToDot ncp
-
-visualise :: GraphData -> IO ()
-visualise gd = runGraphvizCanvas' (toDot (toGr gd)) Xlib
-
-renderToFile :: FilePath -> GraphData -> IO FilePath
-renderToFile path gd = runGraphvizCommand Dot (toDot (toGr gd)) DotOutput path
-
-decorateEdge :: DTran -> Attributes
-decorateEdge (DTran x) = [ Label $ StrLabel $ pack x ]
-
-decorateNode :: DNode -> Attributes
-decorateNode (DNode x) = [ Label $ StrLabel $ pack x ]
-
-

### 2018-05-14 tex/paper.tex
Add embedded Marlowe explanation to paper

@@
 
-In this section, we revisit the escrow example that we discussed briefly in the introduction, and show  how we can make Marlowe contracts that are easier to write, read, and 
-understand, by embedding them into Haskell code, taking advantage of the fact that Marlowe contracts are implemented as Haskell 
-terms.
-
-The example contract we use through this section implements an escrow contract. Let us imagine that participant \haskellinline{1} 
-is buying some item from participant \haskellinline{2}. On one hand, if participant \haskellinline{1} pays before receiving the item, there 
-is a risk that participant \haskellinline{2} will never give \haskellinline{1} the item. On the other hand, if participant \haskellinline{2} 
-gives participant \haskellinline{1} the item before \haskellinline{1} pays, then there is a risk that \haskellinline{1} will never 
-pay.
-
-The escrow mechanism allows participant \haskellinline{1} to deposit the money into a contract, in a way that the money will only 
-be released when two out of three participants agree on whether participant \haskellinline{2} has indeed given participant \haskellinline{1} 
-the item. The third participant (participant \haskellinline{3}) is supposed to be a neutral third party that will decide in case of 
-dispute.
-
-This way, if participants \haskellinline{1} and \haskellinline{2} are honest, they will just agree on the result of the transaction 
-and \haskellinline{3} will not need to do anything.
-If \haskellinline{1} and \haskellinline{2} disagree, participant \haskellinline{3} will be able to choose whether the money must go to \haskellinline{1} or to \haskellinline{2}.
-
-In addition to this, in our implementation, the money paid for the item is 450 ADA, it must be committed by 
-participant \haskellinline{1} before block \haskellinline{10}, and it will be refunded to participant \haskellinline{1} if there is no consensus before block 
-\haskellinline{90}.
+In this section, we revisit the escrow example that we discussed briefly in the introduction, and show how we can make 
+Marlowe contracts that are easier to write, read, and understand, by embedding them into Haskell code, taking advantage 
+of the fact that Marlowe contracts are implemented as Haskell terms. We used Haskell because it is the language in 
+which Marlowe is implemented, but it would be easy to embed Marlowe in any other languages. It would only be necessary 
+to translate its primitives into a data type in that language. In Meadow, we use Fay embedding, which is a subset of 
+Haskell (see Section~\ref{section:tool}).
+
+The example contract we use through this section implements an escrow contract. Let us imagine that participant 
+\haskellinline{1} is buying some item from participant \haskellinline{2}. On one hand, if participant \haskellinline{1} 
+pays before receiving the item, there is a risk that participant \haskellinline{2} will never give \haskellinline{1} the 
+item. On the other hand, if participant \haskellinline{2} gives participant \haskellinline{1} the item before 
+\haskellinline{1} pays, then there is a risk that \haskellinline{1} will never pay.
+
+The escrow mechanism allows participant \haskellinline{1} to deposit the money into a contract, in a way that the money 
+will only be released when two out of three participants agree on whether participant \haskellinline{2} has indeed given 
+participant \haskellinline{1} the item. The third participant (participant \haskellinline{3}) is supposed to be a 
+neutral third party that will decide in case of dispute.
+
+This way, if participants \haskellinline{1} and \haskellinline{2} are honest, they will just agree on the result of the 
+transaction and \haskellinline{3} will not need to do anything. If \haskellinline{1} and \haskellinline{2} disagree, 
+participant \haskellinline{3} will be able to choose whether the money must go to \haskellinline{1} or to 
+\haskellinline{2}.
+
+In addition to this, in our implementation, the money paid for the item is 450 ADA, it must be committed by participant 
+\haskellinline{1} before block \haskellinline{10}, and it will be refunded to participant \haskellinline{1} if there is 
+no consensus before block \haskellinline{90}.
 
@@
 \end{haskellcode}
-Each participant has a say on who deserves the money: either participant \haskellinline{1} deserves a refund, which we represent 
-with the number $0$; or participant \haskellinline{2} deserves a payment, which we represent with the number $1$. Because there is 
-only one choice to make per participant, we use the same \haskellinline{IdentChoice} as their participant id. We can 
-define a function that returns an observation that is true if and only if the participant \haskellinline{per} has 
-chosen the number \haskellinline{c} as follows:
+Each participant has a say on who deserves the money: either participant \haskellinline{1} deserves a refund, which we 
+represent with the number $0$; or participant \haskellinline{2} deserves a payment, which we represent with the number 
+$1$. Because there is only one choice to make per participant, we use the same \haskellinline{IdentChoice} as their 
+participant id. We can define a function that returns an observation that is true if and only if the participant 
+\haskellinline{per} has chosen the number \haskellinline{c} as follows:
 \begin{haskellcode}
@@
 
-The outermost primitive \haskellinline{CommitCash} allows the participant \haskellinline{1} to commit 450 ADA before block 10, 
-with the promise that money will be released on block 100 if they are not claimed before that.
+The outermost primitive \haskellinline{CommitCash} allows the participant \haskellinline{1} to commit 450 ADA before 
+block 10, with the promise that money will be released on block 100 if they are not claimed before that.
 
@@
 If the third option happens first, the money is refunded. Otherwise, there is a \haskellinline{Choice} that will 
-immediately refund the money to participant \haskellinline{1} unless two out of three chose option $1$; in the later case, 
-\haskellinline{Pay} will give participant \haskellinline{2} the opportunity to claim the funds available in \haskellinline{iCC1} 
-before block 100. If the funds are not claimed before block 100 they will also be refunded to participant \haskellinline{1}.
-
-Reflecting on the example we can see that using Haskell definitions has made the contract substantially more comprehensible; it will also support more efficient operation, if we were to replace the repeated  expression \haskellinline{two_chose 1 2 3 1} above by a \haskellinline{where} clause,
+immediately refund the money to participant \haskellinline{1} unless two out of three chose option $1$; in the later 
+case, \haskellinline{Pay} will give participant \haskellinline{2} the opportunity to claim the funds available in 
+\haskellinline{iCC1} before block 100. If the funds are not claimed before block 100 they will also be refunded to 
+participant \haskellinline{1}.
+
+Reflecting on the example we can see that using Haskell definitions has made the contract substantially more 
+comprehensible; it will also support more efficient operation, if we were to replace the repeated  expression 
+\haskellinline{two_chose 1 2 3 1} above by a \haskellinline{where} clause, 
 \begin{haskellcode}
@@
 than a general purpose language would. With this aim, and to aid Marlowe's
-usage by people that does not know is not familiar with its syntax,
+usage by people that does not know or is not familiar with its syntax,
 we have developed Meadow, a web tool that supports the interactive construction,
@@
 Meadow has been mainly written in Haskell and compiled to JavaScript
-by the Haste compiler~\cite{Haste}, it also relies on
-the Blockly library~\cite{Blockly} for providing a visual
-editor for smart contracts written in Marlowe. Blockly's editor allows
-the user to visualise and edit smart-contracts as interlocking blocks
-that can be dragged and dropped like pieces of a jigsaw puzzle (see
-Figure~\ref{fig:detail-of-block}). Meadow also provides functionality
-to generate syntactically correct and formatted code (that is displayed
-in the upper right corner of the application), and to convert the
-code back to its block representation.
+partially by the Haste compiler~\cite{Haste} and partially by GHCJS \cite{GHCJS};
+Meadow also relies on the Blockly library~\cite{Blockly} for providing a visual
+editor for smart contracts written in Marlowe. Embedding support in Meadow
+is provided throught a pruned and bundled version of the Fay compiler \cite{Fay},
+compiled to JavaScript using GHCJS. All text fields that edit and present Meadow and Fay code are
+instances of CodeMirror text editor \cite{CodeMirror}.
+
+Blockly's editor allows the user to visualise and edit
+smart-contracts as interlocking blocks that can be dragged and dropped
+like pieces of a jigsaw puzzle (see Figure~\ref{fig:detail-of-block}).
+Meadow also provides functionality to generate syntactically correct
+and formatted code (that is displayed in the upper right corner of the application),
+and to convert the code back to its block representation.
+
+Additionally, Meadow allows the user to use Meadow's Fay embedding from within the browser (see 
+Figure~\ref{fig:embedded-demo}). Meadow includes a pruned copy of the Fay compiler, bundled with its Prelude module, 
+its foreign interface, and a module with the definitions of the Marlowe primitives and a function to pretty-print 
+Marlowe contracts. This allows the user to compile and execute, inside the browser, Fay
+code that generates and prints a Meadow contract. When the user opens the embedded editor, the current Marlowe contract 
+is embedded into a template of a Fay program that prints it. The user is able to modify this code in the left 
+panel of the editor and use the advanced functionalities provided by Fay like, for example, bindings, 
+list-comprehensions, turing-complete functions definitions, etc.
+
+When the user clicks the execute button, the Fay code is 
+compiled to JavaScript and evaluated; this causes the Marlowe contract generated by the execution of the code to be 
+written to the panel on the right. Generated contracts can then be sent back to the main screen of Meadow, translated 
+to Blockly, and their execution can be simulated. The embedded editor also allows users to save the Fay code to a 
+file in their computers, to load code from a file in their computers, and to temporarily hide the editor while keeping 
+its contents in memory, even while the compilation and execution process is being carried out in the background.
 
@@
 
-In the next section we re-examine the model and the Marlowe language, looking at possible extensions, alternative design decisions and ways in which we can formally analyse Marlowe contracts.
+\begin{figure}
+\begin{centering}
+\includegraphics[width=1\textwidth]{/home/palas/Escritorio/scdsl/tex/pix/embedded.png}
+\par\end{centering}
+\caption{\label{fig:embedded-demo}Screenshot of Meadow's embedded Fay editor.}
+\end{figure}
+
+In the next section, we re-examine the model and the Marlowe language, looking at possible extensions, alternative 
+design decisions and ways in which we can formally analyse Marlowe contracts.
 

### 2018-05-16 tex/paper.tex
fixed path to .../pix/embedded.png
Adapt paper to LLNCS
Word-wrap source tex

@@
-\documentclass[
-      acmsmall
-    , screen
-%    , authordraft
-%    , anonymous
-%    , review=true
-  ]{acmart}
+\documentclass[runningheads]{llncs}
 
@@
 %\setcopyright{acmcopyright}
-\setcopyright{acmlicensed}
+%\setcopyright{acmlicensed}
 %\setcopyright{rightsretained}
@@
 
-\citestyle{acmauthoryear}
+%\citestyle{acmauthoryear}
 
@@
 
-%%% FIXME PACKAGE SETUP %%%
-
-\fxusetheme{color}
-\fxuseenvlayout{color}
-
-\FXRegisterAuthor{pls}{apls}{\color{teal}[Pablo]}
-\FXRegisterAuthor{sjt}{asjt}{\color{magenta}[Simon]}
-
 %%% MINTED COMMANDS FOR HASKELL %%%
@@
 
+\FXRegisterAuthor{pls}{apls}{\color{teal}[Pablo]}
+\FXRegisterAuthor{sjt}{asjt}{\color{magenta}[Simon]}
+
 %%% MACROS %%%
@@
 
+%
+%
 % Title portion. Note the short title for running heads
-\title
-  [Marlowe]
-  {Marlowe: financial contracts on blockchain}
-
-\author{Pablo Lamela Seijas}
-\affiliation{%
-  \institution{University of Kent}
-%  \streetaddress{School of Computing}
-  \city{Canterbury}
-  \state{Kent}
-  \postcode{CT2 7NF}
-  \country{UK}}
-\email{pl292@kent.ac.uk}
-
-\author{Simon Thompson}
-\orcid{0000-0002-2350-301X}
-\affiliation{%
-  \institution{University of Kent}
-%  \streetaddress{School of Computing}
-  \city{Canterbury}
-  \state{Kent}
-  \postcode{CT2 7NF}
-  \country{UK}}
-\email{s.j.thompson@kent.ac.uk}
+\title {Marlowe: financial contracts on blockchain\thanks{IOHK}}
+%
+%\titlerunning{Abbreviated paper title}
+% If the paper title is too long for the running head, you can set
+% an abbreviated paper title here
+%
+\author{Pablo Lamela Seijas\orcidID{0000-0002-1730-1219}\\
+\and Simon Thompson\orcidID{0000-0002-2350-301X}}
+%
+\authorrunning{P. Lamela Seijas and S. Thompson}
+% First names are abbreviated in the running head.
+% If there are more than two authors, 'et al.' is used.
+%
+\institute{University of Kent, Canterbury, CT2 7NF, Kent (UK)}
+\email{\{pl292,s.j.thompson\}@kent.ac.uk}
 
@@
 Blockchains allow the specification of contracts in the form of programs that guarantee their fulfilment. 
-Nevertheless, errors in those programs can cause important, and often irretrievable, monetary loss.
-General-purpose languages provide a platform on which contracts can be built, but by their very generality they have the potential to exhibit behaviours of an unpredictable kind, and are also not easy to read or comprehend for general users.
-
-An alternative solution is provided by domain-specific languages (DSLs), which are designed to express programs in a particular field.
-This paper explores the design of one DSL, Marlowe, targeted at the execution of financial
-contracts in the style of Peyton Jones \emph{et al} on blockchains. We have focused on predictability, enforceability, simplicity, 
-and ease of analysis and use.
+Nevertheless, errors in those programs can cause important, and often irretrievable, monetary loss. General-purpose 
+languages provide a platform on which contracts can be built, but by their very generality they have the potential to 
+exhibit behaviours of an unpredictable kind, and are also not easy to read or comprehend for general users.
+
+An alternative solution is provided by domain-specific languages (DSLs), which are designed to express programs in a 
+particular field. This paper explores the design of one DSL, Marlowe, targeted at the execution of financial 
+contracts in the style of Peyton Jones \emph{et al} on blockchains. We have focused on predictability, enforceability, 
+simplicity, and ease of analysis and use.
 
@@
 
-This paper explores the design of a domain specific language, Marlowe,\footnote{Marlowe is named after Christopher Marlowe, the Elizabethan poet, dramatist and spy, who was born and educated in Canterbury, which is also where our university is based.
-\url{https://en.wikipedia.org/wiki/Christopher_Marlowe}}\footnote{The code for Marlowe is available from \url{https://github.com/input-output-hk/scdsl}
-} targetted at the execution of financial contracts in the style of Peyton Jones, Eber and Seward~\cite{PeytonJones:2000} on blockchains.
-This work is a part of the Cardano project, and is supported by IOHK.\footnote{\url{https://iohk.io}}
-
-In doing this, we are required to refine the model of contracts in a number of ways in order to fit with a radically different context.
-Let us consider the following example of an ``escrow'' contract so that we can explain the motivation more concretely.
-The aim of this contract, written in functional pseudocode in the style of~\cite{PeytonJones:2000} involves three participants \haskellinline{1}, \haskellinline{2} and \haskellinline{3}.
-Participant \haskellinline{1} is to pay an amount of money to participant \haskellinline{2} on receipt of goods from her.
-Participant \haskellinline{1} pays the money into escrow controlled by participant \haskellinline{3}.
-There are two options for the money: if two participants agree to \haskellinline{pay} it to \haskellinline{2}, that goes ahead; if, on the other hand, two of the participants opt to \haskellinline{refund} the money to \haskellinline{1}, that is done instead.
-The \haskellinline{When} construct fires when its first argument -- a condition -- becomes true; the second argument, a \haskellinline{Contract} itself, is then performed, and this makes the payment according to the participants' \haskellinline{Choice}.
+This paper explores the design of a domain specific language, Marlowe,\footnote{Marlowe is named after Christopher 
+Marlowe, the Elizabethan poet, dramatist and spy, who was born and educated in Canterbury, which is also where our 
+university is based. \url{https://en.wikipedia.org/wiki/Christopher_Marlowe}}\footnote{The code for Marlowe is available 
+from \url{https://github.com/input-output-hk/scdsl}} targetted at the execution of financial contracts in the style of 
+Peyton Jones, Eber and Seward~\cite{PeytonJones:2000} on blockchains. This work is a part of the Cardano project, and is 
+supported by IOHK.\footnote{\url{https://iohk.io}}
+
+In doing this, we are required to refine the model of contracts in a number of ways in order to fit with a radically 
+different context. Let us consider the following example of an ``escrow'' contract so that we can explain the motivation 
+more concretely. The aim of this contract, written in functional pseudocode in the style of~\cite{PeytonJones:2000} 
+involves three participants \haskellinline{1}, \haskellinline{2} and \haskellinline{3}. Participant \haskellinline{1} is 
+to pay an amount of money to participant \haskellinline{2} on receipt of goods from her. Participant \haskellinline{1} 
+pays the money into escrow controlled by participant \haskellinline{3}. There are two options for the money: if two 
+participants agree to \haskellinline{pay} it to \haskellinline{2}, that goes ahead; if, on the other hand, two of the 
+participants opt to \haskellinline{refund} the money to \haskellinline{1}, that is done instead. The 
+\haskellinline{When} construct fires when its first argument -- a condition -- becomes true; the second argument, a 
+\haskellinline{Contract} itself, is then performed, and this makes the payment according to the participants' 
+\haskellinline{Choice}.
 
@@
 
-In the traditional model, enforcement of the contract is the responsibility of the legal system.
-If participant \haskellinline{1} does not pay the money into escrow, or participant \haskellinline{3} chooses to keep it for herself, then we can assume that they can be sued for the money and probably damages on top, thus providing both legal and financial incentives for compliance.
-On the other hand, in the decentralised blockchain model, where there is no central authority, the contract needs to be enforced by design.
-
-This means that we require participants to commit money to cover all possible expenditure \emph{in advance of the contract executing}.
-In order to make sure that participants continue to engage with a contract, we ensure urgency by imposing timeouts: money is committed for a finite period, for example, and when waiting for a participant to make a commitment we also impose a timeout to ensure that the contract does not become stuck even if one or more of the participants stop interacting with it.
-We revisit this example in Section~\ref{section:example-escrow} where we show how commitments, timeouts and other features of Marlowe are used to describe this as a self-enforcing contract ``on the blockchain''.
-
-Moreover, there is not a single blockchain model, and we give a brief overview of the different mechanisms for scripting on blockchain (although this is not the primary purpose of this article). 
-A particularly important technical distinction here is that between an account-style model as used by Ethereum~\cite{EthereumRationale}, and the unspent transaction output (UTxO) model of bitcoin~\cite{sok}.
-There is also the question of how precisely the execution of the contract interacts with the blockchain, and how information is to be recorded on the chain.
-Finally, should the execution ``push'' for changes  to happen, or do external actors have to ``pull'' to effect changes enabled by execution of the contract?
-
-The paper describes version 1.2 of Marlowe, and it should be seen as a work in progress. We have designed it to provide financial contracts in Cardano, but also in order to start to answer the twin
- questions of what smart contracts will look like in Cardano, and how those contracts will be implemented in the overall system.
-
-The paper is structured like this: we begin in Section \ref{section:design} with a discussion of the design rationale for Marlowe, including an overview of the various approaches to scripting in blockchain as well as the different designs of the blockchains themselves.
-We conclude Section \ref{section:design} with a set of decisions about the blockchain environment for Marlowe.
-The central part of the paper is Section \ref{section:model} where we introduce the Marlowe model, including the assumptions made in designing
-it, the types of the principal functions and a description of the DSL as an algebraic type, constructor by constructor. 
-Section \ref{section:example-saving} presents an example of a larger Marlowe contract, for incentivised saving.
-Section \ref{section:example-escrow} revisits the escrow example, and shows how it is described using a combination of Marlowe and Haskell constructs: that is, we use Marlowe as an embedded DSL.
-Section
-\ref{section:tool} introduces Meadow, our tool for visualising and interacting with Marlowe contracts.
-Section \ref{section:reflection} reflects on the system design, and observes some alternative designs and Section \ref{section:next-steps} enumerates the next steps for the project.
+In the traditional model, enforcement of the contract is the responsibility of the legal system. If participant 
+\haskellinline{1} does not pay the money into escrow, or participant \haskellinline{3} chooses to keep it for herself, 
+then we can assume that they can be sued for the money and probably damages on top, thus providing both legal and 
+financial incentives for compliance. On the other hand, in the decentralised blockchain model, where there is no central 
+authority, the contract needs to be enforced by design.
+
+This means that we require participants to commit money to cover all possible expenditure \emph{in advance of the 
+contract executing}. In order to make sure that participants continue to engage with a contract, we ensure urgency by 
+imposing timeouts: money is committed for a finite period, for example, and when waiting for a participant to make a 
+commitment we also impose a timeout to ensure that the contract does not become stuck even if one or more of the 
+participants stop interacting with it. We revisit this example in Section~\ref{section:example-escrow} where we show how 
+commitments, timeouts and other features of Marlowe are used to describe this as a self-enforcing contract ``on the 
+blockchain''.
+
+Moreover, there is not a single blockchain model, and we give a brief overview of the different mechanisms for 
+scripting on blockchain (although this is not the primary purpose of this article). 
+A particularly important technical distinction here is that between an account-style model as used by 
+Ethereum~\cite{EthereumRationale}, and the unspent transaction output (UTxO) model of bitcoin~\cite{sok}. 
+There is also the question of how precisely the execution of the contract interacts with the blockchain, and how 
+information is to be recorded on the chain. Finally, should the execution ``push'' for changes  to happen, or do 
+external actors have to ``pull'' to effect changes enabled by execution of the contract?
+
+The paper describes version 1.2 of Marlowe, and it should be seen as a work in progress. We have designed it to provide 
+financial contracts in Cardano, but also in order to start to answer the twin questions of what smart contracts will 
+look like in Cardano, and how those contracts will be implemented in the overall system.
+
+The paper is structured like this: we begin in Section \ref{section:design} with a discussion of the design rationale 
+for Marlowe, including an overview of the various approaches to scripting in blockchain as well as the different designs 
+of the blockchains themselves. We conclude Section \ref{section:design} with a set of decisions about the blockchain 
+environment for Marlowe. The central part of the paper is Section \ref{section:model} where we introduce the Marlowe 
+model, including the assumptions made in designing it, the types of the principal functions and a description of the DSL 
+as an algebraic type, constructor by constructor. Section \ref{section:example-saving} presents an example of a larger 
+Marlowe contract, for incentivised saving. Section \ref{section:example-escrow} revisits the escrow example, and shows 
+how it is described using a combination of Marlowe and Haskell constructs: that is, we use Marlowe as an embedded DSL. 
+Section \ref{section:tool} introduces Meadow, our tool for visualising and interacting with Marlowe contracts.
+Section \ref{section:reflection} reflects on the system design, and observes some alternative designs and Section 
+\ref{section:next-steps} enumerates the next steps for the project.
 
@@
 \noindent
-In this section we look at existing work on smart contracts and blockchains in general, before exploring the rationale for building the Marlowe language.
+In this section we look at existing work on smart contracts and blockchains in general, before exploring the rationale 
+for building the Marlowe language.
 
@@
 
-Blockchains are executed in a replicated form by parties who cannot be guaranteed not to be hostile, either by directly trying to change the contents of the chain, or through trying to affect other properties of the chain by indirect means (such as swamping honest parties with work).
-Programming on the blockchain therefore needs to be constrained in some ways, since it has to be amenable to replication or verification within a reasonable time if the security and integrity of the chain are to be preserved.
-We look at representatives of the main approaches now.
+Blockchains are executed in a replicated form by parties who cannot be guaranteed not to be hostile, either by directly 
+trying to change the contents of the chain, or through trying to affect other properties of the chain by indirect means 
+(such as swamping honest parties with work). Programming on the blockchain therefore needs to be constrained in some 
+ways, since it has to be amenable to replication or verification within a reasonable time if the security and integrity 
+of the chain are to be preserved. We look at representatives of the main approaches now.
 
@@
 
-A common feature of many blockchain platforms is that they provide multiple scripting languages.
-As we noted earlier, Ethereum can be programmed at the EVM level as well as by using the high-level Solidity language.
-Tezos~\cite{tezos-white-paper} supports the stack-based, strongly-typed, functional language Michelson, but also provides a high-level language, Liquidity, that compiles into Michelson.
-Liquidity is also functional and strongly typed, but provides more constructs in a more familiar syntax, namely that of a subset of OCaml.
-
-The \ae{}ternity system~\cite{aeternity} provides multiple languages and VMs:\footnote{As presented by Erik Stenman at CODE BEAM SF, March 2018, \url{https://www.youtube.com/watch?v=VXsqvfPIdWg}} the functional language  Sophia (akin to Reason) and the Functional Typed Warded Virtual Machine (FTWVM) for safe ``system level programming'', the language Varna and the HLM for simple contracts, and a (port of) Solidity and the EVM for compatibility with Ethereum.
-In a similar way, Cardano provides support for IELE~\cite{IELE}, a rational reconstruction of the EVM, and thus for Solidity contracts too.
+A common feature of many blockchain platforms is that they provide multiple scripting languages. As we noted earlier, 
+Ethereum can be programmed at the EVM level as well as by using the high-level Solidity language. 
+Tezos~\cite{tezos-white-paper} supports the stack-based, strongly-typed, functional language Michelson, but also 
+provides a high-level language, Liquidity, that compiles into Michelson. Liquidity is also functional and strongly 
+typed, but provides more constructs in a more familiar syntax, namely that of a subset of OCaml.
+
+The \ae{}ternity system~\cite{aeternity} provides multiple languages and VMs:\footnote{As presented by Erik Stenman at 
+CODE BEAM SF, March 2018, \url{https://www.youtube.com/watch?v=VXsqvfPIdWg}} the functional language  Sophia (akin to 
+Reason) and the Functional Typed Warded Virtual Machine (FTWVM) for safe ``system level programming'', the language 
+Varna and the HLM for simple contracts, and a (port of) Solidity and the EVM for compatibility with Ethereum. In a 
+similar way, Cardano provides support for IELE~\cite{IELE}, a rational reconstruction of the EVM, and thus for Solidity 
+contracts too.
 
@@
 
-Stand-alone DSLs have the advantage of providing appropriate, domain-level error messages when things go wrong, but suffer the disadvantage of having to be implemented from scratch.
-An alternative to this are \emph{embedded} DSLs (EDSLs), which provide a ``little language'' for the particular domain embedded within a general-purpose host language.
-This means that parts of the host -- such as arithmetical expressions, or list idioms --  can be used to extend the expressibility of the DSL.
-A notable example of this is the financial contracts language described by Peyton Jones and his collaborators~\cite{PeytonJones:2000}, which is embedded in Haskell.
+Stand-alone DSLs have the advantage of providing appropriate, domain-level error messages when things go wrong, but 
+suffer the disadvantage of having to be implemented from scratch. An alternative to this are \emph{embedded} DSLs 
+(EDSLs), which provide a ``little language'' for the particular domain embedded within a general-purpose host language. 
+This means that parts of the host -- such as arithmetical expressions, or list idioms --  can be used to extend the 
+expressibility of the DSL. A notable example of this is the financial contracts language described by Peyton Jones and 
+his collaborators~\cite{PeytonJones:2000}, which is embedded in Haskell.
 
@@
 
-The Findel project~\cite{findel} examines financial contracts on the Ethereum platform, based on the seminal~\cite{PeytonJones:2000}, and the authors note that payments need to be bounded; this is made concrete in our account by our notion of commitments.
-They take no account of commitments or timeouts as we do below, but it should be noted that the Ethereum platform is more powerful than the one that we target in this paper.
+The Findel project~\cite{findel} examines financial contracts on the Ethereum platform, based on the 
+seminal~\cite{PeytonJones:2000}, and the authors note that payments need to be bounded; this is made concrete in our 
+account by our notion of commitments. They take no account of commitments or timeouts as we do below, but it should be 
+noted that the Ethereum platform is more powerful than the one that we target in this paper.
 
@@
 
-There is not a single model of how a blockchain operates, and in this section we look at some of the tradeoffs possible in this space.
-This allows us to give a context to some of the design decisions taken in designing Marlowe.
+There is not a single model of how a blockchain operates, and in this section we look at some of the tradeoffs possible 
+in this space. This allows us to give a context to some of the design decisions taken in designing Marlowe.
 
@@
 
-Some chains have a layered design, in particular Cardano~\cite{cardano}.
-The Cardano Settlement Layer (SL) is to support settlement of transactions but nothing more complicated, whereas the Computation Layer (or CL) supports more powerful general computations over the chain.
-Moreover, the Cardano roadmap~\cite{cardano-rationale} envisages the possibility of computation taking place, in part at least, on a sidechain rather than on the main chain itself, which may support only the SL.
-Other off-chain approaches include the Lightning network, overlaid on blockchains (usually Bitcoin) and the channels of \ae{}ternity, introduced below.
+Some chains have a layered design, in particular Cardano~\cite{cardano}. The Cardano Settlement Layer (SL) is to 
+support settlement of transactions but nothing more complicated, whereas the Computation Layer (or CL) supports more 
+powerful general computations over the chain. Moreover, the Cardano roadmap~\cite{cardano-rationale} envisages the 
+possibility of computation taking place, in part at least, on a sidechain rather than on the main chain itself, which 
+may support only the SL. Other off-chain approaches include the Lightning network, overlaid on blockchains (usually 
+Bitcoin) and the channels of \ae{}ternity, introduced below.
 
@@
 Marlowe is described as an algebraic \haskellinline{data} type in Haskell, and as such can be seen as a simple EDSL. 
-Note, however, that the implementation presented here is for the purposes of presenting the semantics and behaviour of the language in the abstract.
-The implementation on the Cardano platform could be stand-alone, but more likely it will be a DSL embedded in Plutus or Plutus Core.
-The details of this remain to be decided, as noted in the introduction.
+Note, however, that the implementation presented here is for the purposes of presenting the semantics and behaviour of 
+the language in the abstract. The implementation on the Cardano platform could be stand-alone, but more likely it will 
+be a DSL embedded in Plutus or Plutus Core. The details of this remain to be decided, as noted in the introduction.
 
@@
 The DSL is modelled as an algebraic type in Haskell, together with an executable small-step semantics. 
-We start by looking at the different types used by the model, then we look at the \haskellinline{Contract} DSL itself, and finally we give its semantics in Haskell.
-Section \ref{section:example-saving} gives an example of Marlowe in action, and Section
+We start by looking at the different types used by the model, then we look at the \haskellinline{Contract} DSL itself, 
+and finally we give its semantics in Haskell. Section \ref{section:example-saving} gives an example of Marlowe in 
+action, and Section
 \ref{section:example-escrow} shows the advantage of seeing Marlowe as a DSL embedded in Haskell.
@@
 
-First, it will need to observe different kinds of varying quantities including, for example, the current time, the current block number and, random numbers, as well as ``real world'' quantities like ``the price of oil'' or ``the exchange rate between currencies A and B''.
-Each instance of such an observable will be recorded on the blockchain to allow the computation to be repeated for verification purposes.
-It is assumed that at each step of the execution of the contract, the values of some observables will be needed, and these values are (together) given by a value of type \haskellinline{OS}.
-Note that these values are not determined by the participants in the contract, but rather by the external environment.
-
-As the examples illustrate, observables come both from aspects of the blockchain (e.g.\ current block number) and externally.
-In the latter case it will be necessary to agree a trusted oracle or beacon giving the value, which can then be stored (or a hash stored) on chain.
+First, it will need to observe different kinds of varying quantities including, for example, the current time, the 
+current block number and, random numbers, as well as ``real world'' quantities like ``the price of oil'' or ``the 
+exchange rate between currencies A and B''. Each instance of such an observable will be recorded on the blockchain to 
+allow the computation to be repeated for verification purposes. It is assumed that at each step of the execution of the 
+contract, the values of some observables will be needed, and these values are (together) given by a value of type 
+\haskellinline{OS}. Note that these values are not determined by the participants in the contract, but rather by the 
+external environment.
+
+As the examples illustrate, observables come both from aspects of the blockchain (e.g.\ current block number) and 
+externally. In the latter case it will be necessary to agree a trusted oracle or beacon giving the value, which can then 
+be stored (or a hash stored) on chain.
 
@@
 
-On the other hand, at each step there are -- potentially, at least -- a variety of inputs available from the participants.
-These include commitments of currency (or ``cash''), redemption of commitments, and claims of payments by a participant; moreover it is also possible for a participant to input an arbitrary value (which we term a ``choice'').
+On the other hand, at each step there are -- potentially, at least -- a variety of inputs available from the 
+participants. These include commitments of currency (or ``cash''), redemption of commitments, and claims of payments by 
+a participant; moreover it is also possible for a participant to input an arbitrary value (which we term a ``choice''). 
 The particular inputs at a given step are described by a value of type \haskellinline{Input}.
@@
 
-While informally we might see a commitment to something as being indefinite, it is important to realise that on blockchain a commitment needs to have a timeout so that progress can be forced in a game.
-After the timeout period the cash can be refunded through the user creating a transaction to reclaim the cash.
-Information about the commitments currently in force forms the \haskellinline{State}, which is (potentially) modified at each execution step.
+While informally we might see a commitment to something as being indefinite, it is important to realise that on 
+blockchain a commitment needs to have a timeout so that progress can be forced in a game. After the timeout period the 
+cash can be refunded through the user creating a transaction to reclaim the cash.
+Information about the commitments currently in force forms the \haskellinline{State}, which is (potentially) modified 
+at each execution step.
 
@@
 same way that cash commitments are redeemed when they expire. The effects of the contract in the blockchain are 
-represented by a list \haskellinline{AS} of \haskellinline{Actions} that is derived from the execution of each step of the semantics.
+represented by a list \haskellinline{AS} of \haskellinline{Actions} that is derived from the execution of each step of 
+the semantics.
 
@@
 \begin{itemize}
-\item It is assumed that cryptographic functions and operations are provided by a layer external to the system, and so they need not be modelled explicitly here.
-\item We assume that time is ``coarse grained'' and measured simply by block number, so that timeouts etc.\ are delimited using nothing finer than block number. 
-\item Because we assume a ``pull'' model, making a commitment is not something that a contract can perform; rather, it can enable that a commitment is made, but that then has to be established externally: hence the input of (a set of) commitments at each step.
-The same observation also applies to redeeming a commitment of cash.
-Commitments that have been made or redeemed need to be recorded somehow.
+\item It is assumed that cryptographic functions and operations are provided by a layer external to the system, and so 
+they need not be modelled explicitly here.
+\item We assume that time is ``coarse grained'' and measured simply by block number, so that timeouts etc.\ are 
+delimited using nothing finer than block number. 
+\item Because we assume a ``pull'' model, making a commitment is not something that a contract can perform; rather, it 
+can enable that a commitment is made, but that then has to be established externally: hence the input of (a set of) 
+commitments at each step. The same observation also applies to redeeming a commitment of cash. Commitments that have 
+been made or redeemed need to be recorded somehow.
 
@@
 ``expression language''.
-\item It is important to be aware that there may be different instances of contracts executed, and for each instance there will typically be different instances of observables (and commitments) observed and recorded.
-\item The model manages the release of funds back to the committer when a cash commitment expires (see discussion of the \haskellinline{stepBlock} function below).
+\item It is important to be aware that there may be different instances of contracts executed, and for each instance 
+there will typically be different instances of observables (and commitments) observed and recorded.
+\item The model manages the release of funds back to the committer when a cash commitment expires (see discussion of 
+the \haskellinline{stepBlock} function below).
 \end{itemize}
@@
    \end{haskellcode}
-%Originally on "Choice": \footnote{Doesn't the step function need to record what observations are made, so they can be archived in the blockchain?
-I don't see how your design will support doing so.}\footnote{Yes, we assume that it is part of the infrastructure of this language to record the values of *observables* that are used.
-However, we have chosen no to make it part of the language itself.}
+%Originally on "Choice": \footnote{Doesn't the step function need to record what observations are made, so they can be 
+%archived in the blockchain? I don't see how your design will support doing so.}\footnote{Yes, we assume that it is 
+%part of the infrastructure of this language to record the values of *observables* that are used. However, we have 
+%chosen no to make it part of the language itself.}
 
@@
 
-The \haskellinline{step} function is total, so that for every contract a result of stepping is defined.
-However, for some kinds of contracts -- commits, redeems or time-shifted contracts -- it is possible that the result of a step is to produce the same contract as the result; such a step is called \emph{quiescent} whereas all other steps \emph{make progress}.
-We use this distinction in the explanation that follows.
-
-The function \haskellinline{step} is wrapped by the \haskellinline{stepAll} function, which calls \haskellinline{step} until it stops having any effects.
-For every block, the \haskellinline{stepBlock} function is called.
-In addition to calling \haskellinline{stepAll}, \haskellinline{stepBlock} will first enable expired cash commitments to be refunded and record any choices made at that step.
+The \haskellinline{step} function is total, so that for every contract a result of stepping is defined. However, for 
+some kinds of contracts -- commits, redeems or time-shifted contracts -- it is possible that the result of a step is to 
+produce the same contract as the result; such a step is called \emph{quiescent} whereas all other steps \emph{make 
+progress}. We use this distinction in the explanation that follows.
+
+The function \haskellinline{step} is wrapped by the \haskellinline{stepAll} function, which calls \haskellinline{step} 
+until it stops having any effects. For every block, the \haskellinline{stepBlock} function is called. In addition to 
+calling \haskellinline{stepAll}, \haskellinline{stepBlock} will first enable expired cash commitments to be refunded and 
+record any choices made at that step.
 
@@
 
-If the cash is committed successfully and the timeout \haskellinline{t2} is reached, then it is impossible to further spend the committed cash, and any unspent funds can be reclaimed by participant \haskellinline{p}.
+If the cash is committed successfully and the timeout \haskellinline{t2} is reached, then it is impossible to further 
+spend the committed cash, and any unspent funds can be reclaimed by participant \haskellinline{p}.
  This is performed by the \haskellinline{stepBlock} function, as noted above.
 
-\item \haskellinline{RedeemCC idCC k} (\haskellinline{CC} stands for cash commitment.) For this contract to make progress, the creator of the cash commitment with identifier \haskellinline{idCC} is required to redeem the unspent funds in that commitment; the contract then continues as \haskellinline{k}, and the action \haskellinline{CommitRedeemed} is produced.
+\item \haskellinline{RedeemCC idCC k} (\haskellinline{CC} stands for cash commitment.) For this contract to make 
+progress, the creator of the cash commitment with identifier \haskellinline{idCC} is required to redeem the unspent 
+funds in that commitment; the contract then continues as \haskellinline{k}, and the action 
+\haskellinline{CommitRedeemed} is produced.
 
@@
 
-Committed cash can only be redeemed once, and an attempt to redeem it a second time will produce a  \haskellinline{DuplicateRedeem} action, continuing as \haskellinline{k}.
-
-If the cash commitment with identifier \haskellinline{idCC} has expired, it becomes possible for the remaining funds to be redeemed by the committer; this can be done by the \haskellinline{stepBlock} function processing the appropriate \haskellinline{Input}, and an \haskellinline{ExpiredCommitRedeemed} action will be produced.
-
-Once the commitment \haskellinline{idCC} has expired and is redeemed, a \haskellinline{RedeemCC idCC k} contract will immediately evolve to \haskellinline{k}.
+Committed cash can only be redeemed once, and an attempt to redeem it a second time will produce a  
+\haskellinline{DuplicateRedeem} action, continuing as \haskellinline{k}.
+
+If the cash commitment with identifier \haskellinline{idCC} has expired, it becomes possible for the remaining funds to 
+be redeemed by the committer; this can be done by the \haskellinline{stepBlock} function processing the appropriate 
+\haskellinline{Input}, and an \haskellinline{ExpiredCommitRedeemed} action will be produced.
+
+Once the commitment \haskellinline{idCC} has expired and is redeemed, a \haskellinline{RedeemCC idCC k} contract will 
+immediately evolve to \haskellinline{k}.
 %\footnote{How does the contract determine what actions resulted from a given step? Or is this not required?}
-%\footnote{If it is required, restructuring the DSL so that the result action appears as the 'value' of executing a step (using a monadic interface as described for id generation below) would be a good idea.}
-%\footnote{We specify the actions generated by each kind of constructor; at each step the list of such actions, of type \haskellinline{AS}, is a component of the result.}.
+%\footnote{If it is required, restructuring the DSL so that the result action appears as the 'value' of executing a 
+%          step (using a monadic interface as described for id generation below) would be a good idea.}
+%\footnote{We specify the actions generated by each kind of constructor; at each step the list of such actions, of type 
+%          \haskellinline{AS}, is a component of the result.}.
 
@@
 
- This contract will result in a \haskellinline{FailedPay} action if the funds are not available; otherwise a \haskellinline{SuccessfulPay} action is generated.
+This contract will result in a \haskellinline{FailedPay} action if the funds are not available; otherwise a 
+\haskellinline{SuccessfulPay} action is generated.
 
@@
 
-Because the model is stateful and produces output actions, to make a step it is necessary to execute a single step of each of  the contracts \haskellinline{k1} and \haskellinline{k2} in sequence: first \haskellinline{k1} then \haskellinline{k2}.
-
-\item \haskellinline{Choice obs k1 k2} behaves as either \haskellinline{k1} or \haskellinline{k2} depending on the (Boolean) result of \haskellinline{obs} at the time that the observation is made, \haskellinline{k1} if it is \haskellinline{True} and \haskellinline{k2} if \haskellinline{False}.
+Because the model is stateful and produces output actions, to make a step it is necessary to execute a single step of 
+each of  the contracts \haskellinline{k1} and \haskellinline{k2} in sequence: first \haskellinline{k1} then 
+\haskellinline{k2}.
+
+\item \haskellinline{Choice obs k1 k2} behaves as either \haskellinline{k1} or \haskellinline{k2} depending on the 
+(Boolean) result of \haskellinline{obs} at the time that the observation is made, \haskellinline{k1} if it is 
+\haskellinline{True} and \haskellinline{k2} if \haskellinline{False}.
 
@@
 
-\item \haskellinline{When obs timeout k def_k} This contract will not progress until \haskellinline{obs} is \haskellinline{True} or until the current block number is greater than or equal to the one specified by timeout.
-In case the timeout applies, the contract will continue as \haskellinline{def_k}, if the timeout does not apply and \haskellinline{obs} is \haskellinline{True}, then the contract continues as \haskellinline{k}.
-Otherwise the contract is quiescent.
+\item \haskellinline{When obs timeout k def_k} This contract will not progress until \haskellinline{obs} is 
+\haskellinline{True} or until the current block number is greater than or equal to the one specified by timeout. In case 
+the timeout applies, the contract will continue as \haskellinline{def_k}, if the timeout does not apply and 
+\haskellinline{obs} is \haskellinline{True}, then the contract continues as \haskellinline{k}. Otherwise the contract is 
+quiescent.
 
@@
 In this section, we show how Marlowe can be used for designing a contract that incentivises saving. We use the 
-following example: participant \haskellinline{2} offers 20 ADA to participant \haskellinline{1} in exchange for saving 100 ADA for 100 blocks. 
-participant \haskellinline{1} will be able to withdraw the savings after block 20, but if that option is chosen, then participant 
- \haskellinline{2} will not pay the extra 20 ADA to participant \haskellinline{1}.
-
-We will start from outer to inner primitives. The first thing is to allow participant \haskellinline{1} to commit the 100 ADA. The 
-primitive says that with identity \haskellinline{IdentCC 1} (this is the identity of the commitment), participant 
-\haskellinline{1} can commit \haskellinline{100} ADA before block \haskellinline{10}, and the commitment must 
-expire on block \haskellinline{200}. But if the money is not committed before block \haskellinline{10}, then the 
+following example: participant \haskellinline{2} offers 20 ADA to participant \haskellinline{1} in exchange for saving 
+100 ADA for 100 blocks. participant \haskellinline{1} will be able to withdraw the savings after block 20, but if that 
+option is chosen, then participant \haskellinline{2} will not pay the extra 20 ADA to participant \haskellinline{1}.
+
+We will start from outer to inner primitives. The first thing is to allow participant \haskellinline{1} to commit the 
+100 ADA. The primitive says that with identity \haskellinline{IdentCC 1} (this is the identity of the commitment), 
+participant \haskellinline{1} can commit \haskellinline{100} ADA before block \haskellinline{10}, and the commitment 
+must expire on block \haskellinline{200}. But if the money is not committed before block \haskellinline{10}, then the 
 contract will continue as \haskellinline{Null}.
@@
 If the commitment is done correctly and on time, we continue as in \haskellinline{step2}. \haskellinline{step2} is 
-another commitment, this time by participant \haskellinline{2}. The primitive says that with identity \haskellinline{IdentCC 2},
-participant \haskellinline{2} can commit \haskellinline{20} ADA before block \haskellinline{20}, and the commitment
-must expire on block \haskellinline{200}. But if the money is not committed before block \haskellinline{20}, then the
-contract will continue as \haskellinline{RedeemCC (IdentCC 1) Null}, which in turn gives participant \haskellinline{1} the 
-opportunity to withdraw all the money committed immediately.
+another commitment, this time by participant \haskellinline{2}. The primitive says that with identity 
+\haskellinline{IdentCC 2}, participant \haskellinline{2} can commit \haskellinline{20} ADA before block 
+\haskellinline{20}, and the commitment must expire on block \haskellinline{200}. But if the money is not committed 
+before block \haskellinline{20}, then the contract will continue as \haskellinline{RedeemCC (IdentCC 1) Null}, which in 
+turn gives participant \haskellinline{1} the opportunity to withdraw all the money committed immediately.
 
@@
 \begin{centering}
-\includegraphics[width=1\textwidth]{/home/palas/Escritorio/scdsl/tex/pix/embedded.png}
+\includegraphics[width=1\textwidth]{pix/embedded.png}
 \par\end{centering}
@@
 
-The model that we have discussed so far is deliberately kept as compact as possible. However there are a number of potential extensions to the model which we introduce here.
+The model that we have discussed so far is deliberately kept as compact as possible. However there are a number of 
+potential extensions to the model which we introduce here.
 
@@
 We can make the \haskellinline{Contract} type monadic, so that new, unique identifiers are automatically generated. 
-This gives two nice properties.
-First, they are guaranteed to be unique; secondly they can be generated only in cases where a commit has been made successfully (within the required time period).
-The
-\haskellinline{do} notation supported by monads allows the generated names to be captured by the \haskellinline{bind} operation and referred to later in the computation.
+This gives two nice properties. First, they are guaranteed to be unique; secondly they can be generated only in cases 
+where a commit has been made successfully (within the required time period). The 
+\haskellinline{do} notation supported by monads allows the generated names to be captured by the \haskellinline{bind} 
+operation and referred to later in the computation.
 
@@
 One possible definition is ``those contracts that do not lead to a \haskellinline{FailedPay} action''.
-Given the mechanisms for making and using commitments are limited by timeouts, it should be possible to write contracts that have this property, perhaps with some assumptions about the value of observables.
+Given the mechanisms for making and using commitments are limited by timeouts, it should be possible to write contracts 
+that have this property, perhaps with some assumptions about the value of observables.
 Looking at this in a little more detail, we would expect the following points to hold.
@@
 \item For a \haskellinline{Pay} contract a sufficient balance not to produce a \haskellinline{FailedPay}.
-\item For \haskellinline{Both} we require a sufficient balance for both contracts combined.
-Need to think here whether the two conjuncts should be independent, or whether a commitment in one half can be used in the other.
-\item For \haskellinline{Choice} we require a sufficient balance for the maximum of the requirements of the individual contracts. 
+\item For \haskellinline{Both} we require a sufficient balance for both contracts combined. Need to think here whether 
+the two conjuncts should be independent, or whether a commitment in one half can be used in the other.
+\item For \haskellinline{Choice} we require a sufficient balance for the maximum of the requirements of the individual 
+contracts. 
 \item For \haskellinline{When}, need to ``age'' the existing commitments relative to the new starting point of the 
@@
 The model does not take time as an observable, but rather it is defined in terms of \haskellinline{BlockNumber} only. 
-We could  use ``slot time'' instead: \emph{``We consider a setting where time is divided into discrete units called slots.
-A ledger, described in more detail below, associates with each time slot (at most) one ledger block.''}\footnote{\url{https://eprint.iacr.org/2016/889.pdf}}
+We could  use ``slot time'' instead: \emph{``We consider a setting where time is divided into discrete units called 
+slots. A ledger, described in more detail below, associates with each time slot (at most) one ledger 
+block.''}\footnote{\url{https://eprint.iacr.org/2016/889.pdf}}
 
@@
 
-In systems where time is significant, it is possible to build ``timelocks'' under which 
-it becomes impossible for time to progress.
-Is it possible for block number to depend on progress of computation?
-We conjecture not, because otherwise it would be possible to produce timelocks, under which it would become impossible for time to progress.
+In systems where time is significant, it is possible to build ``timelocks'' under which it becomes impossible for time 
+to progress. Is it possible for block number to depend on progress of computation? We conjecture not, because otherwise 
+it would be possible to produce timelocks, under which it would become impossible for time to progress.
 
@@
 
-Timeouts are expressed in absolute time: presumably they could be relative (with recording on the blockchain of the times at which they are triggered).
+Timeouts are expressed in absolute time: presumably they could be relative (with recording on the blockchain of the 
+times at which they are triggered).
 
@@
 
-The model is deterministic, with all non-determinism being external, and confined to observables.
-This is surely necessary in order to support computation replay, but does it restrict the expressibility of the model?
+The model is deterministic, with all non-determinism being external, and confined to observables. This is surely 
+necessary in order to support computation replay, but does it restrict the expressibility of the model?
 
@@
 
-The \haskellinline{RedeemCC} construct is intended for premature repayment of a cash commitment, i.e.
-before the commitment has timed out.
-Alternatively, it would be possible for the underlying model to repay all committed but unspent cash automatically at the end of a contract (but before all commitments have expired), but we might want to do the same thing at the end of each ``round'' of  a game, for instance.
+The \haskellinline{RedeemCC} construct is intended for premature repayment of a cash commitment, i.e. before the 
+commitment has timed out. Alternatively, it would be possible for the underlying model to repay all committed but 
+unspent cash automatically at the end of a contract (but before all commitments have expired), but we might want to do 
+the same thing at the end of each ``round'' of  a game, for instance.
 
@@
 
-We have modelled cash as fungible: we make payments using the cash commitments in order of expiry time, with those expiring earliest being used first.
-Building a model on UTxO might be a reason for using a different approach.
+We have modelled cash as fungible: we make payments using the cash commitments in order of expiry time, with those 
+expiring earliest being used first. Building a model on UTxO might be a reason for using a different approach.
 
@@
 
-Some observables have a ``monotonicity'' property: once a choice is made, it is fixed forever; on the other hand, something not yet chosen is only so thus far; do we need to reflect this in the observations that we allow?
+Some observables have a ``monotonicity'' property: once a choice is made, it is fixed forever; on the other hand, 
+something not yet chosen is only so thus far; do we need to reflect this in the observations that we allow?
 
@@
 \end{itemize} 
-Even in Ethereum this distinction is important because single contract execution is limited by gas (and is thus finite), whereas the life of the whole contract may be unbounded and consist of an indefinite number of transactions.
-So far we are assuming contracts defined in Marlowe to be finite in both aspects.
-They could be infinite if we were to define them co-inductively as recursive members of the DSL type.
+Even in Ethereum this distinction is important because single contract execution is limited by gas (and is thus 
+finite), whereas the life of the whole contract may be unbounded and consist of an indefinite number of transactions.  
+So far we are assuming contracts defined in Marlowe to be finite in both aspects. They could be infinite if we were to 
+define them co-inductively as recursive members of the DSL type.
 
@@
 
-If we have contracts that involve, for example, the oil price then we may need to convert a published price in USD into ADA, using a value for the prevailing USD/ADA exchange rate.
-This will require a computation: in this case, a multiplication.
-If Marlowe is to be stand-alone then we need to be extend it with arithmetic and other operations, but we expect Marlowe to be embedded in Plutus (Core) and so to be able to use the facilities of the host language here.
+If we have contracts that involve, for example, the oil price then we may need to convert a published price in USD into 
+ADA, using a value for the prevailing USD/ADA exchange rate. This will require a computation: in this case, a 
+multiplication. If Marlowe is to be stand-alone then we need to be extend it with arithmetic and other operations, but 
+we expect Marlowe to be embedded in Plutus (Core) and so to be able to use the facilities of the host language here.
 
@@
 
-The implementation may split contract execution between the main chain and a sidechain (or chains). This will allow links from Marlowe to work on securing multiparty sidechain games/contracts (MPCs).
+The implementation may split contract execution between the main chain and a sidechain (or chains). This will allow 
+links from Marlowe to work on securing multiparty sidechain games/contracts (MPCs).
 
@@
 
-We plan to continue the work with Marlowe in a number of directions. First, we need to sign off a definition of the language, as the current status of Marlowe 1.2 is that of a proposal.
+We plan to continue the work with Marlowe in a number of directions. First, we need to sign off a definition of the 
+language, as the current status of Marlowe 1.2 is that of a proposal.
 
@@
 
-We should decide on the final form of the language -- for example, should it be monadic or not -- and the relation of Marlowe to Plutus (Core).
-In particular, answering the latter question will require work on the current redeemer/validator model of transactions in Plutus Core.
+We should decide on the final form of the language -- for example, should it be monadic or not -- and the relation of 
+Marlowe to Plutus (Core). In particular, answering the latter question will require work on the current 
+redeemer/validator model of transactions in Plutus Core.
 
@@
 
-An immediate goal is to implement Marlowe in Cardano, noting the observations of Section \ref{section:towards-implementation} and to deploy it in a test network to observe and measure its behaviour in practice.
+An immediate goal is to implement Marlowe in Cardano, noting the observations of Section 
+\ref{section:towards-implementation} and to deploy it in a test network to observe and measure its behaviour in 
+practice.
 
@@
 
-Marlowe contracts could be generated by a compilation process from `vanilla' financial contracts.
-This also suggests that models with off-chain enforcement could be investigated too, with an MPC foundation.
+Marlowe contracts could be generated by a compilation process from `vanilla' financial contracts. This also suggests 
+that models with off-chain enforcement could be investigated too, with an MPC foundation.
 
@@
 
-As we noted in Section \ref{section:analysis} we can develop analyses of various aspects of Marlowe contracts.
-While these might initially be ``pencil and paper'' analyses, we could expect these to be based on a formalisation of the language and its semantics.
+As we noted in Section \ref{section:analysis} we can develop analyses of various aspects of Marlowe contracts. While 
+these might initially be ``pencil and paper'' analyses, we could expect these to be based on a formalisation of the 
+language and its semantics.
 
@@
 
-Marlowe and its semantics could be formalised in a theorem prover such as Isabelle~\cite{IsabelleHOL}, which has been used in the formalisation of language semantics.
-A formalisation could provide a sanity check on the language definition, as well as formalise proofs of important properties, based on analysis of the contract language.
+Marlowe and its semantics could be formalised in a theorem prover such as Isabelle~\cite{IsabelleHOL}, which has been 
+used in the formalisation of language semantics. A formalisation could provide a sanity check on the language 
+definition, as well as formalise proofs of important properties, based on analysis of the contract language.
 
@@
 
-\bibliographystyle{ACM-Reference-Format}
+\bibliographystyle{splncs04}
 \bibliography{paper}

### 2018-05-16 tex/paper.tex
Change tag version and fix two typos

@@
 % Title portion. Note the short title for running heads
-\title {Marlowe: financial contracts on blockchain\thanks{IOHK}}
+\title {Marlowe: financial contracts on blockchain\thanks{This work is a part of the Cardano project, and is
+supported by IOHK}}
 %
@@
 
-The paper describes version 1.2 of Marlowe, and it should be seen as a work in progress. We have designed it to provide 
+The paper describes version 1.3 of Marlowe, and it should be seen as a work in progress. We have designed it to provide 
 financial contracts in Cardano, but also in order to start to answer the twin questions of what smart contracts will 
@@
 \noindent
-In this section we look at existing work on smart contracts and blockchains in general, before exploring the rationale 
+In this section, we look at existing work on smart contracts and blockchains in general, before exploring the rationale 
 for building the Marlowe language.
@@
 
-In designing, Marlowe we have chosen a number of options; this description applies to Marlowe version 1.2
+In designing Marlowe, we have chosen a number of options; this description applies to Marlowe version 1.3.
 
@@
 We have given a concrete model and DSL, namely Marlowe version 
-1.2\footnote{\url{https://github.com/input-output-hk/scdsl/releases/tag/v1.2}}.
+1.3\footnote{\url{https://github.com/input-output-hk/scdsl/releases/tag/v1.3}}.
 In this section we examine how the model can be extended, how contacts can be analysed, reflect on different design 
@@
 We plan to continue the work with Marlowe in a number of directions. First, we need to sign off a definition of the 
-language, as the current status of Marlowe 1.2 is that of a proposal.
+language, as the current status of Marlowe 1.3 is that of a proposal.
 

### 2018-05-16 tex/paper.tex
Change version number and fix some small typos

@@
 \haskellinline{k1}; in the second case no action is generated and the contract continues as 
-\haskellinline{k2}. While neither case holds the contract is quiescent, waiting for the cash to be committed.
+\haskellinline{k2}. While neither case holds, the contract is quiescent, waiting for the cash to be committed.
 
@@
 
-Because the model is stateful and produces output actions, to make a step it is necessary to execute a single step of 
+Because the model is stateful and produces output actions, to make a step, it is necessary to execute a single step of 
 each of  the contracts \haskellinline{k1} and \haskellinline{k2} in sequence: first \haskellinline{k1} then 

### 2018-05-17 tex/paper.tex
moved references in footnotes into bib

@@
 % Title portion. Note the short title for running heads
-\title {Marlowe: financial contracts on blockchain\thanks{This work is a part of the Cardano project, and is
-supported by IOHK}}
+\title {Marlowe: financial contracts on blockchain}
 %
@@
 %
-\institute{University of Kent, Canterbury, CT2 7NF, Kent (UK)}
+\institute{School of Computing, University of Kent, Canterbury, UK}
 \email{\{pl292,s.j.thompson\}@kent.ac.uk}
@@
 
-This paper explores the design of a domain specific language, Marlowe,\footnote{Marlowe is named after Christopher 
-Marlowe, the Elizabethan poet, dramatist and spy, who was born and educated in Canterbury, which is also where our 
-university is based. \url{https://en.wikipedia.org/wiki/Christopher_Marlowe}}\footnote{The code for Marlowe is available 
+This paper explores the design of a domain specific language, Marlowe,\footnote{Named after Christopher 
+Marlowe, the Elizabethan poet, dramatist and spy, who was born and educated in Canterbury,  \url{en.wikipedia.org/wiki/Christopher_Marlowe}}\footnote{Marlowe is available 
 from \url{https://github.com/input-output-hk/scdsl}} targetted at the execution of financial contracts in the style of 
 Peyton Jones, Eber and Seward~\cite{PeytonJones:2000} on blockchains. This work is a part of the Cardano project, and is 
-supported by IOHK.\footnote{\url{https://iohk.io}}
+supported by IOHK,~\url{https://iohk.io}.
 
@@
 
-The \ae{}ternity system~\cite{aeternity} provides multiple languages and VMs:\footnote{As presented by Erik Stenman at 
-CODE BEAM SF, March 2018, \url{https://www.youtube.com/watch?v=VXsqvfPIdWg}} the functional language  Sophia (akin to 
+The \ae{}ternity system~\cite{aeternity} provides multiple languages and VMs~\cite{Stenman-CODE-BEAM}: the functional language  Sophia (akin to 
 Reason) and the Functional Typed Warded Virtual Machine (FTWVM) for safe ``system level programming'', the language 
@@
 slots. A ledger, described in more detail below, associates with each time slot (at most) one ledger 
-block.''}\footnote{\url{https://eprint.iacr.org/2016/889.pdf}}
+block.''}~\cite{Ouroboros}.
 

### 2018-05-17 tex/paper.tex
some removal/editing of figures for Meadow

@@
 \newenvironment{allbluett}{\begin{alltt}\color{myblue}}{\end{alltt}}
+\newcommand{\ignore}[1]{} % for commenting out large portions, surround by \ignore{ and }
 
@@
 model, including the assumptions made in designing it, the types of the principal functions and a description of the DSL 
-as an algebraic type, constructor by constructor. Section \ref{section:example-saving} presents an example of a larger 
-Marlowe contract, for incentivised saving. Section \ref{section:example-escrow} revisits the escrow example, and shows 
+as an algebraic type, constructor by constructor. 
+%Section \ref{section:example-saving} presents an example of a larger Marlowe contract, for incentivised saving. 
+Section \ref{section:example-escrow} revisits the escrow example, and shows 
 how it is described using a combination of Marlowe and Haskell constructs: that is, we use Marlowe as an embedded DSL. 
@@
 \includegraphics[width=0.6\textwidth]{pix/context.png}
-\caption{The context for the contract}
+\caption{The context for a contract}
 \label{fig:context}
@@
 
+\ignore{ % begin ignore
+
 \section{Incentivised saving}
@@
 
-\section{Marlowe as a DSL embedded in Haskell}
+} % end ignore
+
+\section{Marlowe as an embedded DSL in Haskell}
 \label{section:example-escrow}
@@
 smart-contracts as interlocking blocks that can be dragged and dropped
-like pieces of a jigsaw puzzle (see Figure~\ref{fig:detail-of-block}).
+like pieces of a jigsaw puzzle. % (see Figure~\ref{fig:detail-of-block}).
 Meadow also provides functionality to generate syntactically correct
@@
 
-Additionally, Meadow allows the user to use Meadow's Fay embedding from within the browser (see 
-Figure~\ref{fig:embedded-demo}). Meadow includes a pruned copy of the Fay compiler, bundled with its Prelude module, 
+Additionally, Meadow allows the user to use Meadow's Fay embedding from within the browser. % (see Figure~\ref{fig:embedded-demo}). 
+Meadow includes a pruned copy of the Fay compiler, bundled with its Prelude module, 
 its foreign interface, and a module with the definitions of the Marlowe primitives and a function to pretty-print 
@@
 
-\begin{figure}
+\begin{figure}[t]
 \begin{centering}
-\includegraphics[width=1\textwidth]{pix/screenshot1}
+\includegraphics[width=1.1\textwidth]{pix/screenshot1}
 \par\end{centering}
@@
 
-\begin{figure}
-\centering{}\includegraphics[scale=0.5]{pix/detail1}\caption{\label{fig:detail-of-block}Detail of five interlocked 
-blocks.}
-\end{figure}
+%\begin{figure}
+%\centering{}\includegraphics[scale=0.5]{pix/detail1}\caption{\label{fig:detail-of-block}Detail of five interlocked 
+%blocks.}
+%\end{figure}
 
@@
 
-\begin{figure}
-\begin{centering}
-\includegraphics[width=1\textwidth]{pix/embedded.png}
-\par\end{centering}
-\caption{\label{fig:embedded-demo}Screenshot of Meadow's embedded Fay editor.}
-\end{figure}
+%\begin{figure}
+%\begin{centering}
+%\includegraphics[width=1\textwidth]{pix/embedded.png}
+%\par\end{centering}
+%\caption{\label{fig:embedded-demo}Screenshot of Meadow's embedded Fay editor.}
+%\end{figure}
 
@@
 
-Thanks to IOHK staff Manuel Chakravarty, Duncan Coutts, Bernardo David, Charles Hoskinson, Aggelos Kiayias, Darryl 
-McAdams, Bruno Woltzenlogel Paleo, and Phil Wadler for their collaboration, and to colleagues from the Universities of 
+Thanks to IOHK staff Manuel Chakravarty, Duncan Coutts, Bernardo David, Charles Hoskinson, Aggelos Kiayias, % Darryl McAdams, 
+Bruno Woltzenlogel Paleo, Rebecca Valentine and Phil Wadler for their collaboration, and to colleagues from the Universities of 
 Kent and Leicester for their comments on previous presentations of this work.

### 2018-05-17 tex/paper.tex
moved related work to the end; edits/changes to the intro

@@
 In doing this, we are required to refine the model of contracts in a number of ways in order to fit with a radically 
-different context. Let us consider the following example of an ``escrow'' contract so that we can explain the motivation 
+different context. Consider the following example of an ``escrow'' contract so that we can explain the motivation 
 more concretely. The aim of this contract, written in functional pseudocode in the style of~\cite{PeytonJones:2000} 
@@
 \haskellinline{1} does not pay the money into escrow, or participant \haskellinline{3} chooses to keep it for herself, 
-then we can assume that they can be sued for the money and probably damages on top, thus providing both legal and 
+then they can be sued for the money (and probably damages), thus providing both legal and 
 financial incentives for compliance. On the other hand, in the decentralised blockchain model, where there is no central 
-authority, the contract needs to be enforced by design.
-
-This means that we require participants to commit money to cover all possible expenditure \emph{in advance of the 
+authority, the contract needs to be enforced \emph{by design}.
+
+This means that we require participants to \emph{commit} money to cover all possible expenditure \emph{in advance of the 
 contract executing}. In order to make sure that participants continue to engage with a contract, we ensure urgency by 
-imposing timeouts: money is committed for a finite period, for example, and when waiting for a participant to make a 
-commitment we also impose a timeout to ensure that the contract does not become stuck even if one or more of the 
-participants stop interacting with it. We revisit this example in Section~\ref{section:example-escrow} where we show how 
-commitments, timeouts and other features of Marlowe are used to describe this as a self-enforcing contract ``on the 
-blockchain''.
-
-Moreover, there is not a single blockchain model, and we give a brief overview of the different mechanisms for 
-scripting on blockchain (although this is not the primary purpose of this article). 
-A particularly important technical distinction here is that between an account-style model as used by 
-Ethereum~\cite{EthereumRationale}, and the unspent transaction output (UTxO) model of bitcoin~\cite{sok}. 
-There is also the question of how precisely the execution of the contract interacts with the blockchain, and how 
-information is to be recorded on the chain. Finally, should the execution ``push'' for changes  to happen, or do 
-external actors have to ``pull'' to effect changes enabled by execution of the contract?
-
-The paper describes version 1.3 of Marlowe, and it should be seen as a work in progress. We have designed it to provide 
-financial contracts in Cardano, but also in order to start to answer the twin questions of what smart contracts will 
-look like in Cardano, and how those contracts will be implemented in the overall system.
-
-The paper is structured like this: we begin in Section \ref{section:design} with a discussion of the design rationale 
+imposing \emph{timeouts}: money is committed for a finite period only. We also impose a timeout when  waiting for a participant to make a 
+commitment to ensure that the contract does not become stuck even if one of the 
+participants stops interacting with it. We discuss this particular example in more detail in Marlowe in Section~\ref{section:example-escrow} below.
+
+We make the following contributions in this paper.
+\begin{itemize}
+\item
+The design of the Marlowe language as a Haskell data type; Marlowe is made more expressible through its embedding in Haskell.
+\item
+An executable, small-step semantics of Marlowe in Haskell.
+\item
+The Meadow tool that allows users to simulate the operation of Marlowe contracts, including ``embedded'' contracts that use features of Haskell. 
+\end{itemize}
+
+%Moreover, there is not a single blockchain model, and we give a brief overview of the different mechanisms for 
+%scripting on blockchain (although this is not the primary purpose of this article). 
+%A particularly important technical distinction here is that between an account-style model as used by 
+%Ethereum~\cite{EthereumRationale}, and the unspent transaction output (UTxO) model of bitcoin~\cite{sok}. 
+%There is also the question of how precisely the execution of the contract interacts with the blockchain, and how 
+%information is to be recorded on the chain. Finally, should the execution ``push'' for changes  to happen, or do 
+%external actors have to ``pull'' to effect changes enabled by execution of the contract?
+%
+%The paper describes version 1.3 of Marlowe, and it should be seen as a work in progress. We have designed it to provide 
+%financial contracts in Cardano, but also in order to start to answer the twin questions of what smart contracts will 
+%look like in Cardano, and how those contracts will be implemented in the overall system.
+
+The paper is structured like this: REWRITE we begin in Section \ref{section:design} with a discussion of the design rationale 
 for Marlowe, including an overview of the various approaches to scripting in blockchain as well as the different designs 
@@
 
-\subsection{Smart contracts on blockchains}
-
-As more than one wag has observed, many of the programs that run on blockchains are neither smart nor indeed form 
-contracts in any real sense. Nonetheless, this terminology has stuck, and has become shorthand for ``programmed 
-interactions on the blockchain''. In an earlier paper, we presented a survey of the approaches 
-available~\cite{cryptoeprint:2016:1156}; we give an overview here.
-
-Blockchains are executed in a replicated form by parties who cannot be guaranteed not to be hostile, either by directly 
-trying to change the contents of the chain, or through trying to affect other properties of the chain by indirect means 
-(such as swamping honest parties with work). Programming on the blockchain therefore needs to be constrained in some 
-ways, since it has to be amenable to replication or verification within a reasonable time if the security and integrity 
-of the chain are to be preserved. We look at representatives of the main approaches now.
-
-\paragraph{Bitcoin script}
-
-One approach to this is to choose mechanisms such as bitcoin script~\cite{BitcoinWikiScript} which are manifestly 
-non-Turing complete. A bitcoin script, written in a Forth-like language, is essentially linear: it can branch, but the 
-language contains neither looping constructs nor recursion. It is therefore straightforward not only to see that scripts 
-will terminate, but also to give an accurate estimate of the time taken to execute a script.
-
-\paragraph{Ethereum}
-
-On the other hand, the Ethereum system~\cite{wood2014ethereum} provides a Turing-complete language for the EVM virtual 
-machine, and a higher-level programming language, Solidity, that compiles into EVM code. However, EVM and Solidity 
-programs are constrained \emph{post hoc} by two mechanisms: program execution must be paid for using `gas' proportional 
-to the effort expended, and a set of \emph{ad hoc} limits on program execution, e.g.\ on stack size.
-
-\paragraph{Nxt}
-
-In Nxt~\cite{Nxt}, programmability of the system is provided through a ``fat'' high-level 
-API, which is accessible from Nxt clients through a REST interface. The API provides functionality supporting various 
-kinds of transactions. The core software itself does not support any form of scripting language; rather, users are 
-expected to work with the built in transaction types and transactions that support some 250 primitive operations; these 
-can be ``scripted'' in a client (only) using a binding to the API, which is available, for instance, in JavaScript.
-
-\paragraph{Multiple languages}
-
-A common feature of many blockchain platforms is that they provide multiple scripting languages. As we noted earlier, 
-Ethereum can be programmed at the EVM level as well as by using the high-level Solidity language. 
-Tezos~\cite{tezos-white-paper} supports the stack-based, strongly-typed, functional language Michelson, but also 
-provides a high-level language, Liquidity, that compiles into Michelson. Liquidity is also functional and strongly 
-typed, but provides more constructs in a more familiar syntax, namely that of a subset of OCaml.
-
-The \ae{}ternity system~\cite{aeternity} provides multiple languages and VMs~\cite{Stenman-CODE-BEAM}: the functional language  Sophia (akin to 
-Reason) and the Functional Typed Warded Virtual Machine (FTWVM) for safe ``system level programming'', the language 
-Varna and the HLM for simple contracts, and a (port of) Solidity and the EVM for compatibility with Ethereum. In a 
-similar way, Cardano provides support for IELE~\cite{IELE}, a rational reconstruction of the EVM, and thus for Solidity 
-contracts too.
-
-\paragraph{Domain-specific languages}
-
-A domain-specific language or DSL is a high-level language designed to work in a specific field or domain. The intention 
-is that because the users will know about the field, the constructs of the language can be designed to be meaningful to 
-them, and also that, because of its nature, the DSL need not include all the features of a general purpose language. 
-Removing this clutter and having the remaining operations directly reflect the application area is intended to make the 
-language more accessible to domain experts who do not necessarily see themselves as programmers.
-
-Stand-alone DSLs have the advantage of providing appropriate, domain-level error messages when things go wrong, but 
-suffer the disadvantage of having to be implemented from scratch. An alternative to this are \emph{embedded} DSLs 
-(EDSLs), which provide a ``little language'' for the particular domain embedded within a general-purpose host language. 
-This means that parts of the host -- such as arithmetical expressions, or list idioms --  can be used to extend the 
-expressibility of the DSL. A notable example of this is the financial contracts language described by Peyton Jones and 
-his collaborators~\cite{PeytonJones:2000}, which is embedded in Haskell.
-
-\paragraph{Other approaches}
-
-There are a variety of other approaches to scripting, including business process notations, such as the OMG's BPMN; 
-process calculi such as Rholang; logic rules (in Formal Contract Logic), and finite state machines. More details of 
-these are provided in~\cite{cryptoeprint:2016:1156}.
-
-\paragraph{Related work}
-
-The Findel project~\cite{findel} examines financial contracts on the Ethereum platform, based on the 
-seminal~\cite{PeytonJones:2000}, and the authors note that payments need to be bounded; this is made concrete in our 
-account by our notion of commitments. They take no account of commitments or timeouts as we do below, but it should be 
-noted that the Ethereum platform is more powerful than the one that we target in this paper.
-
 \subsection{Blockchain platforms: Cardano and others}
@@
 
-A running contract will interact with its environment in two ways, as shown in Figure \ref{fig:context}.
+A running contract interacts with its environment in two ways, as in Figure \ref{fig:context}.
 
@@
 
-\begin{figure}
+\begin{figure}[t]
 \centering{}\includegraphics[scale=0.5]{pix/detail2}\caption{\label{fig:detail-of-interface}Detail of the interface for 
@@
 \end{itemize}
-\begin{figure}
-\begin{centering}
-\includegraphics[scale=0.5]{pix/detail3}
-\par\end{centering}
-\caption{\label{fig:detail-of-manual-interface}Detail of the manual interface.}
-\end{figure}
-
-\begin{figure}
-\centering{}\includegraphics[scale=0.5]{pix/detail4}\caption{\label{fig:detail-of-smart-interface}Detail of the smart 
-interface.}
-\end{figure}
+%\begin{figure}
+%\begin{centering}
+%\includegraphics[scale=0.5]{pix/detail3}
+%\par\end{centering}
+%\caption{\label{fig:detail-of-manual-interface}Detail of the manual interface.}
+%\end{figure}
+%
+%\begin{figure}
+%\centering{}\includegraphics[scale=0.5]{pix/detail4}\caption{\label{fig:detail-of-smart-interface}Detail of the smart 
+%interface.}
+%\end{figure}
 
@@
 
+\section{Related work}
+
+Blockchains are executed in a replicated form by parties who cannot be guaranteed not to be hostile, either by directly 
+trying to change the contents of the chain, or through trying to affect other properties of the chain by indirect means 
+(such as swamping honest parties with work). Programming on the blockchain therefore needs to be constrained in some 
+ways, since it has to be amenable to replication or verification within a reasonable time if the security and integrity 
+of the chain are to be preserved. We look at representatives of the main approaches now; we discuss some of these in more detail in an earlier paper~\cite{cryptoeprint:2016:1156}.
+
+\paragraph{Bitcoin script}
+
+One approach to this is to choose mechanisms such as bitcoin script~\cite{BitcoinWikiScript} which are manifestly 
+non-Turing complete. A bitcoin script, written in a Forth-like language, is essentially linear: it can branch, but the 
+language contains neither looping constructs nor recursion. It is therefore straightforward not only to see that scripts 
+will terminate, but also to give an accurate estimate of the time taken to execute a script.
+
+\paragraph{Ethereum}
+
+On the other hand, the Ethereum system~\cite{wood2014ethereum} provides a Turing-complete language for the EVM virtual 
+machine, and a higher-level programming language, Solidity, that compiles into EVM code. However, EVM and Solidity 
+programs are constrained \emph{post hoc} by two mechanisms: program execution must be paid for using `gas' proportional 
+to the effort expended, and a set of \emph{ad hoc} limits on program execution, e.g.\ on stack size.
+
+\paragraph{Nxt}
+
+In Nxt~\cite{Nxt}, programmability of the system is provided through a ``fat'' high-level 
+API, which is accessible from Nxt clients through a REST interface. The API provides functionality supporting various 
+kinds of transactions. The core software itself does not support any form of scripting language; rather, users are 
+expected to work with the built in transaction types and transactions that support some 250 primitive operations; these 
+can be ``scripted'' in a client (only) using a binding to the API, which is available, for instance, in JavaScript.
+
+\paragraph{Multiple languages}
+
+A common feature of many blockchain platforms is that they provide multiple scripting languages. As we noted earlier, 
+Ethereum can be programmed at the EVM level as well as by using the high-level Solidity language. 
+Tezos~\cite{tezos-white-paper} supports the stack-based, strongly-typed, functional language Michelson, but also 
+provides a high-level language, Liquidity, that compiles into Michelson. Liquidity is also functional and strongly 
+typed, but provides more constructs in a more familiar syntax, namely that of a subset of OCaml.
+
+The \ae{}ternity system~\cite{aeternity} provides multiple languages and VMs~\cite{Stenman-CODE-BEAM}: the functional language  Sophia (akin to 
+Reason) and the Functional Typed Warded Virtual Machine (FTWVM) for safe ``system level programming'', the language 
+Varna and the HLM for simple contracts, and a (port of) Solidity and the EVM for compatibility with Ethereum. In a 
+similar way, Cardano provides support for IELE~\cite{IELE}, a rational reconstruction of the EVM, and thus for Solidity 
+contracts too.
+
+\paragraph{Domain-specific languages}
+
+A domain-specific language or DSL is a high-level language designed to work in a specific field or domain. The intention 
+is that because the users will know about the field, the constructs of the language can be designed to be meaningful to 
+them, and also that, because of its nature, the DSL need not include all the features of a general purpose language. 
+Removing this clutter and having the remaining operations directly reflect the application area is intended to make the 
+language more accessible to domain experts who do not necessarily see themselves as programmers.
+
+Stand-alone DSLs have the advantage of providing appropriate, domain-level error messages when things go wrong, but 
+suffer the disadvantage of having to be implemented from scratch. An alternative to this are \emph{embedded} DSLs 
+(EDSLs), which provide a ``little language'' for the particular domain embedded within a general-purpose host language. 
+This means that parts of the host -- such as arithmetical expressions, or list idioms --  can be used to extend the 
+expressibility of the DSL. A notable example of this is the financial contracts language described by Peyton Jones and 
+his collaborators~\cite{PeytonJones:2000}, which is embedded in Haskell.
+
+\paragraph{Other approaches}
+
+There are a variety of other approaches to scripting, including business process notations, such as the OMG's BPMN; 
+process calculi such as Rholang; logic rules (in Formal Contract Logic), and finite state machines. More details of 
+these are provided in~\cite{cryptoeprint:2016:1156}.
+
+\paragraph{Findel}
+
+The Findel project~\cite{findel} examines financial contracts on the Ethereum platform, based on the 
+seminal~\cite{PeytonJones:2000}, and the authors note that payments need to be bounded; this is made concrete in our 
+account by our notion of commitments. They take no account of commitments or timeouts as we do below, but it should be 
+noted that the Ethereum platform is more powerful than the one that we target in this paper.
+
 \section{Next steps}
@@
 
-Thanks to IOHK staff Manuel Chakravarty, Duncan Coutts, Bernardo David, Charles Hoskinson, Aggelos Kiayias, % Darryl McAdams, 
+Thanks to Manuel Chakravarty, Duncan Coutts, Bernardo David, Charles Hoskinson, Aggelos Kiayias, % Darryl McAdams, 
 Bruno Woltzenlogel Paleo, Rebecca Valentine and Phil Wadler for their collaboration, and to colleagues from the Universities of 

### 2018-05-17 tex/paper.tex
Pack together screenshots of Meadow

--- 2018-05-17 tex/paper.tex
@@
 \begin{figure}
-\centering{}\includegraphics[scale=0.5]{pix/detail2}\caption{\label{fig:detail-of-interface}Detail of the interface for 
+\centering{}\includegraphics[scale=0.52]{pix/detail}\caption{\label{fig:detail-of-interface}Detail of the interface for 
 contract
@@
 \begin{itemize}
-\item The manual interface (see Figure~\ref{fig:detail-of-manual-interface}),
-provides a template for each of the four possible types of input:
+\item The manual interface, provides a template for each of the four possible types of input:
 commits, redeems, payment claims, and choices.
-\item The smart interface (see Figure~\ref{fig:detail-of-smart-interface}),
+\item The smart interface (shown in Figure~\ref{fig:detail-of-interface}),
 calculates the possible operations that would make sense given the
@@
 \end{itemize}
-\begin{figure}
-\begin{centering}
-\includegraphics[scale=0.5]{pix/detail3}
-\par\end{centering}
-\caption{\label{fig:detail-of-manual-interface}Detail of the manual interface.}
-\end{figure}
-
-\begin{figure}
-\centering{}\includegraphics[scale=0.5]{pix/detail4}\caption{\label{fig:detail-of-smart-interface}Detail of the smart 
-interface.}
-\end{figure}
+%\begin{figure}
+%\begin{centering}
+%\includegraphics[scale=0.5]{pix/detail3}
+%\par\end{centering}
+%\caption{\label{fig:detail-of-manual-interface}Detail of the manual interface.}
+%\end{figure}
+
+%\begin{figure}
+%\centering{}\includegraphics[scale=0.5]{pix/detail4}\caption{\label{fig:detail-of-smart-interface}Detail of the smart 
+%interface.}
+%\end{figure}
 

### 2018-05-17 tex/paper.tex

--- 2018-05-17 tex/paper.tex
@@
 \begin{figure}[t]
-\centering{}\includegraphics[scale=0.5]{pix/detail2}\caption{\label{fig:detail-of-interface}Detail of the interface for 
+\centering{}\includegraphics[scale=0.52]{pix/detail}\caption{\label{fig:detail-of-interface}Detail of the interface for 
 contract

### 2018-05-17 tex/paper.tex
changes to section 2.1

--- 2018-05-17 tex/paper.tex
@@
 
-\section{Design rationale}
-\label{section:design}
-
-\medskip
-\noindent
-In this section, we look at existing work on smart contracts and blockchains in general, before exploring the rationale 
-for building the Marlowe language.
-
-\subsection{Blockchain platforms: Cardano and others}
-
-There is not a single model of how a blockchain operates, and in this section we look at some of the tradeoffs possible 
-in this space. This allows us to give a context to some of the design decisions taken in designing Marlowe.
-
-\paragraph{UTxO and accounts}
-
-The simplest blockchain is Bitcoin. Transactions on Bitcoin are made by spending the (as yet) unspent outputs of 
-previous transactions (`unspent transaction outputs' or UTxOs). The chain need not maintain any state, such as the 
-current value owned by any particular participant: such \emph{account} information is implicit and external to the 
-Bitcoin blockchain. On the other hand, the Ethereum model explicitly keeps track of account values, and so this 
-information needs to be kept as state information on chain. While it is possible to translate between UTxO-based and 
-account-based chains \sjtnote{cite Bruno paper} the UTxO model is more fundamental, and requires less support from the 
-implementation of the chain itself.
-
-\paragraph{Interaction modality}
-
-Contracts can be conceived of as acting in two ways. In a \emph{push} model, contracts are seen to be pro-active, and to 
-make things happen. In the case of blockchain, this would include making payments or transactions happen. 
-Alternatively, in the \emph{pull} model, contracts enable certain things to happen, such as transactions, but the 
-transactions need to be effected by an external actor. The pull model makes fewer demands on the blockchain 
-implementation than the push model, which requires the contract to be triggered automatically.
-
-\paragraph{Layered design and sidechains}
-
-Some chains have a layered design, in particular Cardano~\cite{cardano}. The Cardano Settlement Layer (SL) is to 
-support settlement of transactions but nothing more complicated, whereas the Computation Layer (or CL) supports more 
-powerful general computations over the chain. Moreover, the Cardano roadmap~\cite{cardano-rationale} envisages the 
-possibility of computation taking place, in part at least, on a sidechain rather than on the main chain itself, which 
-may support only the SL. Other off-chain approaches include the Lightning network, overlaid on blockchains (usually 
-Bitcoin) and the channels of \ae{}ternity, introduced below.
-
-\subsection{Designing Marlowe}
-
-In designing Marlowe, we have chosen a number of options; this description applies to Marlowe version 1.3.
-
-\paragraph{Pull}
-
-While earlier versions used a ``push'' model of operations, we have revised the language to give a ``pull'' semantics. 
-This means, for example, that the language can \emph{enable} a payment to be made, but cannot make the payment happen. 
-It is up to the interested party (or parties) to make the change happen. It is therefore possible for the payment not to 
-happen, despite the fact that this is against the interest of the receiving party.
-
-\paragraph{UTxO vs.\ accounts}
-
-The language semantics is described in terms of \emph{commitments} of sums of cryptocurrency -- which we choose to 
-abbreviate to ``cash'' or ``money'' -- and so is agnostic of the underlying implementation.  However, the current 
-redemption mechanism in Plutus Core is not strong enough to support Marlowe, and we are actively investigating the best 
-way to extend this mechanism to implement Marlowe in full on Plutus Core.
-
-\paragraph{DSL or EDSL?}
-
-Marlowe is described as an algebraic \haskellinline{data} type in Haskell, and as such can be seen as a simple EDSL. 
-Note, however, that the implementation presented here is for the purposes of presenting the semantics and behaviour of 
-the language in the abstract. The implementation on the Cardano platform could be stand-alone, but more likely it will 
-be a DSL embedded in Plutus or Plutus Core. The details of this remain to be decided, as noted in the introduction.
-
 \begin{figure}[t]
@@
 The DSL is modelled as an algebraic type in Haskell, together with an executable small-step semantics. 
-We start by looking at the different types used by the model, then we look at the \haskellinline{Contract} DSL itself, 
-and finally we give its semantics in Haskell. Section \ref{section:example-saving} gives an example of Marlowe in 
-action, and Section
-\ref{section:example-escrow} shows the advantage of seeing Marlowe as a DSL embedded in Haskell.
+We start by looking at the different types used by the model, and the assumptions about the infrastructure in which contracts will be run. We then we look at the \haskellinline{Contract} DSL itself, 
+and finally we give its semantics in Haskell. Section 
+%\ref{section:example-saving} gives an example of Marlowe in action, and Section
+\ref{section:example-escrow} revisits the ``escrow'' example using the embedding of Marlowe as a DSL in Haskell.
 
@@
 current block number and, random numbers, as well as ``real world'' quantities like ``the price of oil'' or ``the 
-exchange rate between currencies A and B''. Each instance of such an observable will be recorded on the blockchain to 
-allow the computation to be repeated for verification purposes. It is assumed that at each step of the execution of the 
-contract, the values of some observables will be needed, and these values are (together) given by a value of type 
-\haskellinline{OS}. Note that these values are not determined by the participants in the contract, but rather by the 
-external environment.
-
-As the examples illustrate, observables come both from aspects of the blockchain (e.g.\ current block number) and 
-externally. In the latter case it will be necessary to agree a trusted oracle or beacon giving the value, which can then 
-be stored (or a hash stored) on chain.
-
-\paragraph{Inputs}
+exchange rate between currencies A and B''. 
+As the examples illustrate, observables come both from aspects of the blockchain (e.g.\ the current block number) and 
+externally. In the latter case it will be necessary to agree a trusted oracle or beacon giving the value.
+
+Each instance of such an observable will be observed at a particular time and in a particular context.
+We assume that the system infrastructure ensures that these values are recorded on the blockchain to
+allow the computation to be repeated for verification purposes.
+
+It is assumed that at each step of the execution of the 
+contract, the values of observables will be available if needed, and these values are (together) given by a value of type 
+\haskellinline{OS} (for ``observable set''), where individual observations are described in a 
+``little language'' for that purpose: \haskellinline{Observation}. 
+Note that these values are not determined by the participants in the contract, but rather by the 
+external environment in which the contract is run.
+
+\paragraph{Inputs and Commitments}
 
 On the other hand, at each step there are -- potentially, at least -- a variety of inputs available from the 
-participants. These include commitments of currency (or ``cash''), redemption of commitments, and claims of payments by 
-a participant; moreover it is also possible for a participant to input an arbitrary value (which we term a ``choice''). 
+participants themsleves. These include commitments of currency (or ``cash''), redemption of commitments, and claims of payments by 
+a participant. <oreover it is also possible for a participant to input an arbitrary value (which we term a ``choice''). 
 The particular inputs at a given step are described by a value of type \haskellinline{Input}.
 
-\paragraph{Commitments}
+%\paragraph{Commitments}
 
 While informally we might see a commitment to something as being indefinite, it is important to realise that on 
-blockchain a commitment needs to have a timeout so that progress can be forced in a game. After the timeout period the 
+blockchain a commitment needs to have a timeout so that progress can be forced in a contract. After the timeout period the 
 cash can be refunded through the user creating a transaction to reclaim the cash.
-Information about the commitments currently in force forms the \haskellinline{State}, which is (potentially) modified 
+Information about the commitments currently in force forms the \haskellinline{State}, which can be  modified 
 at each execution step.
+
+\paragraph{Actions}
 
@@
 
-\paragraph{The \haskellinline{step} function}
-
-This analysis justifies the type of the \haskellinline{step} function:
-
-\begin{haskellcode}
-step :: Input -> State -> Contract -> OS -> (State,Contract,AS)
-\end{haskellcode}
-
-which is also shown in Figure \ref{fig:step-function}.
-
 \paragraph{Infrastructure}
@@
 \item It is assumed that cryptographic functions and operations are provided by a layer external to the system, and so 
-they need not be modelled explicitly here.
-\item We assume that time is ``coarse grained'' and measured simply by block number, so that timeouts etc.\ are 
-delimited using nothing finer than block number. 
-\item Because we assume a ``pull'' model, making a commitment is not something that a contract can perform; rather, it 
-can enable that a commitment is made, but that then has to be established externally: hence the input of (a set of) 
-commitments at each step. The same observation also applies to redeeming a commitment of cash. Commitments that have 
-been made or redeemed need to be recorded somehow.
-
-\item Observables are observed at a particular time and in a particular context. We assume that the recording and 
-identification of each (instance of an) observation is provided by the infrastructure of the system. We provide a 
-``little language'' for describing observations: \haskellinline{Observation}. 
+they need not be modelled explicitly.
+\item We assume that time is ``coarse grained'' and measured by block number, so that in particular timeouts are 
+delimited using block numbers. 
 \item 
-Similar properties are true for
-\haskellinline{Money} amounts: their values are calculated depending on the inputs and state of execution of the 
-contract, they can be deterministically established given the context, and they are described through a small 
+Making a commitment is not something that a contract can perform; rather, it 
+can request that a commitment is made, but that then has to be established externally: hence the input of (a set of) 
+commitments at each step.
+\item
+IS THIS NEEDED???
+Amounts of \haskellinline{Money} amounts can be calculated depending on the inputs and state of execution of the 
+contract. They can be deterministically established given the context, and they are described through a small 
 ``expression language''.
-\item It is important to be aware that there may be different instances of contracts executed, and for each instance 
-there will typically be different instances of observables (and commitments) observed and recorded.
 \item The model manages the release of funds back to the committer when a cash commitment expires (see discussion of 
@@
 \subsection{The \haskellinline{step} function}
+
+The \haskellinline{step} function represents a \emph{single computation step} and has this type:
+
+\begin{haskellcode}
+step :: Input -> State -> Contract -> OS -> (State,Contract,AS)
+\end{haskellcode}
+
+which is also shown in Figure \ref{fig:step-function}.
 
@@
 In the next section, we look at how to take advantage of Haskell embedding when implementing Marlowe contracts.
+
+\section{Design rationale}
+\label{section:design}
+
+\medskip
+\noindent
+In this section, we look at existing work on smart contracts and blockchains in general, before exploring the rationale 
+for building the Marlowe language.
+
+\subsection{Blockchain platforms: Cardano and others}
+
+There is not a single model of how a blockchain operates, and in this section we look at some of the tradeoffs possible 
+in this space. This allows us to give a context to some of the design decisions taken in designing Marlowe.
+
+\paragraph{UTxO and accounts}
+
+The simplest blockchain is Bitcoin. Transactions on Bitcoin are made by spending the (as yet) unspent outputs of 
+previous transactions (`unspent transaction outputs' or UTxOs). The chain need not maintain any state, such as the 
+current value owned by any particular participant: such \emph{account} information is implicit and external to the 
+Bitcoin blockchain. On the other hand, the Ethereum model explicitly keeps track of account values, and so this 
+information needs to be kept as state information on chain. While it is possible to translate between UTxO-based and 
+account-based chains \sjtnote{cite Bruno paper} the UTxO model is more fundamental, and requires less support from the 
+implementation of the chain itself.
+
+\paragraph{Interaction modality}
+
+Contracts can be conceived of as acting in two ways. In a \emph{push} model, contracts are seen to be pro-active, and to 
+make things happen. In the case of blockchain, this would include making payments or transactions happen. 
+Alternatively, in the \emph{pull} model, contracts enable certain things to happen, such as transactions, but the 
+transactions need to be effected by an external actor. The pull model makes fewer demands on the blockchain 
+implementation than the push model, which requires the contract to be triggered automatically.
+
+\paragraph{Layered design and sidechains}
+
+Some chains have a layered design, in particular Cardano~\cite{cardano}. The Cardano Settlement Layer (SL) is to 
+support settlement of transactions but nothing more complicated, whereas the Computation Layer (or CL) supports more 
+powerful general computations over the chain. Moreover, the Cardano roadmap~\cite{cardano-rationale} envisages the 
+possibility of computation taking place, in part at least, on a sidechain rather than on the main chain itself, which 
+may support only the SL. Other off-chain approaches include the Lightning network, overlaid on blockchains (usually 
+Bitcoin) and the channels of \ae{}ternity, introduced below.
+
+\subsection{Designing Marlowe}
+
+In designing Marlowe, we have chosen a number of options; this description applies to Marlowe version 1.3.
+
+\paragraph{Pull}
+
+While earlier versions used a ``push'' model of operations, we have revised the language to give a ``pull'' semantics. 
+This means, for example, that the language can \emph{enable} a payment to be made, but cannot make the payment happen. 
+It is up to the interested party (or parties) to make the change happen. It is therefore possible for the payment not to 
+happen, despite the fact that this is against the interest of the receiving party.
+
+\paragraph{UTxO vs.\ accounts}
+
+The language semantics is described in terms of \emph{commitments} of sums of cryptocurrency -- which we choose to 
+abbreviate to ``cash'' or ``money'' -- and so is agnostic of the underlying implementation.  However, the current 
+redemption mechanism in Plutus Core is not strong enough to support Marlowe, and we are actively investigating the best 
+way to extend this mechanism to implement Marlowe in full on Plutus Core.
+
+\paragraph{DSL or EDSL?}
+
+Marlowe is described as an algebraic \haskellinline{data} type in Haskell, and as such can be seen as a simple EDSL. 
+Note, however, that the implementation presented here is for the purposes of presenting the semantics and behaviour of 
+the language in the abstract. The implementation on the Cardano platform could be stand-alone, but more likely it will 
+be a DSL embedded in Plutus or Plutus Core. The details of this remain to be decided, as noted in the introduction.
 

### 2018-05-17 tex/paper.tex
added some notes to the intro about what to include there, and what to do in the body too
Add more detailed explanations to embedded example

@@
 
+TO ADD
+\begin{itemize}
+\item 
+Model vs implementation: add a section on the letter e.g. about distributed computation.
+\item
+In particular we're agnostic about UTxO and accounts, but want to be implementable on both.
+\item
+focus on some of the design decisions, but focus on the important aspects only and cut the stuff that's  more speculative/internal
+\item  
+\ldots
+\end{itemize}
+
 %Moreover, there is not a single blockchain model, and we give a brief overview of the different mechanisms for 
@@
 
-\section{The model}
+\section{The Marlowe model}
 \label{section:model}
@@
 \end{haskellcode}
-We can also create a sub-contract that allows the money to be redeemed by participant \haskellinline{1}:
+We can also create a sub-contract that allows the money from the commitment with identifier \haskellinline{IdentCC 1} to 
+be redeemed:
 \begin{haskellcode}
@@
 \end{haskellcode}
+Once redeemed, the contract continues as \haskellinline{Null} since we expect this to be the last thing that is done.
 Each participant has a say on who deserves the money: either participant \haskellinline{1} deserves a refund, which we 
@@
 $1$. Because there is only one choice to make per participant, we use the same \haskellinline{IdentChoice} as their 
-participant id. We can define a function that returns an observation that is true if and only if the participant 
-\haskellinline{per} has chosen the number \haskellinline{c} as follows:
+participant id. We can define a function that returns an observation that is true if and only if
+\haskellinline{per} has chosen the number \haskellinline{c} for the choice \haskellinline{IdentChoice per} as follows:
 \begin{haskellcode}
@@
 \end{haskellcode}
-After that, we can write a function that returns an observation that is true if and only if at least 
-two of the three participants \haskellinline{p1}, \haskellinline{p2}, and \haskellinline{p3} have chosen the number 
-\haskellinline{c} as follows:
+Building on that, we can now write a function that returns an observation that is true if and only if at least 
+two out of the three participants \haskellinline{p1}, \haskellinline{p2}, and \haskellinline{p3} have agreed in 
+choosing the number \haskellinline{c} as follows:
 \begin{haskellcode}
@@
 \begin{enumerate}
- \item Two out of three choose $0$, for `\haskellinline{pay}'. 
- \item Two out of three choose $1$, for `\haskellinline{refund}'.
- \item The block 90 is added to the blockchain.
+ \item The observation became true because two out of three have chosen $0$. 
+ \item The observation became true because two out of three have chosen $1$.
+ \item The observation remained false until 90 was published in the blockchain.
 \end{enumerate}
 
-If the third option happens first, the money is refunded. Otherwise, there is a \haskellinline{Choice} that will 
+If the third option happens, the money is refunded. Otherwise, there is a \haskellinline{Choice} that will 
 immediately refund the money to participant \haskellinline{1} unless two out of three chose option $1$; in the later 
 case, \haskellinline{Pay} will give participant \haskellinline{2} the opportunity to claim the funds available in 
-\haskellinline{iCC1} before block 100. If the funds are not claimed before block 100 they will also be refunded to 
+the commitment with identifier \haskellinline{iCC1} before block 100 (by using the identifier \haskellinline{iP1} for 
+the claim). If the funds are not claimed before block 100 they will also be refunded to 
 participant \haskellinline{1}.
 
-Reflecting on the example we can see that using Haskell definitions has made the contract substantially more 
+Reflecting on the example, we can see that using Haskell definitions has made the contract substantially more 
 comprehensible; it will also support more efficient operation, if we were to replace the repeated  expression 

### 2018-05-17 tex/paper.tex
describing the step functions

--- 2018-05-17 tex/paper.tex
@@
 
+\paragraph{Computations}
+
+Computation is modelled at two different levels.
+
+The \haskellinline{step} function represents a \emph{single computation step} and has this type:
+\begin{haskellcode}
+step :: Input -> State -> Contract -> OS -> (State,Contract,AS)
+\end{haskellcode}
+which is also illustrated in Figure \ref{fig:step-function}.
+
+The \haskellinline{step} function is total, so that for every contract a result of stepping is defined. However, for 
+some kinds of contracts -- commits, redeems or time-shifted contracts -- it is possible that the result of a step is to 
+produce the same contract as the result; such a step is called \emph{quiescent} whereas all other steps \emph{make 
+progress}. We use this distinction in the explanation that follows.
+
+Execution of a contract will involve multiple blocks, with multiple steps in each block.
+The computation at a single block is given by the \haskellinline{stepBlock} function: this will call the \haskellinline{stepAll} function that calls \haskellinline{step} repeatedly
+until it is quiescent. In addition to 
+calling \haskellinline{stepAll}, \haskellinline{stepBlock} will first enable expired cash commitments to be refunded and record any choices made at that step. 
+The functions \haskellinline{stepAll} and \haskellinline{stepBlock} have the same type as \haskellinline{step} itself.
+
 \subsection{The \haskellinline{Contract} type}
 
-The type of contracts is given by
+The type of contracts is given by the Haskell data type
 
@@
 \subsection{The \haskellinline{step} function}
-
-The \haskellinline{step} function represents a \emph{single computation step} and has this type:
-
-\begin{haskellcode}
-step :: Input -> State -> Contract -> OS -> (State,Contract,AS)
-\end{haskellcode}
-
-which is also shown in Figure \ref{fig:step-function}.
-
-The \haskellinline{step} function is total, so that for every contract a result of stepping is defined. However, for 
-some kinds of contracts -- commits, redeems or time-shifted contracts -- it is possible that the result of a step is to 
-produce the same contract as the result; such a step is called \emph{quiescent} whereas all other steps \emph{make 
-progress}. We use this distinction in the explanation that follows.
-
-The function \haskellinline{step} is wrapped by the \haskellinline{stepAll} function, which calls \haskellinline{step} 
-until it stops having any effects. For every block, the \haskellinline{stepBlock} function is called. In addition to 
-calling \haskellinline{stepAll}, \haskellinline{stepBlock} will first enable expired cash commitments to be refunded and 
-record any choices made at that step.
 

### 2018-05-17 tex/paper.tex
some compacting
added textual description to Sec 2.2, and reformatted Sec 2.3

@@
 \end{center}
+\vspace*{-8mm}
 \end{figure}
@@
 
-\paragraph{Computations}
+\paragraph{Computation}
 
@@
 The computation at a single block is given by the \haskellinline{stepBlock} function: this will call the \haskellinline{stepAll} function that calls \haskellinline{step} repeatedly
-until it is quiescent. In addition to 
+until it is quiescent.
+
+In addition to 
 calling \haskellinline{stepAll}, \haskellinline{stepBlock} will first enable expired cash commitments to be refunded and record any choices made at that step. 
@@
    \end{haskellcode}
-%Originally on "Choice": \footnote{Doesn't the step function need to record what observations are made, so they can be 
-%archived in the blockchain? I don't see how your design will support doing so.}\footnote{Yes, we assume that it is 
-%part of the infrastructure of this language to record the values of *observables* that are used. However, we have 
-%chosen no to make it part of the language itself.}
-
+Informally, this type provides a \haskellinline{Null} contract, which does nothing.
+The next three constructs form contracts that do something, and then continue according to another contract (which is one of the components of the original contract).
+\haskellinline{CommitCash} will wait for a participant to make a commitment,
+\haskellinline{RedeemCC} allows for a commitment to be redeemed, and \haskellinline{Pay} for a payment to be made to another participant.
+
+The remaining constructors form composite contracts from simpler components: \haskellinline{Both} has the behaviour of both its components, \haskellinline{Choice} chooses between two contracts on the basis of an observation, and \haskellinline{When} is quiescent until a condition -- i.e.\ an \haskellinline{Observation} -- becomes true.
+
+What many of the contracts also have are timeouts that also determine their behaviour. 
 \subsection{The \haskellinline{step} function}
 
-The constructors are interpreted like this:
-
-\begin{itemize}
-
-\item \haskellinline{Null} is the null contract; it will always be quiescent:
+In this section we explain the detailed behaviour of \haskellinline{Contracts} through a discussion of the \haskellinline{step} function.
 
 \smallskip
+\noindent
+\haskellinline{Null} is the null contract; it will always be quiescent:
+
 \begin{haskellcode}
@@
 \end{haskellcode}
+
 \smallskip
-
-\item \haskellinline{CommitCash idCC p v t1 t2 k1 k2}  For this contract to make progress,\begin{itemize}
+\noindent
+\haskellinline{CommitCash idCC p v t1 t2 k1 k2}  For this contract to make progress,\begin{itemize}
 \item either before the timeout \haskellinline{t1} the user \haskellinline{p} makes a cash commitment of value 
@@
 
-\smallskip
 \begin{haskellcode}
@@
 \end{haskellcode}
-\smallskip
 In the first case, a \haskellinline{SuccessfulCommit} action is generated and the contract continues as 
@@
 
-\item \haskellinline{RedeemCC idCC k} (\haskellinline{CC} stands for cash commitment.) For this contract to make 
+\medskip
+\noindent
+\haskellinline{RedeemCC idCC k} (\haskellinline{CC} stands for cash commitment.) For this contract to make 
 progress, the creator of the cash commitment with identifier \haskellinline{idCC} is required to redeem the unspent 
@@
 \end{haskellcode}
-\smallskip
-
 Committed cash can only be redeemed once, and an attempt to redeem it a second time will produce a  
@@
 
-\item \haskellinline{Pay p1 p2 v t k} makes it possible, assuming that sufficient funds are available, for 
+\medskip
+\noindent
+\haskellinline{Pay p1 p2 v t k} makes it possible, assuming that sufficient funds are available, for 
 \haskellinline{p1} to claim  a payment of \haskellinline{v} from \haskellinline{p2} before the timeout 
@@
 \end{haskellcode}
-\smallskip
-
 By `available' we mean that sufficient commitments have 
@@
 
-\item \haskellinline{Both k1 k2} enforces the behaviour of both contracts \haskellinline{k1} and \haskellinline{k2}.
-
-\smallskip
+\medskip
+\noindent
+\haskellinline{Both k1 k2} enforces the behaviour of both contracts \haskellinline{k1} and \haskellinline{k2}. 
 \begin{haskellcode}
@@
 \end{haskellcode}
-\smallskip
-
 Because the model is stateful and produces output actions, to make a step, it is necessary to execute a single step of 
@@
 
-\item \haskellinline{Choice obs k1 k2} behaves as either \haskellinline{k1} or \haskellinline{k2} depending on the 
+\medskip
+\noindent
+\haskellinline{Choice obs k1 k2} behaves as either \haskellinline{k1} or \haskellinline{k2} depending on the 
 (Boolean) result of \haskellinline{obs} at the time that the observation is made, \haskellinline{k1} if it is 
@@
 \end{haskellcode}
-\smallskip
-
-\item \haskellinline{When obs timeout k def_k} This contract will not progress until \haskellinline{obs} is 
+
+\medskip
+\noindent
+\haskellinline{When obs timeout k def_k} This contract will not progress until \haskellinline{obs} is 
 \haskellinline{True} or until the current block number is greater than or equal to the one specified by timeout. In case 
@@
   \end{haskellcode}
-\smallskip
-
-\end{itemize}
 We look next at an example of Marlowe in action.
@@
 \begin{haskellcode}
-escrow = ...
-                    (When (OrObs (two_chose 1 2 3 0)
+escrow = ...        (When (OrObs (two_chose 1 2 3 0)
                                  chose_refund)
                           90
-                          (Choice chose_refund
+                          (Choice chose_refund ...
                            ...
-         ...
-         where 
-         chose_refund = two_chose 1 2 3 1                 
+         where chose_refund = two_chose 1 2 3 1                 
 \end{haskellcode}
 then the repeated computation of the expression will be avoided.
-
 In the next section, we cover the online contract visualisation and interaction tool for Marlowe.
@@
 simulating the incentivised deposit contract.}
+\vspace*{-8mm}
 \end{figure}
@@
 execution simulation.}
+\vspace*{-5mm}
 \end{figure}

### 2018-05-17 tex/paper.tex
tidied up contributions; misc other updates

@@
 \item
-The design of the Marlowe language as a Haskell data type; Marlowe is made more expressible through its embedding in Haskell.
+The design of the Marlowe language.
 \item
@@
 \item 
-The Meadow tool that allows users to simulate the operation of Marlowe contracts, including ``embedded'' contracts that use features of Haskell. 
+Extending Marlowe's expressibility by embedding it as a Haskell \haskellinline{data} type.
+\item
+The Meadow tool that allows users to interact with and simulate the operation of embedded Marlowe contracts. 
 \end{itemize}
@@
 
-The paper is structured like this: REWRITE we begin in Section \ref{section:design} with a discussion of the design rationale 
-for Marlowe, including an overview of the various approaches to scripting in blockchain as well as the different designs 
-of the blockchains themselves. We conclude Section \ref{section:design} with a set of decisions about the blockchain 
-environment for Marlowe. The central part of the paper is Section \ref{section:model} where we introduce the Marlowe 
-model, including the assumptions made in designing it, the types of the principal functions and a description of the DSL 
-as an algebraic type, constructor by constructor. 
+The paper begins in Section \ref{section:model} by introducing the Marlowe 
+model, including the assumptions made in designing it, the types of the principal functions and a description of the DSL as an algebraic type, constructor by constructor. 
 %Section \ref{section:example-saving} presents an example of a larger Marlowe contract, for incentivised saving. 
@@
 \ref{section:next-steps} enumerates the next steps for the project.
+
+we begin in Section \ref{section:design} with a discussion of the design rationale 
+for Marlowe, including an overview of the various approaches to scripting in blockchain as well as the different designs 
+of the blockchains themselves. We conclude Section \ref{section:design} with a set of decisions about the blockchain 
+environment for Marlowe.
 
@@
 which is also illustrated in Figure \ref{fig:step-function}. 
-
 The \haskellinline{step} function is total, so that for every contract a result of stepping is defined. However, for 
-some kinds of contracts -- commits, redeems or time-shifted contracts -- it is possible that the result of a step is to 
-produce the same contract as the result; such a step is called \emph{quiescent} whereas all other steps \emph{make 
+some kinds of contracts -- commits, redeems or time-shifted contracts -- it is possible that performing  a step produces the same contract as the result; we call these \emph{quiescent} steps whereas all others \emph{make
 progress}. We use this distinction in the explanation that follows.
@@
 
-\section{Marlowe as an embedded DSL in Haskell}
+\section{Marlowe as an embedded DSL}
 \label{section:example-escrow}
@@
 item. On the other hand, if participant \haskellinline{2} gives participant \haskellinline{1} the item before 
-\haskellinline{1} pays, then there is a risk that \haskellinline{1} will never pay.
+\haskellinline{1} pays,  there is a risk that \haskellinline{1} will never pay.
 
@@
 
-\section{Visualising and interacting with contracts}
+\section{Visualising and interacting with Marlowe contracts}
 

### 2018-05-17 tex/paper.tex
tidied up contributions; misc other updates

@@
 \item
-The design of the Marlowe language.
+Designing a DSL for financial contracts on blockchains: Marlowe.
 \item
-An executable, small-step semantics of Marlowe in Haskell.
+Defining an executable, small-step semantics of Marlowe in Haskell.
 \item 
-Extending Marlowe's expressibility by embedding it as a Haskell \haskellinline{data} type.
+Embedding Marlowe as a Haskell \haskellinline{data} type to extend its expressibility.
 \item
-The Meadow tool that allows users to interact with and simulate the operation of embedded Marlowe contracts. 
+Developing the Meadow tool that allows users to interact with and simulate the operation of embedded Marlowe contracts. 
 \end{itemize}

### 2018-05-17 tex/paper.tex
Unify variable naming between code and description

--- 2018-05-17 tex/paper.tex
@@
 \noindent
-\haskellinline{CommitCash idCC p v t1 t2 k1 k2}  For this contract to make progress,\begin{itemize}
-\item either before the timeout \haskellinline{t1} the user \haskellinline{p} makes a cash commitment of value 
-\haskellinline{v} and timeout \haskellinline{t2}  with the identifier \haskellinline{idCC}, 
+\haskellinline{CommitCash ident person val start_timeout end_timeout con1 con2}  For this contract to make 
+progress,\begin{itemize}
+\item either before the timeout \haskellinline{start_timeout} the user \haskellinline{person} makes a cash commitment of 
+value 
+\haskellinline{val} and timeout \haskellinline{end_timeout}  with the identifier \haskellinline{ident}, 
 \item
-or the timeout \haskellinline{t1} is exceed:
+or the timeout \haskellinline{start_timeout} is exceed:
 \end{itemize}
@@
 In the first case, a \haskellinline{SuccessfulCommit} action is generated and the contract continues as 
-\haskellinline{k1}; in the second case no action is generated and the contract continues as 
-\haskellinline{k2}. While neither case holds, the contract is quiescent, waiting for the cash to be committed.
-
-If the cash is committed successfully and the timeout \haskellinline{t2} is reached, then it is impossible to further 
-spend the committed cash, and any unspent funds can be reclaimed by participant \haskellinline{p}.
+\haskellinline{con1}; in the second case no action is generated and the contract continues as 
+\haskellinline{con2}. While neither case holds, the contract is quiescent, waiting for the cash to be committed.
+
+If the cash is committed successfully and the timeout \haskellinline{end_timeout} is reached, then it is impossible to 
+further spend the committed cash, and any unspent funds can be reclaimed by participant \haskellinline{person}.
  This is performed by the \haskellinline{stepBlock} function, as noted above.
@@
 \noindent
-\haskellinline{RedeemCC idCC k} (\haskellinline{CC} stands for cash commitment.) For this contract to make 
-progress, the creator of the cash commitment with identifier \haskellinline{idCC} is required to redeem the unspent 
-funds in that commitment; the contract then continues as \haskellinline{k}, and the action 
+\haskellinline{RedeemCC ident con} (\haskellinline{CC} stands for cash commitment.) For this contract to make 
+progress, the creator of the cash commitment with identifier \haskellinline{ident} is required to redeem the unspent 
+funds in that commitment; the contract then continues as \haskellinline{con}, and the action 
 \haskellinline{CommitRedeemed} is produced.
@@
 Committed cash can only be redeemed once, and an attempt to redeem it a second time will produce a  
-\haskellinline{DuplicateRedeem} action, continuing as \haskellinline{k}.
-
-If the cash commitment with identifier \haskellinline{idCC} has expired, it becomes possible for the remaining funds to 
-be redeemed by the committer; this can be done by the \haskellinline{stepBlock} function processing the appropriate 
+\haskellinline{DuplicateRedeem} action, continuing as \haskellinline{con}.
+
+If the cash commitment with identifier \haskellinline{ident} has expired, it becomes possible for the remaining funds 
+to be redeemed by the committer; this can be done by the \haskellinline{stepBlock} function processing the appropriate 
 \haskellinline{Input}, and an \haskellinline{ExpiredCommitRedeemed} action will be produced.
 
-Once the commitment \haskellinline{idCC} has expired and is redeemed, a \haskellinline{RedeemCC idCC k} contract will 
-immediately evolve to \haskellinline{k}.
+Once the commitment \haskellinline{ident} has expired and is redeemed, a \haskellinline{RedeemCC ident con} contract 
+will 
+immediately evolve to \haskellinline{con}.
 %\footnote{How does the contract determine what actions resulted from a given step? Or is this not required?}
@@
 \noindent
-\haskellinline{Pay p1 p2 v t k} makes it possible, assuming that sufficient funds are available, for 
-\haskellinline{p1} to claim  a payment of \haskellinline{v} from \haskellinline{p2} before the timeout 
-\haskellinline{t}. The contract continues as \haskellinline{k}.
+\haskellinline{Pay idpay from to val expi con} makes it possible, assuming that sufficient funds are available, for 
+\haskellinline{to} to claim a payment with id \haskellinline{idpay} of \haskellinline{val} from \haskellinline{from} 
+before the timeout \haskellinline{expi}. The contract continues as \haskellinline{con}.
 
@@
 \end{haskellcode}
-By `available' we mean that sufficient commitments have 
-been made and  not yet expired to cover the payment; in this case the payment  uses the currency allocated by the cash 
-commitments made by \haskellinline{p1} that expire the earliest.
+By `available' we mean that sufficient commitments have been made and not yet expired to cover the payment; in this 
+case the payment uses the currency allocated by the cash commitments made by \haskellinline{from} that expire the 
+earliest.
 
@@
 \noindent
-\haskellinline{Both k1 k2} enforces the behaviour of both contracts \haskellinline{k1} and \haskellinline{k2}. 
+\haskellinline{Both con1 con2} enforces the behaviour of both contracts \haskellinline{con1} and \haskellinline{con2}. 
 \begin{haskellcode}
@@
 Because the model is stateful and produces output actions, to make a step, it is necessary to execute a single step of 
-each of  the contracts \haskellinline{k1} and \haskellinline{k2} in sequence: first \haskellinline{k1} then 
-\haskellinline{k2}.
+each of  the contracts \haskellinline{con1} and \haskellinline{con2} in sequence: first \haskellinline{con1} then 
+\haskellinline{con2}.
 
@@
 \noindent
-\haskellinline{Choice obs k1 k2} behaves as either \haskellinline{k1} or \haskellinline{k2} depending on the 
-(Boolean) result of \haskellinline{obs} at the time that the observation is made, \haskellinline{k1} if it is 
-\haskellinline{True} and \haskellinline{k2} if \haskellinline{False}.
+\haskellinline{Choice obs conT conF} behaves as either \haskellinline{conT} or \haskellinline{conF} depending on the 
+(Boolean) result of \haskellinline{obs} at the time that the observation is made, \haskellinline{conT} if it is 
+\haskellinline{True} and \haskellinline{conF} if \haskellinline{False}.
 
@@
 \noindent
-\haskellinline{When obs timeout k def_k} This contract will not progress until \haskellinline{obs} is 
-\haskellinline{True} or until the current block number is greater than or equal to the one specified by timeout. In case 
-the timeout applies, the contract will continue as \haskellinline{def_k}, if the timeout does not apply and 
-\haskellinline{obs} is \haskellinline{True}, then the contract continues as \haskellinline{k}. Otherwise the contract is 
-quiescent.
+\haskellinline{When obs expi con con2} This contract will not progress until \haskellinline{obs} is 
+\haskellinline{True} or until the current block number is greater than or equal to the one specified by timeout 
+\haskellinline{expi}. In case the timeout applies, the contract will continue as \haskellinline{con2}, if the timeout 
+does not apply and \haskellinline{obs} is \haskellinline{True}, then the contract continues as \haskellinline{con}. 
+Otherwise the contract is quiescent.
 

### 2018-05-17 tex/paper.tex
Remove mention to Money
Fix references to the "incentivised deposit" example

@@
 commitments at each step.
-\item
-IS THIS NEEDED???
-Amounts of \haskellinline{Money} amounts can be calculated depending on the inputs and state of execution of the 
-contract. They can be deterministically established given the context, and they are described through a small 
-``expression language''.
 \item The model manages the release of funds back to the committer when a cash commitment expires (see discussion of 
@@
 In Figure~\ref{fig:full-screenshot-demo}, we provide a screenshot
-of Meadow in the middle of simulating the execution of the ``incentivised
-saving'' contract described in Section~\ref{section:example-saving}.
+of Meadow in the middle of simulating the execution of the ``deposit incentive'' contract available
+in the GitHub repository (in the file: \texttt{src/DepositIncentive.hs}) and in the examples section of Meadow (on the 
+bottom right part).
 
@@
 \caption{\label{fig:full-screenshot-demo}Screenshot of Meadow tool
-simulating the incentivised deposit contract.}
+simulating the ``deposit incentive'' contract.}
 \vspace*{-8mm}

### 2018-05-17 tex/paper.tex
implementation section

@@
 \end{itemize}
-
-TO ADD
-\begin{itemize}
-\item 
-Model vs implementation: add a section on the letter e.g. about distributed computation.
-\item
-In particular we're agnostic about UTxO and accounts, but want to be implementable on both.
-\item
-focus on some of the design decisions, but focus on the important aspects only and cut the stuff that's  more speculative/internal
-\item  
-\ldots
-\end{itemize}
+Having established this model and its semantics we are able to do a number of other things, which we discuss in the paper.
+We can explore how the language will be implemented on an existing blockchain, such as Cardano, and our model has shown us that we will need to consider how the evolution of a contract interacts with the blockchain, with miners in the case of a `proof of work'-based chain, and with users participating in contract execution.
+We can also perform analyses of Marlowe contracts, based on its formal semantics.
+Finally, we can explore the design space for Marlowe, and explore potential alternative designs.
 
@@
 
-we begin in Section \ref{section:design} with a discussion of the design rationale 
+TO UPDATE \ldots we begin in Section \ref{section:design} with a discussion of the design rationale 
 for Marlowe, including an overview of the various approaches to scripting in blockchain as well as the different designs 
@@
 
-\section{Design rationale}
-\label{section:design}
-
-\medskip
-\noindent
-In this section, we look at existing work on smart contracts and blockchains in general, before exploring the rationale 
-for building the Marlowe language.
-
-\subsection{Blockchain platforms: Cardano and others}
-
-There is not a single model of how a blockchain operates, and in this section we look at some of the tradeoffs possible 
-in this space. This allows us to give a context to some of the design decisions taken in designing Marlowe.
-
-\paragraph{UTxO and accounts}
-
-The simplest blockchain is Bitcoin. Transactions on Bitcoin are made by spending the (as yet) unspent outputs of 
-previous transactions (`unspent transaction outputs' or UTxOs). The chain need not maintain any state, such as the 
-current value owned by any particular participant: such \emph{account} information is implicit and external to the 
-Bitcoin blockchain. On the other hand, the Ethereum model explicitly keeps track of account values, and so this 
-information needs to be kept as state information on chain. While it is possible to translate between UTxO-based and 
-account-based chains \sjtnote{cite Bruno paper} the UTxO model is more fundamental, and requires less support from the 
-implementation of the chain itself.
-
-\paragraph{Interaction modality}
-
-Contracts can be conceived of as acting in two ways. In a \emph{push} model, contracts are seen to be pro-active, and to 
-make things happen. In the case of blockchain, this would include making payments or transactions happen. 
-Alternatively, in the \emph{pull} model, contracts enable certain things to happen, such as transactions, but the 
-transactions need to be effected by an external actor. The pull model makes fewer demands on the blockchain 
-implementation than the push model, which requires the contract to be triggered automatically.
-
-\paragraph{Layered design and sidechains}
-
-Some chains have a layered design, in particular Cardano~\cite{cardano}. The Cardano Settlement Layer (SL) is to 
-support settlement of transactions but nothing more complicated, whereas the Computation Layer (or CL) supports more 
-powerful general computations over the chain. Moreover, the Cardano roadmap~\cite{cardano-rationale} envisages the 
-possibility of computation taking place, in part at least, on a sidechain rather than on the main chain itself, which 
-may support only the SL. Other off-chain approaches include the Lightning network, overlaid on blockchains (usually 
-Bitcoin) and the channels of \ae{}ternity, introduced below.
-
-\subsection{Designing Marlowe}
-
-In designing Marlowe, we have chosen a number of options; this description applies to Marlowe version 1.3.
-
-\paragraph{Pull}
-
-While earlier versions used a ``push'' model of operations, we have revised the language to give a ``pull'' semantics. 
-This means, for example, that the language can \emph{enable} a payment to be made, but cannot make the payment happen. 
-It is up to the interested party (or parties) to make the change happen. It is therefore possible for the payment not to 
-happen, despite the fact that this is against the interest of the receiving party.
-
-\paragraph{UTxO vs.\ accounts}
-
-The language semantics is described in terms of \emph{commitments} of sums of cryptocurrency -- which we choose to 
-abbreviate to ``cash'' or ``money'' -- and so is agnostic of the underlying implementation.  However, the current 
-redemption mechanism in Plutus Core is not strong enough to support Marlowe, and we are actively investigating the best 
-way to extend this mechanism to implement Marlowe in full on Plutus Core.
-
-\paragraph{DSL or EDSL?}
-
-Marlowe is described as an algebraic \haskellinline{data} type in Haskell, and as such can be seen as a simple EDSL. 
-Note, however, that the implementation presented here is for the purposes of presenting the semantics and behaviour of 
-the language in the abstract. The implementation on the Cardano platform could be stand-alone, but more likely it will 
-be a DSL embedded in Plutus or Plutus Core. The details of this remain to be decided, as noted in the introduction.
-
 } % end ignore
@@
 \par\end{centering}
-\caption{\label{fig:full-screenshot-demo}Screenshot of Meadow tool
+\caption{\label{fig:full-screenshot-demo}The Meadow tool
 simulating the ``deposit incentive'' contract.}
-\vspace*{-8mm}
+\vspace*{-6mm}
 \end{figure}
@@
 
+\section{Implementation on blockchain}
+\label{section:towards-implementation}
+
+Marlowe is defined by its executable semantics, but to be deployed on blockchain it will have to be implemented on an existing distributed infrastructure.
+We intend to deploy it on IOHK's Cardano infrastructure, but it can be implemented in other systems too, as we explain now.
+
+\subsection{Abstraction}
+
+Marlowe abstracts away from a number of concrete details of how blockchains operate.
+In particular, it is agnostic between UTxO-based systems, such as Bitcoin and Cardano SL, and account-based models such as Ethereum.
+
+\paragraph{UTxO and accounts}
+
+The simplest blockchain is Bitcoin. Transactions on Bitcoin are made by spending the (as yet) unspent outputs of 
+previous transactions (`unspent transaction outputs' or UTxOs). The chain need not maintain any state, such as the 
+current value owned by any particular participant: such \emph{account} information is implicit and external to the 
+Bitcoin blockchain. On the other hand, the Ethereum model explicitly keeps track of account values, and so this 
+information needs to be kept as state information on chain. While it is possible to translate between UTxO-based and 
+account-based chains the UTxO model is more fundamental, and requires less support from the 
+implementation of the chain itself.
+
+\paragraph{Interaction modality}
+
+Contracts can be conceived of as acting in two ways.
+In a \emph{push} model, contracts are seen to make things happen.
+In the case of blockchain, this would include making payments or transactions take plade.
+Alternatively, in a \emph{pull} model, contracts enable certain things, such as transactions,  to happen but the 
+transactions need to be effected by an external actor. The pull model makes fewer demands on the blockchain implementation than the push model.
+
+\paragraph{Layered design and sidechains}
+
+Some chains have a layered design, in particular Cardano~\cite{cardano}.
+The Cardano Settlement Layer (SL) is to support settlement of transactions but nothing more complicated, whereas the Computation Layer (or CL) supports more powerful general computations over the chain.
+Moreover, the Cardano roadmap~\cite{cardano-rationale} envisages the possibility of computation taking place, in part at least, on a sidechain rather than on the main chain itself, which
+may support only the SL. Other off-chain approaches include the Lightning network, overlaid on blockchains (usually Bitcoin) and the channels of \ae{}ternity, introduced below.
+
+\subsection{Compilation}
+
+We have designed Marlowe assuming that each contract would be compiled into
+\begin{itemize} 
+\item one or more on-chain programs and 
+\item a client side program or interface that has a limited number of choices to be made at each point in time (including an interface for the observables). 
+\end{itemize}
+The contracts implemented in the Marlowe language have an implicit interface defined for what would correspond to the client side of a distributed application, as shown by the interactive interface. 
+Of course, we probably do not want to give a user a programmatic interface but an actual application with user interface.
+So there is still some work to be done to create interfaces on top of the client side of the contracts.
+
+A contract written in Marlowe could also be potentially
+compiled to a smart-contract written in Ethereum. This could be done in two ways, it could be compiled so that an 
+smart-contract represents a single instance of the Marlowe contract (with a fixed set of participants), or 
+alternatively it would be possible to create a generic smart-contract with an extra ``instantiate'' operation that 
+sets-up a new Marlowe contract instance with a given set of participants (represented by a set of public keys).
+
+\paragraph{Transactions}
+
+In principle, we can directly translate each operation in the model (i.e: cash commitment, payment claim, choice, or
+redemption) into a separate transaction in the blockchain. In the Ethereum model this would work in the same 
+way that is simulated by Meadow: several operations can be issued in parallel and the blockchain subsystem will 
+automatically apply them in some order and integrate them into the next block.
+
+Another approach would be to use a UTxO model with continuations in which a UTxO represents a contract. In this model,
+an operation would correspond to a transaction that spends a UTxO and creates (at least) a new UTxO that represents the 
+contract after the operation.
+
+Usually, in the UTxO model, it is necessary that a UTxO is in the blockchain in order for another transaction to be able to spend it.
+This could potentially limit the number of operations that can be applied in a single block to one, which would make infeasible the approach of using the same UTxO to represent several instances of the same contract (since it would not scale).
+On the other hand, it would prevent non-determinism and, with it, it would remove the possibility of race-conditions.
+A change to the state of a contract before a transaction (and its corresponding operation) is accepted in the blockchain
+would invalidate the transaction and require the user to issue it again.
+
+Even if we only allow one transaction per block, it would still be possible to combine several transactions off-chain 
+and to issue a transaction that contains several operations. However, that would require an offline protocol 
+independent from the blockchain and it would require participants to collaborate.
+
+\paragraph{Cost}
+
+The Cardano platform has a notion of cost, and Marlowe contracts will potentially incur costs of two kinds: The cost of a single transaction execution (e.g.\ a single Plutus validator), and the cost of the whole contract, which may consist of many transactions.
+
+Even in Ethereum this distinction is important because single contract execution is limited by gas (and is thus 
+finite), whereas the life of the whole contract may be unbounded and consist of an indefinite number of transactions.  
+So far we are assuming contracts defined in Marlowe to be finite in both aspects. 
+%They could be infinite if we were to 
+%define them co-inductively as recursive members of the DSL type.
+
+\paragraph{Recording information on chain}
+
+Commitments and values of observables need to be recorded somehow, and this raises the potential issue about bloating 
+the chain with data. In general, it will only be necessary to store a signed hash value of the observable, and this will 
+keep data usage bounded.
+
+\paragraph{General computations}
+
+If we have contracts that involve, for example, the oil price then we may need to convert a published price in USD into 
+ADA, using a value for the prevailing USD/ADA exchange rate. This will require a computation: in this case, a 
+multiplication. If Marlowe is to be stand-alone then we need to be extend it with arithmetic and other operations, but 
+we expect Marlowe to be embedded in a suitable language and so to be able to use the facilities of the host language here.
+
+\paragraph{Sidechains and wallets}
+
+The implementation may split contract execution between the main chain and a sidechain (or chains). This will allow links from Marlowe to work on securing multiparty sidechain games/contracts (MPCs).
+
 \section{Reflection}
@@
 
-\subsection{Towards an implementation in Cardano}
-\label{section:towards-implementation}
-
-The contracts implemented in the Marlowe language have an implicit interface defined for what would correspond to the 
-client side of a distributed application, as shown by the interactive interface.
-
-\paragraph{Compilation}
-
-We have designed Marlowe assuming that each contract would be compiled into
-\begin{itemize} 
-\item one or more smart contracts (potentially written in Plutus) and 
-\item a client side program or interface that has a limited number of choices to be made at each point in time 
-(including an interface for the observables). 
-\end{itemize}
-Of course, we probably do not want to give a user a programmatic interface but an actual application with user 
-interface. So there is still some work to be done to create interfaces on top of the client side of the contracts.
-
-Since the current version of Marlowe follows the `pull' model, a contract written in Marlowe could also be potentially
-compiled to a smart-contract written in Ethereum. This could be done in two ways, it could be compiled so that an 
-smart-contract represents a single instance of the Marlowe contract (with a fixed set of participants), or 
-alternatively it would be possible to create a generic smart-contract with an extra ``instantiate'' operation that 
-sets-up a new Marlowe contract instance with a given set of participants (represented by a set of public keys).
-
-\paragraph{Transactions}
-
-In principle, we can directly translate each operation in the model (i.e: cash commitment, payment claim, choice, or
-redemption) into a separate transaction in the blockchain. In Ethereum model this would work in the same 
-way that is simulated by Meadow: several operations can be issued in parallel and the blockchain subsystem will 
-automatically apply them in some order and integrate them into the next block.
-
-Another approach would be to use a UTXO model with continuations in which a UTXO represents a contract. In this model,
-an operation would correspond to a transaction that spends a UTXO and creates (at least) a new UTXO that represents the 
-contract after the operation.
-
-Usually, in the UTXO model, it is necessary that a UTXO is in the blockchain in order for another transaction to be able 
-to spend it. This could potentially limit the number of operations that can be applied in a single block to one, which 
-would make infeasible the approach of using the same UTXO to represent several instances of the same contract (since it 
-would not scale).
-
-On the other hand, it would prevent non-determinism and, with it, it would remove the possibility of race-conditions. A 
-change to the state of a contract before a transaction (and its corresponding operation) is accepted in the blockchain 
-would invalidate the transaction and require the user to issue it again.
-
-Even if we only allow one transaction per block, it would still be possible to combine several transactions off-chain 
-and to issue a transaction that contains several operations. However, that would require an offline protocol 
-independent from the blockchain and it would require participants to collaborate.
-
-\paragraph{Cost}
-
-The Cardano platform has a notion of cost, and Marlowe contracts will potentially incur costs of two kinds:
-\begin{itemize} 
-\item The cost of a single transaction execution (a single Plutus validator).
-\item The cost of the whole contract, which may consist of many transactions.
-\end{itemize} 
-Even in Ethereum this distinction is important because single contract execution is limited by gas (and is thus 
-finite), whereas the life of the whole contract may be unbounded and consist of an indefinite number of transactions.  
-So far we are assuming contracts defined in Marlowe to be finite in both aspects. They could be infinite if we were to 
-define them co-inductively as recursive members of the DSL type.
-
-\paragraph{Recording information on chain}
-
-Commitments and values of observables need to be recorded somehow, and this raises the potential issue about bloating 
-the chain with data. In general, it will only be necessary to store a signed hash value of the observable, and this will 
-keep data usage bounded.
-
-\paragraph{General computations}
-
-If we have contracts that involve, for example, the oil price then we may need to convert a published price in USD into 
-ADA, using a value for the prevailing USD/ADA exchange rate. This will require a computation: in this case, a 
-multiplication. If Marlowe is to be stand-alone then we need to be extend it with arithmetic and other operations, but 
-we expect Marlowe to be embedded in Plutus (Core) and so to be able to use the facilities of the host language here.
-
-\paragraph{Sidechains and wallets}
-
-The implementation may split contract execution between the main chain and a sidechain (or chains). This will allow 
-links from Marlowe to work on securing multiparty sidechain games/contracts (MPCs).
-
 \section{Related work}

### 2018-05-17 tex/paper.tex
abstract; conclusion; removed reflection

@@
 particular field. This paper explores the design of one DSL, Marlowe, targeted at the execution of financial 
-contracts in the style of Peyton Jones \emph{et al} on blockchains. We have focused on predictability, enforceability, 
-simplicity, and ease of analysis and use.
+contracts in the style of Peyton Jones \emph{et al} on blockchains.
+We present an executable semantics of Marlowe in Haskell, an example of Marlowe in practice, and describe the Meadow tool that allows users to interact in-browser with simulations of Marlowe contracts.
 
@@
 Section \ref{section:tool} introduces Meadow, our tool for visualising and interacting with Marlowe contracts.
-Section \ref{section:reflection} reflects on the system design, and observes some alternative designs and Section 
-\ref{section:next-steps} enumerates the next steps for the project.
-
-TO UPDATE \ldots we begin in Section \ref{section:design} with a discussion of the design rationale 
-for Marlowe, including an overview of the various approaches to scripting in blockchain as well as the different designs 
-of the blockchains themselves. We conclude Section \ref{section:design} with a set of decisions about the blockchain 
-environment for Marlowe.
+Section \ref{sec:abstraction} reflects on the system design, showing how Marlowe can be supported on a variety of blockchains, and Section \ref{sec:compilation} explores how Marlowe can be implemented.
+Section \ref{sec:related} surveys related work and Section
+\ref{section:next-steps} enumerates the next steps for the project after drawing some conclusions.
 
@@
 
-The DSL is modelled as an algebraic type in Haskell, together with an executable small-step semantics. 
+The Marlowe domain-specific language (DSL) is modelled as an algebraic type in Haskell, together with an executable small-step semantics. 
 We start by looking at the different types used by the model, and the assumptions about the infrastructure in which contracts will be run. We then we look at the \haskellinline{Contract} DSL itself, 
@@
 \subsection{Abstraction}
+\label{sec:abstraction}
 
 Marlowe abstracts away from a number of concrete details of how blockchains operate.
-In particular, it is agnostic between UTxO-based systems, such as Bitcoin and Cardano SL, and account-based models such as Ethereum.
+In particular, it is agnostic between UTxO-based systems, such as Bitcoin and Cardano SL, and account-based models such as Ethereum; it can be implemented on push or pull-based systems, and can be executed on or off-chain.
 
@@
 
-The simplest blockchain is Bitcoin. Transactions on Bitcoin are made by spending the (as yet) unspent outputs of 
+Transactions on Bitcoin are made by spending the (as yet) unspent outputs of 
 previous transactions (`unspent transaction outputs' or UTxOs). The chain need not maintain any state, such as the 
 current value owned by any particular participant: such \emph{account} information is implicit and external to the 
-Bitcoin blockchain. On the other hand, the Ethereum model explicitly keeps track of account values, and so this 
-information needs to be kept as state information on chain. While it is possible to translate between UTxO-based and 
-account-based chains the UTxO model is more fundamental, and requires less support from the 
-implementation of the chain itself.
+Bitcoin blockchain. On the other hand, the Ethereum model explicitly keeps track of account values, and  this 
+information needs to be kept on chain. Of the two, the UTxO model is simpler and requires less support from the implementation of the chain itself.
 
@@
 In a \emph{push} model, contracts are seen to make things happen.
-In the case of blockchain, this would include making payments or transactions take plade.
+In the case of blockchain, this would include making payments or transactions take place.
 Alternatively, in a \emph{pull} model, contracts enable certain things, such as transactions,  to happen but the 
@@
 Moreover, the Cardano roadmap~\cite{cardano-rationale} envisages the possibility of computation taking place, in part at least, on a sidechain rather than on the main chain itself, which
-may support only the SL. Other off-chain approaches include the Lightning network, overlaid on blockchains (usually Bitcoin) and the channels of \ae{}ternity, introduced below.
+may support only the SL. Other off-chain approaches include the Lightning network, overlaid on blockchains (usually Bitcoin) and the channels of \ae{}ternity~\cite{aeternity}.
 
 \subsection{Compilation}
+\label{sec:compilation}
 
@@
 Of course, we probably do not want to give a user a programmatic interface but an actual application with user interface. 
-So there is still some work to be done to create interfaces on top of the client side of the contracts.
+%So there is still some work to be done to create interfaces on top of the client side of the contracts.
 
 A contract written in Marlowe could also be potentially
-compiled to a smart-contract written in Ethereum. This could be done in two ways, it could be compiled so that an 
-smart-contract represents a single instance of the Marlowe contract (with a fixed set of participants), or 
-alternatively it would be possible to create a generic smart-contract with an extra ``instantiate'' operation that 
-sets-up a new Marlowe contract instance with a given set of participants (represented by a set of public keys).
+compiled to a smart-contract written in Ethereum. This could be done in two ways: it could be compiled so that a 
+smart contract represents a single instance of the Marlowe contract with a fixed set of participants; 
+alternatively it would be possible to create a generic contract with an extra ``instantiate'' operation that 
+sets up a new Marlowe contract instance with a given set of participants (set of public keys).
 
@@
 multiplication. If Marlowe is to be stand-alone then we need to be extend it with arithmetic and other operations, but 
-we expect Marlowe to be embedded in a suitable language and so to be able to use the facilities of the host language here.
-
-\paragraph{Sidechains and wallets}
-
-The implementation may split contract execution between the main chain and a sidechain (or chains). This will allow links from Marlowe to work on securing multiparty sidechain games/contracts (MPCs).
+we expect Marlowe to be embedded in a suitable language that provides these facilities.
+
+%\paragraph{Sidechains and wallets}
+%
+%The implementation may split contract execution between the main chain and a sidechain (or chains). This will allow links from Marlowe to work on securing multiparty sidechain games/contracts (MPCs).
+
+\ignore{ % begin ignore
 
@@
 
+} % end ignore
+
 \section{Related work}
+\label{sec:related}
 
@@
 
-\section{Next steps}
+\section{Conclusions and Future Work}
 \label{section:next-steps}
 
-We plan to continue the work with Marlowe in a number of directions. First, we need to sign off a definition of the 
-language, as the current status of Marlowe 1.3 is that of a proposal.
-
-\paragraph{Context}
-
-We should decide on the final form of the language -- for example, should it be monadic or not -- and the relation of 
-Marlowe to Plutus (Core). In particular, answering the latter question will require work on the current 
-redeemer/validator model of transactions in Plutus Core.
-
-\paragraph{Implement in Cardano}
-
-An immediate goal is to implement Marlowe in Cardano, noting the observations of Section 
-\ref{section:towards-implementation} and to deploy it in a test network to observe and measure its behaviour in 
-practice.
-
-\paragraph{Compilation}
-
-Marlowe contracts could be generated by a compilation process from `vanilla' financial contracts. This also suggests 
-that models with off-chain enforcement could be investigated too, with an MPC foundation.
-
-\paragraph{Validation and Verification}
-
-Some degree of validation is provided by Meadow presented in Section \ref{section:tool}, albeit for smaller-scale 
-contracts, but there is a general question of how properties of contracts can be inferred. One relatively low-cost is to 
-use QuickCheck-style property-based testing~\cite{quickCheck}, and properties developed here become candidates for 
-fully fledged verification in a formalisation of the (semantics of) Marlowe.
-
-\paragraph{Analysis}
-
-As we noted in Section \ref{section:analysis} we can develop analyses of various aspects of Marlowe contracts. While 
-these might initially be ``pencil and paper'' analyses, we could expect these to be based on a formalisation of the 
-language and its semantics.
-
-\paragraph{Formalisation}
-
-Marlowe and its semantics could be formalised in a theorem prover such as Isabelle~\cite{IsabelleHOL}, which has been 
-used in the formalisation of language semantics. A formalisation could provide a sanity check on the language 
-definition, as well as formalise proofs of important properties, based on analysis of the contract language.
-
-\section{Acknowledgements}
-
-Thanks to Manuel Chakravarty, Duncan Coutts, Bernardo David, Charles Hoskinson, Aggelos Kiayias, % Darryl McAdams, 
+In this paper we have presented Marlowe, a DSL for financial contracts on blockchains, based on earlier work on contracts~\cite{PeytonJones:2000}, together with examples of its use.
+We have seen that to make this operational on blockchain we need to add commitments and timeouts, and to design a semantics that reflects these.
+As we saw in Section~\ref{sec:abstraction}, Marlowe has been designed to make as few demands as possible on the underlying blockchain: it can be implemented on UTxO or account-based blockchains, for example.
+
+We have also presented Meadow, that allows users to interact with and simulate the operation of Marlowe contracts, contributing to the potential adoptability of the system.
+We have also described the design rationale for Marlowe, and sketched ways in which it can be implemented.
+
+We plan to continue the work with Marlowe in a number of directions.
+We will continue to develop the core language, for example considering the automation of generation of identifiers for commitments and others.
+We will then implement this version of Marlowe in Cardano, compiling from Marlowe contracts to on chain contracts and users' wallets, and deploy it in a test network to observe and measure its behaviour in practice.
+
+Building on the operational semantics given here, we will develop analyses of Marlowe contracts -- such as to show that contracts cannot generate \haskellinline{FailedPay} actions in certain circumstances.
+We will also develop  QuickCheck-style property-based testing~\cite{quickCheck}, and properties developed here can become candidates for fully fledged verification in a formalisation of the semantics of Marlowe.
+
+We would like to thank Manuel Chakravarty, Duncan Coutts, Bernardo David, Charles Hoskinson, Aggelos Kiayias, % Darryl McAdams, 
 Bruno Woltzenlogel Paleo, Rebecca Valentine and Phil Wadler for their collaboration, and to colleagues from the Universities of 
-Kent and Leicester for their comments on previous presentations of this work.
+Kent and Leicester for their comments on this work.
 

### 2018-05-18 tex/paper.tex
Fix various details and typos derived from review

@@
 As the examples illustrate, observables come both from aspects of the blockchain (e.g.\ the current block number) and 
-externally. In the latter case it will be necessary to agree a trusted oracle or beacon giving the value.
+externally. In the latter case, it will be necessary to agree a trusted oracle or beacon giving the value.
 
@@
 participants themsleves. These include commitments of currency (or ``cash''), redemption of commitments, and claims of payments by 
-a participant. <oreover it is also possible for a participant to input an arbitrary value (which we term a ``choice''). 
+a participant. Moreover it is also possible for a participant to input an arbitrary value (which we term a ``choice''). 
 The particular inputs at a given step are described by a value of type \haskellinline{Input}.
@@
 
-While informally we might see a commitment to something as being indefinite, it is important to realise that on 
-blockchain a commitment needs to have a timeout so that progress can be forced in a contract. After the timeout period the 
+While informally we might see a commitment to something as being indefinite, it is important to realise that, on
+blockchain, a commitment needs to have a timeout so that progress can be forced in a contract. After the timeout period the 
 cash can be refunded through the user creating a transaction to reclaim the cash.
@@
 they need not be modelled explicitly.
-\item We assume that time is ``coarse grained'' and measured by block number, so that in particular timeouts are 
+\item We assume that time is ``coarse grained'' and measured by block number, so that, in particular, timeouts are 
 delimited using block numbers. 
@@
 In addition to 
-calling \haskellinline{stepAll}, \haskellinline{stepBlock} will first enable expired cash commitments to be refunded and record any choices made at that step. 
+calling \haskellinline{stepAll}, \haskellinline{stepBlock} will first enable expired cash commitments to be refunded and record, in the state, any choices made at that step. 
 The functions \haskellinline{stepAll} and \haskellinline{stepBlock} have the same type as \haskellinline{step} itself.
@@
 
-The type of contracts is given by the Haskell data type
+The type of contracts is given by the following Haskell data type:
 
@@
 \haskellinline{CommitCash} will wait for a participant to make a commitment,
-\haskellinline{RedeemCC} allows for a commitment to be redeemed, and \haskellinline{Pay} for a payment to be made to another participant.
+\haskellinline{RedeemCC} allows for a commitment to be redeemed, and \haskellinline{Pay} for a payment between participants to be claimed by the recipient.
 
@@
 
-What many of the contracts also have are timeouts that also determine their behaviour. 
+Additionally, many of the contracts have timeouts that also determine their behaviour. 
 \subsection{The \haskellinline{step} function}
 
-In this section we explain the detailed behaviour of \haskellinline{Contracts} through a discussion of the \haskellinline{step} function.
+In this section, we explain the detailed behaviour of \haskellinline{Contracts} through a discussion of the \haskellinline{step} function.
 
@@
 progress,\begin{itemize}
-\item either before the timeout \haskellinline{start_timeout} the user \haskellinline{person} makes a cash commitment of 
+\item either, before the timeout \haskellinline{start_timeout}, the user \haskellinline{person} makes a cash commitment of 
 value 
@@
 further spend the committed cash, and any unspent funds can be reclaimed by participant \haskellinline{person}.
- This is performed by the \haskellinline{stepBlock} function, as noted above.
+ This is enforced by the \haskellinline{stepBlock} function, as noted above.
 
@@
 \haskellinline{RedeemCC ident con} (\haskellinline{CC} stands for cash commitment.) For this contract to make 
-progress, the creator of the cash commitment with identifier \haskellinline{ident} is required to redeem the unspent 
+progress, the creator of the cash commitment with identifier \haskellinline{ident} is allowed to redeem the unspent 
 funds in that commitment; the contract then continues as \haskellinline{con}, and the action 
@@
 By `available' we mean that sufficient commitments have been made and not yet expired to cover the payment; in this 
-case the payment uses the currency allocated by the cash commitments made by \haskellinline{from} that expire the 
+case, the payment uses the currency allocated by the cash commitments made by \haskellinline{from} that expire the 
 earliest.
@@
 \end{haskellcode}
-Then we can easily write a function that returns an observation that is true if and only if at least one of the 
-participants \haskellinline{per} and \haskellinline{per'} has chosen the number \haskellinline{c} as follows:
+Then, we can easily write a function that returns an observation that is \haskellinline{True} if and only if at least one of the 
+participants \haskellinline{per} and \haskellinline{per'} has chosen the number \haskellinline{val} as follows:
 \begin{haskellcode}
@@
 \end{haskellcode}
-Building on that, we can now write a function that returns an observation that is true if and only if at least 
+Building on that, we can now write a function that returns an observation that is \haskellinline{True} if and only if at least 
 two out of the three participants \haskellinline{p1}, \haskellinline{p2}, and \haskellinline{p3} have agreed in 
@@
 
-The outermost primitive \haskellinline{CommitCash} allows the participant \haskellinline{1} to commit 450 ADA before 
-block 10, with the promise that money will be released on block 100 if they are not claimed before that.
+The outermost primitive \haskellinline{CommitCash} allows the participant \haskellinline{1} to commit \haskellinline{450} ADA before 
+block \haskellinline{10}, with the promise that money will be released on block \haskellinline{100} if they are not claimed before that.
 
@@
 \begin{enumerate}
- \item The observation became true because two out of three have chosen $0$. 
- \item The observation became true because two out of three have chosen $1$.
- \item The observation remained false until 90 was published in the blockchain.
+	\item The observation became \haskellinline{True} because two out of three have chosen \haskellinline{0}. 
+	\item The observation became \haskellinline{True} because two out of three have chosen \haskellinline{1}.
+	\item The observation remained \haskellinline{False} until \haskellinline{90} was published in the blockchain.
 \end{enumerate}
@@
 If the third option happens, the money is refunded. Otherwise, there is a \haskellinline{Choice} that will 
-immediately refund the money to participant \haskellinline{1} unless two out of three chose option $1$; in the later 
+immediately refund the money to participant \haskellinline{1} unless two out of three chose option \haskellinline{1}; in the later 
 case, \haskellinline{Pay} will give participant \haskellinline{2} the opportunity to claim the funds available in 
-the commitment with identifier \haskellinline{iCC1} before block 100 (by using the identifier \haskellinline{iP1} for 
-the claim). If the funds are not claimed before block 100 they will also be refunded to 
+the commitment with identifier \haskellinline{iCC1} before block \haskellinline{100} (by using the identifier \haskellinline{iP1} for 
+the claim). If the funds are not claimed before block \haskellinline{100} they will also be refunded to 
 participant \haskellinline{1}.
@@
 and formatted code (that is displayed in the upper right corner of the application),
-and to convert the code back to its block representation.
+and to convert the code back to its blockly representation.
 
@@
 
+Nevertheless, there already exists in Bitcoin a mechanism that allows the combination of several chained transactions within a single block while using the UTxO model, even though it is used to solve a different problem and it is not widely spread yet.
+This is the case of "child pays for parent", which allows the recipient of an unpublished transaction to issue another transaction that spends the former, with the aim of including enough fees for both transactions to have enough incentive to be included by a miner in a block (in which case both transactions would be included in the same block).
+
+The same mechanism could be used for allowing miners to aggregate several transactions that act as a chain in the same block.
+The clients would only need to monitor the transaction pool in order to send transactions that can be chained with the existing ones.
+
 \paragraph{Cost}
@@
 
-Commitments and values of observables need to be recorded somehow, and this raises the potential issue about bloating 
+Inputs and values of observables need to be recorded somehow, and this raises the potential issue about bloating 
 the chain with data. In general, it will only be necessary to store a signed hash value of the observable, and this will 
-keep data usage bounded.
+keep data usage bounded; but the full information may still need to be posted in case of dispute.
 
@@
 seminal~\cite{PeytonJones:2000}, and the authors note that payments need to be bounded; this is made concrete in our 
-account by our notion of commitments. They take no account of commitments or timeouts as we do below, but it should be 
+account by our notion of commitments. They take no account of commitments or timeouts as our approach does, but it should be 
 noted that the Ethereum platform is more powerful than the one that we target in this paper.
@@
 
-In this paper we have presented Marlowe, a DSL for financial contracts on blockchains, based on earlier work on contracts~\cite{PeytonJones:2000}, together with examples of its use.
+In this paper, we have presented Marlowe, a DSL for financial contracts on blockchains, based on earlier work on contracts~\cite{PeytonJones:2000}, together with examples of its use.
 We have seen that to make this operational on blockchain we need to add commitments and timeouts, and to design a semantics that reflects these.

### 2018-05-18 tex/paper.tex
Made modes in Section 3 and 7 as proposed by PLS; removed 'other approaches' section from section 6

@@
 \section{Introduction}
+\label{sec:intro}
 
@@
 Marlowe contracts that are easier to write, read, and understand, by embedding them into Haskell code, taking advantage 
-of the fact that Marlowe contracts are implemented as Haskell terms. We used Haskell because it is the language in 
-which Marlowe is implemented, but it would be easy to embed Marlowe in any other languages. It would only be necessary 
-to translate its primitives into a data type in that language. In Meadow, we use Fay embedding, which is a subset of 
+of the fact that Marlowe contracts are implemented as Haskell terms.
+
+We used Haskell because it is the language in 
+which Marlowe is implemented, but it would be easy to embed Marlowe in any other language. It would only be necessary 
+to translate its primitives into a data type in that language. In Meadow we use Fay, which is a subset of 
 Haskell (see Section~\ref{section:tool}).
 
-The example contract we use through this section implements an escrow contract. Let us imagine that participant 
-\haskellinline{1} is buying some item from participant \haskellinline{2}. On one hand, if participant \haskellinline{1} 
-pays before receiving the item, there is a risk that participant \haskellinline{2} will never give \haskellinline{1} the 
-item. On the other hand, if participant \haskellinline{2} gives participant \haskellinline{1} the item before 
-\haskellinline{1} pays,  there is a risk that \haskellinline{1} will never pay.
-
+The example we use through this section implements an escrow contract, as first introduced in Section~\ref{sec:intro}. 
+%
+%Let us imagine that participant 
+%\haskellinline{1} is buying some item from participant \haskellinline{2}. On one hand, if participant \haskellinline{1} 
+%pays before receiving the item, there is a risk that participant \haskellinline{2} will never give \haskellinline{1} the 
+%item. On the other hand, if participant \haskellinline{2} gives participant \haskellinline{1} the item before 
+%\haskellinline{1} pays,  there is a risk that \haskellinline{1} will never pay.
+%
 The escrow mechanism allows participant \haskellinline{1} to deposit the money into a contract, in a way that the money 
 will only be released when two out of three participants agree on whether participant \haskellinline{2} has indeed given 
-participant \haskellinline{1} the item. The third participant (participant \haskellinline{3}) is supposed to be a 
+participant \haskellinline{1} the item.
+
+The third participant (participant \haskellinline{3}) is supposed to be a 
 neutral third party that will decide in case of dispute.
-
 This way, if participants \haskellinline{1} and \haskellinline{2} are honest, they will just agree on the result of the 
@@
 
-In addition to this, in our implementation, the money paid for the item is 450 ADA, it must be committed by participant 
-\haskellinline{1} before block \haskellinline{10}, and it will be refunded to participant \haskellinline{1} if there is 
+In our implementation we make things more specific: the money paid for the item is 450 ADA, and it must be committed by participant 
+\haskellinline{1} before block \haskellinline{10}; it will be refunded to participant \haskellinline{1} if there is 
 no consensus before block \haskellinline{90}.
@@
 Reflecting on the example, we can see that using Haskell definitions has made the contract substantially more 
-comprehensible; it will also support more efficient operation, if we were to replace the repeated  expression 
-\haskellinline{two_chose 1 2 3 1} above by a \haskellinline{where} clause, 
+comprehensible.
+While our current implementation does not do this, it is possible to modify the embedding to support more efficient operation, by preserving \emph{sharing} in the host language.
+For example, if we were to replace the repeated  expression
+\haskellinline{two_chose 1 2 3 1} by a \haskellinline{where} clause, 
 \begin{haskellcode}
@@
 \end{haskellcode}
-then the repeated computation of the expression will be avoided.
-In the next section, we cover the online contract visualisation and interaction tool for Marlowe.
+then the repeated computation of the expression could be avoided.
+%In the next section, we cover the online contract visualisation and interaction tool for Marlowe.
 
@@
 Varna and the HLM for simple contracts, and a (port of) Solidity and the EVM for compatibility with Ethereum. In a 
-similar way, Cardano provides support for IELE~\cite{IELE}, a rational reconstruction of the EVM, and thus for Solidity 
-contracts too.
+similar way, Cardano provides support for IELE~\cite{IELE}, a rational reconstruction of the EVM, and thus for Solidity too.
 
@@
 
-\paragraph{Other approaches}
-
-There are a variety of other approaches to scripting, including business process notations, such as the OMG's BPMN; 
-process calculi such as Rholang; logic rules (in Formal Contract Logic), and finite state machines. More details of 
-these are provided in~\cite{cryptoeprint:2016:1156}.
+%\paragraph{Other approaches}
+%
+%There are a variety of other approaches to scripting, including business process notations, such as the OMG's BPMN; 
+%process calculi such as Rholang; logic rules (in Formal Contract Logic), and finite state machines. More details of 
+%these are provided in~\cite{cryptoeprint:2016:1156}.
 
@@
 
-We would like to thank Manuel Chakravarty, Duncan Coutts, Bernardo David, Charles Hoskinson, Aggelos Kiayias, % Darryl McAdams, 
-Bruno Woltzenlogel Paleo, Rebecca Valentine and Phil Wadler for their collaboration, and to colleagues from the Universities of 
-Kent and Leicester for their comments on this work.
+We would like to thank IOHK for supporting us: not only has this led us to work on Marlowe, but we have benefitted hugely from collaboration with colleagues including Manuel Chakravarty, Duncan Coutts, Bernardo David, Charles Hoskinson, Aggelos Kiayias, % Darryl McAdams,
+Bruno Woltzenlogel Paleo, Rebecca Valentine and Phil Wadler.
+We are  very grateful to Thomas Arts and colleagues from \ae{}ternity for their convivial discussions on blockchain and contracts in G\"oteborg, and to colleagues from the Universities of Kent and Leicester for their comments too.
 

### 2018-05-19 tex/paper.tex
Move acknowledgement and fix layout

--- 2018-05-18 tex/paper.tex
@@
 % Title portion. Note the short title for running heads
-\title {Marlowe: financial contracts on blockchain}
+\title {Marlowe: financial contracts on blockchain\thanks{This work is part of the Cardano project 
+and is supported by IOHK,~\url{https://iohk.io}}}
 %
@@
 from \url{https://github.com/input-output-hk/scdsl}} targetted at the execution of financial contracts in the style of 
-Peyton Jones, Eber and Seward~\cite{PeytonJones:2000} on blockchains. This work is a part of the Cardano project, and is 
-supported by IOHK,~\url{https://iohk.io}.
+Peyton Jones, Eber and Seward~\cite{PeytonJones:2000} on blockchains.
 
@@
 
-\smallskip
+\begin{itemize}
+\item \haskellinline{Null} is the null contract; it will always be quiescent:
+
+\begin{haskellcode}
+step _ st Null _ = (st, Null, [])
+\end{haskellcode}
+
+\medskip
 \noindent
-\haskellinline{Null} is the null contract; it will always be quiescent:
-
-\begin{haskellcode}
-step _ st Null _ = (st, Null, [])
-\end{haskellcode}
-
-\smallskip
-\noindent
-\haskellinline{CommitCash ident person val start_timeout end_timeout con1 con2}  For this contract to make 
+\item \haskellinline{CommitCash ident person val start_timeout end_timeout con1 con2}  For this contract to make 
 progress,\begin{itemize}
@@
 \noindent
-\haskellinline{RedeemCC ident con} (\haskellinline{CC} stands for cash commitment.) For this contract to make 
+\item \haskellinline{RedeemCC ident con} (\haskellinline{CC} stands for cash commitment.) For this contract to make 
 progress, the creator of the cash commitment with identifier \haskellinline{ident} is allowed to redeem the unspent 
@@
 \noindent
-\haskellinline{Pay idpay from to val expi con} makes it possible, assuming that sufficient funds are available, for 
-\haskellinline{to} to claim a payment with id \haskellinline{idpay} of \haskellinline{val} from \haskellinline{from} 
-before the timeout \haskellinline{expi}. The contract continues as \haskellinline{con}.
+\item \haskellinline{Pay idpay from to val expi con} makes it possible, assuming that sufficient funds are available, 
+for \haskellinline{to} to claim a payment with id \haskellinline{idpay} of \haskellinline{val} from 
+\haskellinline{from} before the timeout \haskellinline{expi}. The contract continues as \haskellinline{con}.
 
@@
 \noindent
-\haskellinline{Both con1 con2} enforces the behaviour of both contracts \haskellinline{con1} and \haskellinline{con2}. 
+\item \haskellinline{Both con1 con2} enforces the behaviour of both contracts \haskellinline{con1} and 
+\haskellinline{con2}. 
 \begin{haskellcode}
@@
 \noindent
-\haskellinline{Choice obs conT conF} behaves as either \haskellinline{conT} or \haskellinline{conF} depending on the 
+\item \haskellinline{Choice obs conT conF} behaves as either \haskellinline{conT} or \haskellinline{conF} depending on 
+the 
 (Boolean) result of \haskellinline{obs} at the time that the observation is made, \haskellinline{conT} if it is 
@@
 \noindent
-\haskellinline{When obs expi con con2} This contract will not progress until \haskellinline{obs} is 
+\item \haskellinline{When obs expi con con2} This contract will not progress until \haskellinline{obs} is 
 \haskellinline{True} or until the current block number is greater than or equal to the one specified by timeout 
@@
   \end{haskellcode}
+\end{itemize}
 We look next at an example of Marlowe in action.

### 2018-05-19 tex/paper.tex
minor edits

@@
 Peyton Jones, Eber and Seward~\cite{PeytonJones:2000} on blockchains.
-
 In doing this, we are required to refine the model of contracts in a number of ways in order to fit with a radically 
-different context. Consider the following example of an ``escrow'' contract so that we can explain the motivation 
+different context.
+
+Consider the following example of an ``escrow'' contract so that we can explain the motivation 
 more concretely. The aim of this contract, written in functional pseudocode in the style of~\cite{PeytonJones:2000} 
@@
 
-In this section, we explain the detailed behaviour of \haskellinline{Contracts} through a discussion of the \haskellinline{step} function.
+In this section, we explain the detailed behaviour of contracts by describing how the \haskellinline{step} function operates on each of the constructors of the \haskellinline{Contract}  type.
 

### 2018-07-18 tex/paper.tex
Fix typos and address some reviewer's comments

@@
 particular field. This paper explores the design of one DSL, Marlowe, targeted at the execution of financial 
-contracts in the style of Peyton Jones \emph{et al} on blockchains.
-We present an executable semantics of Marlowe in Haskell, an example of Marlowe in practice, and describe the Meadow tool that allows users to interact in-browser with simulations of Marlowe contracts.
+contracts in the style of Peyton Jones \emph{et al} on blockchains. We present an executable semantics of Marlowe in 
+Haskell, an example of Marlowe in practice, and describe the Meadow tool that allows users to interact in-browser with 
+simulations of Marlowe contracts.
 
@@
 This paper explores the design of a domain specific language, Marlowe,\footnote{Named after Christopher 
-Marlowe, the Elizabethan poet, dramatist and spy, who was born and educated in Canterbury,  \url{en.wikipedia.org/wiki/Christopher_Marlowe}}\footnote{Marlowe is available 
-from \url{https://github.com/input-output-hk/scdsl}} targetted at the execution of financial contracts in the style of 
-Peyton Jones, Eber and Seward~\cite{PeytonJones:2000} on blockchains.
-In doing this, we are required to refine the model of contracts in a number of ways in order to fit with a radically 
-different context.
+Marlowe, the Elizabethan poet, dramatist and spy, who was born and educated in Canterbury,  
+\url{en.wikipedia.org/wiki/Christopher_Marlowe}}\footnote{Marlowe is available from 
+\url{https://github.com/input-output-hk/scdsl}} targeted at the execution of financial contracts in the style of 
+Peyton Jones, Eber and Seward~\cite{PeytonJones:2000} on blockchains. In doing this, we are required to refine the 
+model of contracts in a number of ways in order to fit with a radically different context.
 
@@
 to pay an amount of money to participant \haskellinline{2} on receipt of goods from her. Participant \haskellinline{1} 
-pays the money into escrow controlled by participant \haskellinline{3}. There are two options for the money: if two 
-participants agree to \haskellinline{pay} it to \haskellinline{2}, that goes ahead; if, on the other hand, two of the 
-participants opt to \haskellinline{refund} the money to \haskellinline{1}, that is done instead. The 
-\haskellinline{When} construct fires when its first argument -- a condition -- becomes true; the second argument, a 
-\haskellinline{Contract} itself, is then performed, and this makes the payment according to the participants' 
-\haskellinline{Choice}.
+pays the money into escrow controlled by participant \haskellinline{3}.
+
+There are two options for the money: if two participants agree to \haskellinline{pay} it to \haskellinline{2}, that goes 
+ahead; if, on the other hand, two of the participants opt to \haskellinline{refund} the money to \haskellinline{1}, that 
+is done instead.
+
+The outer primitive \haskellinline{When} waits until the condition -- its first argument -- 
+becomes true; in this case, the condition is the that either two participants choose \haskellinline{refund} or two 
+participants choose \haskellinline{pay}. The second argument of the \haskellinline{When} is itself another 
+\haskellinline{Contract}, which is performed after the condition of the \haskellinline{When} has been met, and it makes 
+the payment if two participants chose \haskellinline{pay}, otherwise it redeems previous money commitments.
 
@@
 
-In the traditional model, enforcement of the contract is the responsibility of the legal system. If participant 
-\haskellinline{1} does not pay the money into escrow, or participant \haskellinline{3} chooses to keep it for herself, 
-then they can be sued for the money (and probably damages), thus providing both legal and 
-financial incentives for compliance. On the other hand, in the decentralised blockchain model, where there is no central 
-authority, the contract needs to be enforced \emph{by design}.
-
-This means that we require participants to \emph{commit} money to cover all possible expenditure \emph{in advance of the 
-contract executing}. In order to make sure that participants continue to engage with a contract, we ensure urgency by 
-imposing \emph{timeouts}: money is committed for a finite period only. We also impose a timeout when  waiting for a participant to make a 
-commitment to ensure that the contract does not become stuck even if one of the 
-participants stops interacting with it. We discuss this particular example in more detail in Marlowe in Section~\ref{section:example-escrow} below.
+We discuss this particular example in more detail in Marlowe in Section~\ref{section:example-escrow} below; but it 
+already gives us an example of how traditional contracts are fundamentally different from contracts that are meant to 
+be run on top of the blockchain. In the traditional model, enforcement of the contract is the responsibility of the 
+legal system. If participant \haskellinline{1} does not pay the money into escrow, or participant \haskellinline{3} 
+chooses to keep it for herself, then they can be sued for the money (and probably damages), thus providing both legal 
+and financial incentives for compliance. On the other hand, in the decentralised blockchain model, where there is no 
+central authority, the contract needs to be enforced \emph{by design}.
+
+This means that we must require participants to \emph{commit} money to cover all possible expenditure \emph{in advance 
+of the contract executing}. In order to make sure that participants continue to engage with a contract, we ensure 
+urgency by imposing \emph{timeouts}: money is committed for a finite period only. We also impose a timeout when waiting 
+for a participant to make a commitment to ensure that the contract does not become stuck even if one of the participants 
+stops interacting with it.
 
@@
 \item 
-Embedding Marlowe as a Haskell \haskellinline{data} type to extend its expressibility.
+Embedding of Marlowe in Haskell to extend its expressibility; that is: Marlowe is defined as a Haskell 
+\haskellinline{data} type so that generating Marlowe contracts from Haskell is straightforward.
 \item
-Developing the Meadow tool that allows users to interact with and simulate the operation of embedded Marlowe contracts. 
+Developing the Meadow tool that allows users to interact with and simulate the operation of Marlowe contracts 
+and embedded Marlowe contracts.
 \end{itemize}
-Having established this model and its semantics we are able to do a number of other things, which we discuss in the paper.
-We can explore how the language will be implemented on an existing blockchain, such as Cardano, and our model has shown us that we will need to consider how the evolution of a contract interacts with the blockchain, with miners in the case of a `proof of work'-based chain, and with users participating in contract execution.
-We can also perform analyses of Marlowe contracts, based on its formal semantics.
-Finally, we can explore the design space for Marlowe, and explore potential alternative designs.
+
+Having established this model and its semantics we are able to do a number of other things, which we discuss in the 
+paper. We can explore how the language will be implemented on an existing blockchain, such as Cardano, and our model has 
+shown us that we will need to consider how the evolution of a contract interacts with the blockchain, with miners in the 
+case of a `proof of work'-based chain, and with users participating in contract execution. We can also perform analyses 
+of Marlowe contracts, based on its formal semantics. Finally, we can explore the design space for Marlowe, and explore 
+potential alternative designs.
 
@@
 The paper begins in Section \ref{section:model} by introducing the Marlowe 
-model, including the assumptions made in designing it, the types of the principal functions and a description of the DSL as an algebraic type, constructor by constructor. 
+model, including the assumptions made in designing it, the types of the principal functions and a description of the 
+DSL as an algebraic type, constructor by constructor. 
 %Section \ref{section:example-saving} presents an example of a larger Marlowe contract, for incentivised saving. 
@@
 Section \ref{section:tool} introduces Meadow, our tool for visualising and interacting with Marlowe contracts.
-Section \ref{sec:abstraction} reflects on the system design, showing how Marlowe can be supported on a variety of blockchains, and Section \ref{sec:compilation} explores how Marlowe can be implemented.
-Section \ref{sec:related} surveys related work and Section
+Section \ref{sec:abstraction} reflects on the system design, showing how Marlowe can be supported on a variety of 
+blockchains, and Section \ref{sec:compilation} explores how Marlowe can be implemented. Section \ref{sec:related} 
+surveys related work and Section 
 \ref{section:next-steps} enumerates the next steps for the project after drawing some conclusions.
@@
 
-The Marlowe domain-specific language (DSL) is modelled as an algebraic type in Haskell, together with an executable small-step semantics. 
-We start by looking at the different types used by the model, and the assumptions about the infrastructure in which contracts will be run. We then we look at the \haskellinline{Contract} DSL itself, 
+The Marlowe domain-specific language (DSL) is modelled as an algebraic type in Haskell, together with an executable 
+small-step semantics. We start by looking at the different types used by the model, and the assumptions about the 
+infrastructure in which contracts will be run. We then we look at the \haskellinline{Contract} DSL itself, 
 and finally we give its semantics in Haskell. Section 
@@
 
-Each instance of such an observable will be observed at a particular time and in a particular context.
-We assume that the system infrastructure ensures that these values are recorded on the blockchain to
-allow the computation to be repeated for verification purposes.
-
-It is assumed that at each step of the execution of the 
-contract, the values of observables will be available if needed, and these values are (together) given by a value of type 
-\haskellinline{OS} (for ``observable set''), where individual observations are described in a 
-``little language'' for that purpose: \haskellinline{Observation}. 
-Note that these values are not determined by the participants in the contract, but rather by the 
-external environment in which the contract is run.
+Each instance of such an observable will be observed at a particular time and in a particular context. We assume that 
+the system infrastructure ensures that these values are recorded on the blockchain to allow the computation to be 
+repeated for verification purposes.
+
+It is assumed that at each step of the execution of the contract, the values of observables will be available if 
+needed, and these values are (together) given by a value of type \haskellinline{OS} (for ``observable set''), where 
+individual observations are described in a ``little language'' for that purpose: \haskellinline{Observation}. 
+Note that these values are not determined by the participants in the contract, but rather by the external environment in 
+which the contract is run.
 
@@
 On the other hand, at each step there are -- potentially, at least -- a variety of inputs available from the 
-participants themsleves. These include commitments of currency (or ``cash''), redemption of commitments, and claims of payments by 
-a participant. Moreover it is also possible for a participant to input an arbitrary value (which we term a ``choice''). 
-The particular inputs at a given step are described by a value of type \haskellinline{Input}.
+participants themselves. These include commitments of currency (or ``cash''), redemption of commitments, and claims of 
+payments by a participant. Moreover it is also possible for a participant to input an arbitrary value (which we term a 
+``choice''). The particular inputs at a given step are described by a value of type \haskellinline{Input}.
 
@@
 
-While informally we might see a commitment to something as being indefinite, it is important to realise that, on
-blockchain, a commitment needs to have a timeout so that progress can be forced in a contract. After the timeout period the 
-cash can be refunded through the user creating a transaction to reclaim the cash.
-Information about the commitments currently in force forms the \haskellinline{State}, which can be  modified 
-at each execution step.
+While informally we might see a commitment to something as being indefinite, it is important to realise that, 
+on blockchain, a commitment needs to have a timeout so that progress can be forced in a contract. After the timeout 
+period the cash can be refunded through the user creating a transaction to reclaim the cash. Information about the 
+commitments currently in force forms the \haskellinline{State}, which can be  modified at each execution step.
 
@@
 The \haskellinline{step} function is total, so that for every contract a result of stepping is defined. However, for 
-some kinds of contracts -- commits, redeems or time-shifted contracts -- it is possible that performing  a step produces the same contract as the result; we call these \emph{quiescent} steps whereas all others \emph{make
+some kinds of contracts -- commits, redeems or time-shifted contracts -- it is possible that performing  a step 
+produces the same contract as the result; we call these \emph{quiescent} steps whereas all others \emph{make 
 progress}. We use this distinction in the explanation that follows.
 
-Execution of a contract will involve multiple blocks, with multiple steps in each block.
-The computation at a single block is given by the \haskellinline{stepBlock} function: this will call the \haskellinline{stepAll} function that calls \haskellinline{step} repeatedly
-until it is quiescent.
-
-In addition to 
-calling \haskellinline{stepAll}, \haskellinline{stepBlock} will first enable expired cash commitments to be refunded and record, in the state, any choices made at that step. 
-The functions \haskellinline{stepAll} and \haskellinline{stepBlock} have the same type as \haskellinline{step} itself.
+Execution of a contract will involve multiple blocks, with multiple steps in each block. The computation at a single 
+block is given by the \haskellinline{stepBlock} function: this will call the \haskellinline{stepAll} function that calls 
+\haskellinline{step} repeatedly until it is quiescent.
+
+In addition to calling \haskellinline{stepAll}, \haskellinline{stepBlock} will first enable expired cash commitments to 
+be refunded and record, in the state, any choices made at that step. The functions \haskellinline{stepAll} and 
+\haskellinline{stepBlock} have the same type as \haskellinline{step} itself.
 
@@
    \end{haskellcode}
-Informally, this type provides a \haskellinline{Null} contract, which does nothing.
-The next three constructs form contracts that do something, and then continue according to another contract (which is one of the components of the original contract).
-\haskellinline{CommitCash} will wait for a participant to make a commitment,
-\haskellinline{RedeemCC} allows for a commitment to be redeemed, and \haskellinline{Pay} for a payment between participants to be claimed by the recipient.
-
-The remaining constructors form composite contracts from simpler components: \haskellinline{Both} has the behaviour of both its components, \haskellinline{Choice} chooses between two contracts on the basis of an observation, and \haskellinline{When} is quiescent until a condition -- i.e.\ an \haskellinline{Observation} -- becomes true.
+Informally, this type provides a \haskellinline{Null} contract, which does nothing. The next three constructs form 
+contracts that do something, and then continue according to another contract (which is one of the components of the 
+original contract). \haskellinline{CommitCash} will wait for a participant to make a commitment, 
+\haskellinline{RedeemCC} allows for a commitment to be redeemed, and \haskellinline{Pay} for a payment between 
+participants to be claimed by the recipient.
+
+The remaining constructors form composite contracts from simpler components: \haskellinline{Both} has the behaviour of 
+both its components, \haskellinline{Choice} chooses between two contracts on the basis of an observation, and 
+\haskellinline{When} is quiescent until a condition -- i.e.\ an \haskellinline{Observation} -- becomes true.
 
@@
 
-In this section, we explain the detailed behaviour of contracts by describing how the \haskellinline{step} function operates on each of the constructors of the \haskellinline{Contract}  type.
+In this section, we explain the detailed behaviour of contracts by describing how the \haskellinline{step} function 
+operates on each of the constructors of the \haskellinline{Contract}  type.
 
@@
 progress,\begin{itemize}
-\item either, before the timeout \haskellinline{start_timeout}, the user \haskellinline{person} makes a cash commitment of 
-value 
+\item either, before the timeout \haskellinline{start_timeout}, the user \haskellinline{person} makes a cash commitment 
+of value 
 \haskellinline{val} and timeout \haskellinline{end_timeout}  with the identifier \haskellinline{ident}, 
@@
 In this section, we revisit the escrow example that we discussed briefly in the introduction, and show how we can make 
-Marlowe contracts that are easier to write, read, and understand, by embedding them into Haskell code, taking advantage 
-of the fact that Marlowe contracts are implemented as Haskell terms.
-
-We used Haskell because it is the language in 
-which Marlowe is implemented, but it would be easy to embed Marlowe in any other language. It would only be necessary 
-to translate its primitives into a data type in that language. In Meadow we use Fay, which is a subset of 
-Haskell (see Section~\ref{section:tool}).
+Marlowe contracts that are easier to write, read, and understand, by embedding them into Haskell code, that is, taking 
+advantage of the fact that Marlowe contracts are implemented as Haskell terms to write Haskell programs that generate 
+Marlowe code, instead of writing Marlowe directly.
+
+We used Haskell because it is the language in which Marlowe is implemented, but it would be easy to embed Marlowe in any 
+other language. It would only be necessary to translate its primitives into a data type in that language. In Meadow we 
+use Fay \cite{Fay}, which is a subset of Haskell (see Section~\ref{section:tool}).
 
@@
 
-In our implementation we make things more specific: the money paid for the item is 450 ADA, and it must be committed by participant 
-\haskellinline{1} before block \haskellinline{10}; it will be refunded to participant \haskellinline{1} if there is 
-no consensus before block \haskellinline{90}.
+In our implementation we make things more specific: the money paid for the item is 450 ADA, and it must be committed by 
+participant \haskellinline{1} before block \haskellinline{10}; it will be refunded to participant \haskellinline{1} if 
+there is no consensus before block \haskellinline{90}.
 
@@
 \end{haskellcode}
-Then, we can easily write a function that returns an observation that is \haskellinline{True} if and only if at least one of the 
-participants \haskellinline{per} and \haskellinline{per'} has chosen the number \haskellinline{val} as follows:
+Then, we can easily write a function that returns an observation that is \haskellinline{True} if and only if at least 
+one of the participants \haskellinline{per} and \haskellinline{per'} has chosen the number \haskellinline{val} as 
+follows:
 \begin{haskellcode}
@@
 \end{haskellcode}
-Building on that, we can now write a function that returns an observation that is \haskellinline{True} if and only if at least 
-two out of the three participants \haskellinline{p1}, \haskellinline{p2}, and \haskellinline{p3} have agreed in 
-choosing the number \haskellinline{c} as follows:
+Building on that, we can now write a function that returns an observation that is \haskellinline{True} if and only if 
+at least two out of the three participants \haskellinline{p1}, \haskellinline{p2}, and \haskellinline{p3} have agreed 
+in choosing the number \haskellinline{c} as follows:
 \begin{haskellcode}
@@
 
-The outermost primitive \haskellinline{CommitCash} allows the participant \haskellinline{1} to commit \haskellinline{450} ADA before 
-block \haskellinline{10}, with the promise that money will be released on block \haskellinline{100} if they are not claimed before that.
+The outermost primitive \haskellinline{CommitCash} allows the participant \haskellinline{1} to commit 
+\haskellinline{450} ADA before block \haskellinline{10}, with the promise that money will be released on block 
+\haskellinline{100} if they are not claimed before that.
 
@@
 If the third option happens, the money is refunded. Otherwise, there is a \haskellinline{Choice} that will 
-immediately refund the money to participant \haskellinline{1} unless two out of three chose option \haskellinline{1}; in the later 
-case, \haskellinline{Pay} will give participant \haskellinline{2} the opportunity to claim the funds available in 
-the commitment with identifier \haskellinline{iCC1} before block \haskellinline{100} (by using the identifier \haskellinline{iP1} for 
-the claim). If the funds are not claimed before block \haskellinline{100} they will also be refunded to 
-participant \haskellinline{1}.
+immediately refund the money to participant \haskellinline{1} unless two out of three chose option \haskellinline{1}; 
+in the later case, \haskellinline{Pay} will give participant \haskellinline{2} the opportunity to claim the funds 
+available in the commitment with identifier \haskellinline{iCC1} before block \haskellinline{100} (by using the 
+identifier \haskellinline{iP1} for the claim). If the funds are not claimed before block \haskellinline{100} they will 
+also be refunded to participant \haskellinline{1}.
 
 Reflecting on the example, we can see that using Haskell definitions has made the contract substantially more 
-comprehensible.
-While our current implementation does not do this, it is possible to modify the embedding to support more efficient operation, by preserving \emph{sharing} in the host language.
-For example, if we were to replace the repeated  expression
+comprehensible. While our current implementation does not do this, it is possible to modify the embedding to support 
+more efficient operation, by preserving \emph{sharing} in the host language. For example, if we were to replace the 
+repeated  expression 
 \haskellinline{two_chose 1 2 3 1} by a \haskellinline{where} clause, 
@@
 editor for smart contracts written in Marlowe. Embedding support in Meadow
-is provided throught a pruned and bundled version of the Fay compiler \cite{Fay},
+is provided through a pruned and bundled version of the Fay compiler \cite{Fay},
 compiled to JavaScript using GHCJS. All text fields that edit and present Meadow and Fay code are
@@
 and formatted code (that is displayed in the upper right corner of the application),
-and to convert the code back to its blockly representation.
+and to convert the code back to its Blockly representation.
 
 Additionally, Meadow allows the user to use Meadow's Fay embedding from within the browser. % (see Figure~\ref{fig:embedded-demo}).
-Meadow includes a pruned copy of the Fay compiler, bundled with its Prelude module, 
-its foreign interface, and a module with the definitions of the Marlowe primitives and a function to pretty-print 
-Marlowe contracts. This allows the user to compile and execute, inside the browser, Fay
-code that generates and prints a Meadow contract. When the user opens the embedded editor, the current Marlowe contract 
-is embedded into a template of a Fay program that prints it. The user is able to modify this code in the left 
-panel of the editor and use the advanced functionalities provided by Fay like, for example, bindings, 
-list-comprehensions, turing-complete functions definitions, etc.
-
-When the user clicks the execute button, the Fay code is 
-compiled to JavaScript and evaluated; this causes the Marlowe contract generated by the execution of the code to be 
-written to the panel on the right. Generated contracts can then be sent back to the main screen of Meadow, translated 
-to Blockly, and their execution can be simulated. The embedded editor also allows users to save the Fay code to a 
-file in their computers, to load code from a file in their computers, and to temporarily hide the editor while keeping 
-its contents in memory, even while the compilation and execution process is being carried out in the background.
+
+The reason we chose Fay instead of Haskell for Meadow is a technical one: we wanted Meadow to be run completely inside 
+the browser, because that makes it easy to deploy, since the server is only required to send the page and not to 
+run any Marlowe-related computation. There exist compilers from Haskell to JavaScript but most of them are not easy to 
+bootstrap into JavaScript. Among other reasons, they often rely on the API of the host OS to read and write files.
+
+Fay compiler, on the other hand, is mostly written in pure Haskell, and it is not hard to compile to JavaScript by 
+using GHCJS. Nevertheless, it does not include a type checker, it relies in GHC's type-checker; and it also 
+tries to read its \haskellinline{Prelude} from disk. We have worked around these issues by disabling type-checking 
+within Meadow embedded editor, and by embedding all the required modules as constant strings in a modified version 
+of the Fay compiler.
+
+In particular, Meadow includes a pruned copy of the Fay compiler, bundled with its Prelude module, its foreign 
+interface, and a module with the definitions of the Marlowe primitives and a function to pretty-print 
+Marlowe contracts. This allows the user to compile and execute, inside Meadow, Fay code that generates and prints 
+a Marlowe contract. When the user opens the embedded editor, the current Marlowe contract is embedded into a template 
+of a Fay program that prints it. The user is able to modify this code in the left panel of the editor and use the 
+advanced functionalities provided by Fay like, for example, bindings, list-comprehensions, turing-complete functions 
+definitions, etc.
+
+When the user clicks the execute button, the Fay code is compiled to JavaScript and evaluated; this causes the Marlowe 
+contract generated by the execution of the code to be written to the panel on the right. Generated contracts can then be 
+sent back to the main screen of Meadow, translated to Blockly, and their execution can be simulated. The embedded editor 
+also allows users to save the Fay code to a file in their computers, to load code from a file in their computers, and to 
+temporarily hide the editor while keeping its contents in memory, even while the compilation and execution process is 
+being carried out in the background.
 
@@
 
-Marlowe is defined by its executable semantics, but to be deployed on blockchain it will have to be implemented on an existing distributed infrastructure.
-We intend to deploy it on IOHK's Cardano infrastructure, but it can be implemented in other systems too, as we explain now.
+Marlowe is defined by its executable semantics, but to be deployed on blockchain it will have to be implemented on an 
+existing distributed infrastructure.  We intend to deploy it on IOHK's Cardano infrastructure, but it can be implemented 
+in other systems too, as we explain now.
 
@@
 
-Marlowe abstracts away from a number of concrete details of how blockchains operate.
-In particular, it is agnostic between UTxO-based systems, such as Bitcoin and Cardano SL, and account-based models such as Ethereum; it can be implemented on push or pull-based systems, and can be executed on or off-chain.
+Marlowe abstracts away from a number of concrete details of how blockchains operate. In particular, it is agnostic 
+between UTxO-based systems, such as Bitcoin and Cardano SL, and account-based models such as Ethereum; it can be 
+implemented on push or pull-based systems, and can be executed on or off-chain.
 
@@
 Bitcoin blockchain. On the other hand, the Ethereum model explicitly keeps track of account values, and  this 
-information needs to be kept on chain. Of the two, the UTxO model is simpler and requires less support from the implementation of the chain itself.
+information needs to be kept on chain. Of the two, the UTxO model is simpler and requires less support from the 
+implementation of the chain itself.
 
@@
 
-Contracts can be conceived of as acting in two ways.
-In a \emph{push} model, contracts are seen to make things happen.
+Contracts can be conceived of as acting in two ways. In a \emph{push} model, contracts are seen to make things happen. 
 In the case of blockchain, this would include making payments or transactions take place. 
 Alternatively, in a \emph{pull} model, contracts enable certain things, such as transactions,  to happen but the 
-transactions need to be effected by an external actor. The pull model makes fewer demands on the blockchain implementation than the push model.
+transactions need to be effected by an external actor. The pull model makes fewer demands on the blockchain 
+implementation than the push model.
 
@@
 
-Some chains have a layered design, in particular Cardano~\cite{cardano}.
-The Cardano Settlement Layer (SL) is to support settlement of transactions but nothing more complicated, whereas the Computation Layer (or CL) supports more powerful general computations over the chain.
-Moreover, the Cardano roadmap~\cite{cardano-rationale} envisages the possibility of computation taking place, in part at least, on a sidechain rather than on the main chain itself, which
-may support only the SL. Other off-chain approaches include the Lightning network, overlaid on blockchains (usually Bitcoin) and the channels of \ae{}ternity~\cite{aeternity}.
+Some chains have a layered design, in particular Cardano~\cite{cardano}. The Cardano Settlement Layer (SL) is to 
+support settlement of transactions but nothing more complicated, whereas the Computation Layer (or CL) supports more 
+powerful general computations over the chain. Moreover, the Cardano roadmap~\cite{cardano-rationale} envisages the 
+possibility of computation taking place, in part at least, on a sidechain rather than on the main chain itself, which 
+may support only the SL. Other off-chain approaches include the Lightning network, overlaid on blockchains (usually 
+Bitcoin) and the channels of \ae{}ternity~\cite{aeternity}.
 
@@
 \item one or more on-chain programs and 
-\item a client side program or interface that has a limited number of choices to be made at each point in time (including an interface for the observables). 
+\item a client side program or interface that has a limited number of choices to be made at each point in time 
+(including an interface for the observables). 
 \end{itemize}
-The contracts implemented in the Marlowe language have an implicit interface defined for what would correspond to the client side of a distributed application, as shown by the interactive interface. 
-Of course, we probably do not want to give a user a programmatic interface but an actual application with user interface. 
+The contracts implemented in the Marlowe language have an implicit interface defined for what would correspond to the 
+client side of a distributed application, as shown by the interactive interface. Of course, we probably do not want to 
+give a user a programmatic interface but an actual application with user interface. 
 %So there is still some work to be done to create interfaces on top of the client side of the contracts.
@@
 
-Usually, in the UTxO model, it is necessary that a UTxO is in the blockchain in order for another transaction to be able to spend it.
-This could potentially limit the number of operations that can be applied in a single block to one, which would make infeasible the approach of using the same UTxO to represent several instances of the same contract (since it would not scale).
-On the other hand, it would prevent non-determinism and, with it, it would remove the possibility of race-conditions.
-A change to the state of a contract before a transaction (and its corresponding operation) is accepted in the blockchain
+Usually, in the UTxO model, it is necessary that a UTxO is in the blockchain in order for another transaction to be 
+able to spend it. This could potentially limit the number of operations that can be applied in a single block to one, 
+which would make infeasible the approach of using the same UTxO to represent several instances of the same contract 
+(since it would not scale).
+On the other hand, it would prevent non-determinism and, with it, it would remove the possibility of race-conditions. A 
+change to the state of a contract before a transaction (and its corresponding operation) is accepted in the blockchain 
 would invalidate the transaction and require the user to issue it again.
@@
 
-Nevertheless, there already exists in Bitcoin a mechanism that allows the combination of several chained transactions within a single block while using the UTxO model, even though it is used to solve a different problem and it is not widely spread yet.
-This is the case of "child pays for parent", which allows the recipient of an unpublished transaction to issue another transaction that spends the former, with the aim of including enough fees for both transactions to have enough incentive to be included by a miner in a block (in which case both transactions would be included in the same block).
-
-The same mechanism could be used for allowing miners to aggregate several transactions that act as a chain in the same block.
-The clients would only need to monitor the transaction pool in order to send transactions that can be chained with the existing ones.
+Nevertheless, there already exists in Bitcoin a mechanism that allows the combination of several chained transactions 
+within a single block while using the UTxO model, even though it is used to solve a different problem and it is not 
+widely spread yet. This is the case of "child pays for parent", which allows the recipient of an unpublished transaction 
+to issue another transaction that spends the former, with the aim of including enough fees for both transactions to have 
+enough incentive to be included by a miner in a block (in which case both transactions would be included in the same 
+block).
+
+The same mechanism could be used for allowing miners to aggregate several transactions that act as a chain in the same 
+block. The clients would only need to monitor the transaction pool in order to send transactions that can be chained 
+with the existing ones.
 
@@
 
-The Cardano platform has a notion of cost, and Marlowe contracts will potentially incur costs of two kinds: The cost of a single transaction execution (e.g.\ a single Plutus validator), and the cost of the whole contract, which may consist of many transactions.
+The Cardano platform has a notion of cost, and Marlowe contracts will potentially incur costs of two kinds: The cost of 
+a single transaction execution (e.g.\ a single Plutus validator), and the cost of the whole contract, which may consist 
+of many transactions.
 
@@
 %
-%The implementation may split contract execution between the main chain and a sidechain (or chains). This will allow links from Marlowe to work on securing multiparty sidechain games/contracts (MPCs).
+%The implementation may split contract execution between the main chain and a sidechain (or chains). This will allow 
+%links from Marlowe to work on securing multiparty sidechain games/contracts (MPCs).
 
@@
 ways, since it has to be amenable to replication or verification within a reasonable time if the security and integrity 
-of the chain are to be preserved. We look at representatives of the main approaches now; we discuss some of these in more detail in an earlier paper~\cite{cryptoeprint:2016:1156}.
+of the chain are to be preserved. We look at representatives of the main approaches now; we discuss some of these in 
+more detail in an earlier paper~\cite{cryptoeprint:2016:1156}.
 
@@
 
-The \ae{}ternity system~\cite{aeternity} provides multiple languages and VMs~\cite{Stenman-CODE-BEAM}: the functional language  Sophia (akin to 
-Reason) and the Functional Typed Warded Virtual Machine (FTWVM) for safe ``system level programming'', the language 
-Varna and the HLM for simple contracts, and a (port of) Solidity and the EVM for compatibility with Ethereum. In a 
-similar way, Cardano provides support for IELE~\cite{IELE}, a rational reconstruction of the EVM, and thus for Solidity too.
+The \ae{}ternity system~\cite{aeternity} provides multiple languages and VMs~\cite{Stenman-CODE-BEAM}: the functional 
+language  Sophia (akin to Reason) and the Functional Typed Warded Virtual Machine (FTWVM) for safe ``system level 
+programming'', the language Varna and the HLM for simple contracts, and a (port of) Solidity and the EVM for 
+compatibility with Ethereum. In a similar way, Cardano provides support for IELE~\cite{IELE}, a rational reconstruction 
+of the EVM, and thus for Solidity too.
 
@@
 seminal~\cite{PeytonJones:2000}, and the authors note that payments need to be bounded; this is made concrete in our 
-account by our notion of commitments. They take no account of commitments or timeouts as our approach does, but it should be 
-noted that the Ethereum platform is more powerful than the one that we target in this paper.
+account by our notion of commitments. They take no account of commitments or timeouts as our approach does, but it 
+should be noted that the Ethereum platform is more powerful than the one that we target in this paper.
 
@@
 
-In this paper, we have presented Marlowe, a DSL for financial contracts on blockchains, based on earlier work on contracts~\cite{PeytonJones:2000}, together with examples of its use.
-We have seen that to make this operational on blockchain we need to add commitments and timeouts, and to design a semantics that reflects these.
-As we saw in Section~\ref{sec:abstraction}, Marlowe has been designed to make as few demands as possible on the underlying blockchain: it can be implemented on UTxO or account-based blockchains, for example.
-
-We have also presented Meadow, that allows users to interact with and simulate the operation of Marlowe contracts, contributing to the potential adoptability of the system.
-We have also described the design rationale for Marlowe, and sketched ways in which it can be implemented.
-
-We plan to continue the work with Marlowe in a number of directions.
-We will continue to develop the core language, for example considering the automation of generation of identifiers for commitments and others.
-We will then implement this version of Marlowe in Cardano, compiling from Marlowe contracts to on chain contracts and users' wallets, and deploy it in a test network to observe and measure its behaviour in practice.
-
-Building on the operational semantics given here, we will develop analyses of Marlowe contracts -- such as to show that contracts cannot generate \haskellinline{FailedPay} actions in certain circumstances.
-We will also develop  QuickCheck-style property-based testing~\cite{quickCheck}, and properties developed here can become candidates for fully fledged verification in a formalisation of the semantics of Marlowe.
-
-We would like to thank IOHK for supporting us: not only has this led us to work on Marlowe, but we have benefitted hugely from collaboration with colleagues including Manuel Chakravarty, Duncan Coutts, Bernardo David, Charles Hoskinson, Aggelos Kiayias, % Darryl McAdams,
-Bruno Woltzenlogel Paleo, Rebecca Valentine and Phil Wadler.
-We are  very grateful to Thomas Arts and colleagues from \ae{}ternity for their convivial discussions on blockchain and contracts in G\"oteborg, and to colleagues from the Universities of Kent and Leicester for their comments too.
+In this paper, we have presented Marlowe, a DSL for financial contracts on blockchains, based on earlier work on 
+contracts~\cite{PeytonJones:2000}, together with examples of its use. We have seen that to make this operational on 
+blockchain we need to add commitments and timeouts, and to design a semantics that reflects these. As we saw in 
+Section~\ref{sec:abstraction}, Marlowe has been designed to make as few demands as possible on the underlying 
+blockchain: it can be implemented on UTxO or account-based blockchains, for example.
+
+We have also presented Meadow, that allows users to interact with and simulate the operation of Marlowe contracts, 
+contributing to the potential adoptability of the system. We have also described the design rationale for Marlowe, and 
+sketched ways in which it can be implemented.
+
+We plan to continue the work with Marlowe in a number of directions. We will continue to develop the core language, for 
+example considering the automation of generation of identifiers for commitments and others. We will then implement this 
+version of Marlowe in Cardano, compiling from Marlowe contracts to on chain contracts and users' wallets, and deploy it 
+in a test network to observe and measure its behaviour in practice.
+
+Building on the operational semantics given here, we will develop analyses of Marlowe contracts -- such as to show that 
+contracts cannot generate \haskellinline{FailedPay} actions in certain circumstances. We will also develop  
+QuickCheck-style property-based testing~\cite{quickCheck}, and properties developed here can become candidates for fully 
+fledged verification in a formalisation of the semantics of Marlowe.
+
+We would like to thank IOHK for supporting us: not only has this led us to work on Marlowe, but we have benefited 
+hugely from collaboration with colleagues including Manuel Chakravarty, Duncan Coutts, Bernardo David, Charles 
+Hoskinson, Aggelos Kiayias, % Darryl McAdams,
+Bruno Woltzenlogel Paleo, Rebecca Valentine and Phil Wadler. We are  very 
+grateful to Thomas Arts and colleagues from \ae{}ternity for their convivial discussions on blockchain and contracts in 
+G\"oteborg, and to colleagues from the Universities of Kent and Leicester for their comments too.
 

### 2018-07-18 tex/paper.tex
Expand comparison in related work

@@
 expressibility of the DSL. A notable example of this is the financial contracts language described by Peyton Jones and 
-his collaborators~\cite{PeytonJones:2000}, which is embedded in Haskell.
+his collaborators~\cite{PeytonJones:2000}, which is embedded in Haskell. Marlowe is similar to the 
+DSL in~\cite{PeytonJones:2000} in the use of Haskell embedding, in that the functionality provided is similar, in the use of 
+composable combinators, and in the declarative style that allows users to describe what needs to be enforced and not how. 
+However the two approaches differ in other aspects like the lack of Marlowe's reliance on the legal system for 
+enforcement, its support for multiple party contracts, its explicitness of choices, and its use of a pull model.
 

### 2018-07-26 tex/paper.tex
Update semantics in paper

@@
   | right_claim =
-    if committed st from bn >= cval
+    if ((committed st from bn >= cval) && (cval >= 0))
       then (newstate, con, [SuccessfulPay idpay from to cval])

### 2018-07-26 tex/paper.tex
Assded reference to split contracts, and discussion in related work. Changed the title of section 5 and 5.1, and added more about design rationale to the introduction. Added new introduction to the Meadow section. Modified the discussion of "embedded" in contributions.

--- 2018-07-18 tex/paper.tex
@@
 \item 
-Embedding of Marlowe in Haskell to extend its expressibility; that is: Marlowe is defined as a Haskell 
-\haskellinline{data} type so that generating Marlowe contracts from Haskell is straightforward.
+Making Marlowe an embedded DSL in Haskell.
+This extends the expressibility of the language, as we can use all the facilities of Haskell in defining Marlowe contracts; we achieve this by defining Marlowe as  a Haskell \haskellinline{data} type.
 \item
@@
 case of a `proof of work'-based chain, and with users participating in contract execution. We can also perform analyses 
-of Marlowe contracts, based on its formal semantics. Finally, we can explore the design space for Marlowe, and explore 
-potential alternative designs.
+of Marlowe contracts, based on its formal semantics.
+
+In designing a language like Marlowe we are constrained by the blockchain domain, but within that we do have a range of choices.
+For example, should we base the language on a system with accounts, like Ethereum, or a UTxO-based model as used by bitcoin?
+We examine this and other choices after introducing the language and examples of its use.
 
@@
 Section \ref{section:tool} introduces Meadow, our tool for visualising and interacting with Marlowe contracts.
-Section \ref{sec:abstraction} reflects on the system design, showing how Marlowe can be supported on a variety of 
+Section \ref{sec:abstraction} reflects on the design rationale for Marlowe, showing how it can be supported on a variety of 
 blockchains, and Section \ref{sec:compilation} explores how Marlowe can be implemented. Section \ref{sec:related} 
@@
 other language. It would only be necessary to translate its primitives into a data type in that language. In Meadow we 
-use Fay \cite{Fay}, which is a subset of Haskell (see Section~\ref{section:tool}).
+use Fay \cite{Fay}, a subset of Haskell that we discuss in more detail in Section~\ref{section:tool}).
 
@@
 
-One of the objectives of Marlowe is to be accessible to a wider audience
-than a general purpose language would. With this aim, and to aid Marlowe's
-usage by people that does not know or is not familiar with its syntax,
+For Marlowe to be usable in practice, users need to be able to understand how contracts will behave once deployed to the blockchain, but without doing the deployment.
+We can do that by simulating their behaviour off-chain, interactively stepping through the evaluation of a contract in a browser.
+We do this in two stages, first transforming an embedded contact (using features of Haskell) to a pure Marlowe contract, and then interactively stepping through that contract.
+
+To achieve this, and to aid Marlowe's take-up
+usage by people that does not know or are not familiar with its syntax,
 we have developed Meadow, a web tool that supports the interactive construction,
-edition, and simulation of smart-contracts written in Marlowe. The
+revision, and simulation of smart-contracts written in Marlowe. The
 tool is publicly available in the url: \url{https://input-output-hk.github.io/scdsl/}.
@@
 
-The reason we chose Fay instead of Haskell for Meadow is a technical one: we wanted Meadow to be run completely inside 
+The reason that we chose Fay instead of Haskell for Meadow is a technical one: we wanted Meadow to be run completely inside 
 the browser, because that makes it easy to deploy, since the server is only required to send the page and not to 
@@
 
-Fay compiler, on the other hand, is mostly written in pure Haskell, and it is not hard to compile to JavaScript by 
+The Fay compiler, on the other hand, is mostly written in pure Haskell, and it is not hard to compile to JavaScript by 
 using GHCJS. Nevertheless, it does not include a type checker, it relies in GHC's type-checker; and it also 
@@
 
-\section{Implementation on blockchain}
+\section{Design and Implementation}
 \label{section:towards-implementation}
@@
 
-\subsection{Abstraction}
+\subsection{Design Rationale}
 \label{sec:abstraction}
+\label{sec:design-rationale}
 
@@
 between UTxO-based systems, such as Bitcoin and Cardano SL, and account-based models such as Ethereum; it can be 
-implemented on push or pull-based systems, and can be executed on or off-chain.
+implemented on ``push'' or ``pull''-based systems, and can be executed on or off-chain.
 
@@
 
+\paragraph{Split contracts}
+An early work on smart contracts raises the issue that practical contractual situations may well not be amenable to complete formalisation, hence giving rise to a \emph{split}~\cite{split-contracts} between an automated part and a non-automated part that mediates, for example, real assets.
+We can foresee that this approach will be needed for Marlowe too, if it achieves read-world adoption.
+
 \paragraph{Bitcoin script}

### 2018-07-26 tex/paper.tex
Split lines

@@
 \item 
-Making Marlowe an embedded DSL in Haskell.
-This extends the expressibility of the language, as we can use all the facilities of Haskell in defining Marlowe contracts; we achieve this by defining Marlowe as  a Haskell \haskellinline{data} type.
+Making Marlowe an embedded DSL in Haskell. This extends the expressibility of the language, as we can use all the 
+facilities of Haskell in defining Marlowe contracts; we achieve this by defining Marlowe as  a Haskell 
+\haskellinline{data} type.
 \item
@@
 
-In designing a language like Marlowe we are constrained by the blockchain domain, but within that we do have a range of choices.
-For example, should we base the language on a system with accounts, like Ethereum, or a UTxO-based model as used by bitcoin?
-We examine this and other choices after introducing the language and examples of its use.
+In designing a language like Marlowe we are constrained by the blockchain domain, but within that we do have a range of 
+choices. For example, should we base the language on a system with accounts, like Ethereum, or a UTxO-based model as 
+used by bitcoin? We examine this and other choices after introducing the language and examples of its use.
 
@@
 
-For Marlowe to be usable in practice, users need to be able to understand how contracts will behave once deployed to the blockchain, but without doing the deployment.
-We can do that by simulating their behaviour off-chain, interactively stepping through the evaluation of a contract in a browser.
-We do this in two stages, first transforming an embedded contact (using features of Haskell) to a pure Marlowe contract, and then interactively stepping through that contract.
+For Marlowe to be usable in practice, users need to be able to understand how contracts will behave once deployed to 
+the blockchain, but without doing the deployment. We can do that by simulating their behaviour off-chain, interactively 
+stepping through the evaluation of a contract in a browser. We do this in two stages, first transforming an embedded 
+contact (using features of Haskell) to a pure Marlowe contract, and then interactively stepping through that contract.
 
@@
 \paragraph{Split contracts}
-An early work on smart contracts raises the issue that practical contractual situations may well not be amenable to complete formalisation, hence giving rise to a \emph{split}~\cite{split-contracts} between an automated part and a non-automated part that mediates, for example, real assets.
-We can foresee that this approach will be needed for Marlowe too, if it achieves read-world adoption.
+An early work on smart contracts raises the issue that practical contractual situations may well not be amenable to 
+complete formalisation, hence giving rise to a \emph{split}~\cite{split-contracts} between an automated part and a 
+non-automated part that mediates, for example, real assets. We can foresee that this approach will be needed for Marlowe 
+too, if it achieves read-world adoption.
 

### 2018-07-26 tex/paper.tex
Give names to participants in escrow contract

@@
 more concretely. The aim of this contract, written in functional pseudocode in the style of~\cite{PeytonJones:2000} 
-involves three participants \haskellinline{1}, \haskellinline{2} and \haskellinline{3}. Participant \haskellinline{1} is 
-to pay an amount of money to participant \haskellinline{2} on receipt of goods from her. Participant \haskellinline{1} 
-pays the money into escrow controlled by participant \haskellinline{3}.
-
-There are two options for the money: if two participants agree to \haskellinline{pay} it to \haskellinline{2}, that goes 
-ahead; if, on the other hand, two of the participants opt to \haskellinline{refund} the money to \haskellinline{1}, that 
-is done instead.
+involves three participants: \haskellinline{alice}, \haskellinline{bob} and \haskellinline{carol}. \haskellinline{alice} 
+is to pay an amount of money to \haskellinline{bob} on receipt of goods from her. \haskellinline{alice} pays the money 
+into escrow controlled by \haskellinline{carol}.
+
+There are two options for the money: if two out of the three participants agree to \haskellinline{pay} it to 
+\haskellinline{bob}, that goes ahead; if, on the other hand, two of the participants opt to \haskellinline{refund} the 
+money to \haskellinline{alice}, that is done instead.
 
@@
 \begin{haskellcode}
-(When (Or (two_chose 1 2 3 refund)
-          (two_chose 1 2 3 pay))
-      (Choice (two_chose 1 2 3 pay)
-              (Pay 1 2 AvailableMoney)
+(When (Or (two_chose alice bob carol refund)
+          (two_chose alice bob carol pay))
+      (Choice (two_chose alice bob carol pay)
+              (Pay alice bob AvailableMoney)
               redeem_original))
@@
 be run on top of the blockchain. In the traditional model, enforcement of the contract is the responsibility of the 
-legal system. If participant \haskellinline{1} does not pay the money into escrow, or participant \haskellinline{3} 
-chooses to keep it for herself, then they can be sued for the money (and probably damages), thus providing both legal 
-and financial incentives for compliance. On the other hand, in the decentralised blockchain model, where there is no 
-central authority, the contract needs to be enforced \emph{by design}.
+legal system. If \haskellinline{alice} does not pay the money into escrow, or \haskellinline{carol} chooses to keep it 
+for herself, then they can be sued for the money (and probably damages), thus providing both legal and financial 
+incentives for compliance. On the other hand, in the decentralised blockchain model, where there is no central 
+authority, the contract needs to be enforced \emph{by design}.
 
@@
 If the cash is committed successfully and the timeout \haskellinline{end_timeout} is reached, then it is impossible to 
-further spend the committed cash, and any unspent funds can be reclaimed by participant \haskellinline{person}.
+further spend the committed cash, and any unspent funds can be reclaimed by \haskellinline{person}.
  This is enforced by the \haskellinline{stepBlock} function, as noted above.
@@
 In this section, we show how Marlowe can be used for designing a contract that incentivises saving. We use the 
-following example: participant \haskellinline{2} offers 20 ADA to participant \haskellinline{1} in exchange for saving 
-100 ADA for 100 blocks. participant \haskellinline{1} will be able to withdraw the savings after block 20, but if that 
-option is chosen, then participant \haskellinline{2} will not pay the extra 20 ADA to participant \haskellinline{1}.
-
-We will start from outer to inner primitives. The first thing is to allow participant \haskellinline{1} to commit the 
+following example: \haskellinline{bob} offers 20 ADA to \haskellinline{alice} in exchange for saving 100 ADA for 100 
+blocks. \haskellinline{alice} will be able to withdraw the savings after block 20, but if that option is chosen, then 
+\haskellinline{bob} will not pay the extra 20 ADA to participant \haskellinline{alice}.
+
+We will start from outer to inner primitives. The first thing is to allow \haskellinline{alice} to commit the 
 100 ADA. The primitive says that with identity \haskellinline{IdentCC 1} (this is the identity of the commitment), 
-participant \haskellinline{1} can commit \haskellinline{100} ADA before block \haskellinline{10}, and the commitment 
+\haskellinline{alice} can commit \haskellinline{100} ADA before block \haskellinline{10}, and the commitment 
 must expire on block \haskellinline{200}. But if the money is not committed before block \haskellinline{10}, then the 
@@
 If the commitment is done correctly and on time, we continue as in \haskellinline{step2}. \haskellinline{step2} is 
-another commitment, this time by participant \haskellinline{2}. The primitive says that with identity 
-\haskellinline{IdentCC 2}, participant \haskellinline{2} can commit \haskellinline{20} ADA before block 
+another commitment, this time by \haskellinline{bob}. The primitive says that with identity 
+\haskellinline{IdentCC 2}, \haskellinline{bob} can commit \haskellinline{20} ADA before block 
 \haskellinline{20}, and the commitment must expire on block \haskellinline{200}. But if the money is not committed 
 before block \haskellinline{20}, then the contract will continue as \haskellinline{RedeemCC (IdentCC 1) Null}, which in 
-turn gives participant \haskellinline{1} the opportunity to withdraw all the money committed immediately.
+turn gives \haskellinline{alice} the opportunity to withdraw all the money committed immediately.
 
@@
 If the commitment is done correctly and on time, we continue as in \haskellinline{step3}. \haskellinline{step3}
-waits until choice with identity \haskellinline{IdentChoice 1} is made by participant \haskellinline{1} or
+waits until choice with identity \haskellinline{IdentChoice 1} is made by \haskellinline{alice} or
 until block \haskellinline{100} (whatever happens first). If the choice is made before block \haskellinline{100}, 
@@
 
-\haskellinline{step4} pays the \haskellinline{20} ADA committed by participant \haskellinline{2} in the commit with 
-identity \haskellinline{IdentPay 1} to participant \haskellinline{1}, if the payment is claimed before block 
+\haskellinline{step4} pays the \haskellinline{20} ADA committed by \haskellinline{bob} in the commit with 
+identity \haskellinline{IdentPay 1} to \haskellinline{alice}, if the payment is claimed before block 
 \haskellinline{200}. Once the payment is claimed or after block 200, it continues as \haskellinline{stepRedeem}.
@@
 %Let us imagine that participant 
-%\haskellinline{1} is buying some item from participant \haskellinline{2}. On one hand, if participant \haskellinline{1} 
-%pays before receiving the item, there is a risk that participant \haskellinline{2} will never give \haskellinline{1} the 
-%item. On the other hand, if participant \haskellinline{2} gives participant \haskellinline{1} the item before 
-%\haskellinline{1} pays,  there is a risk that \haskellinline{1} will never pay.
-%
-The escrow mechanism allows participant \haskellinline{1} to deposit the money into a contract, in a way that the money 
-will only be released when two out of three participants agree on whether participant \haskellinline{2} has indeed given 
-participant \haskellinline{1} the item.
-
-The third participant (participant \haskellinline{3}) is supposed to be a 
-neutral third party that will decide in case of dispute.
-This way, if participants \haskellinline{1} and \haskellinline{2} are honest, they will just agree on the result of the 
-transaction and \haskellinline{3} will not need to do anything. If \haskellinline{1} and \haskellinline{2} disagree, 
-participant \haskellinline{3} will be able to choose whether the money must go to \haskellinline{1} or to 
-\haskellinline{2}.
+%\haskellinline{alice} is buying some item from \haskellinline{bob}. On one hand, if \haskellinline{alice} 
+%pays before receiving the item, there is a risk that \haskellinline{bob} will never give \haskellinline{alice} the 
+%item. On the other hand, if \haskellinline{bob} gives \haskellinline{alice} the item before 
+%\haskellinline{alice} pays,  there is a risk that \haskellinline{alice} will never pay.
+%
+The escrow mechanism allows \haskellinline{alice} to deposit the money into a contract, in a way that the 
+money will only be released when two out of three participants agree on whether \haskellinline{bob} has indeed 
+given \haskellinline{alice} the item.
+
+The escrow participant (\haskellinline{carol}) is supposed to be a neutral third party that will decide in case of 
+dispute. This way, if participants \haskellinline{alice} and \haskellinline{bob} are honest, they will just agree on 
+the result of the transaction and \haskellinline{carol} will not need to do anything. If \haskellinline{alice} and 
+\haskellinline{bob} disagree, \haskellinline{carol} will be able to choose whether the money must go to 
+\haskellinline{alice} or to \haskellinline{bob}.
 
 In our implementation we make things more specific: the money paid for the item is 450 ADA, and it must be committed by 
-participant \haskellinline{1} before block \haskellinline{10}; it will be refunded to participant \haskellinline{1} if 
+\haskellinline{alice} before block \haskellinline{10}; it will be refunded to participant \haskellinline{alice} if 
 there is no consensus before block \haskellinline{90}.
@@
 \end{haskellcode}
-And an identifier for the payment:
+An identifier for the payment:
 \begin{haskellcode}
@@
 \end{haskellcode}
-We can also create a sub-contract that allows the money from the commitment with identifier \haskellinline{IdentCC 1} to 
-be redeemed:
+And we can create identifiers for all the participants:
+\begin{haskellcode}
+alice :: Person
+alice = 1
+
+bob :: Person
+bob = 2
+
+carol :: Person
+carol = 3
+\end{haskellcode}
+We can also create a sub-contract that allows the money from the commitment with identifier \haskellinline{IdentCC 1} 
+to be redeemed:
 \begin{haskellcode}
@@
 Once redeemed, the contract continues as \haskellinline{Null} since we expect this to be the last thing that is done.
-Each participant has a say on who deserves the money: either participant \haskellinline{1} deserves a refund, which we 
-represent with the number $0$; or participant \haskellinline{2} deserves a payment, which we represent with the number 
-$1$. Because there is only one choice to make per participant, we use the same \haskellinline{IdentChoice} as their 
-participant id. We can define a function that returns an observation that is true if and only if
+Each participant has a say on who deserves the money: either \haskellinline{alice} deserves a refund, which 
+we represent with the number $0$; or \haskellinline{bob} deserves a payment, which we represent with the 
+number $1$. Because there is only one choice to make per participant, we use the same \haskellinline{IdentChoice} as 
+their participant id. We can define a function that returns an observation that is true if and only if
 \haskellinline{per} has chosen the number \haskellinline{c} for the choice \haskellinline{IdentChoice per} as follows:
@@
 escrow = CommitCash iCC1 1 (ConstMoney 450) 10 100
-                    (When (OrObs (two_chose 1 2 3 0)
-                                 (two_chose 1 2 3 1))
+                    (When (OrObs (two_chose alice bob carol 0)
+                                 (two_chose alice bob carol 1))
                           90
-                          (Choice (two_chose 1 2 3 1)
-                                  (Pay iP1 1 2 (AvailableMoney iCC1) 100
+                          (Choice (two_chose alice bob carol 1)
+                                  (Pay iP1 alice bob (AvailableMoney iCC1) 100
                                        redeem_original)
@@
 
-The outermost primitive \haskellinline{CommitCash} allows the participant \haskellinline{1} to commit 
+The outermost primitive \haskellinline{CommitCash} allows the \haskellinline{alice} to commit 
 \haskellinline{450} ADA before block \haskellinline{10}, with the promise that money will be released on block 
@@
 If the third option happens, the money is refunded. Otherwise, there is a \haskellinline{Choice} that will 
-immediately refund the money to participant \haskellinline{1} unless two out of three chose option \haskellinline{1}; 
-in the later case, \haskellinline{Pay} will give participant \haskellinline{2} the opportunity to claim the funds 
+immediately refund the money to \haskellinline{alice} unless two out of three chose option 
+\haskellinline{1}; 
+in the later case, \haskellinline{Pay} will give \haskellinline{bob} the opportunity to claim the funds 
 available in the commitment with identifier \haskellinline{iCC1} before block \haskellinline{100} (by using the 
 identifier \haskellinline{iP1} for the claim). If the funds are not claimed before block \haskellinline{100} they will 
-also be refunded to participant \haskellinline{1}.
+also be refunded to \haskellinline{alice}.
 
@@
 repeated  expression 
-\haskellinline{two_chose 1 2 3 1} by a \haskellinline{where} clause, 
-\begin{haskellcode}
-escrow = ...        (When (OrObs (two_chose 1 2 3 0)
+\haskellinline{two_chose alice bob carol 1} by a \haskellinline{where} clause, 
+\begin{haskellcode}
+escrow = ...        (When (OrObs (two_chose alice bob carol 0)
                                  chose_refund)
@@
                            ...
-         where chose_refund = two_chose 1 2 3 1                 
+         where chose_refund = two_chose alice bob carol 1                 
 \end{haskellcode}

### 2018-07-26 src/EscrowV2.hs

@@ -30,16 +30,24 @@ import Semantics
 -- refund = 0
 -- pay    = 1
 
+alice :: Person
+alice = 1
+
+bob :: Person
+bob = 2
+
+carol :: Person
+carol = 3
 
 escrow :: Contract
-escrow = CommitCash iCC1 1
+escrow = CommitCash iCC1 alice
                     (ConstMoney 450)
                     10 100
-                    (When (OrObs (two_chose 1 2 3 0)
-                                 (two_chose 1 2 3 1))
+                    (When (OrObs (two_chose alice bob carol 0)
+                                 (two_chose alice bob carol 1))
                           90
-                          (Choice (two_chose 1 2 3 1)
-                                  (Pay iP1 1 2
+                          (Choice (two_chose alice bob carol 1)
+                                  (Pay iP1 alice bob
                                        (AvailableMoney iCC1)
                                        100
                                        redeem_original)

### 2018-07-26 tex/paper.tex
Added single type decl for names
Fix some typos

@@
 The outer primitive \haskellinline{When} waits until the condition -- its first argument -- 
-becomes true; in this case, the condition is the that either two participants choose \haskellinline{refund} or two 
+becomes true; in this case, the condition is that either two participants choose \haskellinline{refund} or two 
 participants choose \haskellinline{pay}. The second argument of the \haskellinline{When} is itself another 
@@
 participants themselves. These include commitments of currency (or ``cash''), redemption of commitments, and claims of 
-payments by a participant. Moreover it is also possible for a participant to input an arbitrary value (which we term a 
+payments by a participant. Moreover, it is also possible for a participant to input an arbitrary value (which we term a 
 ``choice''). The particular inputs at a given step are described by a value of type \haskellinline{Input}.
@@
 blocks. \haskellinline{alice} will be able to withdraw the savings after block 20, but if that option is chosen, then 
-\haskellinline{bob} will not pay the extra 20 ADA to participant \haskellinline{alice}.
+\haskellinline{bob} will not pay the extra 20 ADA to \haskellinline{alice}.
 
@@
 In our implementation we make things more specific: the money paid for the item is 450 ADA, and it must be committed by 
-\haskellinline{alice} before block \haskellinline{10}; it will be refunded to participant \haskellinline{alice} if 
+\haskellinline{alice} before block \haskellinline{10}; it will be refunded to \haskellinline{alice} if 
 there is no consensus before block \haskellinline{90}.
@@
 \begin{haskellcode}
-alice :: Person
+alice, bob, carol :: Person
 alice = 1
-
-bob :: Person
 bob   = 2
-
-carol :: Person
 carol = 3
@@
 stepping through the evaluation of a contract in a browser. We do this in two stages, first transforming an embedded 
-contact (using features of Haskell) to a pure Marlowe contract, and then interactively stepping through that contract.
+contract (using features of Haskell) to a pure Marlowe contract, and then interactively stepping through that contract.
 
@@
 ADA, using a value for the prevailing USD/ADA exchange rate. This will require a computation: in this case, a 
-multiplication. If Marlowe is to be stand-alone then we need to be extend it with arithmetic and other operations, but 
+multiplication. If Marlowe is to be stand-alone then we need to extend it with arithmetic and other operations, but 
 we expect Marlowe to be embedded in a suitable language that provides these facilities.

### 2019-10-03 semantics-1.0/tex/oracles.tex
Move Marlowe 1.3 to semantics-1.0 folder

\documentclass{article}
\usepackage{hyperref}
\title{Oracles in Blockchain Technologies}
\author{Alexander Nemish}
\begin{document}
\pagenumbering{gobble}
\maketitle
\section{Introduction}

From
\href{https://blog.apla.io/what-is-a-blockchain-oracle-2ccca433c026}{What is a blockchain oracle?}

Oracles are trusted data sources or entities that provide information or sign claims about the state of the world for smart contracts.
They are the link between real world events and the digital world of blockchain platforms.
They don’t make predictions about the future but report events from the past.

Smart contracts essentially request information from an oracle such as a random number or the results of a football game from the ‘real world’ and then store it on the blockchain.
It then continues executing the smart contract based on this new data.

Take for example Bloomberg or Reuters.
These would be considered data feeds instead of oracles in the blockchain world because interfacing with blockchains is required in addition to the ability to obtain data from multiple sources to, in effect achieve, consensus.
This is why oracles are much more that just data feeds or sources.

The most powerful oracle architecture offers a decentralized network of oracles that connects directly to smart contracts, and feeds the data back in a secure manner. With both software and hardware security measures, this ensures security and tamper-proof data that can be trusted

\section{Existing Oracles}

From \url{https://blog.aeternity.com/blockchain-oracles-657f134ffbc0}

Companies working on oracles include Augur, ChainLink and Oraclize
\begin{itemize}
	\item \href{https://www.augur.net/}{Augur}
	\item \href{https://chain.link/}{ChainLink}
	\item \href{http://www.oraclize.it}{Oraclize}
\end{itemize}

\subsection{Augur}

\begin{quote}
    A decentralized oracle and prediction market platform.
\end{quote}

The Augur team have programmed complex smart-contracts on the Ethereum network
that users can use to create markets and select oracles.
The consensus building process for finalizing an oracle response is quite interesting
and involves the staking of Augur’s native ERC20 token called REP (“reputation”).
However, there are challenges and most of them are related to scalability.

Every market that is created using Augur software utilizes a smart-contract on its own.
This means that if there are 1,000 markets, there will be 1,000 smart-contracts
running on the Ethereum network only for the Augur use-case.
This could create scalability issues.
We have already witnessed the impact that CryptoKitties and popular token sales
have had on Ethereum.

The various dispute resolution rounds (1) will require transactions of REP,
which are more costly than ETH transactions.
Keeping in mind that there are also limits to the duration of dispute rounds,
the execution of REP transactions might be expensive during peak on-chain load-times
and/or may fail to execute in time.

These issues relate to the fact that Augur introduces a consensus protocol
on top of another consensus protocol – that of the Ethereum network.
This makes the market consensus dependent on the blockchain consensus
and creates inefficiencies that may jeopardize the business model of Augur’s oracles
and prediction markets.

\subsubsection{Augur Summary}

Basically, it's Ethereum-specific smart contract, not suitable for Cardano.

\subsection{ChainLink}

\begin{quote}
A Decentralized Oracle Network
\end{quote}

From thier \href{https://link.smartcontract.com/whitepaper}{Whitepaper}

As an oracle service, ChainLink nodes return replies to data requests or queries made
by or on behalf of a user contract, which we refer to as requesting contracts and
denote by USER-SC. ChainLink’s on-chain interface to requesting contracts is itself
an on-chain contract that we denote by CHAINLINK-SC.

Behind CHAINLINK-SC, ChainLink has an on-chain component consisting of three
main contracts: a \emph{reputation contract}, an \emph{order-matching contract},
and an \emph{aggregating contract}.

The reputation contract keeps track of oracle-service-provider performance
metrics.

The order-matching smart contract takes a proposed service level agreement,
logs the SLA parameters, and collects bids from oracle providers. It then selects bids
using the reputation contract and finalizes the oracle SLA.

The aggregating contract collects the oracle providers’ responses
and calculates the final collective result of
the ChainLink query. It also feeds oracle provider metrics back into the reputation
contract.

\subsection{Oraclize}

\section{Summary}

Summary.

\end{document}

### 2019-11-27 papers/marlowe2019/paper.tex
Added skeleton in latex. Sections included.

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}

\begin{document}
%
\title{Marlowe: implmenting and analysing financial contracts\thanks{Supported by IOHK.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{
Pablo Lamela Seijas\inst{1} \and
Alex Nemish\inst{1} \and
David Smith\inst{1} \and
Simon Thompson\inst{1,2}\orcidID{0000-0002-2350-301X}}
%
\authorrunning{P. L. Seijas, A. Nemish, et al.}

% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{IOHK, Hong Kong, \url{https://iohk.io} \\
\email{\{alexander.nemish, pablo.lamela, david.smith, simon.thompson\}@iohk.io} \and
School of Computing, University of Kent, UK\\
\email{s.j.thompson@kent.ac.uk}}
%
\maketitle              % typeset the header of the contribution
%

\begin{abstract}
The abstract should briefly summarize the contents of the paper in
150--250 words.

\keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%

\section{Introduction}

Intro
\section{Marlowe overview SJT}

Introduce the main types and informal overview of execution. Contrast with Marlowe 1.3 (as in ISoLA paper). Embedding in Haskell? [All this stuff is in the tutorial already]

\section{Implementation of Marlowe on the mockchain AN}

 [An updated version of the description in the Plutus platform paper]

\section{Analysing Marlowe contracts PLS}

Covers static analysis and Isabelle proofs [New]

\section{The Marlowe Playground DS}

[There’s some stuff in the tutorial]

\section{Related work}

Need to update from ISoLA paper

\section{Future work and conclusions}

\section*{Left in for info \ldots}

\begin{figure}
\includegraphics[width=\textwidth]{fig1.eps}
\caption{A figure caption is always placed below the illustration.
Please note that short captions are centered, while long ones are
justified by the macro package automatically.} \label{fig1}
\end{figure}

\subsection*{Bibliography \ldots}

We'll use a .bib file here and Bibtex.

For citations of references, we prefer the use of square brackets
and consecutive numbers. Citations using labels or the author/year
convention are also acceptable. The following bibliography provides
a sample reference list with entries for journal
articles~\cite{ref_article1}, an LNCS chapter~\cite{ref_lncs1}, a
book~\cite{ref_book1}, proceedings without editors~\cite{ref_proc1},
and a homepage~\cite{ref_url1}. Multiple citations are grouped
\cite{ref_article1,ref_lncs1,ref_book1},
\cite{ref_article1,ref_book1,ref_proc1,ref_url1}.
%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
% \bibliographystyle{splncs04}
% \bibliography{mybibliography}
%
\begin{thebibliography}{8}
\bibitem{ref_article1}
Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)

\bibitem{ref_lncs1}
Author, F., Author, S.: Title of a proceedings paper. In: Editor,
F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
Springer, Heidelberg (2016). \doi{10.10007/1234567890}

\bibitem{ref_book1}
Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
Location (1999)

\bibitem{ref_proc1}
Author, A.-B.: Contribution title. In: 9th International Proceedings
on Proceedings, pp. 1--2. Publisher, Location (2010)

\bibitem{ref_url1}
LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
Oct 2017
\end{thebibliography}
\end{document}

### 2019-11-28 papers/marlowe2019/paper.tex
Belated commit

@@
 
-[There’s some stuff in the tutorial]
+[There’s some stuff on this in the tutorial.]
 
@@
 
-Need to update from ISoLA paper
+Need to update from ISoLA paper \cite{isola-marlowe}.
 
@@
 
-\subsection*{Bibliography \ldots}
-
-We'll use a .bib file here and Bibtex.
-
-For citations of references, we prefer the use of square brackets
-and consecutive numbers. Citations using labels or the author/year
-convention are also acceptable. The following bibliography provides
-a sample reference list with entries for journal
-articles~\cite{ref_article1}, an LNCS chapter~\cite{ref_lncs1}, a
-book~\cite{ref_book1}, proceedings without editors~\cite{ref_proc1},
-and a homepage~\cite{ref_url1}. Multiple citations are grouped
-\cite{ref_article1,ref_lncs1,ref_book1},
-\cite{ref_article1,ref_book1,ref_proc1,ref_url1}.
 %
@@
 %
-% \bibliographystyle{splncs04}
-% \bibliography{mybibliography}
+\bibliographystyle{splncs04}
+\bibliography{paper}
 %
-\begin{thebibliography}{8}
-\bibitem{ref_article1}
-Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)
-
-\bibitem{ref_lncs1}
-Author, F., Author, S.: Title of a proceedings paper. In: Editor,
-F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
-Springer, Heidelberg (2016). \doi{10.10007/1234567890}
-
-\bibitem{ref_book1}
-Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
-Location (1999)
-
-\bibitem{ref_proc1}
-Author, A.-B.: Contribution title. In: 9th International Proceedings
-on Proceedings, pp. 1--2. Publisher, Location (2010)
-
-\bibitem{ref_url1}
-LNCS Homepage, \url{http://www.springer.com/lncs}. Last accessed 4
-Oct 2017
-\end{thebibliography}
 \end{document}

### 2019-11-28 papers/marlowe2019/paper.tex
Add Makefile and change permissions
Mis-spelled title

@@
 %
-\title{Marlowe: implmenting and analysing financial contracts\thanks{Supported by IOHK.}}
+\title{Marlowe: implementing and analysing financial contracts\thanks{Supported by IOHK.}}
 %
@@
 
-[There’s some stuff on this in the tutorial.]
+[There's some stuff on this in the tutorial.]
 
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{fig1.eps}
-\caption{A figure caption is always placed below the illustration.
-Please note that short captions are centered, while long ones are
-justified by the macro package automatically.} \label{fig1}
-\end{figure}
+%\begin{figure}
+%\includegraphics[width=\textwidth]{fig1.eps}
+%\caption{A figure caption is always placed below the illustration.
+%Please note that short captions are centered, while long ones are
+%justified by the macro package automatically.} \label{fig1}
+%\end{figure}
 

### 2019-11-29 papers/marlowe2019/paper.tex
add playground section to 2019 paper

--- 2019-11-28 papers/marlowe2019/paper.tex
@@
 \usepackage{graphicx}
+\usepackage{listings}
+\usepackage{wrapfig}
 % Used for displaying a sample figure. If possible, figure files should
@@
 
-\section{The Marlowe Playground DS}
+\clearpage
+\section{The Marlowe Playground}
 
-[There’s some stuff on this in the tutorial.]
+For Marlowe to be usable in practice, users need to be able to understand how contracts will behave once deployed to the blockchain before actually deploying the contract.
+We can do that by simulating their behavior with a mockchain and interactively stepping through the evaluation of a contract in a browser.
 
+To achieve this, and to aid Marlowe’s take-up usage by people who are not familiar with its syntax, we have developed the Marlowe Playground, a web tool that supports the interactive construction, revision, and simulation of smart-contracts written in Marlowe.
+The tool is publicly available in the url: https://prod.meadow.marlowe.iohkdev.io/.
+It should be mentioned that development of the playground is rapid and the latest unstable version is also publicly available at https://alpha.marlowe.iohkdev.io/.
+
+The playground has an editor where you can write a contract in "pure" Marlowe, that is, not embedded in a host language. The reason for doing this is two-fold:
+\begin{itemize}
+    \item There is a shallower learning curve for users who are new to Haskell or programming languages in general.
+The Marlowe constructs are quite simple so there is no need to learn about Haskell syntax or even variables, functions etc.
+    \item As we step through the execution of a contract, the contract is reduced, it is very useful to be able to view and even edit the reduced contract during this execution.
+\end{itemize}
+This is probably not the best way to write a large contract so the playground contains an additional editor where contracts can be written using Haskell and then transferred as a pure Marlowe data structure into the Marlowe editor.
+
+In addition to pure Marlowe and Haskell editors, contracts can also be written using Google's Blockly visual programming language.
+Again this is not suitable for writing large contracts however it is an easy way to introduce the concepts of Marlowe to users who have no programming knowledge.
+Once a contract has been 'written' in Blockly it is possible to transfer that contract to the Marlowe editor.
+It is also possible to transfer a Marlowe contract to Blockly for further editing.
+
+The Marlowe editor in the unstable version of the playground has an additional feature to aid writing contracts called holes.
+If we enter the contract \lstinline{?mycontract} we will be presented with a dropdown list of values that could be used.
+
+\begin{wrapfigure}[5]{l}{0.25\textwidth}
+    \vspace*{-0.3in}
+    \includegraphics[width=0.25\textwidth]{hole_options.png}
+\end{wrapfigure}
+In our case \lstinline{?mycontract} must be a \lstinline{Contract} of some sort so we can choose one of the \lstinline{Contract} constructors from the dropdown list.
+If we choose \lstinline{Pay} then the Marlowe editor will automatically fill in a skeleton \lstinline{Pay} contract with new holes where we need to provide values.
+\\ \\
+\begin{verbatim}
+    Pay ?accountId_1_1 ?payee_1_2 ?value_1_3 ?contract_1_4
+\end{verbatim}
+New options will be presented, one for each hole, and each will have a dropdown list of all the possible values.
+\begin{wrapfigure}[6]{l}{0.25\textwidth}
+    \vspace*{-0.3in}
+    \includegraphics[scale=0.2]{hole_options_2.png}
+\end{wrapfigure}
+A complete contract can be written in this guided way with the user needing only to fill in strings and numbers by hand.
+This approach to writing holes in your code and "asking" the compiler what you could put in there is easy to implement in a DSL as there are very limited options however is is becoming popular with more complex languages such as Haskell and Idris.
+
+Contracts written in the Marlowe editor are parsed in real-time and if there are no errors (and no holes) then the contract is analyzed to discover which actions a user could take to progress the contract.
+These actions are displayed in the "Input Composer" above the editor.
+Lets take the following example contract:
+\begin{verbatim}
+    When [Case (Deposit (AccountId 0 "investor") "guarantor" (Constant 1000)) 
+    Close] 10 Close
+\end{verbatim}
+In this case, the only action a user can take to progress the contract is to deposit 1000 ADA from the guarantor to the investor's account.
+The playground knows this and so the input composer displays this action in the input composer.
+
+\begin{figure}[]
+    \includegraphics[width=1\textwidth]{input_composer.png}
+\end{figure}
+The user can then choose to add this action to a transaction, ready to be applied. Multiple actions can be added to a transaction before it is applied if such actions are meaningful.
+
+\begin{figure}[]
+    \includegraphics[width=0.8\textwidth]{tx_composer.png}
+\end{figure}
+A user can then apply this transaction and in the example above this would result in the state pane showing a single payment and in addition the Contract in the Marlowe editor will have been reduced to \lstinline{Close}.
+At this point a user can undo the application of the transaction or even reset the contract to it's initial state.
+This enables a user to apply transactions, see what the effect is, step back and apply a transaction with different actions and see how these differences effect the end results.
+They can also change the reduced contract to investigate a different path.
+
+Users can at any point save their contract directly to a Github Gist and of course load a contract from a Github Gist.
+There are also some demo contracts that can be loaded in order to play around with realistic examples.
+
+The final feature that we would like to present is the static analysis of contracts.
+As mentioned in a previous section, it is possible to carry out a symbolic execution of a contract and then use a SMT solver to look for cases that could cause unwanted situations.
+The playground utilizes this to search for situations where contract execution would cause warnings.
+For example, suppose you write a contract that causes a payment of 450 Lovelace from Alice to Bob but the contract allows a situation where Alice has only deposited 350 Lovelace then the static analysis will find this and report it to the playground user.
+\begin{figure}
+    \includegraphics[width=1\textwidth]{static_analysis.png}
+\end{figure}
+\clearpage
 \section{Related work}
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{fig1.eps}
+% \includegraphics[width=\textwidth]{fig1.eps}
 \caption{A figure caption is always placed below the illustration.

### 2019-12-02 papers/marlowe2019/paper.tex
Add first draft about static analysis

@@
 \author{
-Pablo Lamela Seijas\inst{1} \and
+Pablo {Lamela Seijas}\inst{1} \and
 Alex Nemish\inst{1} \and
@@
 %
-\authorrunning{P. L. Seijas, A. Nemish, et al.}
+\authorrunning{P. Lamela Seijas, A. Nemish, et al.}
 
@@
 
-\section{Analysing Marlowe contracts PLS}
+\section{Analysing Marlowe\label{sec:analysis}}
 
-Covers static analysis and Isabelle proofs [New]
+In this section, we give an overview of some of the work we have carried out with the aim of helping users be more confident in that the Marlowe contracts they write do what they expect them to do.
+
+We look at two complementary approaches: static analysis of concrete contracts (Section~\ref{subsec:static}) and formal verification of general properties about Marlowe (Section~\ref{subsec:verification}).
+
+\subsection{Static analysis of contracts\label{subsec:static}}
+
+For some properties, and given a particular contract, a computer can decidably determine whether the contract satisfies the property and, in the cases when it does not, it can provide a counter example.
+
+For turing-complete languages, this process is often undecidable in the general case.
+And for Marlowe contracts it is often NP-complete in the general case.
+However, in practice, SAT solvers are often able to solve moderately large instances of this problem reasonably efficiently.
+
+Our implementation relies on the Haskell library SBV, which in turn relies on existing SAT solvers for proving satisfiability of properties or finding counter examples.
+
+\subsubsection{SBV library}
+
+SBV (SMT Base Verificiation) library provides a high-level API that allows developers to automatically prove properties about Haskell programs, among other functionalities.
+
+The SBV library seamlessly translates these property to SMTLib queries, passes them to one or several SAT solvers, and translates the results back to the format in which the queries were written.
+
+\paragraph{\texttt{SBV} monad.}
+
+SBV provides a monad called \texttt{SBV} which encapsulates potentially unknown values to the program.
+A program that takes arguments wrapped with the \texttt{SBV} monad can be used with normal arguments and it can also be passed symbolic parameters when used as part of a property that is passed to the solver.
+
+For example, the following function:
+
+\begin{verbatim}
+primeFactors :: SBV Integer -> SBV Integer -> SBV Bool
+primeFactors x y = x .> 1 .&& y .> 1 .=> x * y ./= 25723297
+\end{verbatim}
+
+can be used both with normal integers and with symbolic parameters:
+
+\begin{verbatim}
+*Main> primeFactors 98 314
+True
+*Main> prove primeFactors
+Falsifiable. Counter-example:
+    s0 = 6353 :: Integer
+    s1 = 4049 :: Integer
+*Main> primeFactors 6353 4049
+False
+\end{verbatim}
+
+\paragraph{Limitations}
+
+At the time of writing, there are some types of functions and data-types that SBV cannot translate to \texttt{SMTLib}.
+In particular, there are two that affected the way in which we use SBV with Marlowe and that prevent us from using the same version of the semantics for symbolic execution and normal execution.
+
+Firstly, recursive functions that are unbound cannot be translated.
+As part of the process of converting functions to \texttt{SMTLib} terms, SBV unfolds them into a finite data-structure.
+For this reason, if the function is recursive and is not bounded by a non-symbolic value (a value that is known at the time of translation), the translation process will run out of memory.
+
+Secondly, to the best our knowledge, SBV does not currently provide support for custom data types directly (except for non-ary datatypes, i.e: enumerations).
+
+In the following section, we explain how we worked around these limitations and overview how we use SBV to avoid runtime problems with Marlowe contracts.
+
+\subsubsection{Using SBV to analyse Marlowe Contracts}
+
+Marlowe semantics are designed to use types to prevent many non-sensical contracts from being written.
+But there are potential problems which are harder to notice until runtime (e.g: whether there will be enough money to issue all the payments declared in the contract), at which point it may already be too late to fix them (in the case of blockchain).
+Marlowe semantics represents these errors that can be found at runtime as \texttt{Warnings}.
+
+The property that we have implemented using SBV library can be enunciated as: "the given contract will not need to issue warnings in runtime no matter the inputs it receives".
+
+In order to achieve this, we implemented a symbolic version of the semantics that returns a list of the warnings produced by a trace (a list of transactions input to the contract):
+
+\begin{verbatim}
+warningsTraceWB :: Bounds -> SSlotNumber -> SList NTransaction
+                -> Contract -> SList NTransactionWarning
+\end{verbatim}
+
+Where types that start with \texttt{S} like \texttt{SSlotNumber} are abreviations for the symbolic versions of types (in this case \texttt{SBV SlotNumber}).
+
+The types that start with \texttt{N} are \textit{nested} types, which we explain later in the \textit{Custom datatypes} section below.
+
+We can see that this version of the implementation takes two concrete parameters: \texttt{Bounds} (with info about the bounds of the contract) and \texttt{Contract} (the contract we want to analyse); and two symbolic parameters: \texttt{SSlotNumber} (the initial slot number) and \texttt{SList NTransaction} (the input list of transactions); and it returns \texttt{SList NTransactionWarning} (a symbolic list of transaction warnings).
+
+\paragraph{Custom datatypes}
+
+As we mentioned before, SBV does not currently seem to support in general the use of custom datatypes.
+We do not need to represent the \texttt{Contract} type symbolically because we know its value at the time of analysis, but we still have many other types like \texttt{TransactionWarning} that are custom datatypes and we need to use.
+
+Fortunately, SBV does support tuples and the \texttt{Either} type; combinations of those two give us essentially the same functionality as custom datatypes that are not recursive and have no type-parameters.
+We can represent all types that Marlowe requires as combinations of \texttt{Either} and tuples, with the exception of the \texttt{Contract} type, but we do not need a symbolic version of the \texttt{Contract} type.
+
+For example, the TransactionResult type:
+\begin{verbatim}
+data TransactionResult = TransactionProcessed [TransactionWarning]
+                                              [TransactionEffect]
+                                              State
+                       | TransactionError TransactionError
+\end{verbatim}
+
+becomes the nested type synonym \texttt{NTransactionResult}:
+
+\begin{verbatim}
+type NTransactionResult =
+      Either ([NTransactionWarning], [NTransactionEffect], NState)
+             NTransactionError
+\end{verbatim}
+
+Because working with nested types is much more error prone than working with the original data-types, we used template haskell to implement functions that transform the custom datatypes into nested types and generates the appropriate conversion functions.
+
+\paragraph{Returning non-symbolic \texttt{Contract} values}
+
+In the original semantics, each step typically takes a \texttt{Contract}, a \texttt{State}, and a set of \texttt{Input}s; and it returns an updated \texttt{State} and a remaining \texttt{Contract}.
+This is a problem because values that rely on symbolic values, have to be themselves symbolic.
+And the resulting remaining \texttt{Contract} depends on the \texttt{Input}s and \texttt{State} which are both symbolic.
+
+We work around this problem by slightly modifying the signature of the function to receive a \textit{continuation function} instead; and instead of just returning a value, we return the result of applying the \textit{continuation function} to the result we were planning to return.
+
+For example, the original type signature for the \texttt{apply} function was:
+
+\begin{verbatim}
+apply :: Environment -> State -> Input -> Contract -> ApplyResult
+\end{verbatim}
+
+But the symbolic version of the \texttt{apply} function has the following signature:
+
+\begin{verbatim}
+apply :: SymVal a => Bounds -> SEnvironment -> SState -> SInput
+      -> Contract -> (SApplyResult -> DetApplyResult -> SBV a)
+      -> SBV a
+\end{verbatim}
+
+Where \texttt{DetApplyResult} contains the parts of \texttt{ApplyResult} that are not symbolic (like the \texttt{Contract}).
+
+\paragraph{Bounds for the state and the inputs}
+
+The recursion in the execution of the semantics is bounded by the \texttt{Contract}, and because the \texttt{Contract} is not a symbolic parameter, the translation will terminate.
+
+However, both in the input and in the \texttt{State} record there are several lists (or association maps) that are not explicitly bounded in the implementation.
+Some parts of the semantics are bounded by the length of these lists (or maps), like the implementation of \texttt{Close} that refund all the money available in all the accounts.
+
+In order for the symbolic implementation to be finite we need to find a bound for the length of these lists and maps.
+
+With the current semantics of Marlowe all of the lists/maps in State and input can be bound when looking at a particular contract.
+In particular, we can find out quite easily, by looking at the contract, how many parties there are, how many choice identifiers are used, how many accounts are used, and how many let binding identifiers are used.
+
+It is less obvious that we can also set a bound for the input \texttt{Transaction} list as the maximum number of nested \texttt{When}s in the contract (for all the execution paths) plus one.
+We also provide a proof in Isabelle that shows that any longer list of transactions will result in an error, Since transactions that result on errors do not modify the \texttt{State} or the \texttt{Contract}, we know that limiting the analysis to lists of transactions that are shorter than the bound does not lose generality.
+
+\subsection{Formal verification of semantics\label{subsec:verification}}
+
+We can also use proof assistants like Isabelle to demonstrate that Marlowe semantics present certain desirable properties (like that money is preserved and returned to users by the end of the execution of any contract).
+In the same way, we can also use this technique to prove properties for particular contracts, or for particular contract snippets, templates, or classes of contracts.
 

### 2019-12-03 papers/marlowe2019/paper.tex
Editing of section 4.

--- 2019-11-29 papers/marlowe2019/paper.tex
@@
 
-\section{Analysing Marlowe contracts PLS}
+\section{Analysing Marlowe\label{sec:analysis}}
 
-Covers static analysis and Isabelle proofs [New]
+In this section, we give an overview of some of the work we have carried out with the aim of helping users be more confident that the Marlowe contracts they write do what they expect them to do.
+
+We look at two complementary approaches: static analysis of concrete contracts (Section~\ref{subsec:static}) and formal verification of general properties about Marlowe (Section~\ref{subsec:verification}).
+
+\subsection{Static analysis of contracts\label{subsec:static}}
+
+For some properties, given a particular contract, a computer can decidably determine whether the contract satisfies the property and, in the cases when it does not, it can provide a counter-example.
+
+For Turing-complete languages, this process is usually undecidable in the general case, and for Marlowe contracts it is often NP-complete in general.
+However, in practice, SAT and SMT (Satisfaction Modulo Theories) solvers are often able to solve moderately large instances of this problem reasonably efficiently.
+
+Our implementation relies on the Haskell library SBV, which in turn relies on existing SAT and SMT solvers to check satisfiability of properties or finding counter-examples.
+
+\subsubsection{SBV library}
+
+SBV (SMT Base Verification) library provides a high-level API that allows developers to automatically prove properties about Haskell programs, among other functionalities. 
+The SBV library %seamlessly 
+translates these properties to SMTLib queries, passes them to one or several SAT/SMT solvers, and translates the results back to the format in which the queries were written.
+
+\paragraph{\texttt{SBV} monad.}
+
+SBV provides a monad called \texttt{SBV} which encapsulates potentially unknown values to the program.
+A program that takes arguments wrapped with the \texttt{SBV} monad can be used with normal arguments and it can also be passed symbolic parameters when used as part of a property that is passed to the solver.
+For example, the following function:
+
+\begin{verbatim}
+primeFactors :: SBV Integer -> SBV Integer -> SBV Bool
+primeFactors x y = x .> 1 .&& y .> 1 .=> x * y ./= 25723297
+\end{verbatim}
+
+\noindent
+can be used both with normal integers and with symbolic parameters:
+%SJT: would be usweful to say something more about what symbolic parameters mean here: it's essentially that these are parameters that are universally quantified and passed to the solver, I think?
+They are instantiated with counter-examples if those exist.
+\begin{verbatim}
+*Main> primeFactors 98 314
+True
+*Main> prove primeFactors
+Falsifiable. Counter-example:
+    s0 = 6353 :: Integer
+    s1 = 4049 :: Integer
+*Main> primeFactors 6353 4049
+False
+\end{verbatim}
+
+\paragraph{Limitations}
+
+At the time of writing, there are some types of functions and data-types that SBV cannot translate to \texttt{SMTLib}.
+In particular, there are two that affected the way in which we use SBV with Marlowe and that prevent us from using the same version of the semantics for symbolic execution and normal execution.
+
+Firstly, calls to recursive functions that are unbounded cannot be translated.
+As part of the process of converting functions to \texttt{SMTLib} terms, SBV unfolds them into a finite data-structure.
+For this reason, if the function is recursive and is not bounded by a non-symbolic value (a value that is known at the time of translation), the translation process will run out of memory.
+
+Secondly, to the best our knowledge, SBV does not currently provide support for custom data types directly, except for zero-ary datatypes, i.e: enumerations.
+%SJT: i.e. there's no symbolic representation of user defined data types in general?
+
+In the following section, we explain how we worked around these limitations and give an overview of how we use SBV to avoid runtime problems with Marlowe contracts.
+
+\subsubsection{Using SBV to analyse Marlowe Contracts}
+
+Marlowe semantics are designed to use types to prevent many non-sensical contracts from being written.
+But there are potential problems which are harder to detect until runtime, for example, whether there will be enough money to issue all the payments declared in the contract.
+At this point it may already be too late to fix them, and particularly so in the case of blockchain.
+The Marlowe semantics represents these errors that can be found at runtime as \texttt{Warnings}.
+
+The property that we have implemented using SBV library can be enunciated as: ``the given contract will not need to issue warnings at runtime no matter the inputs it receives''.
+
+In order to achieve this, we implemented a symbolic version of the semantics that returns a list of the warnings produced by a trace (a list of transactions input to the contract):
+
+\begin{verbatim}
+warningsTraceWB :: Bounds -> SSlotNumber -> SList NTransaction
+                -> Contract -> SList NTransactionWarning
+\end{verbatim}
+
+\noindent
+where types that begin with \texttt{S}, like \texttt{SSlotNumber}, are abbreviations for the symbolic versions of types: in this case \texttt{SBV SlotNumber}.
+The types that begin with \texttt{N} are \textit{nested} types, which we explain in the \textit{Custom datatypes} section below.
+
+We can see that this version of the implementation takes two concrete parameters: 
+%SJT: need to explain a little more about what the bounds are
+\texttt{Bounds} (with info about the bounds of the contract) and \texttt{Contract} (the contract we want to analyse); and two symbolic parameters: \texttt{SSlotNumber} (the initial slot number) and \texttt{SList NTransaction} (the input list of transactions); and it returns \texttt{SList NTransactionWarning} (a symbolic list of transaction warnings).
+
+\paragraph{Custom datatypes}
+
+As we mentioned before, SBV does not currently seem to support in general the use of custom datatypes.
+We do not need to represent the \texttt{Contract} type symbolically because we know its value at the time of analysis, but we still have a number of other types like \texttt{TransactionWarning} that are custom datatypes and we need to use.
+
+Fortunately, SBV does support tuples and the \texttt{Either} type; combinations of those two give us essentially the same functionality as custom datatypes that are not recursive and have no type-parameters.
+We can represent all types that Marlowe requires as combinations of \texttt{Either} and tuples, with the exception of the \texttt{Contract} type, but we do not need a symbolic version of the \texttt{Contract} type.
+
+\noindent
+For example, the \texttt{TransactionResult} type:
+\begin{verbatim}
+data TransactionResult
+ = TransactionProcessed [TransactionWarning]
+                        [TransactionEffect]
+                        State
+ | TransactionError TransactionError
+\end{verbatim}
+\noindent
+becomes the nested type synonym \texttt{NTransactionResult}:
+
+\begin{verbatim}
+type NTransactionResult =
+  Either ([NTransactionWarning], [NTransactionEffect], NState)
+         NTransactionError
+\end{verbatim}
+
+\noindent
+Because working with nested types is much more error prone than working with the original data-types, we used Template Haskell % SJT: nice! Need to add reference.
+to implement functions that transform the custom datatypes into nested types and generate the appropriate conversion functions.
+
+\paragraph{Returning non-symbolic \texttt{Contract} values}
+
+In the non-symbolic semantics, each step typically takes a \texttt{Contract}, a \texttt{State}, and a set of \texttt{Input}s; and it returns an updated \texttt{State} and a remaining \texttt{Contract}.
+%SJT: a thought: could just give the type here, or cross-refer to an earlier section.
+This is a problem because values that rely on symbolic values have to be themselves symbolic.
+Moreover,  the resulting remaining \texttt{Contract} depends on the \texttt{Input}s and \texttt{State}, which are both symbolic.
+
+We work around this problem by  modifying the signature of the function to receive a \textit{continuation function} instead, and instead of just returning a value, we return the result of applying the \textit{continuation function} to the result we were planning to return.
+
+For example, the original type signature for the \texttt{apply} function was:
+
+\begin{verbatim}
+apply :: Environment -> State -> Input -> Contract -> ApplyResult
+\end{verbatim}
+
+\noindent
+and the symbolic version of the \texttt{apply} function has the following signature:
+
+\begin{verbatim}
+apply :: SymVal a => Bounds -> SEnvironment -> SState -> SInput
+      -> Contract -> (SApplyResult -> DetApplyResult -> SBV a)
+      -> SBV a
+\end{verbatim}
+
+\noindent
+where \texttt{DetApplyResult} contains the parts of \texttt{ApplyResult} that are not symbolic (like the \texttt{Contract}).
+
+\paragraph{Bounds for the state and the inputs}
+
+The recursion in the execution of the semantics is bounded by the \texttt{Contract}, and because the \texttt{Contract} is not a symbolic parameter, the translation will terminate.
+
+However, in both the input and the \texttt{State} record there are several lists (representing finite maps) that are not explicitly bounded in the implementation.
+Some parts of the semantics are bounded by the length of these lists (or maps), such as the implementation of \texttt{Close} that refunds any remaining money in the contract accounts.
+In order for the symbolic implementation to be finite we need to find a bound for the length of these lists and maps.
+
+With the current semantics of Marlowe all the lists/maps in \texttt{State} and input can be bound when looking at a particular contract.
+In particular, we can find out quite easily, by looking at the contract, how many parties there are, how many choice identifiers are used, how many accounts are used, and how many let binding identifiers are used.
+
+It is less obvious that we can also set a bound for the input \texttt{Transaction} list as the maximum number of nested \texttt{When}s in the contract (for all the execution paths) plus one.
+We also provide a proof in Isabelle that shows that any longer list of transactions will result in an error.
+%SJT: have I rephrased this right?
+That is because any transactions that result in errors do not modify the \texttt{State} or the \texttt{Contract}, we know that limiting the analysis to lists of transactions that are shorter than the bound does not lose generality.
+
+\subsection{Formal verification of semantics\label{subsec:verification}}
+
+We can also use proof assistants like Isabelle to demonstrate that Marlowe semantics present certain desirable properties (like that money is preserved and returned to users by the end of the execution of any contract).
+In the same way, we can also use this technique to prove properties for particular contracts, or for particular contract snippets, templates, or classes of contracts.
 

### 2019-12-03 papers/marlowe2019/paper.tex
changed continuation type layout
Fix some typos and address SJT comments

@@
 For Turing-complete languages, this process is usually undecidable in the general case, and for Marlowe contracts it is often NP-complete in general.
-However, in practice, SAT and SMT (Satisfaction Modulo Theories) solvers are often able to solve moderately large instances of this problem reasonably efficiently.
+However, in practice, SMT (Satisfaction Modulo Theories) solvers are often able to solve moderately large instances of this problem reasonably efficiently.
 
-Our implementation relies on the Haskell library SBV, which in turn relies on existing SAT and SMT solvers to check satisfiability of properties or finding counter-examples.
+Our implementation relies on the Haskell library SBV, which in turn relies on existing SMT solvers to check satisfiability of properties or finding counter-examples.
 
@@
 
-SBV (SMT Base Verification) library provides a high-level API that allows developers to automatically prove properties about Haskell programs, among other functionalities. 
-The SBV library %seamlessly 
-translates these properties to SMTLib queries, passes them to one or several SAT/SMT solvers, and translates the results back to the format in which the queries were written.
+SBV \cite{SBV} (SMT Base Verification) library provides a high-level API that allows developers to automatically prove properties about Haskell programs, among other functionalities. 
+The SBV library translates these properties to SMTLib queries, passes them to one or several SMT solvers, and translates the results back to the format in which the queries were written.
 
@@
 \noindent
-can be used both with normal integers and with symbolic parameters:
-%SJT: would be usweful to say something more about what symbolic parameters mean here: it's essentially that these are parameters that are universally quantified and passed to the solver, I think?
-They are instantiated with counter-examples if those exist.
+can be used both with normal integers and with symbolic parameters, that is, parameters whose concrete value is not specified and which the solver will try to replace with values that satisfy or falsify the property:
+
 \begin{verbatim}
@@
 
-Secondly, to the best our knowledge, SBV does not currently provide support for custom data types directly, except for zero-ary datatypes, i.e: enumerations.
-%SJT: i.e. there's no symbolic representation of user defined data types in general?
+Secondly, to the best our knowledge, SBV does not currently directly provide support for representing custom data types symbolically, except for zero-ary datatypes, i.e: enumerations.
 
@@
 
-We can see that this version of the implementation takes two concrete parameters: 
-%SJT: need to explain a little more about what the bounds are
-\texttt{Bounds} (with info about the bounds of the contract) and \texttt{Contract} (the contract we want to analyse); and two symbolic parameters: \texttt{SSlotNumber} (the initial slot number) and \texttt{SList NTransaction} (the input list of transactions); and it returns \texttt{SList NTransactionWarning} (a symbolic list of transaction warnings).
+We can see that this version of the implementation takes two concrete parameters: \texttt{Bounds} (see \textit{Bounds for the state and the inputs} section below) and \texttt{Contract} (the contract we want to analyse); and two symbolic parameters: \texttt{SSlotNumber} (the initial slot number) and \texttt{SList NTransaction} (the input list of transactions); and it returns \texttt{SList NTransactionWarning} (a symbolic list of transaction warnings).
 
@@
 \noindent
-Because working with nested types is much more error prone than working with the original data-types, we used Template Haskell % SJT: nice! Need to add reference.
-to implement functions that transform the custom datatypes into nested types and generate the appropriate conversion functions.
+Because working with nested types is much more error prone than working with the original data-types, we used Template Haskell \cite{sheard2002template} to implement functions that transform the custom datatypes into nested types and generate the appropriate conversion functions.
 
@@
 In the non-symbolic semantics, each step typically takes a \texttt{Contract}, a \texttt{State}, and a set of \texttt{Input}s; and it returns an updated \texttt{State} and a remaining \texttt{Contract}.
-%SJT: a thought: could just give the type here, or cross-refer to an earlier section.
+
 This is a problem because values that rely on symbolic values have to be themselves symbolic.
@@
 \begin{verbatim}
-apply :: SymVal a => Bounds -> SEnvironment -> SState -> SInput
-      -> Contract -> (SApplyResult -> DetApplyResult -> SBV a)
-      -> SBV a
+apply :: SymVal a => Bounds -> 
+      SEnvironment -> SState -> SInput -> Contract -> 
+      (SApplyResult -> DetApplyResult -> SBV a) -> SBV a
 \end{verbatim}
@@
 Some parts of the semantics are bounded by the length of these lists (or maps), such as the implementation of \texttt{Close} that refunds any remaining money in the contract accounts.
-In order for the symbolic implementation to be finite we need to find a bound for the length of these lists and maps.
+In order for the symbolic implementation to be finite, we need to find a bound for the length of these lists and maps.
 
-With the current semantics of Marlowe all the lists/maps in \texttt{State} and input can be bound when looking at a particular contract.
+With the current semantics of Marlowe, all the lists/maps in \texttt{State} and input can be bounded when looking at a particular contract.
 In particular, we can find out quite easily, by looking at the contract, how many parties there are, how many choice identifiers are used, how many accounts are used, and how many let binding identifiers are used.
@@
 It is less obvious that we can also set a bound for the input \texttt{Transaction} list as the maximum number of nested \texttt{When}s in the contract (for all the execution paths) plus one.
-We also provide a proof in Isabelle that shows that any longer list of transactions will result in an error.
-%SJT: have I rephrased this right?
-That is because any transactions that result in errors do not modify the \texttt{State} or the \texttt{Contract}, we know that limiting the analysis to lists of transactions that are shorter than the bound does not lose generality.
+We also provide a proof in Isabelle that shows that any longer list of transactions will result in an error (we discuss this in Section~\ref{subsubsec:bound_max_transaction_number}).
 
@@
 In the same way, we can also use this technique to prove properties for particular contracts, or for particular contract snippets, templates, or classes of contracts.
+
+\subsubsection{Bound on the maximum number of transactions\label{subsubsec:bound_max_transaction_number}}
+
+\emph{ToDo}
+Any transactions that result in errors do not modify the \texttt{State} or the \texttt{Contract}, we know that limiting the analysis to lists of transactions that are shorter than the bound does not lose generality.
 

### 2019-12-03 papers/marlowe2019/paper.tex
Small layout adjustment
Revised Marlowe Playground section.

--- 2019-12-03 papers/marlowe2019/paper.tex
@@
 \usepackage{wrapfig}
+\usepackage{hyperref}
+
+% redefined to give listings in tt font
+\renewcommand{\lstinline}[1]{\texttt{#1}}
+
 % Used for displaying a sample figure. If possible, figure files should
@@
 
-\clearpage
+%\clearpage
 \section{The Marlowe Playground}
 
-For Marlowe to be usable in practice, users need to be able to understand how contracts will behave once deployed to the blockchain before actually deploying the contract.
-We can do that by simulating their behavior with a mockchain and interactively stepping through the evaluation of a contract in a browser.
+For Marlowe to be usable in practice, users need to be able to design and develop Marlowe contracts, and also to understand how contracts will behave once deployed to the blockchain, but without before actually deploying them.
 
-To achieve this, and to aid Marlowe’s take-up usage by people who are not familiar with its syntax, we have developed the Marlowe Playground, a web tool that supports the interactive construction, revision, and simulation of smart-contracts written in Marlowe.
-The tool is publicly available in the url: https://prod.meadow.marlowe.iohkdev.io/.
-It should be mentioned that development of the playground is rapid and the latest unstable version is also publicly available at https://alpha.marlowe.iohkdev.io/.
+The Marlowe Playground, a web-based tool that supports the interactive construction, revision, and simulation of smart contracts written in Marlowe, provides these facilities, as well as access to a static analysis of contracts (as described in the previous section), an online tutorial for Marlowe and a set of example contracts.
+The playground is available at \url{https://prod.meadow.marlowe.iohkdev.io/}.\footnote{Development of the playground is rapid, and the latest, unstable, version is also available at \url{https://alpha.marlowe.iohkdev.io/}.}
 
-The playground has an editor where you can write a contract in "pure" Marlowe, that is, not embedded in a host language. The reason for doing this is two-fold:
+At the top level, the playground offers three panes: the main \emph{Simulation} pane, as well as panes for developing Marlowe contracts, embedded in \emph{Haskell} or using the \emph{Blockly} visual language.
+
+\paragraph{Development}
+
+On the simulation pane, ``pure'' Marlowe contacts can be developed: that is, contracts not embedded in a host language. The reason for doing this is two-fold:
 \begin{itemize}
     \item There is a shallower learning curve for users who are new to Haskell or programming languages in general.
-The Marlowe constructs are quite simple so there is no need to learn about Haskell syntax or even variables, functions etc.
-    \item As we step through the execution of a contract, the contract is reduced, it is very useful to be able to view and even edit the reduced contract during this execution.
+The Marlowe constructs are quite simple, and there is no need, at least initially, to learn about Haskell syntax or even variables, functions etc.
+    \item As we step through the execution of a contract in a simulation, the contract is reduced; it is very useful to be able to view, or even edit, the reduced contract during this execution.
 \end{itemize}
-This is probably not the best way to write a large contract so the playground contains an additional editor where contracts can be written using Haskell and then transferred as a pure Marlowe data structure into the Marlowe editor.
+As contracts become larger it makes sense to use another editor in the \emph{Haskell} pane.
+Here contracts can be written using facilities from \emph{Haskell} to abbreviate and make more readable the description of the contracts.
+These contracts can then be transferred as a pure Marlowe data structure into the simulation pane.
 
-In addition to pure Marlowe and Haskell editors, contracts can also be written using Google's Blockly visual programming language.
-Again this is not suitable for writing large contracts however it is an easy way to introduce the concepts of Marlowe to users who have no programming knowledge.
-Once a contract has been 'written' in Blockly it is possible to transfer that contract to the Marlowe editor.
+Contracts can also be written using Google's \emph{Blockly} visual programming language, as was earlier described in Meadow~\cite{isola-marlowe}.
+Blockly gives an easy way to introduce the concepts of Marlowe to users who have no programming knowledge, and in particular the editor gives users a set of options for each construct as the contract is built.
+Once a contract has been constructed in Blockly it is possible to transfer that contract to the simulation pane.%SJT Changed from "Marlowe editor".
 It is also possible to transfer a Marlowe contract to Blockly for further editing.
 
-The Marlowe editor in the unstable version of the playground has an additional feature to aid writing contracts called holes.
-If we enter the contract \lstinline{?mycontract} we will be presented with a dropdown list of values that could be used.
-
-\begin{wrapfigure}[5]{l}{0.25\textwidth}
+\begin{wrapfigure}[9]{l}{0.25\textwidth}
     \vspace*{-0.3in}
@@
 \end{wrapfigure}
-In our case \lstinline{?mycontract} must be a \lstinline{Contract} of some sort so we can choose one of the \lstinline{Contract} constructors from the dropdown list.
+
+The Marlowe editor in the unstable version of the playground has an additional feature to aid writing contracts called holes. If we enter the contract \lstinline{?mycontract} 
+%SJT: enter it where?
+we will be presented with a dropdown list of values that could be used.
+
+In our case \lstinline{?mycontract} must be a \lstinline{Contract} of some sort, and so we are offered a choice of \lstinline{Contract} constructors from a dropdown list.
 If we choose \lstinline{Pay} then the Marlowe editor will automatically fill in a skeleton \lstinline{Pay} contract with new holes where we need to provide values.
-\\ \\
 \begin{verbatim}
@@
 New options will be presented, one for each hole, and each will have a dropdown list of all the possible values.
+
 \begin{wrapfigure}[6]{l}{0.25\textwidth}
@@
 A complete contract can be written in this guided way with the user needing only to fill in strings and numbers by hand.
-This approach to writing holes in your code and "asking" the compiler what you could put in there is easy to implement in a DSL as there are very limited options however is is becoming popular with more complex languages such as Haskell and Idris.
+This approach to writing holes in your code and ``asking'' the compiler what you could put in there is easy to implement in a DSL because there are very limited options, however is is also becoming popular with more complex languages such as Haskell and Idris.
 
-Contracts written in the Marlowe editor are parsed in real-time and if there are no errors (and no holes) then the contract is analyzed to discover which actions a user could take to progress the contract.
-These actions are displayed in the "Input Composer" above the editor.
-Lets take the following example contract:
+Users can at any point save the current contract directly to a Github Gist, as well as being able to re-load contracts from Github Gists.
+There are also some demo contracts that can be loaded in their Haskell and Marlowe versions.
+
+\paragraph{Simulation}
+
+Contracts written in the Marlowe editor are parsed in real-time and if there are no errors (and no holes) then the contract is analysed to discover which actions a user could take to progress the contract.
+These actions are displayed in the ``Input Composer'' above the editor.
+Consider the following example contract:
 \begin{verbatim}
-    When [Case (Deposit (AccountId 0 "investor") "guarantor" (Constant 1000)) 
-    Close] 10 Close
+    When [Case (Deposit (AccountId 0 "investor") 
+                        "guarantor" 
+                        (Constant 1000)) 
+               Close] 
+         10 
+         Close
 \end{verbatim}
-In this case, the only action a user can take to progress the contract is to deposit 1000 ADA from the guarantor to the investor's account.
-The playground knows this and so the input composer displays this action in the input composer.
+In this case, the only action a user can take to progress the contract is to accept a deposit of 1000 ADA
+%SJT: we need to sort out the Ada/Lovelance mismatch.
+from the guarantor to the investor's account. Because of this, the playground can display this action in the input composer.
 
@@
 \end{figure}
-The user can then choose to add this action to a transaction, ready to be applied. Multiple actions can be added to a transaction before it is applied if such actions are meaningful.
+
+\noindent
+The user can then choose to add this action to a transaction being prepared.
+Once the action is added other inputs become possible; these are displayed in the input composer, and again they can be added to the transaction being composed.
+In this way, multiple actions can be added to a transaction before it is applied.
 
@@
 \end{figure}
-A user can then apply this transaction and in the example above this would result in the state pane showing a single payment and in addition the Contract in the Marlowe editor will have been reduced to \lstinline{Close}.
-At this point a user can undo the application of the transaction or even reset the contract to it's initial state.
-This enables a user to apply transactions, see what the effect is, step back and apply a transaction with different actions and see how these differences effect the end results.
-They can also change the reduced contract to investigate a different path.
 
-Users can at any point save their contract directly to a Github Gist and of course load a contract from a Github Gist.
-There are also some demo contracts that can be loaded in order to play around with realistic examples.
+\noindent
+A user can then apply this transaction and in the example above this would result in the state pane showing a single payment and in addition the contract in the Marlowe editor will have been reduced to \lstinline{Close}.
+
+At any point in the simulation the user can undo any steps made: in this particular case, they can undo the application of the transaction, and iteratively undo more steps.
+At any point they can also reset the contract to its initial state.
+This enables a user to apply transactions, see what the effect is, step back and apply a transaction with different actions and see how these differences affect the end result.
+They can also change the reduced contract to investigate variants of the original.
 
 The final feature that we would like to present is the static analysis of contracts.
-As mentioned in a previous section, it is possible to carry out a symbolic execution of a contract and then use a SMT solver to look for cases that could cause unwanted situations.
-The playground utilizes this to search for situations where contract execution would cause warnings.
+As described in the previous section, it is possible to carry out a symbolic execution of a contract and then use a SMT solver to look for cases that could cause unwanted situations.
+The playground uses this to search for situations where contract execution would cause warnings.
 For example, suppose you write a contract that causes a payment of 450 Lovelace from Alice to Bob but the contract allows a situation where Alice has only deposited 350 Lovelace then the static analysis will find this and report it to the playground user.

### 2019-12-03 papers/marlowe2019/paper.tex

@@
 \begin{verbatim}
-apply :: SymVal a => Bounds -> 
-      SEnvironment -> SState -> SInput -> Contract -> 
-      (SApplyResult -> DetApplyResult -> SBV a) -> SBV a
+apply :: SymVal a => Bounds
+      -> SEnvironment -> SState -> SInput -> Contract
+      -> (SApplyResult -> DetApplyResult -> SBV a) -> SBV a
 \end{verbatim}
@@
 
-%\clearpage
+\subsubsection{Bound on the maximum number of transactions\label{subsubsec:bound_max_transaction_number}}
+
+\emph{ToDo}
+Any transactions that result in errors do not modify the \texttt{State} or the \texttt{Contract}, we know that limiting the analysis to lists of transactions that are shorter than the bound does not lose generality.
+
 \section{The Marlowe Playground}

### 2019-12-04 papers/marlowe2019/paper.tex
Add draft for formal verification section

@@
 
-We can also use proof assistants like Isabelle to demonstrate that Marlowe semantics present certain desirable properties (like that money is preserved and returned to users by the end of the execution of any contract).
+We can also use proof assistants to demonstrate that Marlowe semantics present certain desirable properties (like that money is preserved and returned to users by the end of the execution of any contract).
 In the same way, we can also use this technique to prove properties for particular contracts, or for particular contract snippets, templates, or classes of contracts.
+
+Currently, we have translated the Haskell Marlowe semantics to Isabelle as closely as possible. The main differences between the Isabelle and the Haskell semantics are that:
+\begin{itemize}
+    \item The Isabelle semantics uses \emph{integers for identifiers} and the Haskell semantics uses strings.
+We do this because handling integers in Isabelle is easier and we only compare them for equality and order (their value is irrelevant).
+    \item We use a \emph{custom implementation of maps and sets} that use lists of tuples and lists of elements respectively.
+We do this because Isabelle already provides many theorems that are proven for lists, so it considerably facilitates the automation of certain proofs.
+\end{itemize}
+
+\subsubsection{Termination proof}
+
+As part of the formalisation of the semantics, Isabelle proves termination automatically for most functions.
+The only function for which termination is not trivial is \texttt{reductionLoop}.
+The reason why Isabelle cannot infer termination for this function is that it repeatedly calls \texttt{reduceContractStep} until it returns \texttt{NotReduced}, so it requires proving that \texttt{reduceContractStep} will eventually do so.
+
+The way we proved that this is the case was by establishing a measurment for the size of a pair of \texttt{Contract} and \texttt{State}. And we defined such measurement as follows:
+
+\begin{verbatim}
+fun evalBound :: "State => Contract => nat" where
+"evalBound sta cont = length (accounts sta) + 2 * (size cont)"
+\end{verbatim}
+
+\texttt{size} is a measurement already generated automatically by Isabelle for custom data types.
+
+The reason why we need the number of accounts in the \texttt{State} for the measurement is that \texttt{reduceContractStep}, when called with the contract \texttt{Close}, will refund one of the accounts and the contract will remain as \texttt{Close}.
+
+And the reason why we multiply the size of the \texttt{Contract} by two is that the primitive \texttt{Deposit} may increase the number of accounts in the \texttt{State} by one, but it will also reduce the size of \texttt{Contract} by at least one.
+By multiplying the size of the \texttt{Contract} by two, its decrease compensates the increase in the size of the list of accounts and, thus, it ensures that the measurement will decrease even in that scenario.
+
+\subsubsection{Valid state and positive account preservation}
+
+Since we use lists of pairs of integers as maps in our \texttt{State}, there are some values that are allowed by the type definition of state but, nevertheless, they make no sense given what these lists represent.
+In particular:
+
+\begin{enumerate}
+    \item The lists of tuples are representing a maps, so it does not make sense for two elements of the list to have the same first element (key).
+    \item Additionally, we want to minimise the number of ways in which a map can be represented as a list, so we require the first elements of the elements of the list to be sorted ascendingly.
+This allows us to use standard equality for comparing maps, for example.
+    \item We only expect accounts to store positive amounts money, so the second element of the pairs of the list of accounts must be a positive integer.
+Whenever an account becomes empty we just delete the account from the list.
+\end{enumerate}
+
+The two first invariants we define as \texttt{validState}. And the latest we define as follows:
+
+\begin{verbatim}
+fun positiveMoneyInAccountOrNoAccount
+             :: "AccountId => Token => Accounts => bool" where
+"positiveMoneyInAccountOrNoAccount accId tok mlist =
+    (case MList.lookup (accId, tok) mlist of
+       None => True
+     | Some x => x > 0)"
+\end{verbatim}
+
+And it must be \texttt{True} independently of the \texttt{AccountId} and \texttt{Token}.
+
+We prove that, if the properties are true for a given input \texttt{State}, then the semantics will preserve the properties in the output \texttt{State}.
+
+\subsubsection{Quiescent result}
+
+There is a class of \texttt{Contract}s that will not evolve when passed to the semantics unless a matching \texttt{Input} is passed as well.
+We call this state \textit{quiescent} and we define it as follows:
+
+\begin{verbatim}
+fun isQuiescent :: "Contract => State => bool" where
+"isQuiescent Close state = (accounts state = [])" |
+"isQuiescent (When _ _ _) _ = True" |
+"isQuiescent _ _ = False"
+\end{verbatim}
+
+We prove that if the input \texttt{State} is valid and all accounts are positive, then the result of a transaction is always quiescent.
+
+\subsubsection{Money preservation}
+
+One of the dangers of using smart-contracts is that, a badly written one may lock its funds forever.
+We want to avoid that feature in Marlowe altogether and, with that aim, we proved that Marlowe account system does not create or destroy any money.
+By the end of the contract, all the money should be distributed back somehow to a subset of the participants.
+
+We state the property as follows:
+
+\begin{verbatim}
+lemma computeTransaction_preserves_money :
+"validAndPositive_state state
+  ==> moneyInState state + moneyInInputs (inputs tra)
+        = moneyInTransactionOutput state (inputs tra)
+                (computeTransaction tra state contract)"
+\end{verbatim}
+
+The only consideration is that we only accept positive \texttt{Deposit}s, so we interpret money in \texttt{Deposit} inputs as the maximum of $0$ and the actual value specified in the \texttt{Deposit} input.
 
@@
 
-\emph{ToDo}
-Any transactions that result in errors do not modify the \texttt{State} or the \texttt{Contract}, we know that limiting the analysis to lists of transactions that are shorter than the bound does not lose generality.
+Another property of Marlowe is that any given finite \texttt{Contract} has an implicit finite bound on the maximum number of \texttt{Transaction}s that it accepts.
+This is a convenient property because:
+
+\begin{itemize}
+    \item It reduces the danger of Denegation of Service (DoS) attacks: because the number of valid inputs is limited, an attacker partcipant cannot arbitrarily block the contract by issuing an unbounded amount of useless \texttt{Transaction}s.
+    \item It bounds the length of traces (the number of transactions) that symbolic execution (see Section~\ref{subsec:static}) needs to explore (more on this below).
+\end{itemize}
+
+We state the property as follows:
+
+\begin{verbatim}
+lemma playTrace_only_accepts_maxTransactionsInitialState :
+"playTrace sl c l = TransactionOutput txOut
+  ==> length l <= maxTransactionsInitialState c"
+\end{verbatim}
+
+\noindent
+where \texttt{maxTransactionsInitialState} is essentially the maximum number of nested \texttt{When} in the contract plus one.
+
+This property implies that any trace that is longer than this is guaranteed to produce at least one error.
+And because transactions that produce an error do not alter the state of the contract, such a list of transactions (trace) will be equivalent to a list of transactions that does not have the erroneous transaction.
+
+In other words, for every list of \texttt{Transaction}s, there exists at least one list of \texttt{Transaction}s with length less or equal to \texttt{maxTransactionsInitialState} that has the same effect on the contracts.
+
+Thus, we do not lose generality by only exploring traces with length of at most \texttt{maxTransactionsInitialState}, because any longer traces produce the same effect.
 

### 2019-12-05 papers/marlowe2019/paper.tex
Fix SBV example

@@
 primeFactors :: SBV Integer -> SBV Integer -> SBV Bool
-primeFactors x y = x .> 1 .&& y .> 1 .=> x * y ./= 25723297
+primeFactors x y = x .> 1 .&& y .> 1 .&& x * y .== 25723297
 \end{verbatim}
@@
 *Main> primeFactors 98 314
-True
-*Main> prove primeFactors
-Falsifiable. Counter-example:
+False
+*Main> sat primeFactors 
+Satisfiable. Model:
     s0 = 6353 :: Integer
@@
 *Main> primeFactors 6353 4049
-False
+True    
 \end{verbatim}

### 2019-12-05 papers/marlowe2019/paper.tex
Improve writing of Section 4.2

@@
 We do this because handling integers in Isabelle is easier and we only compare them for equality and order (their value is irrelevant).
-    \item We use a \emph{custom implementation of maps and sets} that use lists of tuples and lists of elements respectively.
+    \item We use a \emph{custom implementation of maps and sets} that uses lists of tuples and lists of elements respectively.
 We do this because Isabelle already provides many theorems that are proven for lists, so it considerably facilitates the automation of certain proofs.
@@
 
-The way we proved that this is the case was by establishing a measurment for the size of a pair of \texttt{Contract} and \texttt{State}. And we defined such measurement as follows:
+The way we proved that this is the case was by establishing a measurment for the size of a pair of \texttt{Contract} and \texttt{State}. We defined such measurement as follows:
 
@@
 
-The reason why we need the number of accounts in the \texttt{State} for the measurement is that \texttt{reduceContractStep}, when called with the contract \texttt{Close}, will refund one of the accounts and the contract will remain as \texttt{Close}.
+The reason why we need the number of accounts in the \texttt{State} for the measurement is that \texttt{reduceContractStep}, when called with the contract \texttt{Close}, will refund one of the accounts and the contract will remain as \texttt{Close}, so its \texttt{size} does not decrease.
 
 And the reason why we multiply the size of the \texttt{Contract} by two is that the primitive \texttt{Deposit} may increase the number of accounts in the \texttt{State} by one, but it will also reduce the size of \texttt{Contract} by at least one.
-By multiplying the size of the \texttt{Contract} by two, its decrease compensates the increase in the size of the list of accounts and, thus, it ensures that the measurement will decrease even in that scenario.
+By multiplying the size of the \texttt{Contract} by two, the decrease in the \texttt{size} of the \texttt{Contract} compensates the increase in the \texttt{length} of the list of accounts and, thus, it ensures that the overall measurement \texttt{evalBound} will decrease even in that scenario.
 
@@
 
-Since we use lists of pairs of integers as maps in our \texttt{State}, there are some values that are allowed by the type definition of state but, nevertheless, they make no sense given what these lists represent.
-In particular:
+Since we use lists of pairs of integers as maps in our \texttt{State}, there are some values that are allowed by the type definition of \texttt{State} still make no sense given what these lists represent:
 
 \begin{enumerate}
-    \item The lists of tuples are representing a maps, so it does not make sense for two elements of the list to have the same first element (key).
-    \item Additionally, we want to minimise the number of ways in which a map can be represented as a list, so we require the first elements of the elements of the list to be sorted ascendingly.
-This allows us to use standard equality for comparing maps, for example.
-    \item We only expect accounts to store positive amounts money, so the second element of the pairs of the list of accounts must be a positive integer.
+    \item The lists of tuples are representing maps, so it does not make sense for two elements of the list to be tuples with the same first element (repeated keys).
+    \item We want to minimise the number of ways in which a map can be represented as a list, so we require the tuples in the list to be sorted ascendingly by their first element.
+This allows us to use standard equality for comparing maps.
+    \item We only expect accounts to store positive amounts money, so the second element of the tuples of the list of accounts must be a positive integer.
 Whenever an account becomes empty we just delete the account from the list.
@@
 
-The two first invariants we define as \texttt{validState}. And the latest we define as follows:
+If a value of \texttt{State} satisfies the first two properties above we call it a \texttt{validState}. And we formalised the third property above as follows:
 
@@
 
-And it must be \texttt{True} independently of the \texttt{AccountId} and \texttt{Token}.
+For a state to have positive accounts, the above function must evaluate to \texttt{True} for any \texttt{AccountId} and \texttt{Token}.
 
-We prove that, if the properties are true for a given input \texttt{State}, then the semantics will preserve the properties in the output \texttt{State}.
+We have proven that if all the three properties are true for a given input \texttt{State}, then the properties will also be true for the resulting output \texttt{State}, when processing any \texttt{Transaction} on any \texttt{Contract}.
 
@@
 
-There is a class of \texttt{Contract}s that will not evolve when passed to the semantics unless a matching \texttt{Input} is passed as well.
-We call this state \textit{quiescent} and we define it as follows:
+There is a class of combinations of \texttt{State} and \texttt{Contract} that will not evolve when passed to the semantics unless a matching \texttt{Input} is passed as well.
+We call these combinations of \texttt{Contract} and \texttt{State} \textit{quiescent}, and we define it as follows:
 
@@
 
-We prove that if the input \texttt{State} is valid and all accounts are positive, then the result of a transaction is always quiescent.
+We have proven that, if an input \texttt{State} is valid and all accounts are positive, then the combination of \texttt{State} and \texttt{Contract} resulting from processing any transaction will always be quiescent.
 
@@
 
-One of the dangers of using smart-contracts is that, a badly written one may lock its funds forever.
+One of the dangers of using smart-contracts is that a badly written one can potentially lock its funds forever.
 We want to avoid that feature in Marlowe altogether and, with that aim, we proved that Marlowe account system does not create or destroy any money.
-By the end of the contract, all the money should be distributed back somehow to a subset of the participants.
+By the end of the contract, all the money paid to the contract must be distributed back, in some way, to a subset of the participants of the contract.
 
-We state the property as follows:
+This requires that money is preserved (not created nor destroyed). We state this property as follows:
 

### 2019-12-05 src/Language/Marlowe/Analysis/FSSemantics.hs
Refine UselessTransaction check

@@ -782,11 +782,11 @@ convertTransactionError (SSAAReduceError reerr) f =
   f (sTransactionError $ sTEReduceError reerr) DTError
 
 -- Try to process a transaction
-processApplyResult :: SymVal a => Bounds -> STransaction -> Contract
+processApplyResult :: SymVal a => Bounds -> STransaction -> SState -> Contract
                    -> (STransactionResult -> DetTransactionResult -> SBV a)
                    -> SApplyAllResult -> DetApplyAllResult
                    -> SBV a
-processApplyResult bnds tra c f saar (DAARNormal ncon numPays numInps) =
+processApplyResult bnds tra sta c f saar (DAARNormal ncon numPays numInps) =
   symCaseApplyAllResult
     (\aar ->
      extractAppliedAll bnds aar
@@ -794,11 +794,15 @@ processApplyResult bnds tra c f saar (DAARNormal ncon numPays numInps) =
 --        let sigs = getSignatures bnds numInps inps in
 --        let outcomes = getOutcomes bnds numPays numInps ef inps in
         if c == ncon
-        then f (sTransactionError sTEUselessTransaction) DTError
+        then (if c /= Close
+              then (f (sTransactionError sTEUselessTransaction) DTError)
+              else (ite (SL.null $ account sta)
+                        (f (sTransactionError sTEUselessTransaction) DTError)
+                        (f (sTransactionProcessed wa ef {- sigs outcomes -} nsta) (DTProcessed ncon))))
         else f (sTransactionProcessed wa ef {- sigs outcomes -} nsta) (DTProcessed ncon)))
     saar
   where inps = inputs tra
-processApplyResult _ _ _ f saar DAARError =
+processApplyResult _ _ _ _ f saar DAARError =
   symCaseApplyAllResult
     (\aar -> convertTransactionError aar f)
     saar
@@ -806,9 +810,9 @@ processApplyResult _ _ _ f saar DAARError =
 processAux :: SymVal a => Bounds -> STransaction -> SState -> Contract
            -> (STransactionResult -> DetTransactionResult -> SBV a)
            -> SSIntervalResult -> SBV a
-processAux bnds tra _ c f (SSIntervalTrimmed env fixSta) =
+processAux bnds tra sta c f (SSIntervalTrimmed env fixSta) =
   applyAll bnds env fixSta c (inputs tra)
-           (\sym det -> processApplyResult bnds tra c f sym det)
[…]

### 2019-12-05 src/Language/Marlowe/Semantics.hs

@@ -143,9 +143,9 @@ data Contract = Close
   deriving (Eq,Ord,Show,Read,Generic,Pretty)
 
 data State = State { accounts    :: Map AccountId Money
-                    , choices     :: Map ChoiceId ChosenNum
-                    , boundValues :: Map ValueId Integer
-                    , minSlot     :: Slot }
+                   , choices     :: Map ChoiceId ChosenNum
+                   , boundValues :: Map ValueId Integer
+                   , minSlot     :: Slot }
   deriving (Eq,Ord,Show,Read)
 
 emptyState :: Slot -> State
@@ -490,7 +490,7 @@ computeTransaction tx state contract = let
     in case fixInterval (txInterval tx) state of
         IntervalTrimmed env fixState -> case applyAllInputs env fixState contract inputs of
             ApplyAllSuccess warnings payments newState cont -> let
-                in  if contract == cont
+                in  if (contract == cont) && ((contract /= Close) || (Map.null $ accounts state))
                     then Error TEUselessTransaction
                     else TransactionOutput { txOutWarnings = warnings
                                             , txOutPayments = payments

### 2019-12-05 papers/marlowe2019/paper.tex
Add timeout closure property to paper

@@
 
-\subsubsection{Money preservation}
+\subsubsection{Money preservation and contract timeout}
 
@@
 
-This requires that money is preserved (not created nor destroyed). We state this property as follows:
+To ensure this is the case we proved two properties:
+
+\paragraph{Money preservation}
+Money is not created or destroyed by the semantics. We state this property as follows:
 
@@
 The only consideration is that we only accept positive \texttt{Deposit}s, so we interpret money in \texttt{Deposit} inputs as the maximum of $0$ and the actual value specified in the \texttt{Deposit} input.
+
+\paragraph{Timeout closes a contract}
+
+For every Marlowe \texttt{Contract} there exists a slot number after which an empty transaction can be issued that will close the contract and refund all the money in its accounts.
+
+A conservative upper bound for the expiration slot number can be calculated efficiently by using the function \texttt{maxTime} (or \texttt{maxTimeContract} in the Isabelle semantics).
+
+We proved that this conservative upper bound works for every contract in a property stated as follows:
+
+\begin{verbatim}
+theorem timedOutTransaction_closes_contract :
+  "validAndPositive_state sta
+   ==> iniSlot >= minSlot sta
+   ==> iniSlot >= maxTimeContract cont
+   ==> endSlot >= iniSlot
+   ==> accounts sta ~= []
+   ==> isClosedAndEmpty
+          (computeTransaction (| interval = (iniSlot, endSlot)
+                               , inputs = [] |) sta cont)"
+\end{verbatim}
+
+\noindent
+
+where \texttt{isClosedAndEmpty} is defined as follows:
+
+\begin{verbatim}
+fun isClosedAndEmpty :: "TransactionOutput => bool" where
+"isClosedAndEmpty (TransactionOutput txOut) =
+   (txOutContract txOut = Close
+      /\ accounts (txOutState txOut) = [])" |
+"isClosedAndEmpty _ = False"
+\end{verbatim}
+
+Because the money is preserved and we know that there is an empty transaction that can be sent at any point after \texttt{maxTime} that will empty the accounts of the contract: we know that money will not be locked in the contract forever.
 

### 2019-12-05 papers/marlowe2019/paper.tex
Fix Isabelle snippets to use unicode

@@
 \documentclass[runningheads]{llncs}
-%
+
+\usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
 \usepackage{graphicx}
@@
 \usepackage{hyperref}
+\usepackage{stmaryrd}
+
+\DeclareUnicodeCharacter{21D2}{$\Rightarrow$}
+\DeclareUnicodeCharacter{2227}{$\wedge$}
+\DeclareUnicodeCharacter{2260}{$\neq$}
+\DeclareUnicodeCharacter{2265}{$\geq$}
+\DeclareUnicodeCharacter{2264}{$\leq$}
+\DeclareUnicodeCharacter{27F9}{$\Longrightarrow$}
+\DeclareUnicodeCharacter{2987}{$\llparenthesis$}
+\DeclareUnicodeCharacter{2988}{$\rrparenthesis$}
 
@@
 \begin{verbatim}
-fun evalBound :: "State => Contract => nat" where
+fun evalBound :: "State ⇒ Contract ⇒ nat" where
 "evalBound sta cont = length (accounts sta) + 2 * (size cont)"    
@@
 fun positiveMoneyInAccountOrNoAccount
-             :: "AccountId => Token => Accounts => bool" where
+             :: "AccountId ⇒ Token ⇒ Accounts ⇒ bool" where
 "positiveMoneyInAccountOrNoAccount accId tok mlist =
     (case MList.lookup (accId, tok) mlist of
-       None => True
-     | Some x => x > 0)"
+       None ⇒ True
+     | Some x ⇒ x > 0)"
 \end{verbatim}
@@
 \begin{verbatim}
-fun isQuiescent :: "Contract => State => bool" where
+fun isQuiescent :: "Contract ⇒ State ⇒ bool" where
 "isQuiescent Close state = (accounts state = [])" |
@@
 "validAndPositive_state state
-  ==> moneyInState state + moneyInInputs (inputs tra)
+  ⟹ moneyInState state + moneyInInputs (inputs tra)
        = moneyInTransactionOutput state (inputs tra)
@@
 "validAndPositive_state sta
-   ==> iniSlot >= minSlot sta
-   ==> iniSlot >= maxTimeContract cont
-   ==> endSlot >= iniSlot
-   ==> accounts sta ~= []
-   ==> isClosedAndEmpty
-          (computeTransaction (| interval = (iniSlot, endSlot)
-                               , inputs = [] |) sta cont)"
+ ⟹ iniSlot ≥ minSlot sta
+ ⟹ iniSlot ≥ maxTimeContract cont
+ ⟹ endSlot ≥ iniSlot
+ ⟹ accounts sta ≠ []
+ ⟹ isClosedAndEmpty
+         (computeTransaction ⦇ interval = (iniSlot, endSlot)
+                             , inputs = [] ⦈ sta cont)"
 \end{verbatim}
@@
 \begin{verbatim}
-fun isClosedAndEmpty :: "TransactionOutput => bool" where
+fun isClosedAndEmpty :: "TransactionOutput ⇒ bool" where
 "isClosedAndEmpty (TransactionOutput txOut) =
    (txOutContract txOut = Close
-      /\ accounts (txOutState txOut) = [])" |
+      ∧ accounts (txOutState txOut) = [])" |
 "isClosedAndEmpty _ = False"
@@
 "playTrace sl c l = TransactionOutput txOut
-  ==> length l <= maxTransactionsInitialState c"
+  ⟹ length l ≤ maxTransactionsInitialState c"
 \end{verbatim}

### 2019-12-08 papers/marlowe2019/paper.tex
Section Implementation of Marlowe on Cardano

@@
 Pablo {Lamela Seijas}\inst{1} \and
-Alex Nemish\inst{1} \and
+Alexander Nemish\inst{1} \and
 David Smith\inst{1} \and
@@
 
-\section{Implementation of Marlowe on the mockchain AN}
+\section{Implementation of Marlowe on Cardano}
 
- [An updated version of the description in the Plutus platform paper]
+\subsection{Cardano and Plutus}
+
+Cardano is a third-generation blockchain that solves the energy usage
+issue by moving to an energy efficient \emph{Proof of Stake} protocol.
+LINK?
+
+Cardano aims to support smart contracs during its Shelley release in 2020.
+Cardano smart contract platform is called \emph{Plutus}, and
+it uses Haskell programming language to generate a form of $System F_{\omega}$,
+called \emph{Plutus Core}, by extending GHC using its plugin
+support~\cite[Section 13.3]{ghcusersguide}.
+
+To implement Marlowe contracts we use the PlutusTx compiler,
+which compiles Haskell code into serialized \emph{Plutus Core} code,
+to create a Cardano \emph{validator script} that ensures the correct execution of the contract.
+This form of implementation relies on the extensions to the UTxO model
+that are described in REFERENCE TO PLUTUS WHAT???.
+
+\subsection{Extended UTxO}
+
+Cardano uses so called UTxO (unspent transaction outputs) model, similar to Bitcoin. LINK
+It extends Bitcoin's model by allowing transaction outputs to hold a \emph{Data Script}.
+As it sounds, it's just a serialized data intended to store a contract state.
+This allows us to create complex multitransactional contracts.
+
+In nutshell, it EUTxO models looks like this:
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=1\textwidth]{figures/Marlowe3-Figures-003.jpeg}
+    \caption{Cardano EUTx0 Model}
+    \label{fig:eutxo}
+\end{figure}
+
+Black circles represent \emph{unspent transaction outputs}, and red lines show \emph{transaction inputs}
+that reference existing \emph{transaction outputs}.
+Each transaction output contains a \emph{Value} it holds, and is protected either by a \emph{public key},
+or by a \emph{Validator}.
+
+In order to spend an existing transaction output protected by a \emph{Validator},
+one must create a transaction (\emph{PendingTx}) that has an \emph{input} that references the transaction output,
+and contains such a \emph{Redeemer}, that
+
+\begin{verbatim}
+    Validator(Data Script, Redeemer, PendingTx)
+\end{verbatim}
+evaluates to $True$.
+
+In order to spend a transaction output protected by a \emph{public key}, one must provide a valid signature
+made with respective private key of the transaction public key.
+
+\subsection{Design space}
+
+There are several ways to implement Marlowe contracts on top of Plutus.
+We could write a Marlowe to Plutus compiler that would convert
+a particular Marlowe contract into a specific Plutus script.
+Instead, we chose to implement an interpreter of Marlowe contracts.
+This approach has a number of advantages:
+\begin{itemize}
+\item It is simple: we implement a single Plutus script that can be used for all Marlowe contracts,
+thus making it easier to implement, review, and test what we have done.
+\item It is close to the semantics of Marlowe, as sketched above and in more detail in~\cite{Marlowe:2018},
+so making it easier to validate.
+\item It means that the same implementation can be used for both on- and off-chain (wallet) execution of Marlowe code.
+\item It allows client-side contract evaluation, where we reuse the same code
+to do contract execution emulation (i.g. in IDE), and compile it to WASM/JavaScript on client side
+(e.g.\ in Plutus Playground or Marlowe Playground).
+\item Having a single interpreter for all (or a particular group of) Marlowe contracts allows
+  to monitor the blockchain for these kinds of contract, if desired.
+\item Finally, there is a potential to special-case this sort of script,
+  and implement a specialized, highly effective interpreter in Cardano node itself.
+\end{itemize}
+
+Marlowe contract execution on the blockchain consists of a chain of transactions
+where, at each stage, the remaining contract and its state are passed through the \emph{Data script},
+and actions and inputs (i.e. \emph{choices} and \emph{money deposits}) are passed as
+\emph{Redeemer}.
+Each step in contract execution is a transaction that spends a Marlowe contract transaction output
+by providing a valid input in $Redeemer$, and produces a transaction output
+with a Marlowe contract as continuation (remaining contract).
+
+We store a remaining contract in the \emph{Data script}, which makes it visible to everyone.
+This simplifies contract reflection and retrospection.
+
+\subsection{Contract lifecycle on extended UTxO model}
+
+As we described above, the Marlowe interpreter is realised as a \emph{Validation script}.
+We can divide the execution of a Marlowe Contract into two phases:
+creation, and execution.
+
+\paragraph{Creation}
+
+Contract creation is realised as a transaction
+with at least one script output,
+with the particular Marlowe contract in the data script, and protected by the Marlowe validator script.
+Note that we do not place any restriction on the transaction inputs, which could use any other transaction
+outputs, including other scripts. This gives this model a great flexibility and composability.
+
+\begin{verbatim}
+data MarloweData = MarloweData {
+    marloweState    :: State,
+    marloweContract :: Contract
+}
+\end{verbatim}
+
+Contract has a state
+
+\begin{verbatim}
+data State = State { accounts    :: Map AccountId Ada
+                   , choices     :: Map ChoiceId ChosenNum
+                   , boundValues :: Map ValueId Integer
+                   , minSlot     :: Slot }
+\end{verbatim}
+
+where \emph{accounts} maps account ids to their balances, \emph{choices} stores user made choice values,
+\emph{boundValues} stores evalutated \emph{Value}'s introduced by \emph{Let} expressions,
+and \emph{minSlot} holds a minimal slot number a contract have seen (to avoid 'travels to the past')
+
+It is possible to initialize a contract with a particular state, containing a number of deposits,
+as shown in figure~\ref{fig:marlowe-init}.
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=1\textwidth]{figures/Marlowe3-Figures-002.jpeg}
+\caption{Marlowe contract initialization}
+\label{fig:marlowe-init}
+\end{figure}
+
+\paragraph{Execution}
+
+Marlowe contract execution consists of a chain of transactions,
+where the remaining contract and state are passed through the \emph{data script},
+and input actions (i.e. \emph{choices}) are passed as \emph{redeemer scripts}.
+
+Each execution step is a transaction that spends a Marlowe contract transaction output by providing
+an expected input in a redeemer script,
+and produces a transaction output with a Marlowe contract as continuation.
+
+The Marlowe interpreter first validates the current contract state.
+That is, we check that the contract locks at least what it should according
+to contract balances (i.e. \emph{accounts} field in \emph{State}),
+and all balances are strictly positive.
+
+We have proven Isabelle proof assistant, that given a state with positive balances,
+it's impossible to result in non-positive balances for all possible contracts and inputs.
+LINK??
+
+Then we compute expected transaction outcomes for given inputs, such as:
+contract continuation, new state, and payments.
+
+\begin{verbatim}
+expectedTxOutputs = computeTransaction txInput marloweState marloweContract
+\end{verbatim}
+
+where
+
+\begin{verbatim}
+data TransactionInput = TransactionInput
+    { txInterval :: SlotInterval
+    , txInputs   :: [Input] }
+
+data TransactionOutput =
+    TransactionOutput
+        { txOutWarnings :: [TransactionWarning]
+        , txOutPayments :: [Payment]
+        , txOutState    :: State
+        , txOutContract :: Contract }
+    | Error TransactionError
+
+computeTransaction :: TransactionInput -> State -> Contract -> TransactionOutput
+\end{verbatim}
+
+Interpreter reduces a contract until it becomes quiescent, meaning it's either evaluates to \emph{Close},
+or it expects a user input in \emph{When} construct.
+This means that all \emph{Pay, If, Let, Close} constructs are evaluated immediately.
+
+The function returns a new contract state, contract continuation, a list of warnings (such as partial payments),
+and a list of expected payments (i.e. all evaluated \emph{Pay} constructs).
+
+On-chain \emph{Validator} code cannot generate transaction outputs,
+but can only validate whatever a user provides in a transaction.
+
+Consider this simple zero coupon bond example.
+
+\begin{verbatim}
+When [ Case (Deposit aliceAccount alicePubKey (Constant 850_000_000))
+    (Pay aliceAccount (Party bobPubKey) (Constant 850_000_000)
+        (When
+            [ Case
+                (Deposit aliceAccount bobPubKey (Constant 1000_000_000))
+                Close
+            ] (Slot 200) Close
+        ))] (Slot 100) Close
+\end{verbatim}
+
+Here we expect Alice to deposit 850 ada (850,000,000 lovelace) into her $aliceAccount$ before slot 100.
+Otherwise, we \emph{Close} the contract.
+
+If Alice deposits the money before slot 100, money immediately go to Bob,
+by requiring a transaction output of 850 ada to Bob's public key address.
+She must produce the following \emph{Redeemer} to satisfy Marlowe validator:
+
+\begin{verbatim}
+[IDeposit aliceAccount alicePubKey 850000000]
+\end{verbatim}
+
+Then Bob is expected to deposit 1000 ada into Alice's account before slot 200.
+If he does, the contract is closed, and all remaining balances must be payed out
+to their respective owners. In our case, 1000 ada must be payed to Alice.
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=1\textwidth]{figures/Marlowe3-Figures-004.jpeg}
+    \caption{Zero Coupon Bond Contract}
+    \label{fig:zero-coupon-bond}
+\end{figure}
+
+Note, that it's possible to provide multiple inputs at a time,
+allowing to merge as many steps of a contract execution as necessary or possible.
+This gives atomicity of some operations, and saves on transaction fees.
+
+Sometimes it's necessary to just evaluate a contract in a given moment in time,
+and not provide any inputs. Consider the following example:
+
+\begin{verbatim}
+If (SlotIntervalStart `ValueGT' 100)
+    When [] (Slot 200)
+        (Pay aliceAccount (Party bobPubKey) (Constant 1000))
+    Close
+\end{verbatim}
+
+Let's say Alice has 1000 lovelace on her account, and Bob wants to receive the payment.
+The payment could be obtained only after \emph{When} is timed out after slot 200.
+Buy if we evaluate the contract after slot 200, \emph{If} reduces to \emph{Close}, and
+Bob won't receive the payment. So he must evaluate the contract before slot 100 to
+reduce \emph{If} to \emph{When}.
+To do that he can spend the contract transaction output with empty list of inputs in \emph{redeemer}.
+The he should do the same after slot 200 to receive the payment.
+
+\paragraph{Ensuring execution validity}
+
+Marlowe validator script checks that a transaction contains a valid continuation
+output in case a contract is not closed, by checking output's validator hash, and state.
+
+The continuation must have the same validator, hence the same hash of the validator script.
+It should contain expected continuation contract and state, evaluated by \emph{computeTransaction}
+function.
+
+\paragraph{Closing a contract}
+
+When a contract evaluates to \emph{Close}, and all remaining balances must payed out
+to respective owners, removing the contract from the set of unspent transaction outputs.
+
+\subsection{Further Work}
+
+Cardano extends its ledger rules to support \emph{forging} of custom currencies and tokens,
+effectively embedding support of Ethereum ERC-20 and ERC-721 contracts.
+We are working on adding this \emph{multicurrency} support to Marlowe.
+
+Simple token creation gives interesting possibilities of representing Marlowe contract parties with tokens.
+Currently, parties are represented as public keys. We expect a valid signature for each input,
+and do payments to respective addresses for party's public key.
+In similar matter, we could check proofs of ownership of a particular token.
+
+This allows us to tokenize contract participants, and abstract away concrete public key into contract \emph{roles}.
+Thus, those roles could be traded independently of a contract.
+Also, we can represent a role with multiple \emph{shares}, allowing automatic distribution of a payment
+proportionally to a number of shares owned.
 

### 2019-12-08 papers/marlowe2019/paper.tex
added references and intro to section 3.

@@
 
+Marlowe is specified by an executable semantics written in Haskell, but to make it usable in practice to execute financial contracts, it needs to be implemented on a blockchain.
+In this section we explain how Marlowe is executed on the Cardano blockchain using an interpreter written in the Plutus programming language.
+
 \subsection{Cardano and Plutus}
@@
 Cardano is a third-generation blockchain that solves the energy usage
-issue by moving to an energy efficient \emph{Proof of Stake} protocol.
-LINK?
+issue by moving to an energy efficient \emph{Proof of Stake} protocol~\cite{Ouroboros-Genesis}.
 
@@
 to create a Cardano \emph{validator script} that ensures the correct execution of the contract.
-This form of implementation relies on the extensions to the UTxO model
-that are described in REFERENCE TO PLUTUS WHAT???.
+This form of implementation relies on the extensions to the UTxO model described in~\cite{PlutusPlatform}.
 
@@
     \centering
-    \includegraphics[width=1\textwidth]{figures/Marlowe3-Figures-003.jpeg}
+    \includegraphics[width=4in]{figures/Marlowe3-Figures-003.jpeg}
     \caption{Cardano EUTx0 Model}

### 2019-12-08 papers/marlowe2019/paper.tex
Editing of section 3.

@@
 
-Cardano uses so called UTxO (unspent transaction outputs) model, similar to Bitcoin. LINK
-It extends Bitcoin's model by allowing transaction outputs to hold a \emph{Data Script}.
-As it sounds, it's just a serialized data intended to store a contract state.
-This allows us to create complex multitransactional contracts.
-
-In nutshell, it EUTxO models looks like this:
+Cardano is a UTxO-based (unspent transaction output) blockchain, similar to Bitcoin~\cite{sok}.
+It extends the Bitcoin model by allowing transaction outputs to hold a \emph{Data Script}.
+As the name suggests, this is  a serialised data value used to store and communicate a contract state.
+This allows us to create complex multi-transactional contracts.
+In a nutshell, the EUTxO model looks like this:
 
@@
 
-Black circles represent \emph{unspent transaction outputs}, and red lines show \emph{transaction inputs}
+\noindent
+where black circles represent \emph{unspent transaction outputs}, and red lines show \emph{transaction inputs}
 that reference existing \emph{transaction outputs}.
-Each transaction output contains a \emph{Value} it holds, and is protected either by a \emph{public key},
+Each transaction output contains a \emph{Value}, and is protected either by a \emph{public key},
 or by a \emph{Validator}.
@@
 In order to spend an existing transaction output protected by a \emph{Validator},
-one must create a transaction (\emph{PendingTx}) that has an \emph{input} that references the transaction output,
-and contains such a \emph{Redeemer}, that
+one must create a transaction (a \emph{PendingTx}) that has an \emph{input} that references the transaction output,
+and contains a \emph{Redeemer}, such that
 
@@
 evaluates to $True$.
-
 In order to spend a transaction output protected by a \emph{public key}, one must provide a valid signature
-made with respective private key of the transaction public key.
+made with the private key corresponding to the transaction public key.
 
@@
 We could write a Marlowe to Plutus compiler that would convert
-a particular Marlowe contract into a specific Plutus script.
-Instead, we chose to implement an interpreter of Marlowe contracts.
+each Marlowe contract into a specific Plutus script.
+Instead, we chose to implement an interpreter of Marlowe contracts as a single Plutus script.
 This approach has a number of advantages:
 \begin{itemize}
-\item It is simple: we implement a single Plutus script that can be used for all Marlowe contracts,
-thus making it easier to implement, review, and test what we have done.
-\item It is close to the semantics of Marlowe, as sketched above and in more detail in~\cite{Marlowe:2018},
+\item It is simple: having a single Plutus script that implements all Marlowe contracts
+makes it easier to implement, review, and test what we have done.
+\item It is close to the semantics of Marlowe, as sketched above and in more detail in~\cite{isola-marlowe},
 so making it easier to validate.
@@
 \item It allows client-side contract evaluation, where we reuse the same code
-to do contract execution emulation (i.g. in IDE), and compile it to WASM/JavaScript on client side
-(e.g.\ in Plutus Playground or Marlowe Playground).
-\item Having a single interpreter for all (or a particular group of) Marlowe contracts allows
-  to monitor the blockchain for these kinds of contract, if desired.
+to do contract execution emulation in an IDE, and compile it to WASM/JavaScript on the client side, 
+e.g.\ in the Marlowe Playground.
+\item Having a single interpreter for all (or a particular group of) Marlowe contracts allows us
+  to monitor the blockchain for these contracts, if required.
 \item Finally, there is a potential to special-case this sort of script,
-  and implement a specialized, highly effective interpreter in Cardano node itself.
+  and implement a specialized, highly effective interpreter in the Cardano node itself.
 \end{itemize}
 
+\noindent
 Marlowe contract execution on the blockchain consists of a chain of transactions
 where, at each stage, the remaining contract and its state are passed through the \emph{Data script},
-and actions and inputs (i.e. \emph{choices} and \emph{money deposits}) are passed as
+and actions and inputs (i.e. \emph{choices} and \emph{money deposits}) are passed to the %SJT: is this irght?
 \emph{Redeemer}.
@@
 
-We store a remaining contract in the \emph{Data script}, which makes it visible to everyone.
+We store the remaining contract in the \emph{Data script}, which makes it visible to everyone.
 This simplifies contract reflection and retrospection.
 
-\subsection{Contract lifecycle on extended UTxO model}
+\subsection{Contract lifecycle on the extended UTxO model}
 
-As we described above, the Marlowe interpreter is realised as a \emph{Validation script}.
+As described above, the Marlowe interpreter is realised as a \emph{Validation script}.
 We can divide the execution of a Marlowe Contract into two phases:
-creation, and execution.
+creation and execution.
 
@@
 Note that we do not place any restriction on the transaction inputs, which could use any other transaction
-outputs, including other scripts. This gives this model a great flexibility and composability.
+outputs, including other scripts. This gives this model optimal flexibility and composability.
 
@@
 \end{verbatim}
-
-Contract has a state
+\noindent
+The contract has a state
 
@@
 
+\noindent%SJT: **general question** using \emph for inline code: why not \texttt ?
 where \emph{accounts} maps account ids to their balances, \emph{choices} stores user made choice values,
 \emph{boundValues} stores evalutated \emph{Value}'s introduced by \emph{Let} expressions,
-and \emph{minSlot} holds a minimal slot number a contract have seen (to avoid 'travels to the past')
+and \emph{minSlot} holds a minimal slot number that a contract has seen, to avoid `time travel to the past'.
 
@@
 
-The Marlowe interpreter first validates the current contract state.
-That is, we check that the contract locks at least what it should according
-to contract balances (i.e. \emph{accounts} field in \emph{State}),
+The Marlowe interpreter first validates the current contract state: i.e.\ 
+we check that the contract locks at least what it should according
+to contract balances (the \emph{accounts} field in \emph{State}),
 and all balances are strictly positive.
+\footnote{
+Using the Isabelle proof assistant, we have formally verified that given a state with positive balances,
+it is impossible for any possible contract and inputs to result in non-positive balances.
+This is described in more detail in Section \ref{sec:analysis}.}
 
-We have proven Isabelle proof assistant, that given a state with positive balances,
-it's impossible to result in non-positive balances for all possible contracts and inputs.
-LINK??
-
-Then we compute expected transaction outcomes for given inputs, such as:
+We then compute expected transaction outcomes for given inputs, such as:
 contract continuation, new state, and payments.
@@
 
+\noindent
 where
@@
 
-Interpreter reduces a contract until it becomes quiescent, meaning it's either evaluates to \emph{Close},
+\noindent
+The interpreter reduces a contract until it becomes quiescent, meaning that either it evaluates to \emph{Close},
 or it expects a user input in \emph{When} construct.
@@
 The function returns a new contract state, contract continuation, a list of warnings (such as partial payments),
-and a list of expected payments (i.e. all evaluated \emph{Pay} constructs).
+and a list of expected payments (i.e. all the evaluated \emph{Pay} constructs).
 
-On-chain \emph{Validator} code cannot generate transaction outputs,
+The on-chain \emph{Validator} code cannot generate transaction outputs,
 but can only validate whatever a user provides in a transaction.
-
 Consider this simple zero coupon bond example.
@@
 
-Here we expect Alice to deposit 850 ada (850,000,000 lovelace) into her $aliceAccount$ before slot 100.
+\noindent
+Here we expect Alice to deposit 850 Ada (850,000,000 Lovelace) into her $aliceAccount$ before slot 100.
 Otherwise, we \emph{Close} the contract.
 
-If Alice deposits the money before slot 100, money immediately go to Bob,
-by requiring a transaction output of 850 ada to Bob's public key address.
-She must produce the following \emph{Redeemer} to satisfy Marlowe validator:
+If Alice deposits the money before slot 100, money immediately goes to Bob,
+by requiring a transaction output of 850 Ada to Bob's public key address.
+Alice must produce the following \emph{Redeemer} to satisfy the Marlowe validator:
 
@@
 
-Then Bob is expected to deposit 1000 ada into Alice's account before slot 200.
-If he does, the contract is closed, and all remaining balances must be payed out
-to their respective owners. In our case, 1000 ada must be payed to Alice.
+\noindent
+Bob is then is expected to deposit 1000 Ada into Alice's account before slot 200.
+If he does, the contract is closed, and all remaining balances must be paid out
+to their respective owners. In our case, 1000 Ada must be paid to Alice.
 
@@
 
-Note, that it's possible to provide multiple inputs at a time,
-allowing to merge as many steps of a contract execution as necessary or possible.
+Note, that it is possible to provide multiple inputs at a time,
+allowing as many steps of a contract execution as necessary to be merged.
 This gives atomicity of some operations, and saves on transaction fees.
 
-Sometimes it's necessary to just evaluate a contract in a given moment in time,
+Sometimes it is necessary to just evaluate a contract in a given moment in time,
 and not provide any inputs. Consider the following example:
@@
 
-Let's say Alice has 1000 lovelace on her account, and Bob wants to receive the payment.
-The payment could be obtained only after \emph{When} is timed out after slot 200.
+Let us say that Alice has 1000 Lovelace on her account, and Bob wants to receive the payment.
+The payment could be obtained only after \emph{When} has timed out after slot 200.
 Buy if we evaluate the contract after slot 200, \emph{If} reduces to \emph{Close}, and
-Bob won't receive the payment. So he must evaluate the contract before slot 100 to
+Bob will not receive the payment. So he must evaluate the contract before slot 100 to
 reduce \emph{If} to \emph{When}.
 To do that he can spend the contract transaction output with empty list of inputs in \emph{redeemer}.
-The he should do the same after slot 200 to receive the payment.
+He should then do the same after slot 200 to receive the payment.
 
@@
 
-Marlowe validator script checks that a transaction contains a valid continuation
-output in case a contract is not closed, by checking output's validator hash, and state.
-
+The Marlowe validator script checks that a transaction contains a valid continuation
+output in case a contract is not closed, by checking the hash of the output validator and state.
 The continuation must have the same validator, hence the same hash of the validator script.
-It should contain expected continuation contract and state, evaluated by \emph{computeTransaction}
+It should contain both the expected continuation contract and then state, evaluated by the \emph{computeTransaction}
 function.
@@
 
-Simple token creation gives interesting possibilities of representing Marlowe contract parties with tokens.
-Currently, parties are represented as public keys. We expect a valid signature for each input,
-and do payments to respective addresses for party's public key.
+Simple token creation gives interesting possibilities of representing Marlowe contract parties by tokens.
+Currently parties are represented as public keys. We expect a valid signature for each input,
+and make payments to  addresses for trhe public key of a party.
 In similar matter, we could check proofs of ownership of a particular token.
 
-This allows us to tokenize contract participants, and abstract away concrete public key into contract \emph{roles}.
+This allows us to tokenize contract participants, and abstract away concrete public keys into contract \emph{roles}.
 Thus, those roles could be traded independently of a contract.

### 2019-12-09 papers/marlowe2019/paper.tex
Started on introduction.:wq

@@
 %
-\title{Marlowe: implementing and analysing financial contracts\thanks{Supported by IOHK.}}
+\title{Marlowe: implementing and analysing financial contracts on blockchain}%\thanks{Supported by IOHK.}}
 %
@@
 David Smith\inst{1} \and
-Simon Thompson\inst{1,2}\orcidID{0000-0002-2350-301X}}
+Simon Thompson\inst{1,2}}%\orcidID{0000-0002-2350-301X}}
 %
@@
 
-Intro
-\section{Marlowe overview SJT}
+Marlowe is a domain-specific language (DSL) for implementing financial contracts on blockchain: our initial target is Cardano, but it can be implemented on most DLT platforms, including Ethereum.
+Our DSL is embedded in Haskell, allowing users selectively to use aspects of Haskell -- typically definitions of constants and simple functions -- to express contracts more readably and succinctly.
+Marlowe was introduced in the paper~\cite{isola-marlowe}: Section \ref{sec:overview} here gives an overview of the language, and the changes made to it since the publication of that paper.
+
+Marlowe is specified by a reference semantics for the language written in Haskell, thus giving us an \emph{executable} semantics for the language, which we can use in a number of ways.
+Obviously, we can interpret Marlowe contracts in Haskell itself, but we can also use that implementation, translated into Plutus~\cite{PlutusPlatform}, to interpret  Marlowe directly on the Cardano blockchain, as explained in Section \ref{sec:implementation}.
+Moreover, we can execute that semantics (translated into PureScript) directly in a browser to give an interactive simulation environment, the Meadow Playground, described in Section \ref{sec:playground}.
+
+DSLs have a number of advantages over general purpose languages. TO COMPLETE.
+
+\section{Marlowe overview}
+\label{sec:overview}
+
+, as well as the online simulation environment Meadow.
+Since publication we have revised the language design, and so we give a brief overview of the current (3.0) version of the language and its informal semantics.\footnote{The complete Marlowe project, including semantics and implementation, is available at \url{https://github.com/input-output-hk/marlowe}.} Meadow has been upgraded to the Marlowe Playground, and that is also introduced here.
 
@@
 \section{Implementation of Marlowe on Cardano}
+\label{sec:implementation}
 
@@
 \section{The Marlowe Playground}
+\label{sec:playground}
 

### 2019-12-09 papers/marlowe2019/paper.tex
Typos
Completed intro

--- 2019-12-08 papers/marlowe2019/paper.tex
@@
 Pablo {Lamela Seijas}\inst{1} \and
-Alexander Nemish\inst{1} \and
+Alexander Nemish\inst{1} \and \\
 David Smith\inst{1} \and
-Simon Thompson\inst{1,2}\orcidID{0000-0002-2350-301X}}
+Simon Thompson\inst{1,2}}%\orcidID{0000-0002-2350-301X}}
 %
@@
 \begin{abstract}
-The abstract should briefly summarize the contents of the paper in
-150--250 words.
+Marlowe is a DSL for financial contracts running on the Cardano blockchain.
+We describe the implementation of Marlowe on blockchain, and the Marlowe Playground web-based development and simulation environment.
 
-\keywords{First keyword \and Second keyword \and Another keyword.}
+Contracts in Marlowe can be exhaustively analysed prior to running them, thus providing strong guarantees to participants in the contracts.
+The Marlowe system itself has been formally verified using the Isabelle theorem prover, establishing such properties as the conservation of money.
+
+\keywords{blockchain \and Cardano \and Haskell \and SMT \and static analysis}
 \end{abstract}
-%
-%
-%
 
@@
 
-Intro
-\section{Marlowe overview SJT}
+Marlowe is a domain-specific language (DSL) for implementing financial contracts on blockchain: our initial target is Cardano, but it can be implemented on many DLT platforms, including Ethereum.
+Marlowe is embedded in Haskell, allowing users selectively to use aspects of Haskell -- typically definitions of constants and simple functions -- to express contracts more readably and succinctly.
+Marlowe was introduced in the paper~\cite{isola-marlowe}: Section \ref{sec:overview} below gives an overview of the language, and the changes made to it since the publication of that paper.
+
+Marlowe is specified by a reference semantics for the language written in Haskell, and we can use that  in a number of ways.
+Obviously, we can interpret Marlowe contracts in Haskell itself, but we can also use that implementation, translated into Plutus~\cite{PlutusPlatform}, to interpret  Marlowe directly on the Cardano blockchain, as explained in Section \ref{sec:implementation}.
+Moreover, we can execute that semantics -- translated into PureScript -- directly in a browser to give an interactive simulation environment, the Meadow Playground, described in Section \ref{sec:playground}.
+
+Because Marlowe is a DSL, we are able to build special purpose tools and techniques to support it.
+Crucially in a financial environment we are able to \emph{exhaustively analyse contracts} without executing them, so that we can, for instance, check whether any particular contract makes all the payments it should: in the case it does not, we get an explicit example of how it can fail.
+This analysis is built into the Marlowe Playground.
+We are also able to use program verification techniques to prove guarantees about the implementation of Marlowe, including a guarantee that ``money in $=$ money out'' for all contracts.
+
+\section{Marlowe overview}
+\label{sec:overview}
+
+Since publication we have revised the language design, and so we give a brief overview of the current (3.0) version of the language and its informal semantics.\footnote{The complete Marlowe project, including semantics and implementation, is available at \url{https://github.com/input-output-hk/marlowe}.} Meadow has been upgraded to the Marlowe Playground, and that is also introduced here.
 

### 2019-12-09 papers/marlowe2019/paper.tex
Added cross reference to sectino 4 to intro.
Weaken timeout precondition

@@
 \DeclareUnicodeCharacter{21D2}{$\Rightarrow$}
-\DeclareUnicodeCharacter{2227}{$\wedge$}
+\DeclareUnicodeCharacter{2227}{$\land$}
+\DeclareUnicodeCharacter{2228}{$\lor$}
 \DeclareUnicodeCharacter{2260}{$\neq$}
@@
 We are also able to use program verification techniques to prove guarantees about the implementation of Marlowe, including a guarantee that ``money in $=$ money out'' for all contracts.
+Section \ref{sec:analysis} covers analysis and verification.
 
@@
  ⟹ endSlot ≥ iniSlot
- ⟹ accounts sta ≠ []
+ ⟹ accounts sta ≠ [] ∨ cont ≠ Close
  ⟹ isClosedAndEmpty

### 2019-12-09 papers/marlowe2019/paper.tex
First attempt at Marlowe intro.

--- 2019-12-09 papers/marlowe2019/paper.tex
@@
 
-Marlowe is a domain-specific language (DSL) for implementing financial contracts on blockchain: our initial target is Cardano, but it can be implemented on many DLT platforms, including Ethereum.
+Marlowe\footnote{The complete Marlowe project, including semantics and implementation, is available at \url{https://github.com/input-output-hk/marlowe}.} is a domain-specific language (DSL) for implementing financial contracts on blockchain: our initial target is Cardano, but it can be implemented on many DLT platforms, including Ethereum.
 Marlowe is embedded in Haskell, allowing users selectively to use aspects of Haskell -- typically definitions of constants and simple functions -- to express contracts more readably and succinctly.
@@
 Crucially in a financial environment we are able to \emph{exhaustively analyse contracts} without executing them, so that we can, for instance, check whether any particular contract makes all the payments it should: in the case it does not, we get an explicit example of how it can fail.
-This analysis is built into the Marlowe Playground.
-We are also able to use program verification techniques to prove guarantees about the implementation of Marlowe, including a guarantee that ``money in $=$ money out'' for all contracts.
+This analysis, reported in Section \ref{sec:analysis}, is built into the Marlowe Playground.
+This section also covers our use of formal verification to prove properties of the implementation of Marlowe, including a guarantee that ``money in $=$ money out'' for all contracts.
 
@@
 
-Since publication we have revised the language design, and so we give a brief overview of the current (3.0) version of the language and its informal semantics.\footnote{The complete Marlowe project, including semantics and implementation, is available at \url{https://github.com/input-output-hk/marlowe}.} Meadow has been upgraded to the Marlowe Playground, and that is also introduced here.
+Since first publication we have revised the language design: this section gives a brief overview of the current (3.0) version of the language and its semantics.
 
-Introduce the main types and informal overview of execution. Contrast with Marlowe 1.3 (as in ISoLA paper). Embedding in Haskell? [All this stuff is in the tutorial already]
+\subsubsection*{The Marlowe model}
+
+Marlowe is designed to support the execution of financial contracts on blockchain, and specifically to work on Cardano.
+Contracts are built by putting together a small number of constructs that in combination can be used to describe many different kinds of financial contract.
+
+Contracts in Marlowe run on a blockchain, but need to interact with the off-chain world.
+The parties to the contract, also called the participants, can engage in various actions: they can be asked to deposit money, or to make a choice between various alternatives.
+A notification of an external / oracle value is the other possible form of input.
+Running a contract will also produce external effects, by making payments to parties in the contract.
+
+The Marlowe model allows for a contract to control money in a number of disjoint accounts: this allows for more explicit control of how the money flows in the contract.
+Each account is owned by a particular party to the contract, and that party receives a refund of any remaining funds in the account when the contract is closed.
+%These accounts are local, in that they only exist as during the execution of the contract, and during that time they are only accessible by parties to the contract.
+
+Marlowe contracts describe a series of steps, typically by describing the first step, together with another (sub-) contract that describes what to do next.
+For example, the contract \texttt{Pay a p v cont} says ``make a payment of \texttt{v} Lovelace to the party \texttt{p} from the account \texttt{a}, and then follow the contract \texttt{cont}''.
+We call \texttt{cont} the continuation of the contract.
+
+In executing a contract we need to keep track of the current contract: after making a step in the example above, the current contract is \texttt{cont}.
+We also have to keep track of some other information, such as how much is held in each account: this information together is the state, and generally this changes at each step.
+A step can also see an action taking place, such as money being deposited, or an effect being produced, e.g.
+a payment.
+%Value on the blockchain resides in the UTxO, which are protected cryptographically by a private key held by the owner.
+These keys can be used to redeem the output, and so to use them as inputs to new transactions, which can be seen as spending the value in the inputs.
+Users typically keep track of their private keys, and the values attached to them, in a cryptographically-secure wallet.
+It is through their wallets that users are able to interact with Marlowe contracts running on the blockchain, making deposits and receiving payments.
+%Note, however, that these are definitely off chain actions that need to be controlled by code running in the user's wallet: they cannot be made to happen by the Marlowe contract itself.
+
+%\subsection{Executing a Marlowe contract}
+%At each slot, a running Marlowe contract will receive a list of inputs in order.
+The contract is executed by evaluating it step by step until it cannot be changed any further without processing any input, a condition that we call being quiescent.
+The first input is then processed, and then the contract is single stepped again until quiescence, and this process is repeated until all the inputs are processed.
+At each step the current contact and the state will change, some input may be processed, and payments made.
+
+\subsubsection*{Marlowe step by step}
+Marlowe has five ways of building contracts.
+Four of these -- \texttt{Pay}, \texttt{Let}, \texttt{If} and \texttt{When} -- build a complex contract from simpler contracts, and the fifth, \texttt{Close}, is a simple contract.
+At each step of execution, as well as returning a new state and continuation contract, it is possible that effects, e.g.\ payments, and warnings can be generated too.
+We also explain Marlowe values, observations and actions, which are used to supply external information and inputs to a running contract to control how it will evolve.
+
+\paragraph{Pay}
+A payment contract \texttt{Pay a p v} cont will make a payment of value \texttt{v} from the account \texttt{a} to a payee \texttt{p}, which will be one of the contract participants or another account in the contract.
+Warnings will be generated if the value \texttt{v} is negative, or if there is not enough in the account to make the payment in full.
+In that case a partial payment (of all the money available) is made.
+The continuation contract is the one given in the contract: \texttt{cont}.
+
+\paragraph{Close}
+A contract \texttt{Close} provides for the contract to be closed (or terminated).
+The only action that is performs is to provide refunds to the owners of accounts that contain a positive balance.
+This is performed one account per step, but all accounts will be refunded in a single transaction.
+
+%Before discussing other forms of contracts, we need to describe values, observations and actions.
+
+\paragraph{Values, observations and actions}
+\emph{Values} include some quantities that change with time, including ``the current slot number'', ``the current balance of an account (in Lovelace)'', and any choices that have already been made; we call these volatile values.
+Values can also be combined using addition, subtraction and negation.
+
+\emph{Observations} are Boolean values derived by comparing values, and can be combined using the standard Boolean operators.
+It is also possible to observe whether any choice has been made (for a particular identified choice).
+
+Observations will have a value at every step of execution. On the other hand, \emph{actions} happen at particular points during execution and can be
+(i) depositing money, (ii)
+making a choice between various alternatives, or (iii)
+notifying an external or oracle value, such as the current price of a particular commodity.
+
+\paragraph{If}
+The conditional \texttt{If obs cont1 cont2} will continue as \texttt{cont1} or \texttt{cont2}, depending on the Boolean value of the observation \texttt{obs} on execution.
+
+\paragraph{When}
+This is the most complex constructor for contracts, with the form \texttt{When cases timeout cont}.
+It is a contract that is triggered on actions, which may or may not happen at any particular slot: what happens when various actions happen is described by the cases in the contract.
+
+In the contract \texttt{When cases timeout cont}, the list \texttt{cases} contains a collection of cases.
+Each case has the form \texttt{Case ac co} where \texttt{ac} is an action and \texttt{co} a continuation (another contract).
+When a particular action, e.g.
+\texttt{ac}, happens, the state is updated accordingly and the contract will continue as the corresponding continuation \texttt{co}.
+
+In order to make sure that the contract makes progress eventually, the contract \texttt{When cases timeout cont} will continue as cont once the \texttt{timeout}, a slot number, is reached.
+
+\paragraph{Let}
+A let contract \texttt{Let id val cont} causes the expression \texttt{val} to be evaluated, and stored with the name \texttt{id}. The contract then continues as \texttt{cont}.
+
+%As well as allowing us to use abbreviations, this mechanism also means that we can capture and save volatile values that might be changing with time, e.g.
+the current price of oil, or the current slot number, at a particular point in the execution of the contract, to be used later on in contract execution.
+
+%%% end overview
 

### 2019-12-09 papers/marlowe2019/paper.tex
Initial incomplete draft of related work.

@@
 \end{figure}
-\clearpage
+
 \section{Related work}
+\label{sec:related}
 
-Need to update from ISoLA paper \cite{isola-marlowe}.
+An earlier paper~\cite{cryptoeprint:2016:1156} reviews smart contracts on blockchain. Here we look at a number of recent systems that bear direct comparison with Marlowe.
+
+The Ethereum system~\cite{wood2014ethereum} provides an exemplar Turing-complete language for  the EVM virtual machine, and a higher-level programming language, Solidity, that compiles into EVM code.
+Both these languages are general-purpose, and so have more akin with Plutus~\cite{PlutusPlatform} rather than Marlowe.
+
+On the other hand,  Nxt~\cite{Nxt}, is special-purpose in providing a ``fat'' high-level 
+API, containing built-in transaction types and transactions that support some 250 primitive operations; these can be ``scripted'' in a client (only) using a binding to the API, which is available, for instance, in JavaScript.
+In providing such specificity this bears comparison with our implementation of contracts from the ACTUS standard~\cite{actus}.
+
+Our work is inspired by the original work of Peyton Jones and others~\cite{PeytonJones:2000} to describe financial contracts using a DSL embedded in Haskell.
+The Findel project~\cite{findel} examines financial contracts on the Ethereum platform, and is also based on~\cite{PeytonJones:2000}.
+The authors note that payments need to be bounded; this is made concrete in our
+account by our notion of commitments.
+They take no account of commitments or timeouts as our approach does, and so are unable to guarantee some of the properties -- such as a finite lifetime -- built into Marlowe by design.
+
+\paragraph{TO DO}
+
+Add stuff on BitML, DAML, \ldots.
 
 \section{Future work and conclusions}
-
-\section*{Left in for info \ldots}
-
-\begin{figure}
-% \includegraphics[width=\textwidth]{fig1.eps}
-\caption{A figure caption is always placed below the illustration.
-Please note that short captions are centered, while long ones are
-justified by the macro package automatically.} \label{fig1}
-\end{figure}
 

### 2019-12-09 papers/marlowe2019/paper.tex
Fixes
Review of overview and introduction

@@
 Marlowe is specified by a reference semantics for the language written in Haskell, and we can use that  in a number of ways.
-Obviously, we can interpret Marlowe contracts in Haskell itself, but we can also use that implementation, translated into Plutus~\cite{PlutusPlatform}, to interpret  Marlowe directly on the Cardano blockchain, as explained in Section \ref{sec:implementation}.
-Moreover, we can execute that semantics -- translated into PureScript -- directly in a browser to give an interactive simulation environment, the Meadow Playground, described in Section \ref{sec:playground}.
+Obviously, we can interpret Marlowe contracts in Haskell itself, but we can also use that implementation, compiled into Plutus~\cite{PlutusPlatform}, to interpret  Marlowe directly on the Cardano blockchain, as explained in Section \ref{sec:implementation}.
+Moreover, we can execute that semantics -- translated into PureScript -- directly in a browser, to give an interactive simulation environment, the Meadow Playground, described in Section \ref{sec:playground}.
 
 Because Marlowe is a DSL, we are able to build special purpose tools and techniques to support it.
-Crucially in a financial environment we are able to \emph{exhaustively analyse contracts} without executing them, so that we can, for instance, check whether any particular contract makes all the payments it should: in the case it does not, we get an explicit example of how it can fail.
+Crucially, in a financial environment we are able to \emph{exhaustively analyse contracts} without executing them, so that we can, for instance, check whether any particular contract is able to make all the payments it should: in the case it is not, we get an explicit example of how it can fail.
 This analysis, reported in Section \ref{sec:analysis}, is built into the Marlowe Playground.
 This section also covers our use of formal verification to prove properties of the implementation of Marlowe, including a guarantee that ``money in $=$ money out'' for all contracts.
+% PLS: Should we divide the analysis/formalisation section in two?
 
@@
 
-Since first publication we have revised the language design: this section gives a brief overview of the current (3.0) version of the language and its semantics.
+Since the first publication we have revised the language design: this section gives a brief overview of the current (3.0) version of the language and its semantics.
 
@@
 The parties to the contract, also called the participants, can engage in various actions: they can be asked to deposit money, or to make a choice between various alternatives.
-A notification of an external / oracle value is the other possible form of input.
-Running a contract will also produce external effects, by making payments to parties in the contract.
+In some cases, any party will be able to trigger the contract just to notify it that some condition has become true (for example, a timeout has occurred).
 
@@
 
-In executing a contract we need to keep track of the current contract: after making a step in the example above, the current contract is \texttt{cont}.
-We also have to keep track of some other information, such as how much is held in each account: this information together is the state, and generally this changes at each step.
+In executing a contract, we need to keep track of the current contract: after making a step in the example above, the current contract would be \texttt{cont}.
+We also have to keep track of some other information, such as how much is held in each account: this information together is the state, which generally changes at each step.
 A step can also see an action taking place, such as money being deposited, or an effect being produced, e.g.
@@
 \paragraph{Pay}
-A payment contract \texttt{Pay a p v} cont will make a payment of value \texttt{v} from the account \texttt{a} to a payee \texttt{p}, which will be one of the contract participants or another account in the contract.
-Warnings will be generated if the value \texttt{v} is negative, or if there is not enough in the account to make the payment in full.
-In that case a partial payment (of all the money available) is made.
-The continuation contract is the one given in the contract: \texttt{cont}.
+A payment contract \texttt{Pay a p v cont} will make a payment of value \texttt{v} from the account \texttt{a} to a payee \texttt{p}, which will be one of the contract participants or another account in the contract.
+Warnings will be generated if the value \texttt{v} is not positive, or if there is not enough in the account to make the payment in full.
+In the first case, nothing will be transferred; in the later case, a partial payment (of all the money available) is made.
+The contract will continue as specified by \texttt{cont}.
 
@@
 A contract \texttt{Close} provides for the contract to be closed (or terminated).
-The only action that is performs is to provide refunds to the owners of accounts that contain a positive balance.
+The only action that is performed is to refund the contents of each account to their respective owners.
 This is performed one account per step, but all accounts will be refunded in a single transaction.
@@
 \paragraph{Values, observations and actions}
-\emph{Values} include some quantities that change with time, including ``the current slot number'', ``the current balance of an account (in Lovelace)'', and any choices that have already been made; we call these volatile values.
+\emph{Values} include some quantities that change with time, including the current slot number, the current balance of an account (in Lovelace), and any choices that have already been made.
+% ; we call these volatile values.
+- PLS: this definition is no longer used
 Values can also be combined using addition, subtraction and negation.
@@
 This is the most complex constructor for contracts, with the form \texttt{When cases timeout cont}.
-It is a contract that is triggered on actions, which may or may not happen at any particular slot: what happens when various actions happen is described by the cases in the contract.
+It is a contract that is triggered on actions, which may or may not happen at any particular slot: the permitted actions and their consequences are described by \texttt{cases}.
 
-In the contract \texttt{When cases timeout cont}, the list \texttt{cases} contains a collection of cases.
-Each case has the form \texttt{Case ac co} where \texttt{ac} is an action and \texttt{co} a continuation (another contract).
-When a particular action, e.g.
-\texttt{ac}, happens, the state is updated accordingly and the contract will continue as the corresponding continuation \texttt{co}.
+The list \texttt{cases} contains a collection of cases of the form \texttt{Case ac co}, where \texttt{ac} is an action and \texttt{co} a continuation (another contract).
+When the action \texttt{ac} is performed, the state is updated accordingly and the contract will continue as described by \texttt{co}.
 
-In order to make sure that the contract makes progress eventually, the contract \texttt{When cases timeout cont} will continue as cont once the \texttt{timeout}, a slot number, is reached.
+In order to make sure that the contract makes progress eventually, the contract \texttt{When cases timeout cont} will continue as \texttt{cont} as soon as any valid transaction is issued after the \texttt{timeout} (a slot number) is reached.
 
@@
     \centering
-    \includegraphics[width=1\textwidth]{figures/Marlowe3-Figures-004.jpeg}
+    \includegraphics[width=0.7\textwidth]{figures/Marlowe3-Figures-004.jpeg}
     \caption{Zero Coupon Bond Contract}
@@
 Currently parties are represented as public keys. We expect a valid signature for each input,
-and make payments to  addresses for trhe public key of a party.
+and make payments to  addresses for the public key of a party.
 In similar matter, we could check proofs of ownership of a particular token.

### 2019-12-09 papers/marlowe2019/paper.tex
Finalising bibliography, Beginning coonclusions.

--- 2019-12-09 papers/marlowe2019/paper.tex
@@
 
-\paragraph{TO DO}
+BitML~\cite{BitML} is a DSL for specifying Marlowe-like contracts that regulate transfers on the Bitcoin blockchain.
+BitML is implemented via a compiler that translates contracts into standard Bitcoin transactions.
+Participants execute a contract by appending these transactions on the Bitcoin blockchain, according to their strategies, which also involve the exchange of bitstrings that guarantee to a very high probability the correctness of contract execution.
+Marlowe is directly implemented by an interpreter which could also be implemented on a covenant-based~\cite{covenants} extension of the Bitcoin blockchain.
 
-Add stuff on BitML, DAML, \ldots.
+\section{Conclusions and future work}
 
-\section{Future work and conclusions}
+Conclusions (draft)
+\begin{itemize}
+\item
+Using a DSL means that we can d o much better with analysis and proof than a general purpose language.
+\item
+Defining the language by means of an executable reference semantics means that we can, as well as directly executing this semantics, generate an on-chain interpreter for it and simulate it in browser using the Haskell-like languages Plutus and PrueScript.
+This is made particularly straightforward by working in the subset of Haskell that directly corresponds to what translates directly into these languages.
+\item
+
+\end{itemize}
 

### 2019-12-09 papers/marlowe2019/paper.tex
Draft conclusions written.

@@
 \item
-Using a DSL means that we can d o much better with analysis and proof than a general purpose language.
+Using a DSL means that we can do much better with analysis and proof than a general purpose language: so much that we can incorporate the analysis in the development environment.
 \item
@@
 \item
+Different development modes: visual and Haskell-aided.
+\item
+Something about ACTUS?
+\item \ldots
+\end{itemize}
 
+\medskip\noindent
+Future Work (draft)
+\begin{itemize}
+\item Tokenising and securitising contracts.
+\item Other implementations e.g. Ethereum
+\item Multi-currency
+\item Launching on Cardano blockchain itself 
+\item \ldots
 \end{itemize}

### 2019-12-09 papers/marlowe2019/paper.tex
New sections,

@@
 Crucially, in a financial environment we are able to \emph{exhaustively analyse contracts} without executing them, so that we can, for instance, check whether any particular contract is able to make all the payments it should: in the case it is not, we get an explicit example of how it can fail.
-This analysis, reported in Section \ref{sec:analysis}, is built into the Marlowe Playground.
-This section also covers our use of formal verification to prove properties of the implementation of Marlowe, including a guarantee that ``money in $=$ money out'' for all contracts.
-% PLS: Should we divide the analysis/formalisation section in two?
+This analysis, reported in Section \ref{sec:static}, is built into the Marlowe Playground.
+We are also able to use formal verification to prove properties of the implementation of Marlowe, including a guarantee that ``money in $=$ money out'' for all contracts; this is covered in Section~\ref{sec:verification}.
+% PLS: Should we divide the analysis/formalisation section in two? SJT: done
 
@@
 it is impossible for any possible contract and inputs to result in non-positive balances.
-This is described in more detail in Section \ref{sec:analysis}.}
+This is described in more detail in Section \ref{sec:static}.}
 
@@
 
-\section{Analysing Marlowe\label{sec:analysis}}
+\section{Static analysis of contracts\label{sec:static}}
 
-In this section, we give an overview of some of the work we have carried out with the aim of helping users be more confident that the Marlowe contracts they write do what they expect them to do.
-
-We look at two complementary approaches: static analysis of concrete contracts (Section~\ref{subsec:static}) and formal verification of general properties about Marlowe (Section~\ref{subsec:verification}).
-
-\subsection{Static analysis of contracts\label{subsec:static}}
+In the next two sections we give an overview of some of the work we have carried out with the aim of helping users be more confident that the Marlowe contracts they write do what they expect them to do.
+We look at two complementary approaches: static analysis of concrete contracts in this section,  and formal verification of general properties about Marlowe in the next.
 
@@
 
-\subsection{Formal verification of semantics\label{subsec:verification}}
+\section{Formal verification of semantics\label{sec:verification}}
 
@@
     \item It reduces the danger of Denegation of Service (DoS) attacks: because the number of valid inputs is limited, an attacker partcipant cannot arbitrarily block the contract by issuing an unbounded amount of useless \texttt{Transaction}s.
-    \item It bounds the length of traces (the number of transactions) that symbolic execution (see Section~\ref{subsec:static}) needs to explore (more on this below).
+    \item It bounds the length of traces (the number of transactions) that symbolic execution (see Section~\ref{sec:static}) needs to explore (more on this below).
 \end{itemize}

### 2019-12-10 papers/marlowe2019/paper.tex
Edit of section 5 (verification).

@@
 
-\section{Formal verification of semantics\label{sec:verification}}
+\section{Formal verification of the Marlowe semantics\label{sec:verification}}
 
-We can also use proof assistants to demonstrate that Marlowe semantics present certain desirable properties (like that money is preserved and returned to users by the end of the execution of any contract).
+We can also use proof assistants to demonstrate that the Marlowe semantics presents certain desirable properties, such as that money is preserved and anything unspent is returned to users by the end of the execution of any contract.
 In the same way, we can also use this technique to prove properties for particular contracts, or for particular contract snippets, templates, or classes of contracts.
@@
     \item We use a \emph{custom implementation of maps and sets} that uses lists of tuples and lists of elements respectively.
-We do this because Isabelle already provides many theorems that are proven for lists, so it considerably facilitates the automation of certain proofs.
+We do this because Isabelle already provides many theorems that are proved for lists, so it considerably facilitates the automation of certain proofs.
 \end{itemize}
@@
 
-As part of the formalisation of the semantics, Isabelle proves termination automatically for most functions.
-The only function for which termination is not trivial is \texttt{reductionLoop}.
-The reason why Isabelle cannot infer termination for this function is that it repeatedly calls \texttt{reduceContractStep} until it returns \texttt{NotReduced}, so it requires proving that \texttt{reduceContractStep} will eventually do so.
+As part of the formalisation process, Isabelle requires functions to be terminating.
+While  Isabelle proves termination automatically for most functions in the semantics, that is not the case for \texttt{reductionLoop}.
+The reason why Isabelle cannot infer termination for this function is that it repeatedly calls \texttt{reduceContractStep} until it returns \texttt{NotReduced}, so proving overall termination  requires a proof that \texttt{reduceContractStep} will eventually do that.
 
-The way we proved that this is the case was by establishing a measurment for the size of a pair of \texttt{Contract} and \texttt{State}. We defined such measurement as follows:
+The way we proved that was by establishing a measure for the size of a pair of \texttt{Contract} and \texttt{State}. We defined such a measure as follows:
 
@@
 \end{verbatim}
+\noindent
+where \texttt{size} is a measure already generated automatically by Isabelle for custom data types.
 
-\texttt{size} is a measurement already generated automatically by Isabelle for custom data types.
+The reason that we need the number of accounts in the \texttt{State} as part of the measure is that \texttt{reduceContractStep}, when called with the contract \texttt{Close}, will refund one of the accounts and the contract will remain as \texttt{Close}, so its \texttt{size} does not decrease.
 
-The reason why we need the number of accounts in the \texttt{State} for the measurement is that \texttt{reduceContractStep}, when called with the contract \texttt{Close}, will refund one of the accounts and the contract will remain as \texttt{Close}, so its \texttt{size} does not decrease.
-
-And the reason why we multiply the size of the \texttt{Contract} by two is that the primitive \texttt{Deposit} may increase the number of accounts in the \texttt{State} by one, but it will also reduce the size of \texttt{Contract} by at least one.
-By multiplying the size of the \texttt{Contract} by two, the decrease in the \texttt{size} of the \texttt{Contract} compensates the increase in the \texttt{length} of the list of accounts and, thus, it ensures that the overall measurement \texttt{evalBound} will decrease even in that scenario.
+The reason that we also multiply the size of the \texttt{Contract} by two is that the primitive \texttt{Deposit} may increase the number of accounts in the \texttt{State} by one, but it will also reduce the size of \texttt{Contract} by at least one.
+By multiplying the size of the \texttt{Contract} by two, the decrease in the \texttt{size} of the \texttt{Contract} compensates the increase in the \texttt{length} of the list of accounts and, thus, it ensures that the overall measure \texttt{evalBound} will decrease even in that scenario.
 
@@
 
-Since we use lists of pairs of integers as maps in our \texttt{State}, there are some values that are allowed by the type definition of \texttt{State} still make no sense given what these lists represent:
+Since we use lists of pairs of integers as maps in our \texttt{State}, there are some values that are allowed by the type definition of \texttt{State} that make no sense as maps:
 
 \begin{enumerate}
-    \item The lists of tuples are representing maps, so it does not make sense for two elements of the list to be tuples with the same first element (repeated keys).
-    \item We want to minimise the number of ways in which a map can be represented as a list, so we require the tuples in the list to be sorted ascendingly by their first element.
+    \item The lists of tuples should represen maps, so it does not make sense for two elements of the list to be tuples with the same first element (repeated keys).
+    \item We want to represent each map in a unique way as a list, and so we require the tuples in the list to be sorted in ascending order by their first element.
 This allows us to use standard equality for comparing maps.
-    \item We only expect accounts to store positive amounts money, so the second element of the tuples of the list of accounts must be a positive integer.
-Whenever an account becomes empty we just delete the account from the list.
+    \item We only expect accounts to store positive amounts of money, so the second element of each tuple in the list must be a positive integer.
+Whenever an account becomes empty we  delete the corresponding pair  from the list.
 \end{enumerate}
-
-If a value of \texttt{State} satisfies the first two properties above we call it a \texttt{validState}. And we formalised the third property above as follows:
+\noindent
+If a value of \texttt{State} satisfies the first two properties above we call it a \texttt{validState}; we also formalised the third property above as follows:
 
@@
 \end{verbatim}
-
+\noindent
 For a state to have positive accounts, the above function must evaluate to \texttt{True} for any \texttt{AccountId} and \texttt{Token}.
 
-We have proven that if all the three properties are true for a given input \texttt{State}, then the properties will also be true for the resulting output \texttt{State}, when processing any \texttt{Transaction} on any \texttt{Contract}.
+We have proved that if all the three properties are true for a given input \texttt{State}, then the properties will also be true for the resulting output \texttt{State}, when processing any \texttt{Transaction} on any \texttt{Contract}.
 
@@
 There is a class of combinations of \texttt{State} and \texttt{Contract} that will not evolve when passed to the semantics unless a matching \texttt{Input} is passed as well.
-We call these combinations of \texttt{Contract} and \texttt{State} \textit{quiescent}, and we define it as follows:
+We call these combinations of \texttt{Contract} and \texttt{State} \textit{quiescent}, and we define them as follows:
 
@@
 \end{verbatim}
-
-We have proven that, if an input \texttt{State} is valid and all accounts are positive, then the combination of \texttt{State} and \texttt{Contract} resulting from processing any transaction will always be quiescent.
+\noindent
+We have proved that, if an input \texttt{State} is valid and all accounts are positive, then the combination of \texttt{State} and \texttt{Contract} resulting from processing any transaction will always be quiescent.
 
@@
 
-One of the dangers of using smart-contracts is that a badly written one can potentially lock its funds forever.
+One of the dangers of using smart contracts is that a badly written one can potentially lock its funds forever.
 We want to avoid that feature in Marlowe altogether and, with that aim, we proved that Marlowe account system does not create or destroy any money.
 By the end of the contract, all the money paid to the contract must be distributed back, in some way, to a subset of the participants of the contract.
-
 To ensure this is the case we proved two properties:
@@
 \end{verbatim}
-
+\noindent
 The only consideration is that we only accept positive \texttt{Deposit}s, so we interpret money in \texttt{Deposit} inputs as the maximum of $0$ and the actual value specified in the \texttt{Deposit} input.
@@
 
-For every Marlowe \texttt{Contract} there exists a slot number after which an empty transaction can be issued that will close the contract and refund all the money in its accounts.
+For every Marlowe \texttt{Contract} there is a slot number after which an empty transaction can be issued that will close the contract and refund all the money in its accounts.
 
-A conservative upper bound for the expiration slot number can be calculated efficiently by using the function \texttt{maxTime} (or \texttt{maxTimeContract} in the Isabelle semantics).
+A conservative upper bound for the expiration slot number can be calculated efficiently by using the function \texttt{maxTime} (or \texttt{maxTimeContract} in the Isabelle semantics), essentially by taking the lowest upper bound of all the timeouts in the contract.
 
@@
 \noindent
-
 where \texttt{isClosedAndEmpty} is defined as follows:
@@
 \end{verbatim}
-
-Because the money is preserved and we know that there is an empty transaction that can be sent at any point after \texttt{maxTime} that will empty the accounts of the contract: we know that money will not be locked in the contract forever.
+\noindent
+Because the money is preserved and we know that there is an empty transaction that can be sent at any point after \texttt{maxTime} that will empty the accounts of the contract, we can conclude that money will not be locked in the contract forever.
 
@@
 
-\begin{itemize}
-    \item It reduces the danger of Denegation of Service (DoS) attacks: because the number of valid inputs is limited, an attacker partcipant cannot arbitrarily block the contract by issuing an unbounded amount of useless \texttt{Transaction}s.
-    \item It bounds the length of traces (the number of transactions) that symbolic execution (see Section~\ref{sec:static}) needs to explore (more on this below).
+\begin{itemize}%SJT: DoS usually means denial of service?
+    \item It reduces the danger of Denegation of Service (DoS) attacks: because the number of valid inputs is limited, an attacker participant cannot arbitrarily block the contract by issuing an unbounded amount of useless \texttt{Transaction}s.
+    \item The number of transactions bounds the length of traces that symbolic execution (see Section~\ref{sec:static}) needs to explore (more on this below).%SJT: is this really below? Or above?
 \end{itemize}
@@
 \noindent
-where \texttt{maxTransactionsInitialState} is essentially the maximum number of nested \texttt{When} in the contract plus one.
+where \texttt{maxTransactionsInitialState} is essentially the maximum number of nested \texttt{When} constructs in the contract plus one.
 
 This property implies that any trace that is longer than this is guaranteed to produce at least one error.
-And because transactions that produce an error do not alter the state of the contract, such a list of transactions (trace) will be equivalent to a list of transactions that does not have the erroneous transaction.
+Because transactions that produce an error do not alter the state of the contract, such a list of transactions (a trace) will be equivalent to a list of transactions that does not have the erroneous transaction.
 
 In other words, for every list of \texttt{Transaction}s, there exists at least one list of \texttt{Transaction}s with length less or equal to \texttt{maxTransactionsInitialState} that has the same effect on the contracts.
-
 Thus, we do not lose generality by only exploring traces with length of at most \texttt{maxTransactionsInitialState}, because any longer traces produce the same effect.

### 2019-12-10 papers/marlowe2019/paper.tex
stuff
Fix position of comma

@@
 %
-\institute{IOHK, Hong Kong, \url{https://iohk.io} \\
+\institute{IOHK, Hong Kong, \\
 \email{\{alexander.nemish, pablo.lamela, david.smith, simon.thompson\}@iohk.io} \and
 School of Computing, University of Kent, UK\\
-\email{s.j.thompson@kent.ac.uk}}
+\email{s.j.thompson@kent.ac.uk}
+}
 %
@@
 Because Marlowe is a DSL, we are able to build special purpose tools and techniques to support it.
-Crucially, in a financial environment we are able to \emph{exhaustively analyse contracts} without executing them, so that we can, for instance, check whether any particular contract is able to make all the payments it should: in the case it is not, we get an explicit example of how it can fail.
+Crucially in a financial environment, we are able to \emph{exhaustively analyse contracts} without executing them, so that we can, for instance, check whether any particular contract is able to make all the payments it should: in the case it is not, we get an explicit example of how it can fail.
 This analysis, reported in Section \ref{sec:static}, is built into the Marlowe Playground.

### 2019-12-10 papers/marlowe2019/paper.tex
suggested changes by david

@@
 \begin{abstract}
-Marlowe is a DSL for financial contracts running on the Cardano blockchain.
-We describe the implementation of Marlowe on blockchain, and the Marlowe Playground web-based development and simulation environment.
+Marlowe is a DSL for financial contracts. We describe the implementation of Marlowe on the Cardano blockchain, and the Marlowe Playground web-based development and simulation environment.
 
-Contracts in Marlowe can be exhaustively analysed prior to running them, thus providing strong guarantees to participants in the contracts.
+Contracts in Marlowe can be exhaustively analysed prior to running them, thus providing strong guarantees to participants in the contract.
 The Marlowe system itself has been formally verified using the Isabelle theorem prover, establishing such properties as the conservation of money.
@@
 
-Marlowe\footnote{The complete Marlowe project, including semantics and implementation, is available at \url{https://github.com/input-output-hk/marlowe}.} is a domain-specific language (DSL) for implementing financial contracts on blockchain: our initial target is Cardano, but it can be implemented on many DLT platforms, including Ethereum.
+Marlowe\footnote{The complete Marlowe project, including semantics and implementation, is available at \url{https://github.com/input-output-hk/marlowe}.} is a domain-specific language (DSL) for implementing financial contracts on blockchain: our initial target is Cardano, but it could be implemented on many DLT platforms, including Ethereum.
 Marlowe is embedded in Haskell, allowing users selectively to use aspects of Haskell -- typically definitions of constants and simple functions -- to express contracts more readably and succinctly.
-Marlowe was introduced in the paper~\cite{isola-marlowe}: Section \ref{sec:overview} below gives an overview of the language, and the changes made to it since the publication of that paper.
+Section \ref{sec:overview} gives an overview of the language, and the changes made to it since it was originally introduced in ~\cite{isola-marlowe}.
 
 Marlowe is specified by a reference semantics for the language written in Haskell, and we can use that in a number of ways.
-Obviously, we can interpret Marlowe contracts in Haskell itself, but we can also use that implementation, compiled into Plutus~\cite{PlutusPlatform}, to interpret  Marlowe directly on the Cardano blockchain, as explained in Section \ref{sec:implementation}.
-Moreover, we can execute that semantics -- translated into PureScript -- directly in a browser, to give an interactive simulation environment, the Meadow Playground, described in Section \ref{sec:playground}.
+We can interpret Marlowe contracts in Haskell itself, but we can also use that implementation, compiled into Plutus~\cite{PlutusPlatform}, to interpret Marlowe directly on the Cardano blockchain, see Section \ref{sec:implementation}.
+We can also execute the semantics -- translated into PureScript -- directly in a browser, to give an interactive simulation environment, see Section \ref{sec:playground}.
 
@@
 This analysis, reported in Section \ref{sec:static}, is built into the Marlowe Playground.
-We are also able to use formal verification to prove properties of the implementation of Marlowe, including a guarantee that ``money in $=$ money out'' for all contracts; this is covered in Section~\ref{sec:verification}.
-% PLS: Should we divide the analysis/formalisation section in two? SJT: done
+Finally, we are able to use formal verification to prove properties of the implementation of Marlowe, including a guarantee that ``money in $=$ money out'' for all contracts; see Section~\ref{sec:verification}.
 
@@
 
-Since the first publication we have revised the language design: this section gives a brief overview of the current (3.0) version of the language and its semantics.
+Since the first publication, we have revised the language design: this section gives a brief overview of the current (3.0) version of the language and its semantics.
 
@@
 
-Marlowe is designed to support the execution of financial contracts on blockchain, and specifically to work on Cardano.
 Contracts are built by putting together a small number of constructs that in combination can be used to describe many different kinds of financial contract.
@@
 \subsubsection*{Marlowe step by step}
-Marlowe has five ways of building contracts.
-Four of these -- \texttt{Pay}, \texttt{Let}, \texttt{If} and \texttt{When} -- build a complex contract from simpler contracts, and the fifth, \texttt{Close}, is a simple contract.
-At each step of execution, as well as returning a new state and continuation contract, it is possible that effects, e.g.\ payments, and warnings can be generated too.
-We also explain Marlowe values, observations and actions, which are used to supply external information and inputs to a running contract to control how it will evolve.
 
-\paragraph{Pay}
-A payment contract \texttt{Pay a p v cont} will make a payment of value \texttt{v} from the account \texttt{a} to a payee \texttt{p}, which will be one of the contract participants or another account in the contract.
+Marlowe has five ways of building contracts, we call these \emph{contract constructs}. \emph{Contract constructs}, in turn, can also contain \emph{values}, \emph{observations} and \emph{actions}.
+
+\paragraph{Values, observations and actions} are used to supply external information and inputs to a running contract to control how it will evolve.
+
+\texttt{Value}s include some quantities that change with time, including the current slot number, the current balance of an account (in Lovelace), and any choices that have already been made.
+% ; we call these volatile values.
+- PLS: this definition is no longer used
+Values can also be combined using addition, subtraction and negation.
+
+\texttt{Observation}s are Boolean expressions that compare values, and can be combined using the standard Boolean operators.
+It is also possible to observe whether any choice has been made (for a particular identified choice).
+Observations will have a value at every step of execution.
+
+\texttt{Action}s happen at particular points during execution and can be (i) depositing money, (ii) making a choice between various alternatives, or (iii) notifying the contract that a certain observation has become true.
+
+\paragraph{Contract constructs} are the main building block of contracts, and there are five of them: four of these -- \texttt{Pay}, \texttt{Let}, \texttt{If} and \texttt{When} -- build a complex contract from simpler contracts, and the fifth, \texttt{Close}, is a simple contract.
+At each step of execution we will obtain a new state and continuation contract and, in some it is possible that effects, like payments and warnings, can be generated too.
+
+\texttt{Pay}: A payment contract \texttt{Pay a p v cont} will make a payment of value \texttt{v} from the account \texttt{a} to a payee \texttt{p}, which will be one of the contract participants or another account in the contract.
 Warnings will be generated if the value \texttt{v} is not positive, or if there is not enough in the account to make the payment in full.
@@
 
-\paragraph{Close}
-A contract \texttt{Close} provides for the contract to be closed (or terminated).
+\texttt{Close}: A contract \texttt{Close} provides for the contract to be closed (or terminated).
 The only action that is performed is to refund the contents of each account to their respective owners.
 This is performed one account per step, but all accounts will be refunded in a single transaction.
+All contracts eventually reduce to \texttt{Close}.
 
@@
 
-\paragraph{Values, observations and actions}
-\emph{Values} include some quantities that change with time, including the current slot number, the current balance of an account (in Lovelace), and any choices that have already been made.
-% ; we call these volatile values.
-- PLS: this definition is no longer used
-Values can also be combined using addition, subtraction and negation.
+\texttt{If}: The conditional \texttt{If obs cont1 cont2} will continue as \texttt{cont1} or \texttt{cont2}, depending on the Boolean value of the observation \texttt{obs} on execution.
 
-\emph{Observations} are Boolean values derived by comparing values, and can be combined using the standard Boolean operators.
-It is also possible to observe whether any choice has been made (for a particular identified choice).
-
-Observations will have a value at every step of execution. On the other hand, \emph{actions} happen at particular points during execution and can be
-(i) depositing money, (ii)
-making a choice between various alternatives, or (iii)
-notifying an external or oracle value, such as the current price of a particular commodity.
-
-\paragraph{If}
-The conditional \texttt{If obs cont1 cont2} will continue as \texttt{cont1} or \texttt{cont2}, depending on the Boolean value of the observation \texttt{obs} on execution.
-
-\paragraph{When}
-This is the most complex constructor for contracts, with the form \texttt{When cases timeout cont}.
+\texttt{When}: This is the most complex constructor for contracts, with the form \texttt{When cases timeout cont}.
 It is a contract that is triggered on actions, which may or may not happen at any particular slot: the permitted actions and their consequences are described by \texttt{cases}.
@@
 
-\paragraph{Let}
-A let contract \texttt{Let id val cont} causes the expression \texttt{val} to be evaluated, and stored with the name \texttt{id}. The contract then continues as \texttt{cont}.
+\texttt{Let}: A let contract \texttt{Let id val cont} causes the expression \texttt{val} to be evaluated, and stored with the name \texttt{id}. The contract then continues as \texttt{cont}.
 
@@
 one must create a transaction (a \emph{PendingTx}) that has an \emph{input} that references the transaction output,
-and contains a \emph{Redeemer}, such that
-
-\begin{verbatim}
-    Validator(Data Script, Redeemer, PendingTx)
-\end{verbatim}
-evaluates to $True$.
+and contains a \emph{Redeemer}, such that \texttt{Validator(Data Script, Redeemer, PendingTx)} evaluates to $True$.
 In order to spend a transaction output protected by a \emph{public key}, one must provide a valid signature
@@
 The continuation must have the same validator, hence the same hash of the validator script.
-It should contain both the expected continuation contract and then state, evaluated by the \emph{computeTransaction}
+It should contain both the expected continuation contract and the state, evaluated by the \emph{computeTransaction}
 function.
@@
 
-When a contract evaluates to \emph{Close}, and all remaining balances must payed out
+When a contract evaluates to \emph{Close}, all remaining balances must payed out
 to respective owners, removing the contract from the set of unspent transaction outputs.
@@
 
-Conclusions (draft)
-\begin{itemize}
-\item
-Using a DSL means that we can do much better with analysis and proof than a general purpose language: so much that we can incorporate the analysis in the development environment.
-\item
-Defining the language by means of an executable reference semantics means that we can, as well as directly executing this semantics, generate an on-chain interpreter for it and simulate it in browser using the Haskell-like languages Plutus and PrueScript.
+Rather than aiming to be general-purpose, Marlowe is a DSL designed to support financial contracts on blockchain.
+We leverage its specificity in our work on static analysis and verification, where we are able to deliver much greater impact and focus than we could for a general-purpose language.
+We are able to shape the development and simulation environment to give stronger user support too.
+Moreover, Marlowe  presents a model for how other DSLs can be built in this space, supporting different domains such as provenance in the supply chain.
+
+Defining the language by means of an executable reference semantics means that we can, as well as directly executing this semantics, generate an on-chain interpreter for it and simulate it in browser using the Haskell-like languages Plutus and PureScript.
 This is made particularly straightforward by working in the subset of Haskell that directly corresponds to what translates directly into these languages.
-\item
-Different development modes: visual and Haskell-aided.
-\item
-Something about ACTUS?
-\item \ldots
-\end{itemize}
 
-\medskip\noindent
-Future Work (draft)
-\begin{itemize}
-\item Tokenising and securitising contracts.
-\item Other implementations e.g. Ethereum
-\item Multi-currency
-\item Launching on Cardano blockchain itself 
-\item \ldots
-\end{itemize}
+Our medium term aim is launching on Cardano blockchain itself, by which time we expect to have added multiple currencies to Marlowe, as well as making (roles in) Marlowe contracts tradeable, through tokenising contract roles.
 

### 2019-12-10 papers/marlowe2019/paper.tex
Restructuring

--- 2019-12-10 papers/marlowe2019/paper.tex
@@
 \subsubsection*{Marlowe step by step}
-Marlowe has five ways of building contracts.
-Four of these -- \texttt{Pay}, \texttt{Let}, \texttt{If} and \texttt{When} -- build a complex contract from simpler contracts, and the fifth, \texttt{Close}, is a simple contract.
-At each step of execution, as well as returning a new state and continuation contract, it is possible that effects, e.g.\ payments, and warnings can be generated too.
-We also explain Marlowe values, observations and actions, which are used to supply external information and inputs to a running contract to control how it will evolve.
 
-\paragraph{Pay}
-A payment contract \texttt{Pay a p v cont} will make a payment of value \texttt{v} from the account \texttt{a} to a payee \texttt{p}, which will be one of the contract participants or another account in the contract.
+Marlowe has five ways of building contracts, we call these \emph{contract constructs}. \emph{Contract constructs}, in turn, can also contain \emph{values}, \emph{observations} and \emph{actions}.
+
+\paragraph{Values, observations and actions} are used to supply external information and inputs to a running contract to control how it will evolve.
+
+\texttt{Value}s include some quantities that change with time, including the current slot number, the current balance of an account (in Lovelace), and any choices that have already been made.
+% ; we call these volatile values.
+- PLS: this definition is no longer used
+Values can also be combined using addition, subtraction and negation.
+
+\texttt{Observation}s are Boolean expressions that compare values, and can be combined using the standard Boolean operators.
+It is also possible to observe whether any choice has been made (for a particular identified choice).
+
+\texttt{Action}s, unlike \texttt{Observation}s, which have a value at every step of execution, happen at particular points during execution and can be:
+(i) depositing money, (ii)
+making a choice between various alternatives, or (iii)
+notifying the contract that a certain observation has become true.
+
+\paragraph{Contract constructs} are the main building block of contracts, and there are five of them: four of these -- \texttt{Pay}, \texttt{Let}, \texttt{If} and \texttt{When} -- build a complex contract from simpler contracts, and the fifth, \texttt{Close}, is a simple contract.
+At each step of execution we will obtain a new state and continuation contract and, in some it is possible that effects, like payments and warnings, can be generated too.
+
+\texttt{Pay}: A payment contract \texttt{Pay a p v cont} will make a payment of value \texttt{v} from the account \texttt{a} to a payee \texttt{p}, which will be one of the contract participants or another account in the contract.
 Warnings will be generated if the value \texttt{v} is not positive, or if there is not enough in the account to make the payment in full.
@@
 
-\paragraph{Close}
-A contract \texttt{Close} provides for the contract to be closed (or terminated).
+\texttt{Close}: A contract \texttt{Close} provides for the contract to be closed (or terminated).
 The only action that is performed is to refund the contents of each account to their respective owners.

### 2019-12-10 papers/marlowe2019/paper.tex
Conclusions and future work.

--- 2019-12-10 papers/marlowe2019/paper.tex
@@
 Because Marlowe is a DSL, we are able to build special purpose tools and techniques to support it.
-Crucially in a financial environment, we are able to \emph{exhaustively analyse contracts} without executing them, so that we can, for instance, check whether any particular contract is able to make all the payments it should: in the case it is not, we get an explicit example of how it can fail.
+Crucially, in a financial environment we are able to \emph{exhaustively analyse contracts} without executing them, so that we can, for instance, check whether any particular contract is able to make all the payments it should: in the case it is not, we get an explicit example of how it can fail.
 This analysis, reported in Section \ref{sec:static}, is built into the Marlowe Playground.

### 2019-12-10 papers/marlowe2019/paper.tex
Add commas

@@
 It is also possible to observe whether any choice has been made (for a particular identified choice).
+Observations will have a value at every step of execution.
 
-\texttt{Action}s, unlike \texttt{Observation}s, which have a value at every step of execution, happen at particular points during execution and can be:
-(i) depositing money, (ii)
-making a choice between various alternatives, or (iii)
-notifying the contract that a certain observation has become true.
+\texttt{Action}s happen at particular points during execution and can be (i) depositing money, (ii) making a choice between various alternatives, or (iii) notifying the contract that a certain observation has become true.
 
@@
 This is performed one account per step, but all accounts will be refunded in a single transaction.
+All contracts eventually reduce to \texttt{Close}.
 
@@
 Marlowe is specified by an executable semantics written in Haskell, but to make it usable in practice to execute financial contracts, it needs to be implemented on a blockchain.
-In this section we explain how Marlowe is executed on the Cardano blockchain using an interpreter written in the Plutus programming language.
+In this section, we explain how Marlowe is executed on the Cardano blockchain using an interpreter written in the Plutus programming language.
 
@@
 
-To implement Marlowe contracts we use the PlutusTx compiler,
+To implement Marlowe contracts, we use the PlutusTx compiler,
 which compiles Haskell code into serialized \emph{Plutus Core} code,

### 2019-12-10 papers/marlowe2019/paper.tex
Some fixes

@@
 makes it easier to implement, review, and test what we have done.
-\item It is close to the semantics of Marlowe, as sketched above and in more detail in~\cite{isola-marlowe},
-so making it easier to validate.
-\item It means that the same implementation can be used for both on- and off-chain (wallet) execution of Marlowe code.
-\item It allows client-side contract evaluation, where we reuse the same code
+\item Implementation is close to the semantics of Marlowe, as sketched above and in more detail in~\cite{isola-marlowe},
+which makes it easier to validate.
+\item The same implementation can be used for both on- and off-chain (wallet) execution of Marlowe code.
+\item It facilitates client-side contract evaluation, where we reuse the same code
 to do contract execution emulation in an IDE, and compile it to WASM/JavaScript on the client side,
@@
   to monitor the blockchain for these contracts, if required.
-\item Finally, there is a potential to special-case this sort of script,
-  and implement a specialized, highly effective interpreter in the Cardano node itself.
+\item Finally, Cardano nodes could potentially use an optimised interpreter (e.g:~native) just for Marlowe contracts, which would save processing time.
 \end{itemize}
@@
 Each step in contract execution is a transaction that spends a Marlowe contract transaction output
-by providing a valid input in $Redeemer$, and produces a transaction output
-with a Marlowe contract as continuation (remaining contract).
+by providing a valid input as \emph{Redeemer}, and produces a transaction output
+with a the remaining Marlowe contract and the updated state.
 
@@
 The Marlowe interpreter first validates the current contract state: i.e.\
-we check that the contract locks at least what it should according
+we check that the contract locks at least what it should, according
 to contract balances (the \emph{accounts} field in \emph{State}),
-and all balances are strictly positive.
+and that all balances are strictly positive.
 \footnote{
@@
 \noindent
-The interpreter reduces a contract until it becomes quiescent, meaning that either it evaluates to \emph{Close},
-or it expects a user input in \emph{When} construct.
-This means that all \emph{Pay, If, Let, Close} constructs are evaluated immediately.
+The interpreter reduces a contract until it becomes quiescent: either it evaluates to \emph{Close},
+or it expects a user input in a \emph{When} construct.
+All \emph{Pay, If, Let, Close} constructs are evaluated immediately.
 
-The function returns a new contract state, contract continuation, a list of warnings (such as partial payments),
-and a list of expected payments (i.e. all the evaluated \emph{Pay} constructs).
+The evaluation function returns a new contract state, contract continuation, a list of warnings (such as partial payments),
+and a list of expected payments (i.e. one for each of the \emph{Pay} constructs evaluated).
 
@@
 \noindent
-Here we expect Alice to deposit 850 Ada (850,000,000 Lovelace) into her $aliceAccount$ before slot 100.
+Here we expect Alice to deposit 850 Ada (850,000,000 Lovelace) into her \texttt{aliceAccount} before slot 100.
 Otherwise, we \emph{Close} the contract.
@@
 \noindent
-Bob is then is expected to deposit 1000 Ada into Alice's account before slot 200.
+Bob is then expected to deposit 1000 Ada into Alice's account before slot 200.
 If he does, the contract is closed, and all remaining balances must be paid out
@@
 allowing as many steps of a contract execution as necessary to be merged.
-This gives atomicity of some operations, and saves on transaction fees.
+This gives atomicity to some operations, and saves on transaction fees.
 
 Sometimes it is necessary to just evaluate a contract in a given moment in time,
-and not provide any inputs. Consider the following example:
+without providing any inputs. Consider the following example:
 

### 2019-12-10 papers/marlowe2019/paper.tex
More changes

@@
 The payment could be obtained only after \emph{When} has timed out after slot 200.
-Buy if we evaluate the contract after slot 200, \emph{If} reduces to \emph{Close}, and
+But if we evaluate the contract after slot 200, \emph{If} reduces to \emph{Close}, and
 Bob will not receive the payment. So he must evaluate the contract before slot 100 to
 reduce \emph{If} to \emph{When}.
-To do that he can spend the contract transaction output with empty list of inputs in \emph{redeemer}.
-He should then do the same after slot 200 to receive the payment.
+To achieve that, Bob must spend the contract transaction output with an empty list of inputs in the \emph{Redeemer}.
+He can then do the same after slot 200 to receive the payment.
 
@@
 
-The Marlowe validator script checks that a transaction contains a valid continuation
-output in case a contract is not closed, by checking the hash of the output validator and state.
-The continuation must have the same validator, hence the same hash of the validator script.
-It should contain both the expected continuation contract and the state, evaluated by the \emph{computeTransaction}
-function.
+Except for the transaction that closes a Marlowe contract, the Marlowe validator script checks that a spending transaction contains a valid continuation
+output, i.e: the hash of the output validator is the same (same hash), and the new state and contract are the expected ones: the ones resulting from applying the \emph{computeTransaction} to the given inputs.
 
@@
 
-When a contract evaluates to \emph{Close}, all remaining balances must payed out
-to respective owners, removing the contract from the set of unspent transaction outputs.
+When a contract evaluates to \emph{Close}, all remaining balances the accounts of the contract are payed out
+to the respective owners of each account, and the contract is removed from the set of unspent transaction outputs.
 
@@
 and make payments to addresses for the public key of a party.
-In similar matter, we could check proofs of ownership of a particular token.
+Analogously, we could instead check proofs of ownership of a particular token.
 
-This allows us to tokenize contract participants, and abstract away concrete public keys into contract \emph{roles}.
-Thus, those roles could be traded independently of a contract.
-Also, we can represent a role with multiple \emph{shares}, allowing automatic distribution of a payment
-proportionally to a number of shares owned.
+This would allows us to tokenize contract participants, and abstract away concrete public keys into contract \emph{roles}.
+In turn, those roles could be traded independently of a contract.
+Additionally, we can split a role in multiple \emph{shares}, which would allow automatic distribution of a payment
+proportionally to the number of shares owned.
 

### 2019-12-10 papers/marlowe2019/paper-long.tex
add long version of 2019 paper

--- 2019-12-10 papers/marlowe2019/paper.tex
@@
 
-\keywords{blockchain \and Cardano \and Haskell \and SMT \and static analysis}
+\keywords{Cardano \and DSL \and functional \and Haskell \and SMT \and static analysis}
 \end{abstract}
@@
 
-Marlowe\footnote{The complete Marlowe project, including semantics and implementation, is available at \url{https://github.com/input-output-hk/marlowe}.} is a domain-specific language (DSL) for implementing financial contracts on blockchain: our initial target is Cardano, but it could be implemented on many DLT platforms, including Ethereum.
+Marlowe\footnote{The complete Marlowe project, including semantics and implementation, is available at \url{https://github.com/input-output-hk/marlowe}.} is a domain-specific language (DSL) for implementing financial contracts on blockchain: our initial target is Cardano, but it could be implemented on many distributed ledgers (DLT platforms), including Ethereum.
 Marlowe is embedded in Haskell, allowing users selectively to use aspects of Haskell -- typically definitions of constants and simple functions -- to express contracts more readably and succinctly.
@@
 We can interpret Marlowe contracts in Haskell itself, but we can also use that implementation, compiled into Plutus~\cite{PlutusPlatform}, to interpret Marlowe directly on the Cardano blockchain, see Section \ref{sec:implementation}.
-We can also execute the semantics -- translated into PureScript -- directly in a browser, to give an interactive simulation environment, see Section \ref{sec:playground}.
+We can also execute the semantics -- translated into PureScript -- directly in a browser, to give an interactive simulation environment, see Section~\ref{sec:playground}.
 

### 2019-12-10 papers/marlowe2019/paper.tex
keywords
Addesses and emails.

@@
 %
-\institute{IOHK, Hong Kong, \\
-\email{\{alexander.nemish, pablo.lamela, david.smith, simon.thompson\}@iohk.io} \and
-School of Computing, University of Kent, UK\\
-\email{s.j.thompson@kent.ac.uk}
-}
+\institute{IOHK, Hong Kong, 
+\email{alexander.nemish@iohk.io, pablo.lamela@iohk.io, simon.thompson@iohk.io,~david.smith@tweag.io},  \and
+School of Computing, University of Kent, UK, 
+\email{s.j.thompson@kent.ac.uk}}
 %
 \maketitle              % typeset the header of the contribution
-%
 
+\vspace{-4mm}
 \begin{abstract}
@@
 
-Marlowe is specified by a reference semantics for the language written in Haskell, and we can use that in a number of ways.
+Marlowe is specified by a reference semantics f or the language written in Haskell, and we can use that in a number of ways.
 We can interpret Marlowe contracts in Haskell itself, but we can also use that implementation, compiled into Plutus~\cite{PlutusPlatform}, to interpret Marlowe directly on the Cardano blockchain, see Section \ref{sec:implementation}.
-We can also execute the semantics -- translated into PureScript -- directly in a browser, to give an interactive simulation environment, see Section \ref{sec:playground}.
+We can also execute the semantics -- translated into PureScript -- directly in a browser, to give an interactive simulation environment, see Section~\ref{sec:playground}.
 

### 2019-12-11 papers/marlowe2019/paper-long.tex
fix david email address
Addesses and emails.
Small fixes and reduce one level in sections 4 and 5

@@
 
-\subsubsection{SBV library}
+\subsection{SBV library}
 
-SBV \cite{SBV} (SMT Base Verification) library provides a high-level API that allows developers to automatically prove properties about Haskell programs, among other functionalities.
+SBV \cite{SBV} (SMT Based Verification) library provides a high-level API that allows developers to automatically prove properties about Haskell programs, among other functionalities.
 The SBV library translates these properties to SMTLib queries, passes them to one or several SMT solvers, and translates the results back to the format in which the queries were written.
 
-\paragraph{\texttt{SBV} monad.}
+\subsubsection{\texttt{SBV} monad}
 
@@
 
-\paragraph{Limitations}
+\subsubsection{Limitations}
 
@@
 
-\subsubsection{Using SBV to analyse Marlowe Contracts}
+\subsection{Using SBV to analyse Marlowe Contracts}
 
@@
 
-\paragraph{Custom datatypes}
+\subsubsection{Custom datatypes}
 
@@
 
-\subsubsection{Termination proof}
+\subsection{Termination proof}
 
@@
 
-\subsubsection{Valid state and positive account preservation}
+\subsection{Valid state and positive account preservation}
 
@@
 
-\subsubsection{Money preservation and contract timeout}
+\subsection{Money preservation and contract timeout}
 
@@
 
-\paragraph{Money preservation}
+\subsubsection{Money preservation}
 Money is not created or destroyed by the semantics. We state this property as follows:
@@
 
-\paragraph{Timeout closes a contract}
+\subsubsection{Timeout closes a contract}
 
@@
 
-\subsubsection{Bound on the maximum number of transactions\label{subsubsec:bound_max_transaction_number}}
+\subsection{Bound on the maximum number of transactions\label{subsubsec:bound_max_transaction_number}}
 

### 2019-12-11 papers/marlowe2019/paper.tex
Remove typo
Fix

@@
 
-Marlowe is specified by a reference semantics f or the language written in Haskell, and we can use that in a number of ways.
+Marlowe is specified by a reference semantics for the language written in Haskell, and we can use that in a number of ways.
 We can interpret Marlowe contracts in Haskell itself, but we can also use that implementation, compiled into Plutus~\cite{PlutusPlatform}, to interpret Marlowe directly on the Cardano blockchain, see Section \ref{sec:implementation}.
@@
 
-In the following section, we explain how we worked around these limitations and give an overview of how we use SBV to avoid runtime problems with Marlowe contracts.
+In the following sections, we explain how we worked around these limitations and give an overview of how we use SBV to avoid runtime problems with Marlowe contracts.
 
-\subsubsection{Using SBV to analyse Marlowe Contracts}
+\subsection{Using SBV to analyse Marlowe Contracts}
 

### 2019-12-11 papers/marlowe2019/paper.tex
Fix typo

@@
 
-In the following sections, we explain how we worked around these limitations and give an overview of how we use SBV to avoid runtime problems with Marlowe contracts.
+In the following section, we explain how we worked around these limitations and give an overview of how we use SBV to avoid runtime problems with Marlowe contracts.
 
@@
 But there are potential problems which are harder to detect until runtime, for example, whether there will be enough money to issue all the payments declared in the contract.
-At this point it may already be too late to fix them, and particularly so in the case of blockchain.
+At this point, it may already be too late to fix them, and particularly so in the case of blockchain.
 The Marlowe semantics represents these errors that can be found at runtime as \texttt{Warnings}.
@@
 
-\paragraph{Returning non-symbolic \texttt{Contract} values}
+\subsubsection{Returning non-symbolic \texttt{Contract} values}
 
@@
 
-\paragraph{Bounds for the state and the inputs}
+\subsubsection{Bounds for the state and the inputs}
 

### 2019-12-11 papers/marlowe2019/paper.tex
Add comma

@@
 
-In the next two sections we give an overview of some of the work we have carried out with the aim of helping users be more confident that the Marlowe contracts they write do what they expect them to do.
+In the next two sections, we give an overview of some of the work we have carried out with the aim of helping users be more confident that the Marlowe contracts they write do what they expect them to do.
 We look at two complementary approaches: static analysis of concrete contracts in this section,  and formal verification of general properties about Marlowe in the next.

### 2019-12-11 papers/marlowe2019/paper.tex
Shorten Section 4

--- 2019-12-11 papers/marlowe2019/paper.tex
@@
 
-In the next two sections we give an overview of some of the work we have carried out with the aim of helping users be more confident that the Marlowe contracts they write do what they expect them to do.
-We look at two complementary approaches: static analysis of concrete contracts in this section,  and formal verification of general properties about Marlowe in the next.
+Marlowe semantics use types to prevent many non-sensical contracts from being written.
+But there are potential problems which are harder to detect until runtime, for example, whether there will be enough money to issue all the payments declared in the contract.
+At that point, it may already be too late to fix them, particularly in the case of blockchain.
 
-For some properties, given a particular contract, a computer can decidably determine whether the contract satisfies the property and, in the cases when it does not, it can provide a counter-example.
-
-For Turing-complete languages, this process is usually undecidable in the general case, and for Marlowe contracts it is often NP-complete in general.
-However, in practice, SMT (Satisfaction Modulo Theories) solvers are often able to solve moderately large instances of this problem reasonably efficiently.
+Fortunately, in the case of Marlowe, a computer can decidably determine whether a particular contract satisfies certain property before executing it, and it can provide a counter-example when it does not.
 
@@
 
-SBV provides a monad called \texttt{SBV} which encapsulates potentially unknown values to the program.
-A program that takes arguments wrapped with the \texttt{SBV} monad can be used with normal arguments and it can also be passed symbolic parameters when used as part of a property that is passed to the solver.
-For example, the following function:
-
-\begin{verbatim}
-primeFactors :: SBV Integer -> SBV Integer -> SBV Bool
-primeFactors x y = x .> 1 .&& y .> 1 .&& x * y .== 25723297
-\end{verbatim}
-
-\noindent
-can be used both with normal integers and with symbolic parameters, that is, parameters whose concrete value is not specified and which the solver will try to replace with values that satisfy or falsify the property:
-
-\begin{verbatim}
-*Main> primeFactors 98 314
-False
-*Main> sat primeFactors
-Satisfiable. Model:
-    s0 = 6353 :: Integer
-    s1 = 4049 :: Integer
-*Main> primeFactors 6353 4049
-True
-\end{verbatim}
-
-\subsubsection{Limitations}
-
-At the time of writing, there are some types of functions and data-types that SBV cannot translate to \texttt{SMTLib}.
-In particular, there are two that affected the way in which we use SBV with Marlowe and that prevent us from using the same version of the semantics for symbolic execution and normal execution.
-
-Firstly, calls to recursive functions that are unbounded cannot be translated.
-As part of the process of converting functions to \texttt{SMTLib} terms, SBV unfolds them into a finite data-structure.
-For this reason, if the function is recursive and is not bounded by a non-symbolic value (a value that is known at the time of translation), the translation process will run out of memory.
-
-Secondly, to the best our knowledge, SBV does not currently directly provide support for representing custom data types symbolically, except for zero-ary datatypes, i.e: enumerations.
-
-In the following section, we explain how we worked around these limitations and give an overview of how we use SBV to avoid runtime problems with Marlowe contracts.
+SBV provides a monad called \texttt{SBV}, a function can use parameters wrapped in this monad to represent symbolic values.
+Functions that take symbolic values can be used as properties and passed to the solver; the solver will then try to replace the symbolic values with concrete values that satisfy or falsify the property.
 
@@
 
-Marlowe semantics are designed to use types to prevent many non-sensical contracts from being written.
-But there are potential problems which are harder to detect until runtime, for example, whether there will be enough money to issue all the payments declared in the contract.
-At this point, it may already be too late to fix them, and particularly so in the case of blockchain.
-The Marlowe semantics represents these errors that can be found at runtime as \texttt{Warnings}.
+Marlowe semantics represents errors that can be found at runtime as \texttt{Warnings}.
 
@@
 
-In order to achieve this, we implemented a symbolic version of the semantics that returns a list of the warnings produced by a trace (a list of transactions input to the contract):
+This property is essentially a symbolic version of the semantics that returns a list of the warnings produced by a symbolic trace (a symbolic list of transactions input to the contract):
 
@@
 
-We can see that this version of the implementation takes two concrete parameters: \texttt{Bounds} (see \textit{Bounds for the state and the inputs} section below) and \texttt{Contract} (the contract we want to analyse); and two symbolic parameters: \texttt{SSlotNumber} (the initial slot number) and \texttt{SList NTransaction} (the input list of transactions); and it returns \texttt{SList NTransactionWarning} (a symbolic list of transaction warnings).
-
 \subsubsection{Custom datatypes}
 
-As we mentioned before, SBV does not currently seem to support in general the use of custom datatypes.
-We do not need to represent the \texttt{Contract} type symbolically because we know its value at the time of analysis, but we still have a number of other types like \texttt{TransactionWarning} that are custom datatypes and we need to use.
-
-Fortunately, SBV does support tuples and the \texttt{Either} type; combinations of those two give us essentially the same functionality as custom datatypes that are not recursive and have no type-parameters.
-We can represent all types that Marlowe requires as combinations of \texttt{Either} and tuples, with the exception of the \texttt{Contract} type, but we do not need a symbolic version of the \texttt{Contract} type.
+SBV does not currently seem to support in general the use of custom datatypes.
+Fortunately, SBV supports tuples and the \texttt{Either} type.
+We can represent all types that Marlowe requires as combinations of \texttt{Either} and tuples, with the exception of the \texttt{Contract} type, but we do not need a symbolic version of the \texttt{Contract} type because we know its value at the time of analysis.
 
@@
 
+\subsubsection{Bounds for the state and the inputs}
+
+The recursion in the execution of the semantics is bounded by the \texttt{Contract}, and because the \texttt{Contract} is not a symbolic parameter, the translation will terminate.
+
+However, in both the input and the \texttt{State} record there are several lists (representing finite maps) that are not explicitly bounded in the implementation.
+Some parts of the semantics are bounded by the length of these lists (or maps), such as the implementation of \texttt{Close}.
+In order for the symbolic implementation to be finite, we need to find a bound for the length of these lists or maps.
+
+Fortunately, we can infer a bound for all this lists quite straightforwardly.
+The least obvious one is the length of the list of transactions; we discuss the proof for this bound in Section~\ref{subsubsec:bound_max_transaction_number}.
+
 \subsubsection{Returning non-symbolic \texttt{Contract} values}
 
-In the non-symbolic semantics, each step typically takes a \texttt{Contract}, a \texttt{State}, and a set of \texttt{Input}s; and it returns an updated \texttt{State} and a remaining \texttt{Contract}.
-
-This is a problem because values that rely on symbolic values have to be themselves symbolic.
-Moreover,  the resulting remaining \texttt{Contract} depends on the \texttt{Input}s and \texttt{State}, which are both symbolic.
+Values that rely on symbolic values have to be themselves symbolic, and the continuation \texttt{Contract} after each step depends on the \texttt{Input}s and \texttt{State}, which are both symbolic.
+But having the contract as a symbolic parameter would be inconvenient since it is recursive, we know it in advance, and we use it to bound the execution of the symbolic semantics.
 
@@
 where \texttt{DetApplyResult} contains the parts of \texttt{ApplyResult} that are not symbolic (like the \texttt{Contract}).
-
-\subsubsection{Bounds for the state and the inputs}
-
-The recursion in the execution of the semantics is bounded by the \texttt{Contract}, and because the \texttt{Contract} is not a symbolic parameter, the translation will terminate.
-
-However, in both the input and the \texttt{State} record there are several lists (representing finite maps) that are not explicitly bounded in the implementation.
-Some parts of the semantics are bounded by the length of these lists (or maps), such as the implementation of \texttt{Close} that refunds any remaining money in the contract accounts.
-In order for the symbolic implementation to be finite, we need to find a bound for the length of these lists and maps.
-
-With the current semantics of Marlowe, all the lists/maps in \texttt{State} and input can be bounded when looking at a particular contract.
-In particular, we can find out quite easily, by looking at the contract, how many parties there are, how many choice identifiers are used, how many accounts are used, and how many let binding identifiers are used.
-
-It is less obvious that we can also set a bound for the input \texttt{Transaction} list as the maximum number of nested \texttt{When}s in the contract (for all the execution paths) plus one.
-We also provide a proof in Isabelle that shows that any longer list of transactions will result in an error (we discuss this in Section~\ref{subsubsec:bound_max_transaction_number}).
 

### 2019-12-11 papers/marlowe2019/paper.tex
Shorten section 5

@@
 We can also use proof assistants to demonstrate that the Marlowe semantics presents certain desirable properties, such as that money is preserved and anything unspent is returned to users by the end of the execution of any contract.
-In the same way, we can also use this technique to prove properties for particular contracts, or for particular contract snippets, templates, or classes of contracts.
 
-Currently, we have translated the Haskell Marlowe semantics to Isabelle as closely as possible. The main differences between the Isabelle and the Haskell semantics are that:
+Currently, we have translated the Haskell Marlowe semantics to Isabelle while keeping both versions as close as possible, but we decided to make them different in two main aspects:
+
 \begin{itemize}
-    \item The Isabelle semantics uses \emph{integers for identifiers} and the Haskell semantics uses strings.
-We do this because handling integers in Isabelle is easier and we only compare them for equality and order (their value is irrelevant).
-    \item We use a \emph{custom implementation of maps and sets} that uses lists of tuples and lists of elements respectively.
-We do this because Isabelle already provides many theorems that are proved for lists, so it considerably facilitates the automation of certain proofs.
+    \item We use \emph{integers for identifiers} because they are easier to handle than strings.
+    \item We use a \emph{custom implementation of maps and sets} that use lists because Isabelle already provides many theorems that are proved for lists.
 \end{itemize}
@@
 
-As part of the formalisation process, Isabelle requires functions to be terminating.
-While  Isabelle proves termination automatically for most functions in the semantics, that is not the case for \texttt{reductionLoop}.
-The reason why Isabelle cannot infer termination for this function is that it repeatedly calls \texttt{reduceContractStep} until it returns \texttt{NotReduced}, so proving overall termination  requires a proof that \texttt{reduceContractStep} will eventually do that.
+Isabelle automatically proves termination for most function.
+This is not the case for \texttt{reductionLoop}.
+This function repeatedly calls \texttt{reduceContractStep} until it returns \texttt{NotReduced}, so proving overall termination requires a proof that \texttt{reduceContractStep} will eventually do that.
 
-The way we proved that was by establishing a measure for the size of a pair of \texttt{Contract} and \texttt{State}. We defined such a measure as follows:
+We proved that by establishing a measure for the size of a pair of \texttt{Contract} and \texttt{State}:
 
@@
 
-The reason that we need the number of accounts in the \texttt{State} as part of the measure is that \texttt{reduceContractStep}, when called with the contract \texttt{Close}, will refund one of the accounts and the contract will remain as \texttt{Close}, so its \texttt{size} does not decrease.
+We need the number of accounts in the \texttt{State} because the size of the contract \texttt{Close} will not decrease when calling \texttt{reduceContractStep}, but the number of accounts will, unless they are all empty.
 
-The reason that we also multiply the size of the \texttt{Contract} by two is that the primitive \texttt{Deposit} may increase the number of accounts in the \texttt{State} by one, but it will also reduce the size of \texttt{Contract} by at least one.
-By multiplying the size of the \texttt{Contract} by two, the decrease in the \texttt{size} of the \texttt{Contract} compensates the increase in the \texttt{length} of the list of accounts and, thus, it ensures that the overall measure \texttt{evalBound} will decrease even in that scenario.
+And we needed to multiply the size of the \texttt{Contract} by two because the primitive \texttt{Deposit} may increase the number of accounts by one, so we need to multiply the effect of the reduction of the size of the contract in order to compensate that.
 
@@
 
-Since we use lists of pairs of integers as maps in our \texttt{State}, there are some values that are allowed by the type definition of \texttt{State} that make no sense as maps:
+There are some values for \texttt{State} that are allowed by its type but make no sense, especially in the case of Isabelle semantics where we use lists instead of maps and sets:
 
 \begin{enumerate}
-    \item The lists of tuples should represen maps, so it does not make sense for two elements of the list to be tuples with the same first element (repeated keys).
-    \item We want to represent each map in a unique way as a list, and so we require the tuples in the list to be sorted in ascending order by their first element.
-This allows us to use standard equality for comparing maps.
-    \item We only expect accounts to store positive amounts of money, so the second element of each tuple in the list must be a positive integer.
-Whenever an account becomes empty we  delete the corresponding pair  from the list.
+    \item The lists represent maps, so they should have no repeated keys.
+    \item We want two maps that are equal to be represented the same, so we force keys to be in ascending order.
+    \item For the map of accounts, we only want to record those that have positive amounts.
 \end{enumerate}
 \noindent
-If a value of \texttt{State} satisfies the first two properties above we call it a \texttt{validState}; we also formalised the third property above as follows:
+We call a value for \texttt{State} \emph{valid} if the two first properties are true. And we say it has \emph{positive accounts} if the third property is true.
 
-\begin{verbatim}
-fun positiveMoneyInAccountOrNoAccount
-             :: "AccountId ⇒ Token ⇒ Accounts ⇒ bool" where
-"positiveMoneyInAccountOrNoAccount accId tok mlist =
-    (case MList.lookup (accId, tok) mlist of
-       None ⇒ True
-     | Some x ⇒ x > 0)"
-\end{verbatim}
-\noindent
-For a state to have positive accounts, the above function must evaluate to \texttt{True} for any \texttt{AccountId} and \texttt{Token}.
-
-We have proved that if all the three properties are true for a given input \texttt{State}, then the properties will also be true for the resulting output \texttt{State}, when processing any \texttt{Transaction} on any \texttt{Contract}.
+We have proved that functions in the semantics preserve all three properties.
 
@@
 
-There is a class of combinations of \texttt{State} and \texttt{Contract} that will not evolve when passed to the semantics unless a matching \texttt{Input} is passed as well.
-We call these combinations of \texttt{Contract} and \texttt{State} \textit{quiescent}, and we define them as follows:
+A contract is \emph{quiescent} if and only if the root construct \texttt{When}, or if the contract is \texttt{Close} and all accounts are empty.
 
-\begin{verbatim}
-fun isQuiescent :: "Contract ⇒ State ⇒ bool" where
-"isQuiescent Close state = (accounts state = [])" |
-"isQuiescent (When _ _ _) _ = True" |
-"isQuiescent _ _ = False"
-\end{verbatim}
-\noindent
-We have proved that, if an input \texttt{State} is valid and all accounts are positive, then the combination of \texttt{State} and \texttt{Contract} resulting from processing any transaction will always be quiescent.
+We have proved that, if an input \texttt{State} is valid and all accounts are positive, then the result from processing any transaction will always be quiescent.
 
@@
 One of the dangers of using smart contracts is that a badly written one can potentially lock its funds forever.
-We want to avoid that feature in Marlowe altogether and, with that aim, we proved that Marlowe account system does not create or destroy any money.
 By the end of the contract, all the money paid to the contract must be distributed back, in some way, to a subset of the participants of the contract.
@@
 \subsubsection{Money preservation}
-Money is not created or destroyed by the semantics. We state this property as follows:
-
-\begin{verbatim}
-lemma computeTransaction_preserves_money :
-"validAndPositive_state state
-  ⟹ moneyInState state + moneyInInputs (inputs tra)
-       = moneyInTransactionOutput state (inputs tra)
-               (computeTransaction tra state contract)"
-\end{verbatim}
-\noindent
-The only consideration is that we only accept positive \texttt{Deposit}s, so we interpret money in \texttt{Deposit} inputs as the maximum of $0$ and the actual value specified in the \texttt{Deposit} input.
+Money is not created or destroyed by the semantics.
+More specifically, the money that comes in plust the money in the contract before the transaction must be equal to the money that comes out plus the contract after the transaction, except in the case of an error, when the contract stays unaltered.
 
@@
 
-A conservative upper bound for the expiration slot number can be calculated efficiently by using the function \texttt{maxTime} (or \texttt{maxTimeContract} in the Isabelle semantics), essentially by taking the lowest upper bound of all the timeouts in the contract.
+A conservative upper bound for the expiration slot number can be calculated efficiently by using the function \texttt{maxTime} (or \texttt{maxTimeContract} in the Isabelle semantics), essentially by taking the maximum of all the timeouts in the contract.
 
-We proved that this conservative upper bound works for every contract in a property stated as follows:
+We proved that this conservative upper bound is general enough for every contract, by showing that, if the contract is not closed and empty, then an empty transaction sent after \texttt{maxTime} will close the contract and leave the accounts empty.
 
-\begin{verbatim}
-theorem timedOutTransaction_closes_contract :
-"validAndPositive_state sta
- ⟹ iniSlot ≥ minSlot sta
- ⟹ iniSlot ≥ maxTimeContract cont
- ⟹ endSlot ≥ iniSlot
- ⟹ accounts sta ≠ [] ∨ cont ≠ Close
- ⟹ isClosedAndEmpty
-         (computeTransaction ⦇ interval = (iniSlot, endSlot)
-                             , inputs = [] ⦈ sta cont)"
-\end{verbatim}
-
-\noindent
-where \texttt{isClosedAndEmpty} is defined as follows:
-
-\begin{verbatim}
-fun isClosedAndEmpty :: "TransactionOutput ⇒ bool" where
-"isClosedAndEmpty (TransactionOutput txOut) =
-   (txOutContract txOut = Close
-      ∧ accounts (txOutState txOut) = [])" |
-"isClosedAndEmpty _ = False"
-\end{verbatim}
-\noindent
-Because the money is preserved and we know that there is an empty transaction that can be sent at any point after \texttt{maxTime} that will empty the accounts of the contract, we can conclude that money will not be locked in the contract forever.
+Because of this property and \emph{Money preservation}, we can conclude that money will not be locked in the contract forever.
 
@@
 
-In other words, for every list of \texttt{Transaction}s, there exists at least one list of \texttt{Transaction}s with length less or equal to \texttt{maxTransactionsInitialState} that has the same effect on the contracts.
 Thus, we do not lose generality by only exploring traces with length of at most \texttt{maxTransactionsInitialState}, because any longer traces produce the same effect.

### 2019-12-11 papers/marlowe2019/paper.tex
Fix typo
Compact Section 3

@@
 Marlowe is specified by an executable semantics written in Haskell, but to make it usable in practice to execute financial contracts, it needs to be implemented on a blockchain.
-In this section, we explain how Marlowe is executed on the Cardano blockchain using an interpreter written in the Plutus programming language.
+In this section, we explain how Marlowe is executed on the Cardano blockchain using an interpreter \footnote{The implementation is available at \url{https://github.com/input-output-hk/plutus/blob/0ca9af4f6614d591de7ebbe4dd759ce122d74efd/marlowe/src/Language/Marlowe/Semantics.hs}.} written in the Plutus programming language.
 
@@
 \noindent%SJT: **general question** using \emph for inline code: why not \texttt ?
-where \emph{accounts} maps account ids to their balances, \emph{choices} stores user made choice values,
-\emph{boundValues} stores evalutated \emph{Value}'s introduced by \emph{Let} expressions,
-and \emph{minSlot} holds a minimal slot number that a contract has seen, to avoid `time travel to the past'.
-
-It is possible to initialize a contract with a particular state, containing a number of deposits,
-as shown in figure~\ref{fig:marlowe-init}.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=1\textwidth]{figures/Marlowe3-Figures-002.jpeg}
-\caption{Marlowe contract initialization}
-\label{fig:marlowe-init}
-\end{figure}
+where \texttt{accounts} maps account ids to their balances, \texttt{choices} stores user made choice values,
+\texttt{boundValues} stores evalutated \texttt{Value}'s introduced by \texttt{Let} expressions,
+and \texttt{minSlot} holds a minimal slot number that a contract has seen, to avoid `time travel to the past'.
 
@@
 where the remaining contract and state are passed through the \emph{data script},
-and input actions (i.e. \emph{choices}) are passed as \emph{redeemer scripts}.
+and input actions (i.e. \texttt{choices}) are passed as \emph{redeemer scripts}.
 
@@
 we check that the contract locks at least what it should, according
-to contract balances (the \emph{accounts} field in \emph{State}),
+to contract balances (the \texttt{accounts} field in \texttt{State}),
 and that all balances are strictly positive.
@@
 \noindent
-The interpreter reduces a contract until it becomes quiescent: either it evaluates to \emph{Close},
-or it expects a user input in a \emph{When} construct.
-All \emph{Pay, If, Let, Close} constructs are evaluated immediately.
+The interpreter reduces a contract until it becomes quiescent: either it evaluates to \texttt{Close},
+or it expects a user input in a \texttt{When} construct.
+All \texttt{Pay, If, Let, Close} constructs are evaluated immediately.
 
 The evaluation function returns a new contract state, contract continuation, a list of warnings (such as partial payments),
-and a list of expected payments (i.e. one for each of the \emph{Pay} constructs evaluated).
+and a list of expected payments (i.e. one for each of the \texttt{Pay} constructs evaluated).
 
@@
 Here we expect Alice to deposit 850 Ada (850,000,000 Lovelace) into her \texttt{aliceAccount} before slot 100.
-Otherwise, we \emph{Close} the contract.
+Otherwise, we \texttt{Close} the contract.
 
@@
 Let us say that Alice has 1000 Lovelace on her account, and Bob wants to receive the payment.
-The payment could be obtained only after \emph{When} has timed out after slot 200.
-But if we evaluate the contract after slot 200, \emph{If} reduces to \emph{Close}, and
+The payment could be obtained only after \texttt{When} has timed out after slot 200.
+But if we evaluate the contract after slot 200, \texttt{If} reduces to \texttt{Close}, and
 Bob will not receive the payment. So he must evaluate the contract before slot 100 to
-reduce \emph{If} to \emph{When}.
+reduce \texttt{If} to \texttt{When}.
 To achieve that, Bob must spend the contract transaction output with an empty list of inputs in the \emph{Redeemer}.
@@
 Except for the transaction that closes a Marlowe contract, the Marlowe validator script checks that a spending transaction contains a valid continuation
-output, i.e: the hash of the output validator is the same (same hash), and the new state and contract are the expected ones: the ones resulting from applying the \emph{computeTransaction} to the given inputs.
+output, i.e: the hash of the output validator is the same (same hash), and the new state and contract are the expected ones: the ones resulting from applying the \texttt{computeTransaction} to the given inputs.
 
@@
 
-When a contract evaluates to \emph{Close}, all remaining balances the accounts of the contract are payed out
+When a contract evaluates to \texttt{Close}, all remaining balances the accounts of the contract are payed out
 to the respective owners of each account, and the contract is removed from the set of unspent transaction outputs.
@@
 effectively embedding support of Ethereum ERC-20 and ERC-721 contracts. 
-We are working on adding this \emph{multicurrency} support to Marlowe.
+Simple token creation gives interesting possibilities of representing Marlowe contract parties by tokens.
+This tokenization of contract participants abstracts away concrete public keys into contract \emph{roles}.
+In turn, those roles could be traded independently of a contract.
 
-Simple token creation gives interesting possibilities of representing Marlowe contract parties by tokens.
-Currently parties are represented as public keys. We expect a valid signature for each input,
-and make payments to addresses for the public key of a party.
-Analogously, we could instead check proofs of ownership of a particular token.
-
-This would allows us to tokenize contract participants, and abstract away concrete public keys into contract \emph{roles}.
-In turn, those roles could be traded independently of a contract.
-Additionally, we can split a role in multiple \emph{shares}, which would allow automatic distribution of a payment
-proportionally to the number of shares owned.
+We are working on adding \emph{multicurrency} or \emph{roles} support to Marlowe.
 
@@
 
-Fortunately, in the case of Marlowe, a computer can decidably determine whether a particular contract satisfies certain property before executing it, and it can provide a counter-example when it does not.
+Fortunately, in the case of Marlowe, a computer can decidedly determine whether a particular contract satisfies certain property before executing it, and it can provide a counter-example when it does not.
 
@@
 \noindent
-We call a value for \texttt{State} \emph{valid} if the two first properties are true. And we say it has \emph{positive accounts} if the third property is true.
+We call a value for \texttt{State} \emph{valid} if the first two properties are true. And we say it has \emph{positive accounts} if the third property is true.
 

### 2019-12-11 papers/marlowe2019/paper.tex
Use pdf for figures for vector graphics

@@
     \centering
-    \includegraphics[width=4in]{figures/Marlowe3-Figures-003.jpeg}
+    \includegraphics[width=4in]{figures/Marlowe3-Figure-1-crop.pdf}
     \caption{Cardano EUTx0 Model}
@@
 
-\noindent%SJT: **general question** using \emph for inline code: why not \texttt ?
+\noindent
 where \texttt{accounts} maps account ids to their balances, \texttt{choices} stores user made choice values,
@@
     \centering
-    \includegraphics[width=0.7\textwidth]{figures/Marlowe3-Figures-004.jpeg}
+    \includegraphics[width=4in]{figures/Marlowe3-Figure-2-crop.pdf}
     \caption{Zero Coupon Bond Contract}

### 2019-12-11 papers/marlowe2019/paper.tex
squish pictures in playground section a little to save space

@@
 
-For Marlowe to be usable in practice, users need to be able to design and develop Marlowe contracts, and also to understand how contracts will behave once deployed to the blockchain, but without before actually deploying them.
+For Marlowe to be usable in practice, users need to be able to design and develop Marlowe contracts, and also to understand how contracts will behave once deployed to the blockchain, but without actually deploying them.
 
@@
     When [Case (Deposit (AccountId 0 "investor") 
-                        "guarantor"
-                        (Constant 1000))
-               Close]
-         10
-         Close
+               "guarantor" (Constant 1000)) Close] 10 Close
 \end{verbatim}
@@
 
-\begin{figure}[]
-    \includegraphics[width=1\textwidth]{input_composer.png}
-\end{figure}
+\begin{wrapfigure}[6]{l}{0.5\textwidth}
+    \vspace*{-0.1in}
+    \includegraphics[scale=0.15]{input_composer.png}
+\end{wrapfigure}
 
@@
 
-\begin{figure}[]
-    \includegraphics[width=0.8\textwidth]{tx_composer.png}
-\end{figure}
+\begin{wrapfigure}[6]{l}{0.5\textwidth}
+    \vspace*{-0.2in}
+    \includegraphics[scale=0.15]{tx_composer.png}
+\end{wrapfigure}
 
@@
 The playground uses this to search for situations where contract execution would cause warnings.
-For example, suppose you write a contract that causes a payment of 450 Lovelace from Alice to Bob but the contract allows a situation where Alice has only deposited 350 Lovelace then the static analysis will find this and report it to the playground user.
-\begin{figure}
-    \includegraphics[width=1\textwidth]{static_analysis.png}
-\end{figure}
+For example, suppose you write a contract that causes a payment of 450 Lovelace from Alice to Bob but the contract allows a situation where Alice has only deposited 350 Lovelace.
+The the static analysis will find this partial payment case and report it to the playground user with an example of how it could occur.
 

### 2019-12-11 papers/marlowe2019/paper.tex
fix image spacing
Rephraze a few things. Removed empty input example

@@
 and contains a \emph{Redeemer}, such that \texttt{Validator(Data Script, Redeemer, PendingTx)} evaluates to $True$.
-In order to spend a transaction output protected by a \emph{public key}, one must provide a valid signature
-made with the private key corresponding to the transaction public key.
+A valid signature is required to spend a \emph{public key} transaction output.
 
@@
 where \texttt{accounts} maps account ids to their balances, \texttt{choices} stores user made choice values,
-\texttt{boundValues} stores evalutated \texttt{Value}'s introduced by \texttt{Let} expressions,
+\texttt{boundValues} stores evaluated \texttt{Value}'s introduced by \texttt{Let} expressions,
 and \texttt{minSlot} holds a minimal slot number that a contract has seen, to avoid `time travel to the past'.
@@
 \begin{verbatim}
+computeTransaction :: TransactionInput -> State -> Contract -> TransactionOutput
+
 data TransactionInput = TransactionInput
@@
     | Error TransactionError
-
-computeTransaction :: TransactionInput -> State -> Contract -> TransactionOutput
 \end{verbatim}
@@
 \noindent
-The interpreter reduces a contract until it becomes quiescent: either it evaluates to \texttt{Close},
+Given a list of \texttt{Input}'s from \emph{Redeemer}, the interpreter reduces a contract until it becomes quiescent: either it evaluates to \texttt{Close},
 or it expects a user input in a \texttt{When} construct.
@@
 This gives atomicity to some operations, and saves on transaction fees.
-
-Sometimes it is necessary to just evaluate a contract in a given moment in time,
-without providing any inputs. Consider the following example:
-
-\begin{verbatim}
-If (SlotIntervalStart `ValueGT' 100)
-    When [] (Slot 200)
-        (Pay aliceAccount (Party bobPubKey) (Constant 1000))
-    Close
-\end{verbatim}
-
-Let us say that Alice has 1000 Lovelace on her account, and Bob wants to receive the payment.
-The payment could be obtained only after \texttt{When} has timed out after slot 200.
-But if we evaluate the contract after slot 200, \texttt{If} reduces to \texttt{Close}, and
-Bob will not receive the payment. So he must evaluate the contract before slot 100 to
-reduce \texttt{If} to \texttt{When}.
-To achieve that, Bob must spend the contract transaction output with an empty list of inputs in the \emph{Redeemer}.
-He can then do the same after slot 200 to receive the payment.
 
@@
 \begin{wrapfigure}[6]{l}{0.25\textwidth}
-    \vspace*{-0.3in}
+    \vspace*{-0in}
     \includegraphics[scale=0.2]{hole_options_2.png}

### 2019-12-11 papers/marlowe2019/paper.tex
Fixes, comments.

@@
 where, at each stage, the remaining contract and its state are passed through the \emph{Data script},
-and actions and inputs (i.e. \emph{choices} and \emph{money deposits}) are passed to the %SJT: is this irght?
+and actions/inputs (i.e. \emph{choices} and \emph{money deposits}) are passed via the 
+%SJT: is this right?
+%AN: yes. [Input] is precisely the redeemer script. We could say that Value and SlotInterval are parts of PendingTx,
+% but I think it's not that important here.
 \emph{Redeemer}.
@@
 
-\begin{itemize}%SJT: DoS usually means denial of service?
-    \item It reduces the danger of Denegation of Service (DoS) attacks: because the number of valid inputs is limited, an attacker participant cannot arbitrarily block the contract by issuing an unbounded amount of useless \texttt{Transaction}s.
+\begin{itemize}
+    \item It reduces the danger of Denial of Service (DoS) attacks: because the number of valid inputs is limited, an attacker participant cannot arbitrarily block the contract by issuing an unbounded amount of useless \texttt{Transaction}s.
     \item The number of transactions bounds the length of traces that symbolic execution (see Section~\ref{sec:static}) needs to explore (more on this below).%SJT: is this really below? Or above?
@@
     When [Case (Deposit (AccountId 0 "investor") 
-               "guarantor" (Constant 1000)) Close] 10 Close
+               "guarantor" (Constant 1000_000000)) Close] 10 Close
 \end{verbatim}
 In this case, the only action a user can take to progress the contract is to accept a deposit of 1000 ADA
-%SJT: we need to sort out the Ada/Lovelance mismatch.
 from the guarantor to the investor's account. Because of this, the playground can display this action in the input composer.

### 2019-12-11 papers/marlowe2019/paper.tex

@@
 
-\texttt{Value}s include some quantities that change with time, including the current slot number, the current balance of an account (in Lovelace), and any choices that have already been made.
+\texttt{Value}s include some quantities that change with time, like the current slot number, the current balance of an account, and any choices that have already been made.
 % ; we call these volatile values.
 - PLS: this definition is no longer used
-Values can also be combined using addition, subtraction and negation.
+Values can be combined using addition, subtraction and negation.
 
@@
 In the first case, nothing will be transferred; in the later case, a partial payment (of all the money available) is made.
-The contract will continue as specified by \texttt{cont}.
+The contract will continue as \texttt{cont}.
 
@@
 
-Marlowe is specified by an executable semantics written in Haskell, but to make it usable in practice to execute financial contracts, it needs to be implemented on a blockchain.
+Marlowe is specified by an executable semantics written in Haskell, but to make it usable in practice with financial contracts, it needs to be implemented on a blockchain.
 In this section, we explain how Marlowe is executed on the Cardano blockchain using an interpreter \footnote{The implementation is available at \url{https://github.com/input-output-hk/plutus/blob/0ca9af4f6614d591de7ebbe4dd759ce122d74efd/marlowe/src/Language/Marlowe/Semantics.hs}.} written in the Plutus programming language.
@@
 each Marlowe contract into a specific Plutus script.
-Instead, we chose to implement an interpreter of Marlowe contracts as a single Plutus script.
+Instead, we chose to implement a Marlowe interpreter as a single Plutus script.
 This approach has a number of advantages:
@@
 The Marlowe interpreter first validates the current contract state: i.e.\
-we check that the contract locks at least what it should, according
-to contract balances (the \texttt{accounts} field in \texttt{State}),
-and that all balances are strictly positive.
+we check that the contract locks at least as much as specified by the
+contract balances (the \texttt{accounts} field in \texttt{State}),
+and that balances are strictly positive.
 \footnote{
@@
 it is impossible for any possible contract and inputs to result in non-positive balances.
-This is described in more detail in Section \ref{sec:static}.}
+This is described in more detail in Section \ref{subsec:valid_state}.}
 
@@
 
-Our implementation relies on the Haskell library SBV, which in turn relies on existing SMT solvers to check satisfiability of properties or finding counter-examples.
+Our implementation relies on the Haskell library SBV, which in turn relies on existing SMT solvers to check satisfiability of properties.
 
@@
 SBV provides a monad called \texttt{SBV}, a function can use parameters wrapped in this monad to represent symbolic values.
-Functions that take symbolic values can be used as properties and passed to the solver; the solver will then try to replace the symbolic values with concrete values that satisfy or falsify the property.
+Functions that take symbolic values can be used as properties and passed to the solver; which will replace the symbolic values with concrete values that satisfy or falsify the property.
 
@@
 This function repeatedly calls \texttt{reduceContractStep} until it returns \texttt{NotReduced}, so proving overall termination requires a proof that \texttt{reduceContractStep} will eventually do that.
-
-We proved that by establishing a measure for the size of a pair of \texttt{Contract} and \texttt{State}:
+In order to prove this, we defined a measure for the size of a pair of \texttt{Contract} and \texttt{State}:
 
@@
 \noindent
-where \texttt{size} is a measure already generated automatically by Isabelle for custom data types.
+where \texttt{size} is a measure already generated automatically by Isabelle.
 
@@
 
-\subsection{Valid state and positive account preservation}
+\subsection{Valid state and positive account preservation\label{subsec:valid_state}}
 
-There are some values for \texttt{State} that are allowed by its type but make no sense, especially in the case of Isabelle semantics where we use lists instead of maps and sets:
+There are some values for \texttt{State} that are allowed by its type but make no sense, especially in the case of Isabelle semantics where we use lists instead of maps:
 
@@
     \item We want two maps that are equal to be represented the same, so we force keys to be in ascending order.
-    \item For the map of accounts, we only want to record those that have positive amounts.
+    \item We only want to record those accounts that have positive amount.
 \end{enumerate}
@@
 A contract is \emph{quiescent} if and only if the root construct \texttt{When}, or if the contract is \texttt{Close} and all accounts are empty.
-
-We have proved that, if an input \texttt{State} is valid and all accounts are positive, then the result from processing any transaction will always be quiescent.
+We have proved that, if an input \texttt{State} is valid and accounts are positive, then the output will be quiescent.
 
@@
 Money is not created or destroyed by the semantics.
-More specifically, the money that comes in plust the money in the contract before the transaction must be equal to the money that comes out plus the contract after the transaction, except in the case of an error, when the contract stays unaltered.
+More specifically, the money that comes in plus the money in the contract before the transaction must be equal to the money that comes out plus the contract after the transaction, except in the case of an error.
 
@@
 
-We proved that this conservative upper bound is general enough for every contract, by showing that, if the contract is not closed and empty, then an empty transaction sent after \texttt{maxTime} will close the contract and leave the accounts empty.
-
-Because of this property and \emph{Money preservation}, we can conclude that money will not be locked in the contract forever.
+We proved that this conservative upper bound is general enough for every contract, by showing that, if the contract is not closed and empty, then an empty transaction sent after \texttt{maxTime} will close the contract and empty the accounts.
 
@@
 Because transactions that produce an error do not alter the state of the contract, such a list of transactions (a trace) will be equivalent to a list of transactions that does not have the erroneous transaction.
-
-Thus, we do not lose generality by only exploring traces with length of at most \texttt{maxTransactionsInitialState}, because any longer traces produce the same effect.
+Thus, we do not lose generality by only exploring shorter traces.
 
@@
 
-On the simulation pane, ``pure'' Marlowe contacts can be developed: that is, contracts not embedded in a host language. The reason for doing this is two-fold:
+On the simulation pane, ``pure'' Marlowe contacts can be developed: that is, contracts not embedded. Two reasons for doing this are:
 \begin{itemize}
@@
 Blockly gives an easy way to introduce the concepts of Marlowe to users who have no programming knowledge, and in particular the editor gives users a set of options for each construct as the contract is built.
-Once a contract has been constructed in Blockly it is possible to transfer that contract to the simulation pane.%SJT Changed from "Marlowe editor".
+Once a contract has been constructed in Blockly it is possible to transfer that contract to the simulation pane.
 It is also possible to transfer a Marlowe contract to Blockly for further editing.
@@
 \begin{wrapfigure}[9]{l}{0.25\textwidth}
-    \vspace*{-0.3in}
+    \vspace*{-0.28in}
     \includegraphics[width=0.25\textwidth]{hole_options.png}
@@
 
-The Marlowe editor in the unstable version of the playground has an additional feature to aid writing contracts called holes. If we enter the contract \lstinline{?mycontract}
+The Marlowe editor in the unstable version of the playground has a feature called \emph{holes} to aid writing contracts. If we enter the contract \lstinline{?mycontract}
 %SJT: enter it where?
@@
 \begin{wrapfigure}[6]{l}{0.25\textwidth}
-    \vspace*{-0in}
+    \vspace*{-0.00in}
     \includegraphics[scale=0.2]{hole_options_2.png}
@@
 
-\begin{wrapfigure}[6]{l}{0.5\textwidth}
-    \vspace*{-0.1in}
+\begin{wrapfigure}[5]{l}{0.5\textwidth}
+    \vspace*{-0.2in}
     \includegraphics[scale=0.15]{input_composer.png}
@@
 
-At any point in the simulation the user can undo any steps made: in this particular case, they can undo the application of the transaction, and iteratively undo more steps.
-At any point they can also reset the contract to its initial state.
-This enables a user to apply transactions, see what the effect is, step back and apply a transaction with different actions and see how these differences affect the end result.
+At any point in the simulation, the user can undo any steps made: in this particular case, they can undo the application of the transaction, and iteratively undo more steps.
+At any point, they can also reset the contract to its initial state.
+This enables users to apply transactions, see their effects, step back, and try different transactions to see the effects of the changes on the result.
 They can also change the reduced contract to investigate variants of the original.
@@
 Defining the language by means of an executable reference semantics means that we can, as well as directly executing this semantics, generate an on-chain interpreter for it and simulate it in browser using the Haskell-like languages Plutus and PureScript.
-This is made particularly straightforward by working in the subset of Haskell that directly corresponds to what translates directly into these languages.
+This is particularly straightforward when working with a subset of Haskell that is represented in the same way on these languages.
 

### 2019-12-11 papers/marlowe2019/paper.tex
Address comments

@@
 SBV provides a monad called \texttt{SBV}, a function can use parameters wrapped in this monad to represent symbolic values.
-Functions that take symbolic values can be used as properties and passed to the solver; which will replace the symbolic values with concrete values that satisfy or falsify the property.
+Functions that take symbolic values can be used as properties and passed to the solver, which will replace the symbolic values with concrete values that satisfy or falsify the property.
 
@@
     \item We want two maps that are equal to be represented the same, so we force keys to be in ascending order.
-    \item We only want to record those accounts that have positive amount.
+    \item We only want to record those accounts that contain a positive amount.
 \end{enumerate}
@@
 
-On the simulation pane, ``pure'' Marlowe contacts can be developed: that is, contracts not embedded. Two reasons for doing this are:
+On the simulation pane, ``pure'' Marlowe contacts can be developed directly, not embedded in another language. Two reasons for doing this are:
 \begin{itemize}

### 2019-12-12 papers/marlowe2019/paper.tex
15pp

@@
 
-Marlowe\footnote{The complete Marlowe project, including semantics and implementation, is available at \url{https://github.com/input-output-hk/marlowe}.} is a domain-specific language (DSL) for implementing financial contracts on blockchain: our initial target is Cardano, but it could be implemented on many distributed ledgers (DLT platforms), including Ethereum.
+Marlowe\footnote{see \url{https://github.com/input-output-hk/marlowe} for complete project} is a domain-specific language (DSL) for implementing financial contracts on blockchain: our initial target is Cardano, but it could be implemented on many distributed ledgers (DLT platforms), including Ethereum.
 Marlowe is embedded in Haskell, allowing users selectively to use aspects of Haskell -- typically definitions of constants and simple functions -- to express contracts more readably and succinctly.
@@
 
-Contracts are built by putting together a small number of constructs that in combination can be used to describe many different kinds of financial contract.
+Contracts are built by putting together a small number of constructs that in combination can  describe many different financial contracts.
 
-Contracts in Marlowe run on a blockchain, but need to interact with the off-chain world.
+%Contracts in Marlowe run on a blockchain, but need to interact with the off-chain world. 
 The parties to the contract, also called the participants, can engage in various actions: they can be asked to deposit money, or to make a choice between various alternatives.
-In some cases, any party will be able to trigger the contract just to notify it that some condition has become true (for example, a timeout has occurred).
+In some cases, any party will be able to trigger the contract just to notify it that some condition has become true (e.g., a timeout has occurred).
 
@@
 
-Cardano aims to support smart contracs during its Shelley release in 2020.
+Cardano aims to support smart contracts during its Shelley release in 2020.
 Cardano smart contract platform is called \emph{Plutus}, and
@@
     \includegraphics[width=4in]{figures/Marlowe3-Figure-1-crop.pdf}
-    \caption{Cardano EUTx0 Model}
-    \label{fig:eutxo}
+    %\caption{Cardano EUTx0 Model}
+    %\label{fig:eutxo}
 \end{figure}
@@
     marloweState    :: State,
-    marloweContract :: Contract
-}
+    marloweContract :: Contract }
 \end{verbatim}
@@
 
-We then compute expected transaction outcomes for given inputs, such as:
-contract continuation, new state, and payments.
+We then apply  \texttt{computeTransaction}  to the contract inputs, the contract continuation, and new state to compute the expected transaction outcomes:
+\begin{verbatim}
+computeTransaction :: 
+  TransactionInput -> State -> Contract -> TransactionOutput
+\end{verbatim}
+%\begin{verbatim}
+%expectedTxOutputs = computeTransaction txInput marloweState marloweContract
+%\end{verbatim}
+%
+%\noindent
+%where
+%
+%\begin{verbatim}
+%computeTransaction :: TransactionInput -> State -> Contract -> TransactionOutput
+%
+%data TransactionInput = TransactionInput
+%    { txInterval :: SlotInterval
+%    , txInputs   :: [Input] }
+%
+%data TransactionOutput =
+%    TransactionOutput
+%        { txOutWarnings :: [TransactionWarning]
+%        , txOutPayments :: [Payment]
+%        , txOutState    :: State
+%        , txOutContract :: Contract }
+%    | Error TransactionError
+%\end{verbatim}
+\noindent
+where a \texttt{TransactionInput} consists of the current slot interval, together with other ontract inputs, and the outputs combine any payments and warnings with the resulting output state and contract.
 
-\begin{verbatim}
-expectedTxOutputs = computeTransaction txInput marloweState marloweContract
-\end{verbatim}
-
-\noindent
-where
-
-\begin{verbatim}
-computeTransaction :: TransactionInput -> State -> Contract -> TransactionOutput
-
-data TransactionInput = TransactionInput
-    { txInterval :: SlotInterval
-    , txInputs   :: [Input] }
-
-data TransactionOutput =
-    TransactionOutput
-        { txOutWarnings :: [TransactionWarning]
-        , txOutPayments :: [Payment]
-        , txOutState    :: State
-        , txOutContract :: Contract }
-    | Error TransactionError
-\end{verbatim}
-
-\noindent
 Given a list of \texttt{Input}'s from \emph{Redeemer}, the interpreter reduces a contract until it becomes quiescent: either it evaluates to \texttt{Close},
@@
     \includegraphics[width=4in]{figures/Marlowe3-Figure-2-crop.pdf}
-    \caption{Zero Coupon Bond Contract}
-    \label{fig:zero-coupon-bond}
+    %\caption{Zero Coupon Bond Contract}
+    %\label{fig:zero-coupon-bond}
 \end{figure}
@@
 
-\subsection{Further Work}
+\paragraph{Future Work}
 
-Cardano extends its ledger rules to support \emph{forging} of custom currencies and tokens,
-effectively embedding support of Ethereum ERC-20 and ERC-721 contracts. 
+Cardano extends its ledger rules to support \emph{forging} of custom currencies and tokens.%effectively embedding support of Ethereum ERC-20 and ERC-721 contracts. 
 Simple token creation gives interesting possibilities of representing Marlowe contract parties by tokens.
@@
 In turn, those roles could be traded independently of a contract. 
-
 We are working on adding \emph{multicurrency} or \emph{roles} support to Marlowe.
@@
 
-Another property of Marlowe is that any given finite \texttt{Contract} has an implicit finite bound on the maximum number of \texttt{Transaction}s that it accepts.
-This is a convenient property because:
+Another property of Marlowe is that any given \texttt{Contract} has an implicit finite bound on the maximum number of \texttt{Transaction}s that it accepts. 
+This is a convenient property for two reasons.
 
-\begin{itemize}
-    \item It reduces the danger of Denial of Service (DoS) attacks: because the number of valid inputs is limited, an attacker participant cannot arbitrarily block the contract by issuing an unbounded amount of useless \texttt{Transaction}s.
-    \item The number of transactions bounds the length of traces that symbolic execution (see Section~\ref{sec:static}) needs to explore (more on this below).%SJT: is this really below? Or above?
-\end{itemize}
-
+First, it reduces the danger of Denial of Service (DoS) attacks, because the number of valid inputs is limited, an attacker participant cannot arbitrarily block the contract by issuing an unbounded amount of useless \texttt{Transaction}s.
+Secondly, the number of transactions bounds the length of traces that symbolic execution (see Section~\ref{sec:static}) needs to explore.% (more on this below).%SJT: is this really below?
+Or above?
 We state the property as follows:
@@
 \begin{wrapfigure}[9]{l}{0.25\textwidth}
-    \vspace*{-0.28in}
+    %\vspace*{-0.28in}
     \includegraphics[width=0.25\textwidth]{hole_options.png}
@@
 \begin{wrapfigure}[6]{l}{0.25\textwidth}
-    \vspace*{-0.00in}
+    \vspace*{-0.25in}
     \includegraphics[scale=0.2]{hole_options_2.png}
@@
 Once the action is added other inputs become possible; these are displayed in the input composer, and again they can be added to the transaction being composed.
-In this way, multiple actions can be added to a transaction before it is applied.
+In this way, multiple %actions can be added to a transaction before it is applied.
 
-\begin{wrapfigure}[6]{l}{0.5\textwidth}
+\begin{wrapfigure}[5]{l}{0.5\textwidth}
     \vspace*{-0.2in}
@@
 \noindent
+actions can be added to a transaction before it is applied.
 A user can then apply this transaction and in the example above this would result in the state pane showing a single payment and in addition the contract in the Marlowe editor will have been reduced to \lstinline{Close}.
@@
 
-An earlier paper~\cite{cryptoeprint:2016:1156} reviews smart contracts on blockchain. Here we look at a number of recent systems that bear direct comparison with Marlowe.
-
-The Ethereum system~\cite{wood2014ethereum} provides an exemplar Turing-complete language for  the EVM virtual machine, and a higher-level programming language, Solidity, that compiles into EVM code.
-Both these languages are general-purpose, and so have more akin with Plutus~\cite{PlutusPlatform} rather than Marlowe.
-
-On the other hand,  Nxt~\cite{Nxt}, is special-purpose in providing a ``fat'' high-level
+An earlier paper~\cite{cryptoeprint:2016:1156} reviews smart contracts on blockchain.
+Here we look at a number of recent systems that bear direct comparison with Marlowe, rather than general purpose languages.
+Nxt~\cite{Nxt}, is special-purpose in providing a ``fat'' high-level
 API, containing built-in transaction types and transactions that support some 250 primitive operations; these can be ``scripted'' in a client (only) using a binding to the API, which is available, for instance, in JavaScript.
@@
 account by our notion of commitments.
-They take no account of commitments or timeouts as our approach does, and so are unable to guarantee some of the properties -- such as a finite lifetime -- built into Marlowe by design.
+They take no account of commitments or timeouts as our approach does, and so are unable to guarantee some properties -- such as a finite lifetime -- built into Marlowe by design.
 
-BitML~\cite{BitML} is a DSL for specifying Marlowe-like contracts that regulate transfers on the Bitcoin blockchain.
-BitML is implemented via a compiler that translates contracts into standard Bitcoin transactions.
-Participants execute a contract by appending these transactions on the Bitcoin blockchain, according to their strategies, which also involve the exchange of bitstrings that guarantee to a very high probability the correctness of contract execution.
+BitML~\cite{BitML} is a DSL for specifying Marlowe-like contracts that regulate transfers on the Bitcoin blockchain, and is implemented via a compiler that translates contracts into  Bitcoin transactions plus strategies.
+Participants execute a contract by appending these transactions on the Bitcoin blockchain, according to their strategies, which involve the exchange of bitstrings that guarantee to a very high probability the correctness of contract execution.
 Marlowe is directly implemented by an interpreter which could also be implemented on a covenant-based~\cite{covenants} extension of the Bitcoin blockchain.

### 2020-01-27 papers/marlowe2019/paper.tex
Responded to referees' comments. The paper is now half a page too long, but will leave shortening until we see final publication instructions.

@@
 
-Marlowe\footnote{see \url{https://github.com/input-output-hk/marlowe} for complete project} is a domain-specific language (DSL) for implementing financial contracts on blockchain: our initial target is Cardano, but it could be implemented on many distributed ledgers (DLT platforms), including Ethereum.
+Marlowe~\cite{marlowe-github} is a domain-specific language (DSL) for implementing financial contracts on blockchain: our initial target is Cardano, but it could be implemented on many distributed ledgers (DLT platforms), including Ethereum.
 Marlowe is embedded in Haskell, allowing users selectively to use aspects of Haskell -- typically definitions of constants and simple functions -- to express contracts more readably and succinctly.
@@
 Crucially in a financial environment, we are able to \emph{exhaustively analyse contracts} without executing them, so that we can, for instance, check whether any particular contract is able to make all the payments it should: in the case it is not, we get an explicit example of how it can fail.
-This analysis, reported in Section \ref{sec:static}, is built into the Marlowe Playground.
-Finally, we are able to use formal verification to prove properties of the implementation of Marlowe, including a guarantee that ``money in $=$ money out'' for all contracts; see Section~\ref{sec:verification}.
+This analysis (Section \ref{sec:static}) is built into the Marlowe Playground.
+Finally (Section \ref{sec:verification}) we are able to use formal verification to prove properties of the implementation of Marlowe, including a guarantee that ``money in $=$ money out'' for all contracts.
 
-\section{Marlowe overview}
+Building on~\cite{isola-marlowe}, the specific contributions of this paper are threefold.
+We describe revised versions of Marlowe and its Playground, we describe the implementation of Marlowe on the Cardano blockchain, and we present both automated static analysis and machine-supported verification for Marlowe.
+
+\section{Marlowe revised: version 3.0}
 \label{sec:overview}
 
-Since the first publication, we have revised the language design: this section gives a brief overview of the current (3.0) version of the language and its semantics.
+Since first publication, we have revised the language design: this section gives a brief overview of the current (3.0) version of the language and its semantics.
 
@@
 If he does, the contract is closed, and all remaining balances must be paid out
-to their respective owners. In our case, 1000 Ada must be paid to Alice.
+to their respective owners; in this case, 1000 Ada must be paid to Alice. If Bob does not pay, then Alice has lost her money, because this is an unsecured loan.
 
@@
 We can also use proof assistants to demonstrate that the Marlowe semantics presents certain desirable properties, such as that money is preserved and anything unspent is returned to users by the end of the execution of any contract.
+Formal statements and proofs of these properties are given in~\cite{marlowe-github}.
 
 Currently, we have translated the Haskell Marlowe semantics to Isabelle while keeping both versions as close as possible, but we decided to make them different in two main aspects:
-
 \begin{itemize}
@@
 For example, suppose you write a contract that causes a payment of 450 Lovelace from Alice to Bob but the contract allows a situation where Alice has only deposited 350 Lovelace.
-The the static analysis will find this partial payment case and report it to the playground user with an example of how it could occur.
+The static analysis will find this partial payment case and report it to the playground user with an example of how it could occur.
 
@@
 
-An earlier paper~\cite{cryptoeprint:2016:1156} reviews smart contracts on blockchain.
-Here we look at a number of recent systems that bear direct comparison with Marlowe, rather than general purpose languages.
-Nxt~\cite{Nxt}, is special-purpose in providing a ``fat'' high-level
-API, containing built-in transaction types and transactions that support some 250 primitive operations; these can be ``scripted'' in a client (only) using a binding to the API, which is available, for instance, in JavaScript.
-In providing such specificity this bears comparison with our implementation of contracts from the ACTUS standard~\cite{actus}.
+Our work is inspired by the work of Peyton Jones et al~\cite{PeytonJones:2000} to describe financial contracts using a DSL embedded in Haskell. 
+In the remainder of this section we look at a number of recent systems that bear direct comparison with Marlowe; an earlier paper~\cite{cryptoeprint:2016:1156} gives a detailed review of smart contracts on blockchain.
 
-Our work is inspired by the original work of Peyton Jones and others~\cite{PeytonJones:2000} to describe financial contracts using a DSL embedded in Haskell.
 The Findel project~\cite{findel} examines financial contracts on the Ethereum platform, and is also based on~\cite{PeytonJones:2000}.
@@
 They take no account of commitments or timeouts as our approach does, and so are unable to guarantee some properties -- such as a finite lifetime -- built into Marlowe by design.
+Nxt~\cite{Nxt}, is special-purpose in providing a ``fat'' high-level
+API, containing built-in transaction types and transactions that support some 250 primitives; these can be ``scripted'' in a client using a binding to the API, which is available, e.g., in JavaScript.
+In providing such specificity this bears comparison with our implementation of contracts from the ACTUS standard~\cite{actus}.
 

### 2020-02-25 papers/marlowe2019/paper.tex
line break
Add note about ongoing work on static analysis

@@
 \noindent
-Here we expect Alice to deposit 850 Ada (850,000,000 Lovelace) into her \texttt{aliceAccount} before slot 100.
+Here we expect Alice to deposit 850 Ada (850,000,000 Lovelace) into her account \texttt{aliceAccount} before slot 100.
 Otherwise, we \texttt{Close} the contract.
@@
 
+\subsection{Future work}
+
+At the time of writing, we are actively trying new optimisations and approaches for implementing static analysis that are more efficient in practise than the approach described in this paper.
+As a consequence, in some cases, the static analysis implementation can deviate considerably from the Haskell semantics.
+Our intention is to keep using property-based testing and formal verification to ensure that both implementations are consistent.
+
 \section{The Marlowe Playground}

### 2020-03-10 papers/marlowe2019/paper.tex
Adjust images

@@
 \institute{IOHK, Hong Kong, 
-\email{alexander.nemish@iohk.io, pablo.lamela@iohk.io, simon.thompson@iohk.io,~david.smith@tweag.io},  \and
+\email{alexander.nemish@iohk.io, pablo.lamela@iohk.io, simon.thompson@iohk.io,~david.smith@tweag.io}  \and
 School of Computing, University of Kent, UK, 
@@
 \begin{wrapfigure}[9]{l}{0.25\textwidth}
-    %\vspace*{-0.28in}
+    \vspace*{-0.24in}
     \includegraphics[width=0.25\textwidth]{hole_options.png}
@@
 
-\begin{wrapfigure}[6]{l}{0.25\textwidth}
-    \vspace*{-0.25in}
-    \includegraphics[scale=0.2]{hole_options_2.png}
+\begin{wrapfigure}[5]{l}{0.20\textwidth}
+%    \vspace*{-0.25in}
+    \includegraphics[scale=0.18]{hole_options_2.png}
 \end{wrapfigure}
@@
 \end{verbatim}
-In this case, the only action a user can take to progress the contract is to accept a deposit of 1000 ADA
-from the guarantor to the investor's account. Because of this, the playground can display this action in the input composer.
+In this case, the only action a user can take to progress the contract is to accept a deposit of 1000 ADA from the guarantor to the investor's account.
+Because of this, the playground can display this action in the input composer.
 
-\begin{wrapfigure}[5]{l}{0.5\textwidth}
-    \vspace*{-0.2in}
+\begin{wrapfigure}[4]{l}{0.47\textwidth}
+    \vspace*{-0.28in}
     \includegraphics[scale=0.15]{input_composer.png}
@@
 
-\noindent
 The user can then choose to add this action to a transaction being prepared.
@@
 In this way, multiple %actions can be added to a transaction before it is applied.
+actions can be added to a transaction before it is applied.
 
-\begin{wrapfigure}[5]{l}{0.5\textwidth}
-    \vspace*{-0.2in}
+\begin{wrapfigure}[5]{l}{0.47\textwidth}
+    \vspace*{-0.32in}
     \includegraphics[scale=0.15]{tx_composer.png}
@@
 
-\noindent
-actions can be added to a transaction before it is applied.
 A user can then apply this transaction and in the example above this would result in the state pane showing a single payment and in addition the contract in the Marlowe editor will have been reduced to \lstinline{Close}.

### 2020-04-09 papers/marlowe2019/paper.tex
Update terminology

@@
 Cardano is a UTxO-based (unspent transaction output) blockchain, similar to Bitcoin~\cite{sok}.
-It extends the Bitcoin model by allowing transaction outputs to hold a \emph{Data Script}.
+It extends the Bitcoin model by allowing transaction outputs to hold a \emph{Dataum}.
 As the name suggests, this is a serialised data value used to store and communicate a contract state.
@@
 In order to spend an existing transaction output protected by a \emph{Validator},
-one must create a transaction (a \emph{PendingTx}) that has an \emph{input} that references the transaction output,
-and contains a \emph{Redeemer}, such that \texttt{Validator(Data Script, Redeemer, PendingTx)} evaluates to $True$.
+one must create a transaction (a \emph{Context}) that has an \emph{input} that references the transaction output,
+and contains a \emph{Redeemer}, such that \texttt{Validator(Datum, Redeemer, Context)} evaluates to $True$.
 A valid signature is required to spend a \emph{public key} transaction output.
@@
 Marlowe contract execution on the blockchain consists of a chain of transactions
-where, at each stage, the remaining contract and its state are passed through the \emph{Data script},
+where, at each stage, the remaining contract and its state are passed through the \emph{Datum},
 and actions/inputs (i.e. \emph{choices} and \emph{money deposits}) are passed via the 
 %SJT: is this right?
-%AN: yes. [Input] is precisely the redeemer script. We could say that Value and SlotInterval are parts of PendingTx,
+%AN: yes. [Input] is precisely the redeemer script. We could say that Value and SlotInterval are parts of Context,
 % but I think it's not that important here.
@@
 
-We store the remaining contract in the \emph{Data script}, which makes it visible to everyone.
+We store the remaining contract in the \emph{Datum}, which makes it visible to everyone.
 This simplifies contract reflection and retrospection.
@@
 with at least one script output,
-with the particular Marlowe contract in the data script, and protected by the Marlowe validator script.
+with the particular Marlowe contract in the \emph{Datum}, and protected by the Marlowe validator script.
 Note that we do not place any restriction on the transaction inputs, which could use any other transaction
@@
 Marlowe contract execution consists of a chain of transactions,
-where the remaining contract and state are passed through the \emph{data script},
+where the remaining contract and state are passed through the \emph{Datum},
 and input actions (i.e. \texttt{choices}) are passed as \emph{redeemer scripts}.

### 2020-04-29 papers/static_analysis2020/paper.tex
Add first draft for static analysis paper

\documentclass[english,runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage{graphicx}
\usepackage{babel}
\begin{document}
\title{Efficient static analysis of Marlowe contracts}

\author{Pablo {Lamela Seijas}\orcidID{0000-0002-1730-1219}\\
\email{pablo.lamela@iohk.io}
\institute{IOHK, Hong Kong}}

\authorrunning{P. Lamela Seijas}

\maketitle

\section{Introduction\label{sec:Introduction}}

Thanks to static analysis, we can automatically check beforehand whether
payments promised by a Marlowe contract can be fulfilled in every
possible execution of the contract. For example, if a Marlowe contract
has passed the static analysis, we will have a very high assurance
that whenever the contract says we will be paid a certain amount of
money, the contract will indeed have enough money to make such payment.

State of the art libraries like SBV \cite{erkok2019sbv} allow straightforward
static analysis of arbitrary Haskell functions, with very few restrictions
on how those functions are implemented.

However, static analysis can be very computationally expensive, because
it essentially needs to check every possible execution of the contract.

For example, in the case of Marlowe, contracts are typically interactive
and must deal with inputs arriving from different users in different
orders, and sometimes not arriving at all. In addition, Marlowe allows
several inputs to be grouped in a single transaction, and different
decisions may be taken depending on the content of those inputs, or
in the combinations of contents of those inputs. All these alternatives
multiply the search space of static analysis and can easily compose
exponentially and thus make the whole approach unusable in practice
even for moderately sized contracts.

This paper contributes a number of approaches that can be used when
optimising static analysis, examples extracted from a case-study where
we applied these approaches, and an overview of the techniques used
to test the correctness of the optimisations.

The case study presented is that of the optimisation of Marlowe\textquoteright s
static analysis. During this work, the techniques applied allowed
us to reduce the analysis time of an implementation that followed
the semantics closely and took a couple of minutes to analyse contracts
of a few kilobytes, to one where the same contract would take less
than a second, and where a four-person crowdfunding contract that
when fully expanded occupies about 19 megabytes, can be analysed in
around 10 minutes using Z3. In the original implementation, we were
not able to analyse the 19 megabyte contract because it ran out of
memory after half an hour.

There are many things that can be done to improve the efficiency of
static analysis. We have classified them as lightweight and heavyweight.

\paragraph*{Lightweight }

modifications are local and can be done without fundamentally changing
the implementation. We consider three main ideas:
\begin{itemize}
\item Removing unnecessary parts from the analysis. If they do not affect
the property that is being verified then we can just remove them from
the implementation and reduce its complexity.
\item Avoiding high level abstractions. High level abstractions aid reasoning
and avoid errors, but also introduce complexity that is not necessarily
needed.
\item Reducing search space by normalising parameters. If there are several
ways of representing some inputs, and the different representations
have no impact on the analysis, then we can remove all but one when
implementing static analysis.
\end{itemize}

\paragraph*{Heavyweight}

modifications are more fundamental approaches that require considerable
changes to the fundamental structure of the implementation. We consider
two main ideas:
\begin{itemize}
\item Reducing search space by using normalised execution paths relevant
to the property. Instead of using search space to model inputs, we
use it to model possible executions, and we also ensure that we only
represent each possible group of equivalent executions once, i.e:
executions with the same normal form.
\item Minimizing the representation of inputs and outputs. We can benefit
from encoding inputs and outputs as concisely as possible, and discarding
information that we can infer in other ways.
\end{itemize}
In the following sections, we introduce the semantics of Marlowe as
a case study (Section~\ref{sec:Marlowe-intro}), we go through a
general approach to static analysis (Section~\ref{sec:Making-Marlowe-symbolic}),
then we explore both lightweight (Section~\ref{subsec:Lightweight-modifications})
and heavyweight (Section~\ref{subsec:Heavyweight-modifications})
optimisation techniques in more detail, and we illustrate them with
examples of how they apply to Marlowe static analysis. In Section~\ref{sec:Testing},
we briefly discuss some ways in which property based testing can be
used to address potential problems that may arise from the implementation
of the most heavyweight optimisations described in the previous section.

\section{Marlowe and how its design helps static analysis and correctness\label{sec:Marlowe-intro} }

Firstly, we will briefly introduce the semantics of Marlowe, the guarantees
that are offered implicitly by the semantics, and how the design choices
facilitate static analysis and make it decidable. A more detailed
explanation of Marlowe can be found in \cite{marlowe}.

\subsection{Structure of Marlowe contracts\label{subsec:Structure-of-Marlowe} }

Marlowe contracts are able to receive payments, store money and tokens,
ask for input from the participants, and redistribute stored money
and tokens back among participants. The decision about which and when
these actions are carried out is taken by the contract on the basis
of the information that is available at each point in time. Participants,
in turn, do not need to be a single person, but they may be represented
by a \texttt{Role}, which in turn would be controlled by another contract.

The Marlowe language is written as a set of mutually recursive Haskell
data types. It regulates interactions between a finite number of participants,
established before the start of the contract.

The main data type is called \texttt{Contract}, and it represents
the logic and actions that the contract allows or enforces. The outmost
constructs of the contract represent the actions that will be enforced
first, and as those constructs become enforced, the contract will
evolve into one of its continuations (sub-contracts), and the process
will continue until only the construct \texttt{Close} remains.

There are 5 constructs of type \texttt{Contract}:
\begin{itemize}
\item \texttt{Close} -- signals the end of the life of a contract. Once
it is reached, all the money and tokens stored in the accounts of
the contract will be refunded to the owner of each of the respective
accounts.
\item \texttt{If} -- immediately decides how the contract must continue,
depending on a given condition that we call an \texttt{Observation}.
The value of an \texttt{Observation} can depend on previous choices
made by participants, on amounts of money and tokens remaining, or
other factors.
\item \texttt{Let} -- immediately stores a \texttt{Value} for use later.
The value of a \texttt{Value} can depend on the information available
to the contract at the time. For example, it may store the minimum
slot of the slot interval of the current transaction, or the money
available in a particular account at a given point in time, for use
at a later time in the contract execution.
\item \texttt{When} -- waits for an external input. The \texttt{When} construct
also specifies a timeout slot: after this slot has been reached, the
\texttt{When} construct expires, and no longer accepts any input.
There are three types of input:
\begin{itemize}
\item \texttt{Deposit} -- Waits for a participant to deposit an amount
of money or tokens (specified as a \texttt{Value}) in the contract.
The amount is added to one of the accounts of the contract.
\item \texttt{Choice} -- Waits for a participant to make a choice. A choice
is represented as an integer number from a set specified by the contract.
\item \texttt{Notify} -- Waits for an \texttt{Observation} to be true.
Because contracts are reactive (they cannot initiate transactions),
it is necessary for an external actor to \texttt{Notify} the contract
that the \texttt{Observation} has become true. However, this notification
does not have to be carried out by a participant, anybody can do it.
\end{itemize}
\item \texttt{Pay} -- immediately makes a payment between accounts of the
contract, or from an account of the contract to a given participant.
The amount transferred is specified as a \texttt{Value}. 
\end{itemize}

\subsection{Semantics\label{subsec:Marlowe-Semantics}}

As we mentioned in Section~\ref{subsec:Structure-of-Marlowe}, Marlowe
contracts are passive: they are executed by transactions being submitted
to the blockchain, with the on chain interpreter checking that the
transaction complies with the contract under execution. These transactions
need to be submitted by participants or their representatives (e.g.
user wallets) and their execution is carried out atomically and deterministically.

Each transaction may include a list of inputs, a set of signatures,
minimum and maximum slot numbers, a set of input UTxOs (incoming money
and tokens), and a set of outputs (outgoing money and tokens). For
a transaction to be valid, the transaction should have the same effect
at any slot within that slot range (be deterministic). For example,
if a transaction has a minimum slot number which is before a timeout,
and a maximum slot which after a timeout then the transaction will
fail with the error \texttt{AmbiguousSlotInterval}, since it would
not be deterministic.

The key aspect of the Marlowe semantics is that it checks that a particular
transaction is valid given the current state and contract. Because
transactions are deterministic, there should be no reason why someone
accidentally sends a transaction that is invalid, since it will only
result in a cost to that participant. The type signature of the transaction
validation function is:

\begin{verbatim}
computeTransaction :: TransactionInput -> State -> Contract
                   -> TransactionOutput 
\end{verbatim}

This function can be factored into four main functions:
\begin{itemize}
\item \texttt{reduceContractStep} - this function simplifies the topmost
construct that does not require an input to be executed (i.e: anything
but a \texttt{When} that has not expired or a \texttt{Close} when
accounts are empty). It only simplifies the \texttt{When} construct
if it has expired (i.e. the timeout specified in the \texttt{When}
is less than or equal to the minimum slot number). In the case of
the \texttt{Close} contract, it only refunds one of the accounts at
a time. 
\item \texttt{reduceContractUntilQuiescent} - this function calls \texttt{reduceContractStep}
repeatedly until it has no further effect (i.e. it reaches a fixed
point). 
\item \texttt{applyInput} - this function processes one input from the head
of the input list. The topmost construct must be a \texttt{When} that
is expecting that particular input and has not expired. 
\item \texttt{applyAllInputs} - this function processes a list of inputs.
It calls \texttt{applyInput} for each of the inputs in the list, and
calls \texttt{reduceContractUntilQuiescent} before and after every
call to \texttt{applyInput}.
\end{itemize}
The \texttt{State} stores information about the amount of money and
tokens in the contract at a given time, together with the choices
made, \texttt{Let} bindings made, and a lower bound for the current
slot:

\begin{verbatim}
data State = State { accounts    :: Map AccountId Money
                   , choices     :: Map ChoiceId ChosenNum
                   , boundValues :: Map ValueId Integer
                   , minSlot     :: Slot }
\end{verbatim}

\subsection{Extra considerations\label{subsec:Extra-considerations}}

Many of the design decisions behind Marlowe have been made with the
aim of preventing potential errors. For example:
\begin{itemize}
\item Account identifiers include an account owner; an account owner is
a participant that will get the money or tokens remaining in an account
when a contract terminates. At the same time, the only construct that
can pause the execution of a contract is the \texttt{When} construct,
which has a timeout, this ensures that all contracts eventually expire
and terminate. Together these properties ensure that no money or tokens
are locked in the contract forever.
\item Classification of money and tokens into accounts separates concerns.
Marlowe will never spend more money or tokens than there are in an
account, even if there is more available in the contract. But when
required to pay more than is in an account, it will pay as much as
is available in the account, in order to remain as close as possible
to the original intention of the contract.
\item Marlowe does not perform deposits or payments of negative amounts
of money or tokens. It treats them as zero. However, at the same time,
if there happens to be a request for deposit with a negative amount,
it continues gracefully, it will still wait for a null deposit to
be made, and it will continue as if everything is correct, this way
the expected execution of the contract is disturbed as little as possible.
\item Marlowe contracts have an upper limit in the number of inputs that
they accept throughout their life. This limit is implied by the maximum
number of nested When constructs in the contract, since only one input
per When can be accepted. At the same time, transactions that have
no effect on the contract are considered invalid, thus there is also
a limit in the maximum number of transactions a contract can accept
throughout its life. This prevents DoS attacks, and makes static analysis
easier. We discuss this in more detail in Section~\ref{sec:Making-Marlowe-symbolic}.
\end{itemize}

\section{Making Marlowe semantics symbolic\label{sec:Making-Marlowe-symbolic}}

In this section, we briefly present and reflect on a technique that
can be used to convert a concrete implementation of a Haskell function
into a symbolic one by using the SBV library, and to use this symbolic
implementation for static analysis. In particular, we explore this
technique in the context of the Marlowe semantics.

This approach corresponds to our first attempt at implementing static
analysis for Marlowe contracts, and it is a systematic approach that
can be carried out with very few assumptions.

\subsection{Overview}

The SBV library supports implementing Haskell functions in a way that
the same implementation can be used:
\begin{itemize}
\item With concrete parameters, as a normal Haskell function
\item With symbolic variables, so that properties can be check for satisfiability
using an SMT solver
\item As part of QuickCheck properties, for testing randomly.
\end{itemize}
Parameters that can be used symbolically are wrapped in a monad called
\texttt{SBV}. Functions that depend on symbolic parameters also return
values wrapped in the \texttt{SBV} monad.

The idea is therefore to convert our implementation to use values
wrapped in the \texttt{SBV} monad, by replacing all operations on
those values with symbolic functions. In order to simplify this work,
\texttt{SBV} provides symbolic versions for many of the most common
built-in functions that are either overloaded or have similar names.

Our semantics transaction processing function would thus become:

\begin{verbatim}
computeTransaction :: SBV TransactionInput -> SBV State
                   -> SBV Contract
                   -> SBV TransactionOutput
\end{verbatim}

We can then define a function that takes a list of transactions instead,
and we would just need to write a property that says that the result
of executing the list of transactions for a given contract does not
have any warnings (the property we are trying to verify). Then we
can pass the property to SBV and it will tell us which transaction
lists break the property.

Unfortunately, there are a couple of issues with this approach that
we review in Section~\ref{subsec:Considerations}.

\subsection{Considerations\label{subsec:Considerations}}

SBV does not currently support complex custom data types. It has some
support for nonary data types, but not for data types with parameters.
It does support the \texttt{Either} type, and the \texttt{Tuple} type.
So it is possible to have conversions between custom data types and
nested \texttt{Either} and \texttt{Tuple} type synonyms. We originally
implemented this conversion using Template Haskell, since by using
this technique the symbolic implementation can remain similar to the
concrete one.

For example, the following data structure:

\begin{verbatim}
data Input = IDeposit AccountId Party Money
           | IChoice ChoiceId ChosenNum
           | INotify
\end{verbatim}

Would be translated to the following type synonym:

\begin{verbatim}
type SInput = SBV (Either (AccountId, Party, Money)
                          (Either (ChoiceId, ChosenNum)
                                  ()))
\end{verbatim}

But this approach cannot address recursive datatypes, let alone mutually
recursive datatypes. And the \texttt{Contract} type is mutually recursive.

Even if we could represent mutually recursive datatypes, it would
not be feasible to directly translate the semantics, because a more
general limitation of SMT solvers is that they cannot, in general,
deal with arbitrarily sized computations. In other words, if termination
of a symbolic function is bounded by a parameter that can be arbitrarily
large, SMT solvers will often fail to terminate when trying to validate
a property, since doing so often requires a proof by induction. We
discuss how to address this problem in Section~\ref{subsec:Adapting-the-semantics}.

\subsection{Adapting the semantics\label{subsec:Adapting-the-semantics}}

In order to guarantee termination of the SMT validation, we need a
concrete bound for the property that we want to validate.

Related work often addresses this problem by manually establishing
a bound on the amount of computation, e.g: limiting the number of
computation steps analysed, the number of times loops are unrolled
\cite{clarke2004tool,gulwani2008program,jackson2000finding}.

In the case of Marlowe, we have natural bounds that we can use because,
given a concrete contract, we can calculate limits for:
\begin{itemize}
\item The maximum number of inputs that can have an effect on the contract
\item The maximum number of transactions that can have an effect on the
contract
\item All of the account, choice, and \texttt{Let} identifiers that will
be used in the contract
\item The number of participants that will participate in the contract
\end{itemize}
From this data we can deduce an upper bound for:
\begin{itemize}
\item The number of times that
\texttt{computeTransaction}, \texttt{reduceContractStep}, and \texttt{applyInput} are
called.
\item The number of accounts that the contract will use, and an
the number of elements in each of the associative maps
that comprise the \texttt{State} of the contract.
\end{itemize}
If we keep the \texttt{Contract} parameter of the semantics concrete,
we can use it to bound the execution, because every call to \texttt{reduceContractStep}
will either make no progress or remove one of the constructs, with
the exception of \texttt{Close}. But for \texttt{Close} we know we
just have to call it as many times as the maximum number of accounts
there can possibly be, since every time we call it it will refund
one of the accounts.

Thus, the symbolic transaction processing function becomes:

\begin{verbatim}
computeTransaction :: SBV TransactionInput -> SBV State
                   -> Contract
                   -> SBV TransactionOutput
\end{verbatim}

However, there is one more problem. The output \texttt{Contract} returned
by the function, which is wrapped inside the \texttt{TransactionOutput},
is symbolic, since it depends on the current \texttt{TransactionInput}
and \texttt{State}, which are both symbolic.

We get around this problem by using a continuation style. Instead
of returning the \texttt{TransactionOutput} directly, we take a continuation
function that takes the concrete contract and the symbolic version
of \texttt{TransactionOutput} without the contract.

Thus, the symbolic transaction processing function will look something
like:

\begin{verbatim}
computeTransaction :: SymVal a => SBV TransactionInput
                   -> SBV State -> Contract
                   -> (SBV TransactionOutput ->
                       Contract -> SBV a)
                   -> SBV a
\end{verbatim}

In practice, we also include some extra information about bounds,
and we make some other parts of \texttt{TransactionOutput} concrete,
in addition to the \texttt{Contract}.

\section{Making static analysis efficient\label{sec:Making-static-analysis-efficient}}

In this section, we explain the techniques that we mentioned in Section~\ref{sec:Introduction}
more in detail, and we illustrate those techniques with examples from
the static analysis implementation of Marlowe.

\subsection{Lightweight modifications\label{subsec:Lightweight-modifications}}

Some of the modifications that we can apply to a static analysis implementation
are localised, and they do not fundamentally affect the overall structure
of the implementation. This means that the implementation can remain
intuitive and close to the original semantics and model that we are
trying to analyze, as well as making sure that the changes introduce
fewer errors than a more serious restructure.

\subsubsection{Removing unnecessary parts from the analysis }

This is probably the most straightforward optimization we can make.
When we use the same or similar code for both the analysis and the
implementation, we may end up including code that is not relevant
to the analysis.

In the case of Marlowe, this was the case of the \texttt{Close} construct.
The \texttt{Close} construct refunds all the money and tokens remaining
in the accounts. At the time of analysis, we do not know how much
is available in each account, because it may depend on the value of
a choice made by the user in runtime. For that reason, the generated
constraints must include comparisons for every account. This makes
the implementation of the \texttt{Close} construct very expensive
to execute symbolically.

As it turns out, we do not need to implement the semantics of the
\texttt{Close} construct at all, because it is impossible for the
\texttt{Close} construct to produce a failed payment: it only pays
whatever there is available. So we can safely remove it from the analysis.

\subsubsection{Avoiding high level abstractions }

High level libraries like SBV, and even standards like SMTLib, support
the use and construction of considerably high level abstractions,
e.g: custom data-types, list, sets\dots{} Unfortunately, even though
high level abstractions make reasoning about code easier, they often
prevent certain optimisations, since they abstract out aspects of
the implementation that in our particular case may be concrete.

For example, in the case of Marlowe static analysis, we initially
implemented a symbolic associative map primitive with the only limitation
that it needed a concrete bound in the number of elements. This is
straightforward to realise using the symbolic implementation of list
and tuple, both provided by SBV. However, this is very costly in a
solver: searching a list symbolically requires constraints that check
the key of every element in the list up to the maximum capacity of
the associative map, because it is not known at the time of symbolic
execution how many elements the list has.

However, in Marlowe we know the values of all the keys that we are
going to use, because the contract is concrete, and only \texttt{Account},
\texttt{Choice}, and \texttt{Let} identifiers that are mentioned in
the contract will ever make it into the \texttt{State}. So we do not
need keys of the associative map to be symbolic, we can use a concrete
associative map with symbolic values.

Nevertheless, using a partially concrete map prevents us from using
it as the result of a function that takes symbolic parameters, so
we will need to use a continuation style insead. An alternative is
to assign numbers to each possible key, and store them in a symbolic
list, this will at least prevent the search from having to compare
a lookup with every key, since we will know in which position of the
list each value is stored.

\subsubsection{Reducing search space by normalising parameters\label{subsec:Reducing-search-space-by-normalising-parameters}}

The higher the number of degrees of freedom of the input, the larger
the search space, and the higher the load we are putting on the SMT
solver. In many cases, we do not need to explore the whole search
space to check satisfiability of a property. In particular, we may
be able to prove that some sets of inputs are equivalent in terms
of the property whose satisfiability we want to test.

For example, in the case of Marlowe\textquoteright s static analysis,
Marlowe allows several inputs to be combined in one single transaction.
This functionality is important because each transaction requires
the issuer to pay fees, so combining several inputs in a single transaction
can translate into important savings. On the other hand, it also means
that static analysis must consider many more possibilities, since
the number of ways of partitioning inputs in transactions grows exponentially
in the number of inputs.

However, we can devise a normal form for input traces, in which there
is a maximum of one input per transaction. We only need to make sure
that, for every trace, if it produces a warning, there exists a trace
with only one input per transaction that also produces a warning.
Using the automated proof assistant Isabelle \cite{nipkow2002isabelle},
we have shown the stronger property that splitting a transaction in
two does not modify the effect on a contract as long as the transaction
does not produce an error\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/isabelle/SingleInputTransactions.thy}
(last visited on 29/04/2020)}.

This optimization reduces the search space considerably, but transactions
may still have no inputs, so it still leaves room for many possible
ways of combining the inputs in transactions. We explain how we reduced
the search space even further in the next section.

\subsection{Heavyweight modifications \label{subsec:Heavyweight-modifications}}

When optimising, if our solution is a local minimum, small changes
to the parameters will not grant any improvement to the value function.
For that reason, in this section, we explore ways of optimizing that
may imply considerable rewriting of our properties, constraints, and
static analysis implementation in general.

\subsubsection{Reducing search space by using normalised execution paths relevant
to the property \label{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}}

One approach we can use that helps improve static analysis efficiency
is to implement the property thinking of the property we want to test
instead of trying to make it as close as possible to the full implementation
of the system. We do not even need to think about the representation
of a counterexample (we will discuss that in the next section), but
only in what are the conditions for the property to be false.

Unfortunately, having static analysis be very different from implementation
makes the approach much more error prone, and in making an argument
for the trustworthiness of the analysis there are many more assumptions
that need to be made and reasoned about. In Section~\ref{sec:Testing}
we explore ways of mitigating this problem.

For example, the main property we want to check is whether there is
any possible execution that produces a failed payment. Thus, we look
at executions instead of looking at inputs. The most complicated construct
in terms of executions is the \texttt{When} construct, since it allows
for two transactions that are separate in time to have different effects
depending on when they are issued, all other constructs will get resolved
instantly in one way or another. But, without loss of generality,
we can structure them as Figure~\ref{fig:Distribution-of-transactions}
shows: we can conceptually break the contract tree into subtrees,
where each subtree has a \texttt{When} construct as its root, with
the exception of the subtree that has the same root as the original
contract tree.

\begin{figure}
\includegraphics[width=1\textwidth]{figures/transactions}

\caption{Distribution of transactions with respect to a contract\label{fig:Distribution-of-transactions}}

\end{figure}

Each level of subtrees will correspond to a transaction, i.e: the
root subtree will correspond to the first transaction, the set of
subtrees that are children of the first subtree (in the original tree)
will correspond to the second transaction, and so on. There may be
some paths which do not require as many transactions/subtrees as other
paths, because they go across fewer \texttt{When} constructs before
finding a \texttt{Close} construct.

We can assign one transaction to each path of execution before a \texttt{When},
because if the maximum slot number of the transaction is lower than
the slot number in the timeout and the transaction has no inputs,
then the execution will stop before the \texttt{When}, so we need
to account for that possibility. In other words, if two parts of the
execution path may have different slot numbers when we go through
them, we need to consider them as separate transactions, because they
may have payments that depend on those values.

Of course, if the minimum slot number of the transaction is greater
or equal than the slot number in the timeout, then the first transaction
will execute past the \texttt{When} and into the timeout branch, and
we will not need the second transaction. In the same way, if we include
an input that is allowed by the \texttt{When} into the first transaction,
then the first transaction will continue into the corresponding \texttt{When}
branch, and we will not need the second transaction.

We cannot just constrain the maximum slot number of the first transaction
to be less than the timeout in the \texttt{When} because, if we do
that, then we will miss executions where one transaction expires several
\texttt{When} in a row, or when it expires one \texttt{When} and provides
the right input for later \texttt{When}. And we cannot not constrain
the maximum slot number of the first transaction because then we will
not know whether the first transaction goes beyond the \texttt{When}
or not.

We get around this problem by allowing the slot numbers of the first
and second transactions to be equal. If this happens, we will find
out that one of the transactions will be marked as \texttt{UselessTransaction}
by the semantics when we look at the counterexample. So we just need
to filter out all transactions that produce \texttt{UselessTransaction}
warnings in the final result.

Note however that, in Section~\ref{subsec:Reducing-search-space-by-normalising-parameters},
we already reduced the number of inputs per transaction to a maximum
of one, and the number of transactions to the maximum number of nested
\texttt{When}s plus one. But now we have also assigned each of the
transactions to a part of the contract, so we do not need a symbolic
list of transactions, we can have a finite concrete list of symbolic
transactions and we know in advance which of the transactions we are
working on at each moment during the symbolic execution.

One subtle detail that we realised during testing is that, even though
the diagram above shows that a \texttt{When} belongs to the next transaction,
because it will still be there after the previous transaction is executed,
the observations in the \texttt{Notify} cases. In the same way, the
values in the \texttt{Deposit} cases will refer to the \texttt{State}
before the \texttt{When} is executed, but to the slot interval of
the transaction that triggers the \texttt{When}, since the amount
to deposit must be calculated without considering the effects of the
deposit itself.

\subsubsection{Minimizing the representation of inputs and outputs }

When we initially implemented the efficient version of static analysis
for Marlowe, we did not pay any attention to the inputs and outputs.
The first version we wrote of the SBV property would not take any
symbolic parameters. Originally, it would simply take a concrete \texttt{Contract}
as input, and it would return a symbolic boolean that said whether
the \texttt{Contract} was valid or not. However, when the \texttt{Contract}
turns out to be invalid we also want to know why, so we later modified
the property to give a counterexample that illustrated what went wrong.
The original implementation still used some intermediate symbolic
variables, but they were anonymous, and they were created during the
exploration of the contract.

A simple way of obtaining a counterexample is to modify the output
of the function to return the offending trace using the symbolic \texttt{Maybe}
type. Surprisingly, this small change increases the amount of time
required by the symbolic analysis severalfold.

The solution to this increase in execution time was to pass as input
a fixed list of transactions, each of which being represented by a
tuple with four symbolic integers:
\begin{enumerate}
\item An integer representing the minimum slot
\item An integer representing the maximum slot
\item An integer representing the \texttt{When} case whose input is being
included in the transaction, where zero represents the timeout branch
(and there being no inputs to the transaction).
\item An integer representing the amount of money or tokens (if the input
is a \texttt{Deposit}), or the number chosen (if the input is a \texttt{Choice})
\end{enumerate}
If we do not need a transaction we just set all four numbers to $-1$.

In order to translate this sequence of numbers into a proper list
of transactions that is human readable and we can use to report the
counter example, we need to iterate through the list and see the evolution
of the contract with each transaction, using the concrete semantics.
This way we can gather the rest of necessary information, such as
whether the transaction input is a \texttt{Deposit}. We can also see
which of the transactions does not have an effect, i.e: they produce
\texttt{UselessTransaction} as we mentioned in Section~\ref{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}.

We use this separation of concerns between the static analysis and
the concrete semantics to our advantage, as an opportunity for applying
property based testing, as we explain in Section~\ref{subsec:Testing-for-consistency}.

\section{Testing for consistency and equivalence\label{sec:Testing}}

Errors often occur when optimising or translating. If the static analysis
implementation is close to the actual implementation, it is much less
likely that a mistake will be made when writing it. On the other hand,
if we have to craft a static analysis implementation that is equivalent
(in some sense) to the original, but is written in a very different
way, it is quite likely that we will make assumptions which do not
hold.

In order to avoid these problems, we use a mixture of automated proof
assistants to verify our assumptions, and property based testing to
test for consistency and equivalence.

\subsection{Testing for consistency\label{subsec:Testing-for-consistency} }

If our static analysis is not replicating all the functionality of
the semantics, we can use potential discrepancies as an opportunity
for testing, as shown in Figure~\ref{fig:Property-based-testing-for-consistency}.
We generate random contracts and we apply the static analysis to them
in order to try to find a counterexample that produces warnings. If
we cannot find any counterexamples then the test passes, but if we
find one we test it on the semantics and see whether the counterexample
indeed produces warnings in the semantics too, if it does not we have
found a problem either on the static analysis or the semantics.

A limitation of this approach is that it only tests for false positives,
and it does not prevent false negatives, those can be covered by testing
for equivalence (see Section~\ref{subsec:Testing-for-equivalence}).

In addition, we can add assertions to the process. In the case of
Marlowe, if the counterexample breaks one of the assertions or is
formed incorrectly it would also mean that there is a problem with
the static analysis. For example, it may be that the counterexample
refers to a \texttt{Case} of a \texttt{When} that does not exist,
or that it has invalid or ambiguous intervals. If it has \texttt{UselessTransactions}
is ok, because we are doing that on purpose, as we mentioned in Section~\ref{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}.

\begin{figure}
\begin{centering}
\includegraphics[width=0.5\textwidth]{figures/consistency_property}
\par\end{centering}
\caption{Property based testing for consistency\label{fig:Property-based-testing-for-consistency}}

\end{figure}

\subsection{Testing for equivalence\label{subsec:Testing-for-equivalence} }

If we have two static analysis implementations, we have another opportunity
for testing. In our case, we have one efficient implementation that
is very different from the semantics and one inefficient implementation
that is much closer to the semantics. We can do this as shown in Figure~\ref{fig:Property-based-testing-for-equivalence}.
We generate random contracts, we feed them to both static analysis
implementations, and we compare the results. If the results are the
same then test passes, if they are different then one of the implementations
is wrong.

\begin{figure}
\begin{centering}
\includegraphics[width=0.45\textwidth]{figures/equivalence_property}
\par\end{centering}
\caption{Property based testing for equivalence\label{fig:Property-based-testing-for-equivalence}}
\end{figure}

This approach covers both types of errors, i.e: false positives and
false negatives (for both of the implementations), but the execution
time of the tests is bounded from below by the slower of the two implementations.
The consistency approach is thus more efficient in finding false positives.

\section{Related work}

Work in \cite{mokhov2019formal} documents a similar effort to ensure
correctness of implementations by using Haskell and SBV library; the
authors also discuss performance of the analysis and apply this approach
to non-functional requirements.

The idea of using constraint solvers for finding bugs is not new,
and there has been a number of initiatives that have explored its
application to the verification of assertions in programs written
using general purpose programming languages \cite{gulwani2008program,jackson2000finding};
as well the compliance with usage protocols \cite{ball2001automatically,xie2005saturn}.

\cite{keshishzadeh2013early} also applies constraint solvers for
detecting problems in the usage of DSLs. The authors observe that
SMT solvers have limited support for non-linear constraints such as
exponentiation. This problem does not affect the current design of
Marlowe because it does not support multiplication by arbitrary variables,
and because all inputs are integer and bounded finitely.

\section{Future work }

In the future, we would like to extend static analysis to cover other
potential problems in Marlowe contracts and to aid their development.
We can envisage that we can use static analysis to locate unreachable
subcontracts, to allow developers to provide custom assertions and
check their satisfiability, and even to allow users to inspect the
possible maximum and minimum values that particular expressions can
reach.

\section{Conclusion}

In this paper, we have summarized the lessons learned from our work
on optimising the static analysis for Marlowe contracts. We have found
that there are two clear approaches to static analysis using SMT,
both have advantages and disadvantages. One is less error prone and
straightforward, but inefficient and hard to test; the other approach
is much more efficient, versatile, and testable, but more error prone.

We have also seen that many specific properties and restrictions characteristic
of the target DSL can be utilized both as optimisation opportunities
and, in our case, for completeness of the analysis. Since symbolic
execution of, for example, a turing complete language, would be intractable,
and would require us to manually set a bound; but this is not in the
case for Marlowe.

In the end, we have illustrated how to counteract the main disadvantage
of the later approach, its propensity to errors, by using property
based testing. This way we have obtained a static analysis implementation
that is efficient, versatile, testable, and reliable.

On the other hand, for the static analysis of Marlowe contracts, we
found out that when running statistics on the equivalence testing
property, most of the bugs were false negatives in the straightforward
implementation, and the optimised implementation seems to be more
reliable thanks to the consistency tests that we run beforehand.

Another advantage of the optimised implementation is that, because
it relies on fewer and simpler features, it is compatible with more
SMT solvers which, in turn, it means that it is less reliant on the
correctness of a single SMT solver. If an SMT solver fails to give
an answer, we can try a different one; if we want to ensure that a
contract is valid, we can test it with several SMT solvers and ensure
all of them agree.

\bibliographystyle{splncs04}
\bibliography{paper}

\end{document}

### 2020-04-30 papers/static_analysis2020/paper.tex
Add abstract and fix layout to keep it 16 pages

@@
 \maketitle
+\begin{abstract}
+SMT solvers can verify properties very efficiently, and they offer increasing flexibility on the ways that properties can be described.
+However, the way in which properties are described can affect the computational cost of verifying them considerably.
+In addition, it is quite hard to predict how different changes will affect the computational cost.
 
+In this paper, we discuss the lessons we have learned while implementing and optimising the static analysis functionality for Marlowe, a domain specific language for describing self-enforcing financial smart-contracts that can be deployed on a blockchain.
+\end{abstract}
 \section{Introduction\label{sec:Introduction}}
@@
 static analysis of arbitrary Haskell functions, with very few restrictions
-on how those functions are implemented.
-
-However, static analysis can be very computationally expensive, because
-it essentially needs to check every possible execution of the contract.
+on how those functions are implemented. However, static analysis can be very
+computationally expensive, because it essentially needs to check every
+possible execution of the contract.
 
@@
 
-The case study presented is that of the optimisation of Marlowe\textquoteright s
+The case study presented is that of the optimisation of Marlowe's
 static analysis. During this work, the techniques applied allowed
@@
 ways of representing some inputs, and the different representations
-have no impact on the analysis, then we can remove all but one when
+have no impact on the analysis, we can remove all but one when
 implementing static analysis.
@@
 to the property. Instead of using search space to model inputs, we
-use it to model possible executions, and we also ensure that we only
+use it to model possible executions, and we ensure that we only
 represent each possible group of equivalent executions once, i.e:
@@
 
-For example, in the case of Marlowe\textquoteright s static analysis,
+For example, in the case of Marlowe's static analysis,
 Marlowe allows several inputs to be combined in one single transaction.

### 2020-05-04 papers/static_analysis2020/paper.tex
Add remaining authors to paper

@@
 
-\author{Pablo {Lamela Seijas}\orcidID{0000-0002-1730-1219}\\
-\email{pablo.lamela@iohk.io}
-\institute{IOHK, Hong Kong}}
+\author{Pablo {Lamela Seijas}\inst{1}\orcidID{0000-0002-1730-1219} \and
+        David Smith\inst{1}\orcidID{0000-0003-1859-8007} \and
+        Simon Thompson\inst{1,2}\orcidID{0000-0002-2350-301X}}
 
-\authorrunning{P. Lamela Seijas}
+\authorrunning{P. Lamela Seijas, et al.}
+
+\institute{IOHK, Hong Kong,
+\email{pablo.lamela@iohk.io, david.smith@tweag.io, simon.thompson@iohk.io}  \and
+School of Computing, University of Kent, UK,
+\email{s.j.thompson@kent.ac.uk}}
 

### 2020-05-04 papers/static_analysis2020/paper.tex
Fix typo
Fix broken sentence

@@
 that, then we will miss executions where one transaction expires several
-\texttt{When} in a row, or when it expires one \texttt{When} and provides
+\texttt{When} in a row, or where it expires one \texttt{When} and provides
 the right input for later \texttt{When}. And we cannot not constrain
@@
 because it will still be there after the previous transaction is executed,
-the observations in the \texttt{Notify} cases. In the same way, the
-values in the \texttt{Deposit} cases will refer to the \texttt{State}
-before the \texttt{When} is executed, but to the slot interval of
-the transaction that triggers the \texttt{When}, since the amount
+the observations in the \texttt{Notify} cases will correspond to the
+\texttt{State} before the \texttt{When} is executed, but to the slot
+interfal of the transactions that triggers the \texttt{When}. The same
+is true for \texttt{Value}s in the \texttt{Deposit} cases, since the amount
 to deposit must be calculated without considering the effects of the

### 2020-05-04 papers/static_analysis2020/paper.tex
Fix typo
Signed-off-by: Pablo Lamela <pablo.lamela@iohk.io>

@@
 as input, and it would return a symbolic boolean that said whether
-the \texttt{Contract} was valid or not. However, when the \texttt{Contract}
-turns out to be invalid we also want to know why, so we later modified
+the \texttt{Contract} was valid or not. However, if a \texttt{Contract}
+turns out to be invalid, we also want to know why, so we later modified
 the property to give a counterexample that illustrated what went wrong.

### 2020-05-04 papers/static_analysis2020/paper.tex
Fix typo

@@
 \end{enumerate}
-If we do not need a transaction we just set all four numbers to $-1$.
+If we do not need a transaction, we just set all four numbers to $-1$.
 
@@
 of the contract with each transaction, using the concrete semantics.
-This way we can gather the rest of necessary information, such as
+This way, we can gather the rest of necessary information, such as
 whether the transaction input is a \texttt{Deposit}. We can also see

### 2020-05-04 papers/static_analysis2020/paper.tex
Fix typo

@@
 Marlowe, if the counterexample breaks one of the assertions or is
-formed incorrectly it would also mean that there is a problem with
+formed incorrectly, it would also mean that there is a problem with
 the static analysis. For example, it may be that the counterexample

### 2020-05-04 papers/static_analysis2020/paper.tex
Fix typo

@@
 using general purpose programming languages \cite{gulwani2008program,jackson2000finding};
-as well the compliance with usage protocols \cite{ball2001automatically,xie2005saturn}.
+as well as for the compliance with usage protocols
+\cite{ball2001automatically,xie2005saturn}.
 

### 2020-05-04 papers/static_analysis2020/paper.tex
Fix typo

@@
 of the target DSL can be utilized both as optimisation opportunities
-and, in our case, for completeness of the analysis. Since symbolic
-execution of, for example, a turing complete language, would be intractable,
+and, in our case, for completeness of the analysis. Symbolic execution
+of, for example, a turing complete language, would be intractable,
 and would require us to manually set a bound; but this is not in the

### 2020-05-04 papers/static_analysis2020/paper.tex
Fix typo
Address reviews by @shmish111

@@
 In this section, we explain the techniques that we mentioned in Section~\ref{sec:Introduction}
-more in detail, and we illustrate those techniques with examples from
+in more detail, and we illustrate those techniques with examples from
 the static analysis implementation of Marlowe.
@@
 intuitive and close to the original semantics and model that we are
-trying to analyze, as well as making sure that the changes introduce
-fewer errors than a more serious restructure.
+trying to analyze, as well as making changes less likely to introduce
+errors than a more serious restructure.
 
@@
 \texttt{Close} construct at all, because it is impossible for the
-\texttt{Close} construct to produce a failed payment: it only pays
-whatever there is available. So we can safely remove it from the analysis.
+\texttt{Close} construct to produce a failed payment
+(we have also proven this using Isabelle\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/isabelle/CloseSafe.thy}
+(last visited on 04/04/2020)}): it only pays whatever there is available.
+So we can safely remove it from the analysis.
 
@@
 
-Of course, if the minimum slot number of the transaction is greater
+On the other hand, if the minimum slot number of the transaction is greater
 or equal than the slot number in the timeout, then the first transaction
@@
 \texttt{State} before the \texttt{When} is executed, but to the slot
-interfal of the transactions that triggers the \texttt{When}. The same
+interval of the transactions that triggers the \texttt{When}. The same
 is true for \texttt{Value}s in the \texttt{Deposit} cases, since the amount
@@
 (in some sense) to the original, but is written in a very different
-way, it is quite likely that we will make assumptions which do not
+way, it is more likely that we will make assumptions which do not
 hold.

### 2020-05-04 papers/static_analysis2020/paper.tex
Clarify paragraph

@@
 One subtle detail that we realised during testing is that, even though
-the diagram above shows that a \texttt{When} belongs to the next transaction,
+the diagram above shows that a \texttt{When} belongs to the new transaction,
 because it will still be there after the previous transaction is executed,
 the observations in the \texttt{Notify} cases will correspond to the
-\texttt{State} before the \texttt{When} is executed, but to the slot
-interval of the transactions that triggers the \texttt{When}. The same
-is true for \texttt{Value}s in the \texttt{Deposit} cases, since the amount
-to deposit must be calculated without considering the effects of the
-deposit itself.
+information there was in the \texttt{State} before the \texttt{When}
+was executed, with the exception of the slot interval, which will still be
+considered to be that of the transaction that triggers the \texttt{When} (the
+new transaction). The same is true for \texttt{Value}s in the \texttt{Deposit}
+cases, since the amount to deposit must be calculated without considering the
+effects of the deposit itself, we need to refer to the \texttt{State} before
+the deposit was made.
 

### 2020-05-18 papers/static_analysis2020/paper.tex
Add template table for measurement section

@@
 \usepackage{babel}
+\usepackage{multirow}
+\usepackage[table,xcdraw]{xcolor}
+
+\DeclareUnicodeCharacter{00D7}{$\times$}
+
 \begin{document}
@@
 
-\paragraph*{Lightweight }
+\paragraph*{Lightweight}
 
@@
 
+\section{Measurements}
+
+\begin{table}[]
+\begin{tabular}{l||lllll}
+\multicolumn{1}{c||}{\textbf{Auction}} & \multicolumn{5}{c}{\textbf{Lightweight optimisations}} \\ \hline \hline
+\textbf{Number of participants}        & \multicolumn{1}{c||}{1}        & \multicolumn{1}{c|}{2}            & \multicolumn{1}{c|}{3}                         & \multicolumn{1}{c|}{4}                        & \multicolumn{1}{c}{5}    \\ \hline
+\textbf{Execution time}                & \multicolumn{1}{c||}{0.2205s}  & \multicolumn{1}{c|}{4m 45.2015s}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \color[HTML]{888888} N/A \\ \hline
+Generation overhead                    & \multicolumn{1}{c||}{76.61\%}  & \multicolumn{1}{c|}{96.40\%}      & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \color[HTML]{888888} N/A \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c||}{0.1689s}  & \multicolumn{1}{c|}{4m 34.9342s}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \color[HTML]{888888} N/A \\ \hline
+Z3 overhead                            & \multicolumn{1}{c||}{23.39\%}  & \multicolumn{1}{c|}{3.60\%}       & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \color[HTML]{888888} N/A \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c||}{0.0516s}  & \multicolumn{1}{c|}{10.2673s}     & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \color[HTML]{888888} N/A \\ \hline
+\textbf{RAM usage peak}                & \multicolumn{1}{c||}{44,248KB} & \multicolumn{1}{c|}{1,885,928KB}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \color[HTML]{888888} N/A \\ \hline
+\end{tabular}
+\end{table}
+
+\begin{table}[]
+\begin{tabular}{l||lllll}
+\multicolumn{1}{c||}{\textbf{Auction}} & \multicolumn{5}{c}{\textbf{Heavyweight optimisations}} \\ \hline \hline
+\textbf{Number of participants}        & \multicolumn{1}{c||}{1}        & \multicolumn{1}{c|}{2}        & \multicolumn{1}{c|}{3}        & \multicolumn{1}{c|}{4}              & \multicolumn{1}{c}{5}    \\ \hline
+\textbf{Execution time}                & \multicolumn{1}{c||}{0.01198s} & \multicolumn{1}{c|}{0.0289s}  & \multicolumn{1}{c|}{1.1138s}  & \multicolumn{1}{c|}{1h 5m 45.1377s} & \color[HTML]{888888} N/A \\ \hline
+Generation overhead                    & \multicolumn{1}{c||}{83.73\%}  & \multicolumn{1}{c|}{70.36\%}  & \multicolumn{1}{c|}{75.91\%}  & \multicolumn{1}{c|}{80.86\%}        & \color[HTML]{888888} N/A \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c||}{0.0100s}  & \multicolumn{1}{c|}{0.0203s}  & \multicolumn{1}{c|}{0.8455s}  & \multicolumn{1}{c|}{53m 10.0383s}   & \color[HTML]{888888} N/A \\ \hline
+Z3 overhead                            & \multicolumn{1}{c||}{16.27\%}  & \multicolumn{1}{c|}{29.64\%}  & \multicolumn{1}{c|}{24.09\%}  & \multicolumn{1}{c|}{19.14\%}        & \color[HTML]{888888} N/A \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c||}{0.0019s}  & \multicolumn{1}{c|}{0.0086s}  & \multicolumn{1}{c|}{0.2683s}  & \multicolumn{1}{c|}{12m 35.0993s}   & \color[HTML]{888888} N/A \\ \hline
+\textbf{RAM usage peak}                & \multicolumn{1}{c||}{17,020KB} & \multicolumn{1}{c|}{17,740KB} & \multicolumn{1}{c|}{49,668KB} & \multicolumn{1}{c|}{2,364,500KB}    & \color[HTML]{888888} N/A
+\end{tabular}
+\end{table}
+
 \section{Related work}
@@
 
-\section{Future work }
+\section{Future work}
 

### 2020-05-18 papers/static_analysis2020/paper.tex
Improve table layout

--- 2020-05-04 papers/static_analysis2020/paper.tex
@@
 
+\section{Measurements}
+
+\begin{table}[]
+\centering
+\scalebox{0.9}{
+\begin{tabular}{l||lllll}
+\multicolumn{6}{l}{\textbf{Auction contract -- \texttt{Auction.hs}}} \\ \hline \hline
+\multicolumn{6}{c}{\textbf{Lightweight optimisations}} \\ \hline \hline
+\textbf{Num.~participants}        & \multicolumn{1}{c||}{1}        & \multicolumn{1}{c|}{2}            & \multicolumn{1}{c|}{3}                         & \multicolumn{1}{c|}{4}                        & \multicolumn{1}{c}{5}    \\ \hline
+\textbf{Execution time}                & \multicolumn{1}{c||}{0.2205s}  & \multicolumn{1}{c|}{4m 45.2015s}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+Generation overhead                    & \multicolumn{1}{c||}{76.61\%}  & \multicolumn{1}{c|}{96.40\%}      & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c||}{0.1689s}  & \multicolumn{1}{c|}{4m 34.9342s}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+Z3 overhead                            & \multicolumn{1}{c||}{23.39\%}  & \multicolumn{1}{c|}{3.60\%}       & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c||}{0.0516s}  & \multicolumn{1}{c|}{10.2673s}     & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\textbf{RAM usage peak}                & \multicolumn{1}{c||}{44,248KB} & \multicolumn{1}{c|}{1,885,928KB}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline \hline
+\multicolumn{6}{c}{\textbf{Heavyweight optimisations}} \\ \hline \hline
+\textbf{Num.~participants}        & \multicolumn{1}{c||}{1}        & \multicolumn{1}{c|}{2}        & \multicolumn{1}{c|}{3}        & \multicolumn{1}{c|}{4}              & \multicolumn{1}{c}{5}    \\ \hline
+\textbf{Execution time}                & \multicolumn{1}{c||}{0.01198s} & \multicolumn{1}{c|}{0.0289s}  & \multicolumn{1}{c|}{1.1138s}  & \multicolumn{1}{c|}{1h 5m 45.1377s} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+Generation overhead                    & \multicolumn{1}{c||}{83.73\%}  & \multicolumn{1}{c|}{70.36\%}  & \multicolumn{1}{c|}{75.91\%}  & \multicolumn{1}{c|}{80.86\%}        & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c||}{0.0100s}  & \multicolumn{1}{c|}{0.0203s}  & \multicolumn{1}{c|}{0.8455s}  & \multicolumn{1}{c|}{53m 10.0383s}   & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+Z3 overhead                            & \multicolumn{1}{c||}{16.27\%}  & \multicolumn{1}{c|}{29.64\%}  & \multicolumn{1}{c|}{24.09\%}  & \multicolumn{1}{c|}{19.14\%}        & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c||}{0.0019s}  & \multicolumn{1}{c|}{0.0086s}  & \multicolumn{1}{c|}{0.2683s}  & \multicolumn{1}{c|}{12m 35.0993s}   & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\textbf{RAM usage peak}                & \multicolumn{1}{c||}{17,020KB} & \multicolumn{1}{c|}{17,740KB} & \multicolumn{1}{c|}{49,668KB} & \multicolumn{1}{c|}{2,364,500KB}    & \multicolumn{1}{c}{\color[HTML]{888888} N/A}
+\end{tabular}
+}
+\caption{Execution cost measurement of auction contract\label{table1}}
+\end{table}
+
 \section{Related work}

### 2020-05-18 papers/static_analysis2020/paper.tex
Fix ordering of percentages

@@
 \textbf{Execution time}                & \multicolumn{1}{c||}{0.01198s} & \multicolumn{1}{c|}{0.0289s}  & \multicolumn{1}{c|}{1.1138s}  & \multicolumn{1}{c|}{1h 5m 45.1377s} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-Generation overhead                    & \multicolumn{1}{c||}{83.73\%}  & \multicolumn{1}{c|}{70.36\%}  & \multicolumn{1}{c|}{75.91\%}  & \multicolumn{1}{c|}{80.86\%}        & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c||}{0.0100s}  & \multicolumn{1}{c|}{0.0203s}  & \multicolumn{1}{c|}{0.8455s}  & \multicolumn{1}{c|}{53m 10.0383s}   & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-Z3 overhead                            & \multicolumn{1}{c||}{16.27\%}  & \multicolumn{1}{c|}{29.64\%}  & \multicolumn{1}{c|}{24.09\%}  & \multicolumn{1}{c|}{19.14\%}        & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c||}{0.0019s}  & \multicolumn{1}{c|}{0.0086s}  & \multicolumn{1}{c|}{0.2683s}  & \multicolumn{1}{c|}{12m 35.0993s}   & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+Generation overhead                    & \multicolumn{1}{c||}{16.27\%}  & \multicolumn{1}{c|}{29.64\%}  & \multicolumn{1}{c|}{24.09\%}  & \multicolumn{1}{c|}{80.86\%}        & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c||}{0.001949s}  & \multicolumn{1}{c|}{0.008566s}  & \multicolumn{1}{c|}{0.268314s}  & \multicolumn{1}{c|}{53m 10.038344s}   & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+Z3 overhead                            & \multicolumn{1}{c||}{83.73\%}  & \multicolumn{1}{c|}{70.36\%}  & \multicolumn{1}{c|}{75.91\%}  & \multicolumn{1}{c|}{19.14\%}        & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c||}{0.010031s}  & \multicolumn{1}{c|}{0.020334s}  & \multicolumn{1}{c|}{0.845486s}  & \multicolumn{1}{c|}{12m 35.099356s}   & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
 \textbf{RAM usage peak}                & \multicolumn{1}{c||}{17,020KB} & \multicolumn{1}{c|}{17,740KB} & \multicolumn{1}{c|}{49,668KB} & \multicolumn{1}{c|}{2,364,500KB}    & \multicolumn{1}{c}{\color[HTML]{888888} N/A}

### 2020-05-18 papers/static_analysis2020/paper.tex
Remove vertical line in table

--- 2020-05-18 papers/static_analysis2020/paper.tex
@@
 \multicolumn{6}{c}{\textbf{Lightweight optimisations}} \\ \hline \hline
-\textbf{Num.~participants}        & \multicolumn{1}{c||}{1}        & \multicolumn{1}{c|}{2}            & \multicolumn{1}{c|}{3}                         & \multicolumn{1}{c|}{4}                        & \multicolumn{1}{c}{5}    \\ \hline
-\textbf{Execution time}                & \multicolumn{1}{c||}{0.2205s}  & \multicolumn{1}{c|}{4m 45.2015s}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-Generation overhead                    & \multicolumn{1}{c||}{76.61\%}  & \multicolumn{1}{c|}{96.40\%}      & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c||}{0.1689s}  & \multicolumn{1}{c|}{4m 34.9342s}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-Z3 overhead                            & \multicolumn{1}{c||}{23.39\%}  & \multicolumn{1}{c|}{3.60\%}       & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c||}{0.0516s}  & \multicolumn{1}{c|}{10.2673s}     & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-\textbf{RAM usage peak}                & \multicolumn{1}{c||}{44,248KB} & \multicolumn{1}{c|}{1,885,928KB}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline \hline
+\textbf{Num.~participants}        & \multicolumn{1}{c|}{1}        & \multicolumn{1}{c|}{2}            & \multicolumn{1}{c|}{3}                         & \multicolumn{1}{c|}{4}                        & \multicolumn{1}{c}{5}    \\ \hline
+\textbf{Execution time}                & \multicolumn{1}{c|}{0.2205s}  & \multicolumn{1}{c|}{4m 45.2015s}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+Generation overhead                    & \multicolumn{1}{c|}{76.61\%}  & \multicolumn{1}{c|}{96.40\%}      & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.1689s}  & \multicolumn{1}{c|}{4m 34.9342s}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+Z3 overhead                            & \multicolumn{1}{c|}{23.39\%}  & \multicolumn{1}{c|}{3.60\%}       & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.0516s}  & \multicolumn{1}{c|}{10.2673s}     & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\textbf{RAM usage peak}                & \multicolumn{1}{c|}{44,248KB} & \multicolumn{1}{c|}{1,885,928KB}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline \hline
 \multicolumn{6}{c}{\textbf{Heavyweight optimisations}} \\ \hline \hline
-\textbf{Num.~participants}        & \multicolumn{1}{c||}{1}        & \multicolumn{1}{c|}{2}        & \multicolumn{1}{c|}{3}        & \multicolumn{1}{c|}{4}              & \multicolumn{1}{c}{5}    \\ \hline
-\textbf{Execution time}                & \multicolumn{1}{c||}{0.01198s} & \multicolumn{1}{c|}{0.0289s}  & \multicolumn{1}{c|}{1.1138s}  & \multicolumn{1}{c|}{1h 5m 45.1377s} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-Generation overhead                    & \multicolumn{1}{c||}{83.73\%}  & \multicolumn{1}{c|}{70.36\%}  & \multicolumn{1}{c|}{75.91\%}  & \multicolumn{1}{c|}{80.86\%}        & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c||}{0.0100s}  & \multicolumn{1}{c|}{0.0203s}  & \multicolumn{1}{c|}{0.8455s}  & \multicolumn{1}{c|}{53m 10.0383s}   & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-Z3 overhead                            & \multicolumn{1}{c||}{16.27\%}  & \multicolumn{1}{c|}{29.64\%}  & \multicolumn{1}{c|}{24.09\%}  & \multicolumn{1}{c|}{19.14\%}        & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c||}{0.0019s}  & \multicolumn{1}{c|}{0.0086s}  & \multicolumn{1}{c|}{0.2683s}  & \multicolumn{1}{c|}{12m 35.0993s}   & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-\textbf{RAM usage peak}                & \multicolumn{1}{c||}{17,020KB} & \multicolumn{1}{c|}{17,740KB} & \multicolumn{1}{c|}{49,668KB} & \multicolumn{1}{c|}{2,364,500KB}    & \multicolumn{1}{c}{\color[HTML]{888888} N/A}
+\textbf{Num.~participants}        & \multicolumn{1}{c|}{1}        & \multicolumn{1}{c|}{2}        & \multicolumn{1}{c|}{3}        & \multicolumn{1}{c|}{4}              & \multicolumn{1}{c}{5}    \\ \hline
+\textbf{Execution time}                & \multicolumn{1}{c|}{0.01198s} & \multicolumn{1}{c|}{0.0289s}  & \multicolumn{1}{c|}{1.1138s}  & \multicolumn{1}{c|}{1h 5m 45.1377s} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+Generation overhead                    & \multicolumn{1}{c|}{16.27\%}  & \multicolumn{1}{c|}{29.64\%}  & \multicolumn{1}{c|}{24.09\%}  & \multicolumn{1}{c|}{80.86\%}        & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.001949s}  & \multicolumn{1}{c|}{0.008566s}  & \multicolumn{1}{c|}{0.268314s}  & \multicolumn{1}{c|}{53m 10.038344s}   & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+Z3 overhead                            & \multicolumn{1}{c|}{83.73\%}  & \multicolumn{1}{c|}{70.36\%}  & \multicolumn{1}{c|}{75.91\%}  & \multicolumn{1}{c|}{19.14\%}        & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.010031s}  & \multicolumn{1}{c|}{0.020334s}  & \multicolumn{1}{c|}{0.845486s}  & \multicolumn{1}{c|}{12m 35.099356s}   & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\textbf{RAM usage peak}                & \multicolumn{1}{c|}{17,020KB} & \multicolumn{1}{c|}{17,740KB} & \multicolumn{1}{c|}{49,668KB} & \multicolumn{1}{c|}{2,364,500KB}    & \multicolumn{1}{c}{\color[HTML]{888888} N/A}
 \end{tabular}

### 2020-05-18 papers/static_analysis2020/paper.tex
Add table for crowdfunding

@@
 
+\begin{table}[]
+\centering
+\scalebox{0.9}{
+\begin{tabular}{l||lllll}
+\multicolumn{6}{l}{\textbf{Crowdfunding contract -- \texttt{CrowdFunding.hs}}} \\ \hline \hline
+\multicolumn{6}{c}{\textbf{Lightweight optimisations}} \\ \hline \hline
+\textbf{Num.~participants}        & \multicolumn{1}{c|}{1}        & \multicolumn{1}{c|}{2}            & \multicolumn{1}{c|}{3}                         & \multicolumn{1}{c|}{4}                        & \multicolumn{1}{c}{5}    \\ \hline
+\textbf{Execution time}                & \multicolumn{1}{c|}{0.6298s}  & \multicolumn{1}{c|}{50m 53.4597s}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+Generation overhead                    & \multicolumn{1}{c|}{85.02\%}  & \multicolumn{1}{c|}{99.82\%}      & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.535456s}  & \multicolumn{1}{c|}{50m 47.963473s	}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+Z3 overhead                            & \multicolumn{1}{c|}{14.98\%}  & \multicolumn{1}{c|}{0.18\%}       & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.094344s}  & \multicolumn{1}{c|}{5.496227s}     & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\textbf{RAM usage peak}                & \multicolumn{1}{c|}{111,980KB} & \multicolumn{1}{c|}{5,641,056KB}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline \hline
+\multicolumn{6}{c}{\textbf{Heavyweight optimisations}} \\ \hline \hline
+\textbf{Num.~participants}        & \multicolumn{1}{c|}{1}        & \multicolumn{1}{c|}{2}        & \multicolumn{1}{c|}{3}        & \multicolumn{1}{c|}{4}              & \multicolumn{1}{c}{5}    \\ \hline
+\textbf{Execution time}                & \multicolumn{1}{c|}{0.0125s} & \multicolumn{1}{c|}{0.041s}  & \multicolumn{1}{c|}{1.0515s}  & \multicolumn{1}{c|}{32m 15.4478s} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+Generation overhead                    & \multicolumn{1}{c|}{16.68\%}  & \multicolumn{1}{c|}{25.36\%}  & \multicolumn{1}{c|}{37.23\%}  & \multicolumn{1}{c|}{69.83\%}        & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.002085s}  & \multicolumn{1}{c|}{0.010398s}  & \multicolumn{1}{c|}{0.391473s}  & \multicolumn{1}{c|}{22m 31.523199s}   & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+Z3 overhead                            & \multicolumn{1}{c|}{83.32\%}  & \multicolumn{1}{c|}{74.64\%}  & \multicolumn{1}{c|}{62.77\%}  & \multicolumn{1}{c|}{30.17\%}        & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.010415s}  & \multicolumn{1}{c|}{0.030602s}  & \multicolumn{1}{c|}{0.660027s}  & \multicolumn{1}{c|}{9m 43.924601s}   & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\textbf{RAM usage peak}                & \multicolumn{1}{c|}{17,016KB} & \multicolumn{1}{c|}{18,768KB} & \multicolumn{1}{c|}{62,108KB} & \multicolumn{1}{c|}{3,715,020KB}    & \multicolumn{1}{c}{\color[HTML]{888888} N/A}
+\end{tabular}
+}
+\caption{Execution cost measurement of crowdfunding contract\label{table1}}
+\end{table}
+
 \section{Related work}

### 2020-05-18 papers/static_analysis2020/paper.tex
Fix layout

@@
 Generation overhead                    & \multicolumn{1}{c|}{85.02\%}   & \multicolumn{1}{c|}{99.82\%}        & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.535456s}  & \multicolumn{1}{c|}{50m 47.963473s	}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.535456s} & \multicolumn{1}{c|}{50m 47.963473s} & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
 Z3 overhead                            & \multicolumn{1}{c|}{14.98\%}   & \multicolumn{1}{c|}{0.18\%}         & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline

### 2020-05-18 papers/static_analysis2020/paper.tex
Add measurements for rent contract

@@
 }
-\caption{Execution cost measurement of crowdfunding contract\label{table1}}
+\caption{Execution cost measurement of crowdfunding contract\label{table2}}
+\end{table}
+
+\begin{table}[]
+\centering
+\scalebox{0.9}{
+\begin{tabular}{l||lllll}
+\multicolumn{6}{l}{\textbf{Rent contract -- \texttt{Rent.hs}}} \\ \hline \hline
+\multicolumn{6}{c}{\textbf{Lightweight optimisations}} \\ \hline \hline
+\textbf{Num.~months}                   & \multicolumn{1}{c|}{1}        & \multicolumn{1}{c|}{2}         & \multicolumn{1}{c|}{3}                         & \multicolumn{1}{c|}{4}                        & \multicolumn{1}{c}{5}    \\ \hline
+\textbf{Execution time}                & \multicolumn{1}{c|}{0.2850s}   & \multicolumn{1}{c|}{3.0303s}   & \multicolumn{1}{c|}{2m 53.2458s} & \multicolumn{1}{c|}{3h 22m 13.0122s} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+Generation overhead                    & \multicolumn{1}{c|}{77.55\%}  & \multicolumn{1}{c|}{91.25\%}   & \multicolumn{1}{c|}{99.18\%}     & \multicolumn{1}{c|}{99.94\%}         & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.2210s}  & \multicolumn{1}{c|}{2.7651s}   & \multicolumn{1}{c|}{2m 51.8252s} & \multicolumn{1}{c|}{3h 22m 5.7324s}  & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+Z3 overhead                            & \multicolumn{1}{c|}{22.45\%}  & \multicolumn{1}{c|}{8.75\%}    & \multicolumn{1}{c|}{0.82\%}      & \multicolumn{1}{c|}{0.06\%}          & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.0640s}  & \multicolumn{1}{c|}{0.2652s}   & \multicolumn{1}{c|}{0m 1.4206s}  & \multicolumn{1}{c|}{7.2798s}         & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\textbf{RAM usage peak}                & \multicolumn{1}{c|}{42,052KB} & \multicolumn{1}{c|}{221,960KB} & \multicolumn{1}{c|}{1,237,092KB} & \multicolumn{1}{c|}{9,160,616KB}     & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline \hline
+\multicolumn{6}{c}{\textbf{Heavyweight optimisations}} \\ \hline \hline
+\textbf{Num.~months}                   & \multicolumn{1}{c|}{1}        & \multicolumn{1}{c|}{2}        & \multicolumn{1}{c|}{3}        & \multicolumn{1}{c|}{4}        & \multicolumn{1}{c}{5}    \\ \hline
+\textbf{Execution time}                & \multicolumn{1}{c|}{0.0114s}  & \multicolumn{1}{c|}{0.0111s}  & \multicolumn{1}{c|}{0.01132s} & \multicolumn{1}{c|}{0.01124s} & \multicolumn{1}{c}{0.01255s} \\ \hline
+Generation overhead                    & \multicolumn{1}{c|}{11.55\%}  & \multicolumn{1}{c|}{11.95\%}  & \multicolumn{1}{c|}{13.65\%}  & \multicolumn{1}{c|}{13.69\%}  & \multicolumn{1}{c}{21.00\%} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.0013s}  & \multicolumn{1}{c|}{0.0013s}  & \multicolumn{1}{c|}{0.0015s}  & \multicolumn{1}{c|}{0.0015s}  & \multicolumn{1}{c}{0.0026s} \\ \hline
+Z3 overhead                            & \multicolumn{1}{c|}{88.45\%}  & \multicolumn{1}{c|}{88.05\%}  & \multicolumn{1}{c|}{86.35\%}  & \multicolumn{1}{c|}{86.31\%}  & \multicolumn{1}{c}{79.00\%} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.0101s}  & \multicolumn{1}{c|}{0.0098s}  & \multicolumn{1}{c|}{0.0098s}  & \multicolumn{1}{c|}{0.0097s}  & \multicolumn{1}{c}{0.0099s} \\ \hline
+\textbf{RAM usage peak}                & \multicolumn{1}{c|}{15,536KB} & \multicolumn{1}{c|}{15,364KB} & \multicolumn{1}{c|}{15,400KB} & \multicolumn{1}{c|}{15,364KB} & \multicolumn{1}{c}{15,540KB}
+\end{tabular}
+}
+\caption{Execution cost measurement of crowdfunding contract\label{table3}}
 \end{table}

### 2020-05-18 papers/static_analysis2020/paper.tex
Add coupon bond measurements

@@
 Generation overhead                    & \multicolumn{1}{c|}{85.02\%}   & \multicolumn{1}{c|}{99.82\%}      & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.535456s} & \multicolumn{1}{c|}{50m 47.963473s} & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.5356s}   & \multicolumn{1}{c|}{50m 47.9635s} & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
 Z3 overhead                            & \multicolumn{1}{c|}{14.98\%}   & \multicolumn{1}{c|}{0.18\%}       & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.094344s} & \multicolumn{1}{c|}{5.496227s}      & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.0943s}   & \multicolumn{1}{c|}{5.4962s}      & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
 \textbf{RAM usage peak}                & \multicolumn{1}{c|}{111,980KB} & \multicolumn{1}{c|}{5,641,056KB}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline \hline
@@
 Generation overhead                    & \multicolumn{1}{c|}{16.68\%}  & \multicolumn{1}{c|}{25.36\%}  & \multicolumn{1}{c|}{37.23\%}  & \multicolumn{1}{c|}{69.83\%}      & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.002085s} & \multicolumn{1}{c|}{0.010398s} & \multicolumn{1}{c|}{0.391473s} & \multicolumn{1}{c|}{22m 31.523199s} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.0021s}  & \multicolumn{1}{c|}{0.0104s}  & \multicolumn{1}{c|}{0.3915s}  & \multicolumn{1}{c|}{22m 31.5232s} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
 Z3 overhead                            & \multicolumn{1}{c|}{83.32\%}  & \multicolumn{1}{c|}{74.64\%}  & \multicolumn{1}{c|}{62.77\%}  & \multicolumn{1}{c|}{30.17\%}      & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.010415s} & \multicolumn{1}{c|}{0.030602s} & \multicolumn{1}{c|}{0.660027s} & \multicolumn{1}{c|}{9m 43.924601s}  & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.0104s}  & \multicolumn{1}{c|}{0.0306s}  & \multicolumn{1}{c|}{0.6600s}  & \multicolumn{1}{c|}{9m 43.9246s}  & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
 \textbf{RAM usage peak}                & \multicolumn{1}{c|}{17,016KB} & \multicolumn{1}{c|}{18,768KB} & \multicolumn{1}{c|}{62,108KB} & \multicolumn{1}{c|}{3,715,020KB}  & \multicolumn{1}{c}{\color[HTML]{888888} N/A}
@@
 Z3 overhead                            & \multicolumn{1}{c|}{22.45\%}  & \multicolumn{1}{c|}{8.75\%}    & \multicolumn{1}{c|}{0.82\%}      & \multicolumn{1}{c|}{0.06\%}          & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.0640s}  & \multicolumn{1}{c|}{0.2652s}   & \multicolumn{1}{c|}{0m 1.4206s}  & \multicolumn{1}{c|}{7.2798s}         & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.0640s}  & \multicolumn{1}{c|}{0.2652s}   & \multicolumn{1}{c|}{1.4206s}  & \multicolumn{1}{c|}{7.2798s}         & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
 \textbf{RAM usage peak}                & \multicolumn{1}{c|}{42,052KB} & \multicolumn{1}{c|}{221,960KB} & \multicolumn{1}{c|}{1,237,092KB} & \multicolumn{1}{c|}{9,160,616KB}     & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline \hline
@@
 \caption{Execution cost measurement of crowdfunding contract\label{table3}}
+\end{table}
+
+\begin{table}[]
+\centering
+\scalebox{0.9}{
+\begin{tabular}{l||lllll}
+\multicolumn{6}{l}{\textbf{Coupon bond contract - ACTUS.hs}} \\ \hline \hline
+\multicolumn{6}{c}{\textbf{Lightweight optimisations}} \\ \hline \hline
+\textbf{Num.~months}                   & \multicolumn{1}{c|}{2}        & \multicolumn{1}{c|}{3}         & \multicolumn{1}{c|}{4}           & \multicolumn{1}{c|}{5}            & \multicolumn{1}{c}{6}    \\ \hline
+\textbf{Execution time}                & \multicolumn{1}{c|}{0.8293s}  & \multicolumn{1}{c|}{5.8887s}   & \multicolumn{1}{c|}{1m 35.3930s} & \multicolumn{1}{c|}{26m 36.4585s} & \multicolumn{1}{c}{9h 50m 22.3418s} \\ \hline
+Generation overhead                    & \multicolumn{1}{c|}{76.57\%}  & \multicolumn{1}{c|}{74.91\%}   & \multicolumn{1}{c|}{89.29\%}     & \multicolumn{1}{c|}{72.34\%}      & \multicolumn{1}{c}{76.66\%} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.6350s}  & \multicolumn{1}{c|}{4.4112s}   & \multicolumn{1}{c|}{1m 25.1764s} & \multicolumn{1}{c|}{19m 14.8781s} & \multicolumn{1}{c}{7h 32m 34.7672s} \\ \hline
+Z3 overhead                            & \multicolumn{1}{c|}{23.43\%}  & \multicolumn{1}{c|}{25.09\%}   & \multicolumn{1}{c|}{10.71\%}     & \multicolumn{1}{c|}{27.66\%}      & \multicolumn{1}{c}{23.34\%} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.1943s}  & \multicolumn{1}{c|}{1.4775s}   & \multicolumn{1}{c|}{10.2166s}    & \multicolumn{1}{c|}{7m 21.5804s}  & \multicolumn{1}{c}{2h 17m 47.5746s} \\ \hline
+\textbf{RAM usage peak}                & \multicolumn{1}{c|}{74,180KB} & \multicolumn{1}{c|}{209,724KB} & \multicolumn{1}{c|}{940,924KB}   & \multicolumn{1}{c|}{3,283,384KB}  & \multicolumn{1}{c}{13,483,908KB} \\ \hline \hline
+\multicolumn{6}{c}{\textbf{Heavyweight optimisations}} \\ \hline \hline
+\textbf{Num.~months}                   & \multicolumn{1}{c|}{2}        & \multicolumn{1}{c|}{3}        & \multicolumn{1}{c|}{4}        & \multicolumn{1}{c|}{5}        & \multicolumn{1}{c}{6}    \\ \hline
+\textbf{Execution time}                & \multicolumn{1}{c|}{0.0092s}  & \multicolumn{1}{c|}{0.0095s}  & \multicolumn{1}{c|}{0.0097s}  & \multicolumn{1}{c|}{0.0102s}  & \multicolumn{1}{c}{0.0105s} \\ \hline
+Generation overhead                    & \multicolumn{1}{c|}{14.51\%}  & \multicolumn{1}{c|}{15.50\%}  & \multicolumn{1}{c|}{15.51\%}  & \multicolumn{1}{c|}{19.71\%}  & \multicolumn{1}{c}{19.69\%} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.0013s}  & \multicolumn{1}{c|}{0.0015s}  & \multicolumn{1}{c|}{0.0015s}  & \multicolumn{1}{c|}{0.0020s}  & \multicolumn{1}{c}{0.0021s} \\ \hline
+Z3 overhead                            & \multicolumn{1}{c|}{85.49\%}  & \multicolumn{1}{c|}{84.50\%}  & \multicolumn{1}{c|}{84.49\%}  & \multicolumn{1}{c|}{80.29\%}  & \multicolumn{1}{c}{80.31\%} \\ \hline
+\multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.0079s}  & \multicolumn{1}{c|}{0.0080s}  & \multicolumn{1}{c|}{0.0082s}  & \multicolumn{1}{c|}{0.0082s}  & \multicolumn{1}{c}{0.0085s} \\ \hline
+\textbf{RAM usage peak}                & \multicolumn{1}{c|}{15,636KB} & \multicolumn{1}{c|}{15,816KB} & \multicolumn{1}{c|}{15,788KB} & \multicolumn{1}{c|}{15,780KB} & \multicolumn{1}{c}{15,760KB}
+\end{tabular}
+}
+\caption{Execution cost measurement of coupon contract\label{table4}}
 \end{table}

### 2020-05-19 papers/static_analysis2020/paper.tex
Add contract size to measurements

@@
 \multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.010031s}  & \multicolumn{1}{c|}{0.020334s} & \multicolumn{1}{c|}{0.845486s} & \multicolumn{1}{c|}{12m 35.099356s} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-\textbf{RAM usage peak}                & \multicolumn{1}{c|}{17,020KB}   & \multicolumn{1}{c|}{17,740KB}  & \multicolumn{1}{c|}{49,668KB}  & \multicolumn{1}{c|}{2,364,500KB}    & \multicolumn{1}{c}{\color[HTML]{888888} N/A}
+\textbf{RAM usage peak}                & \multicolumn{1}{c|}{17,020KB}   & \multicolumn{1}{c|}{17,740KB}  & \multicolumn{1}{c|}{49,668KB}  & \multicolumn{1}{c|}{2,364,500KB}    & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline \hline
+\textbf{Contract size (chars)}         & \multicolumn{1}{c|}{275}        & \multicolumn{1}{c|}{3,399}     & \multicolumn{1}{c|}{89,335}    & \multicolumn{1}{c|}{4,747,361}      & \multicolumn{1}{c}{413,784,559}
 \end{tabular}
@@
 \multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.0104s}  & \multicolumn{1}{c|}{0.0306s}  & \multicolumn{1}{c|}{0.6600s}  & \multicolumn{1}{c|}{9m 43.9246s}  & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-\textbf{RAM usage peak}                & \multicolumn{1}{c|}{17,016KB} & \multicolumn{1}{c|}{18,768KB} & \multicolumn{1}{c|}{62,108KB} & \multicolumn{1}{c|}{3,715,020KB}  & \multicolumn{1}{c}{\color[HTML]{888888} N/A}
+\textbf{RAM usage peak}                & \multicolumn{1}{c|}{17,016KB} & \multicolumn{1}{c|}{18,768KB} & \multicolumn{1}{c|}{62,108KB} & \multicolumn{1}{c|}{3,715,020KB}  & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline \hline
+\textbf{Contract size (chars)}         & \multicolumn{1}{c|}{857}      & \multicolumn{1}{c|}{12,704}   & \multicolumn{1}{c|}{364,824}  & \multicolumn{1}{c|}{19,462,278}   & \multicolumn{1}{c}{1,690,574,798}
 \end{tabular}
@@
 \multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.0101s}  & \multicolumn{1}{c|}{0.0098s}  & \multicolumn{1}{c|}{0.0098s}  & \multicolumn{1}{c|}{0.0097s}  & \multicolumn{1}{c}{0.0099s} \\ \hline
-\textbf{RAM usage peak}                & \multicolumn{1}{c|}{15,536KB} & \multicolumn{1}{c|}{15,364KB} & \multicolumn{1}{c|}{15,400KB} & \multicolumn{1}{c|}{15,364KB} & \multicolumn{1}{c}{15,540KB}
+\textbf{RAM usage peak}                & \multicolumn{1}{c|}{15,536KB} & \multicolumn{1}{c|}{15,364KB} & \multicolumn{1}{c|}{15,400KB} & \multicolumn{1}{c|}{15,364KB} & \multicolumn{1}{c}{15,540KB} \\ \hline \hline
+\textbf{Contract size (chars)}         & \multicolumn{1}{c|}{339}      & \multicolumn{1}{c|}{595}      & \multicolumn{1}{c|}{852}      & \multicolumn{1}{c|}{1,109}    & \multicolumn{1}{c}{1,366}
 \end{tabular}
@@
 \multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.0079s}  & \multicolumn{1}{c|}{0.0080s}  & \multicolumn{1}{c|}{0.0082s}  & \multicolumn{1}{c|}{0.0082s}  & \multicolumn{1}{c}{0.0085s} \\ \hline
-\textbf{RAM usage peak}                & \multicolumn{1}{c|}{15,636KB} & \multicolumn{1}{c|}{15,816KB} & \multicolumn{1}{c|}{15,788KB} & \multicolumn{1}{c|}{15,780KB} & \multicolumn{1}{c}{15,760KB}
+\textbf{RAM usage peak}                & \multicolumn{1}{c|}{15,636KB} & \multicolumn{1}{c|}{15,816KB} & \multicolumn{1}{c|}{15,788KB} & \multicolumn{1}{c|}{15,780KB} & \multicolumn{1}{c}{15,760KB} \\ \hline \hline
+\textbf{Contract size (chars)}         & \multicolumn{1}{c|}{479}      & \multicolumn{1}{c|}{635}      & \multicolumn{1}{c|}{791}      & \multicolumn{1}{c|}{947}      & \multicolumn{1}{c}{1,103}
 \end{tabular}

### 2020-05-20 papers/static_analysis2020/paper.tex
Add measurements section explanation

@@
 
+In our experiments, the heavy weight optimisations considerably reduced
+the requirements of both processing time and memory for Marlowe's static
+analysis. In Tables~\ref{table1},~\ref{table2},~\ref{table3},~and~\ref{table4},
+we present the results of measuring the performance of static analysis on
+four example Marlowe contracts,
+before\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/src/Language/Marlowe/Analysis/FSSemantics.hs} [last visited 19-05-2020]}
+and after\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/src/Language/Marlowe/Analysis/FSSemanticsFastVerbose.hs} [last visited 19-05-2020]}
+the heavyweight optimisations.
+
+Unfortunately, at the time of writing, we do not have a completely unoptimised
+version of the static analysis that we can use to compare, because the semantics
+of Marlowe have changed since we implemented such a version.
+
+Using the \texttt{perf} tool \cite{perf}, we have measured the execution time
+and the overhead of both the generation of the constraints and their solution
+by Z3 \cite{de2008z3}. And we have measured the peak RAM usage of the whole
+process by using GNU's \texttt{time} tool \cite{keppel1998gnu}.
+
+In all the contracts, the implementation that has the heavyweight optimisations
+performs much better and scales further. In the case of the \textit{auction}
+(Table~\ref{table1}) and \textit{crowdfunding} (Table~\ref{table2}) contracts,
+whose size grows exponentially, both approaches quickly overwhelm the resources
+available in the execution environment. On the other, in the case of \textit{rent}
+(Table~\ref{table3}) and \textit{coupon bond} (Table~\ref{table4}), which grow
+linearly, by using the lightweight version the problem becomes intractable much
+faster than with the heavyweight version.
+
+Another conclusion that can be derived from the experiments is that, in the
+version with lightweight optimisations, most of the processing time seems to be
+spent during the generation of the constraints, and solving is done relatively
+quickly by Z3; while the opposite happens for the version with heavyweight
+optimisations.
+
+These results suggest that SBV library is able to generate constraints in a way
+that they are handled efficiently by Z3, but the process itself can be costly
+if we are not careful. Nevertheless, the overall execution time required by Z3
+is also lower in the case of the version with heavyweight optimisations and it
+grows more slowly, which suggests the optimisations described in this paper
+affect both stages of the process.
+
 \section{Related work}

### 2020-05-20 papers/static_analysis2020/paper.tex
Add measurement section mention to introduction
Rewordings and fixes

@@
 \begin{abstract}
-SMT solvers can verify properties very efficiently, and they offer increasing flexibility on the ways that properties can be described.
+SMT solvers can verify properties very efficiently, and they offer increasing flexibility on the ways those properties can be described.
 However, the way in which properties are described can affect the computational cost of verifying them considerably.
@@
 \begin{itemize}
-\item Removing unnecessary parts from the analysis. If they do not affect
+\item \textbf{Removing unnecessary parts from the analysis}. If they do not affect
 the property that is being verified then we can just remove them from
 the implementation and reduce its complexity.
-\item Avoiding high level abstractions. High level abstractions aid reasoning
+\item \textbf{Avoiding high level abstractions}. High level abstractions aid reasoning
 and avoid errors, but also introduce complexity that is not necessarily
 needed.
-\item Reducing search space by normalising parameters. If there are several
+\item \textbf{Reducing search space by normalising parameters}. If there are several
 ways of representing some inputs, and the different representations
@@
 modifications are more fundamental approaches that require considerable
-changes to the fundamental structure of the implementation. We consider
-two main ideas:
+changes to the structure of the implementation. We consider two main ideas:
 \begin{itemize}
-\item Reducing search space by using normalised execution paths relevant
-to the property. Instead of using search space to model inputs, we
+\item \textbf{Reducing search space by using normalised execution paths relevant
+to the property}. Instead of using search space to model inputs, we
 use it to model possible executions, and we ensure that we only
@@
 executions with the same normal form.
-\item Minimizing the representation of inputs and outputs. We can benefit
+\item \textbf{Minimizing the representation of inputs and outputs}. We can benefit
 from encoding inputs and outputs as concisely as possible, and discarding
@@
 used to address potential problems that may arise from the implementation
-of the most heavyweight optimisations described in the previous section.
+of the heavyweight optimisations described in the previous section.
+In Section~\ref{sec:Measurements}, we present empirical results that
+show the effects of heavyweight optimisations on the execution time
+and memory usage of Marlowe's static analysis.
 
@@
 ask for input from the participants, and redistribute stored money
-and tokens back among participants. The decision about which and when
+and tokens back among participants. The decision about which and when of
 these actions are carried out is taken by the contract on the basis
-of the information that is available at each point in time. Participants,
-in turn, do not need to be a single person, but they may be represented
-by a \texttt{Role}, which in turn would be controlled by another contract.
+of the information that is available at each point in time. Participants
+may correspond to either individual public keys or to tokens (\texttt{Role}s).
+In turn, \texttt{Role}s may be controlled by separate independent contracts.
 
 The Marlowe language is written as a set of mutually recursive Haskell
-data types. It regulates interactions between a finite number of participants,
-established before the start of the contract.
+data types. Marlowe contracts regulate interactions between a finite number
+of participants, determined before the contract is deployed.
 
-The main data type is called \texttt{Contract}, and it represents
+The main data type in Marlowe is called \texttt{Contract}, and it represents
 the logic and actions that the contract allows or enforces. The outmost
-constructs of the contract represent the actions that will be enforced
-first, and as those constructs become enforced, the contract will
+constructs of the \texttt{Contract} represent the actions that will be enforced
+first and, as those constructs become enforced, the \texttt{Contract} will
 evolve into one of its continuations (sub-contracts), and the process
@@
 accounts.
-\item \texttt{If} -- immediately decides how the contract must continue,
-depending on a given condition that we call an \texttt{Observation}.
-The value of an \texttt{Observation} can depend on previous choices
-made by participants, on amounts of money and tokens remaining, or
-other factors.
-\item \texttt{Let} -- immediately stores a \texttt{Value} for use later.
-The value of a \texttt{Value} can depend on the information available
-to the contract at the time. For example, it may store the minimum
-slot of the slot interval of the current transaction, or the money
-available in a particular account at a given point in time, for use
-at a later time in the contract execution.
+\item \texttt{If} -- immediately decides how the contract must continue
+from between two possibilities and depending on a given Boolean condition
+that we call an \texttt{Observation}. The value of an \texttt{Observation}
+can depend on previous choices made by participants, on amounts of money
+and tokens remaining, or other factors.
+\item \texttt{Let} -- immediately stores a \texttt{Value} for later use.
+Expressions of type \texttt{Value} in Marlowe are evaluated to integer
+values and they depend on information available to the contract at the
+time of evaluation. For example, a \texttt{Let} construct can be used to
+store the minimum slot of the slot interval of the current transaction,
+or the amount of money available in a particular account at a given point
+in time, for use at a later stage of the contract execution.
 \item \texttt{When} -- waits for an external input. The \texttt{When} construct
@@
 As we mentioned in Section~\ref{subsec:Structure-of-Marlowe}, Marlowe
-contracts are passive: they are executed by transactions being submitted
-to the blockchain, with the on chain interpreter checking that the
-transaction complies with the contract under execution. These transactions
+contracts are passive: their code is executed as part of the validation of
+transactions that are submitted to the blockchain. These transactions
 need to be submitted by participants or their representatives (e.g.
-user wallets) and their execution is carried out atomically and deterministically.
+user wallets) and their validation is carried out atomically and
+deterministically.
 
@@
 and tokens), and a set of outputs (outgoing money and tokens). For
-a transaction to be valid, the transaction should have the same effect
-at any slot within that slot range (be deterministic). For example,
-if a transaction has a minimum slot number which is before a timeout,
-and a maximum slot which after a timeout then the transaction will
-fail with the error \texttt{AmbiguousSlotInterval}, since it would
+a transaction to be valid in Marlowe, the transaction must have the
+same effect for every slot within that slot range (be deterministic).
+For example, if a transaction has a minimum slot number which is before
+a timeout, and a maximum slot which after a timeout then the transaction
+will fail with the error \texttt{AmbiguousSlotInterval}, since it would
 not be deterministic.
 
-The key aspect of the Marlowe semantics is that it checks that a particular
+A key aspect of the Marlowe semantics is that it checks that a particular
 transaction is valid given the current state and contract. Because
 transactions are deterministic, there should be no reason why someone
-accidentally sends a transaction that is invalid, since it will only
-result in a cost to that participant. The type signature of the transaction
-validation function is:
+accidentally sends a transaction that is invalid for a given \texttt{State}
+and \texttt{Contract}, since it will only result in a cost to that participant.
+
+It is possible that, due to a race condition, a participant will send a
+transaction that no longer applies to a running \texttt{Contract} and
+\texttt{State}, but such a transaction would simply be ignored by the
+blockchain.
+
+The type signature of the transaction validation function is:
 
@@
 point). 
-\item \texttt{applyInput} - this function processes one input from the head
-of the input list. The topmost construct must be a \texttt{When} that
+\item \texttt{applyInput} - this function processes one single input.
+The topmost construct must be a \texttt{When} that
 is expecting that particular input and has not expired. 
@@
 \begin{itemize}
-\item Account identifiers include an account owner; an account owner is
+\item \textbf{Account identifiers include an account owner}; an account owner is
 a participant that will get the money or tokens remaining in an account
@@
 which has a timeout, this ensures that all contracts eventually expire
-and terminate. Together these properties ensure that no money or tokens
+and terminate. Together, these properties ensure that no money or tokens
 are locked in the contract forever.
-\item Classification of money and tokens into accounts separates concerns.
+\item \textbf{Classification of money and tokens into accounts} separates concerns.
 Marlowe will never spend more money or tokens than there are in an
@@
 to the original intention of the contract.
-\item Marlowe does not perform deposits or payments of negative amounts
-of money or tokens. It treats them as zero. However, at the same time,
+\item \textbf{No negative deposits or payments}. Marlowe treats negative
+amounts in deposits and payments as zero. However, at the same time,
 if there happens to be a request for deposit with a negative amount,
@@
 the expected execution of the contract is disturbed as little as possible.
-\item Marlowe contracts have an upper limit in the number of inputs that
-they accept throughout their life. This limit is implied by the maximum
+\item \textbf{Upper limit in the number of inputs} that a Marlowe contract
+can accept throughout its lifetime. This limit is implied by the maximum
 number of nested When constructs in the contract, since only one input
@@
 
-\section{Measurements}
+\section{Measurements \label{sec:Measurements} }
 

### 2020-05-20 papers/static_analysis2020/paper.tex
More rewording and fixes

@@
 \begin{itemize}
-\item \textbf{Account identifiers include an account owner}; an account owner is
-a participant that will get the money or tokens remaining in an account
-when a contract terminates. At the same time, the only construct that
-can pause the execution of a contract is the \texttt{When} construct,
-which has a timeout, this ensures that all contracts eventually expire
-and terminate. Together, these properties ensure that no money or tokens
-are locked in the contract forever.
 \item \textbf{Classification of money and tokens into accounts} separates concerns.
@@
 to the original intention of the contract.
+\item \textbf{Account identifiers include an account owner}. An account owner
+is a participant that will get the money or tokens remaining in an account
+when a contract terminates. At the same time, the only construct that
+can pause the execution of a contract is the \texttt{When} construct,
+which has a timeout, this ensures that all contracts eventually expire
+and terminate. Together, these properties ensure that no money or tokens
+are locked in the contract forever.
 \item \textbf{No negative deposits or payments}. Marlowe treats negative
@@
 if there happens to be a request for deposit with a negative amount,
-it continues gracefully, it will still wait for a null deposit to
-be made, and it will continue as if everything is correct, this way
+it continues gracefully: it will still wait for a null deposit to
+be made, and it will continue as if everything is correct; this way,
 the expected execution of the contract is disturbed as little as possible.
@@
 can accept throughout its lifetime. This limit is implied by the maximum
-number of nested When constructs in the contract, since only one input
-per When can be accepted. At the same time, transactions that have
+number of nested \texttt{When} constructs in the contract, since only one input
+per \texttt{When} can be accepted. At the same time, transactions that have
 no effect on the contract are considered invalid, thus there is also
 a limit in the maximum number of transactions a contract can accept
-throughout its life. This prevents DoS attacks, and makes static analysis
+throughout its life. This bound prevents DoS attacks, and it makes static analysis
 easier. We discuss this in more detail in Section~\ref{sec:Making-Marlowe-symbolic}.
@@
 \item With concrete parameters, as a normal Haskell function
-\item With symbolic variables, so that properties can be check for satisfiability
+\item With symbolic variables, so that properties can be checked for satisfiability
 using an SMT solver
-\item As part of QuickCheck properties, for testing randomly.
+\item As part of QuickCheck properties, for random testing.
 \end{itemize}
@@
 Unfortunately, there are a couple of issues with this approach that
-we review in Section~\ref{subsec:Considerations}.
+we review in Section~\ref{subsec:considerations}.
 
-\subsection{Considerations\label{subsec:Considerations}}
+\subsection{Additional considerations\label{subsec:considerations}}
 
-SBV does not currently support complex custom data types. It has some
-support for nonary data types, but not for data types with parameters.
-It does support the \texttt{Either} type, and the \texttt{Tuple} type.
-So it is possible to have conversions between custom data types and
-nested \texttt{Either} and \texttt{Tuple} type synonyms. We originally
-implemented this conversion using Template Haskell, since by using
-this technique the symbolic implementation can remain similar to the
-concrete one.
+SBV does not currently support complex custom data types. At the time
+of writing, SBV has some support for nonary data types, but not for
+data types with parameters.
+It does provide symbolic versions of the \texttt{Either} type and the \texttt{Tuple} type.
+Thus, it is possible to have an isomorphism between custom data types and
+nested \texttt{Either} and \texttt{Tuple} type-synonyms. Our original
+implementation automatically generates conversion functions between these
+by using Template Haskell. By using this technique the symbolic implementation
+can remain similar to the concrete one.
 
@@
 In the case of Marlowe, we have natural bounds that we can use because,
-given a concrete contract, we can calculate limits for:
+given a concrete contract, we can easily infer:
 \begin{itemize}
@@
 \end{itemize}
-From this data we can deduce an upper bound for:
+From this data, we can also deduce an upper bound for:
 \begin{itemize}
 \item The number of times that
-\texttt{computeTransaction}, \texttt{reduceContractStep}, and \texttt{applyInput} are
-called.
+\texttt{computeTransaction}, \texttt{reduceContractStep}, and \texttt{applyInput} may
+be called.
 \item The number of accounts that the contract will use, and an
-the number of elements in each of the associative maps
-that comprise the \texttt{State} of the contract.
+the number of elements there may be in each of the associative maps
+that comprise the \texttt{State} of the contract at any given point
+of the execution.
 \end{itemize}
@@
 
-However, there is one more problem. The output \texttt{Contract} returned
+However, there is one more problem: the output \texttt{Contract} returned
 by the function, which is wrapped inside the \texttt{TransactionOutput},
@@
 of returning the \texttt{TransactionOutput} directly, we take a continuation
-function that takes the concrete contract and the symbolic version
-of \texttt{TransactionOutput} without the contract.
+function that takes the concrete \texttt{Contract} and the symbolic version
+of \texttt{TransactionOutput} without the \texttt{Contract}.
 
@@
 trying to analyze, as well as making changes less likely to introduce
-errors than a more serious restructure.
+errors than a more serious restructuring.
 
-\subsubsection{Removing unnecessary parts from the analysis }
+\subsubsection{Removing unnecessary parts from the analysis.}
 
-This is probably the most straightforward optimization we can make.
+This is probably the most straightforward optimization to implement.
 When we use the same or similar code for both the analysis and the
@@
 \texttt{Close} construct to produce a failed payment
-(we have also proven this using Isabelle\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/isabelle/CloseSafe.thy}
-(last visited on 04/04/2020)}): it only pays whatever there is available.
+(we have also proven this\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/isabelle/CloseSafe.thy} using Isabelle \cite{nipkow2002isabelle})
+(last visited on 04/04/2020)}: it only pays whatever there is available.
 So we can safely remove it from the analysis.
 
-\subsubsection{Avoiding high level abstractions }
+\subsubsection{Avoiding high level abstractions.}
 
@@
 
-For example, in the case of Marlowe static analysis, we initially
+For example, in the case of Marlowe's static analysis, we initially
 implemented a symbolic associative map primitive with the only limitation
@@
 straightforward to realise using the symbolic implementation of list
-and tuple, both provided by SBV. However, this is very costly in a
-solver: searching a list symbolically requires constraints that check
+and tuple, both provided by SBV. However, this can be very costly:
+searching a list symbolically requires constraints that check
 the key of every element in the list up to the maximum capacity of
@@
 
-\subsubsection{Reducing search space by normalising parameters\label{subsec:Reducing-search-space-by-normalising-parameters}}
+\subsubsection{Reducing search space by normalising parameters.
+\label{subsec:Reducing-search-space-by-normalising-parameters}}
 
@@
 Using the automated proof assistant Isabelle \cite{nipkow2002isabelle},
-we have shown the stronger property that splitting a transaction in
-two does not modify the effect on a contract as long as the transaction
-does not produce an error\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/isabelle/SingleInputTransactions.thy}
+we have shown that, indeed, splitting transactions into transactions
+with single inputs and the same slot interval as the original transactions
+does not modify the effect of those transactions on a contract\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/isabelle/SingleInputTransactions.thy}
 (last visited on 29/04/2020)}.
@@
 ways of combining the inputs in transactions. We explain how we reduced
-the search space even further in the next section.
+the search space even further in Section~\ref{subsec:Heavyweight-modifications}.
 
-\subsection{Heavyweight modifications \label{subsec:Heavyweight-modifications}}
+\subsection{Heavyweight modifications. \label{subsec:Heavyweight-modifications}}
 
@@
 \subsubsection{Reducing search space by using normalised execution paths relevant
-to the property \label{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}}
+to the property. \label{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}}
 
@@
 for the trustworthiness of the analysis there are many more assumptions
-that need to be made and reasoned about. In Section~\ref{sec:Testing}
+that need to be made and reasoned about. In Section~\ref{sec:Testing},
 we explore ways of mitigating this problem.
@@
 the diagram above shows that a \texttt{When} belongs to the new transaction,
-because it will still be there after the previous transaction is executed,
-the observations in the \texttt{Notify} cases will correspond to the
+because the \texttt{When} will not be consumed by the previous transaction,
+\texttt{Observation}s in the \texttt{Notify} cases will correspond to the
 information there was in the \texttt{State} before the \texttt{When}
@@
 
-\subsubsection{Minimizing the representation of inputs and outputs }
+\subsubsection{Minimizing the representation of inputs and outputs.}
 
@@
 symbolic parameters. Originally, it would simply take a concrete \texttt{Contract}
-as input, and it would return a symbolic boolean that said whether
+as input, and it would return a symbolic Boolean that determined whether
 the \texttt{Contract} was valid or not. However, if a \texttt{Contract}
-turns out to be invalid, we also want to know why, so we later modified
+turns out to be invalid, we will also want to know why, so we later modified
 the property to give a counterexample that illustrated what went wrong.
@@
 whether the transaction input is a \texttt{Deposit}. We can also see
-which of the transactions does not have an effect, i.e: they produce
+which of the transactions do not have an effect, i.e: they produce
 \texttt{UselessTransaction} as we mentioned in Section~\ref{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}.
@@
 
-In order to avoid these problems, we use a mixture of automated proof
-assistants to verify our assumptions, and property based testing to
+In order to avoid these problems, we combine the use of automated proof
+assistants to verify our assumptions and the use of property based testing to
 test for consistency and equivalence.
@@
 whose size grows exponentially, both approaches quickly overwhelm the resources
-available in the execution environment. On the other, in the case of \textit{rent}
-(Table~\ref{table3}) and \textit{coupon bond} (Table~\ref{table4}), which grow
-linearly, by using the lightweight version the problem becomes intractable much
+available in the execution environment. On the other hand, in the case of \textit{rent}
+(Table~\ref{table3}) and \textit{coupon bond} (Table~\ref{table4}) contracts, which grow
+linearly, by using the lightweight version, the problem becomes intractable much
 faster than with the heavyweight version.
@@
 grows more slowly, which suggests the optimisations described in this paper
-affect both stages of the process.
+affect both stages of the process to some extent.
 
@@
 it relies on fewer and simpler features, it is compatible with more
-SMT solvers which, in turn, it means that it is less reliant on the
+SMT solvers which, in turn, means that it is less reliant on the
 correctness of a single SMT solver. If an SMT solver fails to give

### 2020-05-21 papers/static_analysis2020/paper.tex
Tables reduced and space saved.

@@
 \centering
-\scalebox{0.9}{
+\scalebox{0.8}{
 \begin{tabular}{l||lllll}
@@
 }
-\caption{Execution cost measurement of auction contract\label{table1}}
+%\caption{Execution cost measurement of auction contract\label{table1}}
 \end{table}
@@
 }
-\caption{Execution cost measurement of crowdfunding contract\label{table2}}
+%\caption{Execution cost measurement of crowdfunding contract\label{table2}}
 \end{table}
@@
 }
-\caption{Execution cost measurement of crowdfunding contract\label{table3}}
+%\caption{Execution cost measurement of crowdfunding contract\label{table3}}
 \end{table}
@@
 }
-\caption{Execution cost measurement of coupon contract\label{table4}}
+%\caption{Execution cost measurement of coupon contract\label{table4}}
+\vspace*{-0.6cm}
 \end{table}
@@
 the requirements of both processing time and memory for Marlowe's static
-analysis. In Tables~\ref{table1},~\ref{table2},~\ref{table3},~and~\ref{table4},
-we present the results of measuring the performance of static analysis on
-four example Marlowe contracts,
+analysis. %In Tables~\ref{table1},~\ref{table2},~\ref{table3},~and~\ref{table4},
+We present below the results of measuring the performance of static analysis on
+four example  contracts,
 before\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/src/Language/Marlowe/Analysis/FSSemantics.hs} [last visited 19-05-2020]}
@@
 
+\noindent
 Unfortunately, at the time of writing, we do not have a completely unoptimised

### 2020-05-21 papers/static_analysis2020/paper.tex
Address some reviews by @simonjohnthompson

@@
 \begin{abstract}
-SMT solvers can verify properties very efficiently, and they offer increasing flexibility on the ways those properties can be described.
-However, the way in which properties are described can affect the computational cost of verifying them considerably.
-In addition, it is quite hard to predict how different changes will affect the computational cost.
+SMT solvers can verify properties automatically and efficiently, and they offer increasing
+flexibility on the ways those properties can be described. However, the way in which
+properties are described can affect the computational cost of verifying them considerably. In addition, it is hard to predict how changes on the way of describing a property
+translate into changes to the computational costs.
 
@@
 
-Thanks to static analysis, we can automatically check beforehand whether
+Thanks to static analysis, we can automatically check beforehand whether any
 payments promised by a Marlowe contract can be fulfilled in every
-possible execution of the contract. For example, if a Marlowe contract
-has passed the static analysis, we will have a very high assurance
+possible execution of the contract. If a Marlowe contract has passed
+the static analysis, we will have a very high assurance
 that whenever the contract says we will be paid a certain amount of
-money, the contract will indeed have enough money to make such payment.
+money, any execution of the contract will indeed have enough money to
+make such payment.
 
@@
 This paper contributes a number of approaches that can be used when
-optimising static analysis, examples extracted from a case-study where
+optimising static analysis, with examples extracted from a case-study where
 we applied these approaches, and an overview of the techniques used
@@
 around 10 minutes using Z3. In the original implementation, we were
-not able to analyse the 19 megabyte contract because it ran out of
-memory after half an hour.
+not able to analyse the 19 megabyte contract because the analysis
+ran out of memory after half an hour.
 
@@
 needed.
-\item \textbf{Reducing search space by normalising parameters}. If there are several
+\item \textbf{Reducing the search space by normalising parameters}. If there are several
 ways of representing some inputs, and the different representations
@@
 \begin{itemize}
-\item \textbf{Reducing search space by using normalised execution paths relevant
-to the property}. Instead of using search space to model inputs, we
+\item \textbf{Reducing the search space by using normalised execution paths relevant
+to the property}. Instead of using the search space to model inputs, we
 use it to model possible executions, and we ensure that we only
 represent each possible group of equivalent executions once, i.e:
-executions with the same normal form.
+executions with the same normal form (or equivalence partitioning).
 \item \textbf{Minimizing the representation of inputs and outputs}. We can benefit
@@
 In the following sections, we introduce the semantics of Marlowe as
-a case study (Section~\ref{sec:Marlowe-intro}), we go through a
+a case study (Section~\ref{sec:Marlowe-intro}), we cover a
 general approach to static analysis (Section~\ref{sec:Making-Marlowe-symbolic}),
-then we explore both lightweight (Section~\ref{subsec:Lightweight-modifications})
+we then explore both lightweight (Section~\ref{subsec:Lightweight-modifications})
 and heavyweight (Section~\ref{subsec:Heavyweight-modifications})
@@
 
+The Marlowe language is realised as a set of mutually recursive Haskell
+data types. Marlowe contracts regulate interactions between a finite number
+of participants, determined before the contract is deployed.
+
 Marlowe contracts are able to receive payments, store money and tokens,
@@
 and tokens back among participants. The decision about which and when of
-these actions are carried out is taken by the contract on the basis
+these actions are carried out is determined by the contract on the basis
 of the information that is available at each point in time. Participants
@@
 In turn, \texttt{Role}s may be controlled by separate independent contracts.
-
-The Marlowe language is written as a set of mutually recursive Haskell
-data types. Marlowe contracts regulate interactions between a finite number
-of participants, determined before the contract is deployed.
 
@@
 can depend on previous choices made by participants, on amounts of money
-and tokens remaining, or other factors.
+and tokens stored in the contract, or other factors.
 \item \texttt{Let} -- immediately stores a \texttt{Value} for later use.
@@
 time of evaluation. For example, a \texttt{Let} construct can be used to
-store the minimum slot of the slot interval of the current transaction,
-or the amount of money available in a particular account at a given point
+store the amount of money available in a particular account at a given point
 in time, for use at a later stage of the contract execution.
 \item \texttt{When} -- waits for an external input. The \texttt{When} construct
-also specifies a timeout slot: after this slot has been reached, the
+also specifies a timeout slot\footnote{Slots are blockchain's proxy for time,
+they are added to the blockchain periodically.}: after this slot has been reached, the
 \texttt{When} construct expires, and no longer accepts any input.
-There are three types of input:
+There are three kinds of input:
 \begin{itemize}
 \item \texttt{Deposit} -- Waits for a participant to deposit an amount
-of money or tokens (specified as a \texttt{Value}) in the contract.
-The amount is added to one of the accounts of the contract.
+of money or tokens (specified as a \texttt{Value}) in an account of the contract.
 \item \texttt{Choice} -- Waits for a participant to make a choice. A choice
@@
 
-As we mentioned in Section~\ref{subsec:Structure-of-Marlowe}, Marlowe
+As we mentioned in the previous section, Marlowe
 contracts are passive: their code is executed as part of the validation of
@@
 Each transaction may include a list of inputs, a set of signatures,
-minimum and maximum slot numbers, a set of input UTxOs (incoming money
+a slot interval (in which the transaction can be executed), a set
+of input UTxOs (incoming money
 and tokens), and a set of outputs (outgoing money and tokens). For
@@
 and \texttt{Contract}, since it will only result in a cost to that participant.
-
-It is possible that, due to a race condition, a participant will send a
+However, it is still possible that, due to a race condition, a participant will send a
 transaction that no longer applies to a running \texttt{Contract} and
@@
 \begin{itemize}
-\item \texttt{reduceContractStep} - this function simplifies the topmost
+\item \texttt{reduceContractStep} - this function executes the topmost
 construct that does not require an input to be executed (i.e: anything
@@
 the \texttt{Close} contract, it only refunds one of the accounts at
-a time. 
+each invocation. 
 \item \texttt{reduceContractUntilQuiescent} - this function calls \texttt{reduceContractStep}
-repeatedly until it has no further effect (i.e. it reaches a fixed
-point). 
+repeatedly until it has no further effect,
 \item \texttt{applyInput} - this function processes one single input.
@@
 \item \textbf{Classification of money and tokens into accounts} separates concerns.
-Marlowe will never spend more money or tokens than there are in an
+A Marlowe will never spend more money or tokens than there are in an
 account, even if there is more available in the contract. But when
@@
 can be used to convert a concrete implementation of a Haskell function
-into a symbolic one by using the SBV library, and to use this symbolic
+into a symbolic one by using the SBV library \cite{erkok2019sbv}, and to use this symbolic
 implementation for static analysis. In particular, we explore this
@@
 SBV does not currently support complex custom data types. At the time
-of writing, SBV has some support for nonary data types, but not for
+of writing, SBV has some support for 0-ary data types, but not for
 data types with parameters.
@@
 implementation automatically generates conversion functions between these
-by using Template Haskell. By using this technique the symbolic implementation
-can remain similar to the concrete one.
+by using Template Haskell \cite{sheard2002template}. By using this technique the symbolic
+implementation can remain similar to the concrete one.
 
@@
 \end{itemize}
-If we keep the \texttt{Contract} parameter of the semantics concrete,
+If we keep the \texttt{Contract} parameter concrete,
 we can use it to bound the execution, because every call to \texttt{reduceContractStep}
 will either make no progress or remove one of the constructs, with
-the exception of \texttt{Close}. But for \texttt{Close} we know we
-just have to call it as many times as the maximum number of accounts
-there can possibly be, since every time we call it it will refund
-one of the accounts.
+the exception of \texttt{Close}. But for any particular \texttt{Contract},
+we know we only need to call \texttt{reduceContractStep} at most as many times as
+the maximum number of accounts there can possibly be, since, every time we call
+it, it will refund one of the accounts.
 
@@
 
-However, there is one more problem: the output \texttt{Contract} returned
+There is one more problem: the output \texttt{Contract} returned
 by the function, which is wrapped inside the \texttt{TransactionOutput},
@@
 
-\section{Making static analysis efficient\label{sec:Making-static-analysis-efficient}}
-
-In this section, we explain the techniques that we mentioned in Section~\ref{sec:Introduction}
-in more detail, and we illustrate those techniques with examples from
-the static analysis implementation of Marlowe.
+\section{Making static analysis more efficient\label{sec:Making-static-analysis-efficient}}
+
+In this section, we explain, in more detail, the techniques that we mentioned
+in the introduction, and we illustrate those techniques with
+examples from the static analysis implementation of Marlowe.
 
@@
 are localised, and they do not fundamentally affect the overall structure
-of the implementation. This means that the implementation can remain
-intuitive and close to the original semantics and model that we are
+of the implementation. This means that the implementation can remain close
+to the original semantics and model that we are
 trying to analyze, as well as making changes less likely to introduce
@@
 High level libraries like SBV, and even standards like SMTLib, support
-the use and construction of considerably high level abstractions,
+the use and construction of high level abstractions,
 e.g: custom data-types, list, sets\dots{} Unfortunately, even though
@@
 For example, in the case of Marlowe's static analysis, we initially
-implemented a symbolic associative map primitive with the only limitation
+implemented a \emph{symbolic} associative map primitive with the only limitation
 that it needed a concrete bound in the number of elements. This is
@@
 However, in Marlowe we know the values of all the keys that we are
-going to use, because the contract is concrete, and only \texttt{Account},
+going to use in maps, because the contract is concrete, and only \texttt{Account},
 \texttt{Choice}, and \texttt{Let} identifiers that are mentioned in
@@
 
-\subsubsection{Reducing search space by normalising parameters.
+\subsubsection{Reducing the search space by normalising parameters.
 \label{subsec:Reducing-search-space-by-normalising-parameters}}
@@
 For example, in the case of Marlowe's static analysis,
-Marlowe allows several inputs to be combined in one single transaction.
+Marlowe allows several inputs to be combined into a single transaction.
 This functionality is important because each transaction requires
@@
 This optimization reduces the search space considerably, but transactions
-may still have no inputs, so it still leaves room for many possible
-ways of combining the inputs in transactions. We explain how we reduced
+may still have no inputs, so there are still many ways of distributing
+the inputs in transactiongs. We explain how we reduced
 the search space even further in Section~\ref{subsec:Heavyweight-modifications}.
@@
 When optimising, if our solution is a local minimum, small changes
-to the parameters will not grant any improvement to the value function.
+to the parameters will not grant any improvement to the result.
 For that reason, in this section, we explore ways of optimizing that
@@
 
-\subsubsection{Reducing search space by using normalised execution paths relevant
+\subsubsection{Reducing the search space by using normalised execution paths relevant
 to the property. \label{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}}
@@
 One approach we can use that helps improve static analysis efficiency
-is to implement the property thinking of the property we want to test
-instead of trying to make it as close as possible to the full implementation
+is to focus on the property we want to test instead of trying to make
+it as close as possible to the full implementation
 of the system. We do not even need to think about the representation
@@
 where each subtree has a \texttt{When} construct as its root, with
-the exception of the subtree that has the same root as the original
-contract tree.
+the exception of the subtree at the root of the original tree.
 
@@
 some paths which do not require as many transactions/subtrees as other
-paths, because they go across fewer \texttt{When} constructs before
+paths, because they traverse fewer \texttt{When} constructs before
 finding a \texttt{Close} construct.
 
-We can assign one transaction to each path of execution before a \texttt{When},
-because if the maximum slot number of the transaction is lower than
-the slot number in the timeout and the transaction has no inputs,
-then the execution will stop before the \texttt{When}, so we need
-to account for that possibility. In other words, if two parts of the
-execution path may have different slot numbers when we go through
-them, we need to consider them as separate transactions, because they
-may have payments that depend on those values.
-
-On the other hand, if the minimum slot number of the transaction is greater
+We can assign one transaction to each segment of execution
+before starting at a \texttt{When} or the root of the tree,
+because if the maximum slot number of the
+transaction is lower than the slot number in the timeout and the
+transaction has no inputs, then the execution will stop before the
+\texttt{When}, so we need to account for that possibility. In other
+words, if two parts of the execution path may have different slot
+numbers when we go through them, we need to consider them as separate
+transactions, because they may have payments that depend on those values.
+
+On the other hand, if the minimum slot number of a transaction is greater
 or equal than the slot number in the timeout, then the first transaction
@@
 
-One subtle detail that we realised during testing is that, even though
+One detail that we realised during testing is that, even though
 the diagram above shows that a \texttt{When} belongs to the new transaction,
@@
 for Marlowe, we did not pay any attention to the inputs and outputs.
-The first version we wrote of the SBV property would not take any
-symbolic parameters. Originally, it would simply take a concrete \texttt{Contract}
+The first version would simply take a concrete \texttt{Contract}
 as input, and it would return a symbolic Boolean that determined whether
@@
 
-The solution to this increase in execution time was to pass as input
-a fixed list of transactions, each of which being represented by a
-tuple with four symbolic integers:
+The solution to this increase in execution time was to pass as input, to
+the static analysis implementation, a fixed list, with each element corresponding
+to a transaction and consisting of a tuple with four symbolic integers:
 \begin{enumerate}
@@
 assistants to verify our assumptions and the use of property based testing to
-test for consistency and equivalence.
+test for consistency and equivalence (as we describe in
+Section~\ref{subsec:Testing-for-consistency}).
 
@@
 
-If our static analysis is not replicating all the functionality of
+If our static analysis does not replicate all the functionality of
 the semantics, we can use potential discrepancies as an opportunity
@@
 we cannot find any counterexamples then the test passes, but if we
-find one we test it on the semantics and see whether the counterexample
+find one, we can test it on the semantics and see whether the counterexample
 indeed produces warnings in the semantics too, if it does not we have
-found a problem either on the static analysis or the semantics.
+found a problem in either the static analysis or the semantics.
 
 A limitation of this approach is that it only tests for false positives,
-and it does not prevent false negatives, those can be covered by testing
+and it does not prevent false negatives, but those can be covered by testing
 for equivalence (see Section~\ref{subsec:Testing-for-equivalence}).
@@
 In addition, we can add assertions to the process. In the case of
-Marlowe, if the counterexample breaks one of the assertions or is
+Marlowe, if the counterexample causes errors during the execution or is
 formed incorrectly, it would also mean that there is a problem with
@@
 is very different from the semantics and one inefficient implementation
-that is much closer to the semantics. We can do this as shown in Figure~\ref{fig:Property-based-testing-for-equivalence}.
+that is much closer to the semantics, and we can compare the results of the
+two implementations. We can do this as shown in Figure~\ref{fig:Property-based-testing-for-equivalence}.
 We generate random contracts, we feed them to both static analysis
@@
 
-In our experiments, the heavy weight optimisations considerably reduced
+In our experiments, the heavyweight optimisations considerably reduced
 the requirements of both processing time and memory for Marlowe's static
@@
 version of the static analysis that we can use to compare, because the semantics
-of Marlowe have changed since we implemented such a version.
-
-Using the \texttt{perf} tool \cite{perf}, we have measured the execution time
+of Marlowe have changed since we implemented that version.
+
+Using the \texttt{perf} tool \cite{perf}, we have measured the execution
+time\footnote{The experiments were run on a laptop computer with a
+i9-9900K (3.6GHz) processor and two modules of 16GB of SODIMM DDR4 RAM at 2400MHz.}
 and the overhead of both the generation of the constraints and their solution
@@
 performs much better and scales further. In the case of the \textit{auction}
-(Table~\ref{table1}) and \textit{crowdfunding} (Table~\ref{table2}) contracts,
-whose size grows exponentially, both approaches quickly overwhelm the resources
-available in the execution environment. On the other hand, in the case of \textit{rent}
-(Table~\ref{table3}) and \textit{coupon bond} (Table~\ref{table4}) contracts, which grow
-linearly, by using the lightweight version, the problem becomes intractable much
-faster than with the heavyweight version.
+and \textit{crowdfunding} contracts, whose size grows exponentially, both
+approaches quickly overwhelm the resources available in the execution environment.
+On the other hand, in the case of \textit{rent} and \textit{coupon bond} contracts,
+which grow linearly, by using the lightweight version, the problem becomes 
+intractable much faster than with the heavyweight version.
 
@@
 version with lightweight optimisations, most of the processing time seems to be
-spent during the generation of the constraints, and solving is done relatively
+spent generating of the constraints, and solving is done relatively
 quickly by Z3; while the opposite happens for the version with heavyweight
@@
 
-These results suggest that SBV library is able to generate constraints in a way
+These results suggest that the SBV library is able to generate constraints in a way
 that they are handled efficiently by Z3, but the process itself can be costly

### 2020-05-21 papers/static_analysis2020/paper.tex
Address more reviews by @simonjohnthompson

@@
 
-State of the art libraries like SBV \cite{erkok2019sbv} allow straightforward
-static analysis of arbitrary Haskell functions, with very few restrictions
-on how those functions are implemented. However, static analysis can be very
-computationally expensive, because it essentially needs to check every
-possible execution of the contract.
+Marlowe's static analysis relies on SMT solvers, which are able to check whether a
+set of constraints is satisfiable very efficiently. Thanks to state of the art
+libraries like SBV \cite{erkok2019sbv}, we can describe those constraints in a
+high-level language. In the case of SBV, we can directly write properties as
+Haskell functions, with very few restrictions on how those functions are
+implemented; SBV will automatically translate those functions into a language
+that SMT solvers can understand (namely, SMTLib format \cite{barrett2010smt}).
+
+Nevertheless, static analysis can become very computationally expensive if
+implemented naïvely, because it needs to consider every possible value of the
+property we want to check (in our case the main property we want to check
+is whether a Marlowe contract will have enough money to make all the
+payments described by the contract). Indeed, high-level abstractions
+are often a tradeoff with efficiency.
 
@@
 optimising static analysis, with examples extracted from a case-study where
-we applied these approaches, and an overview of the techniques used
-to test the correctness of the optimisations.
+we applied these approaches, an approach to ensure correctness
+of the optimisations that combines the usage of property-based testing
+and verification using automated theorem provers, and empirical data
+that measures the effect of some of the optimisations on the implementation
+of our case-study.
 
@@
 modifications are more fundamental approaches that require considerable
-changes to the structure of the implementation. We consider two main ideas:
+changes to the structure of the implementation. But these optimisations can also
+translate in important reductions to execution time and memory usage, as
+shown by the experiments we report in Section~\ref{sec:Measurements}.
+We consider two main ideas:
 \begin{itemize}
@@
 Each transaction may include a list of inputs, a set of signatures,
-a slot interval (in which the transaction can be executed), a set
-of input UTxOs (incoming money
-and tokens), and a set of outputs (outgoing money and tokens). For
+a slot interval\footnote{We use a slot interval because it is very difficult to
+know in practice in which slot the transaction is going to be included,
+and we need transactions to be deterministic, so we cannot use individual
+slot numbers.}, a set of input UTxOs (incoming money and tokens), and a
+set of outputs (outgoing money and tokens). For
 a transaction to be valid in Marlowe, the transaction must have the
@@
 deal with arbitrarily sized computations. In other words, if termination
-of a symbolic function is bounded by a parameter that can be arbitrarily
-large, SMT solvers will often fail to terminate when trying to validate
+of a symbolic function is not bounded, or is only bounded by a symbolic
+parameter, SMT solvers will often fail to terminate when trying to validate
 a property, since doing so often requires a proof by induction. We
@@
 \texttt{Close} construct to produce a failed payment
-(we have also proven this\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/isabelle/CloseSafe.thy} using Isabelle \cite{nipkow2002isabelle})
-(last visited on 04/04/2020)}: it only pays whatever there is available.
+(we have also proven this\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/isabelle/CloseSafe.thy} (last visited on 04/04/2020)} using Isabelle \cite{nipkow2002isabelle}): it only pays whatever there is available.
 So we can safely remove it from the analysis.
@@
 \end{enumerate}
-If we do not need a transaction, we just set all four numbers to $-1$.
+If we need fewer transactions, we will just add dummy transactions with all four numbers
+set to $-1$ at the end of the list.
 
@@
 two implementations. We can do this as shown in Figure~\ref{fig:Property-based-testing-for-equivalence}.
-We generate random contracts, we feed them to both static analysis
-implementations, and we compare the results. If the results are the
-same then test passes, if they are different then one of the implementations
-is wrong.
+We generate random contracts using a custom QuickCheck \cite{quickcheck} generator,
+we feed them to both static analysis implementations, and we compare the results.
+If the results are the same then test passes, if they are different then
+one of the implementations is wrong.
 
@@
 version of the static analysis that we can use to compare, because the semantics
-of Marlowe have changed since we implemented that version.
+of Marlowe have changed since we implemented that version. However, our impression
+from manual testing is that the impact of lightweight optimisations was much more modest than the impact of heavyweight optimisations.
 

### 2020-05-22 papers/static_analysis2020/paper.tex
Improve explanation of transaction and segment mapping

@@
 
-On the other hand, if the minimum slot number of a transaction is greater
-or equal than the slot number in the timeout, then the first transaction
-will execute past the \texttt{When} and into the timeout branch, and
-we will not need the second transaction. In the same way, if we include
+On the other hand, this segmentation is not general enough, because there is
+an edge case. A transaction may execute past a \texttt{When} if the minimum
+slot number in the transaction is greater or equal than the timeout of the
+\texttt{When}. If the transaction expires the \texttt{When} we can no longer
+separate the execution before the \texttt{When} and after the \texttt{When}
+into two different transactions, because we know that the timeout branch of the
+\texttt{When} will be executed as part of the first transaction, and thus
+it will use the slot interval of the first
+transaction (same used for the first segment).
+
+This is not only true for the timeout branch, if we include
 an input that is allowed by the \texttt{When} into the first transaction,
 then the first transaction will continue into the corresponding \texttt{When}
-branch, and we will not need the second transaction.
+branch, and we will not use a second transaction.
 
@@
 \texttt{When} in a row, or where it expires one \texttt{When} and provides
-the right input for later \texttt{When}. And we cannot not constrain
+the right input for later \texttt{When}. And we cannot just not constrain
 the maximum slot number of the first transaction because then we will
-not know whether the first transaction goes beyond the \texttt{When}
-or not.
+not know to which segments each transaction applies.
 

### 2020-05-26 src/Language/Marlowe/Examples/CouponBond.hs
Add CouponBond generator to Examples

@@ -1,0 +1,25 @@
+{-# LANGUAGE OverloadedStrings #-}
+module Language.Marlowe.Examples.CouponBond where
+
+import Language.Marlowe.ACTUS.ActusContracts
+import Language.Marlowe.ACTUS
+
+import Data.Time
+import Data.Time.Clock
+import Data.Time.Clock.POSIX
+import Data.Time.Clock.System
+
+cb2 nm ied md notional rate = (emptyContractConfig ied)
+    { maturityDate = Just md
+    , notional = notional
+    , nominalInterestRate = rate
+    , interestPaymentCycle = Just $ Cycle nm Month LongLastStub
+    , premiumDiscountAtIED = 0 }
+
+couponBond numMonths =
+  let td = fromGregorian 2020 5 14 in
+  let tcb = cb2 numMonths td (addGregorianMonthsClip numMonths td) 1000 0.12 in
+  genPrincialAtMaturnityContract "inverstor" "issuer" tcb
+
+contract = couponBond 5
+

### 2020-05-27 papers/static_analysis2020/paper.tex
Fix typo
Fix formatting in table
Add CouponBond generator to Examples

@@
 \item \textbf{Classification of money and tokens into accounts} separates concerns.
-A Marlowe will never spend more money or tokens than there are in an
+A Marlowe contract will never spend more money or tokens than there are in an
 account, even if there is more available in the contract. But when
@@
 be called.
-\item The number of accounts that the contract will use, and an
+\item The number of accounts that the contract will use, and
 the number of elements there may be in each of the associative maps
@@
 \begin{tabular}{l||lllll}
-\multicolumn{6}{l}{\textbf{Coupon bond contract - ACTUS.hs}} \\ \hline \hline
+\multicolumn{6}{l}{\textbf{Coupon bond contract - \texttt{CouponBond.hs}}} \\ \hline \hline
 \multicolumn{6}{c}{\textbf{Lightweight optimisations}} \\ \hline \hline

### 2020-05-27 papers/static_analysis2020/paper.tex
Reduce to 16.5 pages

@@
 SMT solvers can verify properties automatically and efficiently, and they offer increasing
-flexibility on the ways those properties can be described. However, the way in which
-properties are described can affect the computational cost of verifying them considerably. In addition, it is hard to predict how changes on the way of describing a property
-translate into changes to the computational costs.
+flexibility on the ways those properties can be described. But it is hard to predict how those ways of describing the properties affect the computational cost of verifying them.
 
-In this paper, we discuss the lessons we have learned while implementing and optimising the static analysis functionality for Marlowe, a domain specific language for describing self-enforcing financial smart-contracts that can be deployed on a blockchain.
+In this paper, we discuss what we learned while implementing and optimising the static analysis for Marlowe, a domain specific language for self-enforcing financial smart-contracts that can be deployed on a blockchain.
 \end{abstract}
@@
 the static analysis, we will have a very high assurance
-that whenever the contract says we will be paid a certain amount of
-money, any execution of the contract will indeed have enough money to
-make such payment.
+that whenever the contract says it will make a payment, the contract will
+indeed have enough money available.
 
-Marlowe's static analysis relies on SMT solvers, which are able to check whether a
-set of constraints is satisfiable very efficiently. Thanks to state of the art
+Marlowe's static analysis relies on SMT solvers, which are able to check efficiently
+whether a set of constraints is satisfiable. Thanks to state of the art
 libraries like SBV \cite{erkok2019sbv}, we can describe those constraints in a
-high-level language. In the case of SBV, we can directly write properties as
-Haskell functions, with very few restrictions on how those functions are
-implemented; SBV will automatically translate those functions into a language
-that SMT solvers can understand (namely, SMTLib format \cite{barrett2010smt}).
+high level language. In the case of SBV, we can write properties as
+Haskell functions, with few restrictions on how those functions are
+implemented; SBV automatically translates those functions to SMTLib format \cite{barrett2010smt}, a language many SMT solvers understand.
 
 Nevertheless, static analysis can become very computationally expensive if
-implemented naïvely, because it needs to consider every possible value of the
-property we want to check (in our case the main property we want to check
-is whether a Marlowe contract will have enough money to make all the
-payments described by the contract). Indeed, high-level abstractions
-are often a tradeoff with efficiency.
+implemented naïvely, because it is an NP-complete problem in the general case.
+In our case the main property is whether a contract will have enough money for all payments. Indeed, high level abstractions are often a tradeoff with efficiency.
 
-For example, in the case of Marlowe, contracts are typically interactive
-and must deal with inputs arriving from different users in different
-orders, and sometimes not arriving at all. In addition, Marlowe allows
-several inputs to be grouped in a single transaction, and different
-decisions may be taken depending on the content of those inputs, or
-in the combinations of contents of those inputs. All these alternatives
-multiply the search space of static analysis and can easily compose
-exponentially and thus make the whole approach unusable in practice
-even for moderately sized contracts.
+% For example, in the case of Marlowe, contracts are typically interactive
+% and must deal with inputs arriving from different users in different
+% orders, and sometimes not arriving at all. In addition, Marlowe allows
+% several inputs to be grouped in a single transaction, and different
+% decisions may be taken depending on the content of those inputs, or
+% in the combinations of contents of those inputs. All these alternatives
+% multiply the search space of static analysis and can easily compose
+% exponentially and thus make the whole approach unusable in practice
+% even for moderately sized contracts.
 
@@
 that measures the effect of some of the optimisations on the implementation
-of our case-study.
+of our case-study: the optimisation of Marlowe's static analysis.
 
-The case study presented is that of the optimisation of Marlowe's
-static analysis. During this work, the techniques applied allowed
-us to reduce the analysis time of an implementation that followed
-the semantics closely and took a couple of minutes to analyse contracts
-of a few kilobytes, to one where the same contract would take less
-than a second, and where a four-person crowdfunding contract that
-when fully expanded occupies about 19 megabytes, can be analysed in
-around 10 minutes using Z3. In the original implementation, we were
-not able to analyse the 19 megabyte contract because the analysis
-ran out of memory after half an hour.
+The techniques described helped us reduce the analysis time of an implementation
+that followed the semantics closely and took a couple of minutes to analyse contracts
+of a few kilobytes, to one where the same contract takes less
+than a second to analyse, and where a four-person crowdfunding contract that
+fully expanded occupies about 19 megabytes, can be analysed in
+around 10 minutes.
 
-There are many things that can be done to improve the efficiency of
-static analysis. We have classified them as lightweight and heavyweight.
+We have classified optimisation techniques as lightweight and heavyweight.
 
@@
 \item \textbf{Removing unnecessary parts from the analysis}. If they do not affect
-the property that is being verified then we can just remove them from
-the implementation and reduce its complexity.
+the property that is being verified then we can just remove them.
 \item \textbf{Avoiding high level abstractions}. High level abstractions aid reasoning
-and avoid errors, but also introduce complexity that is not necessarily
-needed.
+and avoid errors, but also introduce complexity that may not be necessary.
 \item \textbf{Reducing the search space by normalising parameters}. If there are several
 ways of representing some inputs, and the different representations
-have no impact on the analysis, we can remove all but one when
-implementing static analysis.
+have no impact on the analysis, we can remove all but one for analysis.
 \end{itemize}
@@
 to the property}. Instead of using the search space to model inputs, we
-use it to model possible executions, and we ensure that we only
-represent each possible group of equivalent executions once, i.e:
-executions with the same normal form (or equivalence partitioning).
-\item \textbf{Minimizing the representation of inputs and outputs}. We can benefit
-from encoding inputs and outputs as concisely as possible, and discarding
-information that we can infer in other ways.
+use it to model possible executions, and we only
+represent each equivalence class of executions once, i.e:
+we represent them in a normal form.
+\item \textbf{Minimizing the representation of inputs and outputs}. We encode inputs
+and outputs as concisely as possible, discarding inferable information.
 \end{itemize}
@@
 examples of how they apply to Marlowe static analysis. In Section~\ref{sec:Testing},
-we briefly discuss some ways in which property based testing can be
-used to address potential problems that may arise from the implementation
-of the heavyweight optimisations described in the previous section.
-In Section~\ref{sec:Measurements}, we present empirical results that
-show the effects of heavyweight optimisations on the execution time
-and memory usage of Marlowe's static analysis.
+we illustrate the use of property based testing on the implementation with
+heavyweight optimisations. Finally, we present empirical results that
+show the effect of heavyweight optimisations on the execution time
+and memory usage of Marlowe's static analysis (Section~\ref{sec:Measurements}).
 
-\section{Marlowe and how its design helps static analysis and correctness\label{sec:Marlowe-intro} }
+\section{Marlowe design and semantics\label{sec:Marlowe-intro} }
 
-Firstly, we will briefly introduce the semantics of Marlowe, the guarantees
+Firstly, we introduce the semantics of Marlowe, the guarantees
 that are offered implicitly by the semantics, and how the design choices
 facilitate static analysis and make it decidable. A more detailed
-explanation of Marlowe can be found in \cite{marlowe}.
+explanation can be found in \cite{marlowe}.
 
@@
 Marlowe contracts are able to receive payments, store money and tokens,
-ask for input from the participants, and redistribute stored money
-and tokens back among participants. The decision about which and when of
-these actions are carried out is determined by the contract on the basis
-of the information that is available at each point in time. Participants
-may correspond to either individual public keys or to tokens (\texttt{Role}s).
-In turn, \texttt{Role}s may be controlled by separate independent contracts.
+ask for input from its participants, and redistribute stored money
+and tokens among participants. A contract determines when and which of
+these actions may be carried out. Participants may correspond to either
+individual public keys or to tokens (\texttt{Role}s). In turn, \texttt{Role}s
+may be controlled by other contracts.
 
@@
 \item \texttt{If} -- immediately decides how the contract must continue
-from between two possibilities and depending on a given Boolean condition
-that we call an \texttt{Observation}. The value of an \texttt{Observation}
-can depend on previous choices made by participants, on amounts of money
-and tokens stored in the contract, or other factors.
+from between two possibilities, depending on a given Boolean condition
+that we call an \texttt{Observation}. An \texttt{Observation}
+may depend on previous choices by participants, on amounts of money
+and tokens in the contract, or other factors.
 \item \texttt{Let} -- immediately stores a \texttt{Value} for later use.
@@
 values and they depend on information available to the contract at the
-time of evaluation. For example, a \texttt{Let} construct can be used to
-store the amount of money available in a particular account at a given point
-in time, for use at a later stage of the contract execution.
+time of evaluation. For example, a \texttt{Let} construct record
+the amount of money in an account at a point in time.
 \item \texttt{When} -- waits for an external input. The \texttt{When} construct
@@
 Because contracts are reactive (they cannot initiate transactions),
-it is necessary for an external actor to \texttt{Notify} the contract
-that the \texttt{Observation} has become true. However, this notification
-does not have to be carried out by a participant, anybody can do it.
+it is necessary for an external actor (not necessarily a participant)
+to \texttt{Notify} the contract.
 \end{itemize}
@@
 contracts are passive: their code is executed as part of the validation of
-transactions that are submitted to the blockchain. These transactions
+transactions that are submitted to the blockchain. Transactions
 need to be submitted by participants or their representatives (e.g.
-user wallets) and their validation is carried out atomically and
-deterministically.
+user wallets) and validation is atomic and deterministic.
 
 Each transaction may include a list of inputs, a set of signatures,
-a slot interval\footnote{We use a slot interval because it is very difficult to
-know in practice in which slot the transaction is going to be included,
-and we need transactions to be deterministic, so we cannot use individual
-slot numbers.}, a set of input UTxOs (incoming money and tokens), and a
-set of outputs (outgoing money and tokens). For
-a transaction to be valid in Marlowe, the transaction must have the
-same effect for every slot within that slot range (be deterministic).
-For example, if a transaction has a minimum slot number which is before
-a timeout, and a maximum slot which after a timeout then the transaction
-will fail with the error \texttt{AmbiguousSlotInterval}, since it would
-not be deterministic.
+a slot interval, a set of input UTxOs (incoming money and tokens), and a
+set of outputs (outgoing money and tokens). We use a slot interval because
+it is very difficult to know the exact slot in which the transaction will
+be included in practice. For a transaction to be valid in Marlowe, the
+transaction must have the same effect for every slot within that slot
+interval (be deterministic). For example, if a transaction has a minimum
+slot number lower than timeout, and a maximum slot that is greater, then the transaction will fail with the error \texttt{AmbiguousSlotInterval}.
 
@@
 A Marlowe contract will never spend more money or tokens than there are in an
-account, even if there is more available in the contract. But when
-required to pay more than is in an account, it will pay as much as
-is available in the account, in order to remain as close as possible
+account, even if there is more available in the contract. But a payment
+for more than there is will not fail, it will pay as much as
+is available, in order to remain as close as possible
 to the original intention of the contract.
@@
 \item \textbf{No negative deposits or payments}. Marlowe treats negative
-amounts in deposits and payments as zero. However, at the same time,
-if there happens to be a request for deposit with a negative amount,
-it continues gracefully: it will still wait for a null deposit to
-be made, and it will continue as if everything is correct; this way,
-the expected execution of the contract is disturbed as little as possible.
+amounts in deposits and payments as zero. At the same time,
+if there happens to be a request for a deposit with a negative amount,
+the contract will still wait for a null deposit to
+be made, and it will continue as if everything is correct. This way execution
+of the contract is disturbed as little as possible.
 \item \textbf{Upper limit in the number of inputs} that a Marlowe contract
-can accept throughout its lifetime. This limit is implied by the maximum
-number of nested \texttt{When} constructs in the contract, since only one input
-per \texttt{When} can be accepted. At the same time, transactions that have
-no effect on the contract are considered invalid, thus there is also
-a limit in the maximum number of transactions a contract can accept
-throughout its life. This bound prevents DoS attacks, and it makes static analysis
-easier. We discuss this in more detail in Section~\ref{sec:Making-Marlowe-symbolic}.
+can accept throughout its lifetime. This limit is implied by the path with
+maximum number of nested \texttt{When} constructs in the contract, since only
+one input per \texttt{When} can be accepted. At the same time, transactions
+with no effect on the contract are invalid, thus there is
+a limit on the maximum number of transactions a contract can accept
+throughout its life too. This bound prevents DoS attacks, and it makes static analysis
+easier. We discuss further in Section~\ref{sec:Making-Marlowe-symbolic}.
 \end{itemize}
@@
 implementation for static analysis. In particular, we explore this
-technique in the context of the Marlowe semantics.
+technique in the context of the Marlowe.
 
@@
 \begin{itemize}
-\item With concrete parameters, as a normal Haskell function
-\item With symbolic variables, so that properties can be checked for satisfiability
-using an SMT solver
+\item With concrete parameters, as a normal Haskell function.
+\item With symbolic parameters, so that properties can be checked for satisfiability
+using an SMT solver.
 \item As part of QuickCheck properties, for random testing.
@@
 Parameters that can be used symbolically are wrapped in a monad called
-\texttt{SBV}. Functions that depend on symbolic parameters also return
-values wrapped in the \texttt{SBV} monad.
-
-The idea is therefore to convert our implementation to use values
-wrapped in the \texttt{SBV} monad, by replacing all operations on
-those values with symbolic functions. In order to simplify this work,
-\texttt{SBV} provides symbolic versions for many of the most common
-built-in functions that are either overloaded or have similar names.
+\texttt{SBV}. Values that depend on symbolic values must also be wrapped in the \texttt{SBV} monad.
 
@@
 computeTransaction :: SBV TransactionInput -> SBV State
-                   -> SBV Contract
-                   -> SBV TransactionOutput
+                   -> SBV Contract -> SBV TransactionOutput
 \end{verbatim}
 
-We can then define a function that takes a list of transactions instead,
-and we would just need to write a property that says that the result
-of executing the list of transactions for a given contract does not
-have any warnings (the property we are trying to verify). Then we
-can pass the property to SBV and it will tell us which transaction
-lists break the property.
-
-Unfortunately, there are a couple of issues with this approach that
-we review in Section~\ref{subsec:considerations}.
+We just need a function \texttt{playTrace} that takes a list of
+transactions and calls \texttt{computeTransaction} for each. 
+We can then write our property to state that the output of
+\texttt{playTrace} does not have any failed payments (or other warnings). We
+can ask SBV to find an input transaction list that breaks the property.
+However, there are a couple of issues with this approach,
+we review them in Section~\ref{subsec:considerations}.
 
@@
 
-SBV does not currently support complex custom data types. At the time
-of writing, SBV has some support for 0-ary data types, but not for
-data types with parameters.
-It does provide symbolic versions of the \texttt{Either} type and the \texttt{Tuple} type.
-Thus, it is possible to have an isomorphism between custom data types and
-nested \texttt{Either} and \texttt{Tuple} type-synonyms. Our original
-implementation automatically generates conversion functions between these
-by using Template Haskell \cite{sheard2002template}. By using this technique the symbolic
-implementation can remain similar to the concrete one.
-
-For example, the following data structure:
+At the time of writing, SBV does not fully support complex custom data types,
+but it provides symbolic versions for \texttt{Either} and \texttt{Tuple} types.
+Our original implementation of Marlowe's static analysis overcomes this limitation
+by generating conversion functions using using Template Haskell
+\cite{sheard2002template}. This allows static analysis to remain similar to the semantics. For example, the following data structure:
 
@@
 But this approach cannot address recursive datatypes, let alone mutually
-recursive datatypes. And the \texttt{Contract} type is mutually recursive.
+recursive datatypes. And the \texttt{Contract} definition uses mutual recursion.
 
-Even if we could represent mutually recursive datatypes, it would
-not be feasible to directly translate the semantics, because a more
-general limitation of SMT solvers is that they cannot, in general,
-deal with arbitrarily sized computations. In other words, if termination
-of a symbolic function is not bounded, or is only bounded by a symbolic
-parameter, SMT solvers will often fail to terminate when trying to validate
-a property, since doing so often requires a proof by induction. We
-discuss how to address this problem in Section~\ref{subsec:Adapting-the-semantics}.
+Even if we could use symbolic corecursive datatypes, SMT solvers have another
+general limitation: if termination of a function is not bounded
+by a concrete value, SMT solvers may not terminate when determining
+the satisfiability of a property about the function.
+We discuss how to address this in Section~\ref{subsec:Adapting-the-semantics}.
 
@@
 
-In order to guarantee termination of the SMT validation, we need a
-concrete bound for the property that we want to validate.
+In order to guarantee termination of the analysis, we need a
+concrete bound. Related work often addresses this problem by
+manually establishing an artificial bound on the amount of computation,
+e.g: limiting the number of computation steps analysed,
+the number of times loops are unrolled \cite{clarke2004tool,gulwani2008program,jackson2000finding}.
 
-Related work often addresses this problem by manually establishing
-a bound on the amount of computation, e.g: limiting the number of
-computation steps analysed, the number of times loops are unrolled
-\cite{clarke2004tool,gulwani2008program,jackson2000finding}.
-
-In the case of Marlowe, we have natural bounds that we can use because,
-given a concrete contract, we can easily infer:
+Marlowe has natural bounds, given a concrete contract, we can infer:
 \begin{itemize}
@@
 \end{itemize}
-If we keep the \texttt{Contract} parameter concrete,
-we can use it to bound the execution, because every call to \texttt{reduceContractStep}
-will either make no progress or remove one of the constructs, with
-the exception of \texttt{Close}. But for any particular \texttt{Contract},
-we know we only need to call \texttt{reduceContractStep} at most as many times as
-the maximum number of accounts there can possibly be, since, every time we call
-it, it will refund one of the accounts.
-
-Thus, the symbolic transaction processing function becomes:
+Marlowe \texttt{Contract}s are finite, and every call to
+\texttt{reduceContractStep} will either make no progress or
+remove one of the constructs, with the exception of \texttt{Close}.
+In the case of \texttt{Close}, every call to \texttt{reduceContractStep} refunds
+one account, and the number of accounts is also bounded, since each
+needs to be mentioned in the \texttt{Contract}. Thus, the symbolic transaction
+processing function becomes:
 
@@
 computeTransaction :: SBV TransactionInput -> SBV State
-                   -> Contract
-                   -> SBV TransactionOutput
+                   -> Contract -> SBV TransactionOutput
 \end{verbatim}
@@
 
-We get around this problem by using a continuation style. Instead
-of returning the \texttt{TransactionOutput} directly, we take a continuation
+We get around this by using a continuation style. Instead
+of returning the \texttt{TransactionOutput}, we take a continuation
 function that takes the concrete \texttt{Contract} and the symbolic version
@@
                    -> SBV State -> Contract
-                   -> (SBV TransactionOutput ->
-                       Contract -> SBV a)
+                   -> (SBV TransactionOutput -> Contract -> SBV a)
                    -> SBV a
@@
 In practice, we also include some extra information about bounds,
-and we make some other parts of \texttt{TransactionOutput} concrete,
-in addition to the \texttt{Contract}.
+and we make some other parts of \texttt{TransactionOutput} concrete.
 
@@
 
-In this section, we explain, in more detail, the techniques that we mentioned
-in the introduction, and we illustrate those techniques with
-examples from the static analysis implementation of Marlowe.
+In this section, we explain the optimisation techniques in more detail,
+and we illustrate them with examples of their application to Marlowe's static analysis.
 
@@
 
-Some of the modifications that we can apply to a static analysis implementation
-are localised, and they do not fundamentally affect the overall structure
-of the implementation. This means that the implementation can remain close
-to the original semantics and model that we are
-trying to analyze, as well as making changes less likely to introduce
-errors than a more serious restructuring.
+Lightweight modifications are local, which means it is less likely that we
+will introduce reasoning errors when implementing them.
 
@@
 
-This is probably the most straightforward optimization to implement.
 When we use the same or similar code for both the analysis and the
@@
 The \texttt{Close} construct refunds all the money and tokens remaining
-in the accounts. At the time of analysis, we do not know how much
-is available in each account, because it may depend on the value of
-a choice made by the user in runtime. For that reason, the generated
-constraints must include comparisons for every account. This makes
-the implementation of the \texttt{Close} construct very expensive
-to execute symbolically.
+in the accounts. The number of times that \texttt{reduceContractStep} needs
+to be called depends on how many accounts have money left, and because
+this information is symbolic, there are many potential ways in which execution
+can unfold. All these paths need to be represented as constraints, which
+makes analysing \texttt{Close} very costly.
 
-As it turns out, we do not need to implement the semantics of the
-\texttt{Close} construct at all, because it is impossible for the
-\texttt{Close} construct to produce a failed payment
-(we have also proven this\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/isabelle/CloseSafe.thy} (last visited on 04/04/2020)} using Isabelle \cite{nipkow2002isabelle}): it only pays whatever there is available.
-So we can safely remove it from the analysis.
+Fortunately, as it turns out, we do not need to analyse \texttt{Close} at all.
+Because it is impossible for the \texttt{Close} construct to produce a failed payment
+or any other warning (we have proven this\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/isabelle/CloseSafe.thy} (last visited on 04/04/2020)} using Isabelle \cite{nipkow2002isabelle}).
+\texttt{Close} only pays as much as available,
+so we can safely remove it from the analysis.
 
@@
 e.g: custom data-types, list, sets\dots{} Unfortunately, even though
-high level abstractions make reasoning about code easier, they often
-prevent certain optimisations, since they abstract out aspects of
-the implementation that in our particular case may be concrete.
+high level abstractions aid reasoning about code, they often
+prevent optimisations, since they abstract out aspects that
+in our particular case may be concrete.
 
@@
 straightforward to realise using the symbolic implementation of list
-and tuple, both provided by SBV. However, this can be very costly:
-searching a list symbolically requires constraints that check
-the key of every element in the list up to the maximum capacity of
-the associative map, because it is not known at the time of symbolic
-execution how many elements the list has.
+and tuple, both provided by SBV. However, because we assumed keys
+were symbolic, looking up a single element required constraints that
+compared the element with every key up to the maximum capacity of
+the associative map.
 
-However, in Marlowe we know the values of all the keys that we are
+Nevertheless, in Marlowe we know the values of all the keys that we are
 going to use in maps, because the contract is concrete, and only \texttt{Account},
@@
 
-Nevertheless, using a partially concrete map prevents us from using
-it as the result of a function that takes symbolic parameters, so
-we will need to use a continuation style insead. An alternative is
-to assign numbers to each possible key, and store them in a symbolic
-list, this will at least prevent the search from having to compare
-a lookup with every key, since we will know in which position of the
-list each value is stored.
-
 \subsubsection{Reducing the search space by normalising parameters.
@@
 the search space, and the higher the load we are putting on the SMT
-solver. In many cases, we do not need to explore the whole search
-space to check satisfiability of a property. In particular, we may
-be able to prove that some sets of inputs are equivalent in terms
-of the property whose satisfiability we want to test.
+solver. But, if two or more different inputs have the same effect on the
+property we only need to include one of them.
 
@@
 This functionality is important because each transaction requires
-the issuer to pay fees, so combining several inputs in a single transaction
-can translate into important savings. On the other hand, it also means
+the issuer to pay fees. On the other hand, it also means
 that static analysis must consider many more possibilities, since
-the number of ways of partitioning inputs in transactions grows exponentially
+the number of ways of partitioning inputs is exponential
 in the number of inputs.
@@
 
-This optimization reduces the search space considerably, but transactions
-may still have no inputs, so there are still many ways of distributing
-the inputs in transactiongs. We explain how we reduced
-the search space even further in Section~\ref{subsec:Heavyweight-modifications}.
+This optimization reduces the search space, but transactions may still
+have either one or zero inputs, so there are still many ways of distributing
+the inputs in transactions. We explain how we reduced
+the search space further in Section~\ref{subsec:Heavyweight-modifications}.
 
@@
 to the parameters will not grant any improvement to the result.
-For that reason, in this section, we explore ways of optimizing that
+For that reason, in this section, we explore ways of optimising that
 may imply considerable rewriting of our properties, constraints, and
 static analysis implementation in general.
+
+Unfortunately, not following the concrete implementation closely is
+much more error prone, since there are many
+more assumptions that we need to make and reason about.
+In Section~\ref{sec:Testing}, we explore ways of mitigating this issue.
 
@@
 
-One approach we can use that helps improve static analysis efficiency
-is to focus on the property we want to test instead of trying to make
-it as close as possible to the full implementation
-of the system. We do not even need to think about the representation
-of a counterexample (we will discuss that in the next section), but
-only in what are the conditions for the property to be false.
-
-Unfortunately, having static analysis be very different from implementation
-makes the approach much more error prone, and in making an argument
-for the trustworthiness of the analysis there are many more assumptions
-that need to be made and reasoned about. In Section~\ref{sec:Testing},
-we explore ways of mitigating this problem.
+Instead of modelling the execution symbolically, we can focus on modelling
+the property. We do not even need to consider the
+representation of the counterexamples (we will discuss that in the next section),
+but only in what are the conditions for the property to be false.
 
 For example, the main property we want to check is whether there is
-any possible execution that produces a failed payment. Thus, we look
-at executions instead of looking at inputs. The most complicated construct
-in terms of executions is the \texttt{When} construct, since it allows
-for two transactions that are separate in time to have different effects
-depending on when they are issued, all other constructs will get resolved
-instantly in one way or another. But, without loss of generality,
-we can structure them as Figure~\ref{fig:Distribution-of-transactions}
-shows: we can conceptually break the contract tree into subtrees,
-where each subtree has a \texttt{When} construct as its root, with
-the exception of the subtree at the root of the original tree.
+any possible execution that produces a failed payment. Thus, we only
+write constraints for executions in which this can happen instead of
+modelling all possibilities.
+The most complicated construct in terms of execution is the \texttt{When}
+construct, since it allows for transactions that are separate in time
+to have different effects depending on when they are issued, all other
+constructs will get resolved atomically in one way or another. Without
+loss of generality, we can structure possible executions as shown in
+Figure~\ref{fig:Distribution-of-transactions}: we can conceptually break
+the contract tree into subtrees, where each subtree has a \texttt{When}
+construct as its root, with the exception of the subtree at the root
+of the original tree.
 
 \begin{figure}
-\includegraphics[width=1\textwidth]{figures/transactions}
+\center
+\includegraphics[width=0.7\textwidth]{figures/transactions}
 
@@
 
-Each level of subtrees will correspond to a transaction, i.e: the
+Each level of subtrees corresponds to a potential transaction, i.e: the
 root subtree will correspond to the first transaction, the set of
@@
 will correspond to the second transaction, and so on. There may be
-some paths which do not require as many transactions/subtrees as other
-paths, because they traverse fewer \texttt{When} constructs before
-finding a \texttt{Close} construct.
+paths which require fewer transactions/subtrees because they traverse
+fewer \texttt{When} constructs.
 
-We can assign one transaction to each segment of execution
-before starting at a \texttt{When} or the root of the tree,
-because if the maximum slot number of the
-transaction is lower than the slot number in the timeout and the
-transaction has no inputs, then the execution will stop before the
-\texttt{When}, so we need to account for that possibility. In other
-words, if two parts of the execution path may have different slot
-numbers when we go through them, we need to consider them as separate
-transactions, because they may have payments that depend on those values.
+We split transactions like this because if the maximum slot number of
+a transaction is lower than the slot number in the timeout of a \texttt{When}
+it will stop before executing that \texttt{When}. Since the \texttt{When}
+and its continuation may be executed by a different transaction, the
+slot numbers for that segment of execution may be different, which means
+the values of \texttt{Value}s may be different, which means the amounts
+in payments and deposits may be different, and thus the warnings issued may
+be different as if both segments were executed by the same transaction.
 
-On the other hand, this segmentation is not general enough, because there is
-an edge case. A transaction may execute past a \texttt{When} if the minimum
-slot number in the transaction is greater or equal than the timeout of the
-\texttt{When}. If the transaction expires the \texttt{When} we can no longer
-separate the execution before the \texttt{When} and after the \texttt{When}
-into two different transactions, because we know that the timeout branch of the
-\texttt{When} will be executed as part of the first transaction, and thus
-it will use the slot interval of the first
-transaction (same used for the first segment).
+However, there is an edge case: a transaction may execute past 
+a \texttt{When} if the minimum slot number in the transaction is greater
+or equal than the timeout of the \texttt{When}. If the transaction expires
+the \texttt{When} we can no longer separate the execution before the
+\texttt{When} and after the \texttt{When} into two different transactions,
+because we know that the timeout branch of the \texttt{When} will be executed
+as part of the first transaction, and thus it will use the slot interval
+of the first transaction (same used for the first segment).
 
 This is not only true for the timeout branch, if we include
-an input that is allowed by the \texttt{When} into the first transaction,
-then the first transaction will continue into the corresponding \texttt{When}
+an input that is allowed by a \texttt{When}, then the first transaction
+will continue into the corresponding \texttt{When}
 branch, and we will not use a second transaction.
@@
 \texttt{When} in a row, or where it expires one \texttt{When} and provides
-the right input for later \texttt{When}. And we cannot just not constrain
+the right input for a later \texttt{When}. And we cannot just not constrain
 the maximum slot number of the first transaction because then we will
-not know to which segments each transaction applies.
+be considering impossible executions.
 
@@
 
-Note however that, in Section~\ref{subsec:Reducing-search-space-by-normalising-parameters},
-we already reduced the number of inputs per transaction to a maximum
-of one, and the number of transactions to the maximum number of nested
-\texttt{When}s plus one. But now we have also assigned each of the
-transactions to a part of the contract, so we do not need a symbolic
-list of transactions, we can have a finite concrete list of symbolic
-transactions and we know in advance which of the transactions we are
-working on at each moment during the symbolic execution.
+To sum up, in Section~\ref{subsec:Reducing-search-space-by-normalising-parameters}, we already limitted the number of inputs per transaction
+to a maximum of one. But now we have also assigned each of the
+transactions to a part of the contract, so we no longer need a symbolic
+list of transactions, we can use a finite concrete list of symbolic
+transactions.
 
-One detail that we realised during testing is that, even though
-the diagram above shows that a \texttt{When} belongs to the new transaction,
-because the \texttt{When} will not be consumed by the previous transaction,
-\texttt{Observation}s in the \texttt{Notify} cases will correspond to the
-information there was in the \texttt{State} before the \texttt{When}
-was executed, with the exception of the slot interval, which will still be
-considered to be that of the transaction that triggers the \texttt{When} (the
-new transaction). The same is true for \texttt{Value}s in the \texttt{Deposit}
-cases, since the amount to deposit must be calculated without considering the
-effects of the deposit itself, we need to refer to the \texttt{State} before
-the deposit was made.
+One detail we found out during testing is that, even though a \texttt{When}
+belongs to the begining of a transaction, the environment used by
+the \texttt{Observation}s in the \texttt{Notify} cases of the \texttt{When}
+correspond to the \texttt{State} before the \texttt{When} is executed (except
+for the slot interval). The same is true for the \texttt{Value} in the
+\texttt{Deposit} cases, since the amount to deposit must be calculated without
+considering the effects of the deposit itself.
 
@@
 
-\begin{figure}
+\begin{figure}[h]
+\centering
+\begin{minipage}{.5\textwidth}
     \begin{centering}
-\includegraphics[width=0.5\textwidth]{figures/consistency_property}
+    \includegraphics[width=0.8\textwidth]{figures/consistency_property}
     \par\end{centering}
-\caption{Property based testing for consistency\label{fig:Property-based-testing-for-consistency}}
-
+    \caption{Testing for consistency\label{fig:Property-based-testing-for-consistency}}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+    \begin{centering}
+    \includegraphics[width=0.7\textwidth]{figures/equivalence_property}
+    \par\end{centering}
+    \caption{Testing for equivalence\label{fig:Property-based-testing-for-equivalence}}
+\end{minipage}
 \end{figure}
@@
 
-\begin{figure}
-\begin{centering}
-\includegraphics[width=0.45\textwidth]{figures/equivalence_property}
-\par\end{centering}
-\caption{Property based testing for equivalence\label{fig:Property-based-testing-for-equivalence}}
-\end{figure}
-
 This approach covers both types of errors, i.e: false positives and
@@
 
+In our experiments, the heavyweight optimisations considerably reduced
+the requirements of both processing time and memory for Marlowe's static
+analysis. %In Tables~\ref{table1},~\ref{table2},~\ref{table3},~and~\ref{table4},
+We present below the results of measuring the performance of static analysis on
+four example  contracts,
+before\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/src/Language/Marlowe/Analysis/FSSemantics.hs} [last visited 19-05-2020]}
+and after\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/src/Language/Marlowe/Analysis/FSSemanticsFastVerbose.hs} [last visited 19-05-2020]}
+the heavyweight optimisations.
+
 \begin{table}[]
@@
 \multicolumn{6}{l}{\textbf{Auction contract -- \texttt{Auction.hs}}} \\ \hline \hline
+\textbf{Num.~participants}             & \multicolumn{1}{c|}{1}        & \multicolumn{1}{c|}{2}            & \multicolumn{1}{c|}{3}                         & \multicolumn{1}{c|}{4}                        & \multicolumn{1}{c}{5}    \\ \hline
+\textbf{Contract size (chars)}         & \multicolumn{1}{c|}{275}        & \multicolumn{1}{c|}{3,399}     & \multicolumn{1}{c|}{89,335}    & \multicolumn{1}{c|}{4,747,361}      & \multicolumn{1}{c}{413,784,559} \\ \hline \hline
 \multicolumn{6}{c}{\textbf{Lightweight optimisations}} \\ \hline \hline
-\textbf{Num.~participants}             & \multicolumn{1}{c|}{1}        & \multicolumn{1}{c|}{2}            & \multicolumn{1}{c|}{3}                         & \multicolumn{1}{c|}{4}                        & \multicolumn{1}{c}{5}    \\ \hline
 \textbf{Execution time}                & \multicolumn{1}{c|}{0.2205s}  & \multicolumn{1}{c|}{4m 45.2015s}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A}  & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
@@
 \multicolumn{6}{c}{\textbf{Heavyweight optimisations}} \\ \hline \hline
-\textbf{Num.~participants}             & \multicolumn{1}{c|}{1}          & \multicolumn{1}{c|}{2}         & \multicolumn{1}{c|}{3}         & \multicolumn{1}{c|}{4}              & \multicolumn{1}{c}{5}    \\ \hline
 \textbf{Execution time}                & \multicolumn{1}{c|}{0.01198s}   & \multicolumn{1}{c|}{0.0289s}   & \multicolumn{1}{c|}{1.1138s}   & \multicolumn{1}{c|}{1h 5m 45.1377s} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
@@
 \multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.010031s}  & \multicolumn{1}{c|}{0.020334s} & \multicolumn{1}{c|}{0.845486s} & \multicolumn{1}{c|}{12m 35.099356s} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-\textbf{RAM usage peak}                & \multicolumn{1}{c|}{17,020KB}   & \multicolumn{1}{c|}{17,740KB}  & \multicolumn{1}{c|}{49,668KB}  & \multicolumn{1}{c|}{2,364,500KB}    & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline \hline
-\textbf{Contract size (chars)}         & \multicolumn{1}{c|}{275}        & \multicolumn{1}{c|}{3,399}     & \multicolumn{1}{c|}{89,335}    & \multicolumn{1}{c|}{4,747,361}      & \multicolumn{1}{c}{413,784,559}
+\textbf{RAM usage peak}                & \multicolumn{1}{c|}{17,020KB}   & \multicolumn{1}{c|}{17,740KB}  & \multicolumn{1}{c|}{49,668KB}  & \multicolumn{1}{c|}{2,364,500KB}    & \multicolumn{1}{c}{\color[HTML]{888888} N/A}
 \end{tabular}
@@
 \multicolumn{6}{l}{\textbf{Crowdfunding contract -- \texttt{CrowdFunding.hs}}} \\ \hline \hline
+\textbf{Num.~participants}             & \multicolumn{1}{c|}{1}         & \multicolumn{1}{c|}{2}            & \multicolumn{1}{c|}{3}                        & \multicolumn{1}{c|}{4}                        & \multicolumn{1}{c}{5}    \\
+\hline
+\textbf{Contract size (chars)}         & \multicolumn{1}{c|}{857}      & \multicolumn{1}{c|}{12,704}   & \multicolumn{1}{c|}{364,824}  & \multicolumn{1}{c|}{19,462,278}   & \multicolumn{1}{c}{1,690,574,798} \\ \hline \hline
 \multicolumn{6}{c}{\textbf{Lightweight optimisations}} \\ \hline \hline
-\textbf{Num.~participants}             & \multicolumn{1}{c|}{1}         & \multicolumn{1}{c|}{2}            & \multicolumn{1}{c|}{3}                        & \multicolumn{1}{c|}{4}                        & \multicolumn{1}{c}{5}    \\ \hline
 \textbf{Execution time}                & \multicolumn{1}{c|}{0.6298s}   & \multicolumn{1}{c|}{50m 53.4597s} & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c|}{\color[HTML]{888888} N/A} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
@@
 \multicolumn{6}{c}{\textbf{Heavyweight optimisations}} \\ \hline \hline
-\textbf{Num.~participants}             & \multicolumn{1}{c|}{1}        & \multicolumn{1}{c|}{2}        & \multicolumn{1}{c|}{3}        & \multicolumn{1}{c|}{4}            & \multicolumn{1}{c}{5}    \\ \hline
 \textbf{Execution time}                & \multicolumn{1}{c|}{0.0125s}  & \multicolumn{1}{c|}{0.041s}   & \multicolumn{1}{c|}{1.0515s}  & \multicolumn{1}{c|}{32m 15.4478s} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
@@
 \multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.0104s}  & \multicolumn{1}{c|}{0.0306s}  & \multicolumn{1}{c|}{0.6600s}  & \multicolumn{1}{c|}{9m 43.9246s}  & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
-\textbf{RAM usage peak}                & \multicolumn{1}{c|}{17,016KB} & \multicolumn{1}{c|}{18,768KB} & \multicolumn{1}{c|}{62,108KB} & \multicolumn{1}{c|}{3,715,020KB}  & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline \hline
-\textbf{Contract size (chars)}         & \multicolumn{1}{c|}{857}      & \multicolumn{1}{c|}{12,704}   & \multicolumn{1}{c|}{364,824}  & \multicolumn{1}{c|}{19,462,278}   & \multicolumn{1}{c}{1,690,574,798}
+\textbf{RAM usage peak}                & \multicolumn{1}{c|}{17,016KB} & \multicolumn{1}{c|}{18,768KB} & \multicolumn{1}{c|}{62,108KB} & \multicolumn{1}{c|}{3,715,020KB}  & \multicolumn{1}{c}{\color[HTML]{888888} N/A}
 \end{tabular}
@@
 \multicolumn{6}{l}{\textbf{Rent contract -- \texttt{Rent.hs}}} \\ \hline \hline
+\textbf{Num.~months}                   & \multicolumn{1}{c|}{1}        & \multicolumn{1}{c|}{2}         & \multicolumn{1}{c|}{3}           & \multicolumn{1}{c|}{4}               & \multicolumn{1}{c}{5}    \\ \hline
+\textbf{Contract size (chars)}         & \multicolumn{1}{c|}{339}      & \multicolumn{1}{c|}{595}      & \multicolumn{1}{c|}{852}      & \multicolumn{1}{c|}{1,109}    & \multicolumn{1}{c}{1,366} \\ \hline \hline
 \multicolumn{6}{c}{\textbf{Lightweight optimisations}} \\ \hline \hline
-\textbf{Num.~months}                   & \multicolumn{1}{c|}{1}        & \multicolumn{1}{c|}{2}         & \multicolumn{1}{c|}{3}           & \multicolumn{1}{c|}{4}               & \multicolumn{1}{c}{5}    \\ \hline
 \textbf{Execution time}                & \multicolumn{1}{c|}{0.2850s}  & \multicolumn{1}{c|}{3.0303s}   & \multicolumn{1}{c|}{2m 53.2458s} & \multicolumn{1}{c|}{3h 22m 13.0122s} & \multicolumn{1}{c}{\color[HTML]{888888} N/A} \\ \hline
@@
 \multicolumn{6}{c}{\textbf{Heavyweight optimisations}} \\ \hline \hline
-\textbf{Num.~months}                   & \multicolumn{1}{c|}{1}        & \multicolumn{1}{c|}{2}        & \multicolumn{1}{c|}{3}        & \multicolumn{1}{c|}{4}        & \multicolumn{1}{c}{5}    \\ \hline
 \textbf{Execution time}                & \multicolumn{1}{c|}{0.0114s}  & \multicolumn{1}{c|}{0.0111s}  & \multicolumn{1}{c|}{0.01132s} & \multicolumn{1}{c|}{0.01124s} & \multicolumn{1}{c}{0.01255s} \\ \hline
@@
 \multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.0101s}  & \multicolumn{1}{c|}{0.0098s}  & \multicolumn{1}{c|}{0.0098s}  & \multicolumn{1}{c|}{0.0097s}  & \multicolumn{1}{c}{0.0099s} \\ \hline
-\textbf{RAM usage peak}                & \multicolumn{1}{c|}{15,536KB} & \multicolumn{1}{c|}{15,364KB} & \multicolumn{1}{c|}{15,400KB} & \multicolumn{1}{c|}{15,364KB} & \multicolumn{1}{c}{15,540KB} \\ \hline \hline
-\textbf{Contract size (chars)}         & \multicolumn{1}{c|}{339}      & \multicolumn{1}{c|}{595}      & \multicolumn{1}{c|}{852}      & \multicolumn{1}{c|}{1,109}    & \multicolumn{1}{c}{1,366}
+\textbf{RAM usage peak}                & \multicolumn{1}{c|}{15,536KB} & \multicolumn{1}{c|}{15,364KB} & \multicolumn{1}{c|}{15,400KB} & \multicolumn{1}{c|}{15,364KB} & \multicolumn{1}{c}{15,540KB}
+
 \end{tabular}
@@
 \multicolumn{6}{l}{\textbf{Coupon bond contract - \texttt{CouponBond.hs}}} \\ \hline \hline
+\textbf{Num.~months}                   & \multicolumn{1}{c|}{2}        & \multicolumn{1}{c|}{3}         & \multicolumn{1}{c|}{4}           & \multicolumn{1}{c|}{5}            & \multicolumn{1}{c}{6}    \\ \hline
+\textbf{Contract size (chars)}         & \multicolumn{1}{c|}{479}      & \multicolumn{1}{c|}{635}      & \multicolumn{1}{c|}{791}      & \multicolumn{1}{c|}{947}      & \multicolumn{1}{c}{1,103} \\ \hline \hline
 \multicolumn{6}{c}{\textbf{Lightweight optimisations}} \\ \hline \hline
-\textbf{Num.~months}                   & \multicolumn{1}{c|}{2}        & \multicolumn{1}{c|}{3}         & \multicolumn{1}{c|}{4}           & \multicolumn{1}{c|}{5}            & \multicolumn{1}{c}{6}    \\ \hline
 \textbf{Execution time}                & \multicolumn{1}{c|}{0.8293s}  & \multicolumn{1}{c|}{5.8887s}   & \multicolumn{1}{c|}{1m 35.3930s} & \multicolumn{1}{c|}{26m 36.4585s} & \multicolumn{1}{c}{9h 50m 22.3418s} \\ \hline
@@
 \multicolumn{6}{c}{\textbf{Heavyweight optimisations}} \\ \hline \hline
-\textbf{Num.~months}                   & \multicolumn{1}{c|}{2}        & \multicolumn{1}{c|}{3}        & \multicolumn{1}{c|}{4}        & \multicolumn{1}{c|}{5}        & \multicolumn{1}{c}{6}    \\ \hline
 \textbf{Execution time}                & \multicolumn{1}{c|}{0.0092s}  & \multicolumn{1}{c|}{0.0095s}  & \multicolumn{1}{c|}{0.0097s}  & \multicolumn{1}{c|}{0.0102s}  & \multicolumn{1}{c}{0.0105s} \\ \hline
@@
 \multicolumn{1}{r||}{× execution time} & \multicolumn{1}{c|}{0.0079s}  & \multicolumn{1}{c|}{0.0080s}  & \multicolumn{1}{c|}{0.0082s}  & \multicolumn{1}{c|}{0.0082s}  & \multicolumn{1}{c}{0.0085s} \\ \hline
-\textbf{RAM usage peak}                & \multicolumn{1}{c|}{15,636KB} & \multicolumn{1}{c|}{15,816KB} & \multicolumn{1}{c|}{15,788KB} & \multicolumn{1}{c|}{15,780KB} & \multicolumn{1}{c}{15,760KB} \\ \hline \hline
-\textbf{Contract size (chars)}         & \multicolumn{1}{c|}{479}      & \multicolumn{1}{c|}{635}      & \multicolumn{1}{c|}{791}      & \multicolumn{1}{c|}{947}      & \multicolumn{1}{c}{1,103}
+\textbf{RAM usage peak}                & \multicolumn{1}{c|}{15,636KB} & \multicolumn{1}{c|}{15,816KB} & \multicolumn{1}{c|}{15,788KB} & \multicolumn{1}{c|}{15,780KB} & \multicolumn{1}{c}{15,760KB}
 \end{tabular}
@@
 \end{table}
-
-In our experiments, the heavyweight optimisations considerably reduced
-the requirements of both processing time and memory for Marlowe's static
-analysis. %In Tables~\ref{table1},~\ref{table2},~\ref{table3},~and~\ref{table4},
-We present below the results of measuring the performance of static analysis on
-four example  contracts,
-before\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/src/Language/Marlowe/Analysis/FSSemantics.hs} [last visited 19-05-2020]}
-and after\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/src/Language/Marlowe/Analysis/FSSemanticsFastVerbose.hs} [last visited 19-05-2020]}
-the heavyweight optimisations.
 

### 2020-05-28 papers/static_analysis2020/paper.tex
Tidied up bib and adsded orginal (ISoLA) Marlowe reference.

@@
 Thanks to static analysis, we can automatically check beforehand whether any
-payments promised by a Marlowe contract can be fulfilled in every
+payments promised by a Marlowe~\cite{Marlowe:2018,marlowe} contract can be fulfilled in every
 possible execution of the contract. If a Marlowe contract has passed
@@
 facilitate static analysis and make it decidable. A more detailed
-explanation can be found in \cite{marlowe}.
+explanation can be found in \cite{marlowe2020}.
 

### 2020-05-28 papers/static_analysis2020/paper.tex
Various ods: revised conclusions, put future work into conclusions. Space saving in Section 5 and Section 6.

@@
 Thanks to static analysis, we can automatically check beforehand whether any
-payments promised by a Marlowe~\cite{Marlowe:2018,marlowe} contract can be fulfilled in every
+payments promised by a Marlowe~\cite{marlowe2020,Marlowe:2018} contract can be fulfilled in every
 possible execution of the contract. If a Marlowe contract has passed
@@
 Marlowe's static analysis relies on SMT solvers, which are able to check efficiently
-whether a set of constraints is satisfiable. Thanks to state of the art
+whether a set of constraints is satisfiable. The main property for us is whether a contract will have enough money for all payments to be made in full.
+
+Thanks to state of the art
 libraries like SBV \cite{erkok2019sbv}, we can describe those constraints in a
@@
 implemented; SBV automatically translates those functions to SMTLib format \cite{barrett2010smt}, a language many SMT solvers understand.
+However, high level abstractions often have a tradeoff with efficiency, and static analysis can become very expensive computationally if
+implemented naïvely, because it is NP-complete in the general case.
 
-Nevertheless, static analysis can become very computationally expensive if
-implemented naïvely, because it is an NP-complete problem in the general case.
-In our case the main property is whether a contract will have enough money for all payments. Indeed, high level abstractions are often a tradeoff with efficiency.
+% Indeed, high level abstractions are often a tradeoff with efficiency.
 
@@
 optimising static analysis, with examples extracted from a case-study where
-we applied these approaches, an approach to ensure correctness
-of the optimisations that combines the usage of property-based testing
-and verification using automated theorem provers, and empirical data
+we applied these approaches. Our approach aims to ensure correctness
+of the optimisations by combining the use of property-based testing
+and with verification in an automated theorem prover to establish properties of the optimisations. We also present empirical data
 that measures the effect of some of the optimisations on the implementation
-of our case-study: the optimisation of Marlowe's static analysis.
+of our Marlowe case-study.
 
@@
 around 10 minutes. 
-
 We have classified optimisation techniques as lightweight and heavyweight.
@@
 In the following sections, we introduce the semantics of Marlowe as
-a case study (Section~\ref{sec:Marlowe-intro}), we cover a
-general approach to static analysis (Section~\ref{sec:Making-Marlowe-symbolic}),
-we then explore both lightweight (Section~\ref{subsec:Lightweight-modifications})
-and heavyweight (Section~\ref{subsec:Heavyweight-modifications})
-optimisation techniques in more detail, and we illustrate them with
+a case study (Section~\ref{sec:Marlowe-intro}), and we cover a
+general approach to static analysis (Section~\ref{sec:Making-Marlowe-symbolic}).
+We then, in Section \ref{sec:Making-static-analysis-efficient} explore in Section the lightweight and heavyweight 
+optimisation techniques in more detail, and illustrate them with
 examples of how they apply to Marlowe static analysis. In Section~\ref{sec:Testing},
@@
 
-To sum up, in Section~\ref{subsec:Reducing-search-space-by-normalising-parameters}, we already limitted the number of inputs per transaction
+To sum up, in Section~\ref{subsec:Reducing-search-space-by-normalising-parameters}, we already limited the number of inputs per transaction
 to a maximum of one. But now we have also assigned each of the
@@
 One detail we found out during testing is that, even though a \texttt{When}
-belongs to the begining of a transaction, the environment used by
+belongs to the beginning of a transaction, the environment used by
 the \texttt{Observation}s in the \texttt{Notify} cases of the \texttt{When}
@@
 
-We use this separation of concerns between the static analysis and
-the concrete semantics to our advantage, as an opportunity for applying
-property based testing, as we explain in Section~\ref{subsec:Testing-for-consistency}.
+We use this separation of concerns between static analysis and
+concrete semantics to our advantage, as an opportunity for applying
+property based testing.%, as we explain in Section~\ref{subsec:Testing-for-consistency}.
 
@@
 hold.
-
 In order to avoid these problems, we combine the use of automated proof
 assistants to verify our assumptions and the use of property based testing to
-test for consistency and equivalence (as we describe in
-Section~\ref{subsec:Testing-for-consistency}).
+test for consistency and equivalence. %as we describe in
+%Section~\ref{subsec:Testing-for-consistency}).
 
@@
 
-A limitation of this approach is that it only tests for false positives,
-and it does not prevent false negatives, but those can be covered by testing
+A limitation of this approach is that it only tests for false positives; 
+false negatives can be detected by testing
 for equivalence (see Section~\ref{subsec:Testing-for-equivalence}).
@@
 or that it has invalid or ambiguous intervals. If it has \texttt{UselessTransactions}
-is ok, because we are doing that on purpose, as we mentioned in Section~\ref{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}.
+that is ok, because we are doing that on purpose, as we mentioned in Section~\ref{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}.
 
@@
 \centering
-\begin{minipage}{.5\textwidth}
+\begin{minipage}{.4\textwidth}
     \begin{centering}
@@
 \end{minipage}%
-\begin{minipage}{.5\textwidth}
+\begin{minipage}{.4\textwidth}
     \begin{centering}
@@
 
+\vspace*{-5mm}
+
 \subsection{Testing for equivalence\label{subsec:Testing-for-equivalence} }
 
-If we have two static analysis implementations, we have another opportunity
-for testing. In our case, we have one efficient implementation that
+Given two  implementations of the static analysis, we have another opportunity
+for testing. We have one efficient implementation that
 is very different from the semantics and one inefficient implementation
 that is much closer to the semantics, and we can compare the results of the
-two implementations. We can do this as shown in Figure~\ref{fig:Property-based-testing-for-equivalence}.
+two, as shown in Figure~\ref{fig:Property-based-testing-for-equivalence}.
+
 We generate random contracts using a custom QuickCheck \cite{quickcheck} generator,
-we feed them to both static analysis implementations, and we compare the results.
+ feed them to both implementations, and compare the results.
 If the results are the same then test passes, if they are different then
 one of the implementations is wrong.
-
 This approach covers both types of errors, i.e: false positives and
@@
 
-Using the \texttt{perf} tool \cite{perf}, we have measured the execution
+Using the \texttt{perf} tool \cite{perf}, we measured the execution
 time\footnote{The experiments were run on a laptop computer with a
 i9-9900K (3.6GHz) processor and two modules of 16GB of SODIMM DDR4 RAM at 2400MHz.}
-and the overhead of both the generation of the constraints and their solution
-by Z3 \cite{de2008z3}. And we have measured the peak RAM usage of the whole
-process by using GNU's \texttt{time} tool \cite{keppel1998gnu}.
+and the overhead of the generation of the constraints and their solution
+by Z3 \cite{de2008z3}. We also measured the peak RAM usage of the whole
+process using GNU's \texttt{time} tool \cite{keppel1998gnu}.
 
-In all the contracts, the implementation that has the heavyweight optimisations
+In all cases the implementation with the heavyweight optimisations
 performs much better and scales further. In the case of the \textit{auction}
@@
 approaches quickly overwhelm the resources available in the execution environment.
-On the other hand, in the case of \textit{rent} and \textit{coupon bond} contracts,
-which grow linearly, by using the lightweight version, the problem becomes 
-intractable much faster than with the heavyweight version.
+In the case of \textit{rent} and \textit{coupon bond} contracts,
+which grow linearly, when using the lightweight version the problem becomes 
+intractable much faster than with the heavyweight one.
 
@@
 These results suggest that the SBV library is able to generate constraints in a way
-that they are handled efficiently by Z3, but the process itself can be costly
-if we are not careful. Nevertheless, the overall execution time required by Z3
-is also lower in the case of the version with heavyweight optimisations and it
-grows more slowly, which suggests the optimisations described in this paper
-affect both stages of the process to some extent.
+that they are handled efficiently by Z3, but the process itself can be costly. However, the  execution time required by Z3
+is also lower in the case of the heavyweight optimisations as well as
+growing more slowly, which suggests that the optimisations described here
+affect both parts of the process.
 
@@
 Work in \cite{mokhov2019formal} documents a similar effort to ensure
-correctness of implementations by using Haskell and SBV library; the
+correctness of control software in Haskell using the SBV library; the
 authors also discuss performance of the analysis and apply this approach
@@
 The idea of using constraint solvers for finding bugs is not new,
-and there has been a number of initiatives that have explored its
+and there have been a number of initiatives that have explored its
 application to the verification of assertions in programs written
 using general purpose programming languages \cite{gulwani2008program,jackson2000finding};
-as well as for the compliance with usage protocols
+as well as for the compliance with protocols
 \cite{ball2001automatically,xie2005saturn}.
@@
 
-\section{Future work}
-
-In the future, we would like to extend static analysis to cover other
-potential problems in Marlowe contracts and to aid their development.
-We can envisage that we can use static analysis to locate unreachable
-subcontracts, to allow developers to provide custom assertions and
-check their satisfiability, and even to allow users to inspect the
-possible maximum and minimum values that particular expressions can
-reach.
-
 \section{Conclusion}
 
-In this paper, we have summarized the lessons learned from our work
-on optimising the static analysis for Marlowe contracts. We have found
-that there are two clear approaches to static analysis using SMT,
+In this paper we have summarized our work
+on optimising the static analysis for Marlowe contracts. We have seen
+that there are two distinct approaches to static analysis using SMT,
 both have advantages and disadvantages. One is less error prone and
-straightforward, but inefficient and hard to test; the other approach
+straightforward, but inefficient and hard to test; the other 
 is much more efficient, versatile, and testable, but more error prone.
-
 We have also seen that many specific properties and restrictions characteristic
@@
 and, in our case, for completeness of the analysis. Symbolic execution
-of, for example, a turing complete language, would be intractable,
+of a Turing-complete language, would be intractable,
 and would require us to manually set a bound; but this is not in the
@@
 In the end, we have illustrated how to counteract the main disadvantage
-of the later approach, its propensity to errors, by using property
+of the optimised approach -- its propensity to errors -- by using property
 based testing. This way we have obtained a static analysis implementation
 that is efficient, versatile, testable, and reliable. 
-
 On the other hand, for the static analysis of Marlowe contracts, we
@@
 SMT solvers which, in turn, means that it is less reliant on the
-correctness of a single SMT solver. If an SMT solver fails to give
-an answer, we can try a different one; if we want to ensure that a
-contract is valid, we can test it with several SMT solvers and ensure
-all of them agree.
+correctness or efficiency of a single solver. If one solver fails to give
+an answer, we can try another; if we want further evidence that a
+contract is valid, we can test it with several solvers.
+
+In the future, we would like to extend static analysis to cover other
+potential problems in Marlowe contracts and to aid their development.
+We plan to use static analysis to locate unreachable
+subcontracts, to allow developers to provide custom assertions and
+check their satisfiability, and to allow users to inspect the
+possible maximum and minimum values that particular expressions can
+reach.
 

### 2020-05-28 papers/static_analysis2020/paper.tex
Reduce paper to 16 pages

@@
 \usepackage{multirow}
+\usepackage[bottom]{footmisc}
 \usepackage[table,xcdraw]{xcolor}
@@
 of the function to return the offending trace using the symbolic \texttt{Maybe}
-type. Surprisingly, this small change increases the amount of time
+type. Surprisingly, this change increases the time
 required by the symbolic analysis severalfold.
 
-The solution to this increase in execution time was to pass as input, to
-the static analysis implementation, a fixed list, with each element corresponding
-to a transaction and consisting of a tuple with four symbolic integers:
+An efficient way of extracting counterexample information was to
+pass a symbolic fixed-size list, as input to the static analysis, where
+each element corresponds to a transaction and consists of a tuple with four symbolic integers:
 \begin{enumerate}
@@
 \item An integer representing the \texttt{When} case whose input is being
-included in the transaction, where zero represents the timeout branch
-(and there being no inputs to the transaction).
+included in the transaction, where zero represents the timeout branch (and no input).
 \item An integer representing the amount of money or tokens (if the input
@@
 \end{enumerate}
-If we need fewer transactions, we will just add dummy transactions with all four numbers
-set to $-1$ at the end of the list.
+For short branches (that require fewer transactions) we pad the end of
+the list with dummy transactions with all four numbers set to $-1$.
 
@@
 of transactions that is human readable and we can use to report the
-counter example, we need to iterate through the list and see the evolution
-of the contract with each transaction, using the concrete semantics.
-This way, we can gather the rest of necessary information, such as
-whether the transaction input is a \texttt{Deposit}. We can also see
-which of the transactions do not have an effect, i.e: they produce
-\texttt{UselessTransaction} as we mentioned in Section~\ref{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}.
+counter example, we need to use the concrete semantics together with
+the information obtained from the static analysis to \emph{fill the gaps},
+by iterating through the list and looking at the evolution
+of the contract with each transaction.
+This provides us with the rest of necessary information, such as
+the type of a transaction input (e.g: whether it is a \texttt{Deposit} or a \texttt{Choice}). We also use this process to filter transactions with no effect,
+i.e: they produce \texttt{UselessTransaction} as we mentioned in Section~\ref{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}.
 
-We use this separation of concerns between static analysis and
-concrete semantics to our advantage, as an opportunity for applying
+We also use this separation of concerns between static analysis and
+concrete semantics as an opportunity for applying
 property based testing.%, as we explain in Section~\ref{subsec:Testing-for-consistency}.
@@
 
-Errors often occur when optimising or translating. If the static analysis
-implementation is close to the actual implementation, it is much less
-likely that a mistake will be made when writing it. On the other hand,
-if we have to craft a static analysis implementation that is equivalent
-(in some sense) to the original, but is written in a very different
-way, it is more likely that we will make assumptions which do not
-hold.
-In order to avoid these problems, we combine the use of automated proof
-assistants to verify our assumptions and the use of property based testing to
-test for consistency and equivalence. %as we describe in
-%Section~\ref{subsec:Testing-for-consistency}).
+The more different the static analysis implementation and the concrete
+implementation are, the harder it is to ensure they are consistent with
+each other. To ensure that heavyweight optimisations remain consistent,
+we combine the use of automated proof assistants and the use of property based testing.
 
@@
 \end{minipage}%
+\hspace{0.08\textwidth}
 \begin{minipage}{.4\textwidth}
@@
 
-\begin{table}[]
+Unfortunately, at the time of writing, we do not have a completely unoptimised
+version of the static analysis that we can use to compare, because the semantics
+of Marlowe have changed since we implemented that version. However, our impression
+from manual testing is that the impact of lightweight optimisations was much more modest than the impact of heavyweight optimisations.
+
+Using the \texttt{perf} tool \cite{perf}, we measured the execution
+time\footnote{The experiments were run on a laptop computer with a
+i9-9900K (3.6GHz) processor and two modules of 16GB of SODIMM DDR4 RAM at 2400MHz.}
+and the overhead of the generation of the constraints and their solution
+by Z3 \cite{de2008z3}. We also measured the peak RAM usage of the whole
+process using GNU's \texttt{time} tool \cite{keppel1998gnu}.
+
+In all cases the implementation with the heavyweight optimisations
+performs much better and scales further. In the case of the \textit{auction}
+and \textit{crowdfunding} contracts, whose size grows exponentially, both
+approaches quickly overwhelm the resources available in the execution environment.
+In the case of \textit{rent} and \textit{coupon bond} contracts,
+which grow linearly, when using the lightweight version the problem becomes 
+intractable much faster than with the heavyweight one.
+
+\begin{table}[b]
 \centering
@@
 \noindent
-Unfortunately, at the time of writing, we do not have a completely unoptimised
-version of the static analysis that we can use to compare, because the semantics
-of Marlowe have changed since we implemented that version. However, our impression
-from manual testing is that the impact of lightweight optimisations was much more modest than the impact of heavyweight optimisations.
-
-Using the \texttt{perf} tool \cite{perf}, we measured the execution
-time\footnote{The experiments were run on a laptop computer with a
-i9-9900K (3.6GHz) processor and two modules of 16GB of SODIMM DDR4 RAM at 2400MHz.}
-and the overhead of the generation of the constraints and their solution
-by Z3 \cite{de2008z3}. We also measured the peak RAM usage of the whole
-process using GNU's \texttt{time} tool \cite{keppel1998gnu}.
-
-In all cases the implementation with the heavyweight optimisations
-performs much better and scales further. In the case of the \textit{auction}
-and \textit{crowdfunding} contracts, whose size grows exponentially, both
-approaches quickly overwhelm the resources available in the execution environment.
-In the case of \textit{rent} and \textit{coupon bond} contracts,
-which grow linearly, when using the lightweight version the problem becomes 
-intractable much faster than with the heavyweight one.
 

### 2020-07-06 papers/static_analysis2020/paper.tex
SCP-843 - Address reviews from ISOLA 2020

@@
 that followed the semantics closely and took a couple of minutes to analyse contracts
-of a few kilobytes, to one where the same contract takes less
+with a source code of a few kilobytes, to one where the same contract takes less
 than a second to analyse, and where a four-person crowdfunding contract that
@@
 
-\subsection{Overview}
-
 The SBV library supports implementing Haskell functions in a way that
@@
 the use and construction of high level abstractions,
-e.g: custom data-types, list, sets\dots{} Unfortunately, even though
+e.g: custom data-types, list, sets\ldots{} Unfortunately, even though
 high level abstractions aid reasoning about code, they often
@@
 and after\footnote{\url{https://github.com/input-output-hk/marlowe/blob/master/src/Language/Marlowe/Analysis/FSSemanticsFastVerbose.hs} [last visited 19-05-2020]}
-the heavyweight optimisations.
+the heavyweight optimisations. We could not obtain data for some of the experiments because they
+required more RAM than there was available or too much processing time.
 

### 2021-02-02 papers/actus2021/article.tex
draft

%% LyX 2.3.6.2 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[runningheads]{llncs}
\usepackage[latin9]{inputenc}
\usepackage{url}
\usepackage{amsmath}
\usepackage{graphicx}

\makeatletter
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}

\usepackage{listings}
\renewcommand{\lstlistingname}{Listing}

\makeatother

\usepackage{listings}
\renewcommand{\lstlistingname}{Listing}

\begin{document}
\title{Contribution Title\thanks{Supported by organization x.}}
\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and Second
Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
\authorrunning{F. Author et al.}
\institute{Princeton University, Princeton NJ 08544, USA \and Springer Heidelberg,
Tiergartenstr. 17, 69121 Heidelberg, Germany }
\email{lncs@springer.com}
\institute{\url{http://www.springer.com/gp/computer-science/lncs} \and ABC
Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
 }
\email{\{abc,lncs\}@uni-heidelberg.de}
\maketitle
\begin{abstract}
Crypto-loans are unique and innovative financial instruments that
allow trustless P2P lending, which could provide a safe and convenient
source of liquidity for cryptocurrency holders. In this article, we
explore a smart-contract framework for building standardized crypto-loans
with Marlowe DSL and Algorithmic Contract Types Unified Standards
at its core.

\keywords{cardano \and actus \and blockchain \and finance} 
\end{abstract}

\section{Introduction}

\section{Background}

A loan is a form of debt incurred by an individual or other entity.
The lender advances a sum of money to the borrower. In return, the
borrower agrees to a certain set of terms including any finance charges,
interest, repayment date, and other conditions.

Cryptocurrency-backed loans must have collaterals as there is no trust
between party and counterparty. While loan is usually settled in a
stable-coin currency (e.g. USDT/USDC), the typical collateral is denominated
in a cryptocurrency (e.g. BTC). The purpose of this loan is to give
the borrower access to the fiat value of their crypto-funds without
actually selling them for fiat. Borrower pays interest In exchange
for provided liquidity.

Every loan has a positive net payoff (return minus investment) that
is either rendered as a one-time payment (see ZCB) or as a scheduled
payment of the interest. The rate of interest could be fixed throughout
the lifetime of a contract - for example zero-risk bonds have an fixed
interest proportional to inflation rate.

However, in a generic case - the interest rate is variable and depends
on an external factor agreed in advance, while the interest is periodically
updated by observing the state of that factor. Such loans usually
represent an investment in a particular venture or the industry. As
a somewhat fictional example, one could imagine a cryptocurrency miner
who decided to scale their crypto-farm: the loan (in USD) with variable
interest that directly depends on cryptocurrency prices would be more
attractive for a miner because it would directly correlate with miner's
profits. For example, if the price of the cryptocurrency goes down
in a particular month - miners would have to pay lower interest -
one always pays ``fixed share'' of the profits. In a more traditional
setup - the canonical example would be a power plant taking a loan
with interest depending on electricity prices. In both cases, either
cryptocurrency or electricity prices become a driver for the interest
rate.

Although, one can't just take the bare price of the asset and turn
it into a rate. In order to make ``units of measurement'' compatible
- adjustments should be made. Fluctuations of the interest rate driver
are embedded as:

\noindent 
\begin{equation}
\Delta_{r}=capfloor(driver*multiplier+spread-interestRate_{t-1})
\end{equation}

\noindent 
\begin{equation}
interestRate_{t+1}=capfloor(interestRate_{t-1}+\Delta_{r}),
\end{equation}

Where capfloor is a function that limits the range of fluctuation
(thus limiting lender's risk exposure):

\noindent 
\begin{equation}
capfloor(x)=max(min(x,floor),cap)
\end{equation}

Here, the spread parameter loosely represents the difference between
the average prime rate that the lender expects (benchmark yield) and
the rate imposed by the driver - the higher the spread, the higher
the resulting interest rate is. Multiplier effectively rescales the
interest rate plot in order to represent the difference between units
of measurement: how much percentage points you would get for a usd-to-kilowatt
and so on.

In the context of ACTUS and other event-based frameworks, there is
one more factor influencing interest rates thrrough scaling:

\noindent 
\begin{equation}
interestPayment=interestScalingFactor*interstRate*notional
\end{equation}

This scaling is dynamic and loosely adjusts for variance (volatility)
of the asset that the interest rate driver represents.

\paragraph*{Interest accrual and capitalisation. }

Counterparty might decide to reinvest profit received as interest
from the same loan. In the simplest case, this renders as compound
interest (TODO link). This can be modeled through interest accrual
and capitalisation, for instance contracts from the ACTUS specification
accrue interest between interest payments and can transfer interest
to a notional during interest capitalisation event (IPCL).

Overall, variable interest rates introduce a certain risk for a lender,
thus they are subject to hedging. While any instrument that depends
on the same risk factor (interest rate driver) would suffice - the
most popular way to hedge a variable interest rate loan is interest
rate swap. This instrument allows two (or more) parties to exchange
their incomes - one from fixed interest rate loans, the other one
- from variable interest

\paragraph*{Counterparty risk.}

Trustless setups, especially ones in the cryptocurrency world (e.g.
decentralised smart-contracts and exchanges) require zero trust between
party and counterparty involved in a contract. In case of a loan,
this literally means that counterparty has zero-obligation to pay
the money back, thus rendering the loan useless for a party. Such
risks are usually addressed by introducing collaterals: 
\begin{enumerate}
\item Alice would like to borrow 1000 USD 
\item She has Bitcoin assets cost around 1500 USD, which she intends to
hold throughout a year, so Alice has high confidence in the market
(she expects prices to double or triple) 
\item Bob would like to lend 1000 USD and get an interest higher than traditional
interest rate offered by banks (let's say 15\% instead of 10\%). He
is either bearish or neutral towards Bitcoin. 
\item Alice transfers her BTC as collateral to a contract, and Bob transfers
his USD to Alice 
\item If Alice pays interest and notional on time (and Btc-price doesn't
render collateral worthless) - she can get her collateral back, otherwise
the loan gets liquidated and collateral is transferred to Bob. 
\end{enumerate}

\section{Modeling financial contracts}

\subsection{Marlowe }

Marlowe is a high-level, domain-specific language (DSL) for writing
financial contracts on blockchain. Marlowe is defined by an executable
semantics in Haskell, and has been implemented on the UTxO-based Cardano
blockchain.

The main advantage of using Marlowe to carry ACTUS logic is enhanced
security that static analysis {[}TODO link{]} provides. SMT-solvers
gain increasing popularity as an alternative to fuzzing when it comes
to testing software for vulnerabilities {[}TODO link{]}. Smart contracts
generated from ACTUS specification are quite complicated, so manually
testing every possible execution path for every possible combination
of contract terms is not feasible. Static analysis, in turn, can reduce
the amount of ``dumb'' mistakes that could cost millions to the
users by ensuring obvious properties of a contract. It is not a panacea,
however, unlike with fuzzing, if SA says that certain property holds
- it does it with 100\% assurance.

\subsection{ACTUS}

ACTUS defines the logic embedded in legal agreements that eventually
turn the contract terms into actual cash flows, or more generally
business events. Most of Its basic contract types represent different
variations of lending contracts.

ACTUS relies on a state machine formalism in order to describe the
behaviour of a given contract. Every payoff (transfer of funds in
or out of a contract) can be inferred for any given state. Every state
can be inferred from previous events and observed risk factors:

\noindent 
\begin{equation}
payoff_{i}=POF(state_{i})
\end{equation}

\noindent 
\begin{equation}
path=STF_{1}(ct,ev_{1})\circ STF_{2}(ct,ev_{2}).\circ\ldots\circ STF_{i}(ct,ev_{n})
\end{equation}

\noindent 
\begin{equation}
state_{i}=path(INIT(ct)),
\end{equation}

where $ct$ is contract terms, $INIT$ returns initial state, $sched$
returns scheduled events, $STF$ takes ${contractterms,event,state}$
and returns next state, $POF$ returns payoff.

\subsection{Necessity for Oracles}

In order to support variable interest rates and scaling, ACTUS requires
a smart contract to be able to observe the value of a given risk factor
(such as interest rate) at a particular point in time t:

\noindent 
\begin{equation}
riskfactor_{i}{}_{t}=O_{rf}(i,t)
\end{equation}

This is due to the risk factor's state not being known at instantiation
time. In the case of Cardano blockchain, these values are usually
provided through an ``Oracle'' mechanism. Oracle could be a trusted
party providing necessary data or network of parties under consensus
{[}find a quote{]}. From Marlowe DSL perspective, the exact mechanism
that provides external data is less important, as Marlowe abstracts
over IO by requiring particular type of input (``Choice'') protected
with a cryptographic signature (Choice owner), although Plutus implementation
of Marlowe only supports {[}Cardano Elliptic Curves - which ones?{]}
so adding new types of signatures would require some extra work. As
a result, the event of receiving data from Oracle is treated the same
as receiving integer numbers from user input in other languages

\section{Building executable specification of ACTUS for crypto-loans}

\subsection{Mimicking the specification with Haskell code}

The ACTUS standard is specified in terms of {[}link{]} scheduling,
payoff and state transition functions that are polymorphic on event
and contract type. The specification also follows quite specific naming
conventions that are incompatible with Haskell's conventions. The
executable specification follows original ACTUS conventions as close
as possible in order to ease code base maintenance when faced with
updates of original ACTUS spec (link to github).

Using Haskell itself as a DSL for explicitly encoding formulas without
using advanced language idioms also simplifies code generation, but
in case of ACTUS comes at a cost reduced type-safety, handling nullable
values explicitly introduces risk of exceptions. However this risk
is addressed using QuickCheck generators.

\subsection{Utilizing polymorphism to abstract over basic operations}

In order to keep our executable specification independent of the carrier
(whether its smart-contract engine, proof assistant, analytical framework
or even machine learning model) - we abstract over underlying representations
of state variables and arithmetic operations:

\begin{lstlisting}
--Definitions/ContractState.hs 
data ContractStatePoly a b = ContractStatePoly  
{  
tmd     :: b  
, nt    :: a  
, ipnr  :: a  
, ipac  :: a  
, feac  :: a  
, fac   :: a  
, nsc   :: a  
, isc   :: a  
, prf   :: ContractStatus  
, sd    :: b  
, prnxt :: a  
, ipcb  :: a  
} deriving (Show)

--Ops.hs
class ActusOps a where    
   _min :: a -> a -> a
   _max :: a -> a -> a
   _zero :: a
   _one :: a

class ActusNum a where
   (+) :: a -> a -> a
   (-) :: a -> a -> a
   (*) :: a -> a -> a
   (/) :: a -> a -> a

class YearFractionOps a b where
   _y :: DCC -> a -> a -> a -> b

class DateOps a b where
   _lt :: a -> a -> b --returns pseudo-boolean

class RoleSignOps a where
   _r :: ContractRole -> a
\end{lstlisting}

Thus, every formula in the executable spec could either be instantiated
to: 
\begin{itemize}
\item Formula on some ``atomic'' type (like Double, Day) - could be used
to directly compute cash-flows for analytical purposes or precompute
payoffs for smart contracts that don't depend on oracles 
\item Formula representing piece of AST (Marlowe Value, Observation) - could
be used to generate smart contracts that depend on oracles or to generate
code in another language (like Agda) 
\end{itemize}
This approach of abstracting formulas has a limitation of not allowing
conditionals to be expressed in an abstract way (There is no ActusIf
typeclass) - luckily most of conditional expressions in ACTUS specification
don't depend on variable state of a contract - they depend on ContractTerms
that are known in advance during contract generation. This allows
us to dispatch appropriate formulas during generation rather than
execution.

The only exception to this are rare situations where we need to compare
2 state variables and either choose formula1 or 0 depending on comparison
result:

\begin{equation}
formula(st)=\begin{cases}
formula' & var1(st)<var2(st)\\
0
\end{cases}
\end{equation}

We rely on pseudo-boolean ``less than'' function in order to address
that:

\noindent 
\begin{equation}
formula(st)=pseudoLt(var1(s1),var2(st)))*formula'
\end{equation}

\noindent 
\begin{equation}
pseudoLt(a,b)=Cond(a>b,1,0)
\end{equation}

\subsection{Contract terms representation and explicit applicability}

In order to simplify serialisation/deserialisation of contract terms
across Actus related services maintained by Cardano/IOG - we rely
on \textquotedblleft superposed\textquotedblright{} representation
of contract terms - all contract types are represented with the same
type. Although it allows both encoder and decoder to express any ACTUS
contract terms, it also allows for invalid combinations of terms (for
example PRNXT cannot be applied to PAM-contract), which means contracts
require specific validation that is implemented by means of applicability
function:

Applicability: ContractTerms \textbar -\textgreater{} Bool

Actus has a family of applicability functions polymorphic on contract
type:

Applicability: ContractType x ContractTerm \textbar -\textgreater{}
ApplicabilityType Where applicability is either: none, always, nullable,
multiple -{}-TODO describe

In order to build superposed contract terms type for such function,
we have to resolve conflicting applicability for contract terms merged
by following the resolution rules below:

weaken(none, always) = weaken(always, none) = nullable weaken(a1,
a2) = if (priority(a1) \textgreater{} priority(a2)) a1 else a2 Where
priority(always) = 0 priority(none) = 1 priority(nullable) = 2 priority(multiple)
= 3 a1 - is applicability of a given term of first contract type to
be merged a2 - is applicability of a given term of second contract
type to be merged

\section{Generating Marlowe contracts from standardised ACTUS contract terms}

\subsection{Overall architecture }

Generated contract is essentially a continuation chain of Marlowe
Contracts:

chainlink(eventtype, date) = receive\_data(oracle\_data(date)) -\textgreater{}
calculate\_payoff(eventtype, date) -\textgreater{} process\_payoff(date)

where receive\_data - is asking an Oracle for Marlowe Choice if needed
calculate\_payoff - calculates payoff formula. For fixed-rate contracts
this is optimised into a precalculated constant. process\_payoff -
await Deposit of payoff amount from a party: if deposit is transferred
it redirects the funds to a counterparty - otherwise it transfers
the collateral to a counterparty and closes the contract.

The chain is generated from the fixed (known in advance) schedule
of events:

contract(schedule) = receive\_collaterals -\textgreater{} initialise
state -\textgreater{} fold schedule chainlink

\subsection{Avoiding exponential growth }

Marlowe doesn\textquoteright t allow functions and recursion. Naive
usage of If operator in Marlowe could lead to exponential growth of
a contract, for example:

If (condition) then perform\_something1(); continue(); else perform\_something2;
continue()

would inline contents of \textquotedblleft continue()\textquotedblright{}
twice and given that ACTUS contracts are essentially generated using
continuation as an accumulator - this would lead to exponential explosion
of the size of any actus contract that has conditionals in their state
transition logic.

Example of such logic would be cap/floor limitations on interest rate:

adjusted = max(min(original, floor), cap) Where max(a, b) = if (a
\textgreater{} b) then a else b min(a,b) = if (a \textless{} b) then
a else b

We addressed this issue by introducing Cond(link to Marlowe docs)
construct in order to represent \textquotedblleft if\textquotedblright{}
expressions. So instead of using the \textquotedblleft If\textquotedblright{}
contract to decide the value of some variable - we use conditional
assignment instead as Cond is essentially just a pure function that
returns value depending on condition in contrast to the \textquotedblleft If\textquotedblright{}
contract that performs action instead of returning value. The name
is borrowed from a similar function in Lisp.

\subsection{Limitations due to termination }

Marlowe doesn\textquoteright t allow perpetual contracts even if their
recursion is productive (example: perpetual swap) - thus we can\textquoteright t
support certain contract types from ACTUS specification, namely the
ones that don\textquoteright t have a defined maturity date (like
UMP). In a long-term perspective - there is a possible workaround:
contracts with no maturity date could be represented as actors with
finite amounts of state transitions. We prototyped this approach in
a separate branch, however it does seem more prone to errors comparing
to rendering predefined schedules and more importantly it greatly
affects static analysis as the amount of \textquotedblleft reduction
steps\textquotedblright{} grows from N\_scheduled\_event = count(event)
to N\_state\_transition = max(date(event)) - min(date(event)) / precision.
See https://github.com/dk14/plutus/tree/ACTUS

\subsection{Fixed-point precision }

Marlowe only supports Integers, while ACTUS is expressed in terms
of real numbers. In order to model a real number in such a setup -
we rely on fixed point precision. The algebra looks like this: 
\begin{lstlisting}
   (+)                         = AddValue
   (-)                         = SubValue
   a * b                       = Scale (1 % marloweFixedPoint) $ MulValue a b
   (Constant 0) / (Constant 0) = (Constant 0)
   (Constant x) / (Constant y) = Scale (marloweFixedPoint % 1) $ Constant $ div x y
   x / (Constant y)            = Scale (marloweFixedPoint % y) $ x    
   _ / _                       = undefined --division not supported in Marlowe yet 
\end{lstlisting}

\subsection{Representing Actus state in a Marlowe contract}

Unfortunately, Marlowe DSL doesn\textquoteright t support any notion
of records, the variables can only be of type \textquotedblleft Integer\textquotedblright .
Thus, in order to map contract state (ContractStatePoly) we first
pack a set of Marlowe variables (Value Observation) representing previous
t-1 state into ContractStatePoly, then we apply polymorphic state
transition, and finally unpack ContractState poly into a set of Marlowe
variables into t+1 state:

st\_t+1 = unpack(stateTransition(pack(st\_t-1))) , where pack is a
chain of UseValue and unpack is a chain of Let

Representing \textquotedblleft mutations\textquotedblright{} (e.g.
state transitions) in Marlowe (unfolding)

Marlowe doesn\textquoteright t support mutable variables, so we have
to represent state at t (st\_t) literally by naming convention:

variable\_name(name, t) = concat(name, \textquotedblleft \_\textquotedblright ,
t) generate\_variable\_accessor(name, t) = UseValue variable\_name(name,
t) generate\_variable\_setter(name, t, formula) = Let variable\_name(name,
t) formula

\subsection{ActusLabs}

In order to demonstrate and test the capabilities of Actus generators,
a visual Blockly-based tool was developed for Marlowe Playground.
The Actus Labs tool allows users to construct contract terms visually,
generate a corresponding Marlowe contract and try it out in a simulated
environment.

\section{Tokenization}

Every participant of an ACTUS contract is described by a Role which
is in its turn represented through a special unique non-fungible token.
This makes every Marlowe contract a tradable security, allowing a
participant to sell its \textquotedblleft share\textquotedblright{}
in a contract by selling a corresponding role token.

Role tokens can potentially allow more complex manipulation over shares,
especially when share represents an incoming cash flow: participants
only send funds to a party represented by a given token. Such a token
would represent a positive cashflow which in turn could not just become
tradable but would also allow to derive tokens only representing fractions
of a particular cash flow in a contract.

Moreover, it turns Actus-loans into derivatives, for example contracts
like Interest Rate Swap (and Swaps in general) could be approximated
by an Atomic Swap of tokens representing incoming cash flows from
lending.

For example, if Alice has fixed income from a loan (or some other
investment) and Bob has comparable but variable (fluctuating) income
- Bob can hedge by swapping cash flows with Alice. If Alice\textquoteright s
income is locked with token1 and Bob\textquoteright s income is locked
with token2 - then atomic swap of those tokens is equivalent to a
swap of cash flows.

\section{Verification and testing}

\subsection{QuickCheck for cross-testing}

In order to test the equivalence between Executable Haskell implementation
of ACTUS for smart-contracts and original Java implementation for
analytics - a simple property test was introduced:

Forall ct. get\_cashflows\_haskell(ct, riskfactormodel) == get\_cashflows\_java(ct,
riskfactormodel)

Where get\_cashflows returns a set of \{date, payoff\} tuples.

\subsection{QuickCheck for verification}

QuickCheck contract terms generators also allow to ensure arbitrary
properties of a Marlowe contract. This could be enhanced with Marlowe\textquoteright s
static analysis feature by utilising Assert operator:

ContractTerms := sample(qcgenerator) Contract := generateMarloweContract(ContractTerms)
ContractWithAssert := appendAssertion(Contract, assertion) runStaticAnalysis(ContractWithAssert)

While it doesn\textquoteright t cover all possible contracts it could
guarantee that property holds for a statistically significant fraction
of a contract. Using Static Analysis to check a particular contract
instead of randomly sampling allows: 
\begin{itemize}
\item more balanced sampling: Space of contract terms depends linearly on
the coverage, e.g. if we cover 10\% of all possible contract terms
- we\textquoteright ll cover 10\% of all contracts. Different contracts
have different amount of risk factors thus different search space,
for instance a contract with 3 risk factors (e.g. 3 observations of
an interest rate) would span over n\textasciicircum 3 while a contract
with 10 risk factors would span over n\textasciicircum 10. Meanwhile,
space of all risk factors in all contracts doesn\textquoteright t
linearly depend on the coverage. 
\item dependency tracking: SMT-solver would likely be more aware of execution
paths leading to failure of a test, which could significantly reduce
search space 
\item completeness: If not timed out, SMT-solver is decidable while sampling
is semi-decidable. 
\end{itemize}

\subsection{Securing collateral logic with auto-refund warnings}

By design, the Marlowe interpreter always refunds assets held in a
contract when it terminates. This is done in order to ensure that
no funds are lost forever. The funds are returned to whoever\textquoteright s
internal account actually holds them.

However, this presents as a problem in certain cases where ownership
of the funds could not be automatically determined. For example if
Alice puts collateral in a crypto-loan contract - she would formally
maintain ownership, which means she would get automatically refunded
when Close contract is reached.

This implicit refund is easy to overlook by smart-contract developers
as plain Close is often used as a default action in case of unexpected
behaviour like timeouts(especially Choice timeout). This can easily
lead to multi-million dollar mistakes if malicious Alice decides to
exploit an auto-refund feature in order to get her collateral without
paying back. Here\textquoteright s the example of simple attack vector:
\begin{enumerate}
\item Alice creates a contract where Deposit timeout would lead to Close 
\item Bob doesn\textquoteright t know or test the timeout path. Even if
Bob is a programmer - he might not be aware that Closing contract
would cause collateral refund 
\item Alice puts collateral and gets notional from Bob 
\item Alice doesn\textquoteright t pay for a loan (Deposit timeout) 
\item Alice gets her collateral back 
\item Bob loses notional
\end{enumerate}
In order to prevent this from happening, an additional \textquotedblright Auto-Refund\textquotedblright{}
security check was introduced as part of Static Analysis {[}TODO link{]}
tooling which notifies users about all Close constructs that can lead
to automatic refunds and encourages users to write explicit logic
for edge cases.

\section{Comparison with existing smart-contracts approaches }

Unlike current popular DeFi lending approaches (like Harvest Finance)
- Marlowe ACTUS-based implementation tends to rely on \textquotedblleft good
old\textquotedblright{} matching instead of pooling. While pooling
is proven to be superior to order-book based approaches when it comes
to automated market makers, using it for lending has been shown to
be more susceptible to attacks (for example \textquotedblleft flash
loan\textquotedblright{} (https://www.coindesk.com/harvest-finance-24m-attack-triggers-570m-bank-run-in-latest-defi-exploit)).

Moving trade matching off-chain also improves scalability of the protocol
- every loan is a separate contract, thus there is no global state.

Actus provides additional benefit of being regulatory friendly, ACTUS
foundation provides a set of tools allowing Monte-Carlo simulations
of ACTUS contracts.

\section{Conclusion}

Marlowe language was explicitly designed as a set of building blocks
for financial contracts that could be combined by anyone familiar
with basic programming. Marlowe ACTUS generators improve on that by
providing a way to automatically combine blocks based on standardized
requirements specified by the user.

On top of that, Marlowe ACTUS provides a toolkit for cross-testing
with original ACTUS spec, a framework for adding new contract types,
cash-flow visualisations and verification tooling.

\begin{figure}
\includegraphics[width=1\textwidth]{fig1} \caption{A figure caption is always placed below the illustration. Please note
that short captions are centered, while long ones are justified by
the macro package automatically.}
\label{fig1} 
\end{figure}

For citations of references, we prefer the use of square brackets
and consecutive numbers. Citations using labels or the author/year
convention are also acceptable. The following bibliography provides
a sample reference list with entries for journal articles~\cite{ref_article1},
an LNCS chapter~\cite{ref_lncs1}, a book~\cite{ref_book1}, proceedings
without editors~\cite{ref_proc1}, and a homepage~\cite{ref_url1}.
Multiple citations are grouped \cite{ref_article1,ref_lncs1,ref_book1},
\cite{ref_article1,ref_book1,ref_proc1,ref_url1}. % ---- Bibliography ----
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
% \bibliographystyle{splncs04}
% \bibliography{mybibliography}

\begin{thebibliography}{1}
\bibitem{ref_article1} Author, F.: Article title. Journal \textbf{2}(5),
99--110 (2016)

\bibitem{ref_lncs1} Author, F., Author, S.: Title of a proceedings
paper. In: Editor, F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol.
9999, pp. 1--13. Springer, Heidelberg (2016). \doi{10.10007/1234567890}

\bibitem{ref_book1} Author, F., Author, S., Author, T.: Book title.
2nd edn. Publisher, Location (1999)

\bibitem{ref_proc1} Author, A.-B.: Contribution title. In: 9th International
Proceedings on Proceedings, pp. 1--2. Publisher, Location (2010)

\bibitem{ref_url1} LNCS Homepage, \url{http://www.springer.com/lncs}.
Last accessed 4 Oct 2017 
\end{thebibliography}

\end{document}

### 2021-02-02 papers/actus2021/article.tex
more formulas

@@
 formula' & var1(st)<var2(st)\\
-0
+0 & otherwise
 \end{cases}
@@
 
-Applicability: ContractTerms \textbar -\textgreater{} Bool
+\noindent 
+\begin{equation}
+Applicability:ContractTerms\mapsto Bool
+\end{equation}
 
@@
 
-Applicability: ContractType x ContractTerm \textbar -\textgreater{}
-ApplicabilityType Where applicability is either: none, always, nullable,
-multiple -{}-TODO describe
+\noindent 
+\begin{equation}
+Applicability:ContractType\times ContractTerm\mapsto ApplicabilityType
+\end{equation}
+
+Where applicability could be: none, always, nullable, multiple
 
@@
 we have to resolve conflicting applicability for contract terms merged
-by following the resolution rules below:
+using the following resolution rules:
 
-weaken(none, always) = weaken(always, none) = nullable weaken(a1,
-a2) = if (priority(a1) \textgreater{} priority(a2)) a1 else a2 Where
-priority(always) = 0 priority(none) = 1 priority(nullable) = 2 priority(multiple)
-= 3 a1 - is applicability of a given term of first contract type to
-be merged a2 - is applicability of a given term of second contract
-type to be merged
+\noindent 
+\begin{equation}
+weaken(a_{1},a_{2})=\begin{cases}
+nullable & a_{1}=none\wedge a_{2}=always\\
+nullable & a_{1}=always\wedge a_{2}=none\\
+a_{1} & priority(a_{1})>priority(a_{2})\\
+a_{2} & otherwise
+\end{cases}
+\end{equation}
+
+\noindent 
+\begin{equation}
+priority(x)=\begin{cases}
+0 & x=always\\
+1 & x=none\\
+2 & x=nullable\\
+3 & x=multiple
+\end{cases}
+\end{equation}
+
+Where $a_{1}$ - is applicability of a given term of first contract
+type to be merged $a_{2}$ - is applicability of a given term of second
+contract type to be merged
 
@@
 
-Generated contract is essentially a continuation chain of Marlowe
-Contracts:
+Generated contract is essentially a continuation chain of repetition
+Marlowe Contracts:
 
-chainlink(eventtype, date) = receive\_data(oracle\_data(date)) -\textgreater{}
-calculate\_payoff(eventtype, date) -\textgreater{} process\_payoff(date)
+\noindent 
+\begin{equation}
+chainlink(t)=receiveData(t)\circ calculatePayoff(t)\circ processPayoff(t)
+\end{equation}
 
-where receive\_data - is asking an Oracle for Marlowe Choice if needed
-calculate\_payoff - calculates payoff formula. For fixed-rate contracts
+\noindent 
+\begin{equation}
+contract(ct)=receiveCollaterals(ct)\circ INIT(ct)\circ\prod_{t\in SCHED(ct)}chainlink(t)
+\end{equation}
+
+where $receiveData$ - is asking an Oracle for Marlowe Choice if needed,
+$calculatePayoff$ - calculates payoff formula. For fixed-rate contracts
 this is optimised into a precalculated constant. process\_payoff -
@@
 it redirects the funds to a counterparty - otherwise it transfers
-the collateral to a counterparty and closes the contract.
-
-The chain is generated from the fixed (known in advance) schedule
-of events:
-
-contract(schedule) = receive\_collaterals -\textgreater{} initialise
-state -\textgreater{} fold schedule chainlink
+the collateral to a counterparty and closes the contract. The chain
+is generated from the fixed (known in advance) schedule of events
+using $SCHED()$function from ACTUS.
 
@@
 
-If (condition) then perform\_something1(); continue(); else perform\_something2;
+\begin{lstlisting}
+If (condition) 
+	then 
+		perform_something1()
 		continue() 
+	else 
+		perform_something2 
+		continue()
+\end{lstlisting}
 

### 2021-02-02 papers/actus2021/article.tex
more of actus article (#102)

@@
 \usepackage{graphicx}
+\usepackage[unicode=true,
+ bookmarks=false,
+ breaklinks=false,pdfborder={0 0 1},backref=section,colorlinks=false]
+ {hyperref}
 
@@
 
+\usepackage{listings}
+\renewcommand{\lstlistingname}{Listing}
+
+\usepackage{listings}
+\renewcommand{\lstlistingname}{Listing}
+
 \makeatother
@@
 \title{Contribution Title\thanks{Supported by organization x.}}
+
+\maketitle
+%\titlerunning{Abbreviated paper title}
+% If the paper title is too long for the running head, you can set
+% an abbreviated paper title here
+
 \author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and Second
@@
 \authorrunning{F. Author et al.}
+
+% First names are abbreviated in the running head.
+% If there are more than two authors, 'et al.' is used.
+
 \institute{Princeton University, Princeton NJ 08544, USA \and Springer Heidelberg,
@@
 \email{lncs@springer.com}
-\institute{\url{http://www.springer.com/gp/computer-science/lncs} \and ABC
+\institute{\\
+ \url{http://www.springer.com/gp/computer-science/lncs} \and ABC
 Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
@@
 \email{\{abc,lncs\}@uni-heidelberg.de}
-\maketitle
 \begin{abstract}
@@
 \begin{lstlisting}
---Definitions/ContractState.hs 
+-- Definitions/ContractState.hs 
 data ContractStatePoly a b = ContractStatePoly  
@@
 
---Ops.hs
+-- Ops.hs
 class ActusOps a where    
@@
 \begin{itemize}
-\item Formula on some ``atomic'' type (like Double, Day) - could be used
+\item formula on some ``atomic'' type (like Double, Day) - could be used
 to directly compute cash-flows for analytical purposes or precompute
 payoffs for smart contracts that don't depend on oracles 
-\item Formula representing piece of AST (Marlowe Value, Observation) - could
+\item formula representing piece of AST (Marlowe Value, Observation) - could
 be used to generate smart contracts that depend on oracles or to generate
@@
 In order to simplify serialisation/deserialisation of contract terms
-across Actus related services maintained by Cardano/IOG - we rely
-on \textquotedblleft superposed\textquotedblright{} representation
-of contract terms - all contract types are represented with the same
-type. Although it allows both encoder and decoder to express any ACTUS
-contract terms, it also allows for invalid combinations of terms (for
-example PRNXT cannot be applied to PAM-contract), which means contracts
-require specific validation that is implemented by means of applicability
-function:
+across Actus related services maintained by Cardano - we rely on ``superposed''
+representation of contract terms - all ACTUS contract types are represented
+with the same type. While such representation allows both encoder
+and decoder to express any ACTUS contract terms, it also allows for
+invalid combinations of terms (for example PRNXT cannot be applied
+to PAM-contract), which means contracts require specific validation
+that is implemented by means of applicability function:
 
@@
 
-Actus has a family of applicability functions polymorphic on contract
-type:
+ACTUS defines a family of applicability functions polymorphic on contract
+type (TODO link to ACTUS applicability):
 
@@
 \begin{equation}
-Applicability:ContractType\times ContractTerm\mapsto ApplicabilityType
+Applicability:ContractType\times ContractTerm\mapsto ApplicabilityType,
 \end{equation}
 
-Where applicability could be: none, always, nullable, multiple
+where applicability could be: none, always, nullable, multiple
 
@@
 
+\begin{figure}
+\includegraphics[width=1\textwidth]{modules} \caption{Modules responsible for contract generation.}
+\label{fig1-1} 
+\end{figure}
+
 Generated contract is essentially a continuation chain of repetition
@@
 \begin{equation}
-contract(ct)=receiveCollaterals(ct)\circ INIT(ct)\circ\prod_{t\in SCHED(ct)}chainlink(t)
+contract(ct)=collaterals(ct)\circ INIT(ct)\circ\prod_{t\in SCHED(ct)}chainlink(t)
 \end{equation}
@@
 $calculatePayoff$ - calculates payoff formula. For fixed-rate contracts
-this is optimised into a precalculated constant. process\_payoff -
-await Deposit of payoff amount from a party: if deposit is transferred
+this is optimised into a precalculated constant. $processPayoff$
+- awaits Deposit of payoff amount from a party: if deposit is transferred
 it redirects the funds to a counterparty - otherwise it transfers
@@
 
+\begin{figure}
+\includegraphics[width=1\textwidth]{flowchart} \caption{Modules responsible for contract generation.}
+\label{fig1-1-1} 
+\end{figure}
+
 \subsection{Avoiding exponential growth }
 
-Marlowe doesn\textquoteright t allow functions and recursion. Naive
-usage of If operator in Marlowe could lead to exponential growth of
-a contract, for example:
+Marlowe doesn't allow functions and recursion. Naive usage of If operator
+in Marlowe could lead to exponential growth of a contract, for example:
 
 \begin{lstlisting}
-If (condition) 
+if condition 
 	then 
@@
 
-would inline contents of \textquotedblleft continue()\textquotedblright{}
-twice and given that ACTUS contracts are essentially generated using
-continuation as an accumulator - this would lead to exponential explosion
-of the size of any actus contract that has conditionals in their state
-transition logic.
+would inline contents of $continue()$ twice and given that ACTUS
+contracts are essentially generated using continuation as an accumulator
+- this would lead to exponential explosion of the size of any actus
+contract that has conditionals in their state transition logic.
 
@@
 
-adjusted = max(min(original, floor), cap) Where max(a, b) = if (a
-\textgreater{} b) then a else b min(a,b) = if (a \textless{} b) then
-a else b
+\noindent 
+\begin{equation}
+adjusted=max(min(original,floor),cap)
+\end{equation}
 
-We addressed this issue by introducing Cond(link to Marlowe docs)
-construct in order to represent \textquotedblleft if\textquotedblright{}
-expressions. So instead of using the \textquotedblleft If\textquotedblright{}
+We addressed this issue by introducing ``Cond'' construct in order
+to represent ``if'' expressions. So instead of using the ``If''
 contract to decide the value of some variable - we use conditional
-assignment instead as Cond is essentially just a pure function that
-returns value depending on condition in contrast to the \textquotedblleft If\textquotedblright{}
+assignment instead as ``Cond'' is essentially just a pure function
+that returns value depending on condition in contrast to the ``If''
 contract that performs action instead of returning value. The name
@@
 
-Marlowe doesn\textquoteright t allow perpetual contracts even if their
-recursion is productive (example: perpetual swap) - thus we can\textquoteright t
-support certain contract types from ACTUS specification, namely the
-ones that don\textquoteright t have a defined maturity date (like
-UMP). In a long-term perspective - there is a possible workaround:
-contracts with no maturity date could be represented as actors with
-finite amounts of state transitions. We prototyped this approach in
-a separate branch, however it does seem more prone to errors comparing
-to rendering predefined schedules and more importantly it greatly
-affects static analysis as the amount of \textquotedblleft reduction
-steps\textquotedblright{} grows from N\_scheduled\_event = count(event)
-to N\_state\_transition = max(date(event)) - min(date(event)) / precision.
-See https://github.com/dk14/plutus/tree/ACTUS
+Marlowe doesn't allow perpetual contracts even if their recursion
+is productive (example: perpetual swap) - thus we can't support certain
+contract types from ACTUS specification, namely the ones that don't
+have a defined maturity date (like UMP). In a long-term perspective
+- there is a possible workaround: contracts with no maturity date
+could be represented as actors with finite amounts of state transitions.
+We prototyped this approach in a separate branch, however it does
+seem more prone to errors comparing to rendering predefined schedules
+and more importantly it greatly affects static analysis as the amount
+of ``reduction steps'' grows from $N_{scheduled}=count(event)$
+to $N_{stateTransitions}=\frac{max(date(event))-min(date(event))}{precision}$.
+See \href{https://github.com/dk14/plutus/tree/ACTUS}{https://github.com/dk14/plutus/tree/ACTUS}
 
@@
 a * b = Scale (1 % marloweFixedPoint) $ MulValue a b
-   (Constant 0) / (Constant 0) = (Constant 0)
-   (Constant x) / (Constant y) = Scale (marloweFixedPoint % 1) $ Constant $ div x y
-   x / (Constant y)            = Scale (marloweFixedPoint % y) $ x    
-   _ / _                       = undefined --division not supported in Marlowe yet 
 \end{lstlisting}
@@
 
-Unfortunately, Marlowe DSL doesn\textquoteright t support any notion
-of records, the variables can only be of type \textquotedblleft Integer\textquotedblright .
-Thus, in order to map contract state (ContractStatePoly) we first
-pack a set of Marlowe variables (Value Observation) representing previous
-t-1 state into ContractStatePoly, then we apply polymorphic state
-transition, and finally unpack ContractState poly into a set of Marlowe
-variables into t+1 state:
+Unfortunately, Marlowe DSL doesn't support any notion of records,
+the variables can only be of type ``Integer''. Thus, in order to
+map contract state (ContractStatePoly) we first pack a set of Marlowe
+variables (Value Observation) representing previous $t-1$ state into
+ContractStatePoly, then we apply polymorphic state transition, and
+finally unpack ContractState poly into a set of Marlowe variables
+into $t+1$ state:
 
-st\_t+1 = unpack(stateTransition(pack(st\_t-1))) , where pack is a
-chain of UseValue and unpack is a chain of Let
+\noindent 
+\begin{equation}
+st_{t+1}=unpack(stateTransition(pack(st_{t-1}))),
+\end{equation}
 
-Representing \textquotedblleft mutations\textquotedblright{} (e.g.
-state transitions) in Marlowe (unfolding)
+,where pack is a chain of ``UseValue'' and unpack is a chain of
+``Let''
 
-Marlowe doesn\textquoteright t support mutable variables, so we have
-to represent state at t (st\_t) literally by naming convention:
+\paragraph{Representing state transitions in Marlowe (unfolding)}
 
-variable\_name(name, t) = concat(name, \textquotedblleft \_\textquotedblright ,
-t) generate\_variable\_accessor(name, t) = UseValue variable\_name(name,
-t) generate\_variable\_setter(name, t, formula) = Let variable\_name(name,
-t) formula
+Marlowe doesn't support mutable variables, so we have to represent
+state at t ($st_{t}$) literally through naming convention:
+
+\begin{lstlisting}
+variableName(name, t) = 
+	concat(name, '_', t) 
+generateAccessor(name, t) = 
+	UseValue variableName(name, t) 
+generateSetter(name, t, formula) = 
+	Let variableName(name, t) formula 
+\end{lstlisting}
 
@@
 
+\begin{figure} 
+\includegraphics[width=0.8\textwidth,natwidth=664,natheight=552]{labs.png}
+\caption{Actus Labs - online tool for generating Actus contracts for Marlowe.} 
+\label{fig2} 
+\end{figure}
+
 In order to demonstrate and test the capabilities of Actus generators,
-a visual Blockly-based tool was developed for Marlowe Playground.
+a visual online Blockly-based tool was developed for Marlowe Playground.
 The Actus Labs tool allows users to construct contract terms visually,
@@
 This makes every Marlowe contract a tradable security, allowing a
-participant to sell its \textquotedblleft share\textquotedblright{}
-in a contract by selling a corresponding role token.
+participant to sell its ``share'' in a contract by selling a corresponding
+role token.
 
@@
 investment) and Bob has comparable but variable (fluctuating) income
-- Bob can hedge by swapping cash flows with Alice. If Alice\textquoteright s
-income is locked with token1 and Bob\textquoteright s income is locked
-with token2 - then atomic swap of those tokens is equivalent to a
-swap of cash flows.
+- Bob can hedge by swapping cash flows with Alice. If Alice's income
+is locked with token1 and Bob's income is locked with token2 - then
+atomic swap of those tokens is equivalent to a swap of cash flows.
 
@@
 
-Forall ct. get\_cashflows\_haskell(ct, riskfactormodel) == get\_cashflows\_java(ct,
-riskfactormodel)
+\noindent 
+\begin{equation}
+\forall ct\forall rf.getCashFlows("haskell",ct,rf)\equiv getCashFlows("java",ct,rf)
+\end{equation}
 
-Where get\_cashflows returns a set of \{date, payoff\} tuples.
+Where $ct$ represnts contract terms, $rf$ is risk factor model and
+$getCashFlows()$ returns a set of $(date,payoff)$ tuples.
 
@@
 QuickCheck contract terms generators also allow to ensure arbitrary
-properties of a Marlowe contract. This could be enhanced with Marlowe\textquoteright s
+properties of a Marlowe contract. This could be enhanced with Marlowe's
 static analysis feature by utilising Assert operator:
 
-ContractTerms := sample(qcgenerator) Contract := generateMarloweContract(ContractTerms)
-ContractWithAssert := appendAssertion(Contract, assertion) runStaticAnalysis(ContractWithAssert)
+\begin{lstlisting}
+ContractTerms := sample(qcgenerator) 
+Contract := generateMarloweContract(ContractTerms) 
+ContractWithAssert := appendAssertion(Contract, assertion) 
+runStaticAnalysis(ContractWithAssert)
+\end{lstlisting}
 
-While it doesn\textquoteright t cover all possible contracts it could
-guarantee that property holds for a statistically significant fraction
-of a contract. Using Static Analysis to check a particular contract
-instead of randomly sampling allows: 
+While it doesn't cover all possible contracts it could guarantee that
+property holds for a statistically significant fraction of a contract.
+Using Static Analysis to check a particular contract instead of randomly
+sampling allows: 
 \begin{itemize}
@@
 the coverage, e.g. if we cover 10\% of all possible contract terms
-- we\textquoteright ll cover 10\% of all contracts. Different contracts
-have different amount of risk factors thus different search space,
-for instance a contract with 3 risk factors (e.g. 3 observations of
-an interest rate) would span over n\textasciicircum 3 while a contract
-with 10 risk factors would span over n\textasciicircum 10. Meanwhile,
-space of all risk factors in all contracts doesn\textquoteright t
-linearly depend on the coverage. 
+- we'll cover 10\% of all contracts. Different contracts have different
+amount of risk factors thus different search space, for instance a
+contract with 3 risk factors (e.g. 3 observations of an interest rate)
+would span over n\textasciicircum 3 while a contract with 10 risk
+factors would span over n\textasciicircum 10. Meanwhile, space of
+all risk factors in all contracts doesn't linearly depend on the coverage. 
 \item dependency tracking: SMT-solver would likely be more aware of execution
@@
 contract when it terminates. This is done in order to ensure that
-no funds are lost forever. The funds are returned to whoever\textquoteright s
-internal account actually holds them.
+no funds are lost forever. The funds are returned to whoever's internal
+account actually holds them.
 
@@
 exploit an auto-refund feature in order to get her collateral without
-paying back. Here\textquoteright s the example of simple attack vector:
+paying back. Here's the example of simple attack vector: 
 \begin{enumerate}
 \item Alice creates a contract where Deposit timeout would lead to Close 
-\item Bob doesn\textquoteright t know or test the timeout path. Even if
-Bob is a programmer - he might not be aware that Closing contract
-would cause collateral refund 
+\item Bob doesn't know or test the timeout path. Even if Bob is a programmer
+- he might not be aware that Closing contract would cause collateral
+refund 
 \item Alice puts collateral and gets notional from Bob 
-\item Alice doesn\textquoteright t pay for a loan (Deposit timeout) 
+\item Alice doesn't pay for a loan (Deposit timeout) 
 \item Alice gets her collateral back 
@@
 \end{enumerate}
-In order to prevent this from happening, an additional \textquotedblright Auto-Refund\textquotedblright{}
+In order to prevent this from happening, an additional ''Auto-Refund''
 security check was introduced as part of Static Analysis {[}TODO link{]}
-tooling which notifies users about all Close constructs that can lead
-to automatic refunds and encourages users to write explicit logic
-for edge cases.
+tooling which notifies users about all ``Close'' constructs that
+can lead to automatic refunds and encourages users to write explicit
+logic for edge cases.
 
@@
 Unlike current popular DeFi lending approaches (like Harvest Finance)
-- Marlowe ACTUS-based implementation tends to rely on \textquotedblleft good
-old\textquotedblright{} matching instead of pooling. While pooling
-is proven to be superior to order-book based approaches when it comes
-to automated market makers, using it for lending has been shown to
-be more susceptible to attacks (for example \textquotedblleft flash
-loan\textquotedblright{} (https://www.coindesk.com/harvest-finance-24m-attack-triggers-570m-bank-run-in-latest-defi-exploit)).
+- Marlowe ACTUS-based implementation tends to rely on ``good old''
+matching instead of pooling. While pooling is proven to be superior
+to order-book based approaches when it comes to automated market makers,
+using it for lending has been shown to be more susceptible to attacks
+(for example ``\href{https://www.coindesk.com/harvest-finance-24m-attack-triggers-570m-bank-run-in-latest-defi-exploit}{flash loan}'').
 
@@
 
-\begin{figure}
-\includegraphics[width=1\textwidth]{fig1} \caption{A figure caption is always placed below the illustration. Please note
-that short captions are centered, while long ones are justified by
-the macro package automatically.}
-\label{fig1} 
-\end{figure}
+\section*{TODO}
 

### 2021-02-02 papers/actus2021/article.tex
more

@@
 \begin{document}
-\title{Contribution Title\thanks{Supported by organization x.}}
+\title{Standardized crypto-loans on Cardano blockchain\textquotedblright \thanks{Supported by organization x.}}
 
@@
 \section{Introduction}
+
+List of contributions:
+\begin{itemize}
+\item \textquotedblleft Marlowe Labs\textquotedblright{} editor on Playground
+(Dmytro) 
+\item R Shiny cashflow visualiser for ACTUS (Dmytro) 
+\item ACTUS Utility functions (Dmytro, Quanterall) 
+\item executable ACTUS spec for PAM (Dmytro, Alex) 
+\item executable ACTUS spec for LAM (Wyo Hackathon) 
+\item Library for generating ACTUS contracts (Dmytro) 
+\item ACTUS-like examples on Playground (Dmytro, Alex, Pablo) 
+\item Marlowe DSL enhancements for Actus: Cond, Mul (Dmytro) (draft) 
+\item ACTUS spec generator for Agda (Dmytro) 
+\item Auto-refund static analysis (Pablo) (partial: only QC) 
+\item Verification with QC + SMT (Dmytro) 
+\end{itemize}
 
@@
 over IO by requiring particular type of input (``Choice'') protected
-with a cryptographic signature (Choice owner), although Plutus implementation
-of Marlowe only supports {[}Cardano Elliptic Curves - which ones?{]}
-so adding new types of signatures would require some extra work. As
-a result, the event of receiving data from Oracle is treated the same
-as receiving integer numbers from user input in other languages
+with a cryptographic signature (Choice owner). As a result, the event
+of receiving data from Oracle is treated the same as receiving integer
+numbers from user input in other languages
 
-\section{Building executable specification of ACTUS for crypto-loans}
+\section{Building executable specification of ACTUS}
 
@@
 formula(st)=\begin{cases}
-formula' & var1(st)<var2(st)\\
+formula'(st) & var1(st)<var2(st)\\
 0 & otherwise
@@
 
-Generated contract is essentially a continuation chain of repetition
-Marlowe Contracts:
+Generated contract is essentially a continuation chain of smaller
+contracts:
 
@@
 \begin{figure}
-\includegraphics[width=1\textwidth]{flowchart} \caption{Modules responsible for contract generation.}
+\includegraphics[width=1\textwidth]{flowchart} \caption{Chain of sub-contracts representing ACTUS logic}
 \label{fig1-1-1} 

### 2021-02-03 papers/actus2021/article.tex
Actus article (#103)
Move biblography to .bib file, add Makefile and move images to subfolder

@@
-%% LyX 2.3.6.2 created this file.  For more info, see http://www.lyx.org/.
-%% Do not edit unless you really know what you are doing.
 \documentclass[runningheads]{llncs}
@@
 \begin{document}
-\title{Standardized crypto-loans on Cardano blockchain\textquotedblright \thanks{Supported by organization x.}}
+\title{Standardized crypto-loans on\\ the Cardano blockchain}%\thanks{Supported by Cardano Foundation.}}
 
-\maketitle
-%\titlerunning{Abbreviated paper title}
-% If the paper title is too long for the running head, you can set
-% an abbreviated paper title here
+\author{Dmytro Kondratiuk\inst{1}\orcidID{XXXX} \and
+				Pablo {Lamela Seijas}\inst{1}\orcidID{0000-0002-1730-1219} \and
+                Simon Thompson\inst{1,2}\orcidID{0000-0002-2350-301X}}
 
-\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and Second
-Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
-\authorrunning{F. Author et al.}
+\authorrunning{D. Kondratiuk, et al.}
 
-% First names are abbreviated in the running head.
-% If there are more than two authors, 'et al.' is used.
+\institute{IOHK, Hong Kong,
+\email{dmytro.kondratiuk@iohk.io, pablo.lamela@iohk.io, simon.thompson@iohk.io}  \and
+School of Computing, University of Kent, UK,
+\email{s.j.thompson@kent.ac.uk}}
 
-\institute{Princeton University, Princeton NJ 08544, USA \and Springer Heidelberg,
-Tiergartenstr. 17, 69121 Heidelberg, Germany }
-\email{lncs@springer.com}
-\institute{\\
- \url{http://www.springer.com/gp/computer-science/lncs} \and ABC
-Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
- }
-\email{\{abc,lncs\}@uni-heidelberg.de}
+%
+\maketitle              % typeset the header of the contribution
+%
 \begin{abstract}
@@
 
-\keywords{cardano \and actus \and blockchain \and finance} 
+\keywords{Cardano \and ACTUS \and blockchain \and finance \and Marlowe \and Haskell} 
 \end{abstract}
@@
 borrower agrees to a certain set of terms including any finance charges,
-interest, repayment date, and other conditions.
+interest, repayment date, and other conditions \cite{loan}.
 
@@
 from the same loan. In the simplest case, this renders as compound
-interest (TODO link). This can be modeled through interest accrual
-and capitalisation, for instance contracts from the ACTUS specification
-accrue interest between interest payments and can transfer interest
-to a notional during interest capitalisation event (IPCL).
+interest. This can be modeled through interest accrual and capitalisation,
+for instance contracts from the ACTUS specification accrue interest
+between interest payments and can transfer interest to a notional
+during interest capitalisation event (IPCL).
 
@@
 Marlowe is a high-level, domain-specific language (DSL) for writing
-financial contracts on blockchain. Marlowe is defined by an executable
-semantics in Haskell, and has been implemented on the UTxO-based Cardano
-blockchain.
+financial contracts on blockchain\cite{marlowe}. Marlowe is defined
+by an executable semantics in Haskell, and has been implemented on
+the UTxO-based Cardano blockchain.
 
 The main advantage of using Marlowe to carry ACTUS logic is enhanced
-security that static analysis {[}TODO link{]} provides. SMT-solvers
-gain increasing popularity as an alternative to fuzzing when it comes
-to testing software for vulnerabilities {[}TODO link{]}. Smart contracts
-generated from ACTUS specification are quite complicated, so manually
-testing every possible execution path for every possible combination
-of contract terms is not feasible. Static analysis, in turn, can reduce
-the amount of ``dumb'' mistakes that could cost millions to the
-users by ensuring obvious properties of a contract. It is not a panacea,
-however, unlike with fuzzing, if SA says that certain property holds
-- it does it with 100\% assurance.
+security that static analysis provides. SMT-solvers gain increasing
+popularity as an alternative to fuzzing when it comes to testing software
+for vulnerabilities\cite{smt}. Smart contracts generated from ACTUS
+specification are quite complicated, so manually testing every possible
+execution path for every possible combination of contract terms is
+not feasible. Static analysis, in turn, can reduce the amount of ``dumb''
+mistakes that could cost millions to the users by ensuring obvious
+properties of a contract. It is not a panacea, however, unlike with
+fuzzing, if SA says that certain property holds - it does it with
+100\% assurance.
 
@@
 turn the contract terms into actual cash flows, or more generally
-business events. Most of Its basic contract types represent different
-variations of lending contracts.
+business events \cite{actus}. Most of its basic contract types represent
+different variations of lending contracts.
 
@@
 time. In the case of Cardano blockchain, these values are usually
-provided through an ``Oracle'' mechanism. Oracle could be a trusted
-party providing necessary data or network of parties under consensus
-{[}find a quote{]}. From Marlowe DSL perspective, the exact mechanism
-that provides external data is less important, as Marlowe abstracts
-over IO by requiring particular type of input (``Choice'') protected
-with a cryptographic signature (Choice owner). As a result, the event
-of receiving data from Oracle is treated the same as receiving integer
-numbers from user input in other languages
+provided through an ``Oracle'' mechanism\cite{oracles}. Oracle
+could be a trusted party providing necessary data or network of parties
+under consensus {[}find a quote{]}. From Marlowe DSL perspective,
+the exact mechanism that provides external data is less important,
+as Marlowe abstracts over IO by requiring particular type of input
+(``Choice'') protected with a cryptographic signature (Choice owner).
+As a result, the event of receiving data from Oracle is treated the
+same as receiving integer numbers from user input in other languages
 
@@
 
-The ACTUS standard is specified in terms of {[}link{]} scheduling,
-payoff and state transition functions that are polymorphic on event
-and contract type. The specification also follows quite specific naming
-conventions that are incompatible with Haskell's conventions. The
-executable specification follows original ACTUS conventions as close
-as possible in order to ease code base maintenance when faced with
-updates of original ACTUS spec (link to github).
+The ACTUS standard is specified in terms of scheduling, payoff and
+state transition functions that are polymorphic on event and contract
+type. The specification also follows quite specific naming conventions
+that are incompatible with Haskell's conventions. The executable specification
+follows original ACTUS conventions as close as possible in order to
+ease code base maintenance when faced with updates of original ACTUS
+spec (link to github).
 
@@
 
-\begin{lstlisting}
+\begin{verbatim}
 -- Definitions/ContractState.hs 
@@
    _r :: ContractRole -> a
-\end{lstlisting}
+\end{verbatim}
 
@@
 \begin{figure}
-\includegraphics[width=1\textwidth]{modules} \caption{Modules responsible for contract generation.}
-\label{fig1-1} 
+\includegraphics[width=1\textwidth]{images/modules} \caption{Modules responsible for contract generation.}
+\label{fig1} 
 \end{figure}
@@
 
-where $receiveData$ - is asking an Oracle for Marlowe Choice if needed,
+where $receiveData$ - is asking an oracle for Marlowe Choice if needed,
 $calculatePayoff$ - calculates payoff formula. For fixed-rate contracts
@@
 \begin{figure}
-\includegraphics[width=1\textwidth]{flowchart} \caption{Chain of sub-contracts representing ACTUS logic}
-\label{fig1-1-1} 
+\includegraphics[width=1\textwidth]{images/flowchart} \caption{Chain of sub-contracts representing ACTUS logic}
+\label{fig2} 
 \end{figure}
@@
 
-\begin{lstlisting}
+\begin{verbatim}
 if condition 
@@
 		continue()
-\end{lstlisting}
+\end{verbatim}
 
@@
 to $N_{stateTransitions}=\frac{max(date(event))-min(date(event))}{precision}$.
-See \href{https://github.com/dk14/plutus/tree/ACTUS}{https://github.com/dk14/plutus/tree/ACTUS}
 
@@
 we rely on fixed point precision. The algebra looks like this: 
-\begin{lstlisting}
+\begin{verbatim}
 (+)   = AddValue
@@
 a * b = Scale (1 % marloweFixedPoint) $ MulValue a b
-\end{lstlisting}
+\end{verbatim}
 
@@
 
-\begin{lstlisting}
+\begin{verbatim}
 variableName(name, t) = 
@@
 	Let variableName(name, t) formula 
-\end{lstlisting}
+\end{verbatim}
 
@@
 \begin{figure} 
-\includegraphics[width=0.8\textwidth,natwidth=664,natheight=552]{labs.png}
+\includegraphics[width=0.8\textwidth]{images/labs.png}
 \caption{Actus Labs - online tool for generating Actus contracts for Marlowe.} 
-\label{fig2} 
+\label{fig3} 
 \end{figure}
@@
 
-QuickCheck contract terms generators also allow to ensure arbitrary
-properties of a Marlowe contract. This could be enhanced with Marlowe's
-static analysis feature by utilising Assert operator:
+QuickCheck\cite{qc} contract terms generators also allow to ensure
+arbitrary properties of a Marlowe contract. This could be enhanced
+with Marlowe's static analysis feature by utilising Assert operator:
 
-\begin{lstlisting}
+\begin{verbatim}
 ContractTerms := sample(qcgenerator) 
@@
 runStaticAnalysis(ContractWithAssert)
-\end{lstlisting}
+\end{verbatim}
 
@@
 In order to prevent this from happening, an additional ''Auto-Refund''
-security check was introduced as part of Static Analysis {[}TODO link{]}
-tooling which notifies users about all ``Close'' constructs that
-can lead to automatic refunds and encourages users to write explicit
-logic for edge cases.
+security check was introduced as part of Static Analysis tooling which
+notifies users about all ``Close'' constructs that can lead to automatic
+refunds and encourages users to write explicit logic for edge cases.
 
@@
 
-Unlike current popular DeFi lending approaches (like Harvest Finance)
-- Marlowe ACTUS-based implementation tends to rely on ``good old''
-matching instead of pooling. While pooling is proven to be superior
-to order-book based approaches when it comes to automated market makers,
-using it for lending has been shown to be more susceptible to attacks
-(for example ``\href{https://www.coindesk.com/harvest-finance-24m-attack-triggers-570m-bank-run-in-latest-defi-exploit}{flash loan}'').
+Unlike current popular DeFi lending approaches - Marlowe ACTUS-based
+implementation tends to rely on trade-matching instead of pooling.
+While asset pooling is proven to be superior to order-book based approaches
+when it comes to automated market makers, using it for lending has
+been shown to be more susceptible to attacks\cite{flash-loan}.
 
@@
 
-\section*{TODO}
-
-For citations of references, we prefer the use of square brackets
-and consecutive numbers. Citations using labels or the author/year
-convention are also acceptable. The following bibliography provides
-a sample reference list with entries for journal articles~\cite{ref_article1},
-an LNCS chapter~\cite{ref_lncs1}, a book~\cite{ref_book1}, proceedings
-without editors~\cite{ref_proc1}, and a homepage~\cite{ref_url1}.
-Multiple citations are grouped \cite{ref_article1,ref_lncs1,ref_book1},
-\cite{ref_article1,ref_book1,ref_proc1,ref_url1}. % ---- Bibliography ----
+%
+% ---- Bibliography ----
+%
 % BibTeX users should specify bibliography style 'splncs04'.
 % References will then be sorted and formatted in the correct style.
-% \bibliographystyle{splncs04}
-% \bibliography{mybibliography}
-
-\begin{thebibliography}{1}
-\bibitem{ref_article1} Author, F.: Article title. Journal \textbf{2}(5),
-99--110 (2016)
-
-\bibitem{ref_lncs1} Author, F., Author, S.: Title of a proceedings
-paper. In: Editor, F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol.
-9999, pp. 1--13. Springer, Heidelberg (2016). \doi{10.10007/1234567890}
-
-\bibitem{ref_book1} Author, F., Author, S., Author, T.: Book title.
-2nd edn. Publisher, Location (1999)
-
-\bibitem{ref_proc1} Author, A.-B.: Contribution title. In: 9th International
-Proceedings on Proceedings, pp. 1--2. Publisher, Location (2010)
-
-\bibitem{ref_url1} LNCS Homepage, \url{http://www.springer.com/lncs}.
-Last accessed 4 Oct 2017 
-\end{thebibliography}
+%
+\bibliographystyle{splncs04}
+\bibliography{article}
+%
 

### 2021-02-03 papers/actus2021/article.tex
Tidying up: some formatting of identifiers in math mode; spellings.

@@
 
-\keywords{Cardano \and ACTUS \and blockchain \and finance \and Marlowe \and Haskell} 
+\keywords{ACTUS \and blockchain \and Cardano \and finance \and Haskell \and Marlowe} 
 \end{abstract}
@@
 In the context of ACTUS and other event-based frameworks, there is
-one more factor influencing interest rates thrrough scaling:
+one more factor influencing interest rates thorough scaling:
 
@@
 \begin{equation}
-interestPayment=interestScalingFactor*interstRate*notional
+interestPayment=interestScalingFactor*interestRate*notional
 \end{equation}
@@
 \begin{equation}
-payoff_{i}=POF(state_{i})
+\mbox{\emph{payoff}}_{i}=POF(state_{i})
 \end{equation}
@@
 \begin{equation}
-riskfactor_{i}{}_{t}=O_{rf}(i,t)
+\mbox{\emph{riskfactor}}_{i}{}_{t}=O_{rf}(i,t)
 \end{equation}
@@
 
-\section{Building executable specification of ACTUS}
+\section{Building an executable specification of ACTUS}
 
@@
 $calculatePayoff$ - calculates payoff formula. For fixed-rate contracts
-this is optimised into a precalculated constant. $processPayoff$
+this is optimised into a pre-calculated constant. $processPayoff$
 - awaits Deposit of payoff amount from a party: if deposit is transferred
@@
 
+\smallskip
 \noindent 
@@
 
-Where $ct$ represnts contract terms, $rf$ is risk factor model and
+Where $ct$ represents contract terms, $rf$ is risk factor model and
 $getCashFlows()$ returns a set of $(date,payoff)$ tuples.

### 2021-02-03 papers/actus2021/article.tex
bib, formatting, overview, edited most of section 2.

@@
 \usepackage[latin9]{inputenc}
-\usepackage{url}
+\usepackage{url}[obeyspaces]
 \usepackage{amsmath}
@@
 % \renewcommand\UrlFont{\color{blue}\rmfamily}
+
+% SJT use this to get nicer formatting of textual identifiers in math mode
+\newcommand{\ident}[1]{\mbox{\emph{#1}}}
 
@@
 \begin{abstract}
-Crypto-loans are unique and innovative financial instruments that
-allow trustless P2P lending, which could provide a safe and convenient
-source of liquidity for cryptocurrency holders. In this article, we
-explore a smart-contract framework for building standardized crypto-loans
-with Marlowe DSL and Algorithmic Contract Types Unified Standards
-at its core.
+Crypto-loans are %unique and 
+innovative financial instruments that
+allow trustless peer-to-peer %P2P 
+lending, and potentially providing a safe and convenient
+source of liquidity for cryptocurrency holders. In this paper we
+explore a smart contract framework for building standardized crypto-loans
+using the Marlowe domain-specific language and with the ACTUS standard at its core.
 
@@
 
+\textbf{DONE:}
+\begin{enumerate}
+\item
+Added overview at the end of the intro.
+\item
+Tidying up identifiers in math equations using the command \verb+\ident{...}+
+\item
+Biblio sorted out ordering (added keys). shortened author and editor lists
+\end{enumerate}
+
+\textbf{TO DO:}
+\begin{enumerate}
+\item
+Algorithmic Contract Types Unified Standards (ACTUS) -- for first use. Also reference.
+\item
+Do we include Alex as co-author?
+\item
+What about R (shiny)?
+\end{enumerate}
+
 \section{Introduction}
 
+\bigskip\noindent
 List of contributions:
@@
 
-\section{Background}
+% SJT: added this overview of the paper
+% it is traditional to end the introduction like this,
+% 'signposting' what is in the rest of the paper.
+Section \ref{background} of the paper covers the relevant financial background, and Section \ref{modelling} introduces Marlowe as well as the ACTUS financial standard.
+Section \ref{executable} builds an executable specification of ACTUS in Haskell, and this is used in Section \ref{generation} to generate  the Marlowe code for an ACTUS contract from the terms of the contract.
+Section \ref{tokenization} explains how tokens are used to represent ownership of roles in a running contract, and Section \ref{verification} presents how we provide assurance of the contract correctness.
+Section \ref{related} examines related work and Section \ref{conclusion} concludes.
+
+\section{Financial contracts}
+\label{background}
 
@@
 
-Cryptocurrency-backed loans must have collaterals as there is no trust
-between party and counterparty. While loan is usually settled in a
-stable-coin currency (e.g. USDT/USDC), the typical collateral is denominated
-in a cryptocurrency (e.g. BTC). The purpose of this loan is to give
+Cryptocurrency-backed loans must have \emph{collateral} when there is no trust
+between party and counterparty. While a loan is usually settled in a
+stable-coin currency (e.g. USDT/USDC), collateral is typically denominated
+in a cryptocurrency (e.g. BTC). The purpose of such a loan is to give
 the borrower access to the fiat value of their crypto-funds without
-actually selling them for fiat. Borrower pays interest In exchange
-for provided liquidity.
+actually selling them for fiat. The borrower pays interest in exchange
+for gaining liquidity.
 
 Every loan has a positive net payoff (return minus investment) that
-is either rendered as a one-time payment (see ZCB) or as a scheduled
+is either rendered as a one-time payment -- often called a \emph{zero-coupon bond} (ZCB) -- or by scheduling
 payment of the interest. The rate of interest could be fixed throughout
-the lifetime of a contract - for example zero-risk bonds have an fixed
-interest proportional to inflation rate.
+the lifetime of a contract: for example, zero-risk bonds have a fixed
+interest proportional to the inflation rate.
+However, in the generic case the interest rate is variable and depends
+on an external factor agreed in advance, and the rate is periodically
+updated by observing the state of that factor.
 
-However, in a generic case - the interest rate is variable and depends
-on an external factor agreed in advance, while the interest is periodically
-updated by observing the state of that factor. Such loans usually
-represent an investment in a particular venture or the industry. As
+Such loans often represent an investment in a particular venture or industry. As
 a somewhat fictional example, one could imagine a cryptocurrency miner
-who decided to scale their crypto-farm: the loan (in USD) with variable
+who decided to scale their crypto-farm: a loan (in USD) with variable
 interest that directly depends on cryptocurrency prices would be more
@@
 profits. For example, if the price of the cryptocurrency goes down
-in a particular month - miners would have to pay lower interest -
-one always pays ``fixed share'' of the profits. In a more traditional
-setup - the canonical example would be a power plant taking a loan
-with interest depending on electricity prices. In both cases, either
-cryptocurrency or electricity prices become a driver for the interest
+in a particular month the borrower would have to pay lower interest, and so would always pay a fixed share of the profits. In a more traditional
+setup the interest rate could depend on prices of other commodities: the canonical example would be a power plant taking a loan
+with interest depending on electricity prices.
+
+In both cases, the prices of
+cryptocurrency or electricity become a driver for the interest
 rate.
-
-Although, one can't just take the bare price of the asset and turn
-it into a rate. In order to make ``units of measurement'' compatible
-- adjustments should be made. Fluctuations of the interest rate driver
-are embedded as:
+However, one cannot simply take the bare price of the asset and turn
+it into a rate. In order to make \emph{units of measurement} compatible
+%SJT compatible with what??
+adjustments should be made. Fluctuations of the interest rate driver
+are embedded thus:
 
@@
 \begin{equation}
-\Delta_{r}=capfloor(driver*multiplier+spread-interestRate_{t-1})
+\Delta_{r}=\ident{capfloor}(\ident{driver}*\ident{multiplier}+\ident{spread}-\ident{interestRate}_{t-1})
 \end{equation}
@@
 \begin{equation}
-interestRate_{t+1}=capfloor(interestRate_{t-1}+\Delta_{r}),
+\ident{interestRate}_{t+1}=\ident{capfloor}(\ident{interestRate}_{t-1}+\Delta_{r}),
 \end{equation}
 
-Where capfloor is a function that limits the range of fluctuation
-(thus limiting lender's risk exposure):
+\noindent
+where \emph{capfloor} is a function that limits the range of fluctuation,  
+and so limiting the lender's exposure to risk:
 
@@
 \begin{equation}
-capfloor(x)=max(min(x,floor),cap)
+\ident{capfloor}(x)=\ident{max(min}(x,\ident{floor}),\ident{cap})
 \end{equation}
 
-Here, the spread parameter loosely represents the difference between
-the average prime rate that the lender expects (benchmark yield) and
+\noindent
+The \emph{spread} parameter here loosely represents the difference between
+the average prime 
+% SJT explain "prime" here.
+rate that the lender expects (the benchmark yield) and
 the rate imposed by the driver - the higher the spread, the higher
-the resulting interest rate is. Multiplier effectively rescales the
-interest rate plot in order to represent the difference between units
-of measurement: how much percentage points you would get for a usd-to-kilowatt
+the resulting interest rate. The multiplier %effectively 
+rescales the interest rate value 
+% SJT used value here rather than plot: is that OK?
+in order to represent the changes to be made converting between different units
+of measurement: how many percentage points you would get for a usd-to-kilowatt conversion
 and so on.
@@
 from the same loan. In the simplest case, this renders as compound
-interest. This can be modeled through interest accrual and capitalisation,
+interest. This can be modelled through interest accrual and capitalisation,
 for instance contracts from the ACTUS specification accrue interest
@@
 
-\section{Modeling financial contracts}
+\section{Modelling financial contracts}
+\label{modelling}
 
-\subsection{Marlowe }
+This section introduces the Marlowe DSL and the ACTUS standard for financial contracts.
+
+\subsection{Marlowe}
 
@@
 \section{Building an executable specification of ACTUS}
+\label{executable}
 
@@
 
-\section{Generating Marlowe contracts from standardised ACTUS contract terms}
+\section{Generating Marlowe contracts from ACTUS specs}%standardised ACTUS contract terms}
+\label{generation}
 
@@
 \begin{equation}
-chainlink(t)=receiveData(t)\circ calculatePayoff(t)\circ processPayoff(t)
+chainlink(t)=receiveData(t)\circ \mbox{\emph{calculatePayoff(t)}}\circ \mbox{\emph{processPayoff(t)}}
 \end{equation}
@@
 where $receiveData$ - is asking an oracle for Marlowe Choice if needed,
-$calculatePayoff$ - calculates payoff formula. For fixed-rate contracts
-this is optimised into a pre-calculated constant. $processPayoff$
+$\mbox{\emph{calculatePayoff}}$ - calculates payoff formula. For fixed-rate contracts
+this is optimised into a pre-calculated constant. $\mbox{\emph{processPayoff}}$
 - awaits Deposit of payoff amount from a party: if deposit is transferred
@@
 is generated from the fixed (known in advance) schedule of events
-using $SCHED()$function from ACTUS.
+using the $SCHED()$ function from ACTUS.
 
@@
 \end{verbatim}
-
-would inline contents of $continue()$ twice and given that ACTUS
+\noindent
+would inline contents of \texttt{continue()} twice and given that ACTUS
 contracts are essentially generated using continuation as an accumulator
-- this would lead to exponential explosion of the size of any actus
+- this would lead to exponential explosion of the size of any ACTUS
 contract that has conditionals in their state transition logic.
@@
 
-We addressed this issue by introducing ``Cond'' construct in order
-to represent ``if'' expressions. So instead of using the ``If''
+We addressed this issue by introducing the \texttt{Cond} construct in order
+to represent conditional expressions. So instead of using the \texttt{If}
 contract to decide the value of some variable - we use conditional
-assignment instead as ``Cond'' is essentially just a pure function
-that returns value depending on condition in contrast to the ``If''
+assignment instead as \texttt{Cond} is essentially just a pure function
+that returns value depending on condition in contrast to the \texttt{If}
 contract that performs action instead of returning value. The name
@@
 
-\subsection{Limitations due to termination }
+\subsection{Limitations due to termination}
 
@@
 and more importantly it greatly affects static analysis as the amount
-of ``reduction steps'' grows from $N_{scheduled}=count(event)$
-to $N_{stateTransitions}=\frac{max(date(event))-min(date(event))}{precision}$.
+of ``reduction steps'' grows from 
+\noindent 
+\begin{equation}
+N_{scheduled}=count(event)
+\end{equation}
+to 
+\noindent 
+\begin{equation}
+N_{stateTransitions}=\frac{max(date(event))-min(date(event))}{precision} 
+\end{equation}
 
@@
 \section{Tokenization}
+\label{tokenization}
 
@@
 \section{Verification and testing}
+\label{verification}
 
@@
 
-\section{Comparison with existing smart-contracts approaches }
+\section{Related work}\label{related}
+
+% SJT changed title from {Comparison with existing smart-contracts approaches }
+% it's traditional to call this "relatyed work".
 
@@
 \section{Conclusion}
+\label{conclusion}
 

### 2021-02-03 papers/actus2021/article.tex
revised section 2. Some questions to answer: beginning % SJT ...

@@
 
+% SJT a couple of questions about these equations
+% Q1
+% What is the role of the subscript r in \Delta_r
+% Q2
+% interest rate at t+1 defined from rate at t-1. I'd have expected just one difference, so
+% t+1 defined from t, or t defined from t-1.
+
 \noindent 
@@
 % SJT explain "prime" here.
-rate that the lender expects (the benchmark yield) and
+rate that the lender expects -- the \emph{benchmark yield} -- and
 the rate imposed by the driver - the higher the spread, the higher
@@
 in order to represent the changes to be made converting between different units
-of measurement: how many percentage points you would get for a usd-to-kilowatt conversion
+of measurement: how many percentage points 
+% SJT percentage points of what?
+you would get for a USD-to-kilowatt conversion
 and so on.
 
-In the context of ACTUS and other event-based frameworks, there is
+In the context of ACTUS and other event-based frameworks, 
+% SJT - need to explain "event-based" in this context.
+there is
 one more factor influencing interest rates thorough scaling:
@@
 \end{equation}
-
+\noindent
 This scaling is dynamic and loosely adjusts for variance (volatility)
@@
 
-Counterparty might decide to reinvest profit received as interest
-from the same loan. In the simplest case, this renders as compound
-interest. This can be modelled through interest accrual and capitalisation,
-for instance contracts from the ACTUS specification accrue interest
+A counterparty might decide to reinvest profit received as interest
+from the %same 
+% SJT I think "same" is superfluous here: is that OK?
+loan. In the simplest case, this renders as compound
+interest. This can be modelled through interest accrual and capitalisation;
+% SJT what does "capitalisation" mean?
+for instance, contracts from the ACTUS specification accrue interest
 between interest payments and can transfer interest to a notional
@@
 Overall, variable interest rates introduce a certain risk for a lender,
-thus they are subject to hedging. While any instrument that depends
-on the same risk factor (interest rate driver) would suffice - the
-most popular way to hedge a variable interest rate loan is interest
+thus they can be %are 
+subject to hedging. While any instrument that depends
+on the same risk factor (interest rate driver) would suffice, the
+most popular way to hedge a variable interest rate loan is an interest
 rate swap. This instrument allows two (or more) parties to exchange
-their incomes - one from fixed interest rate loans, the other one
-- from variable interest
+their incomes - one from a fixed interest rate loan, the other from a variable-rate loan.
 
@@
 
-Trustless setups, especially ones in the cryptocurrency world (e.g.
-decentralised smart-contracts and exchanges) require zero trust between
+Trustless setups, especially ones in the cryptocurrency world, including 
+decentralised smart-contracts and exchanges, require no trust between
 party and counterparty involved in a contract. In case of a loan,
-this literally means that counterparty has zero-obligation to pay
+this literally means that counterparty has zero obligation to pay
 the money back, thus rendering the loan useless for a party. Such
-risks are usually addressed by introducing collaterals: 
+risks are usually addressed by introducing collaterals, as in the following scenario. 
 \begin{enumerate}
@@
 his USD to Alice 
-\item If Alice pays interest and notional on time (and Btc-price doesn't
-render collateral worthless) - she can get her collateral back, otherwise
-the loan gets liquidated and collateral is transferred to Bob. 
+\item If Alice pays the interest and notional on time, and the BTC price does not
+render collateral worthless, she can get her collateral back; otherwise
+the loan gets liquidated and the collateral is transferred to Bob. 
 \end{enumerate}

### 2021-02-03 papers/actus2021/article.tex
Revised sectinos 2-4 and bibliography.

@@
 Biblio sorted out ordering (added keys). shortened author and editor lists
+\item
+Revised sections 2--4.
 \end{enumerate}
@@
 \item
-Algorithmic Contract Types Unified Standards (ACTUS) -- for first use. Also reference.
+Deal with inline comments in source code: \verb+% SJT ...+
 \item
-Do we include Alex as co-author?
+Write introduction [SJT]
+\item
+Should we include Alex as co-author?
 \item
@@
 
-This section introduces the Marlowe DSL and the ACTUS standard for financial contracts.
+This section introduces the Marlowe domain-specific language (DSL) and the ACTUS standard for financial contracts.
 
@@
 
-Marlowe is a high-level, domain-specific language (DSL) for writing
-financial contracts on blockchain\cite{marlowe}. Marlowe is defined
+Marlowe~\cite{marlowe} is a high-level DSL for writing
+financial contracts on blockchain. Marlowe is defined
 by an executable semantics in Haskell, and has been implemented on
 the UTxO-based Cardano blockchain. 
+% SJT added this
+Marlowe contracts are finite, and moreover have defined finite lifetime.
+This means that it is possible to perform \emph{static analysis} on them.
+For example, it is possible symbolically to check \emph{all} execution paths without running a contract, and in particular checking that every \texttt{Pay} construct can be executed in full.
 
 The main advantage of using Marlowe to carry ACTUS logic is enhanced
-security that static analysis provides. SMT-solvers gain increasing
+security that such a static analysis provides. SMT-solvers~\cite{smt} are gaining
 popularity as an alternative to fuzzing when it comes to testing software
-for vulnerabilities\cite{smt}. Smart contracts generated from ACTUS
+for vulnerabilities. Smart contracts generated from ACTUS
 specification are quite complicated, so manually testing every possible
@@
 \subsection{ACTUS}
+\label{ACTUS}
 
-ACTUS defines the logic embedded in legal agreements that eventually
+The Algorithmic Contract Types Unified Standards (ACTUS)~\cite{actus} define the logic embedded in legal agreements that eventually
 turn the contract terms into actual cash flows, or more generally
-business events \cite{actus}. Most of its basic contract types represent
+business events. Most of its basic contract types represent
 different variations of lending contracts.
@@
 ACTUS relies on a state machine formalism in order to describe the
-behaviour of a given contract. Every payoff (transfer of funds in
-or out of a contract) can be inferred for any given state. Every state
-can be inferred from previous events and observed risk factors:
+behaviour of a given contract. Every \emph{payoff} -- i.e. transfer of funds in
+or out of a contract -- can be inferred for any given state. Every state
+can be derived from previous events and observed risk factors:
 
@@
 \begin{equation}
-\mbox{\emph{payoff}}_{i}=POF(state_{i})
+\ident{payoff}_{i}=\ident{POF}(\ident{state}_{i})
+\end{equation}
+
+\newcommand{\STF}{\ident{STF}}
+\noindent 
+\begin{equation}
+\ident{path}=\STF_{1}(ct,ev_{1})\circ \STF_{2}(ct,ev_{2}).\circ\ldots\circ \STF_{i}(ct,ev_{n})
 \end{equation}
@@
 \begin{equation}
-path=STF_{1}(ct,ev_{1})\circ STF_{2}(ct,ev_{2}).\circ\ldots\circ STF_{i}(ct,ev_{n})
+\ident{state}_{i}=\ident{path}(\ident{INIT}(ct)),
 \end{equation}
+\noindent
+where \emph{ct} stands for contract terms, \emph{INIT} returns initial state, \emph{sched}
+returns scheduled events, \emph{STF} takes \emph{contractterms}, \emph{event}, and \emph{state}
+and returns the next state, and \emph{POF} returns the payoff in a state.
 
+\subsection{Oracles}
+
+In order to support variable interest rates and scaling, ACTUS requires
+a smart contract to be able to observe the value of a given risk factor,
+such as an interest rate, at a particular point in time $t$. This is due to the state of the risk factor not being known at instantiation
+time.
+
+% SJT - the "i" and "t" subscripts here look a bit odd: is that what's intended?
 \noindent 
 \begin{equation}
-state_{i}=path(INIT(ct)),
+\ident{riskfactor}_{i}{}_{t}=O_{\footnotesize\ident{rf}}(i,t)
 \end{equation}
-
-where $ct$ is contract terms, $INIT$ returns initial state, $sched$
-returns scheduled events, $STF$ takes ${contractterms,event,state}$
-and returns next state, $POF$ returns payoff.
-
-\subsection{Necessity for Oracles}
-
-In order to support variable interest rates and scaling, ACTUS requires
-a smart contract to be able to observe the value of a given risk factor
-(such as interest rate) at a particular point in time t:
-
 \noindent
-\begin{equation}
-\mbox{\emph{riskfactor}}_{i}{}_{t}=O_{rf}(i,t)
-\end{equation}
-
-This is due to the risk factor's state not being known at instantiation
-time. In the case of Cardano blockchain, these values are usually
-provided through an ``Oracle'' mechanism\cite{oracles}. Oracle
+In the case of the Cardano blockchain, these values are usually
+provided through an \emph{oracle} mechanism\cite{oracles}. An oracle
 could be a trusted party providing necessary data or network of parties
-under consensus {[}find a quote{]}. From Marlowe DSL perspective,
+under consensus {[}find a reference{]}. 
+% SJT REF needed
+From a Marlowe DSL perspective,
 the exact mechanism that provides external data is less important,
-as Marlowe abstracts over IO by requiring particular type of input
-(``Choice'') protected with a cryptographic signature (Choice owner).
-As a result, the event of receiving data from Oracle is treated the
-same as receiving integer numbers from user input in other languages
+as Marlowe abstracts over IO by requiring a particular type of input -- a 
+\texttt{Choice} -- that is protected with a cryptographic signature by the source of the choice.
+% SJT changed from "owner" to "source".
+As a result, the event of receiving data from an oracle is treated the
+same as receiving numeric input in other languages
 
@@
 
-\subsection{Mimicking the specification with Haskell code}
+ACTUS is defined in a textual specification\footnote{Available from \url{https://www.actusfrf.org/techspecs}.} which, while expressed in mathematical notation, is essentially informal.
+In this section we describe how this specification is turned into an \emph{executable} version by rendering it in Haskell.
+This translation is in fact a transliteration, since notation, variable names and so forth are respected.
+% SJT: are you happy with this? If not, let's rephrase,
+% but it would be good include something to this effect.
+
+\subsection{Rendering the specification in Haskell}
 
@@
 state transition functions that are polymorphic on event and contract
-type. The specification also follows quite specific naming conventions
+type, as noted in Section \ref{ACTUS} above. The specification also follows quite specific naming conventions
 that are incompatible with Haskell's conventions. The executable specification
-follows original ACTUS conventions as close as possible in order to
-ease code base maintenance when faced with updates of original ACTUS
-spec (link to github).
+follows original ACTUS conventions as closely as possible in order to
+ease code base maintenance when faced with updates of the ACTUS
+spec repository\footnote{\url{https://github.com/actusfrf/actus-techspecs}}.
 
 Using Haskell itself as a DSL for explicitly encoding formulas without
-using advanced language idioms also simplifies code generation, but
-in case of ACTUS comes at a cost reduced type-safety, handling nullable
+using advanced language idioms also simplifies code generation. 
+In case of ACTUS this comes at a cost reduced type-safety, handling nullable
 values explicitly introduces risk of exceptions. However this risk
-is addressed using QuickCheck generators.
+is addressed using property-based testing, and in particular QuickCheck generators. This is discussed in more detail in Section \ref{verification} below.
+% SJT Added x-ref to section 7.
 
@@
 
-In order to keep our executable specification independent of the carrier
-(whether its smart-contract engine, proof assistant, analytical framework
-or even machine learning model) - we abstract over underlying representations
-of state variables and arithmetic operations:
+In order to keep our executable specification independent of the carrier -- whether it is a smart-contract engine, proof assistant, analytical framework
+or even machine learning model -- we abstract over the underlying representation
+of state variables,
 
@@
 } deriving (Show) 
-
+\end{verbatim}
+\noindent
+and arithmetic operations,
+\begin{verbatim}
 -- Ops.hs
@@
 \end{verbatim}
-
-Thus, every formula in the executable spec could either be instantiated
+\noindent
+Thus, every formula in the executable spec could be instantiated
 to: 
 \begin{itemize}
-\item formula on some ``atomic'' type (like Double, Day) - could be used
+\item a formula on some \emph{atomic} type, like \texttt{Double} or \texttt{Day}, which could be used
 to directly compute cash-flows for analytical purposes or precompute
-payoffs for smart contracts that don't depend on oracles 
-\item formula representing piece of AST (Marlowe Value, Observation) - could
+payoffs for smart contracts that do not depend on oracles; or 
+\item a formula representing a piece of \emph{abstract syntax}, e.g. a Marlowe \texttt{Value} or \texttt{Observation}, that could
 be used to generate smart contracts that depend on oracles or to generate
-code in another language (like Agda) 
+code in another language, such as Agda.
 \end{itemize}
 This approach of abstracting formulas has a limitation of not allowing
-conditionals to be expressed in an abstract way (There is no ActusIf
-typeclass) - luckily most of conditional expressions in ACTUS specification
-don't depend on variable state of a contract - they depend on ContractTerms
+conditionals to be expressed in an abstract way: in other words, there is no \texttt{ActusIf}
+typeclass. Luckily most of conditional expressions in ACTUS specification
+don't depend on variable state of a contract, they depend on \texttt{ContractTerms}
 that are known in advance during contract generation. This allows
@@
 
-The only exception to this are rare situations where we need to compare
-2 state variables and either choose formula1 or 0 depending on comparison
+The only exception to this are the rare situations where we need to compare
+2 state variables and choose either $\ident{formula}^\prime$ or 0 depending on the result of the comparison
 result:
@@
 \begin{equation}
-formula(st)=\begin{cases}
-formula'(st) & var1(st)<var2(st)\\
-0 & otherwise
+\ident{formula}(st)=\begin{cases}
+\ident{formula}^\prime(st) & \ident{var}_1(st)<\ident{var}_2(st)\\
+0 & \ident{otherwise}
 \end{cases}
 \end{equation}
-
-We rely on pseudo-boolean ``less than'' function in order to address
+We rely on a pseudo-Boolean \emph{less than} function in order to address
 that:
@@
 \begin{equation}
-formula(st)=pseudoLt(var1(s1),var2(st)))*formula'
+\ident{formula}(st)=\ident{pseudoLt}(\ident{var}_1(st),\ident{var}_2(st)))*\ident{formula}^\prime
 \end{equation}
@@
 \begin{equation}
-pseudoLt(a,b)=Cond(a>b,1,0)
+\ident{pseudoLt}(a,b)=\ident{Cond}(a>b,1,0)
 \end{equation}
@@
 
-In order to simplify serialisation/deserialisation of contract terms
-across Actus related services maintained by Cardano - we rely on ``superposed''
-representation of contract terms - all ACTUS contract types are represented
-with the same type. While such representation allows both encoder
+In order to simplify serialisation and deserialisation of contract terms
+across ACTUS related services maintained by Cardano we rely on ``superposed''
+representation of contract terms: all ACTUS contract types are represented
+with the same type.
+
+While such a representation allows both encoder
 and decoder to express any ACTUS contract terms, it also allows for
-invalid combinations of terms (for example PRNXT cannot be applied
-to PAM-contract), which means contracts require specific validation
-that is implemented by means of applicability function:
+invalid combinations of terms (for example \emph{PRNXT} cannot be applied
+to a PAM contract), which means contracts require specific validation
+that is implemented by means of the applicability function:
 
+% SJT - do you mean \mapsto or \rightarrow here?
 \noindent 
 \begin{equation}
-Applicability:ContractTerms\mapsto Bool
+\ident{Applicability}:\ident{ContractTerms}\mapsto \ident{Bool}
 \end{equation}
@@
 type (TODO link to ACTUS applicability):
+% SJT something needed here.
 
+\smallskip
+% SJT changed ContractTerm to ContractTerms
 \noindent 
 \begin{equation}
-Applicability:ContractType\times ContractTerm\mapsto ApplicabilityType,
+\ident{Applicability} : \ident{ContractType} \times \ident{ContractTerms} \mapsto \ident{ApplicabilityType}
 \end{equation}
+\noindent
+where applicability could be: \emph{none}, \emph{always}, \emph{nullable}, or \emph{multiple}.
 
-where applicability could be: none, always, nullable, multiple
-
-In order to build superposed contract terms type for such function,
-we have to resolve conflicting applicability for contract terms merged
+In order to build superposed contract terms type for such functions,
+we have to resolve conflicting applicability types for merged contract terms
 using the following resolution rules:
@@
 \begin{equation}
-weaken(a_{1},a_{2})=\begin{cases}
-nullable & a_{1}=none\wedge a_{2}=always\\
-nullable & a_{1}=always\wedge a_{2}=none\\
-a_{1} & priority(a_{1})>priority(a_{2})\\
-a_{2} & otherwise
+\ident{weaken}(a_{1},a_{2})=\begin{cases}
+\ident{nullable} & a_{1}=\ident{none}\wedge a_{2}=\ident{always}\\
+\ident{nullable} & a_{1}=\ident{always}\wedge a_{2}=\ident{none}\\
+a_{1} & \ident{priority}(a_{1})>\ident{priority}(a_{2})\\
+a_{2} & \ident{otherwise}
 \end{cases}
@@
 \begin{equation}
-priority(x)=\begin{cases}
-0 & x=always\\
-1 & x=none\\
-2 & x=nullable\\
-3 & x=multiple
+\ident{priority}(x)=\begin{cases}
+0 & x=\ident{always}\\
+1 & x=\ident{none}\\
+2 & x=\ident{nullable}\\
+3 & x=\ident{multiple}
 \end{cases}
 \end{equation}
-
-Where $a_{1}$ - is applicability of a given term of first contract
-type to be merged $a_{2}$ - is applicability of a given term of second
-contract type to be merged
+\noindent
+where $a_{i}$ is the applicability of a given term of the \emph{i}th contract
+type to be merged.
 

### 2021-02-04 papers/actus2021/article.tex
Reviewed Section 5.

@@
 What about R (shiny)?
+\item
+We use \verb+equation+ environment which numbers equations, but the numbering isn't used I think.
+So I'd suggest moving to \verb+equation*+ instead (easy edit).
 \end{enumerate}
@@
 
-\subsection{Contract terms representation and explicit applicability}
+\subsection{Contract term representation and explicit applicability}
 
@@
 
-\section{Generating Marlowe contracts from ACTUS specs}%standardised ACTUS contract terms}
+\section{Generating Marlowe contracts from standardised ACTUS contract terms}
 \label{generation}
+
+In this section we describe how concrete Marlowe contracts are generate from the terms -- i.e.
+parameters -- of standard ACTUS contracts; we also describe the implementation of the system, and reflect on the limitations of generating contracts in Marlowe, where contracts have a a predefined lifetime and collection of interactions with contract participants.
 
@@
 
-Generated contract is essentially a continuation chain of smaller
-contracts:
+A generated contract is essentially a continuation chain of smaller contracts:
 
@@
 \begin{equation}
-chainlink(t)=receiveData(t)\circ \mbox{\emph{calculatePayoff(t)}}\circ \mbox{\emph{processPayoff(t)}}
+\ident{chainlink}(t)=\ident{receiveData}(t)\circ \ident{calculatePayoff(t)}\circ \ident{processPayoff(t)}
 \end{equation}
@@
 \begin{equation}
-contract(ct)=collaterals(ct)\circ INIT(ct)\circ\prod_{t\in SCHED(ct)}chainlink(t)
+\ident{contract}(ct)=\ident{collaterals}(ct)\circ \ident{INIT}(ct)\circ\prod_{\footnotesize t\in \ident{SCHED(ct)}}\ident{chainlink(t)}
 \end{equation}
 
-where $receiveData$ - is asking an oracle for Marlowe Choice if needed,
-$\mbox{\emph{calculatePayoff}}$ - calculates payoff formula. For fixed-rate contracts
-this is optimised into a pre-calculated constant. $\mbox{\emph{processPayoff}}$
-- awaits Deposit of payoff amount from a party: if deposit is transferred
-it redirects the funds to a counterparty - otherwise it transfers
-the collateral to a counterparty and closes the contract. The chain
-is generated from the fixed (known in advance) schedule of events
-using the $SCHED()$ function from ACTUS.
+\noindent
+where the component \emph{receiveData} asks an oracle for Marlowe Choice if needed, and
+\mbox{\emph{calculatePayoff}} calculates the payoff formula. For fixed-rate contracts
+this is optimised into a pre-calculated constant function. The \emph{processPayoff} function
+awaits the \texttt{Deposit} of a payoff amount from a party: if the deposit is made them 
+it directs the funds to a counterparty, otherwise it transfers
+the collateral to the counterparty and closes the contract. The principal components of the system are shown in Figure \ref{fig1}.
+% SJT added this figure ref: OK?
+% Do we need to add a little more to explain Fig1?
+
+The chain
+is generated from the fixed schedule of events, known in advance of execution, 
+using the \ident{SCHED()} function from ACTUS, as shown in Figure \ref{fig2}.
+% SJT added this figure ref: OK?
 
@@
 
-Marlowe doesn't allow functions and recursion. Naive usage of If operator
-in Marlowe could lead to exponential growth of a contract, for example:
+Marlowe contracts are finite, and in particular Marlowe itself does not have constructs for functions or recursion; these \emph{are} available in the Haskell and JavaScript embeddings of Marlowe, but they are unrolled on translation to pure Marlowe.
+Naive usage of the \texttt{If} operator in Marlowe could lead to exponential growth of a contract, as in this pesudocode example:
 
@@
 	then 
-		perform_something1()
+		perform\_something1()
 		continue() 
 	else 
-		perform_something2 
+		perform\_something2() 
 		continue()
@@
 \noindent
-would inline contents of \texttt{continue()} twice and given that ACTUS
-contracts are essentially generated using continuation as an accumulator
-- this would lead to exponential explosion of the size of any ACTUS
+Translating this to Marlowe would inline contents of \texttt{continue()} twice and, given that ACTUS
+contracts are essentially generated using continuation as an accumulator,
+this would lead to exponential explosion of the size of any ACTUS
 contract that has conditionals in their state transition logic.
 
-Example of such logic would be cap/floor limitations on interest rate:
+An example of such logic would be cap/floor limitations on interest rates:
 
@@
 \begin{equation}
-adjusted=max(min(original,floor),cap)
+\ident{adjusted}=\ident{max(min(original,floor),cap)}
 \end{equation}
-
-We addressed this issue by introducing the \texttt{Cond} construct in order
-to represent conditional expressions. So instead of using the \texttt{If}
-contract to decide the value of some variable - we use conditional
-assignment instead as \texttt{Cond} is essentially just a pure function
-that returns value depending on condition in contrast to the \texttt{If}
-contract that performs action instead of returning value. The name
-is borrowed from a similar function in Lisp.
+\noindent
+We addressed this issue by introducing the \texttt{Cond} expression construct in order
+to represent conditional expressions, rather than only conditional contracts as was the case before. Instead of using the \texttt{If}
+contract to decide the value of some variable, we use a conditional expression instead.  
+\texttt{Cond} is  a pure function
+that returns a value depending on a condition, in contrast to the \texttt{If}
+contract that chooses between two continuation contracts. 
+% SJT I don't think we need to say this.
+%The name is borrowed from a similar function in Lisp.
 
@@
 
-Marlowe doesn't allow perpetual contracts even if their recursion
-is productive (example: perpetual swap) - thus we can't support certain
+Marlowe doesn't allow contracts that run indefinitely, even if their recursion
+is productive, as would be the case in a perpetual swap contract, for example. We therefore cannot support certain
 contract types from ACTUS specification, namely the ones that don't
-have a defined maturity date (like UMP). In a long-term perspective
-- there is a possible workaround: contracts with no maturity date
-could be represented as actors with finite amounts of state transitions.
-We prototyped this approach in a separate branch, however it does
-seem more prone to errors comparing to rendering predefined schedules
-and more importantly it greatly affects static analysis as the amount
-of ``reduction steps'' grows from 
+have a defined maturity date (like UMP).
+
+% In a long-term perspective
+There is a possible workaround: contracts with no maturity date
+could be represented as actors with a finite number of state transitions.
+% SJT is this ^^^ rephrasing OK?
+We prototyped this approach, % in a separate branch, 
+however it does
+seem more prone to errors comparing to rendering predefined schedules.
+More importantly, it greatly affects static analysis because the number
+of reduction steps in the contract grows from 
 \noindent 
-\begin{equation}
-N_{scheduled}=count(event)
-\end{equation}
+\begin{equation*}
+N_{\footnotesize \ident{scheduled}}=\ident{count(event)}
+\end{equation*}
 to 
@@
 \begin{equation}
-N_{stateTransitions}=\frac{max(date(event))-min(date(event))}{precision} 
+N_{\footnotesize \ident{stateTransitions}}=\frac{\ident{max(date(event))}-\ident{min(date(event))}}{\ident{precision}} 
 \end{equation}
@@
 
-Marlowe only supports Integers, while ACTUS is expressed in terms
-of real numbers. In order to model a real number in such a setup -
-we rely on fixed point precision. The algebra looks like this: 
+For numeric types Marlowe supports Integers, while ACTUS is expressed in terms
+of real numbers. In order to model a real number in such a setup
+we rely on fixed-point precision. The algebra looks like this: 
 \begin{verbatim}
@@
 \end{verbatim}
+% SJT this needs more explanation here ^^^
+\noindent
+We plan to move Marlowe to fixed-precision numbers for the on-blockchain implementation available later in 2021.
 
@@
 
-Unfortunately, Marlowe DSL doesn't support any notion of records,
-the variables can only be of type ``Integer''. Thus, in order to
-map contract state (ContractStatePoly) we first pack a set of Marlowe
-variables (Value Observation) representing previous $t-1$ state into
-ContractStatePoly, then we apply polymorphic state transition, and
-finally unpack ContractState poly into a set of Marlowe variables
-into $t+1$ state:
-
+The Marlowe DSL does not support any notion of records,
+the variables can only be of type ``Integer''. In order to
+map contract state -- \emph{ContractStatePoly} -- we pack a set of Marlowe
+variables of type \texttt{Value} and \texttt{Observation}, representing the previous ($t-1$) state, passing them into
+\emph{ContractStatePoly}, to apply the polymorphic state transition, and
+finally unpack \emph{ContractStatePoly} into a set of Marlowe variables representing the 
+ state at $t+1$:
+% SJT again I am confused about t-1 to t+1 transition. Why not just one step?  
 \noindent 
@@
 \end{equation}
+\noindent
+where \emph{pack} is a chain of \texttt{UseValue} constructs and \ident{unpack} is a chain of
+\texttt{Let}s.
 
-,where pack is a chain of ``UseValue'' and unpack is a chain of
-``Let''
+\paragraph{Representing state transitions in Marlowe}
 
-\paragraph{Representing state transitions in Marlowe (unfolding)}
-
-Marlowe doesn't support mutable variables, so we have to represent
-state at t ($st_{t}$) literally through naming convention:
+Marlowe is a declarative language, and so in partticular does support mutable variables. We therefore represent the state at stage $t$ ($st_{t}$) literally through this naming convention:
 
@@
 
-\begin{figure} 
+\begin{figure}[t]
 \includegraphics[width=0.8\textwidth]{images/labs.png}
-\caption{Actus Labs - online tool for generating Actus contracts for Marlowe.} 
+\caption{Actus Labs - an online tool for generating Actus contracts for Marlowe.} 
 \label{fig3} 
@@
 In order to demonstrate and test the capabilities of Actus generators,
-a visual online Blockly-based tool was developed for Marlowe Playground.
-The Actus Labs tool allows users to construct contract terms visually,
-generate a corresponding Marlowe contract and try it out in a simulated
+a visual online Blockly-based tool was developed for the Marlowe Playground.
+The Actus Labs tool, shown in Figure \ref{fig3}, allows users to construct contract terms visually
+to generate a corresponding Marlowe contract and then to try it out in a simulation 
 environment.

### 2021-02-04 papers/actus2021/article.tex
fixes

@@
 
+% dk14 
+% A1 it is from the ACTUS spec - I assume "r" means "rate"
+% A2 I meant "before vs after event", but I'll remove it
+
 \noindent 
@@
 \begin{equation}
-\ident{interestRate}_{t+1}=\ident{capfloor}(\ident{interestRate}_{t-1}+\Delta_{r}),
+\ident{interestRate}_{t}=\ident{capfloor}(\ident{interestRate}_{t-1}+\Delta_{r}),
 \end{equation}
@@
 % SJT I think "same" is superfluous here: is that OK?
+% dk14 sure
 loan. In the simplest case, this renders as compound
@@
 % SJT what does "capitalisation" mean?
+% dk14 "conversion of income or assets into capital" it's quite googlable financial terms so idk if I should note it explicitly
 for instance, contracts from the ACTUS specification accrue interest
@@
 % SJT added this
+%dk14 OK
 Marlowe contracts are finite, and moreover have defined finite lifetime.
@@
 % SJT - the "i" and "t" subscripts here look a bit odd: is that what's intended?
+% dk14 - fixed
 \noindent 
 \begin{equation}
-\ident{riskfactor}_{i}{}_{t}=O_{\footnotesize\ident{rf}}(i,t)
+\ident{riskfactor}_{it}=O_{\footnotesize\ident{rf}}(i,t)
 \end{equation}
@@
 % SJT - do you mean \mapsto or \rightarrow here?
+% dk14 - it is a mapping, but could be morphism too
 \noindent 
 \begin{equation}
-\ident{Applicability}:\ident{ContractTerms}\mapsto \ident{Bool}
+\ident{Applicability}:\ident{ContractTerms}\rightarrow \ident{Bool}
 \end{equation}
 
-ACTUS defines a family of applicability functions polymorphic on contract
-type (TODO link to ACTUS applicability):
+ACTUS standard defines a family of applicability functions polymorphic on contract
+type:
 % SJT something needed here.
+% dk14 - we already gave a link to actusfrf, so I think I can skip references
 
@@
 \begin{equation}
-\ident{Applicability} : \ident{ContractType} \times \ident{ContractTerms} \mapsto \ident{ApplicabilityType}
+\ident{Applicability} : \ident{ContractType} \times \ident{ContractTerms} \rightarrow \ident{ApplicabilityType}
 \end{equation}
@@
 it directs the funds to a counterparty, otherwise it transfers
-the collateral to the counterparty and closes the contract. The principal components of the system are shown in Figure \ref{fig1}.
+the collateral to the counterparty and closes the contract.
+The principal components of the system are shown in Figure \ref{fig1}.
+There are 3 categories of components representing ACTUS functions: specification with formulas, formula wiring for fixed rates (produces precomputed payoffs) and formula wiring for variable rates (produces Marlowe-code that computes payoffs).
+Different types of wiring correspond to different implementations of \emph{ActusOps} implemetations (either \emph{Double} or \emph{Value}).
 % SJT added this figure ref: OK?
 % Do we need to add a little more to explain Fig1?
+% dk14 OK, added few sentences about Fig1
 
@@
 % SJT added this figure ref: OK?
+% dk14 OK
 
@@
 % SJT is this ^^^ rephrasing OK?
+% dk14 - sure
 We prototyped this approach, % in a separate branch, 
@@
 \begin{verbatim}
-  (+)   = AddValue
-  (-)   = SubValue
-  a * b = Scale (1 % marloweFixedPoint) $ MulValue a b
+  (+)   = AddValue -- x/n + y/n = (x + y)/n
+  (-)   = SubValue -- x/n - y/n = (x - y)/n
+  a * b = Scale (1 % marloweFixedPoint) $ MulValue a b -- x/n * y/n = (x * y)/n^2
 \end{verbatim}
+We scale all numbers with \emph{marloweFixedPoint} factor - which only requires modification of \emph{MulValue}.
 % SJT this needs more explanation here ^^^
+% dk14 I've added some
 \noindent
@@
 % SJT again I am confused about t-1 to t+1 transition. Why not just one step?  
+% dk14. I'll change that. I imagined t as time of event while t+1 is after event 
 \noindent 
 \begin{equation}
-st_{t+1}=unpack(stateTransition(pack(st_{t-1}))),
+st_{t}=unpack(stateTransition(pack(st_{t-1}))),
 \end{equation}

### 2021-02-04 papers/actus2021/article.tex
comments
more fixes
Revised to end section 7.

--- 2021-02-04 papers/actus2021/article.tex
@@
 \item
-Revised sections 2--4.
+Revised sections 2--7.
 \end{enumerate}
@@
 
+A note on notation: \texttt{typewriter} font will be used for Marlowe constructs
+% SJT and fragments?
+while \emph{math} font will be used for mathematical formulas and pseudo-code.
+% Have I got this right?
+
 % SJT: added this overview of the paper
@@
 Section \ref{executable} builds an executable specification of ACTUS in Haskell, and this is used in Section \ref{generation} to generate  the Marlowe code for an ACTUS contract from the terms of the contract.
-Section \ref{tokenization} explains how tokens are used to represent ownership of roles in a running contract, and Section \ref{verification} presents how we provide assurance of the contract correctness.
+Section \ref{tokenization} explains how tokens are used to represent ownership of roles in a running contract, and Section \ref{assurance} describes how we provide assurance that  contracts behave as they should.
 Section \ref{related} examines related work and Section \ref{conclusion} concludes.
@@
 
-Every participant of an ACTUS contract is described by a Role which
-is in its turn represented through a special unique non-fungible token.
-This makes every Marlowe contract a tradable security, allowing a
-participant to sell its ``share'' in a contract by selling a corresponding
+Every participant of a Marlowe contact %ACTUS contract 
+is described by a \texttt{Role} which
+is in its turn represented through a unique non-fungible token, created at the time that the contract instantiated on the blockchain.
+This makes every ACTUS contract a tradable security, allowing a
+participant to sell its \emph{share} in a contract by selling a corresponding
 role token.
 
-Role tokens can potentially allow more complex manipulation over shares,
-especially when share represents an incoming cash flow: participants
-only send funds to a party represented by a given token. Such a token
-would represent a positive cashflow which in turn could not just become
-tradable but would also allow to derive tokens only representing fractions
+Role tokens can potentially allow more complex manipulation over such shares,
+especially when the share represents an incoming cash flow; in that case, participants
+send funds to a party represented by a given token. Such a token
+would represent a positive cashflow, which in turn could not only become
+tradable but could also allow the derivation of tokens representing \emph{fractional parts}
 of a particular cash flow in a contract.
 
-Moreover, it turns Actus-loans into derivatives, for example contracts
-like Interest Rate Swap (and Swaps in general) could be approximated
-by an Atomic Swap of tokens representing incoming cash flows from
-lending.
-
-For example, if Alice has fixed income from a loan (or some other
-investment) and Bob has comparable but variable (fluctuating) income
-- Bob can hedge by swapping cash flows with Alice. If Alice's income
-is locked with token1 and Bob's income is locked with token2 - then
+Moreover, this process turns ACTUS loans into derivatives. For example, contracts
+like \emph{Interest Rate Swap} (and \emph{Swaps} in general) could be approximated
+by an \emph{Atomic Swap} of tokens representing incoming cash flows from
+loans. 
+For example, if Alice has fixed income from a loan, or some other
+investment, and Bob has comparable but variable (fluctuating) income,
+Bob can hedge by swapping cash flows with Alice. If Alice's income
+is locked with \texttt{token1} and Bob's income is locked with \texttt{token2} then an
 atomic swap of those tokens is equivalent to a swap of cash flows.
 
-\section{Verification and testing}
-\label{verification}
+\section{Assurance}
+\label{assurance}
+
+We provide assurance to users of our Marlowe ACTUS contracts in a number of ways, as described here.
 
@@
 
-In order to test the equivalence between Executable Haskell implementation
-of ACTUS for smart-contracts and original Java implementation for
-analytics - a simple property test was introduced:
+First, we are able to test the executable Haskell implementation
+of ACTUS for smart-contracts by comparing it with an existing implementation written in Java. To do this  
+%original Java implementation for analytics - 
+a simple property-based test in QuickCheck~\cite{qc} was introduced, where we generate contract terms and risk factors randomly to test the property.
 
+\newcommand{\rf}{\ident{rf}}
+\newcommand{\ct}{\ident{ct}}
 \smallskip
@@
 \begin{equation}
-\forall ct\forall rf.getCashFlows("haskell",ct,rf)\equiv getCashFlows("java",ct,rf)
+\forall \ct.\forall \rf. \; \ident{getCashFlows}(\ident{"haskell"},\ct,\rf)\equiv \ident{getCashFlows}(\ident{"java"},\ct,\rf)
 \end{equation}
-
-Where $ct$ represents contract terms, $rf$ is risk factor model and
-$getCashFlows()$ returns a set of $(date,payoff)$ tuples.
+\noindent
+where \emph{ct} represents contract terms, \emph{rf} is risk factor model, and
+\emph{getCashFlows()} returns a set of \emph{(date,payoff)} tuples.
 
@@
 
-QuickCheck\cite{qc} contract terms generators also allow to ensure
-arbitrary properties of a Marlowe contract. This could be enhanced
-with Marlowe's static analysis feature by utilising Assert operator:
+QuickCheck contract terms generators also allow us to check
+other properties of a Marlowe contract. This could be enhanced
+with Marlowe's static analysis feature by utilising the \texttt{Assert} operator:
 
+% SJT what language is this written in?
+% can we turn it into Haskell? or Java?
 \begin{verbatim}
@@
 
-While it doesn't cover all possible contracts it could guarantee that
+While this scenario does not cover all possible contracts, it could guarantee that
 property holds for a statistically significant fraction of a contract.
-Using Static Analysis to check a particular contract instead of randomly
-sampling allows: 
+
+\subsection{Static analysis for verification}
+% SJT added this as a separate sub section, as it covers the static analysis work.
+% references added.
+
+Using static analysis for Marlowe~\cite{} it is possible fully to check a particular contract instead of using random
+sampling. That would allow some refinements: 
 \begin{itemize}
-\item more balanced sampling: Space of contract terms depends linearly on
+\item More balanced sampling: the space of contract terms depends linearly on
 the coverage, e.g. if we cover 10\% of all possible contract terms
 - we'll cover 10\% of all contracts. Different contracts have different
-amount of risk factors thus different search space, for instance a
+sets of risk factors, and so different search spaces. For instance, a
 contract with 3 risk factors (e.g. 3 observations of an interest rate)
-would span over n\textasciicircum 3 while a contract with 10 risk
-factors would span over n\textasciicircum 10. Meanwhile, space of
+would span over $n^3$ values while a contract with 10 risk
+factors would span over $n^{10}$. Meanwhile, the space of
 all risk factors in all contracts doesn't linearly depend on the coverage.
-\item dependency tracking: SMT-solver would likely be more aware of execution
-paths leading to failure of a test, which could significantly reduce
-search space 
-\item completeness: If not timed out, SMT-solver is decidable while sampling
+% SJT explain last sentence. 
+\item Dependency tracking: an SMT-solver is potentially likely to be more aware of execution
+paths that lead to the failure of a test, a feature that could significantly reduce
+search space. 
+\item Completeness: if not timed out, SMT-solving is decidable while sampling
 is semi-decidable. 
@@
 
-By design, the Marlowe interpreter always refunds assets held in a
+By design, the Marlowe interpreter always refunds any assets held in a
 contract when it terminates. This is done in order to ensure that
 no funds are lost forever. The funds are returned to whoever's internal
-account actually holds them.
+account  holds them.
 
 However, this presents as a problem in certain cases where ownership
-of the funds could not be automatically determined. For example if
-Alice puts collateral in a crypto-loan contract - she would formally
+of the funds could not be  determined automatically. For example, if
+Alice puts collateral in a crypto-loan contract she would formally
 maintain ownership, which means she would get automatically refunded
-when Close contract is reached.
+when a \texttt{Close} construct is reached.
 
 This implicit refund is easy to overlook by smart-contract developers
-as plain Close is often used as a default action in case of unexpected
-behaviour like timeouts(especially Choice timeout). This can easily
-lead to multi-million dollar mistakes if malicious Alice decides to
+as plain \texttt{Close} is often used as a default action in case of unexpected
+behaviour like timeouts, and especially a \texttt{Choice} timeout. This can easily
+lead to costly %multi-million dollar 
+mistakes if  Alice maliciously decides to
 exploit an auto-refund feature in order to get her collateral without
-paying back. Here's the example of simple attack vector: 
+paying back, as in the following scenario:
 \begin{enumerate}
-\item Alice creates a contract where Deposit timeout would lead to Close 
-\item Bob doesn't know or test the timeout path. Even if Bob is a programmer
-- he might not be aware that Closing contract would cause collateral
-refund 
-\item Alice puts collateral and gets notional from Bob 
-\item Alice doesn't pay for a loan (Deposit timeout) 
-\item Alice gets her collateral back 
-\item Bob loses notional 
+\item Alice creates a contract where a \texttt{Deposit} timeout would lead to \texttt{Close} 
+\item Bob doesn't know or test the timeout path. Even if Bob is a programmer,
+he might not be aware that \texttt{Close}ing the contract would cause the collateral
+to be refunded to Alice.
+\item Alice puts her collateral in the contract and gets the notional from Bob.
+\item Alice doesn't pay for the loan: i.e. there is a \texttt{Deposit} timeout. 
+\item Alice gets her collateral back. 
+\item Bob loses his notional. 
 \end{enumerate}
-In order to prevent this from happening, an additional ''Auto-Refund''
-security check was introduced as part of Static Analysis tooling which
-notifies users about all ``Close'' constructs that can lead to automatic
+In order to prevent this from happening, an additional \texttt{Auto-Refund}
+security check was introduced as part of static analysis tooling which
+notifies users about all \texttt{Close} constructs that can lead to automatic
 refunds and encourages users to write explicit logic for edge cases.

### 2021-02-04 papers/actus2021/article.tex

@@
 could be a trusted party providing necessary data or network of parties
-under consensus {[}find a reference{]}. 
+under consensus \cite{de-oracles}. 
 % SJT REF needed
+% dk14. I've found this one has explanation for oracles (ChainLink etc)
+
 From a Marlowe DSL perspective,
@@
 % SJT changed from "owner" to "source".
+% dk14 OK
 As a result, the event of receiving data from an oracle is treated the
@@
 % but it would be good include something to this effect.
+% dk14 thanks for adding this
 
@@
 % SJT - do you mean \mapsto or \rightarrow here?
+% dk14 - it is a mapping, but could be function/morphism too
 \noindent 
 \begin{equation}
-\ident{Applicability}:\ident{ContractTerms}\mapsto \ident{Bool}
+\ident{Applicability}:\ident{ContractTerms}\rightarrow \ident{Bool}
 \end{equation}
 
-ACTUS defines a family of applicability functions polymorphic on contract
-type (TODO link to ACTUS applicability):
+ACTUS standard defines a family of applicability functions polymorphic on contract
+type:
 % SJT something needed here.
+% dk14 - we already gave a link to actusfrf, so I think I can skip references
 
@@
 %The name is borrowed from a similar function in Lisp.
+% dk14 - Oh sure. I was going to remove it too :)
 
@@
 % it's traditional to call this "relatyed work".
+% dk14. sure this is better
 

### 2021-02-04 papers/actus2021/article.tex
added reference for static analysis
more

@@
 % Have I got this right?
+% dk14 looks right to me - I didn't know how to highlight Marlowe constructs properly
 
@@
 % SJT explain "prime" here.
+% it's standard financial terminology: https://www.investopedia.com/terms/p/primerate.asp
 rate that the lender expects -- the \emph{benchmark yield} -- and
@@
 % SJT used value here rather than plot: is that OK?
+% dk14 maybe "curve" would be better
 in order to represent the changes to be made converting between different units
-of measurement: how many percentage points 
+of measurement: how many rate percentage points 
 % SJT percentage points of what?
+% dk14 rate
 you would get for a USD-to-kilowatt conversion
@@
 
-In the context of ACTUS and other event-based frameworks, 
+In the context of ACTUS and simillar frameworks, 
 % SJT - need to explain "event-based" in this context.
+% dk14. "Event-driven" but it has no relation to scaling, so I'll remove it
 there is
@@
 % SJT changed ContractTerm to ContractTerms
+% dk14 OK
 \noindent 
@@
 % can we turn it into Haskell? or Java?
+% dk14 it's pseudo-code but I'll make it more Haskell-y
 \begin{verbatim}
-ContractTerms := sample(qcgenerator) 
-Contract := generateMarloweContract(ContractTerms) 
-ContractWithAssert := appendAssertion(Contract, assertion) 
-runStaticAnalysis(ContractWithAssert)
+do
+let contractTerms = sample(qcgenerator) 
+    contract = generateMarloweContract(contractTerms) 
+    contractWithAssert = appendAssertion(contract, assertion) 
+runStaticAnalysis(contractWithAssert)
 \end{verbatim}
@@
 % references added.
+% dk14. Thanks
 

### 2021-02-04 papers/actus2021/article.tex
more

@@
 
-Using static analysis for Marlowe~\cite{} it is possible fully to check a particular contract instead of using random
+Using static analysis for Marlowe~\cite{marlowe-efficient} it is possible fully to check a particular contract instead of using random
 sampling. That would allow some refinements: 
@@
 factors would span over $n^{10}$. Meanwhile, the space of
-all risk factors in all contracts doesn't linearly depend on the coverage.
+all risk factors in all contracts doesn't linearly depend on the coverage - some contracts might have more risk factor observations and some - less.
+So covering 10\% of all contracts doesn't necessary mean covering 10\% of all possible observations.
 % SJT explain last sentence. 
+% dk14 added explanation.
+Maybe it could be phrased better.
+The idea is that if I have contracts A and B as population and only A is covered then coverage is 50%.
+Meanwhile contract A has 9 observatioins and B has 1 so risk factors have 90% coverage.
+Idk how to put it nicely.
 \item Dependency tracking: an SMT-solver is potentially likely to be more aware of execution
@@
 % SJT changed title from {Comparison with existing smart-contracts approaches }
-% it's traditional to call this "relatyed work".
+% it's traditional to call this "related work".
 % dk14. sure this is better

### 2021-02-04 papers/actus2021/article.tex
minor changes

--- 2021-02-04 papers/actus2021/article.tex
@@
 \usepackage{amsmath}
+\usepackage{alltt}
 \usepackage{graphicx}
@@
 
-\begin{verbatim}
+\begin{alltt}
 if condition 
@@
     continue()
-\end{verbatim}
+\end{alltt}
 \noindent

### 2021-02-04 papers/actus2021/article.tex
layout (v. minor)

@@
 \noindent 
-\begin{equation}
+\begin{equation*}
 \Delta_{r}=\ident{capfloor}(\ident{driver}*\ident{multiplier}+\ident{spread}-\ident{interestRate}_{t-1})
-\end{equation}
+\end{equation*}
 
 \noindent 
-\begin{equation}
+\begin{equation*}
 \ident{interestRate}_{t}=\ident{capfloor}(\ident{interestRate}_{t-1}+\Delta_{r}),
-\end{equation}
+\end{equation*}
 
@@
 \noindent 
-\begin{equation}
+\begin{equation*}
 \ident{capfloor}(x)=\ident{max(min}(x,\ident{floor}),\ident{cap})
-\end{equation}
+\end{equation*}
 
@@
 \noindent 
-\begin{equation}
+\begin{equation*}
 interestPayment=interestScalingFactor*interestRate*notional
-\end{equation}
+\end{equation*}
 \noindent
@@
 \noindent 
-\begin{equation}
+\begin{equation*}
 \ident{payoff}_{i}=\ident{POF}(\ident{state}_{i})
-\end{equation}
+\end{equation*}
 
@@
 \noindent 
-\begin{equation}
+\begin{equation*}
 \ident{path}=\STF_{1}(ct,ev_{1})\circ \STF_{2}(ct,ev_{2}).\circ\ldots\circ \STF_{i}(ct,ev_{n})
-\end{equation}
+\end{equation*}
 
 \noindent 
-\begin{equation}
+\begin{equation*}
 \ident{state}_{i}=\ident{path}(\ident{INIT}(ct)),
-\end{equation}
+\end{equation*}
 \noindent
@@
 \noindent 
-\begin{equation}
+\begin{equation*}
 \ident{riskfactor}_{it}=O_{\footnotesize\ident{rf}}(i,t)
-\end{equation}
+\end{equation*}
 \noindent
@@
 
-\begin{equation}
+\begin{equation*}
 \ident{formula}(st)=\begin{cases}
@@
 \end{cases}
-\end{equation}
+\end{equation*}
 We rely on a pseudo-Boolean \emph{less than} function in order to address
@@
 \noindent 
-\begin{equation}
+\begin{equation*}
 \ident{formula}(st)=\ident{pseudoLt}(\ident{var}_1(st),\ident{var}_2(st)))*\ident{formula}^\prime
-\end{equation}
+\end{equation*}
 
 \noindent 
-\begin{equation}
+\begin{equation*}
 \ident{pseudoLt}(a,b)=\ident{Cond}(a>b,1,0)
-\end{equation}
+\end{equation*}
 
@@
 \noindent 
-\begin{equation}
+\begin{equation*}
 \ident{Applicability}:\ident{ContractTerms}\rightarrow \ident{Bool}
-\end{equation}
+\end{equation*}
 
@@
 \noindent 
-\begin{equation}
+\begin{equation*}
 \ident{Applicability} : \ident{ContractType} \times \ident{ContractTerms} \rightarrow \ident{ApplicabilityType}
-\end{equation}
+\end{equation*}
 \noindent
@@
 \noindent 
-\begin{equation}
+\begin{equation*}
 \ident{weaken}(a_{1},a_{2})=\begin{cases}
@@
 \end{cases}
-\end{equation}
+\end{equation*}
 
 \noindent 
-\begin{equation}
+\begin{equation*}
 \ident{priority}(x)=\begin{cases}
@@
 \end{cases}
-\end{equation}
+\end{equation*}
 \noindent
@@
 \noindent 
-\begin{equation}
+\begin{equation*}
 \ident{chainlink}(t)=\ident{receiveData}(t)\circ \ident{calculatePayoff(t)}\circ \ident{processPayoff(t)}
-\end{equation}
+\end{equation*}
 
 \noindent 
-\begin{equation}
+\begin{equation*}
 \ident{contract}(ct)=\ident{collaterals}(ct)\circ \ident{INIT}(ct)\circ\prod_{\footnotesize t\in \ident{SCHED(ct)}}\ident{chainlink(t)}
-\end{equation}
+\end{equation*}
 
@@
 \noindent 
-\begin{equation}
+\begin{equation*}
 \ident{adjusted}=\ident{max(min(original,floor),cap)}
-\end{equation}
+\end{equation*}
 \noindent
@@
 \noindent 
-\begin{equation}
+\begin{equation*}
 N_{\footnotesize \ident{stateTransitions}}=\frac{\ident{max(date(event))}-\ident{min(date(event))}}{\ident{precision}} 
-\end{equation}
+\end{equation*}
 
@@
   (-)   = SubValue -- x/n - y/n = (x - y)/n
-  a * b = Scale (1 % marloweFixedPoint) $ MulValue a b -- x/n * y/n = (x * y)/n^2
+  a * b = Scale (1 % marloweFixedPoint) $ MulValue a b 
+                            -- x/n * y/n = (x * y)/n^2
 \end{verbatim}
@@
 \noindent 
-\begin{equation}
+\begin{equation*}
 st_{t}=unpack(stateTransition(pack(st_{t-1}))),
-\end{equation}
+\end{equation*}
 \noindent
@@
 \noindent 
-\begin{equation}
+\begin{equation*}
 \forall \ct.\forall \rf. \; \ident{getCashFlows}(\ident{"haskell"},\ct,\rf)\equiv \ident{getCashFlows}(\ident{"java"},\ct,\rf)
-\end{equation}
+\end{equation*}
 \noindent

### 2021-02-04 papers/actus2021/article.tex
TO DO tidying

@@
 
-\textbf{DONE:}
-\begin{enumerate}
-\item
-Added overview at the end of the intro.
-\item
-Tidying up identifiers in math equations using the command \verb+\ident{...}+
-\item
-Biblio sorted out ordering (added keys). shortened author and editor lists
-\item
-Revised sections 2--7.
-\end{enumerate}
-
 \textbf{TO DO:}
@@
 Should we include Alex as co-author?
-\item
-What about R (shiny)?
-\item
-We use \verb+equation+ environment which numbers equations, but the numbering isn't used I think.
-So I'd suggest moving to \verb+equation*+ instead (easy edit).
 \end{enumerate}

### 2021-02-04 papers/actus2021/article.tex
Responding to DK response to SJT comments.

@@
 However, one cannot simply take the bare price of the asset and turn
-it into a rate. In order to make \emph{units of measurement} compatible
+it into a rate. In order to make \emph{units of measurement} compatible with each other
 %SJT compatible with what??
@@
 the resulting interest rate. The multiplier %effectively 
-rescales the interest rate value 
+rescales the interest rate curve 
 % SJT used value here rather than plot: is that OK?
@@
 
-In the context of ACTUS and simillar frameworks, 
+In the context of ACTUS and similar frameworks, 
 % SJT - need to explain "event-based" in this context.
@@
 loan. In the simplest case, this renders as compound
-interest. This can be modelled through interest accrual and capitalisation;
+interest. This can be modelled through interest accrual and capitalisation (conversion of income or assets into capital);
 % SJT what does "capitalisation" mean?
@@
 the collateral to the counterparty and closes the contract.
+
 The principal components of the system are shown in Figure \ref{fig1}.
-There are 3 categories of components representing ACTUS functions: specification with formulas, formula wiring for fixed rates (produces precomputed payoffs) and formula wiring for variable rates (produces Marlowe-code that computes payoffs).
-Different types of wiring correspond to different implementations of \emph{ActusOps} implemetations (either \emph{Double} or \emph{Value}).
+There are three categories of components representing ACTUS functions: specification with formulas, formula wiring for fixed rates (produces precomputed payoffs) and formula wiring for variable rates, which  produces Marlowe-code that computes payoffs.
+Different types of wiring correspond to different implementations of \emph{ActusOps} implementations (either \emph{Double} or \emph{Value}).
 % SJT added this figure ref: OK?
@@
 
-Marlowe is a declarative language, and so in partticular does support mutable variables. We therefore represent the state at stage $t$ ($st_{t}$) literally through this naming convention:
+Marlowe is a declarative language, and so in particular it does not support mutable variables. We therefore represent the state at stage $t$ ($st_{t}$) literally through this naming convention:
 

### 2021-02-04 papers/actus2021/article.tex
isda thoughts

@@
 factors would span over $n^{10}$. Meanwhile, the space of
-all risk factors in all contracts doesn't linearly depend on the coverage - some contracts might have more risk factor observations and some - less.
+all risk factors in all contracts doesn't linearly depend on the coverage - 
+some contracts might have more risk factor observations and some - less. 
 So covering 10\% of all contracts doesn't necessary mean covering 10\% of all possible observations.
@@
 
+There are frameworks like ISDA Common Domain Model \cite{isda-cdm} providing 
+more precise representation of business processes involved in institutional trading 
+as well as code-generation capabilities. However, due to it's structural simplicity ACTUS 
+is much more suited for generating code in a financial domain specific language 
+like Marlowe rather than general-purpose one like Java or Haskell.
+
 \section{Conclusion}

### 2021-02-04 papers/actus2021/article.tex
minor change
edit list of contributions

@@
 (Dmytro) 
-\item R Shiny cashflow visualiser for ACTUS (Dmytro) 
 \item ACTUS Utility functions (Dmytro, Quanterall) 
@@
 \item ACTUS-like examples on Playground (Dmytro, Alex, Pablo) 
-\item Marlowe DSL enhancements for Actus: Cond, Mul (Dmytro) (draft) 
+\item Marlowe DSL enhancements for Actus: Cond, Mul (Dmytro)
 \item ACTUS spec generator for Agda (Dmytro) 
-\item Auto-refund static analysis (Pablo) (partial: only QC) 
+\item Auto-refund static analysis (Pablo)
 \item Verification with QC + SMT (Dmytro) 
@@
 
-Unlike current popular DeFi lending approaches - Marlowe ACTUS-based
+Unlike current mainstream DeFi lending approaches - Marlowe ACTUS-based
 implementation tends to rely on trade-matching instead of pooling.

### 2021-02-04 papers/actus2021/article.tex
fix

@@
 finally unpack \emph{ContractStatePoly} into a set of Marlowe variables representing the 
- state at $t+1$:
+ state at $t$:
 % SJT again I am confused about t-1 to t+1 transition. Why not just one step?  

### 2021-02-04 papers/actus2021/article.tex
Written first draft of intro; moved Marlowe intro to sec 1, moved ACTUS to sec 2.

@@
 
-\author{Dmytro Kondratiuk\inst{1}\orcidID{XXXX} \and
+\author{Dmytro Kondratiuk\inst{1} \and
 				Pablo {Lamela Seijas}\inst{1}\orcidID{0000-0002-1730-1219} \and
+				Alexander Nemish\inst{1} \and
                 Simon Thompson\inst{1,2}\orcidID{0000-0002-2350-301X}}
@@
 \institute{IOHK, Hong Kong,
-\email{dmytro.kondratiuk@iohk.io, pablo.lamela@iohk.io, simon.thompson@iohk.io}  \and
+\email{dmytro.kondratiuk@iohk.io, pablo.lamela@iohk.io, alexander.nemish@iohk.io, simon.thompson@iohk.io}  \and
 School of Computing, University of Kent, UK,
@@
 
-\keywords{ACTUS \and blockchain \and Cardano \and finance \and Haskell \and Marlowe} 
+\keywords{ACTUS \and blockchain \and Cardano \and finance \and Haskell \and Marlowe
+\and smart contract \and static analysis} 
 \end{abstract}
-
-\textbf{TO DO:}
-\begin{enumerate}
-\item
-Deal with inline comments in source code: \verb+% SJT ...+
-\item
-Write introduction [SJT]
-\item
-Should we include Alex as co-author?
-\end{enumerate}
 
@@
 
+Smart contracts -- programs that run in a blockchain environment --  can be defined in a variety of ways~\cite{cryptoeprint:2016:1156}.
+Many such languages are general purpose, and can be used to program any kind of contract it makes sense to run on a blockchain; moreover, they tend to be expressive enough to be Turing complete (in some cases with restrictions on the runtime environment).
+For example, Plutus, the general-purpose language running on the Cardano blockchain~\cite{utxo-vs-accounts}, is a dialect of Haskell.
+Another approach is to develop special-purpose or \emph{domain-specific languages} (DSLs) which embody a particular application domain: Marlowe~\cite{marlowe} is a high-level DSL for writing
+financial contracts on the Cardano blockchain.
+
+In this paper we explore ways in which contracts described in ACTUS (Algorithmic Contract Types Unified Standards) can be defined in Marlowe, Plutus and Haskell.
+Of course, Plutus or Haskell are able to express these contracts, but rendering them in Marlowe brings extra advantages.
+Marlowe is defined to provide a range of guarantees by design: a Marlowe contract will only make a finite number of interactions with its environment, and its lifetime can be read off from the code for a contract; moreover, when the contract terminates, any assets held by the contract will automatically be returned to the participants.
+None of these guarantees can be provided by a general purpose language.
+
+Each Marlowe contract has a finite set of possible execution paths, and so it is possible to analyse the complete behaviour of a contract without running it.
+Such \emph{static analysis}, based on SMT solving~\cite{smt}, can be used to check properties of a contract; for example, it is possible to check whether a contract will honour all \texttt{Pay} constructs, however it is executed.
+In the case that a \texttt{Pay} can fail, the analysis gives an example trace showing how that failure can happen.
+The language design and static analysis provide assurance that Marlowe contracts are much less likely to ``misbehave'' than contracts written in a general-purpose language like Solidity.
+
+In implementing ACTUS we are able to provide further assurance in three other ways.
+First, we are able to use the declarative nature of Haskell to transliterate ACTUS formulae term-by-term into an executable form.
+Secondly, we are able to use random, property-based testing to validate the Haskell implementation against a Java one.
+Finally, we are able to automatically \emph{generate} ACTUS contracts in Marlowe from the terms -- i.e.
+parameters -- of the contracts; for a simple loan these would include the start and end dates of the loan and the amount loaned (the `principal').
+The generated contracts use the executable spec in calculating values of cash flows in the Marlowe contracts.
+
+Implementing ACTUS provides a suitable benchmark against which to assess Marlowe, and we illustrate how the implementation has led to the addition of a conditional expression construct to the language.
+
 A note on notation: \texttt{typewriter} font will be used for Marlowe constructs
@@
 % 'signposting' what is in the rest of the paper.
-Section \ref{background} of the paper covers the relevant financial background, and Section \ref{modelling} introduces Marlowe as well as the ACTUS financial standard.
+Section \ref{background} of the paper covers the relevant financial background together with the ACTUS financial standard.
 Section \ref{executable} builds an executable specification of ACTUS in Haskell, and this is used in Section \ref{generation} to generate  the Marlowe code for an ACTUS contract from the terms of the contract.
@@
 \label{background}
+
+In this section we give a brief introduction to financial contracts, and loans in particular, and then describe the ACTUS financial standard.
+
+\subsection{Crypto-loans}
 
@@
 
-\section{Modelling financial contracts}
-\label{modelling}
+%\section{Modelling financial contracts}
+%\label{modelling}
 
-This section introduces the Marlowe domain-specific language (DSL) and the ACTUS standard for financial contracts.
-
-\subsection{Marlowe}
-
-Marlowe~\cite{marlowe} is a high-level DSL for writing
-financial contracts on blockchain. Marlowe is defined
-by an executable semantics in Haskell, and has been implemented on
-the UTxO-based Cardano blockchain. 
-% SJT added this
-%dk14 OK
-Marlowe contracts are finite, and moreover have defined finite lifetime.
+%This section introduces the Marlowe domain-specific language (DSL) and the ACTUS standard for financial contracts.
+%
+%\subsection{Marlowe}
+%
+%Marlowe~\cite{marlowe} is a high-level DSL for writing
+%financial contracts on blockchain. Marlowe is defined
+%by an executable semantics in Haskell, and has been implemented on
+%the UTxO-based Cardano blockchain. 
+%% SJT added this
+%%dk14 OK
+%Marlowe contracts are finite, and moreover have defined finite lifetime.
 This means that it is possible to perform \emph{static analysis} on them.
 For example, it is possible symbolically to check \emph{all} execution paths without running a contract, and in particular checking that every \texttt{Pay} construct can be executed in full.
-
-The main advantage of using Marlowe to carry ACTUS logic is enhanced
-security that such a static analysis provides. SMT-solvers~\cite{smt} are gaining
-popularity as an alternative to fuzzing when it comes to testing software
-for vulnerabilities. Smart contracts generated from ACTUS
-specification are quite complicated, so manually testing every possible
-execution path for every possible combination of contract terms is
-not feasible. Static analysis, in turn, can reduce the amount of ``dumb''
-mistakes that could cost millions to the users by ensuring obvious
-properties of a contract. It is not a panacea, however, unlike with
-fuzzing, if SA says that certain property holds - it does it with
-100\% assurance.
+%
+%The main advantage of using Marlowe to carry ACTUS logic is enhanced
+%security that such a static analysis provides. SMT-solvers~\cite{smt} are gaining
+%popularity as an alternative to fuzzing when it comes to testing software
+%for vulnerabilities. Smart contracts generated from ACTUS
+%specification are quite complicated, so manually testing every possible
+%execution path for every possible combination of contract terms is
+%not feasible. Static analysis, in turn, can reduce the amount of ``dumb''
+%mistakes that could cost millions to the users by ensuring obvious
+%properties of a contract. It is not a panacea, however, unlike with
+%fuzzing, if SA says that certain property holds - it does it with
+%100\% assurance.
 
@@
 values explicitly introduces risk of exceptions. However this risk
-is addressed using property-based testing, and in particular QuickCheck generators. This is discussed in more detail in Section \ref{verification} below.
+is addressed using property-based testing, and in particular QuickCheck generators. This is discussed in more detail in Section \ref{assurance} below.
 % SJT Added x-ref to section 7.
@@
 
-There are frameworks like ISDA Common Domain Model \cite{isda-cdm} providing 
-more precise representation of business processes involved in institutional trading 
-as well as code-generation capabilities. However, due to it's structural simplicity ACTUS 
+There are frameworks, like ISDA Common Domain Model \cite{isda-cdm}, providing a
+more precise representation of the business processes involved in institutional trading 
+as well as code-generation capabilities. However, due to its structural simplicity, ACTUS 
 is much more suited for generating code in a financial domain specific language 
@@
 
+We are grateful to Finley and Kegan McIlwaine of the University of Wyoming for their implementation of the LAM contract in ActusLabs, and to colleagues at Quanterall\footnote{\url{https://quanterall.com/}} for their contribution of utility functions to our ACTUS codebase.
+
 %

### 2021-02-05 papers/actus2021/article.tex
Completed introduction. Some tidying up of text and bibliography.

@@
 
-\bigskip\noindent
-List of contributions:
-\begin{itemize}
-\item \textquotedblleft Marlowe Labs\textquotedblright{} editor on Playground
-(Dmytro) 
-\item ACTUS Utility functions (Dmytro, Quanterall) 
-\item executable ACTUS spec for PAM (Dmytro, Alex) 
-\item executable ACTUS spec for LAM (Wyo Hackathon) 
-\item Library for generating ACTUS contracts (Dmytro) 
-\item ACTUS-like examples on Playground (Dmytro, Alex, Pablo) 
-\item Marlowe DSL enhancements for Actus: Cond, Mul (Dmytro)
-\item ACTUS spec generator for Agda (Dmytro) 
-\item Auto-refund static analysis (Pablo)
-\item Verification with QC + SMT (Dmytro) 
-\end{itemize}
-
 Smart contracts -- programs that run in a blockchain environment --  can be defined in a variety of ways~\cite{cryptoeprint:2016:1156}.
@@
 
-In this paper we explore ways in which contracts described in ACTUS (Algorithmic Contract Types Unified Standards) can be defined in Marlowe, Plutus and Haskell.
+In this paper we explore ways in which contracts described in ACTUS (Algorithmic Contract Types Unified Standards) can be defined in the contract langauges Marlowe, Plutus and Haskell.
 Of course, Plutus or Haskell are able to express these contracts, but rendering them in Marlowe brings extra advantages.
@@
 
-In implementing ACTUS we are able to provide further assurance in three other ways.
-First, we are able to use the declarative nature of Haskell to transliterate ACTUS formulae term-by-term into an executable form.
-Secondly, we are able to use random, property-based testing to validate the Haskell implementation against a Java one.
+In implementing ACTUS on Cardano we are able to provide further assurance in three other ways.
+First, we are able to use the declarative nature of Haskell to transliterate ACTUS formulas term-by-term into an executable form in Haskell.
+Secondly, we are able to use random, property-based testing to validate the Haskell implementation against another written in Java.
 Finally, we are able to automatically \emph{generate} ACTUS contracts in Marlowe from the terms -- i.e.
@@
 
-Implementing ACTUS provides a suitable benchmark against which to assess Marlowe, and we illustrate how the implementation has led to the addition of a conditional expression construct to the language.
+The contribution of our work is to show that that it is possible to implement financial contracts on blockchain in a way that \emph{multiple forms of assurance} are provided: from the language itself, from the static analysis, and from custom property verification.
+The development environment for Marlowe, the Marlowe Playground~\footnote{\url{https://alpha.marlowe.iohkdev.io/\#/}}, also provides a simulation environment for contracts for stepping forwards (and backwards) through contract execution, and thus allowing users to validate that contracts perform as they should.
+In addition, implementing ACTUS provides a suitable benchmark against which to assess the design of Marlowe; we illustrate how the implementation has led to the addition of a conditional expression construct to the language.
+
+In the remainder of the paper, Section \ref{background} covers the relevant financial background, including the ACTUS financial standard.
+Section \ref{executable} builds an executable specification of ACTUS in Haskell, and this is used in Section \ref{generation} to generate  the Marlowe code for an ACTUS contract from the contract terms.
+Section \ref{tokenization} explains how tokens are used to represent ownership of roles in a running contract, and Section \ref{assurance} describes how we provide assurance that  contracts behave as they should.
+Section \ref{related} examines related work and Section \ref{conclusion} concludes.
 
 A note on notation: \texttt{typewriter} font will be used for Marlowe constructs
-% SJT and fragments?
 while \emph{math} font will be used for mathematical formulas and pseudo-code.
-% Have I got this right?
-% dk14 looks right to me - I didn't know how to highlight Marlowe constructs properly
-
-% SJT: added this overview of the paper
-% it is traditional to end the introduction like this,
-% 'signposting' what is in the rest of the paper.
-Section \ref{background} of the paper covers the relevant financial background together with the ACTUS financial standard.
-Section \ref{executable} builds an executable specification of ACTUS in Haskell, and this is used in Section \ref{generation} to generate  the Marlowe code for an ACTUS contract from the terms of the contract.
-Section \ref{tokenization} explains how tokens are used to represent ownership of roles in a running contract, and Section \ref{assurance} describes how we provide assurance that  contracts behave as they should.
-Section \ref{related} examines related work and Section \ref{conclusion} concludes.
 
@@
 
-In this section we give a brief introduction to financial contracts, and loans in particular, and then describe the ACTUS financial standard.
+In this section we give a brief introduction to financial contracts, and to loans in particular, and then describe the ACTUS financial standard.
 
@@
 different variations of lending contracts. 
+% SJT moved from related work section.
+ACTUS provides additional benefit of being regulatory friendly, and the ACTUS
+foundation provides a set of tools allowing Monte-Carlo simulations
+of ACTUS contracts.
 
@@
 
-We provide assurance to users of our Marlowe ACTUS contracts in a number of ways, as described here.
+This section explains how we provide assurance to users of our Marlowe ACTUS contracts by means of property-based testing and SMT-based static analysis.
 
@@
 when it comes to automated market makers, using it for lending has
-been shown to be more susceptible to attacks\cite{flash-loan}.
+been shown to be more susceptible to attacks~\cite{flash-loan}.
 
@@
 - every loan is a separate contract, thus there is no global state.
-
-Actus provides additional benefit of being regulatory friendly, ACTUS
-foundation provides a set of tools allowing Monte-Carlo simulations
-of ACTUS contracts.
 
@@
 
-We are grateful to Finley and Kegan McIlwaine of the University of Wyoming for their implementation of the LAM contract in ActusLabs, and to colleagues at Quanterall\footnote{\url{https://quanterall.com/}} for their contribution of utility functions to our ACTUS codebase.
+We are very grateful to Finley and Kegan McIlwaine of the University of Wyoming for their implementation of the LAM contract in ActusLabs, and to colleagues at Quanterall%\footnote{\url{https://quanterall.com/}}
+for their contributions to the Marlowe ACTUS codebase.
 

### 2021-02-24 papers/actus2021/article.tex
typo fixed

@@
 
-In this paper we explore ways in which contracts described in ACTUS (Algorithmic Contract Types Unified Standards) can be defined in the contract langauges Marlowe, Plutus and Haskell.
+In this paper we explore ways in which contracts described in ACTUS (Algorithmic Contract Types Unified Standards) can be defined in the contract languages Marlowe, Plutus and Haskell.
 Of course, Plutus or Haskell are able to express these contracts, but rendering them in Marlowe brings extra advantages.

### 2021-02-24 papers/actus2021/article.tex
tidying up formulas

@@
 \begin{equation*}
-\ident{path}=\STF_{1}(ct,ev_{1})\circ \STF_{2}(ct,ev_{2}).\circ\ldots\circ \STF_{i}(ct,ev_{n})
+\ident{path}_{i}=\STF(ct,ev_{1})\circ \STF(ct,ev_{2})\circ\ldots\circ \STF(ct,ev_{i})
 \end{equation*}
@@
 \begin{equation*}
-\ident{state}_{i}=\ident{path}(\ident{INIT}(ct)),
+\ident{state}_{i}=\ident{path}_{i}(\ident{INIT}(ct)),
 \end{equation*}
@@
 where \emph{ct} stands for contract terms, \emph{INIT} returns initial state, \emph{sched}
-returns scheduled events, \emph{STF} takes \emph{contractterms}, \emph{event}, and \emph{state}
-and returns the next state, and \emph{POF} returns the payoff in a state.
+returns scheduled events, \emph{STF} takes \emph{contract terms}, \emph{event}, and \emph{state}
+and returns the next state, and \emph{POF} returns the \emph{payoff} from a state.
 

### 2021-02-26 papers/actus2021/article.tex
Typos

@@
 returns scheduled events, \emph{STF} takes \emph{contract terms}, \emph{event}, and \emph{state}
-and returns the next state, and \emph{POF} returns the \emph{payoff} from a state.
+and returns the next state, and \emph{POF} returns the \emph{payoff} in a state.
 

### 2021-03-24 papers/actus2021/article.tex
Minor edits and tidyings up prior to submission of version for publication.

@@
 source of liquidity for cryptocurrency holders. In this paper we
-explore a smart contract framework for building standardized crypto-loans
-using the Marlowe domain-specific language and with the ACTUS standard at its core.
+explore a smart contract framework for building standardised crypto-loans
+using the Marlowe domain-specific language and the ACTUS standard for financial contracts.
 
@@
 Smart contracts -- programs that run in a blockchain environment --  can be defined in a variety of ways~\cite{cryptoeprint:2016:1156}.
-Many such languages are general purpose, and can be used to program any kind of contract it makes sense to run on a blockchain; moreover, they tend to be expressive enough to be Turing complete (in some cases with restrictions on the runtime environment).
+Many such approaches are general purpose, and can be used to program any kind of contract it makes sense to run on a blockchain; moreover, they tend to be expressive enough to be Turing complete (in some cases with restrictions on the runtime environment).
 For example, Plutus, the general-purpose language running on the Cardano blockchain~\cite{utxo-vs-accounts}, is a dialect of Haskell.
@@
 Each Marlowe contract has a finite set of possible execution paths, and so it is possible to analyse the complete behaviour of a contract without running it.
-Such \emph{static analysis}, based on SMT solving~\cite{smt}, can be used to check properties of a contract; for example, it is possible to check whether a contract will honour all \texttt{Pay} constructs, however it is executed.
-In the case that a \texttt{Pay} can fail, the analysis gives an example trace showing how that failure can happen.
+Such \emph{static analysis}, based on SMT solving~\cite{smt}, can be used to check properties of a contract; for example, it is possible to check whether a contract will honour all the \texttt{Pay} constructs that it contains, however it is executed.
+In the case that a \texttt{Pay} can fail, the analysis gives an example trace showing how that failure happens.
 The language design and static analysis provide assurance that Marlowe contracts are much less likely to ``misbehave'' than contracts written in a general-purpose language like Solidity.
@@
 in a particular month the borrower would have to pay lower interest, and so would always pay a fixed share of the profits. In a more traditional
-setup the interest rate could depend on prices of other commodities: the canonical example would be a power plant taking a loan
+setup the interest rate could depend on prices of other commodities: a canonical example would be a power plant taking a loan
 with interest depending on electricity prices.
@@
 rate that the lender expects -- the \emph{benchmark yield} -- and
-the rate imposed by the driver - the higher the spread, the higher
+the rate imposed by the driver: the higher the spread, the higher
 the resulting interest rate. The multiplier %effectively 
@@
 
-\subsection{Utilizing polymorphism to abstract over basic operations}
+\subsection{Utilising polymorphism to abstract over basic operations}
 
@@
 
-In this section we describe how concrete Marlowe contracts are generate from the terms -- i.e.
-parameters -- of standard ACTUS contracts; we also describe the implementation of the system, and reflect on the limitations of generating contracts in Marlowe, where contracts have a a predefined lifetime and collection of interactions with contract participants.
+In this section we describe how concrete Marlowe contracts are generated from the terms -- i.e.
+parameters -- of standard ACTUS contracts.
+We also describe the implementation of the system, and reflect on the limitations of generating contracts in Marlowe, where contracts have a predefined lifetime and a predefined collection of interactions with contract participants.
 
@@
 
+\medskip
 \noindent 
@@
 
-\subsection{ActusLabs}
+\subsection{Actus Labs}
 
@@
 
+\smallskip
 \newcommand{\rf}{\ident{rf}}
@@
 
-Unlike current mainstream DeFi lending approaches - Marlowe ACTUS-based
-implementation tends to rely on trade-matching instead of pooling.
+Unlike current mainstream DeFi lending approaches, our Marlowe ACTUS implementation relies on trade-matching instead of pooling.
 While asset pooling is proven to be superior to order-book based approaches
@@
 been shown to be more susceptible to attacks~\cite{flash-loan}.
-
 Moving trade matching off-chain also improves scalability of the protocol
@@
 as well as code-generation capabilities. However, due to its structural simplicity, ACTUS 
-is much more suited for generating code in a financial domain specific language 
+is more suited for generating code in a financial domain specific language 
 like Marlowe rather than general-purpose one like Java or Haskell.
@@
 
-Marlowe language was explicitly designed as a set of building blocks
+The Marlowe language was explicitly designed as a set of building blocks
 for financial contracts that could be combined by anyone familiar
-with basic programming. Marlowe ACTUS generators improve on that by
-providing a way to automatically combine blocks based on standardized
-requirements specified by the user.
-
-On top of that, Marlowe ACTUS provides a toolkit for cross-testing
-with original ACTUS spec, a framework for adding new contract types,
+with basic programming. The Marlowe ACTUS generators improve on that by
+providing a way to automatically combine blocks based on standardised
+requirements specified by the user. Marlowe ACTUS also provides a toolkit for cross-testing
+against the original ACTUS spec, a framework for adding new contract types,
 cash-flow visualisations and verification tooling.
 
-We are very grateful to Finley and Kegan McIlwaine of the University of Wyoming for their implementation of the LAM contract in ActusLabs, and to colleagues at Quanterall%\footnote{\url{https://quanterall.com/}}
-for their contributions to the Marlowe ACTUS codebase.
+We are very grateful to colleagues at Quanterall and Finley and Kegan McIlwaine of the University of Wyoming for their %implementation of the LAM contract in Actus Labs, and to colleagues at Quanterall %\footnote{\url{https://quanterall.com/}}
+contributions to this project.
 

### 2021-04-08 papers/actus2021/article.tex
tidy up

@@
 				Alexander Nemish\inst{1} \and
-                Simon Thompson\inst{1,2}\orcidID{0000-0002-2350-301X}}
+                Simon Thompson\thanks{Corresponding author.}\inst{1,2}\orcidID{0000-0002-2350-301X}}
 

### 2022-10-01 isabelle/generated/SemanticsGuarantees.hs
Fix Serialization proofs

@@ -15,8 +15,10 @@ import qualified Orderings;
 less_eq_Party :: SemanticsTypes.Party -> SemanticsTypes.Party -> Bool;
 less_eq_Party (SemanticsTypes.Address uu) (SemanticsTypes.Role uv) = True;
 less_eq_Party (SemanticsTypes.Role uw) (SemanticsTypes.Address ux) = False;
-less_eq_Party (SemanticsTypes.Address x) (SemanticsTypes.Address y) = x <= y;
-less_eq_Party (SemanticsTypes.Role x) (SemanticsTypes.Role y) = x <= y;
+less_eq_Party (SemanticsTypes.Address x) (SemanticsTypes.Address y) =
+  Serialisation.less_eq_ByteString x y;
+less_eq_Party (SemanticsTypes.Role x) (SemanticsTypes.Role y) =
+  Serialisation.less_eq_ByteString x y;
 
 less_eq_Partya :: SemanticsTypes.Party -> SemanticsTypes.Party -> Bool;
 less_eq_Partya a b = less_eq_Party a b;

### 2022-10-01 isabelle/generated/Serialisation.hs

@@ -1,6 +1,6 @@
 {-# LANGUAGE EmptyDataDecls, RankNTypes, ScopedTypeVariables #-}
 
-module Serialisation(less_BS) where {
+module Serialisation(less_BS, less_eq_ByteString) where {
 
 import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
   (>>=), (>>), (=<<), (&&), (||), (^), (^^), (.), ($), ($!), (++), (!!), Eq,
@@ -11,5 +11,8 @@ import qualified Prelude;
 
 less_BS :: String -> String -> Bool;
 less_BS a b = not (b <= a);
+
+less_eq_ByteString :: String -> String -> Bool;
+less_eq_ByteString a b = a <= b;
 
 }

### 2022-10-04 isabelle/generated/ByteString.hs
Moved Serialisation to ByteString

@@ -1,6 +1,6 @@
 {-# LANGUAGE EmptyDataDecls, RankNTypes, ScopedTypeVariables #-}
 
-module Serialisation(less_BS, less_eq_ByteString) where {
+module ByteString(less_BS, less_eq_ByteString) where {
 
 import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
   (>>=), (>>), (=<<), (&&), (||), (^), (^^), (.), ($), ($!), (++), (!!), Eq,

### 2022-10-04 isabelle/generated/SemanticsGuarantees.hs

@@ -8,7 +8,7 @@ import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
   zip, null, takeWhile, dropWhile, all, any, Integer, negate, abs, divMod,
   String, Bool(True, False), Maybe(Nothing, Just));
 import qualified Prelude;
-import qualified Serialisation;
+import qualified ByteString;
 import qualified SemanticsTypes;
 import qualified Orderings;
 
@@ -16,9 +16,9 @@ less_eq_Party :: SemanticsTypes.Party -> SemanticsTypes.Party -> Bool;
 less_eq_Party (SemanticsTypes.Address uu) (SemanticsTypes.Role uv) = True;
 less_eq_Party (SemanticsTypes.Role uw) (SemanticsTypes.Address ux) = False;
 less_eq_Party (SemanticsTypes.Address x) (SemanticsTypes.Address y) =
-  Serialisation.less_eq_ByteString x y;
+  ByteString.less_eq_ByteString x y;
 less_eq_Party (SemanticsTypes.Role x) (SemanticsTypes.Role y) =
-  Serialisation.less_eq_ByteString x y;
+  ByteString.less_eq_ByteString x y;
 
 less_eq_Partya :: SemanticsTypes.Party -> SemanticsTypes.Party -> Bool;
 less_eq_Partya a b = less_eq_Party a b;
@@ -45,8 +45,8 @@ instance Orderings.Linorder SemanticsTypes.Party where {
 
 less_eq_Tok :: SemanticsTypes.Token -> SemanticsTypes.Token -> Bool;
 less_eq_Tok (SemanticsTypes.Token a b) (SemanticsTypes.Token c d) =
-  (if Serialisation.less_BS a c then True
-    else (if Serialisation.less_BS c a then False else b <= d));
+  (if ByteString.less_BS a c then True
+    else (if ByteString.less_BS c a then False else b <= d));
 
 less_eq_Token :: SemanticsTypes.Token -> SemanticsTypes.Token -> Bool;
 less_eq_Token a b = less_eq_Tok a b;
@@ -79,7 +79,7 @@ less_eq_ValueId a b = less_eq_ValId a b;
 
 less_ValId :: SemanticsTypes.ValueId -> SemanticsTypes.ValueId -> Bool;
 less_ValId (SemanticsTypes.ValueId a) (SemanticsTypes.ValueId b) =
[…]

### 2022-10-12 isabelle/Doc/Specification/document/json.tex
Configure literate haskell for the json spec

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{module}\;\Conid{\Conid{Language}.\Conid{Marlowe}.Json}\;\mathbf{where}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{testLiterate}\mathbin{::}\Conid{IO}\;(){}\<[E]%
\\
\>[B]{}\Varid{testLiterate}\mathrel{=}\Varid{putStrLn}\;\text{\ttfamily \char34 Yes,~this~is~literate~haskell.~Thanks~sed!\char34}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

### 2022-10-13 isabelle/Doc/Specification/document/json.tex
Moved json instances to isabelle

@@
+The Json specification for Marlowe Core is defined in Literate Haskell using the Aeson library. In order
+to fully understand the specification, some knowledge of the library is recommended. For each type we
+define a way to parse the JSON value to the datatype (FromJSON instances) and a way to serialize the
+datatype to JSON (ToJSON instances).
+
+\isamarkupsubsection{Party}
+
 \begin{hscode}\SaveRestoreHook
 \column{B}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
+\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
 \column{E}{@{}>{\hspre}l<{\hspost}@{}}%
-\>[B]{}\mathbf{module}\;\Conid{\Conid{Language}.\Conid{Marlowe}.Json}\;\mathbf{where}{}\<[E]%
+\>[B]{}\mathbf{instance}\;\Conid{ToJSON}\;\Conid{Party}\;\mathbf{where}{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{toJSON}\;(\Conid{Address}\;\Varid{address})\mathrel{=}{}\<[E]%
+\\
+\>[5]{}\hsindent{4}{}\<[9]%
+\>[9]{}\Varid{object}\;[\mskip1.5mu \text{\ttfamily \char34 address\char34}\mathbin{.=}\Varid{address}\mskip1.5mu]{}\<[E]%
+\\
+\>[B]{}\hsindent{5}{}\<[5]%
+\>[5]{}\Varid{toJSON}\;(\Conid{Role}\;\Varid{name})\mathrel{=}{}\<[E]%
+\\
+\>[5]{}\hsindent{4}{}\<[9]%
+\>[9]{}\Varid{object}\;[\mskip1.5mu \text{\ttfamily \char34 role\char95 token\char34}\mathbin{.=}\Varid{name}\mskip1.5mu]{}\<[E]%
 \\[\blanklineskip]%
-\>[B]{}\Varid{testLiterate}\mathbin{::}\Conid{IO}\;(){}\<[E]%
+\>[B]{}\Varid{addressExample}\mathbin{::}\Conid{Party}{}\<[E]%
 \\
-\>[B]{}\Varid{testLiterate}\mathrel{=}\Varid{putStrLn}\;\text{\ttfamily \char34 Yes,~this~is~literate~haskell.~Thanks~sed!\char34}{}\<[E]%
+\>[B]{}\Varid{addressExample}\mathrel{=}\Conid{Address}\;\text{\ttfamily \char34 example~address\char34}{}\<[E]%
+\\[\blanklineskip]%
+\>[B]{}\Varid{roleExample}\mathbin{::}\Conid{Party}{}\<[E]%
+\\
+\>[B]{}\Varid{roleExample}\mathrel{=}\Conid{Role}\;\text{\ttfamily \char34 example~role\char34}{}\<[E]%
 \ColumnHook
 \end{hscode}\resethooks
+\par
+
+The example address is serialized as \ensuremath{\{\mskip1.5mu \text{\ttfamily \char34 address\char34}\mathbin{:}\text{\ttfamily \char34 example~address\char34}\mskip1.5mu\}}, and the role
+example as \ensuremath{\{\mskip1.5mu \text{\ttfamily \char34 role\char95 token\char34}\mathbin{:}\text{\ttfamily \char34 example~role\char34}\mskip1.5mu\}}

### 2022-10-13 isabelle/haskell/CoreOrphanEq.hs

@@ -1,0 +1,25 @@
+{-# LANGUAGE OverloadedStrings #-}
+
+-- This module has some missing Eq instances that werent exported by isabelle
+-- directly. Ideally we should fix this from the Isabelle CodeExports.thy
+-- theory.
+module CoreOrphanEq where
+
+import SemanticsTypes (Contract, Payee, equal_Contract, equal_Payee)
+import Semantics (TransactionWarning, Payment, equal_TransactionWarning, equal_Payment)
+
+instance Eq Contract where {
+  a == b = equal_Contract a b;
+};
+
+instance Eq TransactionWarning where {
+  a == b = equal_TransactionWarning a b;
+}
+
+instance Eq Payment where {
+  a == b = equal_Payment a b;
+}
+
+instance Eq Payee where {
+  a == b = equal_Payee a b;
+}

### 2022-10-13 isabelle/test/Spec.hs

@@ -24,6 +24,7 @@ module Main
 import Test.Tasty (TestTree, defaultMain, testGroup)
 import Test.Tasty.HUnit (assertBool, assertFailure, testCase, (@=?))
 import qualified Spec.Core.Examples.Swap (tests)
+import qualified Spec.Core.Serialization.Json (tests)
 
 -- | Entry point for the tests.
 main :: IO ()
@@ -34,6 +35,6 @@ main = defaultMain tests
 tests :: TestTree
 tests =
   testGroup "Marlowe Spec"
-    [
-      Spec.Core.Examples.Swap.tests
+    [ Spec.Core.Examples.Swap.tests
+    , Spec.Core.Serialization.Json.tests
     ]

### 2022-10-13 isabelle/test/Spec/Core/Examples/Swap.hs

@@ -7,27 +7,14 @@ module Spec.Core.Examples.Swap (tests) where
 import qualified Arith;
 import Test.Tasty (TestTree, testGroup)
 import Test.Tasty.HUnit (assertBool, assertFailure, testCase, (@?=))
-import SemanticsTypes (Contract (..), Party (..), Payee(..), Token(..), Value(..), Case(..), Action(..), Input(..), equal_Contract)
-import Semantics (TransactionOutput(..), Transaction_ext(..), TransactionWarning(..), Payment(..), playTrace, txOutContract, txOutWarnings, txOutPayments, equal_TransactionWarning, equal_Payment)
+import CoreOrphanEq
+import SemanticsTypes (Contract (..), Party (..), Payee(..), Token(..), Value(..), Case(..), Action(..), Input(..))
+import Semantics (TransactionOutput(..), Transaction_ext(..), TransactionWarning(..), Payment(..), playTrace, txOutContract, txOutWarnings, txOutPayments)
 import qualified Examples.Swap
 
 -- FIXME: Isabelle doesn't export type synonims by default, see if we can fix that or
 --        make a common module that acts as a wrapper with the aliases and the orphan instances.
 type Timeout = Arith.Int
-
--- FIXME: I wasn't able to export some Eq instances (it does export the equal_X functions). See if
---        we can resolve that or at least put the manual instances in a separate module.
-instance Eq Contract where {
-  a == b = equal_Contract a b;
-};
-
-instance Eq TransactionWarning where {
-  a == b = equal_TransactionWarning a b;
-}
-
-instance Eq Payment where {
-  a == b = equal_Payment a b;
-}
 
 ------- Contract definition -------
 

### 2022-10-13 isabelle/test/Spec/Core/Serialization/Json.hs

@@ -1,0 +1,104 @@
+{-# LANGUAGE CPP #-}
+{-# LANGUAGE OverloadedStrings #-}
+
+module Spec.Core.Serialization.Json (tests) where
+
+import Data.Aeson (decode, encode)
+import qualified Data.ByteString.Lazy as L
+import MarloweCoreJson (addressExample, roleExample, dolarToken, internalPayeeExample)
+import Test.Tasty (TestTree, TestName, testGroup)
+import Test.Tasty.HUnit (testCase, (@?=))
+import Test.Tasty.Golden (goldenVsStringDiff)
+import System.FilePath ((</>))
+
+diffCommand :: FilePath -> FilePath -> [String]
+diffCommand golden actual = ["diff", "-u", "--color=always", actual, golden]
+
+goldenPath :: FilePath
+goldenPath = "test" </> "Spec" </> "Core" </> "Serialization" </> "golden"
+
+goldenTest :: TestName -> FilePath -> IO L.ByteString -> TestTree
+goldenTest name path io = goldenVsStringDiff name diffCommand path io
+
+tests :: TestTree
+tests = testGroup "Core Json"
+    [ partyTests
+    , tokenTests
+    , payeeTests
+    ]
+
+-- === Party ===
+
+partyTests :: TestTree
+partyTests = testGroup "Party"
+    [ testCase "Party Roundtrip" testExampleAddressRoundtrip
+    , testGoldenPartyAddress
+    , testGoldenPartyRole
+    ]
+
+
+-- TODO: Generate Gen instances and change for a property based test.
+testExampleAddressRoundtrip :: IO ()
+testExampleAddressRoundtrip =
+    let
+        roundtrip = decode $ encode addressExample
+    in
+        roundtrip @?= Just addressExample
+
+testGoldenPartyAddress :: TestTree
+testGoldenPartyAddress = goldenTest
+    "Golden Party Address"
+    (goldenPath </> "party-address.golden")
+    (pure $ encode addressExample)
+
+testGoldenPartyRole :: TestTree
+testGoldenPartyRole = goldenTest
+    "Golden Party Role"
+    (goldenPath </> "party-role.golden")
+    (pure $ encode roleExample)
+
+
+-- === Token ===
+tokenTests :: TestTree
+tokenTests = testGroup "Token"
+    [ testCase "Token Roundtrip" testTokenRoundtrip
+    , testGoldenDolarToken
+    ]
+
+
[…]

### 2022-10-26 isabelle/generated/Semantics.hs
wip8

@@ -2,14 +2,14 @@
 
 module
   Semantics(Payment(..), ReduceEffect, ReduceWarning, ReduceResult,
-             TransactionWarning(..), ApplyAllResult, OpenTimeInterval,
-             ReduceStepResult, TransactionError(..),
-             TransactionOutputRecord_ext, TransactionOutput(..),
-             Transaction_ext(..), evalValue, evalObservation, txOutWarnings,
-             txOutPayments, reductionLoop, reduceContractUntilQuiescent,
-             applyAllInputs, computeTransaction, playTrace, getOutcomes,
-             maxTimeContract, getSignatures, calculateTimeInterval,
-             equal_Payment, txOutState, txOutContract, equal_TransactionWarning)
+             TransactionWarning(..), ApplyAllResult, ReduceStepResult,
+             TransactionError(..), TransactionOutputRecord_ext,
+             TransactionOutput(..), Transaction_ext(..), evalValue,
+             evalObservation, txOutWarnings, txOutPayments, reductionLoop,
+             reduceContractUntilQuiescent, applyAllInputs, computeTransaction,
+             playTrace, getOutcomes, maxTimeContract, getSignatures,
+             calculateNonAmigousInterval, equal_Payment, txOutState,
+             txOutContract, equal_TransactionWarning)
   where {
 
 import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
@@ -18,14 +18,15 @@ import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
   zip, null, takeWhile, dropWhile, all, any, Integer, negate, abs, divMod,
   String, Bool(True, False), Maybe(Nothing, Just));
 import qualified Prelude;
+import qualified OptBoundTimeInterval;
 import qualified List;
+import qualified Orderings;
 import qualified Option;
 import qualified MList;
 import qualified HOL;
 import qualified Product_Lexorder;
 import qualified Product_Type;
 import qualified ListTools;
-import qualified Orderings;
 import qualified SList;
 import qualified SemanticsGuarantees;
[…]

### 2022-10-26 isabelle/generated/Semantics.hs
Updated TimeRange and fixed typo

@@ -8,7 +8,7 @@ module
              evalObservation, txOutWarnings, txOutPayments, reductionLoop,
              reduceContractUntilQuiescent, applyAllInputs, computeTransaction,
              playTrace, getOutcomes, maxTimeContract, getSignatures,
-             calculateNonAmigousInterval, equal_Payment, txOutState,
+             calculateNonAmbiguousInterval, equal_Payment, txOutState,
              txOutContract, equal_TransactionWarning)
   where {
 
@@ -692,39 +692,39 @@ maxTimeContract (SemanticsTypes.Assert vb contract) = maxTimeContract contract;
 getSignatures :: [SemanticsTypes.Input] -> [SemanticsTypes.Party];
 getSignatures l = List.foldl addSig SList.empty l;
 
-calculateNonAmigousInterval ::
+calculateNonAmbiguousInterval ::
   Maybe Arith.Int ->
     Arith.Int ->
       SemanticsTypes.Contract ->
         (OptBoundTimeInterval.BEndpoint, OptBoundTimeInterval.BEndpoint);
-calculateNonAmigousInterval uu uv SemanticsTypes.Close =
+calculateNonAmbiguousInterval uu uv SemanticsTypes.Close =
   (OptBoundTimeInterval.Unbounded, OptBoundTimeInterval.Unbounded);
-calculateNonAmigousInterval n t (SemanticsTypes.Pay uw ux uy uz c) =
-  calculateNonAmigousInterval n t c;
-calculateNonAmigousInterval n t (SemanticsTypes.If va ct cf) =
-  OptBoundTimeInterval.intersectInterval (calculateNonAmigousInterval n t ct)
-    (calculateNonAmigousInterval n t cf);
-calculateNonAmigousInterval n t (SemanticsTypes.When [] timeout tcont) =
+calculateNonAmbiguousInterval n t (SemanticsTypes.Pay uw ux uy uz c) =
+  calculateNonAmbiguousInterval n t c;
+calculateNonAmbiguousInterval n t (SemanticsTypes.If va ct cf) =
+  OptBoundTimeInterval.intersectInterval (calculateNonAmbiguousInterval n t ct)
+    (calculateNonAmbiguousInterval n t cf);
+calculateNonAmbiguousInterval n t (SemanticsTypes.When [] timeout tcont) =
   (if Arith.less_int t timeout
     then (OptBoundTimeInterval.Unbounded,
[…]

### 2022-10-26 isabelle/generated/Semantics.hs
PR review notes (#144)
Co-authored-by: Pablo Lamela <pablo.lamela@iohk.io>

@@ -8,8 +8,8 @@ module
              evalObservation, txOutWarnings, txOutPayments, reductionLoop,
              reduceContractUntilQuiescent, applyAllInputs, computeTransaction,
              playTrace, getOutcomes, maxTimeContract, getSignatures,
-             calculateTimeInterval, equal_Payment, txOutState, txOutContract,
-             equal_TransactionWarning)
+             calculateNonAmbiguousInterval, equal_Payment, txOutState,
+             txOutContract, equal_TransactionWarning)
   where {
 
 import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
@@ -18,14 +18,15 @@ import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
   zip, null, takeWhile, dropWhile, all, any, Integer, negate, abs, divMod,
   String, Bool(True, False), Maybe(Nothing, Just));
 import qualified Prelude;
+import qualified OptBoundTimeInterval;
 import qualified List;
+import qualified Orderings;
 import qualified Option;
 import qualified MList;
 import qualified HOL;
 import qualified Product_Lexorder;
 import qualified Product_Type;
 import qualified ListTools;
-import qualified Orderings;
 import qualified SList;
 import qualified SemanticsGuarantees;
 import qualified SemanticsTypes;
@@ -112,16 +113,6 @@ addSig acc (SemanticsTypes.IDeposit uu p uv uw) = SList.insert p acc;
 addSig acc (SemanticsTypes.IChoice (SemanticsTypes.ChoiceId ux p) uy) =
   SList.insert p acc;
 addSig acc SemanticsTypes.INotify = acc;
-
-maxOpt :: Maybe Arith.Int -> Maybe Arith.Int -> Maybe Arith.Int;
-maxOpt Nothing y = y;
-maxOpt (Just v) Nothing = Just v;
-maxOpt (Just x) (Just y) = Just (Orderings.max x y);
-
-minOpt :: Maybe Arith.Int -> Maybe Arith.Int -> Maybe Arith.Int;
-minOpt Nothing y = y;
-minOpt (Just v) Nothing = Just v;
-minOpt (Just x) (Just y) = Just (Orderings.min x y);
 
 gtIfNone :: Maybe Arith.Int -> Arith.Int -> Bool;
 gtIfNone Nothing uu = True;
[…]

### 2022-12-05 marlowe-spec-test/src/Marlowe/Spec/Core/Arbitrary.hs
Add implementation for GenerateRandomValue for party

@@ -5,8 +5,8 @@
 
 module Marlowe.Spec.Core.Arbitrary where
 
-import SemanticsTypes (Token(..))
-import QuickCheck.GenT (GenT)
+import SemanticsTypes (Token(..), Party)
+import QuickCheck.GenT (GenT, frequency)
 import Marlowe.Spec.Interpret (InterpretJsonRequest, Request (..), parseValidResponse)
 import Data.Data (Proxy(..))
 import Marlowe.Spec.TypeId (TypeId(..))
@@ -45,4 +45,13 @@ arbitraryToken interpret = liftIO do
   case parseValidResponse res of
     Left err -> throwIO $ GenerateRandomValueException err
     Right (UnknownType _) -> throwIO $ GenerateRandomValueException "Client process doesn't know how to generate Core.Token"
+    Right (RandomValue t) -> pure t
+
+
+arbitraryParty :: InterpretJsonRequest -> GenT IO Party
+arbitraryParty interpret = liftIO do
+  res <- interpret (GenerateRandomValue $ TypeId "Core.Party" (Proxy :: Proxy Party))
+  case parseValidResponse res of
+    Left err -> throwIO $ GenerateRandomValueException err
+    Right (UnknownType _) -> throwIO $ GenerateRandomValueException "Client process doesn't know how to generate Core.Party"
     Right (RandomValue t) -> pure t

### 2022-12-05 marlowe-spec-test/src/Marlowe/Spec/Core/Serialization/Json.hs

@@ -16,12 +16,12 @@ import Data.Proxy (Proxy(..))
 import MarloweCoreJson
 import GHC.Stack (HasCallStack)
 import Test.Tasty (TestTree, testGroup)
-import Marlowe.Spec.Interpret (Response (..), InterpretJsonRequest, exactMatch, Request (..), testResponse)
+import Marlowe.Spec.Interpret (Response (..), InterpretJsonRequest, Request (..), testResponse)
 import Marlowe.Spec.TypeId (TypeId(..), HasTypeId (..))
 import Test.Tasty.HUnit (Assertion, assertBool, testCase, (@?=))
 import qualified SemanticsTypes as C
 import QuickCheck.GenT (runGenT)
-import Marlowe.Spec.Core.Arbitrary (arbitraryToken)
+import Marlowe.Spec.Core.Arbitrary (arbitraryToken, arbitraryParty)
 import Test.QuickCheck (generate)
 
 
@@ -58,6 +58,7 @@ tests i = testGroup "Json Serialization"
   , observationTests i
   , invalidType i
   , tokenTest i
+  , partyTest i
   ]
 
 valueTests :: InterpretJsonRequest -> TestTree
@@ -105,9 +106,16 @@ observationTests i = testGroup "Observation examples"
 -- SCP-4696 Improve thread usage
 tokenTest :: InterpretJsonRequest -> TestTree
 tokenTest i = testCase "Token test" $ do
-  -- Any token that is randomly generated should pass the roundtrip test
+  -- Any token that is randomly generated by the interpreter should also pass the roundtrip test
   token <- join $ generate $ runGenT $ arbitraryToken i
   roundtripTest i token
+
+partyTest :: InterpretJsonRequest -> TestTree
+partyTest i = testCase "Party test" $ do
+  -- Any party that is randomly generated by the interpreter should also pass the roundtrip test
+  party <- join $ generate $ runGenT $ arbitraryParty i
+  roundtripTest i party
+
 
 invalidType :: InterpretJsonRequest -> TestTree
 invalidType i = testResponse i "Invalid type"

### 2022-12-05 marlowe-spec-test/src/Marlowe/Spec/LocalInterpret.hs

@@ -11,7 +11,7 @@ import Marlowe.Spec.TypeId (TypeId (..), fromTypeName)
 import Marlowe.Spec.Core.Serialization.Json
 import Data.Data (Proxy)
 import Data.Aeson (Result (..),FromJSON,ToJSON)
-import SemanticsTypes (Token(Token))
+import SemanticsTypes (Token(Token), Party (..))
 import Test.QuickCheck (Gen, frequency, Arbitrary (arbitrary), generate)
 import qualified Marlowe.Spec.Core.Arbitrary as RandomResponse
 
@@ -37,8 +37,8 @@ interpretLocal (GenerateRandomValue t@(TypeId name _)) =
     <$> JSON.toJSON
     <$> case name of
       "Core.Token" -> RandomResponse.RandomValue <$> JSON.toJSON <$> generate arbitraryToken
+      "Core.Party" -> RandomResponse.RandomValue <$> JSON.toJSON <$> generate arbitraryParty
       _ -> pure $ RandomResponse.UnknownType t
-
 
 arbitraryToken :: Gen Token
 arbitraryToken =
@@ -46,6 +46,51 @@ arbitraryToken =
     [(50, pure $ Token "" "")
     ,(50, Token <$> arbitrary <*> arbitrary)
     ]
+
+
+-- | Some role names.
+randomRoleNames :: [String]
+randomRoleNames =
+  [
+    "Cy"
+  , "Noe"
+  , "Sten"
+  , "Cara"
+  , "Alene"
+  , "Hande"
+  , ""
+  , "I"
+  , "Zakkai"
+  , "Laurent"
+  , "Prosenjit"
+  , "Dafne Helge Mose"
+  , "Nonso Ernie Blanka"
+  , "Umukoro Alexander Columb"
+  , "Urbanus Roland Alison Ty Ryoichi"
+  , "Alcippe Alende Blanka Roland Dafne"  -- NB: Too long for Cardano ledger.
+  ]
+
+-- | Part of the Fibonacci sequence.
+fibonaccis :: Num a => [a]
+fibonaccis = [2, 3, 5, 8, 13, 21, 34, 55, 89, 144, 233, 377, 610, 987, 1597, 2584]
+
+
+-- | Inverse-Fibanoncci frequencies.
+fibonacciFrequencies :: Integral a => [a]
+fibonacciFrequencies = (1000000 `div`) <$> fibonaccis
+
+
+-- | Select an element of a list with propability proportional to inverse-Fibonacci weights.
[…]

### 2023-03-01 isabelle/generated/Examples/Escrow.hs
SCP-4908: Added Escrow contract

@@ -1,0 +1,564 @@
+{-# LANGUAGE EmptyDataDecls, RankNTypes, ScopedTypeVariables #-}
+
+module
+  Examples.Escrow(EscrowArgs_ext, escrowExample, confirmClaimPayments,
+                   dismissClaimPayments, confirmProblemPayments,
+                   confirmClaimTransactions, dismissClaimTransactions,
+                   confirmProblemTransactions, everythingIsAlrightPayments,
+                   everythingIsAlrightTransactions)
+  where {
+
+import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
+  (>>=), (>>), (=<<), (&&), (||), (^), (^^), (.), ($), ($!), (++), (!!), Eq,
+  error, id, return, not, fst, snd, map, filter, concat, concatMap, reverse,
+  zip, null, takeWhile, dropWhile, all, any, Integer, negate, abs, divMod,
+  String, Bool(True, False), Maybe(Nothing, Just));
+import qualified Prelude;
+import qualified Semantics;
+import qualified Stringa;
+import qualified SemanticsTypes;
+import qualified Arith;
+
+data EscrowArgs_ext a =
+  EscrowArgs_ext SemanticsTypes.Value SemanticsTypes.Token SemanticsTypes.Party
+    SemanticsTypes.Party SemanticsTypes.Party Arith.Int Arith.Int Arith.Int
+    Arith.Int a
+  deriving (Prelude.Read, Prelude.Show);
+
+paymentDeadline :: forall a. EscrowArgs_ext a -> Arith.Int;
+paymentDeadline
+  (EscrowArgs_ext price token seller buyer mediator paymentDeadline
+    complaintDeadline disputeDeadline mediationDeadline more)
+  = paymentDeadline;
+
+seller :: forall a. EscrowArgs_ext a -> SemanticsTypes.Party;
+seller
+  (EscrowArgs_ext price token seller buyer mediator paymentDeadline
+    complaintDeadline disputeDeadline mediationDeadline more)
+  = seller;
+
+price :: forall a. EscrowArgs_ext a -> SemanticsTypes.Value;
+price (EscrowArgs_ext price token seller buyer mediator paymentDeadline
+        complaintDeadline disputeDeadline mediationDeadline more)
+  = price;
+
+buyer :: forall a. EscrowArgs_ext a -> SemanticsTypes.Party;
[…]

### 2023-03-01 marlowe-spec-test/src/Marlowe/Spec/Core/Examples.hs

@@ -2,10 +2,12 @@
 
 module Marlowe.Spec.Core.Examples (tests) where
 import Test.Tasty (TestTree, testGroup)
+import qualified Marlowe.Spec.Core.Examples.Escrow
 import qualified Marlowe.Spec.Core.Examples.Swap
 import Marlowe.Spec.Interpret (InterpretJsonRequest)
 
 tests :: InterpretJsonRequest -> TestTree
 tests i = testGroup "Contract examples"
   [ Marlowe.Spec.Core.Examples.Swap.tests i
+  , Marlowe.Spec.Core.Examples.Escrow.tests i
   ]

### 2023-03-01 marlowe-spec-test/src/Marlowe/Spec/Core/Examples/Escrow.hs

@@ -1,0 +1,43 @@
+{-# LANGUAGE OverloadedStrings #-}
+
+module Marlowe.Spec.Core.Examples.Escrow (tests) where
+
+import Test.Tasty (TestTree, testGroup)
+import Test.Tasty.HUnit (assertFailure, testCase, (@?=))
+import Marlowe.Spec.Interpret (InterpretJsonRequest, Request (..), parseValidResponse)
+import Semantics (TransactionOutput(..), txOutContract, txOutWarnings, txOutPayments, Transaction_ext (..), Payment (..))
+import SemanticsTypes (Contract(..))
+import Examples.Escrow (confirmProblemPayments, dismissClaimPayments, confirmClaimPayments, escrowExample, confirmClaimTransactions, dismissClaimTransactions, confirmProblemTransactions, everythingIsAlrightPayments, everythingIsAlrightTransactions)
+
+tests :: InterpretJsonRequest -> TestTree
+tests i = testGroup "Escrow"
+    [ everythingIsAlright i
+    , confirmProblem i
+    , dismissClaim i
+    , confirmClaim i
+    ]
+
+everythingIsAlright :: InterpretJsonRequest -> TestTree
+everythingIsAlright = runScenario "Everything is alright" everythingIsAlrightTransactions everythingIsAlrightPayments
+
+confirmProblem :: InterpretJsonRequest -> TestTree
+confirmProblem = runScenario "Confirm Problem" confirmProblemTransactions confirmProblemPayments
+
+dismissClaim :: InterpretJsonRequest -> TestTree
+dismissClaim = runScenario "Dismiss Claim" dismissClaimTransactions dismissClaimPayments
+
+confirmClaim :: InterpretJsonRequest -> TestTree
+confirmClaim = runScenario "Confirm Claim" confirmClaimTransactions confirmClaimPayments
+
+runScenario :: String -> [Transaction_ext ()] -> [Payment] -> InterpretJsonRequest -> TestTree
+runScenario name transactions payments interpret = testCase name
+    ( do
+        res <- interpret $ PlayTrace 0 escrowExample transactions
+        case parseValidResponse res of
+            Left err -> assertFailure err
+            Right (TransactionError err) -> assertFailure $ "Transaction error: " ++ show err
+            Right (TransactionOutput o) -> do
+              txOutContract o @?= Close
+              txOutWarnings o @?= []
+              txOutPayments o @?= payments
+    )

### 2023-03-27 marlowe-spec-test/src/Marlowe/Spec/Core/Semantics.hs
PLT-3313: `playTrace` from spec-test

@@ -50,7 +50,6 @@ import Semantics
     evalObservation,
     fixInterval,
     isQuiescent,
-    playTrace,
     reduceContractUntilQuiescent,
     txOutWarnings,
     maxTimeContract,
@@ -226,10 +225,13 @@ traceToSingleInputIsEquivalentTest interpret = reproducibleProperty "Single inpu
     transactions <- run $ generateT $ (listOf $ genTransaction interpret) `suchThat` \t -> t /= traceListToSingleInput t
 
     let
-      multipleInputsResponse = RequestResponse $ toJSON $ playTrace startTime contract transactions
-      singletonInputResponse = RequestResponse $ toJSON $ playTrace startTime contract (traceListToSingleInput transactions)
+        multipleInputs = PlayTrace (integer_of_int startTime) contract transactions
+        singletonInput = PlayTrace (integer_of_int startTime) contract (traceListToSingleInput transactions)
 
-    assert $ multipleInputsResponse == singletonInputResponse
+    RequestResponse resMultipleInputs <- run $ liftIO $ interpret multipleInputs
+    RequestResponse resSingletonInput <- run $ liftIO $ interpret singletonInput
+
+    assert $ resMultipleInputs == resSingletonInput
 
 integer_of_int :: Arith.Int -> Integer
 integer_of_int (Arith.Int_of_integer k) = k

### 2023-03-27 marlowe-spec-test/src/Marlowe/Spec/Core/Arbitrary.hs
PLT-3313: multi-element lists for inputs

@@ -18,6 +18,7 @@ module Marlowe.Spec.Core.Arbitrary
   , genState
   , genEnvironment
   , genContract
+  , genGoldenContract
   , genInterval
   , genTransaction
   , genAction
@@ -28,6 +29,7 @@ module Marlowe.Spec.Core.Arbitrary
   , genTransactionError
   , genTransactionOutput
   , arbitraryValidInputs
+  , arbitraryValidStep
   , arbitraryNonnegativeInteger
   , arbitraryFibonacci
   , arbitraryPositiveInteger
@@ -244,11 +246,11 @@ genParty interpret = do
       Right (RandomValue t) -> pure t
 
 genContract :: InterpretJsonRequest -> GenT IO Contract
-genContract interpret = frequency [(90, gen =<< liftGen arbitrary), (10, goldenContracts interpret)]
+genContract interpret = frequency [(90, gen =<< liftGen arbitrary), (10, genGoldenContract interpret)]
   where gen context = sized \size -> arbitraryContractSized (min (size `div` 6) 5) context interpret -- Keep tests from growing too large to execute by capping the maximum contract depth at 5 (default size is 30)
 
-goldenContracts :: InterpretJsonRequest -> GenT IO Contract
-goldenContracts interpret =
+genGoldenContract :: InterpretJsonRequest -> GenT IO Contract
+genGoldenContract interpret =
   frequency
     [ (50, escrow),
       (50, swap)
@@ -534,14 +536,17 @@ arbitraryValidStep state (When cases timeout _) =
       else
         do
           times <- arbitraryTimeIntervalBefore minTime' timeout
-          case' <- QC.elements $ filter (not . isEmptyChoice . getAction) cases
-          i <- case getAction case' of
+          Case action cont <- QC.elements $ filter (not . isEmptyChoice . getAction) cases
+          i <- case action of
                  Deposit a p t v -> pure . IDeposit a p t $ evalValue (Environment_ext times ()) state v
                  Choice n bs     -> do
                                       Bound lower upper <- QC.elements bs
                                       IChoice n <$> chooseinteger (lower, upper)
                  Notify _        -> pure INotify
-          pure $ Transaction_ext times [i] ()
+          case cont of
+            Close -> pure $ Transaction_ext times [i] ()
+            _ -> do Transaction_ext _ inputs _ <- arbitraryValidStep state cont
+                    pure $ Transaction_ext times (i:inputs) ()
   where
     getAction :: Case -> Action
     getAction (Case a _) = a

### 2023-03-27 marlowe-spec-test/src/Marlowe/Spec/Core/Semantics.hs

@@ -11,8 +11,7 @@ import Control.Monad.IO.Class (MonadIO (..))
 import Data.Aeson (ToJSON (..))
 import qualified Data.Aeson as JSON
 import Marlowe.Spec.Core.Arbitrary
-  ( arbitraryNonnegativeInteger,
-    arbitraryTimeIntervalAfter,
+  ( arbitraryTimeIntervalAfter,
     arbitraryValidInputs,
     genContract,
     genEnvironment,
@@ -38,7 +37,7 @@ import Marlowe.Spec.Reproducible
 import Marlowe.Utils (showAsJson)
 import Orderings (Ord (..))
 import PositiveAccounts (validAndPositive_state)
-import QuickCheck.GenT (listOf, suchThat)
+import QuickCheck.GenT (suchThat, liftGen)
 import Semantics
   ( ReduceResult (..),
     TransactionOutput (..),
@@ -220,9 +219,11 @@ playTrace_only_accepts_maxTransactionsInitialStateTest interpret = reproducibleP
 --    "playTrace sn co tral = playTrace sn co (traceListToSingleInput tral)"
 traceToSingleInputIsEquivalentTest :: InterpretJsonRequest -> TestTree
 traceToSingleInputIsEquivalentTest interpret = reproducibleProperty "Single input transactions"  do
-    contract <- run $ generateT $ genContract interpret `suchThat` (/=Close) -- arbitraryValidInputs returns [] for the `Close` contract
-    startTime <- run $ generate arbitraryNonnegativeInteger
-    transactions <- run $ generateT $ (listOf $ genTransaction interpret) `suchThat` \t -> t /= traceListToSingleInput t
+    (contract, State_ext _ _ _ startTime _, transactions) <- run $ generateT $ (do
+        c <- genContract interpret `suchThat` (/=Close) -- arbitraryValidInputs returns [] for the `Close` contract
+        s <- genState interpret
+        t <- liftGen $ arbitraryValidInputs s c
+        pure (c,s,t)) `suchThat` \(_,_,t) -> (t /= traceListToSingleInput t)
 
     let
         multipleInputs = PlayTrace (integer_of_int startTime) contract transactions

### 2023-03-27 marlowe-spec-test/src/Marlowe/Spec/Core/Arbitrary.hs
PLT-3313: Unused field

@@ -53,9 +53,7 @@ import Control.Monad (liftM2)
 import Control.Monad.IO.Class (liftIO)
 import Data.Aeson (FromJSON (..), ToJSON (..), object, withObject, (.:), (.=))
 import Data.Data (Proxy (..))
-import Data.Function (on)
-import Data.List (nub, nubBy)
-import Data.Map (Map, fromList)
+import Data.List (nub)
 import Marlowe.Spec.Interpret (InterpretJsonRequest, Request (..), parseValidResponse)
 import Marlowe.Spec.TypeId (TypeId (..))
 import Orderings (Ord (..), max)
@@ -666,11 +664,7 @@ data Context =
   , valueIds     :: [ValueId]                    -- ^ Universe of value identifiers.
   , values       :: [Arith.Int]                  -- ^ Universe of values.
   , times        :: [Arith.Int]                  -- ^ Universe of times.
-  , cboundValues :: Map ValueId Arith.Int        -- ^ Bound values for state.
   }
-
-instance Prelude.Ord ValueId where
-   compare (ValueId a) (ValueId b) = compare a b
 
 instance Arbitrary Context where
   arbitrary =
@@ -681,7 +675,6 @@ instance Arbitrary Context where
       valueIds <- arbitrary
       values <- listOf arbitraryInteger
       times <- listOf arbitraryPositiveInteger
-      cboundValues <- fromList . nubBy ((==) `on` fst) <$> listOf ((,) <$> perturb arbitrary valueIds <*> perturb arbitraryInteger values)
       pure Context{..}
   shrink context@Context{..} =
       [context {amounts = amounts'} | amounts' <- shrink amounts]
@@ -690,7 +683,6 @@ instance Arbitrary Context where
       ++ [context {valueIds = valueIds'} | valueIds' <- shrink valueIds]
       ++ [context {values = values'} | values' <- shrink values]
       ++ [context {times = times'} | times' <- shrink times]
-      ++ [context {cboundValues = cboundValues'} | cboundValues' <- shrink cboundValues]
 
 instance Arbitrary Arith.Int where
   arbitrary = arbitraryInteger

### 2023-03-27 marlowe-spec-test/src/Marlowe/Spec/Core/Arbitrary.hs
PLT-3313: Non-negative start time for intervals

@@ -171,7 +171,7 @@ arbitraryPositiveInteger = Arith.Int_of_integer <$>
 arbitraryTimeInterval :: Gen (Arith.Int, Arith.Int)
 arbitraryTimeInterval =
   do
-    start <- arbitraryInteger
+    start <- arbitraryNonnegativeInteger
     duration <- arbitraryNonnegativeInteger
     pure (start, start + duration)
 
@@ -179,7 +179,7 @@ arbitraryTimeInterval =
 arbitraryTimeIntervalAround :: Arith.Int -> Gen (Arith.Int, Arith.Int)
 arbitraryTimeIntervalAround limit =
   do
-    start <- arbitraryInteger `suchThat` greater_eq limit
+    start <- arbitraryNonnegativeInteger `suchThat` greater_eq limit
     duration <- ((limit - start) +) <$> arbitraryNonnegativeInteger
     pure (start, start + duration)
 
@@ -187,7 +187,7 @@ arbitraryTimeIntervalAround limit =
 arbitraryTimeIntervalBefore :: Arith.Int -> Arith.Int -> Gen (Arith.Int, Arith.Int)
 arbitraryTimeIntervalBefore lower upper =
   do
-    start <- arbitraryInteger `suchThat` greater_eq lower
+    start <- arbitraryNonnegativeInteger `suchThat` greater_eq lower
     duration <- chooseinteger (0, upper - start - 1)
     pure (start, start + duration)
 
@@ -195,7 +195,7 @@ arbitraryTimeIntervalBefore lower upper =
 arbitraryTimeIntervalAfter :: Arith.Int -> Gen (Arith.Int, Arith.Int)
 arbitraryTimeIntervalAfter lower =
   do
-    start <- arbitraryInteger `suchThat` less_eq lower
+    start <- arbitraryNonnegativeInteger `suchThat` less_eq lower
     duration <- arbitraryNonnegativeInteger
     pure (start, start + duration)
 

### 2023-03-27 marlowe-spec-test/src/Marlowe/Spec/Core/Semantics.hs
PLT-3313: formatting

@@ -223,7 +223,7 @@ traceToSingleInputIsEquivalentTest interpret = reproducibleProperty "Single inpu
         c <- genContract interpret `suchThat` (/=Close) -- arbitraryValidInputs returns [] for the `Close` contract
         s <- genState interpret
         t <- liftGen $ arbitraryValidInputs s c
-        pure (c,s,t)) `suchThat` \(_,_,t) -> (t /= traceListToSingleInput t)
+        pure (c,s,t)) `suchThat` \(_,_,t) -> t /= traceListToSingleInput t
 
     let
         multipleInputs = PlayTrace (integer_of_int startTime) contract transactions

### 2023-03-27 response-to-audit-report.md
PLT-3382 Drafted template for response to audit report.

# DRAFT Response to Audit Report

TODO:

- [ ] Add prefatory text.
- [ ] Add responses and mitigation commits to Isabelle-related findings.
- [ ] Add responses and mitigation commits to Haskell-related findings.
- [ ] Fix display of equations or replace ones that cannot be fixed with ellipses ("...").
- [ ] Replace broken links and cross references, or use elipses ("...").
- [ ] Verify the section numbering is correct.
- [ ] Add table of contents.
- [ ] Proofread.

# 2 Findings

## 2.1 Main concerns

### 2.1.1 Negative deposits allow stealing funds *(Severity: High)*

> **File `marlowe-cardano-specification.md`, `Constraint 6`**
>
> The income from deposits is computed by adding up the deposit inputs, regardless of whether they are negative, while the semantics considers them as zero deposits. Combined with the absence of a balance check on the ending Marlowe state, this allows the ending balance to differ from the value paid to the Marlowe validator.
> 
> This disagreement can be exploited to steal money from a flawed Marlowe contract that allows a negative deposit. The issue is demonstrated in .

### 2.1.2 Contracts vulnerable to double satisfaction attacks *(Severity: High)*

> **File `marlowe-cardano-specification.md`, `Constraint 15`**
> 
> No datum is required for outputs fulfilling payments to addresses generated by the evaluation of a Marlowe contract. This implies that these outputs are vulnerable to double satisfaction in transactions involving other contracts that pay to the same wallets. An example is discussed in .
> 
> One way to strengthen the implementation is for the Marlowe validator to demand that outputs paid to addresses contain a datum that identifies the contract instance, like the `TxOutRef` of the validator UTxO being spent. Then cooperation with other contracts is possible without double satisfaction if the validators of the other contracts demand a different datum for their outputs.

### 2.1.3 Missing constructor in equality instance *(Severity: High)*

> **File `Semantics.hs`, Class instance `Eq ReduceWarning`, line *845***
> 
> The constructor `ReduceAssertionFailed` is not mentioned and compares `False` against itself. This might cause validators to fail checking the presence of this particular warning.

### 2.1.4 Inaccurate formulation of Money preservation *(Severity: High)*

> **File `specification-v3-rc1.pdf`, Section `3.1 Money preservation`, page *29***
> 
> As the property stands, it is permitted to make deposits in one currency and return payments in a different currency. As long as the sums of the amounts match, the equality is satisfied. Yet it is unlikely that the participants of the contract would agree that money has been preserved.
> 
> Money preservation is a property stated with an equality. The left hand side is the sum of the deposits done by a list of transactions. The right hand side of the equality is the sum of all the payments done in the same list of transactions. Each sum, in turn, is represented as a single integer which aggregates the amounts of the various payments and deposits, irrespective of what currencies correspond to these amounts.

### 2.1.5 Insufficient documentation of Money preservation *(Severity: Medium)*

> **File `specification-v3-rc1.pdf`, Section `3.1 Money preservation`, page *29***
> 
> Money preservation is formulated in terms of functions that are not discussed in the specification. It is necessary to explain the meaning of these functions in sufficient detail so readers can understand the property.

### 2.1.6 Missing description of Merkleization *(Severity: High)*

> **File `specification-v3-rc1.pdf`, `Merkleization`**
> 
> There is no property about merkleization, but merkleization is implemented in the Cardano integration.
> 
> Some relevant properties could be:
> 
> 1.  The merkleized contract produces the same payments as the analogous regular contract.
> 2.  If a merkleized case input is applied successfully, it implies that the contract hash in the input corresponds to the continuation of the contract.
> 3.  Merkleizing and unmerkleizing a contract gives back the original contract.

### 2.1.7 Positive balances are not checked for the output state *(Severity: High)*

> **File `marlowe-cardano-specification.md`, `Constraint 13`** 
> 
> *Positive balances* are only checked for the input, not for the output Marlowe state. If the semantics are flawed, a transaction can produce an unspendable output that does not satisfy this constraint.
> 
> If such a transaction is accepted, no further evaluation will be possible since all subsequent transactions will be rejected due to the very same Constraint 13. This is an hypothetical attack vector, where a malicious actor could send a transaction to block a contract.

### 2.1.8 Non-validated Marlowe states *(Severity: High)*

> **File `marlowe-cardano-specification.md`, `Missing constraint`** 
> 
> The validator is not specified to check that the Marlowe states in the input and output datums are valid. This condition is necessary for the lemmas about the Marlowe semantics to be applicable. The Marlowe state could become invalid if there is a flaw in the implementation of the semantics.
> 
> It also could be possible for the Marlowe state to be invalid if someone pays an output to the Marlowe validator with an invalid Marlowe state. Though this problem could be addressed with off-chain checks that prevent sending transactions that spend outputs with invalid Marlowe states. If off-chain checks are used, a note in the specification about how this is handled would be helpful.
> 
> An example showing betrayed user expectations is discussed in .
> 
> For a valid Marlowe state, the association lists for bound values, accounts, and choices have keys sorted and without duplicates.

### 2.1.9 Total balance of ending state uncomputed *(Severity: High)*

> **File `marlowe-cardano-specification.md`, `Constraint 6`** 
> 
> The constraint says
> 
> > The beginning balance plus the deposits equals the ending balance plus
> > the payments.
> 
> However, the Marlowe validator never computes the total balance of the accounts in the ending Marlowe state. Instead, the ending balance is assumed to be whatever value is paid by the transaction to the Marlowe validator. The natural language should describe precisely what is being checked.

### 2.1.10 Unchecked ending balance *(Severity: High)*

> **File `marlowe-cardano-specification.md`, `Constraint 5`** 
> 
> The balance of the starting Marlowe state is checked to match the value in the input. However, the validator does not check that the ending balance matches the value in the output paid to the Marlowe validator. Similarly to Issue [\[main:positive-balances-unchecked\]](#main:positive-balances-unchecked){reference-type="ref" reference="main:positive-balances-unchecked"}, if there are flaws in the semantics that cause the ending balance to differ from the actual value paid to the validator, this constraint would prevent any transaction from spending the output.
> 
> The specification should at least discuss why the check is absent together with the other similar checks that are not implemented (checking that ending accounts have positive balances, checking that the ending Marlowe state is valid).

### 2.1.11 Partial functions used outside their domain *(Severity: Medium)*

> **File `MoneyPreservation.thy`, various functions ``**
> 
> `moneyInRefundOneResult`, `moneyInApplyResult`, `moneyInApplyAllResult`, `moneyInTransactionOutput`, and `moneyInPlayTraceResult` have strange meanings when the result is an error. Arguably, on error there is no money to retrieve, so the return type should be `(Token \times int) option` instead.
> 
> Some lemmas rely on this behavior to have equalities hold even in cases of errors, but the cost is that the meaning is so surprising that the reader may be confused by it. It would be more reliable to have explicit and weaker lemmas that assert equalities only when there are no errors.

### 2.1.12 Different insertion functions used in Isabelle and Haskell code *(Severity: High)*

> **File `Semantics.hs`, Several functions ``** 
> 
> Where `MList.insert` is used in the Isabelle semantics, [AssocMap.insert](https://github.com/input-output-hk/plutus/blob/v1.0.0/plutus-tx/src/PlutusTx/AssocMap.hs#L147-L148) is used in the Cardano implementation. However, the functions are not equivalent as demonstrated by the following examples:
> 
> ``` text
> AssocMap.insert 'a' 1 [('b', 0)] == [('b', 0), ('a', 1)]
> -- whereas
> MList.insert 'a' 1 [('b', 0)] == [('a', 1), ('b', 0)]
> ```
> 
> ``` text
> AssocMap.insert 'b' 1 [('a', 0), (‘b’, 0), (‘b’, 0)] == [('a', 0), ('b', 1), (‘b’, 1)]
> -- whereas
> MList.insert 'b' 1 [(‘a’, 0), ('b', 0), ('b', 0)] == [('a', 0), ('b', 1), ('b', 0)]
> ```
> 
> This renders the Isabelle lemmas inapplicable for the Cardano integration. The lemmas need to demand some properties of an `insert` function without fully spelling it out, or the Cardano integration needs to use `MList.insert` instead of `AssocMap.insert`.
> 
> Similarly, functions `AssocMap.delete` and `MList.delete` differ in behavior when the input map is not sorted:
> 
> ``` text
> AssocMap.delete 'a' [(‘b’, 0), (‘a’, 0)] == [('b', 0)]
> -- whereas
> MList.delete 'a' [(‘b’, 0), ('a', 0)] == [('b', 0), ('a', 0)]
> ```
> 
> Functions `AssocMap.lookup` and `MList.lookup` also differ in behavior when the input map is not sorted:
> 
> ``` text
> AssocMap.lookup 'a' [(‘b’, 0), (‘a’, 0)] == Just 0
> -- whereas
> MList.lookup 'a' [(‘b’, 0), ('a', 0)] == Nothing
> ```
> 
> The following usage places were found:
> -   Line 395, `evalValue` depends on `moneyInAccount` which depends on `AssocMap.lookup`.
> -   Line 413, `evalValue` depends on `AssocMap.lookup`.
> -   Line 428, `evalObservation` depends on `AssocMap.member`.
> -   Line 456, function `updateMoneyInAccount` relies on `AssocMap.delete` and `AssocMap.insert`.
> -   Line 482, function `reduceContractStep` relies on `AssocMap.insert`.
> -   Line 567, function `applyAction` relies on `AssocMap.insert`.

### 2.1.13 Missing specification tests *(Severity: Medium)*

> **File `Spec/Marlowe/Semantics/Compute.hs`, ``** 
> 
> There are no tests for the properties in Section 3 of `specification-v3-rc1.pdf`. Besides checking that there are no translation mistakes, these properties would also help contrasting the assumptions in the Isabelle and the Haskell sides, like the meaning of validity of an association list, which is focused in the previous issue.

## 2.2 Marlowe specification

### 2.2.1 Lack of explanation regarding changing choices *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.1.4 Choices`, page *10***
> 
> Choices can only be changed when evaluating `When` statements. This is something only evident after looking at the implementation of `computeTransaction`. It needs to be discussed when first introducing choices and the `When` contract.

### 2.2.2 Undefined reference *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.1.7 Contracts`, page *13***
> 
> There is an undefined reference.

### 2.2.3 Lack of explanation for necessity of `Environment` type *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.1.8 State and Environment`, page *14***
> 
> An `Environment` type is introduced, but it is unclear why it is needed as it is defined as a synonym for time intervals.

### 2.2.4 Unclear meaning of execution environment *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.1.8 State and Environment`, page *14***
> 
> The meaning of the execution environment of the transaction is unclear. This is due to the concept of *transaction* being assumed by the specification and never formally introduced.
> 
> The specification reads
> 
> > The execution environment of a Marlowe contract simply consists of the (inclusive) time interval within which the transaction is occurring.
> 
> One has to infer that evaluating a Marlowe contract is undefined if it does not happen within a transaction, as otherwise the description of the execution environment would not make sense. It would be necessary to establish more explicitly the relationship between the contract evaluation and the notion of transaction.

### 2.2.5 Unexplained interval data types *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.1.8 State and Environment`, page *14***
> 
> The meaning of the data types `IntervalError` and `IntervalResult` needs to be explained.

### 2.2.6 Incomplete explanation for `TransactionOutput` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.1 Compute Transaction`, page *15***
> 
> The meaning of the data type `TransactionOutput` needs to be explained. More generally, the meaning of the return types of most functions has to be explained. Currently, the meaning can only be inferred from looking at how the types are used, which makes it harder to identify if they are used as intended.
> 
> The purpose of these types needs to be made explicit so it can be checked if the code is doing what is intended.

### 2.2.7 Code snippets switch languages *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.1 Compute Transaction`, page *15***
> 
> The specification changes from using Isabelle to using Haskell henceforth. Making the reader aware of the criteria for the language change would help maintaining the document.

### 2.2.8 Repeated definition of `IntervalResult` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Sections `2.1.8 State and Environment, 2.2.2 Fix Interval`, pages *14, 16***
> 
> The `IntervalResult` type is defined twice in the specification. One should be removed.

### 2.2.9 Poorly named variable `newAccount` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.6 Reduce Contract Step`, page *19***
> 
> In the implementation of the function `reduceContractStep`, the variable `newAccount` should be named `newAccounts`.

### 2.2.10 Poorly named variable `acc` in specification *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.8 Apply Cases`, page *22***
> 
> On the last equation of `applyCases`, `acc` should be named `input`.

### 2.2.11 Inaccurate specification of `giveMoney` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.9 Utilities`, page *22***
> 
> It says
> 
> > The *giveMoney* function transfers funds internally between accounts.
> 
> which is not accurate. It should say instead
> 
> > The *giveMoney* function deposits funds to an internal account.
> 
> This function is confusing in that it takes the account identifier of the paying account which is not used for anything other than filling a field in the returned value.

### 2.2.12 Redundant evaluation in `addMoneyToAccount` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.9 Utilities`, page *22***
> 
> `addMoneyToAccount` is redundantly evaluating `money <= 0` when invoking `updateMoneyInAccount`. The else branch could be replaced instead with `insert (accId, token) money accountsV`.

### 2.2.13 Redundant statement regarding addition *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.10 Evaluate Value`, page *24***
> 
> It says that addition is associative and commutative. This is true but it is already implied by the equation preceding the statement. Maybe change to
> 
> > Note that addition is associative and commutative.
> 
> or remove the redundant statement.

### 2.2.14 Missing implementation for negation case of `evalValue` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.10 Evaluate Value`, page *24***
> 
> Negation for `evalValue` does not show the implementation, just one lemma about `NegValue`, which is inconsistent with how other operations are presented.

### 2.2.15 Missing parentheses in `div` specification *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.10 Evaluate Value`, page *25***
> 
> On page 25 formula $$c \neq 0 \Rightarrow c \mathbin{∗} a \mathbin{\mathrm{div}} (c \mathbin{∗} b) = a \mathbin{\mathrm{div}} b$$ needs additional parentheses around the term $c \mathbin{∗} a$, otherwise it can be parsed as $$c \neq 0 \Rightarrow c \mathbin{∗} (a \mathbin{\mathrm{div}} (c \mathbin{∗} b)) = a \mathbin{\mathrm{div}} b$$ which does not hold (Counter-example: $c=2, a=3, b=2$). The lemma `divMultiply` in the file `Semantics.thy` does use extra parentheses around $c \mathbin{∗} a$.

### 2.2.16 Unclear division explanation *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.10 Evaluate Value`, page *25***
> 
> It says
> 
> > Division is a special case because we only evaluate to natural numbers.
> 
> The meaning of this statement needs to be further explained, since the arguments of `DivValue` could evaluate to negative numbers.

### 2.2.17 Discrepancy with `evalValue` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.10 Evaluate Value`, pages *23--26***
> 
> The order of some cases for `evalValue` is different in the specification text and in the actual Isabelle code, and several cases (for example, `NegValue`) are missing from the specification entirely.
> 
> Moreover, the definition of `evalValue` is juxtaposed with some lemmas about its behavior (for example, `AddValue` being associative and commutative), making it harder to match the specification text with the Isabelle code.

### 2.2.18 Missing `evalValue` lemmas in specification *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.10 Evaluate Value`, pages *23--26***
> 
> Not all lemmas about `evalValue` are listed in the specification. The absent lemmas include `evalDoubleNegValue`, `evalMulValue`, `evalSubValue`, and all division lemmas.

### 2.2.19 Typo in **Use Value** case of `evalValue` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.10 Evaluate Value`, page *26***
> 
> The **Use Value** case mentions `TimeIntervalEnd` instead of `UseValue`.

### 2.2.20 Unexplained parameters of `playTrace` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `3 Marlowe Guarantees`, page *28***
> 
> The parameters of the function `playTrace` need to be explained.

### 2.2.21 Type parameter discrepancy in `playTrace` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `3 Marlowe Guarantees`, page *28***
> 
> The first parameter of `playTrace` in the specification is `int`, while it is `POSIXTime` in the code. Even though the latter is an alias for the former, it is beneficial to use the `POSIXTime` name both for consistency and readability.

### 2.2.22 Money preservation on failing transactions not specified *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `3.1 Money preservation`, page *29***
> 
> Money preservation is expressed with an equality. This equality, however, only ensures money preservation for those lists of transactions that produce no error. In other words, there is no guarantee that money will be preserved for those lists of transactions that fail.
> 
> This is not a concern in practice because the lists of transactions that fail to evaluate are not accepted in the blockchain. However, this should be made explicit in the explanation of the property.

### 2.2.23 Complicated definition of `allAccountsPositive` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `3.3 Possitive Accounts`, page *30***
> 
> The definition of\ `allAccountsPositive` is complicated and can be refactored as `all ((_, money) -> money > 0)`.

### 2.2.24 Discrepancy with Isabelle code for `allAccountsPositive` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Sections `3.3 Positive Accounts`, page *30***
> 
> The `allAccountsPositive` function is defined differently in the specification and in the Isabelle code, although both definitions show the same behavior. These definitions need to be consolidated.

### 2.2.25 Misleading or incorrect formula for contract not holding funds *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `3.6.3 Contract Does Not Hold Funds After it Closes`, page *32***
> 
> The statement in natural language looks unconnected from the proposed formula. Otherwise, it is unclear how not holding funds forever is a consequence of producing no warnings.

### 2.2.26 Different format for lemma statement *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Sections `3.6.2 All Contracts Have a Maximum Time`, page *32***
> 
> The lemma is stated using the proof derivation tree format as opposed to the rest of the specification and the Isabelle code.

### 2.2.27 Function `isClosedAndEmpty` is unexplained *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `3.6.2 All Contracts Have a Maximum Time`, page *32***
> 
> The function `isClosedAndEmpty` needs to be explained.

### 2.2.28 Top-down definitions *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2`** 
> 
> In Section 2, the order of definitions is reversed, and the reader is thus faced with functions which call other functions that have not been introduced yet, despite the claim in Section 1.3 that the definitions will be presented bottom-up.

### 2.2.29 No mention of Isabelle lemmas in specification *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Multiple sections ``** 
> 
> Generally, readability can be improved by mentioning the Isabelle lemma names alongside their statements. This way, it would be much easier to search for the actual Isabelle code and proofs matching the informal specification text, and compare the two.

## 2.3 Lemmas and proofs

### 2.3.1 Unnecessarily large proofs *(Severity: Medium)*

> **Several Isabelle files, several lemmas**
> 
> Some Isabelle proofs are written with long apply-scripts, where Isar would document the proof better. Proofs could also be split using more auxiliary lemmas.
> 
> As the proofs stand, it is hard to figure out why a proof step fails, after changes elsewhere required a proof to be updated. Since the newly-failing proof step was designed with specific goals in mind, and changes in the code may lead to it facing a different set of goals, the maintainer might need to reconstruct the whole structure of the proof from an older version to infer state that Isabelle produces at each step.
> 
> What Isar brings is making the intention of the author explicit at every step of the proof. This helps the maintainer of the proofs and fixes the concerns mentioned above.
> 
>  will likely have to update the proofs. We conjecture that it will happen at least every time they target a new platform. In the case of Cardano, they need to extend the semantics to explain Merkleization. Another action that would make long proofs easier to understand is to split them using more auxiliary lemmas, thus feeding the information to the reader in smaller chunks.
> 
> Some examples of large proofs:
> -   in `MoneyPreservation.thy`, lemmas `reduceContractStep_preserves_money` and\ `reductionLoop_preserves_money`
> -   in `SingleInputTransactions.thy`, lemmas `applyAllInputsPrefix_aux`,\ `computeTransactionIterative`, and `computeTransactionStepEquivalence_error`

### 2.3.2 Long lines in lemmas *(Severity: Low)*

> **Several Isabelle files, several lemmas**
> 
> Lines are sometimes long which makes it difficult to understand the lemmas. Lemmas need to be formulated expressing one hypothesis per line and the conclusion on a separate line. Complex hypotheses need to be indented using several lines to expose their structure.
> 
> Besides the effort of scrolling the text horizontally, the hypotheses are hard to separate visually, and so is the conclusion. Furthermore, when a hypothesis is a nested implication it is difficult to see where it ends without further indentation.
> 
> Some examples of lemmas with long lines or non-trivial hypothesis follow.
> -   in `CloseSafe.thy`, lemmas `closeIsSafe_reduceContractUntilQuiescent`, and `closeIsSafe_reductionLoop`
> -   in `MoneyPreservation.thy`, lemmas `reductionLoop_preserves_money_Payment_not_ReduceNoWarning`, `reductionLoop_preserves_money_Payment` and `reduceContractStep_preserves_money_acc_to_party`
> -   in `SingleInputTransactions.thy`, lemma `applyAllLoop_longer_doesnt_grow`
> -   in `TimeRange.thy`, lemmas `reduceStep_ifCaseLtCt` and `reduceLoop_ifCaseLtCt`
> -   in `ValidState.thy`, lemma `reductionLoop_preserves_valid_state_aux`

### 2.3.3 Confusing auxiliary lemmas *(Severity: Low)*

> **Several Isabelle files, several lemmas**
> 
> Some Isabelle proofs resort to declaring auxiliary lemmas with names suffixed with *\_aux*. Sometimes these lemmas are not expressed succinctly, and look more like a punctual copy of the state of some particular proof that is later developed. For the sake of maintaining the proofs, it would be necessary to structure them in a way that presents the information piecewise to the reader. More generally, even auxiliary lemmas should have a well-defined meaning.
> 
> We found this problem at least in the following:
> -   in `QuiescentResult.thy`, lemmas `reduceContractStepPayIsQuiescent`, `reductionLoopIsQuiescent_aux`, and `applyAllInputsLoopIsQuiescent_loop`
> -   in `PositiveAccounts.thy`, lemma `positiveMoneyInAccountOrNoAccountImpliesAllPositive_aux2`
> -   in `SingleInputTransactions.thy`, lemma `applyAllInputsPrefix_aux`

### 2.3.4 Undescriptive variable names *(Severity: Low)*

> **Several Isabelle files, several lemmas**
> 
> Many Isabelle proof statements and proofs use uninformative variable names. The most common example occurs with variables named $\mathit{x11}, \mathit{x12}$, etc. These inhibit the reader from easily understanding the lemma statements, and often require looking back at constructors to understand what these variables represent.
> 
> Some examples of lemmas with these uninformative variable names follow:
> -   in `QuiescentResult.thy`, lemma `reductionLoopIsQuiescent_aux`
> -   in `SingleInputTransactions.thy`, lemmas `beforeApplyAllLoopIsUseless` and\ `applyAllInputsPrefix_aux`
> -   in `ValidState.thy`, lemma `reductionLoop_preserves_valid_state_aux`
> -   in `TimeRange.thy`, lemmas `resultOfReduceIsCompatibleToo`, `resultOfReductionLoopIsCompatibleToo`, `resultOfReduceUntilQuiescentIsCompatibleToo`, `reduceLoop_ifCaseLtCt`, and\ `reduceContractUntilQuiescent_ifCaseLtCt`

### 2.3.5 Involved proof of `insert_valid` *(Severity: Low)*

> **File `MList.thy`, theorem `insert_valid`, line *66***
> 
> The proof of `insert_valid` sprouts three other lemmas of difficult characterization: `insert_valid_aux`, `insert_valid_aux2`, and `insert_valid_aux3`. These lemmas make assumptions with implications that get in the way of understanding them in isolation.
> 
> An alternative to make the proof pieces more reusable is to use instead the following set of lemmas, which also offers insight on how function `insert` interacts with predicates `sorted` and `distinct`:
> 
> ``` {.text escapeinside="||" mathescape="true"}
>   lemma insert_set:
>     "set (map fst (insert a b xs)) = set (map fst xs) |$\cup$| { a }"
> 
>   lemma insert_sorted:
>     "List.sorted (map fst c) |$\Longrightarrow$| List.sorted (map fst (insert a b c))"
> 
>   lemma insert_distinct :
>     "List.distinct (map fst c)
>      |$\Longrightarrow$|
>      List.sorted (map fst c)
>      |$\Longrightarrow$|
>      List.distinct (map fst (MList.insert a b c))"
> ```
> 
> which then can be combined in the proof of `insert_valid` as follows:
> 
> ``` {.text escapeinside="||" mathescape="true"}
>   theorem insert_valid2 : "valid_map c |$\Longrightarrow$| valid_map (MList.insert a b c)"
>     using insert_sorted[of c a b] insert_distinct[of c a b] by fastforce
> ```
> 
> The proofs of the lemmas can be found in .

### 2.3.6 Repeated verbose expression *(Severity: Low)*

> **File `MoneyPreservation.thy`, lemma `removeMoneyFromAccount_preservation`, line *202***
> 
> The expression
> 
> ``` text
>   giveMoney
>     accId
>     (Party p)
>     tok
>     paidMoney
>     (updateMoneyInAccount accId tok (balance - paidMoney) accs)
> ```
> 
> is large and used in other lemmas as well. It would need to be moved to a separate function to save the effort of reading it repeteadly.

### 2.3.7 Inconsistent variable name `valTrans` *(Severity: Low)*

> **File `MoneyPreservation.thy`, lemma `transferMoneyBetweenAccounts_preserves_aux`, line *257***
> 
> The lemma uses a variable `valTrans` where other proofs use the name `paidMoney`. To convey the meaning of the variable faster, the same name should be used consistently in all places.

### 2.3.8 Unused binding `interAccs` *(Severity: Low)*

> **File `MoneyPreservation.thy`, lemma `transferMoneyBetweenAccounts_preserves_aux`, line *263***
> 
> The binding `interAccs` was probably intended to be used on this line. It should either be used or removed from the premise.

### 2.3.9 Undescriptive variable name `acc` *(Severity: Low)*

> **File `MoneyPreservation.thy`, lemma `transferMoneyBetweenAccounts_preserves`, line *295***
> 
> This lemma has a variable `acc` that is used together with `tok2`. It would be more descriptive to call it `accId2`.

### 2.3.10 Misleading indentation *(Severity: Low)*

> **File `MoneyPreservation.thy`, lemmas `reductionLoop_preserves_money_NoPayment_not_ReduceNoWarning, reductionLoop_preserves_money_NoPayment`, lines *430, 439***
> 
> The indentation is misleading: the premises on these lines are indented as if they are a part of the previous functional premise.

### 2.3.11 Missing theorem regarding `playTrace` *(Severity: Low)*

> **File `PositiveAccounts.thy`, `playTrace preserves valid and positive state`** 
> 
> There is no theorem that `playTrace` keeps the state valid and positive when given a state which is valid and positive. This trivially follows from `playTraceAux_preserves_validAndPositive_state` but no such theorem is present.

### 2.3.12 Unconcise goal in `reduceContractStepPayIsQuiescent` *(Severity: Low)*

> **File `QuiescentResult.thy`, lemma `reduceContractStepPayIsQuiescent`, line *8***
> 
> This lemma does not express its goal concisely, as it makes no mention of `reduceContractStep` in the formulation. Changing the first assumption to $\texttt{reduceContractStep}\ \mathit{env}\ \mathit{sta}\ (\texttt{Pay}\ \mathit{x21}\ \mathit{x22}\ \mathit{tok}\ \mathit{x23}\ \mathit{x24})$ makes more explicit in which contexts this lemma can be useful. Modifying this assumption requires an additional `apply simp` to be added to the proof (before line 30) for the lemma to go through. Further, an additional `apply simp` will need to be added in lemmas `reduceContractStepIsQuiescent` (before line 44) and `timedOutReduce_only_quiescent_in_close` (`Timeout.thy`, before line 128) as well.

### 2.3.13 Misleading lemma names *(Severity: Low)*

> **File `PositiveAccounts.thy`, lemma `reduceOne_gtZero`, line *80***
> 
> This lemma should be renamed as `refundOne_gtZero`.
> 
> **File `QuiescentResult.thy`, lemma `reduceOneIsSomeIfNotEmptyAndPositive`, line *32***
> 
> This lemma should be renamed as `refundOneIsSomeIfNotEmptyAndPositive`.
> 
> **File `TransactionBound.thy`, lemma `computeTransaction_decreases_maxTransaction_aux`, line *240***
> 
> This lemma should be renamed as `applyAllInputs_decreases_maxTransactions` or `applyAllInputs_reduced_decreases_maxTransactions`.

### 2.3.14 Misleading variable name `reduced` *(Severity: Low)*

> **File `QuiescentResult.thy`, lemmas `reductionLoop_reduce_monotonic, reduceContractUntilQuiescent_ifDifferentReduced`, lines *138, 153***
> 
> The boolean variable name `reduce` would be better named `reduced` as it is signifying that the contract has been reduced.

### 2.3.15 Undescriptive name `beforeApplyAllLoopIsUseless` *(Severity: Low)*

> **File `SingleInputTransactions.thy`, lemma `beforeApplyAllLoopIsUseless`, line *270***
> 
> This lemma seems to say that `reduceContractUntilQuiescent` has no effect when composed with `applyAllLoop`, because `applyAllLoop` evaluates `reduceContractUntilQuiescent`, and `reduceContractUntilQuiescent` is idempotent.
> 
> A more descriptive name for this lemma could be `reduceContractUntilQuiescent_hasNoEffect_before_applyAllLoop`

### 2.3.16 Unused and undocumented lemmas *(Severity: Low)*

> **Several Isabelle files, several lemmas**
> 
> Some lemmas are never used, and they would need comments motivating their presence:
> 1.  In file `MoneyPreservation.thy`, line 257, lemma `transferMoneyBetweenAccounts_preserves_aux`.
> 2.  In file `QuiescentResult.thy`
>     1.  Line 5, lemma `reduceOne_onlyIfNonEmptyState`
>     2.  Line 153, lemma `reduceContractUntilQuiescent_ifDifferentReduced`
> 3.  In file `PositiveAccounts.thy`, line 66, lemma `positiveMoneyInAccountOrNoAccount_sublist_gtZero`. Furthermore, it is identical to `positiveMoneyInAccountOrNoAccount_gtZero_preservation`, but with an additional assumption `money > 0`.
> 4.  In file `ValidState.thy`
>     1.  Line 9, lemma `valid_state_valid_choices`
>     2.  Line 13, lemma `valid_state_valid_valueBounds`
> 5.  In file `SingleInputTransactions.thy`, line 1214, lemma `traceListToSingleInput_isSingleInput`. It is mentioned in a commented out line in `StaticAnalysis.thy`. Furthermore, the lemma can be expressed more concisely as $$\llparenthesis \mathit{interval} = \mathit{inte}, \mathit{inputs} = \mathit{inp\_h}\ \#\ \mathit{inp\_t} \rrparenthesis\ \#\ t = \mathit{traceListToSingleInput}\ \mathit{t2} % \mathrel{% \mbox{\fontfamily{cmr}\fontencoding{OT1}\selectfont=}}% \joinrel\Rightarrow\mathit{inp\_t} = []$$

### 2.3.17 Redundant `reduceContractStep` lemmas *(Severity: Low)*

> **File `MoneyPreservation.thy`, lemma `reduceContractStep_preserves_money_acc_to_acc_aux`, line *310***
> 
> This lemma is weaker than `transferMoneyBetweenAccounts_preserves`. If we replace its usage at line 351 with `transferMoneyBetweenAccounts_preserves`, the proof goes through.

### 2.3.18 Redundant `transferMoneyBetweenAccounts_preserves` *(Severity: Low)*

> **File `MoneyPreservation.thy`, lemma `reduceContractStep_preserves_money_acc_to_acc`, line *332***
> 
> This lemma is weaker than `transferMoneyBetweenAccounts_preserves`. We can replace its usage site in line 376
> 
> ``` text
>   using
>     reduceContractStep_preserves_money_acc_to_acc
>     validAndPositive_state.simps
>    by blast
> ```
> 
> with
> 
> ``` text
>   using transferMoneyBetweenAccounts_preserves validAndPositive_state.simps by auto
> ```

### 2.3.19 Duplicated lemmas *(Severity: Low)*

> **File `PositiveAccounts.thy`, theorems `computeTransaction_gtZero, accountsArePositive`, lines *257, 369***
> 
> These theorems are identical (modulo variable names), and one of them should be removed.
> 
> **File `PositiveAccounts.thy, ValidState.thy`, lemma `valid_state_valid_accounts`, lines *381, 5***
> 
> This lemma is defined twice, once in each of these files. One of them should be removed.

### 2.3.20 Redundant `computeTransaction` lemmas *(Severity: Low)*

> **File `ValidState.thy`, lemmas `computeTransaction_preserves_valid_state_aux, computeTransaction_preserve_valid_state`, lines *160, 176***
> 
> If `computeTransaction_preserves_valid_state_aux` is rewritten to have the same formulation as `computeTransaction_preserves_valid_state`, then the lemma (with the exact same proof) is still accepted, and these lemmas become duplicates of each other. Thus, no auxiliary lemma is needed.

### 2.3.21 Complicated formulation of `updateMoneyInAccount_money2_aux` *(Severity: Low)*

> **File `MoneyPreservation.thy`, lemma `updateMoneyInAccount_money2_aux`, line *159***
> 
> `updateMoneyInAccount_money2_aux` could be expressed simpler by removing the hypothesis `moneyToPay >= 0`, leaving
> 
> ``` {.text escapeinside="||" mathescape="true"}
>   valid_map (((thisAccId, tok), money) # tail) |$\Longrightarrow$|
>   allAccountsPositive (((thisAccId, tok), money) # tail) |$\Longrightarrow$|
>   moneyInAccount thisAccId tok (((thisAccId, tok), money) # tail) > 0"
> ```
> 
> The proof of `updateMoneyInAccount_money2` can then in turn be trivially adjusted so it still works, by changing the step cases
> 
> ``` {.text escapeinside="||" mathescape="true"}
>   cases "moneyInAccount accId tok ((thisAccIdTok, money) # tail) + moneyToPay |$\leq$| 0"
> ```
> 
> to
> 
> ``` {.text escapeinside="||" mathescape="true"}
>   cases "moneyInAccount accId tok ((thisAccIdTok, money) # tail) |$\leq$| 0"
> ```

### 2.3.22 Complicated proofs that can be simplified *(Severity: Low)*

> **File `MoneyPreservation.thy`, lemma `moneyInInput_is_positive`, line *53***
> 
> The proof could be more general with `apply (cases x; simp)` instead of using `metis`.
> 
> **File `MoneyPreservation.thy`, lemma `reductionLoop_preserves_money_NoPayment_not_ReduceNoWarning`, line *434***
> 
> This lemma can be proved directly with `metis reductionLoop_preserves_money_NoPayment`, and reversing the order in which the lemmas are defined.
> 
> **File `TimeRange.thy`, lemma `inIntervalIdempotentToIntersectInterval`, line *5***
> 
> The lemma can use a shorter proof: `apply (cases min2;cases max2;auto) done`.
> 
> **File `TimeRange.thy`, lemma `inIntervalIdempotency1, inIntervalIdempotency2`, lines *20, 36***
> 
> These lemmas use the `smt` tactic and `metis` where a simpler Isar proof would work, for example:
> 
> ``` text
>   lemma inIntervalIdempotency1 :
>     assumes "inInterval (x, y) (intersectInterval b c)"
>     shows "inInterval (x, y) b"
>   proof (cases b)
>     case [simp]:(Pair b1 b2)
>     thus ?thesis proof (cases c)
>       case (Pair c1 c2)
>       thus ?thesis using assms by (cases c1; cases c2; cases b1;cases b2; simp)
>     qed
>   qed
> ```
> 
> **File `SemanticsGuarantees.thy`, `Various lemmas/instantiations`**
> 
> Multiple lemmas and `linorder` instantiations in this file repeat auxiliary facts within the proof that are not necessary. For example, in the `linorder` instantiation for `Party`, lines 51--53 state
> 
> ``` {.text escapeinside="||" mathescape="true"}
>   have "(x < y) = (x |$\le$| y |$\land$| |$\neg$| y |$\le$| (x :: Token))"
>     by (meson less_Tok.simps less_Token_def less_eq_Token_def linearToken)
>   thus "(x < y) = (x |$\le$| y |$\land$| |$\neg$| y |$\le$| x)" by simp
> ```
> 
> This can be rewritten to avoid repeating the fact as
> 
> ``` {.text escapeinside="||" mathescape="true"}
>   show "(x < y) = (x |$\le$| y |$\land$| |$\neg$| y |$\le$| (x :: Token))"
>     by (meson less_Tok.simps less_Token_def less_eq_Token_def linearToken)
> ```
> 
> This pattern appears many times in this file. For example, in the `Party` instantation alone, it is present on lines 51 -- 53, 56 -- 57, 77 -- 80, and 83 -- 84.

### 2.3.23 Inconsistent style when applying constructor *(Severity: Low)*

> **File `SingleInputTransactions.thy`, lemmas `beforeApplyAllLoopIsUseless, fixIntervalOnlySummary`, lines *275, 398***
> 
> The lines mentioned in these lemmas display the resulting constructor before the function application, which differs from the general style in the rest of the codebase.

### 2.3.24 Unsimplified boolean formulas *(Severity: Low)*

> **File `SingleInputTransactions.thy`, lemma `computeTransactionIterative_aux2`, lines *708, 710***
> 
> In multiple places, this lemma formulation includes top-level negation in front of nontrivial conjunctions and disjunctions. These negations should be distributed. Otherwise, the reader is taxed with the chore to mentally distribute the negation to understand the lemma.

### 2.3.25 Typo with "independet" in multiple lemmas *(Severity: Low)*

> **File `SingleInputTransactions.thy`, lemmas `applyAllLoop_independet_of_acc_error1, applyAllLoop_independet_of_acc_error2`, lines *977, 987***
> 
> In both of these lemmas, there is a typo with the word "independet".

### 2.3.26 Poorly named `acc` lemmas *(Severity: Low)*

> **File `SingleInputTransactions.thy`, lemmas `applyAllLoop_independet_of_acc_error1, applyAllLoop_independet_of_acc_error2`, lines *977, 987***
> 
> It is unclear what `acc` refers to in these lemma names, as the lemmas are about the independence of warnings and payments, not accounts.

### 2.3.27 Verbose lemma statement `playTraceAuxIterative_base_case` *(Severity: Low)*

> **File `SingleInputTransactions.thy`, lemma `playTraceAuxIterative_base_case`, line *1063***
> 
> The statement of this lemma is very verbose. A more natural (and slightly stronger) formulation could be $$\begin{aligned} &\mathit{playTraceAux}\ \mathit{txOut}\ [\ \llparenthesis \mathit{interval} = \mathit{inte}, \mathit{inputs} = [h] \rrparenthesis, \llparenthesis \mathit{interval} = \mathit{inte}, \mathit{inputs} = t \rrparenthesis \ ] \\ =\ &\mathit{playTraceAux}\ \mathit{txOut}\ [\ \llparenthesis \mathit{interval} = \mathit{inte}, \mathit{inputs} = h\ \#\ t \rrparenthesis \ ]\end{aligned}$$

### 2.3.28 `playTrace_only_accepts_maxTransactionsInitialState` not written as `theorem` *(Severity: Low)*

> **File `TransactionBound.thy`, lemma `playTrace_only_accepts_maxTransactionsInitialState`, line *316***
> 
> This lemma seems like the main result of this file. Assuming it is an important result, we recommend writing it as a `theorem` rather than a `lemma`.

### 2.3.29 Inconsistent style with assumptions *(Severity: Low)*

> **File `Timeout.thy`, lemmas `timedOutReduceContractUntilQuiescent_closes_contract, timedOutReduceContractStep_empties_accounts`, lines *201/204, 211/214***
> 
> These lemmas use the hypothesis $\mathit{minTime}\ \mathit{sta} \leq \mathit{iniTime}$ and build a state $\mathit{sta}\ \llparenthesis \mathit{minTime} := \mathit{iniTime} \rrparenthesis)$ while other lemmas simply say $\mathit{minTime}\ \mathit{sta} = \mathit{iniTime}$. Readability would be improved by presenting these lemmas in the same style as the others, or documenting the need for these distinct presentations via code comments.

### 2.3.30 Function `validTimeInterval` unnecessarily unfolded in lemma *(Severity: Low)*

> **File `TimeRange.thy`, lemma `reduceStep_ifCaseLtCt_aux`, line *234***
> 
> For consistency, $a \leq b$ should be replaced by $\texttt{validTimeInterval}$.

### 2.3.31 Overly specific auxiliary lemma *(Severity: Low)*

> **File `ValidState.thy`, lemma `reductionLoop_preserves_valid_state_aux`, line *73***
> 
> This lemma on its own is very specific, and is only used in `reductionLoop_preserves_valid_state`. If possible, we recommend this lemma to be generalized or broken down into smaller lemmas, in order to present the arguments to the reader in smaller pieces.

### 2.3.32 `playTrace_preserves_valid_state` not written as `theorem` *(Severity: Low)*

> **File `ValidState.thy`, lemma `playTrace_preserves_valid_state`, line *194***
> 
> This lemma seems like the main result of this file. Assuming it is an important result, we recommend writing it as a `theorem` instead.

### 2.3.33 Unnecessary assumptions *(Severity: Low)*

> **File `PositiveAccounts.thy`, lemmas `addMoneyToAccountPositive_match, addMoneyToAccountPositive_noMatch`, lines *12, 23***
> 
> The assumptions $$\forall x\ \mathit{tok}.\ \mathit{positiveMoneyInAccountOrNoAccount}\ x\ \mathit{tok}\ \mathit{accs}$$ in `addMoneyToAccountPositive_match` and $$\mathit{money > 0}$$ in `addMoneyToAccountPositive_noMatch` are unnecessary.
> 
> **File `PositiveAccounts.thy`, lemma `reduceContractStep_gtZero_Refund`, line *93***
> 
> The lemma has an assumption that is mostly redundant.
> 
> ``` {.text escapeinside="||" mathescape="true"}
>   Reduced
>     ReduceNoWarning
>     (ReduceWithPayment (Payment party (Party party) tok2 money))
>     (state|$\llparenthesis$|accounts := newAccount|$\rrparenthesis$|) Close
>     =
>   Reduced wa eff newState newCont
> ```
> 
> A stronger lemma would be valid, which results from eliminating the assumption and changing the conclusion to `positiveMoneyInAccountOrNoAccount y tok3 newAccount`.
> 
> **File `QuiescentResult.thy`, lemma `reduceContractStepPayIsQuiescent`, line *17***
> 
> The assumption $\mathit{cont} = \texttt{Pay}\ \mathit{x21}\ \mathit{x22}\ \mathit{tok}\ \mathit{x23}\ \mathit{x24}$ is unnecessary.
> 
> **File `Timeout.thy`, lemma `timedOutReduce_only_quiescent_in_close_When`, line *43***
> 
> The assumption $\mathit{minTime}\ \mathit{sta} \leq \mathit{iniTime}$ is unnecessary.
> 
> **File `Timeout.thy`, lemma `timedOutReduce_only_quiescent_in_close`, line *122***
> 
> The assumption\ $\mathit{minTime}\ \mathit{sta} \leq \mathit{iniTime}$ is unnecessary. However, removing it will require the later proof `timedOutReduceContractLoop_closes_contract` to be adjusted.
> 
> **File `Timeout.thy`, lemma `timedOutReduceContractLoop_closes_contract`, lines *170, 173***
> 
> The assumptions $\mathit{minTime}\ \mathit{sta} \leq \mathit{iniTime}$ and $\mathit{minTime} \mathit{sta} = \mathit{iniTime}$ are both present. The former is redundant.
> 
> **File `TimeRange.thy`, lemma `reduceStep_ifCaseLtCt_aux`, line *234***
> 
> The assumption\ $\mathit{env} = \llparenthesis \mathit{timeInterval} = (a, b) \rrparenthesis$ is unnecessary.
> 
> **File `ValidState.thy`, lemma `reductionStep_preserves_valid_state_Refund`, line *29***
> 
> The assumption $$\begin{aligned} \mathit{newState} = \llparenthesis &\mathit{accounts} = \mathit{newAccounts}, \mathit{choices} = \mathit{newChoices}, \\ &\mathit{boundValues} = \mathit{newBoundValues}, \mathit{minTime} = \mathit{newMinTime} \rrparenthesis\end{aligned}$$ is unnecessary.

## 2.4 Isabelle implementation

### 2.4.1 Variable shadowing in `applyAllLoop` *(Severity: Medium)*

> **File `Semantics.thy`, function `applyAllLoop`, line *575***
> 
> The $\texttt{cont}$ variable introduced by the pattern match shadows another $\texttt{cont}$ variable, coming from the pattern match of an outer case expression, making the function harder to follow while also making it more error-prone to future changes.

### 2.4.2 Undescriptive name `moneyInPayment` *(Severity: Low)*

> **File `MoneyPreservation.thy`, function `moneyInPayment`, line *5***
> 
> The name of the function can be more precise. Perhaps `moneyInPaymentToParty` or `moneyInExternalPayment` would work.

### 2.4.3 Typo in section name *(Severity: Low)*

> **File `OptBoundTimeInterval.thy`, line `37`** 
> 
> Typo in section name: "Interval intesection".

### 2.4.4 Typo in comment *(Severity: Low)*

> **File `OptBoundTimeInterval.thy`, line `42`** 
> 
> Typo in comment: "endpoits".

### 2.4.5 Unclear need for multiple formulations for positive accounts *(Severity: Low)*

> **File `PositiveAccounts.thy`, `Throughout`** 
> 
> It is unclear what the use is for multiple formulations (and lemmas about) positive accounts. The first formulation (with the theorems `playTraceAux_gtZero` and `playTrace_gtZero`) is not used in any other modules but the alternative formulation is used instead. If both formulations are relevant, then it should be explained why.

### 2.4.6 Variable name discrepancy in `reductionLoop` *(Severity: Low)*

> **File `Semantics.thy`, function `reductionLoop`** 
> 
> When comparing this function against `specification-v3-rc1.pdf`, different names are used for a let-bound variable. It is `a` in the pdf and `newPayments` in the file `Semantics.thy`. There are similar issues in the function `reduceContractStep` in the equation for the `If` case, and in the function `giveMoney`.

### 2.4.7 Typo in constructor *(Severity: Low)*

> **File `Semantics.thy`, function `applyCases`, line *505***
> 
> Apparent typo in the error message constructor: the party mentioned should be $\texttt{party2}$.

### 2.4.8 Unclear function name `calculateNonAmbiguousInterval` *(Severity: Low)*

> **File `Semantics.thy`, function `calculateNonAmbiguousInterval`, line *725***
> 
> The meaning of the function is not obvious. It needs a comment to explain it.

### 2.4.9 Non-modularized file `SingleInputTransactions.thy` *(Severity: Low)*

> **File `SingleInputTransactions.thy`, `Splitting File`** 
> 
> This file is very long, and it covers more than just single-input transactions. For instance, about 530 lines at the beginning are rather dedicated to idempotence of certain operations. Then, the lemmas around lines 530 -- 700 focus on "well-foundedness" of the recursion on contract steps. Then there is also a clear block of lemmas about "distributivity" of semantics over transaction lists.
> 
> Splitting the module, grouping the related lemmas, would help understanding the relationships between the groups.

### 2.4.10 Misleading function names *(Severity: Low)*

> **File `SingleInputTransactions.thy`, function `inputsToTransactions`, line *9***
> 
> This function name is not very descriptive of its meaning. It takes a transaction (both a time interval and a list of inputs) and returns a list of transactions at the same interval containing a single input each. A name like `splitTransactionIntoSingleInputTransactions` would convey better what the input and the output are.
> 
> Moreover, the code would be cleaner if the function takes a single argument of type `Transaction`, instead of asking the caller to rip apart its fields.
> 
> **File `SingleInputTransactions.thy`, function `traceListToSingleInput`, line *18***
> 
> This function name is not descriptive of what it does. Perhaps a more telling name could be `splitTransactionsIntoSingleInputTransactions`.
> 
> **File `SingleInputTransactions.thy`, function `isSingleInput`, line *1222***
> 
> This function should be renamed or repurposed. If renamed, `allAreSingleInput` more accurately reflects the meaning of the function. If repurposed, it should check that a single transaction has a single input, and `all isSingleInput` can be used to express the current behavior.

### 2.4.11 Unused parameter in `maxTransactionCaseList` *(Severity: Low)*

> **File `TransactionBound.thy`, function `maxTransactionCaseList`, line *16***
> 
> This function has a parameter of type `State` that is completely unused and can be removed.

### 2.4.12 Duplicated `isValidInterval` function *(Severity: Low)*

> **File `TimeRange.thy`, function `isValidInterval`, line *231***
> 
> This function duplicates $\texttt{validTimeInterval}$ from $\texttt{OptBoundTimeInterval.thy}$, and the latter has certain additional properties proven about it specifically, so it makes sense to use the latter in both cases.

## 2.5 marlowe-cardano specification

### 2.5.1 Lack of guidelines for creating cooperating contracts *(Severity: Medium)*

> **File `marlowe-cardano-specification.md`, Section `Life Cycle of a Marlowe Contract`** 
> 
> Given that transactions are expected to work with Marlowe and non-Marlowe contracts simultaneously, it would be helpful to offer some guidelines for other contracts to avoid double satisfaction. Some degree of cooperation between the contracts that can appear in the same transaction is unavoidable.
> 
> One measure could be to ask every cooperating contract to refrain from paying to the payout validator. In this way, double satisfaction can not affect the payments of the Marlowe contract, if the Marlowe contract only pays to roles rather than addresses.
> 
> Another alternative would be to demand other contracts' outputs to use datums that are different from the roles used by the Marlowe contract for payments.

### 2.5.2 No reference to creating a minting policy *(Severity: Low)*

> **File `marlowe-cardano-specification.md`, Section `Monetary Policy for Role Tokens`** 
> 
> The minting policy is not specified, but a reference needs to be offered to explain how to create one.

### 2.5.3 Argument for Contract in `txInfoData` not specified *(Severity: Low)*

> **File `marlowe-cardano-specification.md`, Section `Types`** 
> 
> The argument by which the `Contract` in the `txInfoData` list corresponds to the given hash needs to be made explicit.

### 2.5.4 Merkleization section not detailed enough *(Severity: Low)*

> **File `marlowe-cardano-specification.md`, Section `Merkleization`**
> 
> This section is too terse. It needs to explain what Merkleization is, and to motivate why it is needed.
> 
> When explaining how it works, it needs to make explicit that only the `Case` type is modified, and that in the semantics, only the `Input` type is modified. It needs to explain why the `Input` type needs to carry a hash and a contract, and why the evaluation of the contract is changed as described.

### 2.5.5 Unnecessary constraint *(Severity: Low)*

> **File `marlowe-cardano-specification.md`, `Constraint 12. Merkleized continuations`** 
> 
> This constraint is unnecessary to have in the Marlowe validator, since the construction of the arguments for evaluation of the Marlowe contract would fail. However, it would be useful to have it appear in the specification for users to be aware of it when crafting transactions. A note to motivate the presence of the constraint could be helpful.

### 2.5.6 Asymmetry between role and wallet payouts *(Severity: Low)*

> **File `marlowe-cardano-specification.md`, `Constraint 15`, **
> 
> The marlowe validator allows multiple outputs to be paid to a wallet, but it demands that a single output exists when paying to a role instead. The motivation to use different approaches needs to be documented. This is implemented in `Scripts.hs` at line 371, in function `payoutConstraints`.

### 2.5.7 Incorrect description of `rolePayoutValidator` *(Severity: Low)*

> **File `marlowe-cardano-specification.md`, Section `Plutus Validator for Marlowe Payouts`** 
> 
> The description of the Marlowe payout validator in the specification states that it is parameterized by the currency symbol. However, this is not correct as the validator is unparameterized; rather, the datum type of the validator includes the currency symbol (as well as token name). The description should be modified to reflect this.

### 2.5.8 Unspecified initial state *(Severity: Low)*

> **File `marlowe-cardano-specification.md`, Section `Life Cycle of a Marlowe Contract`** 
> 
> The specification should say what the initial state of a Marlowe contract should be. In particular, creating a contract requires giving the minimum Ada to some account in the Marlowe state. Otherwise, Constraint 5 will reject the transactions that try to spend the output.

### 2.5.9 Unspecified behavior when multiple cases can apply *(Severity: Low)*

> **File `Semantics.hs`, Function `applyCases`, line *597***
> 
> If multiple cases in a case list can apply, the first one is taken. This behavior should be better communicated in the specification.

## 2.6 Haskell implementation

### 2.6.1 Name shadowing in `applyAllInputs` *(Severity: Medium)*

> **File `Semantics.hs`, Function `applyAllInputs`, line *658***
> 
> The binding `cont` from the `Applied` constructor shadows the `cont` variable coming from the pattern match in an enclosing case expression. This makes the code error-prone to subsequent changes and refactorings.

### 2.6.2 Non-isomorphic types in `playTraceAux` *(Severity: Medium)*

> **File `Semantics.hs`, Function `playTraceAux`, line *710***
> 
> The function in the Isabelle code takes a `TransactionOutputRecord` while the Haskell version takes a `TransactionOutput`. This means `TransactionError` cannot be an input to `playTraceAux` in Isabelle, possibly invalidating proofs about its properties.

### 2.6.3 Variable names differ from Isabelle code *(Severity: Low)*

> **File `Semantics.hs`, Several functions ``** 
> 
> Different variable names in Isabelle and Haskell make comparison harder. It is less of an issue when only one variable has been renamed in a function, but multiple variable renames require carefully mapping between the names to avoid confusion. For example, the code of `reduceContractStep` in line 482 (`Pay` case) is hard to compare.
> 
> Other name changes include:
> -   Line 456, function `updateMoneyInAccount` uses variable `money` where the Isabelle code uses `amount` and omits naming the last parameter.
> -   Line 473, function `giveMoneyToPay` uses variables `amount` and `accounts` instead of `money` and `accountsV` as in the Isabelle code.
> -   Line 541, function `reductionLoop` uses variable `con` instead of `ncontract`.
> -   Line 300, the data type `TransactionInput` corresponds to the type `Transaction` in the Isabelle code.
> -   Line 313, the data type `TransactionOutput` is isomorphic but not identical to the homonymous data type in the Isabelle code.
> -   Line 439, function `refundOne` uses a variable `balance` where the Isabelle code uses `money`.
> -   Line 463, function `addMoneyToAccount` uses variable `accounts` where the Isabelle code uses `accountsV`.

### 2.6.4 Naming of functions and variables *(Severity: Low)*

> **File `Several files`, `several functions`** 
> 
> Several functions or variables could have more descriptive or precise names, for example:
> -   `Scripts.hs:193`: `validateBalances` could be called `allBalancesArePositive`.
> -   `Scripts.hs:206`: `validateInputs` could be called `allInputsAreAuthorized`.
> -   `Scripts.hs:324`: `checkScriptOutputAny` could be called `noOutputPaysToOwnAddress`, as it checks that *no* outputs pay to the script address.
> -   `Semantics.hs:439`: `refundOne` is named somewhat confusingly, and understanding the name requires the context of `reduceContractStep` where the function is called. Perhaps a better name would be `dropWhileNonPositiveAndUncons`.
> -   `Semantics.hs:597`: the binding `tailCase` should rather be named `tailCases`.

### 2.6.5 Unused functions *(Severity: Low)*

> **File `Several files`, `several functions`** 
> 
> Several functions are unused and perhaps should be removed:
> -   `Semantics.hs:741`: `contractLifespanUpperBound` does not seem to be used anywhere, including tests.
> -   `Semantics.hs:680`: `isClose` is not used in the rest of the codebase (besides checking its behavior via testing). It should either be removed, or comments justifying its existence should be included.
> 
> In addition to that, the functions `validateBalances` and `totalBalance` (defined at `Semantics.hs:755` and `:762`) are only used in `Scripts.hs` and never reused, so they should probably be moved to `Scripts.hs`.

### 2.6.6 Comments *(Severity: Low)*

> **File `Semantics.hs`, Function `refundOne`, line *439***
> 
> The comment describing the function is overly concise, as it does not mention the function removing all non-positive accounts before the first positive one, and effectively `uncons`-ing the list.
> 
> **File `Semantics.hs`, Function `addMoneyToAccount`, line *461***
> 
> There is a typo in the comment: `accoun` is written instead of `account`.

### 2.6.7 Record updates in `playTraceAux` *(Severity: Low)*

> **File `Semantics.hs`, Function `playTraceAux`, line *710***
> 
> The function could have followed the Isabelle code more closely if it used a record update instead of creating a new `TransactionOutput` record from scratch.

### 2.6.8 Potential simplifications *(Severity: Low)*

> **File `Semantics.hs`, Function `totalBalance`, line *755***
> 
> The function uses `foldMap f . AssocMap.toList`. Here, `AssocMap.toList` is redundant.

> **File `Types.hs`, Class instance `Eq Contract`, line *873***
> 
> The equality of cases for the `When` constructor would be simplified by using `cases1 == cases2`. If there is a reason for the more verbose equality condition, it should be outlined in a comment.

### 2.6.9 `computeTransaction` differs from the Isabelle implementation *(Severity: Low)*

> **Helper function `evalValue, evalObservation`, lines *391 (Semantics.hs), 34 (Semantics.thy)***
> 
> `evalValue` and `evalObservation` differ from the Isabelle implementation in the introduction of auxiliary variables to abbreviate the recursive calls. The comparison would be simpler if both definitions were consolidated.
> 
> **Helper function `evalValue`, lines *395 (Semantics.hs), 35 (Semantics.thy)***
> 
> The Isabelle implementation should use the helper function `moneyInAccount` instead of inlining its definition, so as to maintain consistency with the Haskell implementation.
> 
> **Helper function `applyCases`, lines *596 (Semantics.hs), 498 (Semantics.thy)***
> 
> The structure of function `applyCases` differs between the Haskell and Isabelle files. Specifically, the Haskell implementation has an additional function `applyAction` where the Isabelle implementation does not. A comment motivating the discrepancy would be needed. This is likely due to the lack of Merkleization in the Isabelle implementation.
> 
> **Helper function `convertReduceWarnings`, lines *617 (Semantics.hs), 537 (Semantics.thy)***
> 
> The Haskell function is implemented using `foldr`, while the Isabelle function uses explicit recursion, making a one-to-one comparison less obvious. If there is a reason for this discrepancy, such as `foldr` yielding some optimizations, this should be outlined in a comment.

### 2.6.10 Constraint implementations differ from description *(Severity: Low)*

> **File `marlowe-cardano-specification.md`, Section `Plutus Validator for Marlowe Semantics`** 
> 
> Some constraints mentioned in the specification are written in a different structure than the corresponding constraint in `Scripts.hs`. While such a discrepancy may be useful to minimize verbosity, a unified structure when possible would alleviate a side-by-side comparison. Examples of these differing structures include Constraint 6 and Constraint 14.

### 2.6.11 Missing argument of `computeTransaction` *(Severity: Low)*

> **File `marlowe-cardano-specification.md`, Section `Relationship between Marlowe Validator and Semantics`** 
> 
> The specification mentions the output datum as the (fifth) argument for the `computeTransaction` function, while it is not an argument to it.

### 2.6.12 Missing `smallMarloweValidator` *(Severity: Low)*

> **File `marlowe-cardano-specification.md`, Various sections ``** 
> 
> The specification mentions `smallMarloweValidator` in a few places, but it is never mentioned in the source code.

### 2.6.13 Incorrect constraint reference *(Severity: Low)*

> **File `Scripts.hs`, Function `mkRolePayoutValidator`, line *150***
> 
> This line should refer to Constraint 17 rather than Constraint 16.

### 2.6.14 `MarloweParams` differs from the specification *(Severity: Low)*

> **File `Semantics.hs`, type `MarloweParams`, line *355***
> 
> The specification defines `MarloweParams` to contain just the payout validator hash, while the definition in the Haskell code contains just the roles currency symbol.

### 2.6.15 Timeout boundary differs from the specification *(Severity: Low)*

> **File `Semantics.hs`, type `reduceContractStep`, line *518***
> 
> The specification mentions
> 
> > If a valid Transaction is computed with a TimeInterval with a start time bigger than the Timeout t, the contingency continuation c is evaluated.
> 
> where "bigger" implies strict inequality, while the code makes non-strict comparison. This difference needs to be acknowledged and further explained in the specification.

## 2.7 Haskell tests

### 2.7.1 More precise failure checks *(Severity: Medium)*

> **File `Spec/Marlowe/Plutus/Specification.hs`, `Various tests`** 
> 
> The tests use the functions `checkSemanticsTransaction` and `checkPayoutTransaction` to verify that various error conditions cause transactions to be rejected. These functions test that a transaction passes or fails, but when it fails, the functions do not consider the error cause. Checking the exact cause is necessary to ensure the transaction is rejected because of the intended reason and not because of some other error condition arising in a particular test case by coincidence.
> 
> The absence of this information makes it easier to accidentally produce a test that is not testing what is intended.

### 2.7.2 Missing tests *(Severity: Medium)*

> **File `Spec/Marlowe/Semantics/Compute.hs`, ``** 
> 
> The following properties could additionally be tested for `computeTransaction`:
> -   payment subtracts from an account,
> -   deposit adds to an account,
> -   `INotify` input produces the expected continuation,
> -   `IChoice` input produces the expected continuation,
> -   the hash of a successfully applied merkleized input matches the hash of the merkleized case.
> 
> Some of these are tested in `Spec/Marlowe/Semantics/Functions.hs` already for auxiliary functions.
> 
> **File `Spec/Marlowe/Semantics/Functions.hs`, `Missing merkleization tests`** 
> 
> The properties in this module do not seem to be tested with merkleized contracts or inputs except for `checkGetContinuation`. More merkleization tests should be added.
> 
> **File `Spec/Marlowe/Semantics/Compute.hs`, function `checkFixInterval`, lines *100-102***
> 
> The test `checkFixInterval` is never instantiated with an invalid interval that is in the past, meaning the function `fixInterval` is never tested for that case.

### 2023-03-27 response-to-audit-report.md
PLT-3382 Added more space between sections.
PLT-3382 Copied commit information into audit-response report.

@@
 - [ ] Add responses and mitigation commits to Haskell-related findings.
+    - [ ] Convert the commit messages to narrative, and just link to the commit.
 - [ ] Fix display of equations or replace ones that cannot be fixed with ellipses ("...").
@@
 
+## Overview of Isabelle Changes
+
+## Overview of Marlowe-Cardano Specification and Haskell (Plutus) Implementation
+
+TODO: @bwbush will add text here.
+
+```console
+commit fca49124f9ebad285bb08d643a3c5fcf5e1df618
+Merge: 27a3e32bb 6a0517770
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Wed Mar 15 13:50:51 2023 -0600
+
+    Merge pull request #524 from input-output-hk/PLT-3064
+
+    PLT-3064 Cumulative validator changes based on audit report
+```
+
 # 2 Findings
@@
 > 
-> This disagreement can be exploited to steal money from a flawed Marlowe contract that allows a negative deposit. The issue is demonstrated in .
+> This disagreement can be exploited to steal money from a flawed Marlowe contract that allows a negative deposit. The issue is demonstrated in ….
+
+TODO: @bwbush will add text here.
+
+```console
+commit 15446073131c4098fed781cdcc701b89b77f6bd0
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Wed Mar 1 16:56:31 2023 -0700
+
+    SCP-5123 Property-based tests for contracts with negative deposits.
+
+commit 4b81ee96b1e7c74c4964cf25f6bb599c80d07a52
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Wed Mar 1 16:56:31 2023 -0700
+
+    SCP-5123 Property-based tests for contracts with negative deposits.
+
+commit 9d5443c5e9306eaf49ac5389df455e07555c9e04
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Thu Mar 2 10:54:48 2023 -0700
+
+    SCP-5123 Improved negative-deposit test case.
+
+commit 20bf9f13fe33acf9612ec26fe640d27975b0dc59
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Wed Mar 1 17:10:11 2023 -0700
+
+    SCP-5123 Rephrased comment.
+
+commit bbb9b8fc21ad93b7bcdb3b60be0f2f9c921f6bec
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Wed Mar 1 16:56:31 2023 -0700
+
+    SCP-5123 Property-based tests for contracts with negative deposits.
+
+commit 597281c176bcb1f326329ba40b0d968d5e7eb84b
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Wed Mar 1 11:36:56 2023 -0700
+
+    SCP-5123 Updated validator script size and hash tests.
+
+commit 68791facc195068717cbc6e55d1e4fdbe1f4a521
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Wed Mar 1 11:20:27 2023 -0700
+
+    SCP-5123 Ensured that negative deposits do not corrupt accounting equation.
+```
 
@@
 
+TODO: @bwbush will add text here.
+
+```console
+commit 0bf75ac775e53507db4e313f705a7a667f98aac1
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Sat Mar 4 09:46:24 2023 -0700
+
+    SCP-5125 Added property-based tests for multiple validators.
+
+commit 9f4fc9cde581b6dbbb939ec9d8e6303b8b67350e
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Fri Mar 3 08:54:31 2023 -0700
+
+    SCP-5141 Property-based tests for duplicate detection in validator.
+
+commit 435ac680eb8119db1c43ec968cf8ea4ef08158b6
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Sat Mar 4 09:46:24 2023 -0700
+
+    SCP-5125 Added property-based tests for multiple validators.
+
+commit 39fcc8aa9d61ed97dfa56e7ce078bf5e4c4cd6cc
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Fri Mar 3 08:54:31 2023 -0700
+
+    SCP-5141 Property-based tests for duplicate detection in validator.
+
+commit 6f6331b463e7f2fddce7ec0f64b37b030f2f795b
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Fri Mar 3 08:54:31 2023 -0700
+
+    SCP-5141 Property-based tests for duplicate detection in validator.
+
+commit 3f222bc363fcf3e9cd2fa0cb5206933ef597a619
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Sat Mar 4 09:46:24 2023 -0700
+
+    SCP-5125 Added property-based tests for multiple validators.
+
+commit 5f673c477ccc9b38626dca505a399962b22c2675
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Sat Mar 4 09:45:32 2023 -0700
+
+    SCP-5125 Included implicit close payments in satisfaction restriction.
+
+commit 4adf115dbf795e4b014eb6ccab0acb20a73e74ed
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Fri Mar 3 13:54:52 2023 -0700
+
+    SCP-5125 Prevented double-satisfaction.
+```
+
 ### 2.1.3 Missing constructor in equality instance *(Severity: High)*
@@
 > The constructor `ReduceAssertionFailed` is not mentioned and compares `False` against itself. This might cause validators to fail checking the presence of this particular warning.
+
+TODO: @bwbush will add text here.
+
+```console
+commit 04805a393c0815fe0f2a1a6d7f76d9867ffe0c14
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Thu Mar 2 08:27:25 2023 -0700
+
+    SCP-5129 Hardened `Eq` instances in validator files.
+
+    See https://github.com/input-output-hk/marlowe-cardano/pull/513#discussion_r1123062500
+    for motivation.
+
+commit 84d65a70c29320fab501ad759bbc71855d1b63fc
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Wed Mar 1 17:21:39 2023 -0700
+
+    SCP-5129 Fixed `(==)` for `ReduceAssertionFailed`.
+```
 
@@
 
+TODO: @bwbush will add text here.
+
+```console
+commit 8855feae473882b82643db792327423152e45b5c
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Thu Mar 2 15:27:42 2023 -0700
+
+    SCP-5141 SCP-5142 Enforced initial and final state validity in validator.
+
+commit ebab31d9ba6dd322d374b34cad3995050a0e476e
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Thu Mar 2 11:52:04 2023 -0700
+
+    SCP-5124 Added property-based test for new final-balance constraint.
+```
+
 ### 2.1.8 Non-validated Marlowe states *(Severity: High)*
@@
 
+TODO: @bwbush will add text here.
+
+```console
+commit 8855feae473882b82643db792327423152e45b5c
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Thu Mar 2 15:27:42 2023 -0700
+
+    SCP-5141 SCP-5142 Enforced initial and final state validity in validator.
+
+commit ebab31d9ba6dd322d374b34cad3995050a0e476e
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Thu Mar 2 11:52:04 2023 -0700
+
+    SCP-5124 Added property-based test for new final-balance constraint.
+```
+
 ### 2.1.9 Total balance of ending state uncomputed *(Severity: High)*
@@
 > 
-> > The beginning balance plus the deposits equals the ending balance plus
-> > the payments.
+> > The beginning balance plus the deposits equals the ending balance plus the payments.
 > 
@@
 
+TODO: @bwbush will add text here.
+
+```console
+commit 8855feae473882b82643db792327423152e45b5c
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Thu Mar 2 15:27:42 2023 -0700
+
+    SCP-5141 SCP-5142 Enforced initial and final state validity in validator.
+
+commit ebab31d9ba6dd322d374b34cad3995050a0e476e
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Thu Mar 2 11:52:04 2023 -0700
+
+    SCP-5124 Added property-based test for new final-balance constraint.
+
+commit d514bcd075732b07e1c6a73e2e3c68afa01acf2c
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Thu Mar 2 11:52:04 2023 -0700
+
+    SCP-5124 Added property-based test for new final-balance constraint.
+
+commit 7f562545d013dd17472b692b550ffdc7f8a383f3
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Thu Mar 2 11:52:04 2023 -0700
+
+    SCP-5124 Added property-based test for new final-balance constraint.
+```
+
 ### 2.1.10 Unchecked ending balance *(Severity: High)*
@@
 > The specification should at least discuss why the check is absent together with the other similar checks that are not implemented (checking that ending accounts have positive balances, checking that the ending Marlowe state is valid).
+
+```console
+commit 0a890845c44ccfe4df97e765e9b9eb743ce7d580
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Thu Mar 2 09:32:09 2023 -0700
+
+    SCP-5124 Ensured script output is consistent with output state.
+
+commit 201c5df9de29b3b142be5ff7d0a9b56d1b378204
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 13:35:40 2023 -0700
+
+    PLT-4169 Optimized checking of script output in validator.
+
+commit 26f024e80cea0dd4ebf8bff0f1a10b97aea87894
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Thu Mar 9 12:43:12 2023 -0700
+
+    PLT-3064 Made suggested changes to `examineScripts` in validator.
+
+    See https://github.com/input-output-hk/marlowe-cardano/pull/524#discussion_r1127270744.
+```
 
@@
 
+TODO: @bwbush will add text here.
+
+```console
+commit 9e068b6a60103f2a40604a7eaea57c5f29b59180
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Wed Mar 1 11:00:45 2023 -0700
+
+    SCP-5126 Aligned `Gen MList` with semantics.
+
+commit b65ccfec3708484ff52514671e0e227c13f084f3
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Wed Mar 1 10:52:17 2023 -0700
+
+    SCP-5126 Hardened equivalence testing for `MList` vs `AssocMap`.
+
+commit a95a616aca59bc7159d3ffa92487f8f021820bff
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Wed Mar 1 10:28:54 2023 -0700
+
+    SCP-5126 Elaborated comments on `MList` property tests.
+
+commit e2277677987f359dcd8754275eccd5b7c9a880a7
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Wed Mar 1 10:23:16 2023 -0700
+
+    SCP-5126 Added note on invariants in unpack and repacking `AssocMap`.
+
+commit af380a292c0369bcc77d8584d81ea80a31326b61
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Feb 28 13:49:44 2023 -0700
+
+    SCP-5126 Property-based tests for `MList` vs `AssocMap`.
+
+commit a2ff6aa334efb7b7955408d29cf0e2edf06bca08
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Feb 28 12:05:32 2023 -0700
+
+    SCP-5126 Annotated Marlowe validator regarding association lists.
+```
+
 ### 2.1.13 Missing specification tests *(Severity: Medium)*
@@
 Besides checking that there are no translation mistakes, these properties would also help contrasting the assumptions in the Isabelle and the Haskell sides, like the meaning of validity of an association list, which is focused in the previous issue.
+
+TODO: @bwbush will add text here.
 
@@
 
+TODO: @bwbush will add text here.
+
+```console
+commit 00010ea48439892ff9eec54fb93929f56bcb6c7b
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 11:20:14 2023 -0700
+
+    PLT-4169 Added text on running Plutus with Marlowe to specification doc.
+
+    This addresses the following audit comment:
+
+    > File marlowe-cardano-specification.md, Section Life Cycle of a
+    > Marlowe Contract Given that transactions are expected to work
+    > with Marlowe and non-Marlowe contracts simultaneously, it would
+    > be helpful to offer some guidelines for other contracts to avoid
+    > double satisfaction. Some degree of cooperation between the
+    > contracts that can appear in the same transaction is unavoidable.
+    >
+    > One measure could be to ask every cooperating contract to refrain
+    > from paying to the payout validator. In this way, double satisfaction
+    > can not affect the payments of the Marlowe contract. Another
+    > alternative would be to demand other contracts’ outputs to use datums
+    > that are different from the roles used by the Marlowe contract for
+    > payments.
+```
+
 ### 2.5.2 No reference to creating a minting policy *(Severity: Low)*
@@
 
+TODO: @bwbush will add text here.
+
+```console
+commit fe00af7d622bbd17c3327363d4650ee544d1fff7
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 11:25:09 2023 -0700
+
+    PLT-4169 Added text to specification on how to mint role tokens.
+
+    This addresses the audit comment:
+
+    > File marlowe-cardano-specification.md, Section Monetary Policy
+    > for Role Tokens The minting policy is not specified, but a
+    > reference needs to be offered to explain how to create one.
+```
+
 ### 2.5.3 Argument for Contract in `txInfoData` not specified *(Severity: Low)*
@@
 > The argument by which the `Contract` in the `txInfoData` list corresponds to the given hash needs to be made explicit.
+
+TODO: @bwbush will add text here.
+
+```console
+commit b6366efc9f8df38bfe9232f81e6caf2c7f11ffa8
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 11:31:11 2023 -0700
+
+    PLT-4169 Made explicit the relationship between continuation and its hash.
+
+    This addresses the audit comment:
+
+    > File marlowe-cardano-specification.md, Section Types The argument
+    > by which the Contract in the txInfoData list corresponds to the
+    > given hash needs to be made explicit.
+```
 
@@
 It needs to explain why the `Input` type needs to carry a hash and a contract, and why the evaluation of the contract is changed as described.
+
+TODO: @bwbush will add text here.
+
+```console
+commit a9d4043230cb4f240fab2d816bed6a3cab4f5053
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 11:44:32 2023 -0700
+
+    PLT-4169 Elaborated explanation of and motivation for merkleization.
+
+    This addresses the audit comment:
+
+    > File marlowe-cardano-specification.md, Section Merkleization
+    > This section is too terse. It needs to explain what Merkleization
+    > is, and to motivate why it is needed. When explaining how it works,
+    > it needs to make explicit that only the Case type is modified, and
+    > that in the semantics, only the Input type is modified. It needs
+    > to explain why the Input type needs to carry a hash and a contract,
+    > and why the evaluation of the contract is changed as described.
+```
 
@@
 
+TODO: @bwbush will add text here.
+
+```console
+commit fda1d3e862fc8f8c31ff479fbd338226e661eb0b
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 11:50:20 2023 -0700
+
+    PLT-4169 Clarified nature of Constraint 12 in specification.
+
+    This addresses the auditor comment:
+
+    > File marlowe-cardano-specification.md, Constraint 12. Merkleized
+    > continuations This constraint is unnecessary to have in the
+    > Marlowe validator, since the construction of the arguments for
+    > evaluation of the Marlowe contract would fail. However, it would
+    > be useful to have it appear in the specification, for users to
+    > be made aware of it when crafting transactions. A note to motivate
+    > the presence of the constraint could be helpful.
+```
+
 ### 2.5.6 Asymmetry between role and wallet payouts *(Severity: Low)*
@@
 
+TODO: @bwbush will add text here.
+
+```console
+commit 4e81470de90ddd6858c3ad8aace96e90df5a86ae
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Wed Mar 1 17:40:38 2023 -0700
+
+    SCP-5128 Commented on rationale for allowing multiple outputs to an address.
+```
+
 ### 2.5.7 Incorrect description of `rolePayoutValidator` *(Severity: Low)*
@@
 
+TODO: @bwbush will add text here.
+
+```console
+commit 1d16de1f1b7a167dc4eb4a057a3df2c24d0194c9
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 11:12:36 2023 -0700
+
+    PLT-4169 Fixed example call to `rolePayoutValidator`.
+
+    This addresses the following audit finding:
+
+    > File marlowe-cardano-specification.md, Section Plutus Validator
+    > for Marlowe Payouts The type signature of rolePayoutValidator
+    > is missing the datum argument. It should be: rolePayoutValidator
+    > :: CurrencySymbol -> TokenName -> () -> ScriptContext -> Bool
+```
+
 ### 2.5.8 Unspecified initial state *(Severity: Low)*
@@
 
+TODO: @bwbush will add text here.
+
+```console
+commit e150e30867dfa5466639f46bf39656aee2888e38
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 12:34:38 2023 -0700
+
+    PLT-4169 Elaborated in specification on the required initial state.
+
+    This addresses the audit comment:
+
+    > File marlowe-cardano-specification.md, Section Life Cycle of a
+    > Marlowe Contract The specification should say what the initial
+    > state of a Marlowe contract should be. In particular, creating
+    > a contract requires giving the minimum Ada to some account in
+    > the Marlowe state. Otherwise, Constraint 5 will reject the
+    > transactions that try to spend the output.
+```
+
 ### 2.5.9 Unspecified behavior when multiple cases can apply *(Severity: Low)*
@@
 > If multiple cases in a case list can apply, the first one is taken. This behavior should be better communicated in the specification.
+
+TODO: @bwbush will add text here.
+
+```console
+commit 08d6f34ac91da7aa34e7414c5bae0e33e9235d74
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 12:06:10 2023 -0700
+
+    PLT-4169 Improved documentation of `applyCases`.
+
+    This addresses the audit comment:
+
+    > File Semantics.hs, Function applyCases, line 597 If multiple
+    > cases in a case list can apply, the first one is taken. This
+    > behavior should be better communicated in the specification.
+```
 
@@
 
+TODO: @bwbush will add text here.
+
+```console
+```
+
 ### 2.6.2 Non-isomorphic types in `playTraceAux` *(Severity: Medium)*
@@
 This means `TransactionError` cannot be an input to `playTraceAux` in Isabelle, possibly invalidating proofs about its properties.
+
+TODO: @bwbush will add text here.
+
+```console
+```
 
@@
 
+TODO: @bwbush will add text here.
+
+```console
+commit 1c96edfe7e22ecc499ed75dd175490a4f2427bdc
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 13:24:06 2023 -0700
+
+    PLT-4169 Renamed functions, per audit recommendations.
+```
+
 ### 2.6.4 Naming of functions and variables *(Severity: Low)*
@@
 
+TODO: @bwbush will add text here.
+
+```console
+commit 1c96edfe7e22ecc499ed75dd175490a4f2427bdc
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 13:24:06 2023 -0700
+
+    PLT-4169 Renamed functions, per audit recommendations.
+```
+
 ### 2.6.5 Unused functions *(Severity: Low)*
@@
 
+TODO: @bwbush will add text here.
+
+```console
+```
+
 ### 2.6.6 Comments *(Severity: Low)*
@@
 
+TODO: @bwbush will add text here.
+
+```console
+```
+
 ### 2.6.7 Record updates in `playTraceAux` *(Severity: Low)*
@@
 
+TODO: @bwbush will add text here.
+
+```console
+```
+
 ### 2.6.8 Potential simplifications *(Severity: Low)*
@@
 > The equality of cases for the `When` constructor would be simplified by using `cases1 == cases2`. If there is a reason for the more verbose equality condition, it should be outlined in a comment.
+
+TODO: @bwbush will add text here.
+
+```console
+```
 
@@
 The comparison would be simpler if both definitions were consolidated.
-> 
+
 > **Helper function `evalValue`, lines *395 (Semantics.hs), 35 (Semantics.thy)***
@@
 > The Isabelle implementation should use the helper function `moneyInAccount` instead of inlining its definition, so as to maintain consistency with the Haskell implementation.
-> 
+
 > **Helper function `applyCases`, lines *596 (Semantics.hs), 498 (Semantics.thy)***
@@
 This is likely due to the lack of Merkleization in the Isabelle implementation.
-> 
+
+TODO: @bwbush will add text here.
+
+```console
+commit d0a3834ae8484bb6c7d0ecd3aeec60871b3fd165
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 13:00:28 2023 -0700
+
+    PLT-4169 Added comment about merkleization to `applyCases`.
+
+    This addresses the audit comment:
+
+    > Helper function applyCases, lines 596 (Semantics.hs), 498
+    > (Semantics.thy) The structure of function applyCases differs
+    > between the Haskell and Isabelle files. Specifically, the
+    > Haskell implementation has an additional function applyAction
+    > where the Isabelle implementation does not. A comment
+    > motivating the discrepancy would be needed. This is likely
+    > due to the lack of Merkleization in the Isabelle implementation.
+```
+
 > **Helper function `convertReduceWarnings`, lines *617 (Semantics.hs), 537 (Semantics.thy)***
@@
 If there is a reason for this discrepancy, such as `foldr` yielding some optimizations, this should be outlined in a comment.
+
+TODO: @bwbush will add text here.
+
+```console
+commit d14776101c22eac7fbc67893020f37cbc2a9e339
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 13:02:40 2023 -0700
+
+    PLT-4169 Added note on use of `foldr` in semantics.
+
+    This addresses the audit comment:
+
+    > Helper function convertReduceWarnings, lines 617 (Semantics.hs),
+    > 537 (Semantics.thy) The Haskell function is implemented using
+    > foldr, while the Isabelle function uses explicit recursion,
+    > making a oneto-one comparison less obvious. If there is a
+    > reason for this discrepancy, such as foldr yielding some
+    > optimizations, this should be outlined in a comment.
+```
 
@@
 
+TODO: @bwbush will add text here.
+
+```console
+```
+
 ### 2.6.11 Missing argument of `computeTransaction` *(Severity: Low)*
@@
 
+TODO: @bwbush will add text here.
+
+```console
+commit 00a0c2406665f06753aecf6dbb63933edb275436
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 12:39:22 2023 -0700
+
+    PLT-4169 Fixed correspondence between semantics and validator.
+
+    This addresses the audit comment:
+
+    > File marlowe-cardano-specification.md, Section Relationship
+    > between Marlowe Validator and Semantics The specification
+    > mentions the output datum as the (fifth) argument for the
+    > computeTransaction function, while it is not an argument to it.
+```
+
 ### 2.6.12 Missing `smallMarloweValidator` *(Severity: Low)*
@@
 
+TODO: @bwbush will add text here.
+
+```console
+commit 87cf548d9ae7934a702ea1a1c85d7d9bd4993aa9
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 12:37:21 2023 -0700
+
+    PLT-4169 Fixed references to `smallMarloweValidator`.
+
+    This addresses the audit comment:
+
+    > File marlowe-cardano-specification.md, Various sections
+    > The specification mentions smallMarloweValidator in a few
+    > places, but it is never mentioned in the source code.
+```
+
 ### 2.6.13 Incorrect constraint reference *(Severity: Low)*
@@
 
+TODO: @bwbush will add text here.
+
+```console
+commit 3fac0d1749d107f13743756aa50c9723e776e391
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 12:42:01 2023 -0700
+
+    PLT-4169 Fixed incorrect constraint reference.
+
+    This addresses the audit comment:
+
+    > File Scripts.hs, Function mkRolePayoutValidator, line 150 This line should refer to Constraint 17
+    rather than Constraint 16.
+```
+
 ### 2.6.14 `MarloweParams` differs from the specification *(Severity: Low)*
@@
 
+TODO: @bwbush will add text here.
+
+```console
+commit 8691f3351e55758cc750eee0800b1458f0a8c554
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 12:44:53 2023 -0700
+
+    PLT-4169 Fixed `MarloweParams` in specification.
+
+    This addresses the audit comment:
+
+    > File Semantics.hs, type MarloweParams, line 355 The specification
+    > defines MarloweParams to contain just the payout validator hash,
+    > while the definition in the Haskell code contains just the roles
+    > currency symbol.
+```
+
 ### 2.6.15 Timeout boundary differs from the specification *(Severity: Low)*
@@
 > where "bigger" implies strict inequality, while the code makes non-strict comparison. This difference needs to be acknowledged and further explained in the specification.
+
+TODO: @bwbush will add text here.
+
+```console
+```
 
@@
 > The absence of this information makes it easier to accidentally produce a test that is not testing what is intended.
+
+TODO: @bwbush will add text here.
+
+```console
+```
 
@@
 > Some of these are tested in `Spec/Marlowe/Semantics/Functions.hs` already for auxiliary functions.
-> 
+
+TODO: @bwbush will add text here.
+
+```console
+commit ee2e52220e25736af29a1f01144d5fafbe9eed76
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 07:36:59 2023 -0700
+
+    PLT-4168 Tests that choice and deposit continue as expected.
+
+commit 333eee293aa3afe95d0aa7f678c7e7b7bad660ea
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 07:18:27 2023 -0700
+
+    PLT-4168 Test that choice produces expected continuation.
+
+commit 477284c78095482eba63adc2e6090ea68553db9d
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 06:54:14 2023 -0700
+
+    PLT-4168 Tested that deposits add to account.
+
+commit d241669bfb7ec8dad1cc92a7ca6185d08229817e
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Mon Mar 6 14:38:09 2023 -0700
+
+    PLT-4168 Tested that payout substracts from account.
+
+commit f33fc7206522031c93253323f3b8ca083d8ca8c7
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Tue Mar 7 09:56:30 2023 -0700
+
+    PLT-4168 Tested that Merkleization continues as expected.
+```
+
 > **File `Spec/Marlowe/Semantics/Functions.hs`, `Missing merkleization tests`** 
@@
 > The properties in this module do not seem to be tested with merkleized contracts or inputs except for `checkGetContinuation`. More merkleization tests should be added.
-> 
+
+TODO: @bwbush will add text here.
+
+```console
+commit 2a936bf1bbb890cc811b966b46a10fd38314dc25
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Mon Mar 6 11:40:56 2023 -0700
+
+    PLT-4168 Increased coverage of merkleization in validator tests.
+
+    This addresses the following audit finding:
+
+    > File Spec/Marlowe/Semantics/Functions.hs, Missing
+    > merkleization tests The properties in this module
+    > do not seem to be tested with merkleized contracts
+    > or inputs except for checkGetContinuation, and
+    > they probably should.
+
+commit 8f8183d7f14c6bdc3841a1999077ef01ebac4144
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Mon Mar 6 11:40:56 2023 -0700
+
+    PLT-4168 Increased coverage of merkleization in validator tests.
+
+    This addresses the following audit finding:
+
+    > File Spec/Marlowe/Semantics/Functions.hs, Missing
+    > merkleization tests The properties in this module
+    > do not seem to be tested with merkleized contracts
+    > or inputs except for checkGetContinuation, and
+    > they probably should.
+```
+
 > **File `Spec/Marlowe/Semantics/Compute.hs`, function `checkFixInterval`, lines *100-102***
@@
 > The test `checkFixInterval` is never instantiated with an invalid interval that is in the past, meaning the function `fixInterval` is never tested for that case.
+
+TODO: @bwbush will add text here.
+
+```console
+commit 811f048ba93db10972641c0c235864db1d6014e7
+Author: Brian W Bush <brian.bush@iohk.io>
+Date:   Mon Mar 6 10:16:08 2023 -0700
+
+    PLT-4168 Added test for invalid interval in past.
+
+    This addresses the following audit finding:
+
+    > File Spec/Marlowe/Semantics/Compute.hs, function checkFixInterval,
+    > lines 100-102 The test checkFixInterval is never instantiated
+    > with an invalid interval that is in the past, meaning the function
+    > fixInterval is never tested for that case.
+```

### 2023-03-29 response-to-audit-report.md
PLT-3382 Completed documentation for all but one audit mitigation.

--- 2023-03-27 response-to-audit-report.md
@@
 
-- [ ] Add prefatory text.
+- [ ] Add prefatory text and executive summary.
 - [ ] Add responses and mitigation commits to Isabelle-related findings.
-- [ ] Add responses and mitigation commits to Haskell-related findings.
+- [x] Add responses and mitigation commits to Haskell-related findings.
 - [ ] Fix display of equations or replace ones that cannot be fixed with ellipses ("...").
@@
 
-# 2 Findings
+## Executive Summary
+
+TODO: @bwbush will add text here after everything else is written.
+
+## Acknowledgement
+
+TODO: @bwbush will add text thanking the audit team.
+
+## Overview of Isabelle Changes
+
+## Overview of Changes to Marlowe-Cardano Specification and Haskell (Plutus) Implementation
+
+The audit report's main and high-severity concerns for the Plutus validators for Marlowe on Cardano fell into to four clusters:
+
+1. Incorrect handling of negative deposits
+    - "2.1.1 Negative deposits allow stealing funds"
+2. Incomplete prevention of double satisfaction
+    - "2.1.2 Contracts vulnerable to double satisfaction attacks"
+3. Lax enforcement of invariants in Marlowe state
+    - "2.1.7 Positive balances are not checked for the output state"
+    - "2.1.8 Non-validated Marlowe states"
+    - "2.1.9 Total balance of ending state uncomputed"
+    - "2.1.10 Unchecked ending balance"
+4. Difference semantics of association lists in Isabelle and Plutus
+    - "2.1.12 Different insertion functions used in Isabelle and Haskell code"
+
+Incorrect handling of negative deposits (item 1 above) was mitigated by the addition of a guard against this in Marlowe's Plutus validator so that its behavior was consistent with Marlowe semantics.
+Property-based tests were also added to check the correctness of the mitigation.
+
+Although the Marlowe validator had prevented double-satisfaction among multiple copies of the Marlowe validator script running in the same transaction, it did not prevent double-satisfaction in cases where the Marlowe validator run alongside another Plutus validator in the same transaction (item 2 above).
+Double satisfaction has now been prevented by enforcing that the Marlowe validator is the only Plutus script that runs in transactions that make payments to parties.
+This allows Marlowe contracts to coordinate with other Plutus scripts, but only under conditions where double satisfaction is impossible.
+Once again, property-based tests were added to check the correctness of this mitigation.
+
+The Marlowe validator had made optimistic assumptions about its own correct operation and hence did not check certain invariants in order to save on Plutus execution costs (item 3 above).
+The Marlowe semantics validator has now been thoroughly hardened against corruption of initial or final state so that it ensures that the three state-invariants of positive accounts, non-duplication of state entries (accounts, choices, and bound values), and a total internal value matching the script UTxO's value.
+This hardening was completed without unduly increasing the size (in bytes) of the validator or its Plutus execution cost: such was verified with simulated and on-chain measurement of execution costs for a library of real-life Marlowe contracts.
+Notably, the enforcement of the final invariants protects against funds becoming permanently locked in a Marlowe contract due to a garbled state: even in cases where the implementation of Marlowe semantics or parts of the Plutus validator are flawed, at least some of the execution paths of a Marlowe contract remain viable, so funds could be released even under nominally impossible but catastrophic difficulties.
+The property-based test suite was significantly enhanced to
+
+Mitigation of the difference between association lists (item 4 above) in Isabelle (`MList`) and Plutus (`AssocMap`) was handled by the aforementioned enforcement of invariants, by line-by-line code inspection and annotation, and by thoroughly enhanced property-based testing.
+Work is in progress on a formal proof of the equivalence of `MList` and `AssocMap` under pre-conditions that hold within Marlowe semantics.
+Note that porting Isabelle's `MList` implementation to Plutus would have enlarged the Marlowe semantics validator by at least 2000 bytes and rendered it too costly to execute within the Cardano ledger's present execution-cost limits.
+
+The audit report for Marlowe on Cardano was based on the commit hash [523f3d56](https://github.com/input-output-hk/marlowe-cardano/commit/523f3d56f22bf992ddb0b0c8a52bb7a5a188f9e9).
+The revisions and mitigations discussed here apply to [69468d66](https://github.com/input-output-hk/marlowe-cardano/commit/69468d6623e24a4ccd6659e378ae012c38ca01ce) of that repository.
+The appendices to this report list the differences between pre- and post-audit specification and validator code for Marlowe on Cardano.
+
+The medium-severity concerns in the audit report center around missing tests or name shadowing.
+With the exception of "2.7.1 More precise failure checks", these have all been remedied either by code changes (in the case of the shadowing) or by the addition of a significant augmentation of the property-based test suite.
+The audit's low-severity concerns generally relate to insufficiently detailed text in the Marlowe Cardano specification, the need for more elaborate comments in the code, naming of variable bindings, or typographical errors.
+Nearly all of these finds have been addressed, with only the exception of a few cosmetic recommendations that were not adopted.
+
+# 2 Responses to Specific Findings
 
@@
 > 
-> This disagreement can be exploited to steal money from a flawed Marlowe contract that allows a negative deposit. The issue is demonstrated in .
+> This disagreement can be exploited to steal money from a flawed Marlowe contract that allows a negative deposit. The issue is demonstrated in […].
+
+This vulnerability was mitigated in commit [68791fac](https://github.com/input-output-hk/marlowe-cardano/commit/68791facc195068717cbc6e55d1e4fdbe1f4a521) by adding a guard against negative deposits in the Marlowe semantics validator.
+That guard ensures that the validator's semantics for negative deposits matches Marlowe's Isabelle semantics: namely a deposit of a negative quantity is treated as a deposit of zero.
+Thus a negative deposit will not decrement any of the account balances in the Plutus datum and the total of the internal balances will match the value held in the UTxO output to the Marlowe semantics script address.
+
+A new golden test for negative deposits was added in commit [bbb9b8fc](https://github.com/input-output-hk/marlowe-cardano/commit/bbb9b8fc21ad93b7bcdb3b60be0f2f9c921f6bec) and property-based tests for negative deposits were added in commits [bbb9b8fc](https://github.com/input-output-hk/marlowe-cardano/commit/bbb9b8fc21ad93b7bcdb3b60be0f2f9c921f6bec), [9d5443c5](https://github.com/input-output-hk/marlowe-cardano/commit/9d5443c5e9306eaf49ac5389df455e07555c9e04), [4b81ee96](https://github.com/input-output-hk/marlowe-cardano/commit/4b81ee96b1e7c74c4964cf25f6bb599c80d07a52), and [15446073](https://github.com/input-output-hk/marlowe-cardano/commit/15446073131c4098fed781cdcc701b89b77f6bd0).
 
@@
 This implies that these outputs are vulnerable to double satisfaction in transactions involving other contracts that pay to the same wallets.
-An example is discussed in .
+An example is discussed in […].
 > 
@@
 
+The Marlowe semantics validator had prevented a limited form of double-satisfaction attack by preventing two Marlowe validators to execute in the same transaction.
+The audit report correctly highlights that such an attack could be feasible in cases where another Plutus validator (not the same version of the Marlowe semantics validator) executes in a transactions where a payment to an address or to the Marlowe role-payout validator satisfies both the Marlowe semantics validator and the other validator.
+
+Although the audit report's suggestion of including a unique datum (such as that identifying the contract instance) in any output payment from the contract would likely prevent double satisfaction, implementing such a change in the Marlowe validators would have entailed such extensive changes to the specification and implementation that the overall applicability of the audit findings might subsequently be called into question.
+Instead, a minimal change to the specification and implementation mitigates this vulnerability: the specification was amended with a "Constraint 20.
+Single satisfaction" that requires the Marlowe semantics validator be the only Plutus script running in the transaction *if payments to parties are being made in the transaction*.
+Thus, only allowing one script to validate in case of external payments completely eliminates the possibility that one payment would satisfy two scripts.
+Other contracts are permitted to cooperate with Marlowe in cases such as deposits, choices, and notifications where Marlowe is not making payments.
+
+Commits [4adf115d](https://github.com/input-output-hk/marlowe-cardano/commit/4adf115dbf795e4b014eb6ccab0acb20a73e74ed) and [5f673c47](https://github.com/input-output-hk/marlowe-cardano/commit/5f673c477ccc9b38626dca505a399962b22c2675) augment the Marlowe-Cardano specification and the implementation of the Marlowe semantics validator.
+Commits [3f222bc3](https://github.com/input-output-hk/marlowe-cardano/commit/3f222bc363fcf3e9cd2fa0cb5206933ef597a619), [6f6331b4](https://github.com/input-output-hk/marlowe-cardano/commit/6f6331b463e7f2fddce7ec0f64b37b030f2f795b), [39fcc8aa](https://github.com/input-output-hk/marlowe-cardano/commit/39fcc8aa9d61ed97dfa56e7ce078bf5e4c4cd6cc), [435ac680](https://github.com/input-output-hk/marlowe-cardano/commit/435ac680eb8119db1c43ec968cf8ea4ef08158b6), and [9f4fc9cd](https://github.com/input-output-hk/marlowe-cardano/commit/9f4fc9cde581b6dbbb939ec9d8e6303b8b67350e) implement property-based tests for the new single-satisfaction constraint.
+
 ### 2.1.3 Missing constructor in equality instance *(Severity: High)*
@@
 > The constructor `ReduceAssertionFailed` is not mentioned and compares `False` against itself. This might cause validators to fail checking the presence of this particular warning.
+
+Although this `Eq` instance is not used by the Marlowe validators (so it does not have implications for validator security), the equality test has been fixed because it is a liability for off-chain code and for potential future versions of validators that might perform such an equality test.
+Commit [84d65a70](https://github.com/input-output-hk/marlowe-cardano/commit/84d65a70c29320fab501ad759bbc71855d1b63fc) fixes this and commit [04805a39](https://github.com/input-output-hk/marlowe-cardano/commit/04805a393c0815fe0f2a1a6d7f76d9867ffe0c14) hardens all of the other `Eq` instances to prevent similar problems re-arising if sum types are augmented in the future.
 
@@
 
+The Plutus code for the Marlowe semantics validator had been implemented with the optimistic assumption that semantics would not be flawed and that the additional execution cost of checking the output state was not warranted.
+However, as the audit report points out, such a flaw would result in the contract being forever blocked from further execution.
+Adding a check on the final output's validity would prevent a contract from ever reaching a fully blocked state.
+Even if the semantics or Plutus code were flawed, at least some execution paths in the contract would remain viable, so that the contract could eventually terminate and would not permanently lock funds.
+
+Commits [0a890845](https://github.com/input-output-hk/marlowe-cardano/commit/0a890845c44ccfe4df97e765e9b9eb743ce7d580), [8855feae](https://github.com/input-output-hk/marlowe-cardano/commit/8855feae473882b82643db792327423152e45b5c), [26f024e8](https://github.com/input-output-hk/marlowe-cardano/commit/26f024e80cea0dd4ebf8bff0f1a10b97aea87894), and [201c5df9](https://github.com/input-output-hk/marlowe-cardano/commit/201c5df9de29b3b142be5ff7d0a9b56d1b378204) augment the Marlowe Cardano specification to require positive balances upon output and implements the corresponding check in the Marlowe Semantics validator.
+Commit [d514bcd0](https://github.com/input-output-hk/marlowe-cardano/commit/d514bcd075732b07e1c6a73e2e3c68afa01acf2c), [7f562545](https://github.com/input-output-hk/marlowe-cardano/commit/7f562545d013dd17472b692b550ffdc7f8a383f3), and [ebab31d9](https://github.com/input-output-hk/marlowe-cardano/commit/ebab31d9ba6dd322d374b34cad3995050a0e476e) adds a property-based test for this.
+
 ### 2.1.8 Non-validated Marlowe states *(Severity: High)*
@@
 > 
-> An example showing betrayed user expectations is discussed in .
+> An example showing betrayed user expectations is discussed in […].
 > 
@@
 
+Prompted on this finding in the audit report, the Marlowe semantics validator has been altered to enforce all invariants (positive accounts; non-duplication of accounts, choices, and bound values; a total internal account balance that exactly matches the value in the script's UTxO) on both the initial and final states.
+This prevents execution of a Marlowe contract that was inadvertently or purposefully created by off-chain code with an invalid initial state.
+It also prevents transactions that corrupt the final state; such corruption could only happen if there were a flaw in the Plutus validator or Marlowe semantics implementation, in which case that execution-path of the contract would be blocked.
+The [Marlowe Best Practices Guide](https://github.com/input-output-hk/marlowe-cardano/blob/main/marlowe/best-practices.md) and [Marlowe Security Guide](https://github.com/input-output-hk/marlowe-cardano/blob/main/marlowe/security.md) warn of the importance of off-chain code creating valid initial state for Marlowe contracts.
+
+Commits [0a890845](https://github.com/input-output-hk/marlowe-cardano/commit/0a890845c44ccfe4df97e765e9b9eb743ce7d580), [8855feae](https://github.com/input-output-hk/marlowe-cardano/commit/8855feae473882b82643db792327423152e45b5c), [26f024e8](https://github.com/input-output-hk/marlowe-cardano/commit/26f024e80cea0dd4ebf8bff0f1a10b97aea87894), and [201c5df9](https://github.com/input-output-hk/marlowe-cardano/commit/201c5df9de29b3b142be5ff7d0a9b56d1b378204) enforce that both the initial and final states in the respective incoming and outgoing datum obey the aforementioned three invariants of total value, non-duplication, and positivity.
+That commit also adds "Constraint 19.
+No duplicates" to the Marlowe Cardano specification.
+Property-based tests in commit [d514bcd0](https://github.com/input-output-hk/marlowe-cardano/commit/d514bcd075732b07e1c6a73e2e3c68afa01acf2c), [7f562545](https://github.com/input-output-hk/marlowe-cardano/commit/7f562545d013dd17472b692b550ffdc7f8a383f3), and [ebab31d9](https://github.com/input-output-hk/marlowe-cardano/commit/ebab31d9ba6dd322d374b34cad3995050a0e476e) check the implementation.
+
 ### 2.1.9 Total balance of ending state uncomputed *(Severity: High)*
@@
 
+The mitigation of the previous audit-report finding ("2.1.8 Non-validated Marlowe states") also mitigates this finding: in the validator, `checkOwnOutputConstraint marloweData finalBalance` ensures that the computed final balance of accounts matches the value output to the script's UTxO and `checkState "o" finalBalance txOutState` ensures that it matches the sum of value in the internal accounts.
+
+Again as previously, the relevant commits are [0a890845](https://github.com/input-output-hk/marlowe-cardano/commit/0a890845c44ccfe4df97e765e9b9eb743ce7d580), [8855feae](https://github.com/input-output-hk/marlowe-cardano/commit/8855feae473882b82643db792327423152e45b5c), [26f024e8](https://github.com/input-output-hk/marlowe-cardano/commit/26f024e80cea0dd4ebf8bff0f1a10b97aea87894), and [201c5df9](https://github.com/input-output-hk/marlowe-cardano/commit/201c5df9de29b3b142be5ff7d0a9b56d1b378204) for specification and implementation and [d514bcd0](https://github.com/input-output-hk/marlowe-cardano/commit/d514bcd075732b07e1c6a73e2e3c68afa01acf2c), [7f562545](https://github.com/input-output-hk/marlowe-cardano/commit/7f562545d013dd17472b692b550ffdc7f8a383f3), and [ebab31d9](https://github.com/input-output-hk/marlowe-cardano/commit/ebab31d9ba6dd322d374b34cad3995050a0e476e) for property-based tests.
+
 ### 2.1.10 Unchecked ending balance *(Severity: High)*
@@
 However, the validator does not check that the ending balance matches the value in the output paid to the Marlowe validator.
-Similarly to Issue [\[main:positive-balances-unchecked\]](#main:positive-balances-unchecked){reference-type="ref" reference="main:positive-balances-unchecked"}, if there are flaws in the semantics that cause the ending balance to differ from the actual value paid to the validator, this constraint would prevent any transaction from spending the output.
+Similarly to Issue […], if there are flaws in the semantics that cause the ending balance to differ from the actual value paid to the validator, this constraint would prevent any transaction from spending the output.
 > 
 > The specification should at least discuss why the check is absent together with the other similar checks that are not implemented (checking that ending accounts have positive balances, checking that the ending Marlowe state is valid).
+
+As the audit report recognizes, this concern is closely related to the three previous concerns.
+The revised validator's enforcement that of the three invariants for the final state ensures that the ending balance of the internal accounts in the state matches the actual output paid to the script.
+
+Again as previously, the relevant commits are [0a890845](https://github.com/input-output-hk/marlowe-cardano/commit/0a890845c44ccfe4df97e765e9b9eb743ce7d580), [8855feae](https://github.com/input-output-hk/marlowe-cardano/commit/8855feae473882b82643db792327423152e45b5c), [26f024e8](https://github.com/input-output-hk/marlowe-cardano/commit/26f024e80cea0dd4ebf8bff0f1a10b97aea87894), and [201c5df9](https://github.com/input-output-hk/marlowe-cardano/commit/201c5df9de29b3b142be5ff7d0a9b56d1b378204) for specification and implementation and [d514bcd0](https://github.com/input-output-hk/marlowe-cardano/commit/d514bcd075732b07e1c6a73e2e3c68afa01acf2c), [7f562545](https://github.com/input-output-hk/marlowe-cardano/commit/7f562545d013dd17472b692b550ffdc7f8a383f3), and [ebab31d9](https://github.com/input-output-hk/marlowe-cardano/commit/ebab31d9ba6dd322d374b34cad3995050a0e476e) for property-based tests.
 
@@
 
+The porting of `MList` from Isabelle to Plutus would have increased the size and execution cost of the Marlowe semantics validator beyond the present limits of the Cardano ledger.
+Work is underway to modify the Isabelle semantics and proofs to use a data structure similar to Plutus's `AssocMap`.
+Instead of modifying the Marlowe semantics validator, all usages of `AssocMap` were manually reviewed to verify that no vulnerability was introduced by its use (as compared to `MList`) *provided that the precondition of no duplicate entries in the `AssocMap` held*: the source code was annotated with informal reasoning documenting the safety of `AssocMap`'s use.
+That manual review was augmented by a comprehensive set of property-based tests to check the behavior of `MList` (under the precondition that it was ordered and contained no duplicates) against `AssocMap` (under the precondition that it contained not duplicates) for all functions used in Marlowe semantics or in the Marlowe validator.
+
+Commit [a2ff6aa3](https://github.com/input-output-hk/marlowe-cardano/commit/a2ff6aa334efb7b7955408d29cf0e2edf06bca08) annotates the Marlowe semantics validator with reasoning as to the safe use of `AssocMap`.
+Commits [af380a29](https://github.com/input-output-hk/marlowe-cardano/commit/af380a292c0369bcc77d8584d81ea80a31326b61), [e2277677](https://github.com/input-output-hk/marlowe-cardano/commit/e2277677987f359dcd8754275eccd5b7c9a880a7), [a95a616a](https://github.com/input-output-hk/marlowe-cardano/commit/a95a616aca59bc7159d3ffa92487f8f021820bff), [b65ccfec](https://github.com/input-output-hk/marlowe-cardano/commit/b65ccfec3708484ff52514671e0e227c13f084f3), and [9e068b6a](https://github.com/input-output-hk/marlowe-cardano/commit/9e068b6a60103f2a40604a7eaea57c5f29b59180) implement the property-based tests for the behavior of the Plutus `AssocMap` against the Isabelle `MList`.
+
 ### 2.1.13 Missing specification tests *(Severity: Medium)*
@@
 Besides checking that there are no translation mistakes, these properties would also help contrasting the assumptions in the Isabelle and the Haskell sides, like the meaning of validity of an association list, which is focused in the previous issue.
+
+Subsequent to the commencement of the audit, a "test oracle" was developed that checks implementations of Marlowe semantics (such as the Plutus validator) against a reference implementation generated directly from the Isabelle specification.
+That generated Haskell implementation is backed by Isabelle proofs of the correctness of Haskell code generation by Isabelle.
+The test oracle is now applied to the Marlowe semantics implementation in Plutus that the validator use.
+The test oracle uses the refference implementation to generates test cases to challenge the Plutus implementation and then it checks the Plutus results against its own.
+This oracle provides sufficient coverage to address the audit-reports concerns regarding the lack of property-based tests for Section 3 of the Marlowe specification.
 
@@
 > 
-> The proofs of the lemmas can be found in .
+> The proofs of the lemmas can be found in […].
 
@@
 
+Commit [00010ea4](https://github.com/input-output-hk/marlowe-cardano/commit/00010ea48439892ff9eec54fb93929f56bcb6c7b) adds a paragraph of guidance to the Marlowe Cardano specification.
+The mitigation above for "2.1.2 Contracts vulnerable to double satisfaction attacks" implements validator changes to address this.
+
 ### 2.5.2 No reference to creating a minting policy *(Severity: Low)*
@@
 
+The monetary policy is not a concern of the Marlowe validator, but commit [fe00af7d](https://github.com/input-output-hk/marlowe-cardano/commit/fe00af7d622bbd17c3327363d4650ee544d1fff7) adds a sentance referencing the discussion of monetary policies for role tokens in the [Marlowe Best Practices Guide](https://github.com/input-output-hk/marlowe-cardano/blob/main/marlowe/best-practices.md).
+
 ### 2.5.3 Argument for Contract in `txInfoData` not specified *(Severity: Low)*
@@
 > The argument by which the `Contract` in the `txInfoData` list corresponds to the given hash needs to be made explicit.
+
+Commit [b6366efc](https://github.com/input-output-hk/marlowe-cardano/commit/b6366efc9f8df38bfe9232f81e6caf2c7f11ffa8) makes this hash correspondence explicit in the Marlowe Cardano specification and adds a code snippet concretizing it, namely that the hash and continuation must be present in the script context datum map.
 
@@
 It needs to explain why the `Input` type needs to carry a hash and a contract, and why the evaluation of the contract is changed as described.
+
+Commit [a9d40432](https://github.com/input-output-hk/marlowe-cardano/commit/a9d4043230cb4f240fab2d816bed6a3cab4f5053) adds to the Marlowe Cardano specification a paragraph elaborating on mechanics and motivation for Merkleization, namely the ability to handle Marlowe contract that are too large to store in a datum.
 
@@
 
+Commit [fda1d3e8](https://github.com/input-output-hk/marlowe-cardano/commit/fda1d3e862fc8f8c31ff479fbd338226e661eb0b) to the Marlowe Cardano specification clarifies that the constraint is supplementary (informational) in nature.
+
 ### 2.5.6 Asymmetry between role and wallet payouts *(Severity: Low)*
@@
 
+The asymmetric is a convenience and practicality related to the manner in which coin selection and transaction balancing typically occurs nowadays in wallets.
+Commit [4e81470d](https://github.com/input-output-hk/marlowe-cardano/commit/4e81470de90ddd6858c3ad8aace96e90df5a86ae) adds to the Marlowe Cardano specification a paragraph justifying this asymmetry in payout style and a comment to the Marlowe semantics validator along the same lines.
+
 ### 2.5.7 Incorrect description of `rolePayoutValidator` *(Severity: Low)*
@@
 
+This incorrect signature is fixed in commit [1d16de1f](https://github.com/input-output-hk/marlowe-cardano/commit/1d16de1f1b7a167dc4eb4a057a3df2c24d0194c9).
+
 ### 2.5.8 Unspecified initial state *(Severity: Low)*
@@
 
+Commit [e150e308](https://github.com/input-output-hk/marlowe-cardano/commit/e150e30867dfa5466639f46bf39656aee2888e38) details the three invariants that the initial state must satisfy: positivity of accounts, non-duplication of state entries (accounts, choices, bound values), and matching the total value in the internal state to the value of the script's UTxO.
+
 ### 2.5.9 Unspecified behavior when multiple cases can apply *(Severity: Low)*
@@
 > If multiple cases in a case list can apply, the first one is taken. This behavior should be better communicated in the specification.
+
+Although this behavior is required by the Isabelle specification and implemented in the Plutus validator, the comment on `applyCases` has been editing in commit [08d6f34a](https://github.com/input-output-hk/marlowe-cardano/commit/08d6f34ac91da7aa34e7414c5bae0e33e9235d74) to reinforce that the first applicable `Case` is taken for a `When`.
 
@@
 
+The binding of `cont` has been renamed to `cont'` in commit [1c96edfe](https://github.com/input-output-hk/marlowe-cardano/commit/1c96edfe7e22ecc499ed75dd175490a4f2427bdc).
+
 ### 2.6.2 Non-isomorphic types in `playTraceAux` *(Severity: Medium)*
@@
 This means `TransactionError` cannot be an input to `playTraceAux` in Isabelle, possibly invalidating proofs about its properties.
+
+The `playTrace` and `playTraceAux` functions are not used by the Marlowe validators, so no mitigation was made for this audit-report finding.
 
@@
 
+These cosmetic recommendations of the audit report have not been implemented in the Marlowe semantics validator, as the discrepancies do not strongly impact comparison.
+
 ### 2.6.4 Naming of functions and variables *(Severity: Low)*
@@
 
+The aforementioned bindings `validateBalances`, `validateInputs`, and `tailCase` have been renamed to `allBalancesArePositive`, `allInputsAreAuthorized`, and `tailCases` in commit [1c96edfe](https://github.com/input-output-hk/marlowe-cardano/commit/1c96edfe7e22ecc499ed75dd175490a4f2427bdc).
+For historical reasons, `refundOne` has not been renamed.
+The function `checkScriptOutputAny` is no longer present due to other mitigations made to the validator.
+
 ### 2.6.5 Unused functions *(Severity: Low)*
@@
 
+Although the validator does not use these functions, off-chain code in other Haskell modules does.
+For the time being, the `contractLifespanUpperBound` and `isClose` have been retained in `Language.Marlowe.Core.V1.Semantics`.
+They will be relocated in the future if validator and non-validator code is separated into different modules.
+
 ### 2.6.6 Comments *(Severity: Low)*
@@
 
+Commit [1c96edfe](https://github.com/input-output-hk/marlowe-cardano/commit/1c96edfe7e22ecc499ed75dd175490a4f2427bdc) fixes this typo.
+
 ### 2.6.7 Record updates in `playTraceAux` *(Severity: Low)*
@@
 
+The `playTrace` and `playTraceAux` functions are not used by the Marlowe validators, so no mitigation was made for this audit-report finding.
+
 ### 2.6.8 Potential simplifications *(Severity: Low)*
@@
 
+Actually, the `PlutusTx.AssocMap.Map`'s implementation of `foldMap` does not permit this: type checking fails for this proposed change.
+
 > **File `Types.hs`, Class instance `Eq Contract`, line *873***
@@
 > The equality of cases for the `When` constructor would be simplified by using `cases1 == cases2`. If there is a reason for the more verbose equality condition, it should be outlined in a comment.
+
+A comment explaining the rationale for this equality test has been added in commit [faaae175](https://github.com/input-output-hk/marlowe-cardano/commit/faaae1751957b4f750e3c11ffdc946de07baf0cd).
 
@@
 The comparison would be simpler if both definitions were consolidated.
-> 
+
+The Plutus implementation differs slightly from the Isabelle for efficiency and to conform to Plutus's different restrictions on recursive calls.
+
 > **Helper function `evalValue`, lines *395 (Semantics.hs), 35 (Semantics.thy)***
@@
 > The Isabelle implementation should use the helper function `moneyInAccount` instead of inlining its definition, so as to maintain consistency with the Haskell implementation.
-> 
+
+TODO: @hrajchert will add text here.
+
 > **Helper function `applyCases`, lines *596 (Semantics.hs), 498 (Semantics.thy)***
@@
 This is likely due to the lack of Merkleization in the Isabelle implementation.
-> 
+
+Commit [d0a3834a](https://github.com/input-output-hk/marlowe-cardano/commit/d0a3834ae8484bb6c7d0ecd3aeec60871b3fd165) adds a comment that the Plutus implementation differs from the Isabelle one due to the former's support for merkleization.
+
 > **Helper function `convertReduceWarnings`, lines *617 (Semantics.hs), 537 (Semantics.thy)***
@@
 If there is a reason for this discrepancy, such as `foldr` yielding some optimizations, this should be outlined in a comment.
+
+Commit [d1477610](https://github.com/input-output-hk/marlowe-cardano/commit/d14776101c22eac7fbc67893020f37cbc2a9e339) adds a comment that the use of `foldr` in the Plutus implementation is for efficiency.
 
@@
 
+The discrepancies of implementation between the constraints in the specification and the Plutus code result from the emphasis on readability in the specification and of efficiency in the implementation.
+The need to minimize of validator size and execution cost in the implementation has sometimes resulted in less consise or readible code.
+The Plutus compile can be quite sensitive to the precise expression of a constraint as Haskell code.
+
 ### 2.6.11 Missing argument of `computeTransaction` *(Severity: Low)*
@@
 
+Commit [00a0c240](https://github.com/input-output-hk/marlowe-cardano/commit/00a0c2406665f06753aecf6dbb63933edb275436) moves the comment about the non-existant fifth argument to a separate sentance following the list of arguments.
+
 ### 2.6.12 Missing `smallMarloweValidator` *(Severity: Low)*
@@
 
+This reference is changed to `marloweValidator` in commit [87cf548d](https://github.com/input-output-hk/marlowe-cardano/commit/87cf548d9ae7934a702ea1a1c85d7d9bd4993aa9).
+
 ### 2.6.13 Incorrect constraint reference *(Severity: Low)*
@@
 
+This incorrect reference is fixed in commit [3fac0d17](https://github.com/input-output-hk/marlowe-cardano/commit/3fac0d1749d107f13743756aa50c9723e776e391).
+
 ### 2.6.14 `MarloweParams` differs from the specification *(Severity: Low)*
@@
 
+This mistatement is remedied in commit [8691f335](https://github.com/input-output-hk/marlowe-cardano/commit/8691f3351e55758cc750eee0800b1458f0a8c554).
+
 ### 2.6.15 Timeout boundary differs from the specification *(Severity: Low)*
@@
 > where "bigger" implies strict inequality, while the code makes non-strict comparison. This difference needs to be acknowledged and further explained in the specification.
+
+TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 > The absence of this information makes it easier to accidentally produce a test that is not testing what is intended.
+
+TODO: @bwbush will add text here.
 
@@
 > Some of these are tested in `Spec/Marlowe/Semantics/Functions.hs` already for auxiliary functions.
-> 
+
+Commit [ee2e5222](https://github.com/input-output-hk/marlowe-cardano/commit/ee2e52220e25736af29a1f01144d5fafbe9eed76) adds a property-based test that choice and deposit continue as expected; commit [333eee29](https://github.com/input-output-hk/marlowe-cardano/commit/333eee293aa3afe95d0aa7f678c7e7b7bad660ea) tests that choice produces expected continuation; commit [477284c7](https://github.com/input-output-hk/marlowe-cardano/commit/477284c78095482eba63adc2e6090ea68553db9d) tests that that deposits add to an account; commit [d241669b](https://github.com/input-output-hk/marlowe-cardano/commit/d241669bfb7ec8dad1cc92a7ca6185d08229817e) tests that payout substracts from an account; and commit [f33fc720](https://github.com/input-output-hk/marlowe-cardano/commit/f33fc7206522031c93253323f3b8ca083d8ca8c7) tests that Merkleization continues as expected.
+
 > **File `Spec/Marlowe/Semantics/Functions.hs`, `Missing merkleization tests`** 
@@
 > The properties in this module do not seem to be tested with merkleized contracts or inputs except for `checkGetContinuation`. More merkleization tests should be added.
-> 
+
+Commits [2a936bf1](https://github.com/input-output-hk/marlowe-cardano/commit/2a936bf1bbb890cc811b966b46a10fd38314dc25) and [8f8183d7](https://github.com/input-output-hk/marlowe-cardano/commit/8f8183d7f14c6bdc3841a1999077ef01ebac4144) significantly increase the coverage of merkleization in validator tests by utilizing a much larger set of test contracts and adding a probability that almost any test will involve a merkleized contract.
+
 > **File `Spec/Marlowe/Semantics/Compute.hs`, function `checkFixInterval`, lines *100-102***
@@
 > The test `checkFixInterval` is never instantiated with an invalid interval that is in the past, meaning the function `fixInterval` is never tested for that case.
+
+Commit [811f048b](https://github.com/input-output-hk/marlowe-cardano/commit/811f048ba93db10972641c0c235864db1d6014e7) adds a property-based test for an invalid in terval in the past.
+
+## Appendices
+
+### Post-Audit Changes in Marlowe Specification
+
+```bash
+git diff 523f3d56f22bf992ddb0b0c8a52bb7a5a188f9e9 69468d6623e24a4ccd6659e378ae012c38ca01ce marlowe/specification/marlowe-cardano-specification.md
+```
+
+```diff
+diff --git a/marlowe/specification/marlowe-cardano-specification.md b/marlowe/specification/marlowe-cardano-specification.md
+index 326f6f321..312ccf361 100644
+--- a/marlowe/specification/marlowe-cardano-specification.md
++++ b/marlowe/specification/marlowe-cardano-specification.md
+@@ -3,7 +3,7 @@
+
+ ## Scope
+
+-This document defines the specification for Marlowe semantics's interface with the Cardano blockchain.
+Marlowe utilizes three Plutus scripts: (i) a monetary policy for the roles currency used in some Marlowe contracts; (ii) the Marlowe application validator script that enforces Marlowe semantics; and (iii) the Marlowe payout-validator script that allows the holder of a role token to withdraw funds paid by the Marlowe application.
+Marlowe semantics are defined in the Isabelle language and specified in [Marlowe Specification, Version 3 (draft)](marlowe-isabelle-specification-4f9fa249fa51ec09a4f286099d5399eb4301ed49.pdf).
++This document defines the specification for Marlowe semantics's interface with the Cardano blockchain.
+Marlowe utilizes three Plutus scripts: (i) a monetary policy for the roles currency used in some Marlowe contracts; (ii) the Marlowe application validator script that enforces Marlowe semantics; and (iii) the Marlowe payout-validator script that allows the holder of a role token to withdraw funds paid by the Marlowe application.
+Marlowe semantics are defined in the Isabelle language and specified in [Marlowe Specification, Version 3](marlowe-isabelle-specification-4f9fa249fa51ec09a4f286099d5399eb4301ed49.pdf).
+
+ ## Contents
+@@ -30,9 +30,14 @@ Marlowe contracts identify each participant by either a *public-key hash (PKH)*
+ The execution of a Marlowe contract instance proceeds as a sequence of applications of inputs at the contract's script address.
+ 1. Role tokens are typically minted prior to or within the creation transaction of the contract instance, though this is not enforced on-chain.
+ 2. The creation transaction for the contract stores the state of the contract instance in the datum at the contract's script address.
+-3. Each transaction that interacts with the contract instance updates the state/datum at that same address.
+-4. If the contract instance pays funds to a role during application of inputs, those funds are sent to the address of the Marlowe payout-validator script, with a datum equal to the role name.
+-5. When a contract instance closes, there is no output to the contract's script address.
++    - The creation transaction must contain sufficient lovelace to meet the [minimum UTxO requirement](../best-practices.md#minimum-utxo-requirement) *for every future state of the contract.*
++    - The initial state conforms to the state invariants required by Isabelle semantics and the Marlowe validator:
++	    - positive account balances,
++	    - no duplicate accounts, choices, or bound values, and
++	    - a total value that matches the value of the UTxO for the creation transaction.
++1. Each transaction that interacts with the contract instance updates the state/datum at that same address.
++2. If the contract instance pays funds to a role during application of inputs, those funds are sent to the address of the Marlowe payout-validator script, with a datum equal to the role name.
++3. When a contract instance closes, there is no output to the contract's script address.
+
+ Thus each Marlowe contract instance is a finite sequence of continuations at the script address, from creation to closure.
+The following three UTxO diagrams illustrate the three typical types of Marlowe transactions.
+
+@@ -42,11 +47,15 @@ Thus each Marlowe contract instance is a finite sequence of continuations at the
+
+ ![A typical transaction to withdraw funds from the Marlowe payout validator script address](redeem.svg)
+
++Other Plutus scripts may run when the Marlowe semantics validator runs in a transaction *provided that the Marlowe transaction does not pay out funds.* Restricting other Plutus script to non-payout Marlowe transactions eliminates the possibility of double-satisfaction attacks.
+Examples of Plutus scripts that would be usefully run during Marlowe transactions are (1) an oracle which checks that the correct value is being reported as an `IChoice` or (2) a minting script that mints and deposits tokens into the Marlowe contract; in general such Plutus scripts would examine the Marlowe datum and redeemer to determine whether they validate the action that they are coordinating with Marlowe.
++
+
+ ## Monetary Policy for Role Tokens
+
+ Any Cardano monetary policy may be used to mint the role tokens used in a Marlowe contract instance.
+For security in standard Marlowe use cases, a one-time or locked minting policy such as `Plutus.Contracts.Currency.OneShotCurrency` is recommended.
+Exotic use cases might employ other monetary policies.
+It is the responsibility both of the developer of the off-chain code managing a contract instance and also of the user of the contract instance to verify that the monetary policy of the role tokens meets their security requirements.
+
++The off-chain [Marlowe Runtime](../marlowe-runtime/doc/) services use a safe one-shot monetary policy to mint role tokens if the user requests.
+The [Marlowe Best Practices Guide](../best-practices.md) discussed security requirements for role tokens.
++
+
+ ## Representation of Marlowe Semantics in Plutus
+
+@@ -97,10 +106,23 @@ data Input =
+ In the above, the `BuiltinByteString` is the hash of the serialized continuation `Contract`.
+
++### Variations from Isabelle Semantics
++
++Marlowe's semantics validator uses Plutus's association list `PlutuxTx.AssocMap` that differs from the `MList` in Isabelle semantics in several respects:
++
++1.  The `insert`, `delete`, `member`, and `lookup` functions of `AssocMap` haven an equality (`Eq`) constraint whereas `MList` has a ordering (`Ord`) constraint.
++2.  If the elements of an `MList` are not in lexicographic order, then result of `insert`, `delete`, `member`, and `lookup` may differ from that in `AssocMap`.
++3.  The ordering of the result of `toList` may differ between `AssocMap` and `MList`.
++
++Provided that the initial state of the Marlowe contract does not contain duplicate entries in `accounts`, `choices`, or `boundValues` and also provided that the behavior of Marlowe's semantics validator is compared to Isabelle semantics with the Isabelle initial state for `accounts`, `choices`, and `boundValues` lexicographic ordered, the behavior of the Marlowe semantics validator does not differ from Isabelle semantics aside from the ordering of `accounts`, `choices`, and `boundValues` in Marlowe's state in the datum.
+Typically, Marlowe contracts start with an empty state, so this is not an issue, but one can in principle start a Marlowe contract with a non-empty state, even with one that contains duplicate entries.
+Internally to the validator, `computeTransaction` might order payments upon `Close` differently from Isabelle semantics, but that is undetectable at the validation or transaction level.
++
++
+ ## Merkelization
+
+ A contract can be represented as a tree of continuations ("sub-contracts") where each vertex is a contract and each edge follows either an `InputContent` made by a participant or a timeout.
+A `When` contract includes terms that are either (1) a `Case` which contains the `Action` that matches a particular `InputContent` via `NormalInput` and that explicitly includes the `Contract` continuation or (2) a `MerkleizedCase` which contains the `Action` that matches a particular `InputContent` via `MerkleizedInput` and that implicitly includes the continuation by reference to its Merkle hash.
+In the latter case `MerkleizedInput` includes both the Merkle hash of the continuation and the continuation `Contract` itself.
+The Plutus code must verify that the hash in the contract matches the hash in the input before it proceeds to use the continuation that was provided as input.
+
++Merkleization enables large contracts to be succinctly represented on chain: instead of storing gigabytes of data representing a huge contract, only the hash of that contract need be represented on the chain.
+For example, a contract `When [Case action timeout continuation] contract'` that validates input `TransactionInput interval [NormalInput input]` may be "merkleized" to a contract `When [MerkleizedCase action hash] contract'` that will validate the input `TransactionInput interval [MerkleizedInput input hash contract]` provided that `DatumHash hash ≡ datumHash .
+Datum .
+toBuiltinData $ contract`.
+When a contract is merkelized the information `(hash, contract)` is stored off chain and provide to the Marlowe validator if needed for input.
+The Cardano node computes the hash and provides the `(hash, contract)` information in the script context.
++
+ The Isabelle semantics do not include merkleization of Marlowe contracts, but the Haskell implementation does.
+
+@@ -108,7 +130,7 @@ The Isabelle semantics do not include merkleization of Marlowe contracts, but th
+
+ The Marlowe validator is an unparameterized interpreter of Marlowe semantics. Thus, the script address of the Marlowe validator is independent of the particular contract instance and roles currency.
+ ```haskell
+-smallMarloweValidator :: MarloweData -> MarloweInput -> ScriptContext -> Bool
++marloweValidator :: MarloweData -> MarloweInput -> ScriptContext -> Bool
+ ```
+
+@@ -127,7 +149,7 @@ data MarloweData =
+ data MarloweParams =
+     MarloweParams
+     {
+-      rolePayoutValidatorHash :: ValidatorHash
++      rolesCurrency :: CurrencySymbol
+     }
+ ```
+
+@@ -139,14 +161,17 @@ data MarloweTxInput =
+     Input InputContent
+   | MerkleizedTxInput InputContent BuiltinByteString
+ ```
+-In the above, the `BuiltinByteString` is the hash of the serialized continuation of the contract.
+If `MerkleizedTxInput` is supplied in a redeemer, then the `ScriptContext` for the transaction must also contain an extra entry in its `txInfoData .
+scriptContextTxInfo` map from `DatumHash` to `Datum` for the serialized continuation of the contract.
++In the above, the `BuiltinByteString` is the hash of the serialized continuation of the contract.
+If `MerkleizedTxInput` is supplied in a redeemer, then the `ScriptContext` for the transaction must also contain an extra entry in its `txInfoData .
+scriptContextTxInfo` map from `DatumHash` to `Datum` for the serialized continuation of the contract.
+Thus the `continuation` for a given `hash` is computed as follows:
++```haskell
++Just continuation = fromBuiltinData =<< lookup hash (txInfoData $ scriptContextTxInfo scriptContext) :: Maybe Contract
++```
+
+ ### A Note about `Plutus.V1.Ledger.Value`
+
+ This specification relies on the following properties of `Plutus.V1.Ledger.Value` in the `plutus-ledger-api` package:
+ 1. `instance Monoid Value`, where `mempty` is zero value for all tokens and `mappend` sums the amounts of corresponding token types.
+-2.
+`leq` is a partial ordering requiring that quantity of each token in the first operand is less than or equal to quanity of the corresponding token in the second operand, where a missing token in one operand represents a zero quantity.
++2.
+`leq` is a partial ordering requiring that quantity of each token in the first operand is less than or equal to quantity of the corresponding token in the second operand, where a missing token in one operand represents a zero quantity.
+
+ ### Relationship between Marlowe Validator and Semantics
+@@ -156,9 +181,10 @@ The arguments of `computeTransaction` must be constructed as follows:
+ 2. The `txInputs` of `TransactionInput` is derived from the `MarloweInput` provided as the `Redeemer` and the `txInfoData . scriptContextTxInfo` of the `ScriptContext`, as detailed below.
+ 3. The `State` is the `marloweState` of the `MarloweData` provided as the `Datum`.
+ 4. The `Contract` is the `marloweContract` of the `MarloweData` provided as the `Datum`.
+-5. The new `Datum` at the script address is the `MarloweData` with the same `marloweParams` as originally, but with the new `txOutState` and `txOutContract` of the `TransactionOutput`.
+
+-In the diagram below, the upper three rectangles represent functions: the untyped `Validator`, the typed validator `smallMarloweValidator`, or the semantics's `computeTransaction`.
+The data values (ovals) are marshalled or passed from the untyped representation over to the semantics.
+The `computeTransaction` function validates the semantics and returns the required new state and contract instance.
+Conceptually, success or failure is passed to the untyped validator.
+The bottom rectangle packages the `Datum` that is provided as output for the continued progression of the Marlowe contract instance.
++Additionally, the new `Datum` at the script address is the `MarloweData` with the same `marloweParams` as originally, but with the new `txOutState` and `txOutContract` of the `TransactionOutput`.
++
++In the diagram below, the upper three rectangles represent functions: the untyped `Validator`, the typed validator `marloweValidator`, or the semantics's `computeTransaction`.
+The data values (ovals) are marshalled or passed from the untyped representation over to the semantics.
+The `computeTransaction` function validates the semantics and returns the required new state and contract instance.
+Conceptually, success or failure is passed to the untyped validator.
+The bottom rectangle packages the `Datum` that is provided as output for the continued progression of the Marlowe contract instance.
+
+ ![Relationship between Marlowe validator and semantics.](semantics2plutus.svg)
+
+@@ -167,7 +193,7 @@ In the diagram below, the upper three rectangles represent functions: the untype
+
+ Consider the application of the Marlowe validator and Marlowe semantics:
+ ```haskell
+-validationResult = smallMarloweValidator marloweData marloweInput scriptContext
++validationResult = marloweValidator marloweData marloweInput scriptContext
+
+ transactionOutput = computeTransaction transactionInput inState inContract
+ ```
+@@ -241,12 +267,14 @@ inValue ≡ foldMap toValue (accounts $ marloweState marloweData)
+
+ #### *Constraint 6.* Output value to script
+
+-The beginning balance plus the deposits equals the ending balance plus the payments.
++The beginning balance plus the deposits (with negative deposits treated as zero, as required by the semantics) equals the ending balance plus the payments.
+ ```haskell
+ inValue + foldMap valueOfDeposit (fmap getInputContent transactionInput) ≡ outValue + foldMap valueOfPayment (txOutPayments transactionOutput)
+   where getInputContent (Input inputContent) = inputContent
+         getInputContent (MerkleizedTxInput inputContent _) = inputContent
+-        valueOfDeposit (IDeposit _ _ (Token currency name) count) = Value.singleton currency name count
++        valueOfDeposit (IDeposit _ _ (Token currency name) count)
++          | count > 0 = Value.singleton currency name count
++          | otherwise = mempty
+         valueOfDeposit _ = mempty
+         valueOfPayment (Payment _ (Party _) value)) = value
+         valueOfPayment _ = mempty
+@@ -302,12 +330,16 @@ all demerkleizes transactionInput ≡ True
+         demerkleizes (Input _) = True
+ ```
+
++Note that this constraint is not strictly necessary and is included here for informational purposes, since a valid transaction could not be constructed if it were removed.
++
+
+ #### *Constraint 13.* Positive balances
+
+-All accounts in the state have positive balances.
++All accounts in the initial and final states have positive balances.
+ ```haskell
+-all ((> 0) . snd) (accounts $ marloweState marloweData) ≡ True
++positiveBalances $ marloweState marloweData     ≡ True
++positiveBalances $ txOutState transactionOutput ≡ True
++  where positiveBalances state = all ((> 0) . snd) (accounts state)
+ ```
+
+@@ -345,6 +377,38 @@ all sufficient (txOutPayments transactionOutput)
+
+ Note that the comparison is `leq` instead of equality because additional Ada may be required in the payment in order to satisfy the ledger's `minimum UTxO` constraint.
+
++The payment to a role must be in a single output, but the payment to an address may be split in multiple outputs, so long as the total output to the address is sufficient.
+(Note that allowing the multiple outputs to an address may increase transaction fees and wallet fragmentation, but the magnitude of this is severely limited by the Plutus execution budget, which strongly depends upon the number of UTxOs output.
+The flexibily of multiple outputs accommodates wallet-related practicalities (e.g., limitations in coin-selection and transaction balancing) such as the change and the return of the role token being in separate UTxOs in situations where a contract is also paying to the address where that change and that role token are sent.)
++
++
++#### *Constraint 18.* Final balance
++
++The value output to the script's UTxO must equal the total value in the output state.
++```haskell
++outValue ≡ foldMap toValue (accounts $ txOutState transactionOutput)
++  where toValue ((_, Token currency name), count)) = Value.singleton currency name count
++```
++
++
++#### *Constraint 19.* No duplicates
++
++The initial and final state must not have duplicate keys in their account entries, choices, or bound values.
++```haskell
++hasDuplicateKey (accounts    inState ) ≡ False
++hasDuplicateKey (choices     inState ) ≡ False
++hasDuplicateKey (boundValues inState ) ≡ False
++hasDuplicateKey (accounts    outState) ≡ False
++hasDuplicateKey (choices     outState) ≡ False
++hasDuplicateKey (boundValues outState) ≡ False
++  where
++    outState = txOutState transactionOutput
++    hasDuplicateKey am = let ks = fst <$> toList am in length ks /= length (nub ks)
++```
++
++
++#### *Constraint 20.* Single satisfaction
++
++If the contract makes payments, then the Marlowe validator must be the only validator executing in the transaction.
++
+
+ ## Plutus Validator for Marlowe Payouts
+
+@@ -359,7 +423,7 @@ rolePayoutValidator :: (CurrencySymbol, TokenName) -> () -> ScriptContext -> Boo
+
+ Consider the application of the Marlowe payout validator:
+ ```haskell
+-validationResult' = rolePayoutValidator rolesCurrency' role () scriptContext
++validationResult' = rolePayoutValidator (rolesCurrency', role) () scriptContext
+ ```
+
+ The validation fails (via returning `False` for `validationResult'` or via the throwing of an error) if any of the following constraints does not hold.
+```
+
+### Post-Audit Changes in Marlowe Validator
+
+```bash
+git diff 523f3d56f22bf992ddb0b0c8a52bb7a5a188f9e9 69468d6623e24a4ccd6659e378ae012c38ca01ce marlowe/src/Language/Marlowe/{Core/V1/Semantics*,Scripts.hs}
+```
+
+```diff
+diff --git a/marlowe/src/Language/Marlowe/Core/V1/Semantics.hs b/marlowe/src/Language/Marlowe/Core/V1/Semantics.hs
+index e31b8c478..82b57b903 100644
+--- a/marlowe/src/Language/Marlowe/Core/V1/Semantics.hs
++++ b/marlowe/src/Language/Marlowe/Core/V1/Semantics.hs
+@@ -94,12 +94,12 @@ module Language.Marlowe.Core.V1.Semantics
+   , TransactionError(..)
+   , TransactionWarning(..)
+     -- * Utility Functions
++  , allBalancesArePositive
+   , contractLifespanUpperBound
+   , isClose
+   , notClose
+   , paymentMoney
+   , totalBalance
+-  , validateBalances
+     -- * Serialisation
+   , currencySymbolFromJSON
+   , currencySymbolToJSON
+@@ -217,6 +217,26 @@ import Text.PrettyPrint.Leijen (comma, hang, lbrace, line, rbrace, space, text,
+ data Payment = Payment AccountId Payee Token Integer
+   deriving stock (Haskell.Eq, Haskell.Show)
+
++instance ToJSON Payment where
++  toJSON (Payment accountId payee token amount) =
++    object
++      [
++        "payment_from" .= accountId
++      , "to" .= payee
++      , "token" .= token
++      , "amount" .= amount
++      ]
++
++instance FromJSON Payment where
++  parseJSON =
++    withObject "Payment"
++      $ \o ->
++        Payment
++          <$> o .: "payment_from"
++          <*> o .: "to"
++          <*> o .: "token"
++          <*> o .: "amount"
++
+
+ -- | Extract the money value from a payment.
+ paymentMoney :: Payment -> Money
+@@ -293,7 +313,28 @@ data TransactionError = TEAmbiguousTimeIntervalError
+                       | TEUselessTransaction
+                       | TEHashMismatch
+   deriving stock (Haskell.Show, Generic, Haskell.Eq)
+-  deriving anyclass (ToJSON, FromJSON)
++
++instance ToJSON TransactionError where
++  toJSON TEAmbiguousTimeIntervalError = JSON.String "TEAmbiguousTimeIntervalError"
++  toJSON TEApplyNoMatchError = JSON.String "TEApplyNoMatchError"
++  toJSON (TEIntervalError intervalError) = object ["error" .= JSON.String "TEIntervalError", "context" .= intervalError]
++  toJSON TEUselessTransaction = JSON.String "TEUselessTransaction"
++  toJSON TEHashMismatch = JSON.String "TEHashMismatch"
++
++instance FromJSON TransactionError where
++  parseJSON (JSON.String s) =
++    case s of
++      "TEAmbiguousTimeIntervalError" -> return TEAmbiguousTimeIntervalError
++      "TEApplyNoMatchError" -> return TEApplyNoMatchError
++      "TEUselessTransaction" -> return TEUselessTransaction
++      "TEHashMismatch" -> return TEHashMismatch
++      _ -> Haskell.fail "Failed parsing TransactionError"
++  parseJSON (JSON.Object o) = do
++                                err <- o .: "error"
++                                if err Haskell.== ("TEIntervalError" :: Haskell.String)
++                                  then TEIntervalError <$> o .: "context"
++                                  else Haskell.fail "Failed parsing TransactionError"
++  parseJSON _ = Haskell.fail "Failed parsing TransactionError"
+
+ -- | Marlowe transaction input.
+@@ -319,6 +360,31 @@ data TransactionOutput =
+     | Error TransactionError
+   deriving stock (Haskell.Show)
+
++instance ToJSON TransactionOutput where
++  toJSON TransactionOutput{..} =
++    object
++      [
++        "warnings" .= txOutWarnings
++      , "payments" .= txOutPayments
++      , "state" .= txOutState
++      , "contract" .= txOutContract
++      ]
++  toJSON (Error err) = object ["transaction_error" .= err]
++
++instance FromJSON TransactionOutput where
++  parseJSON =
++    withObject "TransactionOutput"
++      $ \o ->
++        let
++          asTransactionOutput =
++            TransactionOutput
++              <$> o .: "warnings"
++              <*> o .: "payments"
++              <*> o .: "state"
++              <*> o .: "contract"
++          asError = Error <$> o .: "transaction_error"
++        in
++          asTransactionOutput <|> asError
+
+ -- | Parse a validator hash from JSON.
+ validatorHashFromJSON :: JSON.Value -> Parser ValidatorHash
+@@ -404,12 +470,22 @@ evalValue env state value = let
+                                    then 0
+                                    else n `Builtins.quotientInteger` d
+         ChoiceValue choiceId ->
++            -- SCP-5126: Given the precondition that `choices` contains no
++            -- duplicate entries, this lookup behaves identically to
++            -- Marlowe's Isabelle semantics given the precondition that
++            -- the initial state's `choices` in Isabelle was sorted and
++            -- did not contain duplicate entries.
+             case Map.lookup choiceId (choices state) of
+                 Just x  -> x
+                 Nothing -> 0
+         TimeIntervalStart    -> getPOSIXTime (fst (timeInterval env))
+         TimeIntervalEnd      -> getPOSIXTime (snd (timeInterval env))
+         UseValue valId       ->
++            -- SCP-5126: Given the precondition that `boundValues` contains
++            -- no duplicate entries, this lookup behaves identically to
++            -- Marlowe's Isabelle semantics given the precondition that
++            -- the initial state's `boundValues` in Isabelle was sorted
++            -- and did not contain duplicate entries.
+             case Map.lookup valId (boundValues state) of
+                 Just x  -> x
+                 Nothing -> 0
+@@ -425,6 +501,11 @@ evalObservation env state obs = let
+         AndObs lhs rhs          -> evalObs lhs && evalObs rhs
+         OrObs lhs rhs           -> evalObs lhs || evalObs rhs
+         NotObs subObs           -> not (evalObs subObs)
++                                   -- SCP-5126: Given the precondition that `choices` contains no
++                                   -- duplicate entries, this membership test behaves identically
++                                   -- to Marlowe's Isabelle semantics given the precondition that
++                                   -- the initial state's `choices` in Isabelle was sorted and did
++                                   -- not contain duplicate entries.
+         ChoseSomething choiceId -> choiceId `Map.member` choices state
+         ValueGE lhs rhs         -> evalVal lhs >= evalVal rhs
+         ValueGT lhs rhs         -> evalVal lhs > evalVal rhs
+@@ -439,6 +520,12 @@ evalObservation env state obs = let
+ refundOne :: Accounts -> Maybe ((Party, Token, Integer), Accounts)
+ refundOne accounts = case Map.toList accounts of
+     [] -> Nothing
++    -- SCP-5126: The return value of this function differs from
++    -- Isabelle semantics in that it returns the least-recently
++    -- added account-token combination rather than the first
++    -- lexicographically ordered one. Also, the sequence
++    -- `Map.fromList . tail . Map.toList` preserves the
++    -- invariants of order and non-duplication.
+     ((accId, token), balance) : rest ->
+         if balance > 0
+         then Just ((accId, token, balance), Map.fromList rest)
+@@ -447,18 +534,30 @@ refundOne accounts = case Map.toList accounts of
+
+ -- | Obtains the amount of money available an account.
+ moneyInAccount :: AccountId -> Token -> Accounts -> Integer
+-moneyInAccount accId token accounts = case Map.lookup (accId, token) accounts of
+-    Just x  -> x
+-    Nothing -> 0
++moneyInAccount accId token accounts =
++    -- SCP-5126: Given the precondition that `accounts` contains
++    -- no duplicate entries, this lookup behaves identically to
++    -- Marlowe's Isabelle semantics given the precondition that
++    -- the initial state's `accounts` in Isabelle was sorted and
++    -- did not contain duplicate entries.
++    case Map.lookup (accId, token) accounts of
++      Just x  -> x
++      Nothing -> 0
+
+ -- | Sets the amount of money available in an account.
+ updateMoneyInAccount :: AccountId -> Token -> Integer -> Accounts -> Accounts
+ updateMoneyInAccount accId token amount =
++    -- SCP-5126: Given the precondition that `accounts` contains
++    -- no duplicate entries, this deletion or insertion behaves
++    -- identically (aside from internal ordering) to Marlowe's
++    -- Isabelle semantics given the precondition that the initial
++    -- state's `accounts` in Isabelle was sorted and did not
++    -- contain duplicate entries.
+     if amount <= 0 then Map.delete (accId, token) else Map.insert (accId, token) amount
+
+--- | Add the given amount of money to an accoun (only if it is positive).
++-- | Add the given amount of money to an account (only if it is positive).
+ --   Return the updated Map.
+ addMoneyToAccount :: AccountId -> Token -> Integer -> Accounts -> Accounts
+ addMoneyToAccount accId token amount accounts = let
+@@ -482,6 +581,14 @@ giveMoney accountId payee token amount accounts = let
+ reduceContractStep :: Environment -> State -> Contract -> ReduceStepResult
+ reduceContractStep env state contract = case contract of
+
++    -- SCP-5126: Although `refundOne` refunds accounts-token combinations
++    -- in least-recently-added order and Isabelle semantics requires that
++    -- they be refunded in lexicographic order, `reduceContractUntilQuiescent`
++    -- ensures that the `Close` pattern will be executed until `accounts`
++    -- is empty. Thus, the net difference between the behavior here and the
++    -- Isabelle semantics is that the `ContractQuiescent` resulting from
++    -- `reduceContractUntilQuiescent` will contain payments in a different
++    -- order.
+     Close -> case refundOne (accounts state) of
+         Just ((party, token, amount), newAccounts) -> let
+             newState = state { accounts = newAccounts }
+@@ -522,7 +629,17 @@ reduceContractStep env state contract = case contract of
+     Let valId val cont -> let
+         evaluatedValue = evalValue env state val
+         boundVals = boundValues state
++        -- SCP-5126: Given the precondition that `boundValues` contains
++        -- no duplicate entries, this insertion behaves identically
++        -- (aside from internal ordering) to Marlowe's Isabelle semantics
++        -- given the precondition that the initial state's `boundValues`
++        -- in Isabelle was sorted and did not contain duplicate entries.
+         newState = state { boundValues = Map.insert valId evaluatedValue boundVals }
++        -- SCP-5126: Given the precondition that `boundValues` contains
++        -- no duplicate entries, this lookup behaves identically to
++        -- Marlowe's Isabelle semantics given the precondition that the
++        -- initial state's `boundValues` in Isabelle was sorted and did
++        -- not contain duplicate entries.
+         warn = case Map.lookup valId boundVals of
+               Just oldVal -> ReduceShadowing valId oldVal evaluatedValue
+               Nothing     -> ReduceNoWarning
+@@ -575,6 +692,11 @@ applyAction env state (IDeposit accId1 party1 tok1 amount) (Deposit accId2 party
+     else NotAppliedAction
+ applyAction _ state (IChoice choId1 choice) (Choice choId2 bounds) =
+     if choId1 == choId2 && inBounds choice bounds
++    -- SCP-5126: Given the precondition that `choices` contains no
++    -- duplicate entries, this insertion behaves identically (aside
++    -- from internal ordering) to Marlowe's Isabelle semantics
++    -- given the precondition that the initial state's `choices`
++    -- in Isabelle was sorted and did not contain duplicate entries.
+     then let newState = state { choices = Map.insert choId1 choice (choices state) }
+          in AppliedAction ApplyNoWarning newState
+     else NotAppliedAction
+@@ -592,18 +714,20 @@ getContinuation (MerkleizedInput _ inputContinuationHash continuation) (Merkleiz
+     else Nothing
+ getContinuation _ _ = Nothing
+
+--- | Try to apply an input to a list of cases.
++-- | Try to apply an input to a list of cases, accepting the first match.
+ applyCases :: Environment -> State -> Input -> [Case Contract] -> ApplyResult
+-applyCases env state input (headCase : tailCase) =
++applyCases env state input (headCase : tailCases) =
+     let inputContent = getInputContent input :: InputContent
+         action = getAction headCase :: Action
+         maybeContinuation = getContinuation input headCase :: Maybe Contract
+     in case applyAction env state inputContent action of
+          AppliedAction warning newState ->
++           -- Note that this differs from Isabelle semantics because
++           -- the Cardano semantics includes merkleization.
+            case maybeContinuation of
+              Just continuation -> Applied warning newState continuation
+              Nothing           -> ApplyHashMismatch
+-         NotAppliedAction -> applyCases env state input tailCase
++         NotAppliedAction -> applyCases env state input tailCases
+ applyCases _ _ _ [] = ApplyNoMatchError
+
+@@ -615,7 +739,7 @@ applyInput _ _ _ _                          = ApplyNoMatchError
+
+ -- | Propagate 'ReduceWarning' to 'TransactionWarning'.
+ convertReduceWarnings :: [ReduceWarning] -> [TransactionWarning]
+-convertReduceWarnings = foldr (\warn acc -> case warn of
++convertReduceWarnings = foldr (\warn acc -> case warn of  -- Note that `foldr` is used here for efficiency, differing from Isabelle.
+     ReduceNoWarning -> acc
+     ReduceNonPositivePay accId payee tok amount ->
+         TransactionNonPositivePay accId payee tok amount : acc
+@@ -655,12 +779,12 @@ applyAllInputs env state contract inputs = let
+                     curState
+                     cont
+                 (input : rest) -> case applyInput env curState input cont of
+-                    Applied applyWarn newState cont ->
++                    Applied applyWarn newState cont' ->
+                         applyAllLoop
+                             True
+                             env
+                             newState
+-                            cont
++                            cont'
+                             rest
+                             (warnings' ++ convertApplyWarning applyWarn)
+                             payments'
+@@ -759,8 +883,8 @@ totalBalance accounts = foldMap
+
+ -- | Check that all accounts have positive balance.
+-validateBalances :: State -> Bool
+-validateBalances State{..} = all (\(_, balance) -> balance > 0) (Map.toList accounts)
++allBalancesArePositive :: State -> Bool
++allBalancesArePositive State{..} = all (\(_, balance) -> balance > 0) (Map.toList accounts)
+
+ instance FromJSON TransactionInput where
+@@ -845,20 +969,26 @@ instance Eq Payment where
+ instance Eq ReduceWarning where
+     {-# INLINABLE (==) #-}
+     ReduceNoWarning == ReduceNoWarning = True
+-    (ReduceNonPositivePay acc1 p1 tn1 a1) == (ReduceNonPositivePay acc2 p2 tn2 a2) =
++    ReduceNoWarning == _ = False
++    ReduceNonPositivePay acc1 p1 tn1 a1 == ReduceNonPositivePay acc2 p2 tn2 a2 =
+         acc1 == acc2 && p1 == p2 && tn1 == tn2 && a1 == a2
+-    (ReducePartialPay acc1 p1 tn1 a1 e1) == (ReducePartialPay acc2 p2 tn2 a2 e2) =
++    ReduceNonPositivePay{} == _ = False
++    ReducePartialPay acc1 p1 tn1 a1 e1 == ReducePartialPay acc2 p2 tn2 a2 e2 =
+         acc1 == acc2 && p1 == p2 && tn1 == tn2 && a1 == a2 && e1 == e2
+-    (ReduceShadowing v1 old1 new1) == (ReduceShadowing v2 old2 new2) =
++    ReducePartialPay{} == _ = False
++    ReduceShadowing v1 old1 new1 == ReduceShadowing v2 old2 new2 =
+         v1 == v2 && old1 == old2 && new1 == new2
+-    _ == _ = False
++    ReduceShadowing{} == _ = False
++    ReduceAssertionFailed == ReduceAssertionFailed = True
++    ReduceAssertionFailed == _ = False
+
+ instance Eq ReduceEffect where
+     {-# INLINABLE (==) #-}
+     ReduceNoPayment == ReduceNoPayment           = True
++    ReduceNoPayment == _                         = False
+     ReduceWithPayment p1 == ReduceWithPayment p2 = p1 == p2
+-    _ == _                                       = False
++    ReduceWithPayment _ == _                     = False
+
+ -- Lifting data types to Plutus Core
+diff --git a/marlowe/src/Language/Marlowe/Core/V1/Semantics/Types.hs b/marlowe/src/Language/Marlowe/Core/V1/Semantics/Types.hs
+index a333b8fbb..bf495773b 100644
+--- a/marlowe/src/Language/Marlowe/Core/V1/Semantics/Types.hs
++++ b/marlowe/src/Language/Marlowe/Core/V1/Semantics/Types.hs
+@@ -108,7 +108,7 @@ import PlutusTx.Lift (makeLift)
+ import PlutusTx.Prelude hiding (encodeUtf8, mapM, (<$>), (<*>), (<>))
+ import Prelude (mapM, (<$>))
+ import qualified Prelude as Haskell
+-import Text.PrettyPrint.Leijen (text)
++import Text.PrettyPrint.Leijen (parens, text)
+
+ -- Functions that used in Plutus Core must be inlinable,
+@@ -126,8 +126,8 @@ data Party =
+   deriving stock (Generic,Haskell.Eq,Haskell.Ord)
+
+ instance Pretty Party where
+-  prettyFragment (Address network address) = text $ "Address " ++ Haskell.show (serialiseAddressBech32 network address)
+-  prettyFragment (Role role)               = text $ "Role "    ++ Haskell.show role
++  prettyFragment (Address network address) = parens $ text "Address " Haskell.<> prettyFragment (serialiseAddressBech32 network address)
++  prettyFragment (Role role)               = parens $ text "Role "    Haskell.<> prettyFragment role
+
+ instance Haskell.Show Party where
+   showsPrec _ (Address network address) = Haskell.showsPrec 11 $ Haskell.show (serialiseAddressBech32 network address)
+@@ -307,6 +307,13 @@ data State = State { accounts    :: Accounts
+ newtype Environment = Environment { timeInterval :: TimeInterval }
+   deriving stock (Haskell.Show,Haskell.Eq,Haskell.Ord)
+
++instance FromJSON Environment where
++  parseJSON = withObject "Environment"
++    (\v -> Environment <$> (posixIntervalFromJSON =<< v .: "timeInterval"))
++
++instance ToJSON Environment where
++  toJSON Environment{..} = object
++    [ "timeInterval" .= posixIntervalToJSON timeInterval]
+
+ -- | Input for a Marlowe contract. Correspond to expected 'Action's.
+ data InputContent = IDeposit AccountId Party Token Integer
+@@ -389,23 +396,24 @@ data IntervalError = InvalidInterval TimeInterval
+                    | IntervalInPastError POSIXTime TimeInterval
+   deriving stock (Haskell.Show, Generic, Haskell.Eq)
+
++
+ instance ToJSON IntervalError where
+   toJSON (InvalidInterval (s, e)) = A.object
+-    [ ("invalidInterval", toJSON (posixTimeToJSON s, posixTimeToJSON e)) ]
++    [ ("invalidInterval", A.object [("from", posixTimeToJSON s), ("to", posixTimeToJSON e)]) ]
+   toJSON (IntervalInPastError t (s, e)) = A.object
+-    [ ("intervalInPastError", toJSON (posixTimeToJSON t, posixTimeToJSON s, posixTimeToJSON e)) ]
++    [ ("intervalInPastError", A.object [("minTime", posixTimeToJSON t), ("from", posixTimeToJSON s), ("to", posixTimeToJSON e)]) ]
+
+ instance FromJSON IntervalError where
+   parseJSON (JSON.Object v) =
+     let
+       parseInvalidInterval = do
+-        (s, e) <- v .: "invalidInterval"
+-        InvalidInterval <$> ((,) <$> posixTimeFromJSON s <*> posixTimeFromJSON e)
++        o <- v .: "invalidInterval"
++        InvalidInterval <$> posixIntervalFromJSON o
+       parseIntervalInPastError = do
+-        (t, s, e) <- v .: "intervalInPastError"
++        o <- v .: "intervalInPastError"
+         IntervalInPastError
+-          <$> posixTimeFromJSON t
+-          <*> ((,) <$> posixTimeFromJSON s <*> posixTimeFromJSON e)
++          <$> (posixTimeFromJSON =<< o .: "minTime")
++          <*> posixIntervalFromJSON o
+     in
+       parseIntervalInPastError <|> parseInvalidInterval
+   parseJSON invalid =
+@@ -422,11 +430,18 @@ posixTimeFromJSON = \case
+   invalid ->
+       JSON.prependFailure "parsing POSIXTime failed, " (JSON.typeMismatch "Number" invalid)
+
++posixIntervalFromJSON :: A.Object -> Parser TimeInterval
++posixIntervalFromJSON o = (,) <$> (posixTimeFromJSON =<< o .: "from") <*> (posixTimeFromJSON =<< o .: "to")
+
+ -- | Serialise time as a JSON value.
+ posixTimeToJSON :: POSIXTime -> JSON.Value
+ posixTimeToJSON (POSIXTime n) = JSON.Number $ scientific n 0
+
++posixIntervalToJSON :: TimeInterval -> JSON.Value
++posixIntervalToJSON (from, to) = object
++  [ "from" .= posixTimeToJSON from
++  , "to" .= posixTimeToJSON to
++  ]
+
+ -- | Result of 'fixInterval'
+ data IntervalResult = IntervalTrimmed Environment State
+@@ -778,23 +793,24 @@ instance ToJSON Contract where
+ instance Eq Party where
+     {-# INLINABLE (==) #-}
+     Address n1 a1 == Address n2 a2 = n1 == n2 && a1 == a2
++    Address _ _   == _             = False
+     Role r1       == Role r2       = r1 == r2
+-    _             == _             = False
++    Role _        == _             = False
+
+ instance Eq ChoiceId where
+     {-# INLINABLE (==) #-}
+-    (ChoiceId n1 p1) == (ChoiceId n2 p2) = n1 == n2 && p1 == p2
++    ChoiceId n1 p1 == ChoiceId n2 p2 = n1 == n2 && p1 == p2
+
+ instance Eq Token where
+     {-# INLINABLE (==) #-}
+-    (Token n1 p1) == (Token n2 p2) = n1 == n2 && p1 == p2
++    Token n1 p1 == Token n2 p2 = n1 == n2 && p1 == p2
+
+ instance Eq ValueId where
+     {-# INLINABLE (==) #-}
+-    (ValueId n1) == (ValueId n2) = n1 == n2
++    ValueId n1 == ValueId n2 = n1 == n2
+
+ instance Pretty ValueId where
+@@ -804,78 +820,107 @@ instance Pretty ValueId where
+ instance Eq Payee where
+     {-# INLINABLE (==) #-}
+     Account acc1 == Account acc2 = acc1 == acc2
++    Account{}    == _            = False
+     Party p1 == Party p2         = p1 == p2
+-    _ == _                       = False
++    Party{}  == _                = False
+
+ instance Eq a => Eq (Value a) where
+     {-# INLINABLE (==) #-}
+-    AvailableMoney acc1 tok1 == AvailableMoney acc2 tok2 =
+-        acc1 == acc2 && tok1 == tok2
++    AvailableMoney acc1 tok1 == AvailableMoney acc2 tok2 = acc1 == acc2 && tok1 == tok2
++    AvailableMoney _    _    == _                        = False
+     Constant i1 == Constant i2 = i1 == i2
++    Constant{}  == _           = False
+     NegValue val1 == NegValue val2 = val1 == val2
++    NegValue{}    == _             = False
+     AddValue val1 val2 == AddValue val3 val4 = val1 == val3 && val2 == val4
++    AddValue{}         == _                  = False
+     SubValue val1 val2 == SubValue val3 val4 = val1 == val3 && val2 == val4
++    SubValue{}         == _                  = False
+     MulValue val1 val2 == MulValue val3 val4 = val1 == val3 && val2 == val4
++    MulValue{}         == _                  = False
+     DivValue val1 val2 == DivValue val3 val4 = val1 == val3 && val2 == val4
++    DivValue{}         == _                  = False
+     ChoiceValue cid1 == ChoiceValue cid2 = cid1 == cid2
++    ChoiceValue{}    == _                = False
+     TimeIntervalStart == TimeIntervalStart = True
+-    TimeIntervalEnd   == TimeIntervalEnd   = True
++    TimeIntervalStart == _                 = False
++    TimeIntervalEnd == TimeIntervalEnd = True
++    TimeIntervalEnd == _               = False
+     UseValue val1 == UseValue val2 = val1 == val2
+-    Cond obs1 thn1 els1 == Cond obs2 thn2 els2 =  obs1 == obs2 && thn1 == thn2 && els1 == els2
+-    _ == _ = False
++    UseValue{}    == _             = False
++    Cond obs1 thn1 els1 == Cond obs2 thn2 els2 = obs1 == obs2 && thn1 == thn2 && els1 == els2
++    Cond{}              == _                   = False
+
+ instance Eq Observation where
+     {-# INLINABLE (==) #-}
+     AndObs o1l o2l == AndObs o1r o2r           = o1l == o1r && o2l == o2r
++    AndObs{}       == _                        = False
+     OrObs  o1l o2l == OrObs  o1r o2r           = o1l == o1r && o2l == o2r
++    OrObs{}        == _                        = False
+     NotObs ol == NotObs or                     = ol == or
++    NotObs{}  == _                             = False
+     ChoseSomething cid1 == ChoseSomething cid2 = cid1 == cid2
++    ChoseSomething _    == _                   = False
+     ValueGE v1l v2l == ValueGE v1r v2r         = v1l == v1r && v2l == v2r
++    ValueGE{}       == _                       = False
+     ValueGT v1l v2l == ValueGT v1r v2r         = v1l == v1r && v2l == v2r
++    ValueGT{}       == _                       = False
+     ValueLT v1l v2l == ValueLT v1r v2r         = v1l == v1r && v2l == v2r
++    ValueLT{}       == _                       = False
+     ValueLE v1l v2l == ValueLE v1r v2r         = v1l == v1r && v2l == v2r
++    ValueLE{}       == _                       = False
+     ValueEQ v1l v2l == ValueEQ v1r v2r         = v1l == v1r && v2l == v2r
++    ValueEQ{}       == _                       = False
+     TrueObs  == TrueObs                        = True
++    TrueObs  == _                              = False
+     FalseObs == FalseObs                       = True
+-    _ == _                                     = False
++    FalseObs == _                              = False
+
+ instance Eq Action where
+     {-# INLINABLE (==) #-}
+     Deposit acc1 party1 tok1 val1 == Deposit acc2 party2 tok2 val2 =
+         acc1 == acc2 && party1 == party2 && tok1 == tok2 && val1 == val2
++    Deposit{}       == _ = False
+     Choice cid1 bounds1 == Choice cid2 bounds2 =
+         cid1 == cid2 && length bounds1 == length bounds2 && let
+             bounds = zip bounds1 bounds2
+             checkBound (Bound low1 high1, Bound low2 high2) = low1 == low2 && high1 == high2
+             in all checkBound bounds
++    Choice{}   == _ = False
+     Notify obs1 == Notify obs2 = obs1 == obs2
+-    _ == _ = False
++    Notify{} == _ = False
+
+ instance Eq a => Eq (Case a) where
+     {-# INLINABLE (==) #-}
+     Case acl cl == Case acr cr                       = acl == acr && cl == cr
++    Case{}      == _                                 = False
+     MerkleizedCase acl bsl == MerkleizedCase acr bsr = acl == acr && bsl == bsr
+-    _ == _                                           = False
++    MerkleizedCase{}       == _                      = False
+
+ instance Eq Contract where
+     {-# INLINABLE (==) #-}
+     Close == Close = True
++    Close == _ = False
+     Pay acc1 payee1 tok1 value1 cont1 == Pay acc2 payee2 tok2 value2 cont2 =
+         acc1 == acc2 && payee1 == payee2 && tok1 == tok2 && value1 == value2 && cont1 == cont2
++    Pay{} == _ = False
+     If obs1 cont1 cont2 == If obs2 cont3 cont4 =
+         obs1 == obs2 && cont1 == cont3 && cont2 == cont4
+-    When cases1 timeout1 cont1 == When cases2 timeout2 cont2 =
++    If{} == _ = False
++    When cases1 timeout1 cont1 == When cases2 timeout2 cont2 =  -- The sequences of tests are ordered for efficiency.
+         timeout1 == timeout2 && cont1 == cont2
+         && length cases1 == length cases2
+         && and (zipWith (==) cases1 cases2)
++    When{} == _ = False
+     Let valId1 val1 cont1 == Let valId2 val2 cont2 =
+         valId1 == valId2 && val1 == val2 && cont1 == cont2
++    Let{} == _ = False
+     Assert obs1 cont1 == Assert obs2 cont2 = obs1 == obs2 && cont1 == cont2
+-    _ == _ = False
++    Assert{}  == _ = False
+
+ instance Eq State where
+diff --git a/marlowe/src/Language/Marlowe/Scripts.hs b/marlowe/src/Language/Marlowe/Scripts.hs
+index 6027202b5..aa397d5fa 100644
+--- a/marlowe/src/Language/Marlowe/Scripts.hs
++++ b/marlowe/src/Language/Marlowe/Scripts.hs
+@@ -47,7 +47,9 @@ module Language.Marlowe.Scripts
+   , alternateMarloweValidator
+   , alternateMarloweValidatorHash
+   , marloweValidator
++  , marloweValidatorCompiled
+   , marloweValidatorHash
++  , mkMarloweValidator
+     -- * Payout Validator
+   , TypedRolePayoutValidator
+   , rolePayoutValidator
+@@ -61,8 +63,9 @@ import GHC.Generics (Generic)
+ import Language.Marlowe.Core.V1.Semantics as Semantics
+ import Language.Marlowe.Core.V1.Semantics.Types as Semantics
+ import Language.Marlowe.Pretty (Pretty(..))
++import Ledger.Typed.Scripts (unsafeMkTypedValidator)
+ import qualified Plutus.Script.Utils.Typed as Scripts
+-import Plutus.Script.Utils.V2.Typed.Scripts (mkTypedValidator, mkUntypedValidator)
++import Plutus.Script.Utils.V2.Typed.Scripts (mkTypedValidator)
+ import qualified Plutus.Script.Utils.V2.Typed.Scripts as Scripts
+ import qualified Plutus.V1.Ledger.Address as Address (scriptHashAddress)
+ import qualified Plutus.V1.Ledger.Value as Val
+@@ -92,11 +95,22 @@ import PlutusTx (makeIsDataIndexed, makeLift)
+ import qualified PlutusTx
+ import qualified PlutusTx.AssocMap as AssocMap
+ import PlutusTx.Plugin ()
+-import PlutusTx.Prelude as PlutusTxPrelude
++import PlutusTx.Prelude as PlutusTxPrelude hiding (traceError, traceIfFalse)
+ import qualified Prelude as Haskell
+ import Unsafe.Coerce (unsafeCoerce)
+
++-- Suppress traces, in order to save bytes.
++
++{-# INLINABLE traceError #-}
++traceError :: BuiltinString -> a
++traceError _ = error ()
++
++{-# INLINABLE traceIfFalse #-}
++traceIfFalse :: BuiltinString -> a -> a
++traceIfFalse _ = id
++
++
+ -- | Input to a Marlowe transaction.
+ type MarloweInput = [MarloweTxInput]
+
+@@ -136,7 +150,7 @@ rolePayoutValidator = mkTypedValidator @TypedRolePayoutValidator
+   $$(PlutusTx.compile [|| mkRolePayoutValidator ||])
+   $$(PlutusTx.compile [|| wrap ||])
+   where
+-    wrap = Scripts.mkUntypedValidator @(CurrencySymbol, TokenName) @()
++    wrap = Scripts.mkUntypedValidator @_ @(CurrencySymbol, TokenName) @()
+
+ {-# INLINABLE rolePayoutValidator #-}
+@@ -147,7 +161,7 @@ mkRolePayoutValidator :: (CurrencySymbol, TokenName)  -- ^ The datum is the curr
+                       -> Bool                         -- ^ Whether the transaction validated.
+ mkRolePayoutValidator (currency, role) _ ctx =
+     -- The role token for the correct currency must be present.
+-    -- [Marlowe-Cardano Specification: "16. Payment authorized".]
++    -- [Marlowe-Cardano Specification: "17. Payment authorized".]
+     Val.singleton currency role 1 `Val.leq` valueSpent (scriptContextTxInfo ctx)
+
+@@ -188,12 +202,10 @@ mkMarloweValidator
+             case closeInterval $ txInfoValidRange scriptContextTxInfo of
+                 Just interval' -> interval'
+                 Nothing        -> traceError "a"
+-    -- The incoming balance of each account must be positive.
+-    -- [Marlowe-Cardano Specification: "Constraint 13. Positive balances".]
+-    let positiveBalances = traceIfFalse "b" $ validateBalances marloweState
+
+     -- Find Contract continuation in TxInfo datums by hash or fail with error.
+     let inputs = fmap marloweTxInputToInput marloweTxInputs
++
+     {-  We do not check that a transaction contains exact input payments.
+         We only require an evidence from a party, e.g. a signature for PubKey party,
+         or a spend of a 'party role' token.  This gives huge flexibility by allowing
+@@ -201,22 +213,19 @@ mkMarloweValidator
+         Then, we check scriptOutput to be correct.
+      -}
+     let inputContents = fmap getInputContent inputs
++
+     -- Check that the required signatures and role tokens are present.
+     -- [Marlowe-Cardano Specification: "Constraint 14. Inputs authorized".]
+-    let inputsOk = validateInputs inputContents
+-
+-    -- Since individual balances were validated to be positive,
+-    -- the total balance is also positive.
+-    let inputBalance = totalBalance (accounts marloweState)
++    let inputsOk = allInputsAreAuthorized inputContents
+
+     -- [Marlowe-Cardano Specification: "Constraint 5. Input value from script".]
+-    -- The total incoming balance must match the actual script value being spent.
+-    let balancesOk = traceIfFalse "v" $ inputBalance == scriptInValue
+-
+-    let preconditionsOk = positiveBalances && balancesOk
++    -- [Marlowe-Cardano Specification: "Constraint 13. Positive balances".]
++    -- [Marlowe-Cardano Specification: "Constraint 19. No duplicates".]
++    -- Check that the initial state obeys the Semantic's invariants.
++    let preconditionsOk = checkState "i" scriptInValue marloweState
+
+-    -- Package the inputs to be applied in the semantics.
+     -- [Marlowe-Cardano Specification: "Constraint 0. Input to semantics".]
++    -- Package the inputs to be applied in the semantics.
+     let txInput = TransactionInput {
+             txInterval = interval,
+             txInputs = inputs }
+@@ -246,16 +255,25 @@ mkMarloweValidator
+
+                 checkContinuation = case txOutContract of
+                     -- [Marlowe-Cardano Specification: "Constraint 4. No output to script on close".]
+-                    Close -> traceIfFalse "c" checkScriptOutputAny
++                    Close -> traceIfFalse "c" hasNoOutputToOwnScript
+                     _ -> let
+                         totalIncome = foldMap collectDeposits inputContents
+                         totalPayouts = foldMap snd payoutsByParty
+-                        finalBalance = inputBalance + totalIncome - totalPayouts
+-                        -- The total account balance must be paid to a single output to the script.
+-                        -- [Marlowe-Cardano Specification: "Constraint 3. Single Marlowe output".]
+-                        -- [Marlowe-Cardano Specification: "Constraint 6. Output value to script."]
+-                        in checkOwnOutputConstraint marloweData finalBalance
++                        finalBalance = scriptInValue + totalIncome - totalPayouts
++                        in
++                             -- [Marlowe-Cardano Specification: "Constraint 3. Single Marlowe output".]
++                             -- [Marlowe-Cardano Specification: "Constraint 6. Output value to script."]
++                             -- Check that the single Marlowe output has the correct datum and value.
++                             checkOwnOutputConstraint marloweData finalBalance
++                             -- [Marlowe-Cardano Specification: "Constraint 18. Final balance."]
++                             -- [Marlowe-Cardano Specification: "Constraint 13. Positive balances".]
++                             -- [Marlowe-Cardano Specification: "Constraint 19. No duplicates".]
++                             -- Check that the final state obeys the Semantic's invariants.
++                          && checkState "o" finalBalance txOutState
+             preconditionsOk && inputsOk && payoutsOk && checkContinuation
++              -- [Marlowe-Cardano Specification: "20. Single satsifaction".]
++              -- Either there must be no payouts, or there must be no other validators.
++              && traceIfFalse "z" (null payoutsByParty || noOthers)
+         Error TEAmbiguousTimeIntervalError -> traceError "i"
+         Error TEApplyNoMatchError -> traceError "n"
+         Error (TEIntervalError (InvalidInterval _)) -> traceError "j"
+@@ -274,24 +292,73 @@ mkMarloweValidator
+         find (\TxInInfo{txInInfoOutRef} -> txInInfoOutRef == txOutRef) txInfoInputs
+     findOwnInput _ = Nothing
+
+-    -- The inputs being spent by this script.
+     -- [Marlowe-Cardano Specification: "2. Single Marlowe script input".]
++    -- The inputs being spent by this script, and whether other validators are present.
+     ownInput :: TxInInfo
+-    ownInput@TxInInfo{txInInfoResolved=TxOut{txOutAddress=ownAddress}} =
++    noOthers :: Bool
++    (ownInput@TxInInfo{txInInfoResolved=TxOut{txOutAddress=ownAddress}}, noOthers) =
+         case findOwnInput ctx of
+-            Just ownTxInInfo ->
+-                case filter (sameValidatorHash ownTxInInfo) (txInfoInputs scriptContextTxInfo) of
+-                    [i] -> i
+-                    _   -> traceError "w" -- Multiple Marlowe contract inputs with the same address are forbidden.
++            Just ownTxInInfo -> examineScripts (sameValidatorHash ownTxInInfo) Nothing True (txInfoInputs scriptContextTxInfo)
+             _ -> traceError "x" -- Input to be validated was not found.
+
++    -- Check for the presence of multiple Marlowe validators or other Plutus validators.
++    examineScripts
++      :: (ValidatorHash -> Bool)  -- Test for this validator.
++      -> Maybe TxInInfo           -- The input for this validator, if found so far.
++      -> Bool                     -- Whether no other validator has been found so far.
++      -> [TxInInfo]               -- The inputs remaining to be examined.
++      -> (TxInInfo, Bool)         -- The input for this validator and whehter no other validators are present.
++    -- This validator has not been found.
++    examineScripts _ Nothing _ [] = traceError "x"
++    -- This validator has been found, and other validators may have been found.
++    examineScripts _ (Just self) noOthers [] = (self, noOthers)
++    -- Found both this validator and another script, so we short-cut.
++    examineScripts _ (Just self) False _ = (self, False)
++     -- Found one script.
++    examineScripts f mSelf noOthers (tx@TxInInfo{txInInfoResolved=TxOut{txOutAddress=Ledger.Address (ScriptCredential vh) _}} : txs)
++      -- The script is this validator.
++      | f vh = case mSelf of
++                 -- We hadn't found it before, so we save it in `mSelf`.
++                 Nothing -> examineScripts f (Just tx) noOthers txs
++                 -- We already had found this validator before
++                 Just _ -> traceError "w"
++      -- The script is something else, so we set `noOther` to `False`.
++      | otherwise = examineScripts f mSelf False txs
++    -- An input without a validator is encountered.
++    examineScripts f self others (_ : txs) = examineScripts f self others txs
++
+     -- Check if inputs are being spent from the same script.
+-    sameValidatorHash:: TxInInfo -> TxInInfo -> Bool
+-    sameValidatorHash
+-        TxInInfo{txInInfoResolved=TxOut{txOutAddress=Ledger.Address (ScriptCredential vh1) _}}
+-        TxInInfo{txInInfoResolved=TxOut{txOutAddress=Ledger.Address (ScriptCredential vh2) _}} = vh1 == vh2
++    sameValidatorHash:: TxInInfo -> ValidatorHash -> Bool
++    sameValidatorHash TxInInfo{txInInfoResolved=TxOut{txOutAddress=Ledger.Address (ScriptCredential vh1) _}} vh2 = vh1 == vh2
+     sameValidatorHash _ _ = False
+
++    -- Check a state for the correct value, positive accounts, and no duplicates.
++    checkState :: BuiltinString -> Val.Value -> State -> Bool
++    checkState tag expected State{..} =
++      let
++        positiveBalance :: (a, Integer) -> Bool
++        positiveBalance (_, balance) = balance > 0
++        noDuplicates :: Eq k => AssocMap.Map k v -> Bool
++        noDuplicates am =
++          let
++            test [] = True           -- An empty list has no duplicates.
++            test (x : xs)            -- Look for a duplicate of the head in the tail.
++              | elem x xs = False    -- A duplicate is present.
++              | otherwise = test xs  -- Continue searching for a duplicate.
++          in
++            test $ AssocMap.keys am
++      in
++           -- [Marlowe-Cardano Specification: "Constraint 5. Input value from script".]
++           -- and/or
++           -- [Marlowe-Cardano Specification: "Constraint 18. Final balance."]
++           traceIfFalse ("v"  <> tag) (totalBalance accounts == expected)
++           -- [Marlowe-Cardano Specification: "Constraint 13. Positive balances".]
++        && traceIfFalse ("b"  <> tag) (all positiveBalance $ AssocMap.toList accounts)
++           -- [Marlowe-Cardano Specification: "Constraint 19. No duplicates".]
++        && traceIfFalse ("ea" <> tag) (noDuplicates accounts)
++        && traceIfFalse ("ec" <> tag) (noDuplicates choices)
++        && traceIfFalse ("eb" <> tag) (noDuplicates boundValues)
++
+     -- Look up the Datum hash for specific data.
+     findDatumHash' :: PlutusTx.ToData o => o -> Maybe DatumHash
+     findDatumHash' datum = findDatumHash (Datum $ PlutusTx.toBuiltinData datum) scriptContextTxInfo
+@@ -301,7 +368,7 @@ mkMarloweValidator
+     checkOwnOutputConstraint ocDatum ocValue =
+         let hsh = findDatumHash' ocDatum
+         in traceIfFalse "d" -- "Output constraint"
+-        $ checkScriptOutput ownAddress hsh ocValue getContinuingOutput
++        $ checkScriptOutput (==) ownAddress hsh ocValue getContinuingOutput
+
+     getContinuingOutput :: TxOut
+     getContinuingOutput = case filter (\TxOut{txOutAddress} -> ownAddress == txOutAddress) allOutputs of
+@@ -309,20 +376,14 @@ mkMarloweValidator
+         _     -> traceError "o" -- No continuation or multiple Marlowe contract outputs is forbidden.
+
+     -- Check that address, value, and datum match the specified.
+-    checkScriptOutput :: Ledger.Address -> Maybe DatumHash -> Val.Value -> TxOut -> Bool
+-    checkScriptOutput addr hsh value TxOut{txOutAddress, txOutValue, txOutDatum=OutputDatumHash svh} =
+-                    txOutValue == value && hsh == Just svh && txOutAddress == addr
+-    checkScriptOutput _ _ _ _ = False
+-
+-    -- Check that address and datum match the specified, and that value is at least that required.
+-    checkScriptOutputRelaxed :: Ledger.Address -> Maybe DatumHash -> Val.Value -> TxOut -> Bool
+-    checkScriptOutputRelaxed addr hsh value TxOut{txOutAddress, txOutValue, txOutDatum=OutputDatumHash svh} =
+-                    txOutValue `Val.geq` value && hsh == Just svh && txOutAddress == addr
+-    checkScriptOutputRelaxed _ _ _ _ = False
++    checkScriptOutput :: (Val.Value -> Val.Value -> Bool) -> Ledger.Address -> Maybe DatumHash -> Val.Value -> TxOut -> Bool
++    checkScriptOutput comparison addr hsh value TxOut{txOutAddress, txOutValue, txOutDatum=OutputDatumHash svh} =
++                    txOutValue `comparison` value && hsh == Just svh && txOutAddress == addr
++    checkScriptOutput _ _ _ _ _ = False
+
+     -- Check for any output to the script address.
+-    checkScriptOutputAny :: Bool
+-    checkScriptOutputAny = all ((/= ownAddress) . txOutAddress) allOutputs
++    hasNoOutputToOwnScript :: Bool
++    hasNoOutputToOwnScript = all ((/= ownAddress) . txOutAddress) allOutputs
+
+     -- All of the script outputs.
+     allOutputs :: [TxOut]
+@@ -339,8 +400,8 @@ mkMarloweValidator
+     marloweTxInputToInput (Input input) = NormalInput input
+
+     -- Check that inputs are authorized.
+-    validateInputs :: [InputContent] -> Bool
+-    validateInputs = all validateInputWitness
++    allInputsAreAuthorized :: [InputContent] -> Bool
++    allInputsAreAuthorized = all validateInputWitness
+       where
+         validateInputWitness :: InputContent -> Bool
+         validateInputWitness input =
+@@ -356,8 +417,10 @@ mkMarloweValidator
+
+     -- Tally the deposits in the input.
+     collectDeposits :: InputContent -> Val.Value
+-    collectDeposits (IDeposit _ _ (Token cur tok) amount) = Val.singleton cur tok amount
+-    collectDeposits _                                     = zero
++    collectDeposits (IDeposit _ _ (Token cur tok) amount)
++      | amount > 0    = Val.singleton cur tok amount  -- SCP-5123: Semantically negative deposits
++      | otherwise     = zero                          -- do not remove funds from the script's UTxO.
++    collectDeposits _ = zero
+
+     -- Extract the payout to a party.
+     payoutByParty :: Payment -> AssocMap.Map Party Val.Value
+@@ -374,12 +437,17 @@ mkMarloweValidator
+       where
+         payoutToTxOut :: (Party, Val.Value) -> Bool
+         payoutToTxOut (party, value) = case party of
++            -- [Marlowe-Cardano Specification: "Constraint 15. Sufficient Payment".]
++            -- SCP-5128: Note that the payment to an address may be split into several outputs but the payment to a role must be
++            -- a single output. The flexibily of multiple outputs accommodates wallet-related practicalities such as the change and
++            -- the return of the role token being in separate UTxOs in situations where a contract is also paying to the address
++            -- where that change and that role token are sent.
+             Address _ address  -> traceIfFalse "p" $ value `Val.leq` valuePaidToAddress address  -- At least sufficient value paid.
+             Role role -> let
+                 hsh = findDatumHash' (rolesCurrency, role)
+                 addr = Address.scriptHashAddress rolePayoutValidatorHash
+                 -- Some output must have the correct value and datum to the role-payout address.
+-                in traceIfFalse "r" $ any (checkScriptOutputRelaxed addr hsh value) allOutputs
++                in traceIfFalse "r" $ any (checkScriptOutput Val.geq addr hsh value) allOutputs
+
+     -- The key for the address must have signed.
+     txSignedByAddress :: Ledger.Address -> Bool
+@@ -402,7 +470,7 @@ marloweValidator :: Scripts.TypedValidator TypedMarloweValidator
+ marloweValidator =
+   let
+     mkUntypedMarloweValidator :: ValidatorHash -> BuiltinData -> BuiltinData -> BuiltinData -> ()
+-    mkUntypedMarloweValidator rp = mkUntypedValidator (mkMarloweValidator rp)
++    mkUntypedMarloweValidator rp = Scripts.mkUntypedValidator (mkMarloweValidator rp)
+
+     untypedValidator :: Scripts.Validator
+     untypedValidator = mkValidatorScript $
+@@ -410,10 +478,20 @@ marloweValidator =
+         `PlutusTx.applyCode` PlutusTx.liftCode rolePayoutValidatorHash
+
+     typedValidator :: Scripts.TypedValidator Scripts.Any
+-    typedValidator = Scripts.unsafeMkTypedValidator untypedValidator
++    typedValidator = unsafeMkTypedValidator (Scripts.Versioned untypedValidator Scripts.PlutusV2)
+   in
+     unsafeCoerce typedValidator
+
++
++marloweValidatorCompiled :: PlutusTx.CompiledCode (ValidatorHash -> PlutusTx.BuiltinData -> PlutusTx.BuiltinData -> PlutusTx.BuiltinData -> ())
++marloweValidatorCompiled =
++  let
++    mkUntypedMarloweValidator :: ValidatorHash -> PlutusTx.BuiltinData -> PlutusTx.BuiltinData -> PlutusTx.BuiltinData -> ()
++    mkUntypedMarloweValidator rp = Scripts.mkUntypedValidator (mkMarloweValidator rp)
++  in
++    $$(PlutusTx.compile [|| mkUntypedMarloweValidator ||])
++
++
+ -- | The hash of the Marlowe semantics validator.
+ marloweValidatorHash :: ValidatorHash
+ marloweValidatorHash = Scripts.validatorHash marloweValidator
+@@ -432,7 +510,7 @@ alternateMarloweValidator = Scripts.mkTypedValidator
+           $$(PlutusTx.compile [|| mkMarloweValidator ||])
+             `PlutusTx.applyCode`
+               PlutusTx.liftCode rolePayoutValidatorHash
+-        mkArgsValidator = mkUntypedValidator @MarloweData @MarloweInput
++        mkArgsValidator = Scripts.mkUntypedValidator @_ @MarloweData @MarloweInput
+         compiledArgsValidator =
+           $$(PlutusTx.compile [|| mkArgsValidator ||])
+```
+
+### Post-Audit Changes in Property-Based Tests for Marlowe Validator
+
+```bash
+git diff 523f3d56f22bf992ddb0b0c8a52bb7a5a188f9e9 69468d6623e24a4ccd6659e378ae012c38ca01ce marlowe-test/src/
+```
+
+```diff
+diff --git a/marlowe-test/src/Data/Jsonable.hs b/marlowe-test/src/Data/Jsonable.hs
+new file mode 100644
+index 000000000..b16e3e044
+--- /dev/null
++++ b/marlowe-test/src/Data/Jsonable.hs
+@@ -0,0 +1,126 @@
++-----------------------------------------------------------------------------
++--
++-- Module      :  $Headers
++-- License     :  Apache 2.0
++--
++-- Stability   :  Experimental
++-- Portability :  Portable
++--
++-- | Wrapping values that can be converted to and from JSON.
++--
++-----------------------------------------------------------------------------
++
++
++{-# LANGUAGE ExistentialQuantification #-}
++{-# LANGUAGE OverloadedStrings #-}
++{-# LANGUAGE RecordWildCards #-}
++{-# LANGUAGE ScopedTypeVariables #-}
++
++
++module Data.Jsonable
++  ( -- * Types
++    Jsonable(..)
++  , JsonableType(..)
++  , KnownJsonable
++    -- * Conversion
++  , fromJsonable
++  , isKnownJson
++  , toJsonable
++    -- * Testing
++  , arbitraryJsonable
++  , checkRoundTripJsonable
++  , generateJsonable
++  , roundTripJsonable
++  ) where
++
++
++import Data.Aeson (FromJSON(..), Result(..), ToJSON(..), Value, fromJSON)
++import Data.Proxy (Proxy(..))
++import Test.Tasty.QuickCheck (Arbitrary(..), Gen)
++
++
++-- | A list of known types with `FromJSON`, `ToJSON`, and `Arbitrary` instances.
++type KnownJsonable = [JsonableType]
++
++
++-- | A type with `FromJSON`, `ToJSON`, and `Arbitrary` instances.
++data JsonableType = forall a. (ToJSON a, FromJSON a, Arbitrary a) => JsonableType {typeKey :: String, typeValue :: Proxy a}
++
++
++-- | A value with `FromJSON`, `ToJSON`, and `Arbitrary` instances.
++data Jsonable = forall a . (ToJSON a, FromJSON a, Arbitrary a) => Jsonable {jsonableType :: String, jsonableValue :: a}
++
++
++-- | Deserialize a JSON value.
++toJsonable
++  :: KnownJsonable  -- ^ A list of known types with `FromJSON`, `ToJSON`, and `Arbitrary` instances.
++  -> String   -- ^ The key for the type.
++  -> Value  -- ^ The JSON value.
++  -> Either String Jsonable  -- ^ The deserialized value, or an error message.
++toJsonable known s v =
++  let
++    fromJson' :: forall a . FromJSON a => Proxy a -> Either String a
++    fromJson' _ =
++      case fromJSON v :: Result a of
++        Success x -> Right x
++        Error msg -> Left msg
++  in
++    case filter ((s ==) . typeKey) known of
++      [JsonableType{..}] -> Jsonable typeKey <$> fromJson' typeValue
++      _                  -> Left $ "JSON serialization not supported for " <> s <> "."
++
++
++-- | Check for a known JSON type.
++isKnownJson
++  :: KnownJsonable  -- ^ A list of known types with `FromJSON`, `ToJSON`, and `Arbitrary` instances.
++  -> String   -- ^ The key for the type.
++  -> Bool  -- ^ Whether the type is in the list of known JSON types.
++isKnownJson known s = length (filter ((s ==) . typeKey) known) == 1
++
++
++-- | Serialize a JSON value.
++fromJsonable
++  :: Jsonable  -- ^ The value.
++  -> Value  -- ^ The JSON.
++fromJsonable Jsonable{..} = toJSON jsonableValue
++
++
++-- | Peform round-trip deserialization and serialization of a JSON value.
++roundTripJsonable
++  :: KnownJsonable  -- ^ A list of known types with `FromJSON`, `ToJSON`, and `Arbitrary` instances.
++  -> String  -- ^ The key for the type.
++  -> Value  -- ^ The JSON value.
++  -> Either String Value  -- ^ There the re-serialized value, or an error message.
++roundTripJsonable known = (fmap fromJsonable .) . toJsonable known
++
++
++-- | Check that round-trip re-serialization doesn't alter a JSON value.
++checkRoundTripJsonable
++  :: KnownJsonable  -- ^ A list of known types with `FromJSON`, `ToJSON`, and `Arbitrary` instances.
++  -> String  -- ^ The key for the type.
++  -> Value  -- ^ The JSON value.
++  -> Bool  -- ^ Whether the JSON value is unchanged.
++checkRoundTripJsonable known s v = roundTripJsonable known s v == Right v
++
++
++-- | Generate an arbitrary value.
++arbitraryJsonable
++  :: JsonableType  -- ^ The type to generate.
++  -> Gen Jsonable  -- ^ The generator.
++arbitraryJsonable JsonableType{..} =
++  let
++    gen :: Arbitrary a => Proxy a -> Gen a
++    gen _ = arbitrary
++  in
++    Jsonable typeKey <$> gen typeValue
++
++
++-- | Generate an arbitrary value.
++generateJsonable
++  :: KnownJsonable  -- ^ A list of known types with `FromJSON`, `ToJSON`, and `Arbitrary` instances.
++  -> String  -- ^ The key for the type.
++  -> Either String (Gen Value)  -- ^ The JSON value.
++generateJsonable known s =
++  case filter ((s ==) . typeKey) known of
++    [jt] -> Right . fmap fromJsonable $ arbitraryJsonable jt
++    _    -> Left $ "JSON serialization not supported for " <> s <> "."
+diff --git a/marlowe-test/src/Spec/Marlowe/Common.hs b/marlowe-test/src/Spec/Marlowe/Common.hs
+index f52715b5f..77f7ef7cd 100644
+--- a/marlowe-test/src/Spec/Marlowe/Common.hs
++++ b/marlowe-test/src/Spec/Marlowe/Common.hs
+@@ -12,7 +12,6 @@
+
+ {-# LANGUAGE DataKinds #-}
+-{-# LANGUAGE NamedFieldPuns #-}
+ {-# LANGUAGE OverloadedStrings #-}
+ {-# LANGUAGE ScopedTypeVariables #-}
+
+@@ -490,5 +489,3 @@ pangramContract = let
+                 Close)
+         , Case (Notify (AndObs (TimeIntervalStart `ValueLT` TimeIntervalEnd) TrueObs)) Close
+         ] (Ledger.POSIXTime 100) Close
+-
+-
+diff --git a/marlowe-test/src/Spec/Marlowe/Marlowe.hs b/marlowe-test/src/Spec/Marlowe/Marlowe.hs
+index f98e1c899..264b1da93 100644
+--- a/marlowe-test/src/Spec/Marlowe/Marlowe.hs
++++ b/marlowe-test/src/Spec/Marlowe/Marlowe.hs
+@@ -29,7 +29,6 @@ module Spec.Marlowe.Marlowe
+   ) where
+
+-
+ import Control.Exception (SomeException, catch)
+ import Control.Monad (when)
+ import Data.Aeson (decode, eitherDecode, encode)
+@@ -151,14 +150,14 @@ alternateMarloweValidatorSize :: IO ()
+ alternateMarloweValidatorSize = do
+     let validator = Scripts.validatorScript alternateMarloweValidator
+     let vsize = SBS.length . SBS.toShort . LB.toStrict $ Serialise.serialise validator
+-    assertBool ("alternateMarloweValidator is too large " <> show vsize) (vsize < 15030)
++    assertBool ("alternateMarloweValidator is too large " <> show vsize) (vsize <= 14821)
+
+ -- | Test that the untyped validator is not too large.
+ marloweValidatorSize :: IO ()
+ marloweValidatorSize = do
+     let validator = Scripts.validatorScript marloweValidator
+     let vsize = SBS.length . SBS.toShort . LB.toStrict $ Serialise.serialise validator
+-    assertBool ("marloweValidator is too large " <> show vsize) (vsize < 12510)
++    assertBool ("marloweValidator is too large " <> show vsize) (vsize <= 12296)
+
+ -- | Test `extractNonMerkleizedContractRoles`.
+diff --git a/marlowe-test/src/Spec/Marlowe/Plutus.hs b/marlowe-test/src/Spec/Marlowe/Plutus.hs
+index f1a44eaea..ccccdf3c1 100644
+--- a/marlowe-test/src/Spec/Marlowe/Plutus.hs
++++ b/marlowe-test/src/Spec/Marlowe/Plutus.hs
+@@ -17,9 +17,11 @@ module Spec.Marlowe.Plutus
+   ) where
+
++import Spec.Marlowe.Reference (ReferencePath)
+ import Test.Tasty (TestTree, testGroup)
+
+ import qualified Spec.Marlowe.Plutus.AssocMap (tests)
++import qualified Spec.Marlowe.Plutus.MList (tests)
+ import qualified Spec.Marlowe.Plutus.Prelude (tests)
+ import qualified Spec.Marlowe.Plutus.ScriptContext (tests)
+ import qualified Spec.Marlowe.Plutus.Specification (tests)
+@@ -27,14 +29,14 @@ import qualified Spec.Marlowe.Plutus.Value (tests)
+
+ -- | Run tests.
+-tests :: TestTree
+-tests =
++tests :: [ReferencePath] -> TestTree
++tests referencePaths =
+   testGroup "Plutus"
+     [
+       Spec.Marlowe.Plutus.Prelude.tests
+     , Spec.Marlowe.Plutus.AssocMap.tests
++    , Spec.Marlowe.Plutus.MList.tests
+     , Spec.Marlowe.Plutus.Value.tests
+     , Spec.Marlowe.Plutus.ScriptContext.tests
+-    , Spec.Marlowe.Plutus.Specification.tests
++    , Spec.Marlowe.Plutus.Specification.tests referencePaths
+     ]
+-
+diff --git a/marlowe-test/src/Spec/Marlowe/Plutus/MList.hs b/marlowe-test/src/Spec/Marlowe/Plutus/MList.hs
+new file mode 100644
+index 000000000..7edb3424e
+--- /dev/null
++++ b/marlowe-test/src/Spec/Marlowe/Plutus/MList.hs
+@@ -0,0 +1,312 @@
++-----------------------------------------------------------------------------
++--
++-- Module      :  $Headers
++-- License     :  Apache 2.0
++--
++-- Stability   :  Experimental
++-- Portability :  Portable
++--
++-- | Compare the `PlutusTx.AssocMap` to Isabelle's `MList`.
++--
++-----------------------------------------------------------------------------
++
++
++module Spec.Marlowe.Plutus.MList
++  ( -- * Testing
++    tests
++  ) where
++
++
++import Data.Function (on)
++import Data.List (nubBy, sortBy)
++import Data.Maybe (fromMaybe, isJust)
++import Prelude hiding (lookup)
++import Test.Tasty (TestTree, testGroup)
++import Test.Tasty.HUnit (Assertion, assertBool, testCase)
++import Test.Tasty.QuickCheck (Arbitrary(..), Gen, Property, elements, forAll, property, testProperty)
++
++import qualified PlutusTx.AssocMap as AM
++  (Map, delete, empty, fromList, insert, lookup, member, null, singleton, toList, unionWith)
++import qualified PlutusTx.Eq as P (Eq)
++
++
++-- | An association list in Isabelle.
++type MList a b = [(a, b)]
++
++
++-- | Empty Isabelle `MList`.
++--
++-- @
++-- definition empty :: "('a::linorder \<times> 'b) list" where
++--   "empty = Nil"
++-- @
++empty :: MList a b
++empty = []
++
++
++{-# ANN insert "HLint: ignore Use guards" #-}
++{-# ANN insert "HLint: ignore Use list literal" #-}
++
++-- | Insert into Isabelle `MList`.
++--
++-- @
++-- fun insert :: "'a::linorder \<Rightarrow> 'b \<Rightarrow> ('a \<times> 'b) list \<Rightarrow> ('a \<times> 'b) list" where
++--   "insert a b Nil = Cons (a, b) Nil" |
++--   "insert a b (Cons (x, y) z) =
++--     (if a < x
++--      then (Cons (a, b) (Cons (x, y) z))
++--      else (if a > x
++--            then (Cons (x, y) (insert a b z))
++--            else (Cons (x, b) z)))"
++-- @
++insert :: Ord a => a -> b -> MList a b -> MList a b
++insert a b [] = (a, b) : []
++insert a b ((x, y) : z) =
++  if a < x
++    then (a, b) : (x, y) : z
++    else if a > x
++      then (x, y) : insert a b z
++      else (x, b) : z
++
++
++{-# ANN delete "HLint: ignore Use guards" #-}
++
++-- | Delete from Isabelle `MList`.
++--
++-- @
++-- fun delete :: "'a::linorder \<Rightarrow> ('a \<times> 'b) list \<Rightarrow> ('a \<times> 'b) list" where
++--   "delete a Nil = Nil" |
++--   "delete a (Cons (x, y) z) =
++--     (if a = x
++--      then z
++--      else (if a > x
++--            then (Cons (x, y) (delete a z))
++--            else (Cons (x, y) z)))"
++-- @
++delete :: Ord a => a -> MList a b -> MList a b
++delete _a [] = []
++delete a ((x, y) : z) =
++  if a == x
++    then z
++    else if a > x
++      then (x, y) : delete a z
++      else (x, y) : z
++
++
++{-# ANN lookup "HLint: ignore Use guards" #-}
++
++-- | Lookup in Isabelle `MList`.
++--
++-- @
++-- fun lookup :: "'a::linorder \<Rightarrow> ('a \<times> 'b) list \<Rightarrow> 'b option" where
++--   "lookup a Nil = None" |
++--   "lookup a (Cons (x, y) z) =
++--     (if a = x
++--      then Some y
++--      else (if a > x
++--            then lookup a z
++--            else None))"
++-- @
++lookup :: Ord a => a -> MList a b -> Maybe b
++lookup _a [] = Nothing
++lookup a ((x, y) : z) =
++  if a == x
++    then Just y
++    else if a > x
++      then lookup a z
++      else Nothing
++
++
++{-# ANN unionWith "HLint: ignore Use guards" #-}
++
++-- | Union with Isabelle `MList`.
++--
++-- @
++-- fun unionWith :: "('b \<Rightarrow> 'b \<Rightarrow> 'b) \<Rightarrow> ('a \<times> 'b) list \<Rightarrow>
++--                   ('a \<times> 'b) list \<Rightarrow> (('a::linorder) \<times> 'b) list" where
++--   "unionWith f (Cons (x, y) t) (Cons (x2, y2) t2) =
++--     (if x < x2
++--      then Cons (x, y) (unionWith f t (Cons (x2, y2) t2))
++--      else (if x > x2
++--            then Cons (x2, y2) (unionWith f (Cons (x, y) t) t2)
++--            else Cons (x, f y y2) (unionWith f t t2)))" |
++--   "unionWith f Nil l = l" |
++--   "unionWith f l Nil = l"
++-- @
++unionWith :: Ord a => (b -> b -> b) -> MList a b -> MList a b -> MList a b
++unionWith f ((x, y) : t) ((x2, y2) : t2) =
++  if x < x2
++    then (x, y) : unionWith f t ((x2, y2) : t2)
++    else if x > x2
++      then (x2, y2) : unionWith f ((x, y) : t) t2
++      else (x, f y y2) : unionWith f t t2
++unionWith _f [] l = l
++unionWith _f l [] = l
++
++
++-- | Find with default for Isabelle `MList`.
++--
++-- @
++-- fun findWithDefault :: "'b \<Rightarrow> 'a \<Rightarrow> (('a::linorder) \<times> 'b) list \<Rightarrow> 'b" where
++--   "findWithDefault d k l = (case lookup k l of
++--                               None \<Rightarrow> d
++--                             | Some x \<Rightarrow> x)"
++-- @
++findWithDefault :: Ord a => b -> a -> MList a b -> b
++findWithDefault d k l =
++  case lookup k l of
++    Nothing -> d
++    Just x -> x
++
++
++-- | Run tests.
++tests :: TestTree
++tests =
++  testGroup "MList vs AssocMap"
++    [
++      testCase     "`empty` is equivalent"           checkEmpty
++    , testProperty "`null` is equivalent"            checkNull
++    , testProperty "`singleton` is equivalent"       checkSingleton
++    , testProperty "`insert` is equivalent"          checkInsert
++    , testProperty "`delete` is equivalent"          checkDelete
++    , testProperty "`lookup` is equivalent"          checkLookup
++    , testProperty "`member` is equivalent"          checkMember
++    , testProperty "`unionWith` is equivalent"       checkUnionWith
++    , testProperty "`findWithDefault` is equivalent" checkFindWithDefault
++    ]
++
++
++-- | Key type for testing.
++type Key = Integer
++
++
++-- | Value type for testing.
++type Value = [()]
++
++
++-- | Generate a sorted `MList` with no duplicate keys.
++arbitraryMList :: Gen (MList Key Value, AM.Map Key Value)
++arbitraryMList =
++  do
++    items <- nubBy ((==) `on` fst) <$> arbitrary
++    pure (sortBy (compare `on` fst) items, AM.fromList items)
++
++
++-- | Compare an `MList` to an `AssocMap`, ignoring ordering.
++equivalent :: Ord a => P.Eq a => Eq b => MList a b -> AM.Map a b -> Bool
++equivalent mlist assocmap =
++  mlist == sortBy (compare `on` fst) (AM.toList assocmap)
++    && and [AM.lookup a assocmap == Just b | (a, b) <- mlist]
++    && and [lookup a mlist == Just b | (a, b) <- AM.toList assocmap]
++
++
++-- | Compare `empty` for `MList` and `AssocMap`, provided the `MList` is sorted and neither contains duplicate keys.
++checkEmpty :: Assertion
++checkEmpty = assertBool "Empty MList and AssocMap" $ (empty :: MList Key Value) `equivalent` AM.empty
++
++
++-- | Compare `null` for `MList` and `AssocMap`, provided the `MList` is sorted and neither contains duplicate keys.
++checkNull :: Property
++checkNull = property $ do
++  let
++    gen = do
++      (mlist, assocmap) <- arbitraryMList
++      pure (mlist, assocmap)
++  forAll gen
++    $ \(mlist, assocmap) -> (== empty) mlist == AM.null assocmap
++
++
++-- | Compare `singleton` for `MList` and `AssocMap`, provided the `MList` is sorted and neither contains duplicate keys.
++checkSingleton :: Property
++checkSingleton = property $ do
++  let
++    gen = do
++      a <- arbitrary :: Gen Key
++      b <- arbitrary :: Gen Value
++      pure (a, b)
++  forAll gen
++    $ \(a, b) -> [(a, b)] `equivalent` AM.singleton a b
++
++
++-- | Compare `insert` for `MList` and `AssocMap`, provided the `MList` is sorted and neither contains duplicate keys.
++checkInsert :: Property
++checkInsert = property $ do
++  let
++    gen = do
++      (mlist, assocmap) <- arbitraryMList
++      a <- arbitrary
++      b <- arbitrary
++      pure (mlist, assocmap, a, b)
++  forAll gen
++    $ \(mlist, assocmap, a, b) -> insert a b mlist `equivalent` AM.insert a b assocmap
++
++
++-- | Compare `delete` for `MList` and `AssocMap`, provided the `MList` is sorted and neither contains duplicate keys.
++checkDelete :: Property
++checkDelete = property $ do
++  let
++    gen = do
++      (mlist, assocmap) <- arbitraryMList
++      a <- arbitrary
++      pure (mlist, assocmap, a)
++  forAll gen
++    $ \(mlist, assocmap, a) -> delete a mlist `equivalent` AM.delete a assocmap
++
++
++-- | Compare `lookup` for `MList` and `AssocMap`, provided the `MList` is sorted and neither contains duplicate keys.
++checkLookup :: Property
++checkLookup = property $ do
++  let
++    gen = do
++      (mlist, assocmap) <- arbitraryMList
++      a <- arbitrary
++      pure (mlist, assocmap, a)
++  forAll gen
++    $ \(mlist, assocmap, a) -> lookup a mlist == AM.lookup a assocmap
++
++
++-- | Compare `member` for `MList` and `AssocMap`, provided the `MList` is sorted and neither contains duplicate keys.
++checkMember :: Property
++checkMember = property $ do
++  let
++    gen = do
++      (mlist, assocmap) <- arbitraryMList
++      a <- arbitrary
++      pure (mlist, assocmap, a)
++  forAll gen
++    $ \(mlist, assocmap, a) -> isJust (lookup a mlist) == AM.member a assocmap
++
++
++-- | Compare `unionWith` for `MList` and `AssocMap`, provided the `MList` is sorted and neither contains duplicate keys.
++checkUnionWith :: Property
++checkUnionWith = property $ do
++  let
++    gen = do
++      (mlist, assocmap) <- arbitraryMList
++      (mlist', assocmap') <- arbitraryMList
++      function <- elements ["concat", "shortest", "longest", "first", "second"]
++      pure (mlist, assocmap, mlist', assocmap', function)
++  forAll gen
++    $ \(mlist, assocmap, mlist', assocmap', function) ->
++      let
++        f = case function of
++              "shortest" -> \x y -> if length x < length y then x else y
++              "longest"  -> \x y -> if length x > length y then x else y
++              "first"    -> const
++              "second"   -> const id
++              _          -> (<>)
++      in
++        unionWith f mlist mlist' `equivalent` AM.unionWith f assocmap assocmap'
++
++
++-- | Compare `findWithDefault` for `MList` and `AssocMap`, provided the `MList` is sorted and neither contains duplicate keys.
++checkFindWithDefault :: Property
++checkFindWithDefault = property $ do
++  let
++    gen = do
++      (mlist, assocmap) <- arbitraryMList
++      a <- arbitrary
++      b <- arbitrary
++      pure (mlist, assocmap, a, b)
++  forAll gen
++    $ \(mlist, assocmap, a, b) -> findWithDefault b a mlist == fromMaybe b (AM.lookup a assocmap)
+diff --git a/marlowe-test/src/Spec/Marlowe/Plutus/Script.hs b/marlowe-test/src/Spec/Marlowe/Plutus/Script.hs
+index ccfab1c9a..9ced8d81e 100644
+--- a/marlowe-test/src/Spec/Marlowe/Plutus/Script.hs
++++ b/marlowe-test/src/Spec/Marlowe/Plutus/Script.hs
+@@ -11,7 +11,9 @@
+ -----------------------------------------------------------------------------
+
++{-# LANGUAGE NumericUnderscores #-}
+ {-# LANGUAGE OverloadedStrings #-}
++{-# LANGUAGE RecordWildCards #-}
+
+ module Spec.Marlowe.Plutus.Script
+@@ -53,6 +55,7 @@ import Plutus.V2.Ledger.Api
+   , Data
+   , Datum(Datum)
+   , DatumHash
++  , ExBudget(..)
+   , ToData(toBuiltinData)
+   , TokenName
+   , Validator(getValidator)
+@@ -64,6 +67,11 @@ import qualified Data.ByteString.Short as SBS (ShortByteString, toShort)
+ import qualified Data.Map.Strict as M (fromList)
+
++-- | Check the Plutus execution budget.
++enforceBudget :: Bool
++enforceBudget = False
++
++
+ -- | Run the Plutus evaluator on the Marlowe semantics validator.
+ evaluateSemantics :: Data                    -- ^ The datum.
+                   -> Data                    -- ^ The redeemer.
+@@ -72,9 +80,11 @@ evaluateSemantics :: Data                    -- ^ The datum.
+ evaluateSemantics datum redeemer context =
+   case evaluationContext of
+     Left message -> This message
+-    Right ec     -> case evaluateScriptCounting PlutusV2 (ProtocolVersion 7 0) Verbose ec serialiseSemanticsValidator [datum, redeemer, context] of
+-                      (logOutput, Right _     ) -> That logOutput
+-                      (logOutput, Left message) -> These (show message) logOutput
++    Right ec     -> case evaluateScriptCounting PlutusV2 (ProtocolVersion 8 0) Verbose ec serialiseSemanticsValidator [datum, redeemer, context] of
++                      (logOutput, Right ex@ExBudget{..}) -> if enforceBudget && (exBudgetCPU > 10_000_000_000 || exBudgetMemory > 14_000_000)
++                                                              then These ("Exceeded Plutus budget: " <> show ex) logOutput
++                                                              else That logOutput
++                      (logOutput, Left message         ) -> These (show message) logOutput
+
+ -- | Run the Plutus evaluator on the Marlowe payout validator.
+@@ -321,12 +331,12 @@ costModel =
+     , ("unListData-memory-arguments", 32)
+     , ("unMapData-cpu-arguments", 38314)
+     , ("unMapData-memory-arguments", 32)
+-    , ("verifyEcdsaSecp256k1Signature-cpu-arguments", 20000000000)
+-    , ("verifyEcdsaSecp256k1Signature-memory-arguments", 20000000000)
+-    , ("verifyEd25519Signature-cpu-arguments-intercept", 9462713)
+-    , ("verifyEd25519Signature-cpu-arguments-slope", 1021)
++    , ("verifyEcdsaSecp256k1Signature-cpu-arguments", 35892428)
++    , ("verifyEcdsaSecp256k1Signature-memory-arguments", 10)
++    , ("verifyEd25519Signature-cpu-arguments-intercept", 57996947)
++    , ("verifyEd25519Signature-cpu-arguments-slope", 18975)
+     , ("verifyEd25519Signature-memory-arguments", 10)
+-    , ("verifySchnorrSecp256k1Signature-cpu-arguments-intercept", 20000000000)
+-    , ("verifySchnorrSecp256k1Signature-cpu-arguments-slope", 0)
+-    , ("verifySchnorrSecp256k1Signature-memory-arguments", 20000000000)
++    , ("verifySchnorrSecp256k1Signature-cpu-arguments-intercept", 38887044)
++    , ("verifySchnorrSecp256k1Signature-cpu-arguments-slope", 32947)
++    , ("verifySchnorrSecp256k1Signature-memory-arguments", 10)
+     ]
+diff --git a/marlowe-test/src/Spec/Marlowe/Plutus/Specification.hs b/marlowe-test/src/Spec/Marlowe/Plutus/Specification.hs
+index e02c42079..fc8d92b8f 100644
+--- a/marlowe-test/src/Spec/Marlowe/Plutus/Specification.hs
++++ b/marlowe-test/src/Spec/Marlowe/Plutus/Specification.hs
+@@ -23,9 +23,10 @@ module Spec.Marlowe.Plutus.Specification
+   ) where
+
+-import Control.Lens (use, uses, (%=), (<>=), (^.))
++import Control.Lens (use, uses, (%=), (<>=), (<~), (^.))
+ import Control.Monad.State (lift)
+ import Data.Bifunctor (bimap)
++import Data.List (nub)
+ import Data.Maybe (maybeToList)
+ import Data.Proxy (Proxy(..))
+ import Data.These (These(That, These, This))
+@@ -35,6 +36,7 @@ import Language.Marlowe.Core.V1.Semantics
+   , Payment(Payment)
+   , TransactionInput(..)
+   , TransactionOutput(txOutContract, txOutPayments, txOutState)
++  , totalBalance
+   )
+ import Language.Marlowe.Core.V1.Semantics.Types
+   ( ChoiceId(ChoiceId)
+@@ -54,7 +56,7 @@ import Plutus.V2.Ledger.Api
+   , BuiltinData(BuiltinData)
+   , Credential(PubKeyCredential)
+   , Data(B, Constr, List)
+-  , Datum(Datum)
++  , Datum(..)
+   , DatumHash(DatumHash)
+   , FromData(..)
+   , OutputDatum(..)
+@@ -71,12 +73,14 @@ import Plutus.V2.Ledger.Api
+   , singleton
+   , toData
+   )
++import Spec.Marlowe.Plutus.Lens ((<><~))
+ import Spec.Marlowe.Plutus.Script
+   (evaluatePayout, evaluateSemantics, payoutAddress, payoutScriptHash, semanticsAddress, semanticsScriptHash)
+ import Spec.Marlowe.Plutus.Transaction
+   ( ArbitraryTransaction
+   , arbitraryPayoutTransaction
+   , arbitrarySemanticsTransaction
++  , isScriptTxIn
+   , merkleize
+   , noModify
+   , noVeto
+@@ -97,26 +101,41 @@ import Spec.Marlowe.Plutus.Types
+   , output
+   , role
+   )
++import Spec.Marlowe.Reference (ReferencePath)
+ import Spec.Marlowe.Semantics.Arbitrary (arbitraryPositiveInteger)
+ import Test.Tasty (TestTree, testGroup)
+-import Test.Tasty.QuickCheck (Arbitrary(..), Gen, Property, forAll, property, suchThat, testProperty, (===))
++import Test.Tasty.QuickCheck
++  ( Arbitrary(..)
++  , Gen
++  , Property
++  , chooseInteger
++  , elements
++  , forAll
++  , listOf1
++  , oneof
++  , property
++  , suchThat
++  , testProperty
++  , (===)
++  )
+
+ import qualified Language.Marlowe.Core.V1.Semantics as M (MarloweData(marloweParams))
+-import qualified Language.Marlowe.Core.V1.Semantics.Types as M (Party(Address))
+-import qualified PlutusTx.AssocMap as AM (fromList, insert, toList)
++import qualified Language.Marlowe.Core.V1.Semantics.Types as M (Party(Address), State(..))
++import qualified PlutusTx.AssocMap as AM (Map, fromList, insert, keys, null, toList)
++import qualified Test.Tasty.QuickCheck as Q (shuffle)
+
+ -- | Run tests.
+-tests :: TestTree
+-tests =
++tests :: [ReferencePath] -> TestTree
++tests referencePaths =
+   testGroup "Marlowe On-Chain Specification"
+     [
+       testGroup "Semantics Validator"
+         [
+           testGroup "Valid transaction succeeds"
+             [
+-              testProperty "Noiseless" $ checkSemanticsTransaction noModify noModify noVeto True False
+-            , testProperty "Noisy"     $ checkSemanticsTransaction noModify noModify noVeto True True
++              testProperty "Noiseless" $ checkSemanticsTransaction referencePaths noModify noModify noVeto True False True
++            , testProperty "Noisy"     $ checkSemanticsTransaction referencePaths noModify noModify noVeto True True  True
+             ]
+         , testGroup "Constraint 1. Typed validation"
+             [
+@@ -129,23 +148,23 @@ tests =
+             ]
+         , testGroup "Constraint 2. Single Marlowe script input"
+             [
+-              testProperty "Invalid attempt to run two Marlowe scripts" checkDoubleInput
++              testProperty "Invalid attempt to run two Marlowe scripts" $ checkDoubleInput referencePaths
+             ]
+         , testGroup "Constraint 3. Single Marlowe output"
+             [
+-              testProperty "Invalid attempt to split Marlowe output" checkMultipleOutput
++              testProperty "Invalid attempt to split Marlowe output" $ checkMultipleOutput referencePaths
+             ]
+         , testGroup "Constraint 4. No output to script on close"
+             [
+-              testProperty "Invalid attempt to output to Marlowe on close" checkCloseOutput
++              testProperty "Invalid attempt to output to Marlowe on close" $ checkCloseOutput referencePaths
+             ]
+         , testGroup "Constraint 5. Input value from script"
+             [
+-              testProperty "Invalid mismatch between state and script input" checkValueInput
++              testProperty "Invalid mismatch between state and script input" $ checkValueInput referencePaths
+             ]
+         , testGroup "Constraint 6. Output value to script"
+             [
+-              testProperty "Invalid mismatch between state and script input" checkValueOutput
++              testProperty "Invalid mismatch between expected and actual output to script" $ checkValueOutput referencePaths
+             ]
+         , testGroup "Constraint 7. Input state"
+             [
+@@ -157,39 +176,53 @@ tests =
+             ]
+         , testGroup "Constraint 9. Marlowe parameters"
+             [
+-              testProperty "Invalid alteration of parameters." checkParamsOutput
++              testProperty "Invalid alteration of parameters" $ checkParamsOutput referencePaths
+             ]
+         , testGroup "Constraint 10. Output state"
+             [
+-              testProperty "Invalid mismatch between state and script output" checkStateOutput
++              testProperty "Invalid mismatch between state and script output's state" $ checkStateOutput referencePaths
+             ]
+         , testGroup "Constraint 11. Output contract"
+             [
+-              testProperty "Invalid mismatch between contract and script output" checkContractOutput
++              testProperty "Invalid mismatch between contract and script output" $ checkContractOutput referencePaths
+             ]
+         , testGroup "Constraint 12. Merkleized continuations"
+             [
+-              testProperty "Valid merkleization"   $ checkMerkleization True
+-            , testProperty "Invalid merkleization" $ checkMerkleization False
++              testProperty "Valid merkleization"   $ checkMerkleization referencePaths True
++            , testProperty "Invalid merkleization" $ checkMerkleization referencePaths False
+             ]
+         , testGroup "Constraint 13. Positive balances"
+             [
+-              testProperty "Invalid non-positive balance" checkPositiveAccounts
++              testProperty "Invalid non-positive balance" $ checkPositiveAccounts referencePaths
++              -- TODO: This test on the output state requires instrumenting the Plutus script. For now, this constraint is enforced manually by code inspection.
+             ]
+         , testGroup "Constraint 14. Inputs authorized"
+             [
+-              testProperty "Invalid missing authorization" checkAuthorization
++              testProperty "Invalid missing authorization" $ checkAuthorization referencePaths
+             ]
+         , testGroup "Constraint 15. Sufficient payment"
+             [
+-              testProperty "Invalid insufficient payment" checkPayment
++              testProperty "Invalid insufficient payment" $ checkPayment referencePaths
++            ]
++        , testGroup "Constraint 18. Final balance"
++            [
++              testProperty "Invalid mismatch between output value and state" $ checkOutputConsistency referencePaths
++            ]
++        , testGroup "Constraint 19. No duplicates"
++            [
++              testProperty "Invalid duplicate accounts in input state" $ checkInputDuplicates referencePaths
++              -- TODO: This test on the output state requires instrumenting the Plutus script. For now, this constraint is enforced manually by code inspection.
++            ]
++        , testGroup "Constraint 20. Single satisfaction"
++            [
++              testProperty "Invalid other validators during payment" $ checkOtherValidators referencePaths
+             ]
+         , testProperty "Script hash matches reference hash"
+             $ checkValidatorHash semanticsScriptHash
+               -- DO NOT ALTER THE FOLLOWING VALUE UNLESS YOU ARE COMMITTING
+               -- APPROVED CHANGES TO MARLOWE'S SEMANTICS VALIDATOR. THIS HASH
+               -- HAS IMPLICATIONS FOR VERSIONING, AUDIT, AND CONTRACT DISCOVERY.
+-              "6a9391d6aa51af28dd876ebb5565b69d1e83e5ac7861506bd29b56b0"
++              "2ed2631dbb277c84334453c5c437b86325d371f0835a28b910a91a6e"
+         ]
+     , testGroup "Payout Validator"
+         [
+@@ -217,7 +250,7 @@ tests =
+               -- DO NOT ALTER THE FOLLOWING VALUE UNLESS YOU ARE COMMITTING
+               -- APPROVED CHANGES TO MARLOWE'S ROLE VALIDATOR. THIS HASH HAS
+               -- IMPLICATIONS FOR VERSIONING, AUDIT, AND CONTRACT DISCOVERY.
+-              "49076eab20243dc9462511fb98a9cfb719f86e9692288139b7c91df3"
++              "e165610232235bbbbeff5b998b233daae42979dec92a6722d9cda989"
+         ]
+     ]
+
+@@ -260,19 +293,21 @@ check1Invalid _ allowEmptyList allowByteString allowUnit =
+
+ -- | Check that a semantics transaction succeeds.
+-checkSemanticsTransaction :: ArbitraryTransaction SemanticsTransaction ()      -- ^ Modifications to make before building the valid transaction.
++checkSemanticsTransaction :: [ReferencePath]                                   -- ^ The reference execution paths from which to choose.
++                          -> ArbitraryTransaction SemanticsTransaction ()      -- ^ Modifications to make before building the valid transaction.
+                           -> ArbitraryTransaction SemanticsTransaction ()      -- ^ Modifications to make after building the valid transaction.
+                           -> (PlutusTransaction SemanticsTransaction -> Bool)  -- ^ Whether to discard the transaction from the testing.
+                           -> Bool                                              -- ^ Whether the transaction should test as valid.
+                           -> Bool                                              -- ^ Whether to add noise to the script context.
++                          -> Bool                                              -- ^ Whether to allow merkleization.
+                           -> Property                                          -- ^ The test property.
+-checkSemanticsTransaction modifyBefore modifyAfter condition valid noisy =
++checkSemanticsTransaction referencePaths modifyBefore modifyAfter condition valid noisy allowMerkleization =
+   property
+-    . forAll (arbitrarySemanticsTransaction modifyBefore modifyAfter noisy `suchThat` condition)
++    . forAll (arbitrarySemanticsTransaction referencePaths modifyBefore modifyAfter noisy allowMerkleization `suchThat` condition)
+     $ \PlutusTransaction{..} ->
+       case evaluateSemantics (toData _datum) (toData _redeemer) (toData _scriptContext) of
+-        This  e   -> not valid || (error $ show e)
+-        These e l -> not valid || (error $ show e <> ": " <> show l)
++        This  e   -> not valid || error (show e)
++        These e l -> not valid || error (show e <> ": " <> show l)
+         That    _ -> valid
+
+@@ -288,14 +323,14 @@ checkPayoutTransaction modifyBefore modifyAfter condition valid noisy =
+     . forAll (arbitraryPayoutTransaction modifyBefore modifyAfter noisy `suchThat` condition)
+     $ \PlutusTransaction{..} ->
+       case evaluatePayout (toData _datum) (toData _redeemer) (toData _scriptContext) of
+-        This  e   -> not valid || (error $ show e)
+-        These e l -> not valid || (error $ show e <> ": " <> show l)
++        This  e   -> not valid || error (show e)
++        These e l -> not valid || error (show e <> ": " <> show l)
+         That    _ -> valid
+
+ -- | Check that validation fails if two Marlowe scripts are run.
+-checkDoubleInput :: Property
+-checkDoubleInput =
++checkDoubleInput :: [ReferencePath] -> Property
++checkDoubleInput referencePaths =
+   let
+     modifyAfter =
+       do
+@@ -313,7 +348,7 @@ checkDoubleInput =
+         infoData <>= AM.fromList [(inDatumHash, inDatum)]
+         shuffle
+   in
+-    checkSemanticsTransaction noModify modifyAfter noVeto False False
++    checkSemanticsTransaction referencePaths noModify modifyAfter noVeto False False False
+
+ -- | Split a value in half.
+@@ -341,9 +376,19 @@ notCloses :: PlutusTransaction SemanticsTransaction -> Bool
+ notCloses = not . doesClose
+
++-- | Ensure that the contract makes payments.
++hasPayouts :: PlutusTransaction SemanticsTransaction -> Bool
++hasPayouts =
++  let
++    isPayout (Payment _ (Party _) _ i) = i > 0
++    isPayout _ = False
++  in
++    any isPayout . txOutPayments . (^. output)
++
++
+ -- | Check that validation fails if there is more than one Marlowe output.
+-checkMultipleOutput :: Property
+-checkMultipleOutput =
++checkMultipleOutput :: [ReferencePath] -> Property
++checkMultipleOutput referencePaths =
+   let
+     modifyAfter =
+       do
+@@ -356,12 +401,12 @@ checkMultipleOutput =
+         infoOutputs %= concatMap splitOwnOutput
+         shuffle
+   in
+-    checkSemanticsTransaction noModify modifyAfter notCloses False False
++    checkSemanticsTransaction referencePaths noModify modifyAfter notCloses False False False
+
+ -- | Check that validation fails if there is one Marlowe output upon close.
+-checkCloseOutput :: Property
+-checkCloseOutput =
++checkCloseOutput :: [ReferencePath] -> Property
++checkCloseOutput referencePaths =
+   let
+     modifyAfter =
+       do
+@@ -374,12 +419,12 @@ checkCloseOutput =
+         infoOutputs <>= (txInInfoResolved <$> inScript)
+         shuffle
+   in
+-    checkSemanticsTransaction noModify modifyAfter doesClose False False
++    checkSemanticsTransaction referencePaths noModify modifyAfter doesClose False False False
+
+ -- | Check that value input to a script matches its input state.
+-checkValueInput :: Property
+-checkValueInput =
++checkValueInput :: [ReferencePath] -> Property
++checkValueInput referencePaths =
+   let
+     modifyAfter =
+       do
+@@ -391,29 +436,90 @@ checkValueInput =
+         -- Update the inputs with the incremented script input.
+         infoInputs %= fmap incrementOwnInput
+     in
+-      checkSemanticsTransaction noModify modifyAfter noVeto False False
++      checkSemanticsTransaction referencePaths noModify modifyAfter noVeto False False False
+
+--- | Check that value output to a script matches its output state.
+-checkValueOutput :: Property
+-checkValueOutput =
++-- | Check that value output to a script matches its expectation.
++checkValueOutput :: [ReferencePath] -> Property
++checkValueOutput referencePaths =
+   let
+     modifyAfter =
+       do
++        delta <- lift $ oneof [chooseInteger (-5, -1), chooseInteger (1, 5), arbitrary `suchThat` (/= 0)]  -- Ensure small non-zero integers.
+         let
+-          -- Add one lovelace to the output to the script.
++          -- Add or subtract some lovelace to the output to the script.
+           incrementOwnOutput txOut@(TxOut address value _ _)
+-            | address == semanticsAddress = txOut {txOutValue = value <> singleton adaSymbol adaToken 1}
+-            | otherwise                  = txOut
++            | address == semanticsAddress = txOut {txOutValue = value <> singleton adaSymbol adaToken delta}
++            | otherwise                   = txOut
+         -- Update the outputs with the incremented script output.
+         infoOutputs %= fmap incrementOwnOutput
+   in
+-    checkSemanticsTransaction noModify modifyAfter notCloses False False
++    checkSemanticsTransaction referencePaths noModify modifyAfter notCloses False False False
++
++
++-- | Check the consistency of the output value with the output state.
++checkOutputConsistency :: [ReferencePath] -> Property
++checkOutputConsistency referencePaths =
++  property
++    . forAll (arbitrarySemanticsTransaction referencePaths noModify noModify False True)
++    $ \tx ->
++      let
++        findOwnOutput (TxOut address value _ _)
++          | address == semanticsAddress = value
++          | otherwise                   = mempty
++        outValue = foldMap findOwnOutput $ tx ^. infoOutputs
++        finalBalance = totalBalance . accounts . txOutState $ tx ^. output
++        valid = outValue == finalBalance
++      in
++        checkSemanticsTransaction referencePaths noModify noModify notCloses valid False False
++
++
++-- | Add a duplicate entry to an assocation list.
++addDuplicate :: Arbitrary v => AM.Map k v -> Gen (AM.Map k v)
++addDuplicate am =
++  do
++    let
++      am' = AM.toList am
++    key <- elements $ fst <$> am'
++    value <- arbitrary
++    AM.fromList <$> Q.shuffle ((key, value) : am')
++
++
++-- | Check for the detection of duplicates in input state
++checkInputDuplicates :: [ReferencePath] -> Property
++checkInputDuplicates referencePaths =
++  let
++    hasDuplicates tx =
++      let
++        hasDuplicate am = length (AM.keys am) /= length (nub $ AM.keys am)
++        M.State{..} = tx ^. inputState
++      in
++           hasDuplicate accounts
++        || hasDuplicate choices
++        || hasDuplicate boundValues
++    makeDuplicates am =
++      if AM.null am
++        then pure am
++        else oneof [pure am, addDuplicate am]
++    modifyBefore =
++      do
++        M.State{..} <- use inputState
++        inputState <~
++          lift
++            (
++              M.State
++                <$> makeDuplicates accounts
++                <*> makeDuplicates choices
++                <*> makeDuplicates boundValues
++                <*> pure minTime
++            )
++  in
++    checkSemanticsTransaction referencePaths modifyBefore noModify hasDuplicates False False False
+
+ -- | Check that output datum to a script matches its semantic output.
+-checkDatumOutput :: (MarloweData -> Gen MarloweData) -> Property
+-checkDatumOutput perturb =
++checkDatumOutput :: [ReferencePath] -> (MarloweData -> Gen MarloweData) -> Property
++checkDatumOutput referencePaths perturb =
+   let
+     modifyAfter =
+       do
+@@ -432,13 +538,24 @@ checkDatumOutput perturb =
+         -- Update the data with the modification.
+         infoData %= AM.fromList . fmap perturbOwnOutputDatum . AM.toList
+   in
+-    checkSemanticsTransaction noModify modifyAfter notCloses False False
++    checkSemanticsTransaction referencePaths noModify modifyAfter notCloses False False False
++
++
++-- | Check that other validators are forbidden during payments.
++checkOtherValidators :: [ReferencePath] -> Property
++checkOtherValidators referencePaths =
++  let
++    modifyAfter =
++      -- Add an extra script input.
++      infoInputs <><~ lift (listOf1 $ arbitrary `suchThat` isScriptTxIn)
++  in
++    checkSemanticsTransaction referencePaths noModify modifyAfter hasPayouts False False False
+
+ -- | Check that parameters in the datum are not changed by the transaction.
+-checkParamsOutput :: Property
+-checkParamsOutput =
+-  checkDatumOutput
++checkParamsOutput :: [ReferencePath] -> Property
++checkParamsOutput referencePaths =
++  checkDatumOutput referencePaths
+     $ \marloweData ->
+       do
+         -- Replace the output parameters with a random one.
+@@ -448,9 +565,9 @@ checkParamsOutput =
+
+ -- | Check that state output to a script matches its semantic output.
+-checkStateOutput :: Property
+-checkStateOutput =
+-  checkDatumOutput
++checkStateOutput :: [ReferencePath] -> Property
++checkStateOutput referencePaths =
++  checkDatumOutput referencePaths
+     $ \marloweData ->
+       do
+         -- Replace the output state with a random one.
+@@ -460,9 +577,9 @@ checkStateOutput =
+
+ -- | Check that contract output to a script matches its semantic output.
+-checkContractOutput :: Property
+-checkContractOutput =
+-  checkDatumOutput
++checkContractOutput :: [ReferencePath] -> Property
++checkContractOutput referencePaths =
++  checkDatumOutput referencePaths
+     $ \marloweData ->
+       do
+         -- Replace the output ccontact with a random one.
+@@ -482,8 +599,8 @@ hasMerkleizedInput =
+
+ -- | Check than an invalid merkleization is rejected.
+-checkMerkleization :: Bool -> Property
+-checkMerkleization valid =
++checkMerkleization :: [ReferencePath] -> Bool -> Property
++checkMerkleization referencePaths valid =
+   let
+     -- Merkleizedd the contract and its input.
+     modifyBefore = merkleize
+@@ -499,12 +616,12 @@ checkMerkleization valid =
+                hashes <- input `uses` (concatMap merkleHash . txInputs)
+                infoData %= (AM.fromList . filter ((`notElem` hashes) . fst) . AM.toList)
+   in
+-    checkSemanticsTransaction modifyBefore modifyAfter hasMerkleizedInput valid False
++    checkSemanticsTransaction referencePaths modifyBefore modifyAfter hasMerkleizedInput valid False False
+
+ -- | Check that non-positive accounts are rejected.
+-checkPositiveAccounts :: Property
+-checkPositiveAccounts =
++checkPositiveAccounts :: [ReferencePath] -> Property
++checkPositiveAccounts referencePaths =
+   let
+     modifyBefore =
+       do
+@@ -515,7 +632,7 @@ checkPositiveAccounts =
+         -- Add the non-positive entry to the accounts.
+         inputState %= (\state -> state {accounts = AM.insert (account, token) amount' $ accounts state})
+   in
+-    checkSemanticsTransaction modifyBefore noModify noVeto False False
++    checkSemanticsTransaction referencePaths modifyBefore noModify noVeto False False False
+
+ -- | Compute the authorization for an input.
+@@ -547,8 +664,8 @@ deleteFirst f z =
+
+ -- | Check that a missing authorization causes failure.
+-checkAuthorization :: Property
+-checkAuthorization =
++checkAuthorization :: [ReferencePath] -> Property
++checkAuthorization referencePaths =
+   let
+     modifyAfter =
+       do
+@@ -565,7 +682,7 @@ checkAuthorization =
+         -- Remove the first PKH signatory.
+         infoSignatories %= deleteFirst matchPkh
+   in
+-    checkSemanticsTransaction noModify modifyAfter hasAuthorizations False False
++    checkSemanticsTransaction referencePaths noModify modifyAfter hasAuthorizations False False False
+
+ -- | Determine whether there are any external payments in a transaction.
+@@ -585,8 +702,8 @@ decrementValue = foldMap (\(c, n, i) -> singleton c n (i - 1)) . flattenValue
+
+ -- | Check that an insufficient payment causes failure.
+-checkPayment :: Property
+-checkPayment =
++checkPayment :: [ReferencePath] -> Property
++checkPayment referencePaths =
+   let
+     modifyAfter =
+       do
+@@ -600,7 +717,7 @@ checkPayment =
+         -- Update the outputs.
+         infoOutputs %= fmap decrementPayment
+   in
+-    checkSemanticsTransaction noModify modifyAfter hasExternalPayments False False
++    checkSemanticsTransaction referencePaths noModify modifyAfter hasExternalPayments False False False
+
+ -- | Remove a role input UTxOs from the transaction.
+diff --git a/marlowe-test/src/Spec/Marlowe/Plutus/Transaction.hs b/marlowe-test/src/Spec/Marlowe/Plutus/Transaction.hs
+index 9141d071f..88201e361 100644
+--- a/marlowe-test/src/Spec/Marlowe/Plutus/Transaction.hs
++++ b/marlowe-test/src/Spec/Marlowe/Plutus/Transaction.hs
+@@ -27,6 +27,7 @@ module Spec.Marlowe.Plutus.Transaction
+   , noModify
+   , shuffle
+     -- * Conditions
++  , isScriptTxIn
+   , noVeto
+   ) where
+
+@@ -61,7 +62,7 @@ import Plutus.Script.Utils.Scripts (datumHash)
+ import Plutus.V1.Ledger.Value (gt)
+ import Plutus.V2.Ledger.Api
+   ( Address(Address)
+-  , Credential(PubKeyCredential)
++  , Credential(..)
+   , CurrencySymbol
+   , Datum(..)
+   , DatumHash(..)
+@@ -76,7 +77,7 @@ import Plutus.V2.Ledger.Api
+   , ToData(toBuiltinData)
+   , TxInInfo(..)
+   , TxInfo(..)
+-  , TxOut(TxOut)
++  , TxOut(TxOut, txOutAddress)
+   , UpperBound(UpperBound)
+   , Value
+   )
+@@ -105,10 +106,11 @@ import Spec.Marlowe.Plutus.Types
+   , role
+   , scriptPurpose
+   )
++import Spec.Marlowe.Reference (ReferencePath, arbitraryReferenceTransaction)
+ import Spec.Marlowe.Semantics.Arbitrary (arbitraryGoldenTransaction, arbitraryPositiveInteger)
+ import Spec.Marlowe.Semantics.Golden (GoldenTransaction)
+ import Spec.Marlowe.Semantics.Merkle (deepMerkleize, merkleizeInputs)
+-import Test.Tasty.QuickCheck (Arbitrary(..), Gen, elements, suchThat)
++import Test.Tasty.QuickCheck (Arbitrary(..), Gen, elements, frequency, listOf, suchThat)
+
+ import qualified Language.Marlowe.Core.V1.Semantics.Types as M (Party(Address))
+ import qualified Plutus.V1.Ledger.Value as V (adaSymbol, adaToken, singleton)
+@@ -185,11 +187,15 @@ makeDeposit :: Input
+ makeDeposit input' =
+   do
+     ref <- lift arbitrary
+-    address' <- lift arbitrary
++    address' <- lift $ arbitrary `suchThat` notScriptAddress
+     pure
+       $ case getInputContent input' of
+-          IDeposit _ (M.Address _ address) (Token c n) i -> pure . TxInInfo ref $ TxOut address  (V.singleton c n i) NoOutputDatum  Nothing
+-          IDeposit _ (Role _             ) (Token c n) i -> pure . TxInInfo ref $ TxOut address' (V.singleton c n i) NoOutputDatum  Nothing
++          IDeposit _ (M.Address _ address) (Token c n) i -> if i > 0
++                                                              then pure . TxInInfo ref $ TxOut address  (V.singleton c n i) NoOutputDatum  Nothing
++                                                              else mempty
++          IDeposit _ (Role _             ) (Token c n) i -> if i > 0
++                                                              then pure . TxInInfo ref $ TxOut address' (V.singleton c n i) NoOutputDatum  Nothing
++                                                              else mempty
+           _                                              -> mempty
+
+ -- | Create role input for a Marlowe semantics transaction.
+@@ -199,7 +205,7 @@ makeRoleIn input' =
+   do
+     MarloweParams currencySymbol <- use marloweParams
+     ref <- lift arbitrary
+-    address  <- lift arbitrary
++    address  <- lift $ arbitrary `suchThat` notScriptAddress
+     pure
+       $ case getInputContent input' of
+           IDeposit _ (Role role') _ _         -> pure . TxInInfo ref $ TxOut address (V.singleton currencySymbol role' 1) NoOutputDatum Nothing
+@@ -276,6 +282,21 @@ makeSpendSignatory (TxInInfo _ (TxOut (Address (PubKeyCredential pkh) _) _ _ _))
+ makeSpendSignatory _                                                             = mempty
+
++-- | Combine payments with the same address and hash.
++consolidatePayments :: [TxOut] -> [TxOut]
++consolidatePayments ps =
++  let
++    extractAddressHash (TxOut address _ hash _) = (address, hash)
++    extractValue (TxOut _ value _ _) = value
++  in
++    [
++      TxOut address value hash Nothing
++    |
++      ah@(address, hash) <- nub $ extractAddressHash <$> ps
++    , let value = foldMap extractValue $ filter ((== ah) . extractAddressHash) ps
++    ]
++
++
+ -- | Generate a valid Marlowe semantics transaction.
+ validSemanticsTransaction :: Bool                                          -- ^ Whether to add noise to the script context.
+                           -> ArbitraryTransaction SemanticsTransaction ()  -- ^ The generator.
+@@ -313,7 +334,7 @@ validSemanticsTransaction noisy =
+     MarloweParams currencySymbol <- use marloweParams
+     -- Add the payments.
+     (payments, paymentData) <- fmap (bimap mconcat mconcat . unzip) . mapM (makePayment currencySymbol) =<< (output `uses` txOutPayments)
+-    infoOutputs <>= payments
++    infoOutputs <>= consolidatePayments payments
+     infoData <>= AM.fromList paymentData
+
+     -- Add the signatories.
+@@ -336,13 +357,20 @@ validSemanticsTransaction noisy =
+
+ -- | Generate an arbitrary, valid Marlowe semantics transaction: datum, redeemer, and script context.
+-arbitrarySemanticsTransaction :: ArbitraryTransaction SemanticsTransaction ()  -- ^ Modifications to make before building the valid transaction.
++arbitrarySemanticsTransaction :: [ReferencePath]                               -- ^ The reference execution paths from which to choose.
++                              -> ArbitraryTransaction SemanticsTransaction ()  -- ^ Modifications to make before building the valid transaction.
+                               -> ArbitraryTransaction SemanticsTransaction ()  -- ^ Modifications to make after building the valid transaction.
+                               -> Bool                                          -- ^ Whether to add noise to the script context.
++                              -> Bool                                          -- ^ Whether to allow merkleization.
+                               -> Gen (PlutusTransaction SemanticsTransaction)  -- ^ The generator.
+-arbitrarySemanticsTransaction modifyBefore modifyAfter noisy =
++arbitrarySemanticsTransaction referencePaths modifyBefore modifyAfter noisy allowMerkleization =
+   do
+-    golden <- arbitraryGoldenTransaction
++    golden <-
++      frequency
++        [
++          (1, arbitraryGoldenTransaction allowMerkleization)  -- Manually vetted transactions.
++        , (5, arbitraryReferenceTransaction referencePaths)   -- Transactions generated using `getAllInputs` and `computeTransaction`.
++        ]
+     start <- bareSemanticsTransaction golden
+     (modifyBefore >> validSemanticsTransaction noisy >> modifyAfter)
+       `execStateT` start
+@@ -377,7 +405,7 @@ makePayoutRoleIn :: ArbitraryTransaction PayoutTransaction TxInInfo
+ makePayoutRoleIn =
+   do
+     ref <- lift arbitrary
+-    address <- lift arbitrary
++    address <- lift $ arbitrary `suchThat` notScriptAddress
+     value <- V.singleton <$> marloweParamsPayout `uses` rolesCurrency <*> use role <*> pure 1
+     pure
+       . TxInInfo ref
+@@ -445,15 +473,50 @@ validPayoutTransaction noisy =
+     infoFee <~ V.singleton V.adaSymbol V.adaToken <$> lift arbitraryPositiveInteger
+
+     -- Add noise.
+-    when noisy addNoise
++    when noisy addNoise'
+
+     -- Shuffle.
+     shuffle
+
++-- | Check that an address is not for a script.
++notScriptAddress :: Address -> Bool
++notScriptAddress (Address (ScriptCredential _) _) = False
++notScriptAddress _ = True
++
++
++-- | Check that an input is not from a script.
++notScriptTxIn :: TxInInfo -> Bool
++notScriptTxIn = not . isScriptTxIn
++
++
++-- | Check that an input is from a script.
++isScriptTxIn :: TxInInfo -> Bool
++isScriptTxIn TxInInfo{txInInfoResolved=TxOut{txOutAddress=Address (ScriptCredential _) _}} = True
++isScriptTxIn _ = False
++
++
+ -- | Add noise to the inputs, outputs, and data in a Plutus transaction.
+-addNoise :: ArbitraryTransaction a ()
++addNoise :: ArbitraryTransaction SemanticsTransaction ()
+ addNoise =
++  do
++    let
++      isPayout (Payment _ (Party _) _ i) = i > 0
++      isPayout _ = False
++    hasPayments <- any isPayout . txOutPayments <$> use output
++    let
++      arbitraryInput =
++        if hasPayments
++          then arbitrary `suchThat` notScriptTxIn
++          else arbitrary
++    infoInputs  <><~ lift (listOf arbitraryInput `suchThat` ((< 5) . length))
++    infoOutputs <><~ lift (arbitrary `suchThat` ((< 5) . length))
++    infoData    <><~ lift (fmap AM.fromList $ arbitrary `suchThat` ((< 5) . length))
++
++
++-- | Add noise to the inputs, outputs, and data in a Plutus transaction.
++addNoise' :: ArbitraryTransaction PayoutTransaction ()
++addNoise' =
+   do
+     infoInputs  <><~ lift (arbitrary `suchThat` ((< 5) . length))
+     infoOutputs <><~ lift (arbitrary `suchThat` ((< 5) . length))
+diff --git a/marlowe-test/src/Spec/Marlowe/Reference.hs b/marlowe-test/src/Spec/Marlowe/Reference.hs
+new file mode 100644
+index 000000000..4c3291e6a
+--- /dev/null
++++ b/marlowe-test/src/Spec/Marlowe/Reference.hs
+@@ -0,0 +1,156 @@
++-----------------------------------------------------------------------------
++--
++-- Module      :  $Headers
++-- License     :  Apache 2.0
++--
++-- Stability   :  Experimental
++-- Portability :  Portable
++--
++-- | Function to generate all valid transactions for contracts in JSON files.
++--
++-----------------------------------------------------------------------------
++
++
++{-# LANGUAGE DeriveGeneric #-}
++{-# LANGUAGE FlexibleContexts #-}
++{-# LANGUAGE LambdaCase #-}
++{-# LANGUAGE NumericUnderscores #-}
++{-# LANGUAGE OverloadedStrings #-}
++{-# LANGUAGE RecordWildCards #-}
++
++
++module Spec.Marlowe.Reference
++  ( -- * Types
++    ReferencePath(..)
++  , ReferenceTransaction(..)
++    -- * Analysis
++  , processContract
++    -- * Testing
++  , arbitraryReferenceTransaction
++  , readReferencePaths
++  , referenceFolder
++  ) where
++
++
++import Control.Monad (forM)
++import Control.Monad.Except (ExceptT(..), lift, throwError)
++import Data.Aeson (FromJSON, ToJSON, eitherDecodeFileStrict, encodeFile)
++import Data.Bifunctor (first)
++import Data.List (isSuffixOf)
++import GHC.Generics (Generic)
++import Language.Marlowe.Core.V1.Semantics (TransactionInput, TransactionOutput(..), computeTransaction)
++import Language.Marlowe.Core.V1.Semantics.Types (Contract, Party(Role), State(..), Token(..))
++import Language.Marlowe.FindInputs (getAllInputs)
++import Plutus.V2.Ledger.Api (POSIXTime)
++import Spec.Marlowe.Semantics.Golden (GoldenTransaction)
++import System.Directory (listDirectory)
++import System.FilePath ((</>))
++import Test.Tasty.QuickCheck (Gen, elements)
++
++import qualified PlutusTx.AssocMap as AM (empty, singleton)
++
++
++referenceFolder :: FilePath
++referenceFolder = "reference" </> "data"
++
++
++readReferencePaths :: IO [ReferencePath]
++readReferencePaths =
++  do
++    pathFiles <- fmap (referenceFolder </>) . filter (".paths" `isSuffixOf`) <$> listDirectory referenceFolder
++    fmap concat
++      . forM pathFiles
++      $ \pathFile ->
++        eitherDecodeFileStrict pathFile
++          >>= \case
++            Right paths -> pure $ filter (not . null . transactions) paths
++            Left msg -> error $ "Failed parsing " <> pathFile <> ": " <> msg <> "."
++
++
++arbitraryReferenceTransaction :: [ReferencePath] -> Gen GoldenTransaction
++arbitraryReferenceTransaction paths =
++  do
++    ReferencePath{..} <- elements paths
++    if length transactions > 1
++      then do
++             (ReferenceTransaction _ prior, ReferenceTransaction{..}) <- elements $ zip (init transactions) (tail transactions)
++             pure (txOutState prior, txOutContract prior, input, output)
++      else let
++             ReferenceTransaction{..} = head transactions
++            in
++              pure (state, contract, input, output)
++
++
++data ReferencePath =
++  ReferencePath
++  {
++    contract :: Contract
++  , state :: State
++  , transactions :: [ReferenceTransaction]
++  }
++    deriving (Generic, Show)
++
++instance FromJSON ReferencePath
++
++instance ToJSON ReferencePath
++
++
++data ReferenceTransaction =
++  ReferenceTransaction
++  {
++    input :: TransactionInput
++  , output :: TransactionOutput
++  }
++    deriving (Generic, Show)
++
++instance FromJSON ReferenceTransaction
++
++instance ToJSON ReferenceTransaction
++
++
++processContract
++  :: FilePath
++  -> FilePath
++  -> ExceptT String IO ()
++processContract contractFile pathsFile =
++  do
++    contract <- ExceptT $ first show <$> eitherDecodeFileStrict contractFile
++    traces <- ExceptT $ first show <$> getAllInputs contract
++    paths <- runTransactions contract `mapM` traces
++    lift $ encodeFile pathsFile paths
++
++
++runTransactions
++  :: Contract
++  -> (POSIXTime, [TransactionInput])
++  -> ExceptT String IO ReferencePath
++runTransactions contract (startTime, inputs) =
++  do
++    let
++      state = makeState startTime
++    transactions <- runTransaction contract state inputs
++    pure ReferencePath{..}
++
++
++runTransaction
++  :: Contract
++  -> State
++  -> [TransactionInput]
++  -> ExceptT String IO [ReferenceTransaction]
++runTransaction _ _ [] = pure []
++runTransaction contract state (input : inputs) =
++  case computeTransaction input state contract of
++    Error err -> throwError $ show err
++    output@TransactionOutput{..} -> (ReferenceTransaction{..} :) <$> runTransaction txOutContract txOutState inputs
++
++
++makeState
++  :: POSIXTime
++  -> State
++makeState minTime =
++  let
++    accounts = AM.singleton (Role "", Token "" "") 30_000_000  -- Note that 30 ada exceeds min-UTxO for current protocol parameters.
++    choices = AM.empty
++    boundValues = AM.empty
++  in
++    State{..}
+diff --git a/marlowe-test/src/Spec/Marlowe/Semantics.hs b/marlowe-test/src/Spec/Marlowe/Semantics.hs
+index dfd167f53..a15face41 100644
+--- a/marlowe-test/src/Spec/Marlowe/Semantics.hs
++++ b/marlowe-test/src/Spec/Marlowe/Semantics.hs
+@@ -24,6 +24,7 @@ import qualified Spec.Marlowe.Semantics.Entropy (tests)
+ import qualified Spec.Marlowe.Semantics.Functions (tests)
+ import qualified Spec.Marlowe.Semantics.Golden (tests)
+
++
+ -- | Run the tests.
+ tests :: TestTree
+ tests =
+diff --git a/marlowe-test/src/Spec/Marlowe/Semantics/Arbitrary.hs b/marlowe-test/src/Spec/Marlowe/Semantics/Arbitrary.hs
+index b05563b29..3ab7ab9df 100644
+--- a/marlowe-test/src/Spec/Marlowe/Semantics/Arbitrary.hs
++++ b/marlowe-test/src/Spec/Marlowe/Semantics/Arbitrary.hs
+@@ -22,7 +22,8 @@
+
+ module Spec.Marlowe.Semantics.Arbitrary
+   ( -- * Types
+-    IsValid(..)
++    Context
++  , IsValid(..)
+   , SemiArbitrary(..)
+     -- * Generators
+   , arbitraryAssocMap
+@@ -30,6 +31,7 @@ module Spec.Marlowe.Semantics.Arbitrary
+   , arbitraryContractWeighted
+   , arbitraryFibonacci
+   , arbitraryGoldenTransaction
++  , arbitraryNonnegativeInteger
+   , arbitraryPositiveInteger
+   , arbitraryTimeIntervalAround
+   , arbitraryValidInput
+@@ -54,7 +56,15 @@ import Control.Monad (replicateM)
+ import Data.Function (on)
+ import Data.List (nub, nubBy)
+ import Language.Marlowe.Core.V1.Semantics
+-  (TransactionInput(..), TransactionOutput(..), computeTransaction, evalObservation, evalValue)
++  ( Payment(..)
++  , TransactionError(..)
++  , TransactionInput(..)
++  , TransactionOutput(..)
++  , TransactionWarning(..)
++  , computeTransaction
++  , evalObservation
++  , evalValue
++  )
+ import Language.Marlowe.Core.V1.Semantics.Types
+   ( Accounts
+   , Action(..)
+@@ -67,6 +77,7 @@ import Language.Marlowe.Core.V1.Semantics.Types
+   , Environment(..)
+   , Input(..)
+   , InputContent(..)
++  , IntervalError(..)
+   , Observation(..)
+   , Party(..)
+   , Payee(..)
+@@ -77,6 +88,7 @@ import Language.Marlowe.Core.V1.Semantics.Types
+   , ValueId(..)
+   , getAction
+   )
++import Plutus.Script.Utils.Scripts (dataHash)
+ import Plutus.V2.Ledger.Api
+   ( Credential(..)
+   , CurrencySymbol(..)
+@@ -87,11 +99,13 @@ import Plutus.V2.Ledger.Api
+   , ValidatorHash(..)
+   , adaSymbol
+   , adaToken
++  , toBuiltinData
+   )
+ import PlutusTx.Builtins (BuiltinByteString, appendByteString, lengthOfByteString, sliceByteString)
+ import Spec.Marlowe.Semantics.Golden (GoldenTransaction, goldenContracts, goldenTransactions)
++import Spec.Marlowe.Semantics.Merkle (merkleizeInputs, shallowMerkleize)
+ import Test.Tasty.QuickCheck
+-  (Arbitrary(..), Gen, chooseInteger, elements, frequency, listOf, shrinkList, sized, suchThat, vectorOf)
++  (Arbitrary(..), Gen, chooseInteger, elements, frequency, listOf, oneof, shrinkList, sized, suchThat, vectorOf)
+
+ import qualified Plutus.V2.Ledger.Api as Ledger (Address(..))
+ import qualified PlutusTx.AssocMap as AM (Map, delete, empty, fromList, keys, toList)
+@@ -880,11 +894,11 @@ goldenContract = (,) <$> elements goldenContracts <*> pure (State AM.empty AM.em
+
+ instance Arbitrary Contract where
+   arbitrary = frequency [(95, semiArbitrary =<< arbitrary), (5, fst <$> goldenContract)]
+-  shrink (Pay a p t x c) = [Pay a' p t x c | a' <- shrink a] ++ [Pay a p' t x c | p' <- shrink p] ++ [Pay a p t' x c | t' <- shrink t] ++ [Pay a p t x' c | x' <- shrink x] ++ [Pay a p t x c' | c' <- shrink c]
+-  shrink (If o x y) = [If o' x y | o' <- shrink o] ++ [If o x' y | x' <- shrink x] ++ [If o x y' | y' <- shrink y]
+-  shrink (When a t c) = [When a' t c | a' <- shrink a] ++ [When a t' c | t' <- shrink t] ++ [When a t c' | c' <- shrink c]
+-  shrink (Let v x c) = [Let v' x c | v' <- shrink v] ++ [Let v x' c | x' <- shrink x] ++ [Let v x c' | c' <- shrink c]
+-  shrink (Assert o c) = [Assert o' c | o' <- shrink o] ++ [Assert o c' | c' <- shrink c]
++  shrink (Pay a p t x c) = [c] ++ [Pay a' p t x c | a' <- shrink a] ++ [Pay a p' t x c | p' <- shrink p] ++ [Pay a p t' x c | t' <- shrink t] ++ [Pay a p t x' c | x' <- shrink x] ++ [Pay a p t x c' | c' <- shrink c]
++  shrink (If o x y) = [x, y] ++ [If o' x y | o' <- shrink o] ++ [If o x' y | x' <- shrink x] ++ [If o x y' | y' <- shrink y]
++  shrink (When a t c) = [c] ++ [When a' t c | a' <- shrink a] ++ [When a t' c | t' <- shrink t] ++ [When a t c' | c' <- shrink c]
++  shrink (Let v x c) = [c] ++ [Let v' x c | v' <- shrink v] ++ [Let v x' c | x' <- shrink x] ++ [Let v x c' | c' <- shrink c]
++  shrink (Assert o c) = [c] ++ [Assert o' c | o' <- shrink o] ++ [Assert o c' | c' <- shrink c]
+   shrink _ = []
+
+@@ -1047,10 +1061,22 @@ instance SemiArbitrary InputContent where
+ instance Arbitrary Input where
+   arbitrary = semiArbitrary =<< arbitrary
+   shrink (NormalInput i)         = NormalInput <$> shrink i
+-  shrink (MerkleizedInput i b c) = [MerkleizedInput i' b c | i' <- shrink i]
++  shrink (MerkleizedInput i b c) =
++    [NormalInput i]
++      <> [MerkleizedInput i' b c | i' <- shrink i]
++      <> [MerkleizedInput i (dataHash $ toBuiltinData c) c' | c' <- shrink c]
+
+ instance SemiArbitrary Input where
+-  semiArbitrary context = NormalInput <$> semiArbitrary context
++  semiArbitrary context =
++    frequency
++      [
++        (9, NormalInput <$> semiArbitrary context)
++      , (1, do
++              input <- semiArbitrary context
++              contract <- semiArbitrary context
++              pure $ MerkleizedInput input (dataHash $ toBuiltinData contract) contract
++        )
++      ]
+
+ instance Arbitrary TransactionInput where
+@@ -1073,7 +1099,7 @@ arbitraryValidStep :: State                 -- ^ The state of the contract.
+                    -> Gen TransactionInput  -- ^ Generator for a transaction input for a single step.
+ arbitraryValidStep _ (When [] timeout _) =
+   TransactionInput <$> arbitraryTimeIntervalAfter timeout <*> pure []
+-arbitraryValidStep state@State{..} (When cases timeout _) =
++arbitraryValidStep state@State{..} contract@(When cases timeout _) =
+   do
+     let
+       isEmptyChoice (Choice _ []) = True
+@@ -1090,7 +1116,13 @@ arbitraryValidStep state@State{..} (When cases timeout _) =
+                                          Bound lower upper <- elements bs
+                                          IChoice n <$> chooseInteger (lower, upper)
+                     Notify _        -> pure INotify
+-             pure $ TransactionInput times [NormalInput i]
++             is <-
++               frequency
++                 [
++                   (9, pure [NormalInput i])
++                 , (1, pure [MerkleizedInput i (dataHash $ toBuiltinData contract) contract])
++                 ]
++             pure $ TransactionInput times is
+ arbitraryValidStep State{minTime} contract =
+ {-
+   NOTE: Alternatively, if semantics should allow applying `[]` to a non-quiescent contract
+@@ -1144,10 +1176,66 @@ arbitraryValidInputs state contract =
+
+ -- | Generate an arbitrary golden transaction.
+-arbitraryGoldenTransaction :: Gen GoldenTransaction
+-arbitraryGoldenTransaction =
++arbitraryGoldenTransaction :: Bool -> Gen GoldenTransaction
++arbitraryGoldenTransaction allowMerkleization =
+   do
++    let
++      perhapsMerkleize gt@(state, contract, input, _) =
++        let
++          (contract', continuations) = shallowMerkleize contract
++          input' = merkleizeInputs continuations state contract' input
++        in
++          case input' of
++            Nothing      -> pure gt
++            Just input'' -> frequency [(9, pure gt), (1, pure (state, contract', input'', computeTransaction input'' state contract'))]
+     equalContractWeights <- frequency [(1, pure True), (5, pure False)]
+-    if equalContractWeights
+-      then elements =<< elements goldenTransactions
+-      else elements $ concat goldenTransactions
++    (if allowMerkleization then perhapsMerkleize else pure)
++      =<< if equalContractWeights
++            then elements =<< elements goldenTransactions
++            else elements $ concat goldenTransactions
++
++
++instance Arbitrary Payment where
++  arbitrary = Payment <$> arbitrary <*> arbitrary <*> arbitrary <*> arbitrary
++
++
++instance Arbitrary IntervalError where
++  arbitrary =
++    oneof
++      [
++        InvalidInterval <$> arbitrary
++      , IntervalInPastError <$> arbitrary <*> arbitrary
++      ]
++
++
++instance Arbitrary TransactionError where
++  arbitrary =
++    oneof
++      [
++        pure TEAmbiguousTimeIntervalError
++      , pure TEApplyNoMatchError
++      , TEIntervalError <$> arbitrary
++      , pure TEUselessTransaction
++      , pure TEHashMismatch
++      ]
++
++
++instance Arbitrary TransactionWarning where
++  arbitrary =
++    oneof
++      [
++        TransactionNonPositiveDeposit <$> arbitrary <*> arbitrary <*> arbitrary <*> arbitrary
++      , TransactionNonPositivePay <$> arbitrary <*> arbitrary <*> arbitrary <*> arbitrary
++      , TransactionPartialPay <$> arbitrary <*> arbitrary <*> arbitrary <*> arbitrary <*> arbitrary
++      , TransactionShadowing <$> arbitrary <*> arbitrary <*> arbitrary
++      , pure TransactionAssertionFailed
++      ]
++
++
++instance Arbitrary TransactionOutput where
++  arbitrary =
++    oneof
++      [
++        TransactionOutput <$> arbitrary <*> arbitrary <*> arbitrary <*> arbitrary
++      , Error <$> arbitrary
++      ]
+diff --git a/marlowe-test/src/Spec/Marlowe/Semantics/Compute.hs b/marlowe-test/src/Spec/Marlowe/Semantics/Compute.hs
+index 402460ed2..7aa9ed803 100644
+--- a/marlowe-test/src/Spec/Marlowe/Semantics/Compute.hs
++++ b/marlowe-test/src/Spec/Marlowe/Semantics/Compute.hs
+@@ -13,6 +13,7 @@
+
+ {-# LANGUAGE ConstraintKinds #-}
+ {-# LANGUAGE FlexibleContexts #-}
++{-# LANGUAGE LambdaCase #-}
+ {-# LANGUAGE RankNTypes #-}
+ {-# LANGUAGE RecordWildCards #-}
+
+@@ -31,6 +32,7 @@ import Control.Applicative (liftA2)
+ import Control.Lens.Getter (Getter, to, view)
+ import Control.Monad.Except (MonadError(throwError), unless, when)
+ import Control.Monad.Reader (ReaderT(runReaderT))
++import Data.Bifunctor (second)
+ import Data.Default (Default(..))
+ import Data.Function (on)
+ import Data.List (sort)
+@@ -49,29 +51,34 @@ import Language.Marlowe.Core.V1.Semantics.Types
+   ( AccountId
+   , Accounts
+   , Action(Choice, Deposit, Notify)
+-  , Case
++  , Bound(..)
++  , Case(..)
+   , ChoiceId
+   , ChosenNum
+   , Contract(..)
+   , Environment(Environment)
+-  , Input
++  , Input(..)
+   , InputContent(IChoice, IDeposit, INotify)
+   , IntervalError(IntervalInPastError, InvalidInterval)
+-  , Observation
++  , Observation(TrueObs)
+   , Payee(Party)
+   , State(..)
+   , TimeInterval
+   , Token(..)
+-  , Value
++  , Value(Constant)
+   , ValueId
+   , getAction
+   , getInputContent
+   )
+ import Language.Marlowe.FindInputs (getAllInputs)
+-import Plutus.V2.Ledger.Api (CurrencySymbol, POSIXTime(..), TokenName)
++import Plutus.Script.Utils.Scripts (datumHash)
++import Plutus.V2.Ledger.Api (CurrencySymbol, Datum(..), DatumHash(..), POSIXTime(..), TokenName, toBuiltinData)
+ import Spec.Marlowe.Semantics.Arbitrary
+-  ( SemiArbitrary(semiArbitrary)
++  ( Context
++  , SemiArbitrary(semiArbitrary)
+   , arbitraryContractWeighted
++  , arbitraryGoldenTransaction
++  , arbitraryPositiveInteger
+   , assertContractWeights
+   , closeContractWeights
+   , defaultContractWeights
+@@ -85,7 +92,19 @@ import Spec.Marlowe.Semantics.Orphans ()
+ import System.IO.Unsafe (unsafePerformIO)
+ import Test.Tasty (TestTree, testGroup)
+ import Test.Tasty.QuickCheck
+-  (Arbitrary(..), Gen, Testable(property), discard, elements, forAll, forAllShrink, frequency, suchThat, testProperty)
++  ( Arbitrary(..)
++  , Gen
++  , Testable(property)
++  , chooseInteger
++  , discard
++  , elements
++  , forAll
++  , forAllShrink
++  , frequency
++  , shuffle
++  , suchThat
++  , testProperty
++  )
+
+ import qualified PlutusTx.AssocMap as AM
+
+@@ -161,18 +180,145 @@ arbitraryMarloweContext w =
+ -- | Generate an arbitrary valid Marlowe transaction context.
+ arbitraryValid :: Gen MarloweContext
+ arbitraryValid =
+-    do
+-      mcContract <- arbitrary `suchThat` (/= Close)
+-      (time, inputs') <-
+-        case unsafePerformIO $ getAllInputs mcContract of
+-          Right candidates -> elements candidates
+-          Left _           -> discard
+-      let
+-        -- TODO: Generalize to arbitrary starting state.
+-        mcState = State AM.empty AM.empty AM.empty time
+-        mcInput = head inputs'
+-        mcOutput = computeTransaction mcInput mcState mcContract
+-      pure MarloweContext{..}
++  let
++    randomContext =
++      do
++        mcContract <- arbitrary `suchThat` (/= Close)
++        (time, inputs') <-
++          case unsafePerformIO $ getAllInputs mcContract of
++            Right candidates -> elements candidates
++            Left _           -> discard
++        let
++          -- TODO: Generalize to arbitrary starting state.
++          mcState = State AM.empty AM.empty AM.empty time
++          mcInput = head inputs'
++          mcOutput = computeTransaction mcInput mcState mcContract
++        pure MarloweContext{..}
++    goldenContext =
++      do
++        (mcState, mcContract, mcInput, mcOutput) <- arbitraryGoldenTransaction True
++        pure MarloweContext{..}
++  in
++    frequency [(1, randomContext), (2, goldenContext)]
++
++
++-- | Generate a simple valid transaction.
++simpleTransaction
++  :: Context
++  -> (TimeInterval -> Contract -> Gen Contract)
++  -> (State -> Gen State)
++  -> (TransactionInput -> Gen TransactionInput)
++  -> Gen MarloweContext
++simpleTransaction context modifyContract modifyState modifyInput =
++  do
++    mcState <- modifyState =<< semiArbitrary context
++    intervalStart <- (getPOSIXTime (minTime mcState) +) <$> arbitraryPositiveInteger
++    intervalEnd <- (intervalStart +) <$> arbitraryPositiveInteger
++    let
++      interval = (POSIXTime intervalStart, POSIXTime intervalEnd)
++    timeout <- (intervalEnd +) <$> arbitraryPositiveInteger
++    mcContract <- modifyContract interval . When [] (POSIXTime timeout) =<< semiArbitrary context
++    mcInput <- modifyInput $ TransactionInput interval []
++    let
++      mcOutput = computeTransaction mcInput mcState mcContract
++    pure MarloweContext{..}
++
++
++-- | Generate a simple payment.
++simplePayment :: Gen MarloweContext
++simplePayment =
++  do
++    context <- arbitrary
++    balance <- arbitraryPositiveInteger
++    payment <- chooseInteger (1, balance)
++    account <- semiArbitrary context
++    payee <- semiArbitrary context
++    token <- semiArbitrary context
++    let
++      modifyContract _ contract = pure $ Pay account (Party payee) token (Constant payment) contract
++      modifyState state =
++        do
++          accounts' <-
++            fmap AM.fromList . shuffle . AM.toList
++              . AM.insert (account, token) balance
++              . AM.delete (account, token)
++              $ accounts state
++          pure state {accounts = accounts'}
++    simpleTransaction context modifyContract modifyState pure
++
++
++-- | Generate a simple deposit.
++simpleDeposit :: Gen MarloweContext
++simpleDeposit =
++  do
++    context <- arbitrary
++    deposit <- arbitraryPositiveInteger
++    account <- semiArbitrary context
++    payee <- semiArbitrary context
++    token <- semiArbitrary context
++    let
++      modifyContract (_, POSIXTime intervalEnd) contract =
++        pure
++          $ When [Case (Deposit account payee token $ Constant deposit) contract]
++            (POSIXTime $ intervalEnd + 1) contract
++      modifyInput (TransactionInput interval _) = pure $ TransactionInput interval [NormalInput $ IDeposit account payee token deposit]
++    simpleTransaction context modifyContract pure modifyInput
++
++
++-- | Generate a simple notification.
++simpleChoice :: Gen MarloweContext
++simpleChoice =
++  do
++    context <- arbitrary
++    choiceId <- semiArbitrary context
++    value <- semiArbitrary context
++    bound <- Bound <$> ((value -) <$> arbitraryPositiveInteger) <*> ((value +) <$> arbitraryPositiveInteger)
++    let
++      modifyContract (_, POSIXTime intervalEnd) contract =
++        do
++          let
++            timeout = POSIXTime $ intervalEnd + 1
++          contract' <- When [] timeout <$> semiArbitrary context
++          pure $ When [Case (Choice choiceId [bound]) contract'] timeout contract
++      modifyInput (TransactionInput interval _) = pure $ TransactionInput interval [NormalInput $ IChoice choiceId value]
++    simpleTransaction context modifyContract pure modifyInput
++
++
++-- | Generate a simple notification.
++simpleNotify :: Gen MarloweContext
++simpleNotify =
++  do
++    context <- arbitrary
++    let
++      modifyContract (_, POSIXTime intervalEnd) contract =
++        do
++          let
++            timeout = POSIXTime $ intervalEnd + 1
++          contract' <- When [] timeout <$> semiArbitrary context
++          pure $ When [Case (Notify TrueObs) contract'] timeout contract
++      modifyInput (TransactionInput interval _) = pure $ TransactionInput interval [NormalInput INotify]
++    simpleTransaction context modifyContract pure modifyInput
++
++
++-- | Generate a simple notification.
++simpleMerkleization :: Gen MarloweContext
++simpleMerkleization =
++  do
++    context <- arbitrary
++    mcState <- semiArbitrary context
++    intervalStart <- (getPOSIXTime (minTime mcState) +) <$> arbitraryPositiveInteger
++    intervalEnd <- (intervalStart +) <$> arbitraryPositiveInteger
++    let
++      interval = (POSIXTime intervalStart, POSIXTime intervalEnd)
++    timeout <- (intervalEnd +) <$> arbitraryPositiveInteger
++    contract <- When [] (POSIXTime timeout) <$> semiArbitrary context
++    let
++      DatumHash hash = datumHash . Datum $ toBuiltinData contract
++      mcInput = TransactionInput interval [MerkleizedInput INotify hash contract]
++    mcContract <- When [MerkleizedCase (Notify TrueObs) hash] (POSIXTime timeout) <$> semiArbitrary context
++    let
++      mcOutput = computeTransaction mcInput mcState mcContract
++    pure MarloweContext{..}
+
+ -- | Recompute the output of a Marlowe transaction in an transaction context.
+@@ -543,6 +689,27 @@ requireAmbiguousTimeout =
+   >> requireNextTimeout `requireLE` view latestTime
+
++-- | Require a payment to a party.
++requirePayout :: Testify ()
++requirePayout =
++  let
++    isPayout (Payment _ (Party _) _ i) = i > 0
++    isPayout _ = False
++  in
++    view payments >>= "Positive payout" `require` any isPayout
++
++
++-- | Require not deposits.
++requireNoDeposits :: Testify ()
++requireNoDeposits =
++  let
++    isDeposit (NormalInput (IDeposit _ _ _ i)) = i > 0
++    isDeposit (MerkleizedInput (IDeposit _ _ _ i) _ _) = i > 0
++    isDeposit _ = False
++  in
++    view inputs >>= "No deposits" `require` (not . any isDeposit)
++
++
+ -- | Throw an error unless a condition holds.
+ throwUnless :: MonadError String m
+             => String
+@@ -671,30 +838,31 @@ checkContinuation expected =
+ data TransactionTest =
+   TransactionTest
+   {
+-    name          :: String
+-  , generator     :: Gen MarloweContext
+-  , precondition  :: Testify ()
+-  , invariant     :: [Invariant]
+-  , postcondition :: Testify ()
++    name           :: String
++  , generator      :: Gen MarloweContext
++  , precondition   :: Testify ()
++  , invariant      :: [Invariant]
++  , postcondition  :: Testify ()
++  , allowShrinkage :: Bool
+   }
+
+ instance Default TransactionTest where
+   def =
+     TransactionTest
+     {
+-      name          = mempty
+-    , generator     = arbitrary
+-    , precondition  = pure ()
+-    , invariant     = mempty
+-    , postcondition = pure ()
++      name           = mempty
++    , generator      = arbitrary
++    , precondition   = pure ()
++    , invariant      = mempty
++    , postcondition  = pure ()
++    , allowShrinkage = True
+     }
+
+ -- | Test a Marlowe transaction.
+-test :: Bool             -- ^ Whether to perform shrinkage of generated values.
+-     -> TransactionTest  -- ^ The test.
++test :: TransactionTest  -- ^ The test.
+      -> TestTree         -- ^ The result.
+-test doShrink TransactionTest{..} =
++test TransactionTest{..} =
+   testProperty name
+     . property
+     $ let
+@@ -707,7 +875,7 @@ test doShrink TransactionTest{..} =
+             Right () -> True   -- Test passed.
+         gen = generator `suchThat` preResolve precondition
+       in
+-        (if doShrink then forAllShrink gen shrink else forAll gen)
++        (if allowShrinkage then forAllShrink gen shrink else forAll gen)
+           . postResolve
+           $ mapM_ checkInvariant invariant >> postcondition
+
+@@ -858,11 +1026,128 @@ anyInput =
+   }
+
++-- | Test that payments substract value from internal accounts.
++payingSubtractsFromAccount :: TransactionTest
++payingSubtractsFromAccount =
++  def
++  {
++    name           = "Paying subtracts from account"
++  , allowShrinkage = False
++  , generator      = simplePayment
++  , postcondition  = do
++                       delta <-
++                         fmap (AM.filter (/= 0))
++                           $ AM.unionWith (+)
++                           <$> view preAccounts
++                           <*> (AM.fromList . fmap (second negate) . AM.toList <$> view postAccounts)
++                       require "Only one balance changes." ((== 1) . length . AM.toList) delta
++                       require "Some balance decreases." (all (> 0) . AM.elems) delta
++  }
++
++
++-- | Test that deposits add value to internal accounts.
++depositAddsToAccount :: TransactionTest
++depositAddsToAccount =
++  def
++  {
++    name           = "Deposit adds to account"
++  , allowShrinkage = False
++  , generator      = simpleDeposit
++  , postcondition  = do
++                       delta <-
++                         fmap (AM.filter (/= 0))
++                           $ AM.unionWith (+)
++                           <$> view postAccounts
++                           <*> (AM.fromList . fmap (second negate) . AM.toList <$> view preAccounts)
++                       require "Only one balance changes." ((== 1) . length . AM.toList) delta
++                       require "Some balance increases." (all (> 0) . AM.elems) delta
++  }
++
++
++-- | Test that notify contines as expected
++notifyContinues :: TransactionTest
++notifyContinues =
++  def
++  {
++    name           = "Notify continues as expected"
++  , allowShrinkage = False
++  , generator      = simpleNotify
++  , postcondition  = view preContract >>=
++                       \case
++                         When [Case (Notify TrueObs) contract] _ _ -> checkContinuation contract
++                         _ -> throwError "Test setup failed."
++  }
++
++
++-- | Test that choice continues as expected.
++choiceContinues :: TransactionTest
++choiceContinues =
++  def
++  {
++    name           = "Choice continues as expected"
++  , allowShrinkage = False
++  , generator      = simpleChoice
++  , postcondition  = view preContract >>=
++                       \case
++                         When [Case (Choice _ _) contract] _ _ -> checkContinuation contract
++                         _ -> throwError "Test setup failed."
++  }
++
++
++-- | Test that deposit continues as expected.
++depositContinues :: TransactionTest
++depositContinues =
++  def
++  {
++    name           = "Deposit continues as expected"
++  , allowShrinkage = False
++  , generator      = simpleDeposit
++  , postcondition  = view preContract >>=
++                       \case
++                         When [Case Deposit{} contract] _ _ -> checkContinuation contract
++                         _ -> throwError "Test setup failed."
++  }
++
++
++-- | Test that deposit continues as expected.
++merkleizationContinues :: TransactionTest
++merkleizationContinues =
++  def
++  {
++    name           = "Merkleization continues as expected"
++  , allowShrinkage = False
++  , generator      = simpleMerkleization
++  , postcondition  = view inputs >>=
++                       \case
++                         [MerkleizedInput INotify _ contract] -> checkContinuation contract
++                         _ -> throwError "Test setup failed."
++  }
++
++
++-- | Test that deposit continues as expected.
++choiceSets :: TransactionTest
++choiceSets =
++  def
++  {
++    name           = "Choice records the value of the choice"
++  , allowShrinkage = False
++  , generator      = simpleChoice
++  , postcondition  = view inputs >>=
++                       \case
++                         [NormalInput (IChoice choiceId value)] ->
++                           do
++                             choices' <- choices <$> view postState
++                             unless (AM.lookup choiceId choices' == Just value)
++                               $ throwError "Choice missing from state"
++                         _ -> throwError "Test setup failed."
++  }
++
++
+ -- | Run the tests.
+ tests :: TestTree
+ tests =
+   testGroup "Compute Transaction"
+-    $ fmap (test True)
++    $ test <$>
+     [
+       invalidInterval
+     , tooEarly
+@@ -875,4 +1160,11 @@ tests =
+     , ifBranches
+     , assertWarns
+     , anyInput
++    , payingSubtractsFromAccount
++    , depositAddsToAccount
++    , depositContinues
++    , choiceContinues
++    , notifyContinues
++    , choiceSets
++    , merkleizationContinues
+     ]
+diff --git a/marlowe-test/src/Spec/Marlowe/Semantics/Functions.hs b/marlowe-test/src/Spec/Marlowe/Semantics/Functions.hs
+index 25fbef841..9a3c8acdc 100644
+--- a/marlowe-test/src/Spec/Marlowe/Semantics/Functions.hs
++++ b/marlowe-test/src/Spec/Marlowe/Semantics/Functions.hs
+@@ -100,6 +100,7 @@ tests =
+         testProperty "Invalid interval" $ checkFixInterval True  False
+       , testProperty "Interval in past" $ checkFixInterval False True
+       , testProperty "Interval trimmed" $ checkFixInterval False False
++      , testProperty "Invalid interval in past" $ checkFixInterval True True
+       ]
+     , testGroup "evalValue"
+       [
+@@ -218,14 +219,22 @@ forAll' = flip forAllShrink shrink
+ --   generating arbitrary valid and invalid intervals and checking that
+ --   results or errors re reported correctly.
+ checkFixInterval :: Bool      -- ^ Whether the validity interval should be invalid.
+-                 -> Bool      -- ^ Whethe the validity interval should be in the past.
++                 -> Bool      -- ^ Whether the validity interval should be in the past.
+                  -> Property  -- ^ The test.
+ checkFixInterval invalid inPast =
+   property $ do
+   let gen = do
+         state <- arbitrary
+-        end   <- arbitrary `suchThat` (\t -> (t < minTime state) == inPast)
+-        start <- arbitrary `suchThat` (\t -> (t > end) == invalid && (t < minTime state) == inPast)
++        (start, end) <-
++          case (invalid, inPast) of
++            (True, True) -> do
++                              start' <- arbitrary `suchThat` (< minTime state)
++                              end'   <- arbitrary `suchThat` (< start')
++                              pure (start', end')
++            _            -> do
++                              end'   <- arbitrary `suchThat` (\t -> (t < minTime state) == inPast)
++                              start' <- arbitrary `suchThat` (\t -> (t > end') == invalid && (t < minTime state) == inPast)
++                              pure (start', end')
+         pure ((start, end), state)
+   forAll' gen $ \(interval, state) ->
+     case fixInterval interval state of
+diff --git a/marlowe-test/src/Spec/Marlowe/Semantics/Golden.hs b/marlowe-test/src/Spec/Marlowe/Semantics/Golden.hs
+index 9025906ad..72d052fb7 100644
+--- a/marlowe-test/src/Spec/Marlowe/Semantics/Golden.hs
++++ b/marlowe-test/src/Spec/Marlowe/Semantics/Golden.hs
+@@ -40,6 +40,7 @@ import Test.Tasty.HUnit (assertBool, testCase)
+
+ import qualified PlutusTx.AssocMap as AM (empty)
+ import qualified Spec.Marlowe.Semantics.Golden.Escrow as Escrow (contract, invalids, valids)
++import qualified Spec.Marlowe.Semantics.Golden.Negative as Negative (contract, invalids, valids)
+ import qualified Spec.Marlowe.Semantics.Golden.Pangram as Pangram (contract, invalids, valids)
+ import qualified Spec.Marlowe.Semantics.Golden.Swap as Swap (contract, invalids, valids)
+ import qualified Spec.Marlowe.Semantics.Golden.Trivial as Trivial (contract, invalids, valids)
+@@ -59,6 +60,7 @@ goldenContracts =
+   , Swap.contract
+   , Trivial.contract
+   , ZCB.contract
++  , Negative.contract
+   -- Note that Pangram is omitted because `getAllInputs` takes 30 minutes for it.
+   ]
+
+@@ -68,11 +70,12 @@ tests :: TestTree
+ tests =
+   testGroup "Golden"
+     [
+-      testGolden "Escrow"           _GENERATE_TEST_CASES_ Escrow.contract  Escrow.valids  Escrow.invalids
+-    , testGolden "Pangram"          _GENERATE_TEST_CASES_ Pangram.contract Pangram.valids Pangram.invalids
+-    , testGolden "Swap"             _GENERATE_TEST_CASES_ Swap.contract    Swap.valids    Swap.invalids
+-    , testGolden "Trivial"          _GENERATE_TEST_CASES_ Trivial.contract Trivial.valids Trivial.invalids
+-    , testGolden "Zero Coupon Bond" _GENERATE_TEST_CASES_ ZCB.contract     ZCB.valids     ZCB.invalids
++      testGolden "Escrow"           _GENERATE_TEST_CASES_ Escrow.contract   Escrow.valids   Escrow.invalids
++    , testGolden "Pangram"          _GENERATE_TEST_CASES_ Pangram.contract  Pangram.valids  Pangram.invalids
++    , testGolden "Swap"             _GENERATE_TEST_CASES_ Swap.contract     Swap.valids     Swap.invalids
++    , testGolden "Trivial"          _GENERATE_TEST_CASES_ Trivial.contract  Trivial.valids  Trivial.invalids
++    , testGolden "Zero Coupon Bond" _GENERATE_TEST_CASES_ ZCB.contract      ZCB.valids      ZCB.invalids
++    , testGolden "Negative"         _GENERATE_TEST_CASES_ Negative.contract Negative.valids Negative.invalids
+     ]
+
+@@ -125,7 +128,7 @@ testInvalid :: Contract -> [GoldenCase] -> IO ()
+ testInvalid = testValidity False
+
+--- | Test erroneous transactions for the Pangram contract.
++-- | Test erroneous transactions for a contract.
+ testValidity :: Bool          -- ^ Whether the test cases should pass.
+              -> Contract      -- ^ The contract.
+              -> [GoldenCase]  -- ^ The test cases.
+@@ -148,11 +151,12 @@ goldenTransactions :: [[GoldenTransaction]]
+ goldenTransactions =
+   uncurry validTransactions
+     <$> [
+-          (Escrow.contract , Escrow.valids )
+-        , (Pangram.contract, Pangram.valids)
+-        , (Swap.contract   , Swap.valids   )
+-        , (Trivial.contract, Trivial.valids)
+-        , (ZCB.contract    , ZCB.valids    )
++          (Escrow.contract  , Escrow.valids  )
++        , (Pangram.contract , Pangram.valids )
++        , (Swap.contract    , Swap.valids    )
++        , (Trivial.contract , Trivial.valids )
++        , (ZCB.contract     , ZCB.valids     )
++        , (Negative.contract, Negative.valids)
+         ]
+
+diff --git a/marlowe-test/src/Spec/Marlowe/Semantics/Golden/Negative.hs b/marlowe-test/src/Spec/Marlowe/Semantics/Golden/Negative.hs
+new file mode 100644
+index 000000000..eea8fa414
+--- /dev/null
++++ b/marlowe-test/src/Spec/Marlowe/Semantics/Golden/Negative.hs
+@@ -0,0 +1,151 @@
++-----------------------------------------------------------------------------
++--
++-- Module      :  $Headers
++-- License     :  Apache 2.0
++--
++-- Stability   :  Experimental
++-- Portability :  Portable
++--
++-- | Reference golden output for a contract with negative deposits.
++--
++-----------------------------------------------------------------------------
++
++
++{-# LANGUAGE NumericUnderscores #-}
++{-# LANGUAGE OverloadedStrings #-}
++{-# LANGUAGE TypeSynonymInstances #-}
++
++
++module Spec.Marlowe.Semantics.Golden.Negative
++  ( -- * Contracts
++    contract
++    -- * Test cases
++  , invalids
++  , valids
++  ) where
++
++
++import Language.Marlowe.Core.V1.Semantics
++  (Payment(Payment), TransactionInput(..), TransactionOutput(..), TransactionWarning(TransactionNonPositiveDeposit))
++import Language.Marlowe.Core.V1.Semantics.Types
++  ( Action(Choice, Deposit, Notify)
++  , Bound(..)
++  , Case(Case)
++  , ChoiceId(..)
++  , Contract(Close, Let, When)
++  , Input(NormalInput)
++  , InputContent(IChoice, IDeposit, INotify)
++  , Observation(ValueEQ, ValueGT, ValueLT)
++  , Party
++  , Payee(Party)
++  , State(State, accounts, boundValues, choices, minTime)
++  , Token(..)
++  , Value(ChoiceValue, Constant)
++  )
++import Language.Marlowe.Util ()
++import Plutus.V2.Ledger.Api (POSIXTime(..))
++
++import qualified PlutusTx.AssocMap as AM (Map, fromList)
++
++
++party :: Party
++party = "Party"
++
++counterparty :: Party
++counterparty = "Counterparty"
++
++ada :: Token
++ada = Token "" ""
++
++
++-- | The Zero-Coupon Bond contract.
++contract :: Contract
++contract =
++  let
++    initial = Constant 2_000_000
++    choice = ChoiceId "Deposit" party
++    value = ChoiceValue choice
++    initialDeadline = 1000
++    choiceDeadline = 2000
++    notifyDeadline = 3000
++    depositDeadline = 4000
++  in
++    When
++      [
++        Case (Deposit counterparty counterparty ada initial)
++          $ When
++              [
++                Case (Choice choice [Bound (-1) 1])
++                  $ When
++                    [
++                      Case (Deposit party counterparty ada value)
++                        Close
++                    , Case (Notify $ ValueGT value $ Constant 0)
++                       $ Let "Positive" (Constant 1)
++                       $ When
++                         [
++                           Case (Deposit party counterparty ada value)
++                             Close
++                         ]
++                         depositDeadline
++                         Close
++                    , Case (Notify $ ValueEQ value $ Constant 0)
++                       $ Let "Neutral" (Constant 1)
++                       $ When
++                         [
++                           Case (Deposit party party ada value)
++                             Close
++                         ]
++                         depositDeadline
++                         Close
++                    , Case (Notify $ ValueLT value $ Constant 0)
++                       $ Let "Negative" (Constant 1)
++                       $ When
++                         [
++                           Case (Deposit counterparty party ada value)
++                             Close
++                         ]
++                         depositDeadline
++                         Close
++                    ]
++                    notifyDeadline
++                    Close
++              ]
++              choiceDeadline
++              Close
++      ]
++      initialDeadline
++      Close
++
++
++-- | A wrapper to assist parsing of test cases.
++newtype Map k v = Map {unMap :: [(k, v)]}
++
++
++-- | A function to assist parsing of test cases.
++toAM :: Map k v -> AM.Map k v
++toAM = AM.fromList . unMap
++
++
++-- | A list of test cases and results that should succeed, generated from `Language.Marlowe.FindInputs.getAllInputs`.
++valids :: [(POSIXTime, [TransactionInput], TransactionOutput)]
++valids =
++  [
++    (POSIXTime {getPOSIXTime = 0}, [TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 1000},POSIXTime {getPOSIXTime = 1000}), txInputs = []}], TransactionOutput {txOutWarnings = [], txOutPayments = [], txOutState = State {accounts = toAM $ Map {unMap = []}, choices = toAM $ Map {unMap = []}, boundValues = toAM $ Map {unMap = []}, minTime = POSIXTime {getPOSIXTime = 1000}}, txOutContract = Close})
++  , (POSIXTime {getPOSIXTime = 0}, [TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IDeposit "Counterparty" "Counterparty" (Token "" "") 2000000)]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 2000},POSIXTime {getPOSIXTime = 2000}), txInputs = []}], TransactionOutput {txOutWarnings = [], txOutPayments = [Payment "Counterparty" (Party "Counterparty") (Token "" "") 2000000], txOutState = State {accounts = toAM $ Map {unMap = []}, choices = toAM $ Map {unMap = []}, boundValues = toAM $ Map {unMap = []}, minTime = POSIXTime {getPOSIXTime = 2000}}, txOutContract = Close})
++  , (POSIXTime {getPOSIXTime = 0}, [TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IDeposit "Counterparty" "Counterparty" (Token "" "") 2000000)]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IChoice (ChoiceId "Deposit" "Party") 0)]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 3000},POSIXTime {getPOSIXTime = 3000}), txInputs = []}], TransactionOutput {txOutWarnings = [], txOutPayments = [Payment "Counterparty" (Party "Counterparty") (Token "" "") 2000000], txOutState = State {accounts = toAM $ Map {unMap = []}, choices = toAM $ Map {unMap = [(ChoiceId "Deposit" "Party",0)]}, boundValues = toAM $ Map {unMap = []}, minTime = POSIXTime {getPOSIXTime = 3000}}, txOutContract = Close})
++  , (POSIXTime {getPOSIXTime = 0}, [TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IDeposit "Counterparty" "Counterparty" (Token "" "") 2000000)]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IChoice (ChoiceId "Deposit" "Party") 0)]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IDeposit "Party" "Counterparty" (Token "" "") 0)]}], TransactionOutput {txOutWarnings = [TransactionNonPositiveDeposit "Counterparty" "Party" (Token "" "") 0], txOutPayments = [Payment "Counterparty" (Party "Counterparty") (Token "" "") 2000000], txOutState = State {accounts = toAM $ Map {unMap = []}, choices = toAM $ Map {unMap = [(ChoiceId "Deposit" "Party",0)]}, boundValues = toAM $ Map {unMap = []}, minTime = POSIXTime {getPOSIXTime = 0}}, txOutContract = Close})
++  , (POSIXTime {getPOSIXTime = 0}, [TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IDeposit "Counterparty" "Counterparty" (Token "" "") 2000000)]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IChoice (ChoiceId "Deposit" "Party") 1)]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput INotify]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 4000},POSIXTime {getPOSIXTime = 4000}), txInputs = []}], TransactionOutput {txOutWarnings = [], txOutPayments = [Payment "Counterparty" (Party "Counterparty") (Token "" "") 2000000], txOutState = State {accounts = toAM $ Map {unMap = []}, choices = toAM $ Map {unMap = [(ChoiceId "Deposit" "Party",1)]}, boundValues = toAM $ Map {unMap = [("Positive",1)]}, minTime = POSIXTime {getPOSIXTime = 4000}}, txOutContract = Close})
++  , (POSIXTime {getPOSIXTime = 0}, [TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IDeposit "Counterparty" "Counterparty" (Token "" "") 2000000)]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IChoice (ChoiceId "Deposit" "Party") 1)]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput INotify]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IDeposit "Party" "Counterparty" (Token "" "") 1)]}], TransactionOutput {txOutWarnings = [], txOutPayments = [Payment "Counterparty" (Party "Counterparty") (Token "" "") 2000000,Payment "Party" (Party "Party") (Token "" "") 1], txOutState = State {accounts = toAM $ Map {unMap = []}, choices = toAM $ Map {unMap = [(ChoiceId "Deposit" "Party",1)]}, boundValues = toAM $ Map {unMap = [("Positive",1)]}, minTime = POSIXTime {getPOSIXTime = 0}}, txOutContract = Close})
++  , (POSIXTime {getPOSIXTime = 0}, [TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IDeposit "Counterparty" "Counterparty" (Token "" "") 2000000)]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IChoice (ChoiceId "Deposit" "Party") 0)]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput INotify]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 4000},POSIXTime {getPOSIXTime = 4000}), txInputs = []}], TransactionOutput {txOutWarnings = [], txOutPayments = [Payment "Counterparty" (Party "Counterparty") (Token "" "") 2000000], txOutState = State {accounts = toAM $ Map {unMap = []}, choices = toAM $ Map {unMap = [(ChoiceId "Deposit" "Party",0)]}, boundValues = toAM $ Map {unMap = [("Neutral",1)]}, minTime = POSIXTime {getPOSIXTime = 4000}}, txOutContract = Close})
++  , (POSIXTime {getPOSIXTime = 0}, [TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IDeposit "Counterparty" "Counterparty" (Token "" "") 2000000)]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IChoice (ChoiceId "Deposit" "Party") 0)]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput INotify]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IDeposit "Party" "Party" (Token "" "") 0)]}], TransactionOutput {txOutWarnings = [TransactionNonPositiveDeposit "Party" "Party" (Token "" "") 0], txOutPayments = [Payment "Counterparty" (Party "Counterparty") (Token "" "") 2000000], txOutState = State {accounts = toAM $ Map {unMap = []}, choices = toAM $ Map {unMap = [(ChoiceId "Deposit" "Party",0)]}, boundValues = toAM $ Map {unMap = [("Neutral",1)]}, minTime = POSIXTime {getPOSIXTime = 0}}, txOutContract = Close})
++  , (POSIXTime {getPOSIXTime = 0}, [TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IDeposit "Counterparty" "Counterparty" (Token "" "") 2000000)]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IChoice (ChoiceId "Deposit" "Party") (-1))]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput INotify]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 4000},POSIXTime {getPOSIXTime = 4000}), txInputs = []}], TransactionOutput {txOutWarnings = [], txOutPayments = [Payment "Counterparty" (Party "Counterparty") (Token "" "") 2000000], txOutState = State {accounts = toAM $ Map {unMap = []}, choices = toAM $ Map {unMap = [(ChoiceId "Deposit" "Party",-1)]}, boundValues = toAM $ Map {unMap = [("Negative",1)]}, minTime = POSIXTime {getPOSIXTime = 4000}}, txOutContract = Close})
++  , (POSIXTime {getPOSIXTime = 0}, [TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IDeposit "Counterparty" "Counterparty" (Token "" "") 2000000)]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IChoice (ChoiceId "Deposit" "Party") (-1))]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput INotify]},TransactionInput {txInterval = (POSIXTime {getPOSIXTime = 0},POSIXTime {getPOSIXTime = 0}), txInputs = [NormalInput (IDeposit "Counterparty" "Party" (Token "" "") (-1))]}], TransactionOutput {txOutWarnings = [TransactionNonPositiveDeposit "Party" "Counterparty" (Token "" "") (-1)], txOutPayments = [Payment "Counterparty" (Party "Counterparty") (Token "" "") 2000000], txOutState = State {accounts = toAM $ Map {unMap = []}, choices = toAM $ Map {unMap = [(ChoiceId "Deposit" "Party",-1)]}, boundValues = toAM $ Map {unMap = [("Negative",1)]}, minTime = POSIXTime {getPOSIXTime = 0}}, txOutContract = Close})
++  ]
++
++
++-- | A list of test cases and results that should fail.
++invalids :: [(POSIXTime, [TransactionInput], TransactionOutput)]
++invalids =
++  [
++  ]
+diff --git a/marlowe-test/src/Spec/Marlowe/Semantics/Merkle.hs b/marlowe-test/src/Spec/Marlowe/Semantics/Merkle.hs
+index 747c053fa..ebe286cb3 100644
+--- a/marlowe-test/src/Spec/Marlowe/Semantics/Merkle.hs
++++ b/marlowe-test/src/Spec/Marlowe/Semantics/Merkle.hs
+@@ -14,7 +14,6 @@
+ {-# LANGUAGE FlexibleContexts #-}
+ {-# LANGUAGE RankNTypes #-}
+ {-# LANGUAGE RecordWildCards #-}
+-{-# LANGUAGE TupleSections #-}
+
+ module Spec.Marlowe.Semantics.Merkle
+diff --git a/marlowe-test/src/Spec/Marlowe/Semantics/Oracle.hs b/marlowe-test/src/Spec/Marlowe/Semantics/Oracle.hs
+deleted file mode 100644
+index 081e88034..000000000
+--- a/marlowe-test/src/Spec/Marlowe/Semantics/Oracle.hs
++++ /dev/null
+@@ -1,261 +0,0 @@
+------------------------------------------------------------------------------
+---
+--- Module      :  $Headers
+--- License     :  Apache 2.0
+---
+--- Stability   :  Experimental
+--- Portability :  Portable
+---
+--- | Tests using external semantics oracle.
+---
+------------------------------------------------------------------------------
+-
+-
+-
+-{-# LANGUAGE DeriveGeneric #-}
+-{-# LANGUAGE NumericUnderscores #-}
+-{-# LANGUAGE OverloadedStrings #-}
+-{-# LANGUAGE RecordWildCards #-}
+-
+-
+-module Spec.Marlowe.Semantics.Oracle
+-  ( -- * Tests
+-    tests
+-  ) where
+-
+-
+-import Data.Aeson
+-  ( FromJSON(parseJSON)
+-  , KeyValue((.=))
+-  , ToJSON(toJSON)
+-  , Value(Array)
+-  , eitherDecode
+-  , encode
+-  , object
+-  , withArray
+-  , withObject
+-  , (.:)
+-  )
+-import Data.Aeson.Types (Parser)
+-import Data.Function (on)
+-import Data.List (sort)
+-import Data.String (fromString)
+-import Data.Vector ((!))
+-import GHC.Generics (Generic)
+-import Language.Marlowe.Core.V1.Semantics
+-  ( Payment(..)
+-  , TransactionInput
+-  , TransactionOutput(TransactionOutput, txOutContract, txOutPayments, txOutState, txOutWarnings)
+-  , computeTransaction
+-  )
+-import Language.Marlowe.Core.V1.Semantics.Types (Contract, State(..), Token(Token))
+-import Language.Marlowe.FindInputs (getAllInputs)
+-import System.Directory (doesFileExist)
+-import System.IO
+-  (BufferMode(LineBuffering), Handle, IOMode(ReadMode, WriteMode), hGetLine, hPutStrLn, hSetBuffering, openFile)
+-import Test.Tasty (TestTree)
+-import Test.Tasty.HUnit (assertBool, testCaseSteps)
+-
+-import qualified Data.ByteString.Lazy.Char8 as LBS8 (pack, unpack)
+-import qualified Data.Vector as V (fromList)
+-import qualified Language.Marlowe.Extended.V1 as E (Contract(Close), Value(Constant), toCore)
+-import qualified PlutusTx.AssocMap as AM (Map, empty, toList)
+-
+-import qualified Marlowe.Contracts.Escrow as Escrow (escrow)
+-import qualified Marlowe.Contracts.Forward as Forward (forward)
+-import qualified Marlowe.Contracts.Futures as Futures (future)
+-import qualified Marlowe.Contracts.Swap as Swap (swap)
+-import qualified Marlowe.Contracts.Trivial as Trivial (trivial)
+-import qualified Marlowe.Contracts.ZeroCouponBond as ZeroCouponBond (zeroCouponBond)
+-
+-
+--- | Compare contract execution to the semantics oracle.
+---
+--- The function reads from a named pipe "orin" and writes to a
+--- named pipe "orout". Those same pipes must be connected to the
+--- semantics oracle. For example:
+---
+--- > git clone git@github.com:input-output-hk/marlowe-cardano.git -b 472170bdf77dc9d89bc77861da851d2476b2aa47
+--- > nix-shell
+--- > cd marlowe-purescript-spec
+--- > spago install
+--- > spago build
+--- > npm install
+--- > node index.js <orout >orin
+-tests :: TestTree
+-tests =
+-  testCaseSteps "Semantics Oracle"
+-    $ \step ->
+-      do
+-      okay <- and <$> mapM doesFileExist ["orin", "orout"]
+-      if okay
+-        then do
+-               step "Connecting to semantics oracle . . ."
+-               hToOracle   <- openFile "orout" WriteMode
+-               hFromOracle <- openFile "orin"  ReadMode
+-               hSetBuffering hToOracle   LineBuffering
+-               hSetBuffering hFromOracle LineBuffering
+-               sequence_
+-                 [
+-                   do
+-                     step $ "Comparing " <> name <> " to oracle . . ."
+-                     compareToOracle hToOracle hFromOracle name contract
+-                 |
+-                   (name, contract) <- contracts
+-                 ]
+-        else step "No semantics oracle found on pipes \"orin\" and \"orout\"."
+-
+-
+--- | Contracts to be tested.
+-contracts :: [(String, E.Contract)]
+-contracts =
+-  let
+-    alice = "Alice"
+-    bob = "Bob"
+-    charlie = "Charlie"
+-    v = E.Constant
+-    a = Token "aaaa" "A"
+-    b = Token "bbbb" "B"
+-    c = Token "cccc" "C"
+-  in
+-    [
+-      ("Escrow"          , Escrow.escrow (v 900_000_000) alice bob charlie 100 200 300 400)
+-    , ("Forward"         , Forward.forward alice a (v 10) 100 bob b (v 20) 200 300)
+-    , ("Future"          , Futures.future alice bob (v 100) (v 200) 100 [200] 500)
+-    , ("Swap"            , Swap.swap alice a (v 100) 100 bob b (v 200) 200 E.Close)
+-    , ("Trivial"         , Trivial.trivial alice 100_000_000 70_000_000 0)
+-    , ("Zero-Coupon Bond", ZeroCouponBond.zeroCouponBond alice bob 100 200 (v 100_000_000) (v 200_000_000) c E.Close)
+-    ]
+-
+-
+--- | A request to the semantics oracle.
+-data OracleRequest =
+-  OracleRequest
+-  {
+-    request          :: String
+-  , state            :: State
+-  , transactionInput :: TransactionInput
+-  , coreContract     :: Contract
+-  }
+-  deriving (Generic, Show)
+-
+-instance FromJSON OracleRequest
+-
+-instance ToJSON OracleRequest
+-
+-
+--- | A response from the semantics oracle.
+-data OracleResponse =
+-    TransactionSuccess TransactionOutput
+-  | TransactionError
+-  deriving (Generic, Show)
+-
+-instance FromJSON OracleResponse where
+-  parseJSON =
+-    withObject "OracleResponse"
+-      $ \o ->
+-        do
+-          success <- o .: "transaction_success"
+-          txOutWarnings <- success .: "txOutWarnings"
+-          payments      <- success .: "txOutPayments"
+-          txOutState    <- success .: "txOutState"
+-          txOutContract <- success .: "txOutContract"
+-          txOutPayments <- mapM paymentFromJSON payments
+-          pure . TransactionSuccess $ TransactionOutput{..}
+-
+-
+-instance ToJSON OracleResponse where
+-  toJSON (TransactionSuccess TransactionOutput{..}) =
+-    object
+-      [
+-        "transaction_success" .= object
+-                                   [
+-                                     "txOutWarnings" .= toJSON txOutWarnings
+-                                   , "txOutPayments" .= fmap paymentToJSON txOutPayments
+-                                   , "txOutState"    .= toJSON txOutState
+-                                   , "txOutContract" .= toJSON txOutContract
+-                                   ]
+-      ]
+-  toJSON _ = "transaction_error"
+-
+-
+--- | Deserialize a payment from JSON.
+-paymentFromJSON :: Value -> Parser Payment
+-paymentFromJSON =
+-  withArray "Payment"
+-    $ \v ->
+-      do
+-        a <- parseJSON $ v ! 0
+-        p <- parseJSON $ v ! 1
+-        [(c, [(n, i)])] <- parseJSON (v ! 2)
+-        pure $ Payment a p (Token (fromString c) (fromString n)) i
+-
+-
+--- | Serialize a payment to JSON.
+-paymentToJSON :: Payment -> Value
+-paymentToJSON (Payment a p t i) =
+-  Array
+-    $ V.fromList
+-    [
+-      toJSON a
+-    , toJSON p
+-    , toJSON t
+-    , toJSON i
+-    ]
+-
+-
+--- | Equality test for transaction output.
+-eqTransactionOutput :: TransactionOutput -> TransactionOutput -> Bool
+-eqTransactionOutput x@TransactionOutput{} y@TransactionOutput{} =
+-     txOutWarnings x == txOutWarnings y
+-  && txOutPayments x == txOutPayments y
+-  && txOutState    x ~~ txOutState    y
+-  && txOutContract x == txOutContract y
+-eqTransactionOutput _ _ = False
+-
+-
+--- | Equality test for state.
+-(~~) :: State -> State -> Bool
+-x ~~ y =
+-     accounts    x `eqAssocMap` accounts    y
+-  && choices     x `eqAssocMap` choices     y
+-  && boundValues x `eqAssocMap` boundValues y
+-  && minTime     x ==           minTime     y
+-
+-
+--- | Equality test for association lists.
+-eqAssocMap :: Ord k => Ord v => AM.Map k v -> AM.Map k v -> Bool
+-eqAssocMap = (==) `on` (sort . AM.toList)
+-
+-
+--- | Use static analysis to generate paths through a contract, and then compare execution to the oracle.
+-compareToOracle :: Handle -> Handle -> String -> E.Contract -> IO ()
+-compareToOracle hToOracle hFromOracle name extended =
+-  do
+-    Just core <- pure $ E.toCore extended
+-    Right paths <- getAllInputs core
+-    let
+-      accounts = AM.empty
+-      choices = AM.empty
+-      boundValues = AM.empty
+-      request = "compute-transaction"
+-    sequence_
+-      [
+-        let
+-          go _ _ _ [] = pure ()
+-          go j state coreContract (transactionInput : transactionInputs) =
+-            do
+-              hPutStrLn hToOracle "```"
+-              hPutStrLn hToOracle $ LBS8.unpack $ encode OracleRequest{..}
+-              hPutStrLn hToOracle "```"
+-              result <- hGetLine hFromOracle
+-              actual@TransactionOutput{..} <- pure $ computeTransaction transactionInput state coreContract
+-              Right (TransactionSuccess expected) <- pure $ eitherDecode $ LBS8.pack result
+-              let match = actual `eqTransactionOutput` expected
+-              assertBool (name <> ", path " <> show (i :: Int) <> ", step " <> show (j :: Int)) match
+-              go (j + 1) txOutState txOutContract transactionInputs
+-        in
+-          go 1 State{..} core inputs
+-      |
+-        (i, (minTime, inputs)) <- zip [1..] paths
+-      ]
+diff --git a/marlowe-test/src/Spec/Marlowe/Serialization/CoreJson.hs b/marlowe-test/src/Spec/Marlowe/Serialization/CoreJson.hs
+index a3c456b9d..3420569d3 100644
+--- a/marlowe-test/src/Spec/Marlowe/Serialization/CoreJson.hs
++++ b/marlowe-test/src/Spec/Marlowe/Serialization/CoreJson.hs
+@@ -9,12 +9,16 @@
+ -- | This suite tests the Json serialization of the Marlowe core module
+ --
+ -----------------------------------------------------------------------------
++
++
+ {-# LANGUAGE OverloadedStrings #-}
+
++
+ module Spec.Marlowe.Serialization.CoreJson
+   ( tests
+   ) where
+
++
+ import Control.Arrow (Arrow((***)))
+ import Data.Aeson (decode, encode)
+ import Data.ByteString (ByteString)
+@@ -28,6 +32,7 @@ import Test.QuickCheck.Property (forAll, forAllShrink, withMaxSuccess)
+ import Test.Tasty (TestTree, testGroup)
+ import Test.Tasty.QuickCheck (testProperty)
+
++
+ tests :: TestTree
+ tests = testGroup "Core Contract Serialization"
+     [ testProperty "Serialise deserialise Contract loops" prop_contractJsonLoops
+@@ -40,6 +45,7 @@ tests = testGroup "Core Contract Serialization"
+ contractJsonLoops :: Contract -> Property
+ contractJsonLoops cont = decode (encode cont) === Just cont
+
++
+ -- | Test that JSON decoding inverts encoding for contracts.
+ prop_contractJsonLoops :: Property
+ prop_contractJsonLoops = withMaxSuccess 1000 $ forAllShrink contractGen shrinkContract contractJsonLoops
+@@ -50,6 +56,7 @@ prop_contractJsonLoops = withMaxSuccess 1000 $ forAllShrink contractGen shrinkCo
+ marloweParamsJsonLoops :: MarloweParams -> Property
+ marloweParamsJsonLoops mp = decode (encode mp) === Just mp
+
++
+ -- | Test that JSON decoding inverts encoding for Marlowe parameters.
+ prop_marloweParamsJsonLoops :: Property
+ prop_marloweParamsJsonLoops = withMaxSuccess 1000 $ forAll gen marloweParamsJsonLoops
+@@ -58,6 +65,7 @@ prop_marloweParamsJsonLoops = withMaxSuccess 1000 $ forAll gen marloweParamsJson
+       c <- toBuiltin <$> (arbitrary :: Gen ByteString)
+       return $ MarloweParams (CurrencySymbol c)
+
++
+ -- | Test that JSON decoding inverts encoding for an interval error.
+ intervalErrorJsonLoops :: IntervalError -> Property
+ intervalErrorJsonLoops ie = decode (encode ie) === Just ie
+diff --git a/marlowe-test/src/Spec/Marlowe/Serialization/ExtendedJson.hs b/marlowe-test/src/Spec/Marlowe/Serialization/ExtendedJson.hs
+index b93984ddd..0c357bbf4 100644
+--- a/marlowe-test/src/Spec/Marlowe/Serialization/ExtendedJson.hs
++++ b/marlowe-test/src/Spec/Marlowe/Serialization/ExtendedJson.hs
+@@ -9,18 +9,23 @@
+ -- | This suite tests the Json serialization of the Marlowe extended module
+ --
+ -----------------------------------------------------------------------------
++
++
+ {-# LANGUAGE OverloadedStrings #-}
+
++
+ module Spec.Marlowe.Serialization.ExtendedJson
+   ( tests
+   ) where
+
++
+ import Data.Aeson (eitherDecodeFileStrict)
+ import Language.Marlowe.Extended.V1 (Contract, Module)
+ import System.FilePath ((</>))
+ import Test.Tasty (TestTree, testGroup)
+ import Test.Tasty.HUnit (assertFailure, testCase)
+
++
+ tests :: TestTree
+ tests = testGroup "Extended Contract Serialization"
+     [ testCase "Golden Swap contract" testGoldenSwapContract
+@@ -29,9 +34,11 @@ tests = testGroup "Extended Contract Serialization"
+
+ -- TODO: Do a small non-property round-trip with example contracts.
+
++
+ goldenPath :: FilePath
+ goldenPath = "test" </> "Spec" </> "Marlowe" </> "Serialization" </> "golden"
+
++
+ -- | Checks that we can decode the Golden JSON Contract for Swap
+ testGoldenSwapContract :: IO ()
+ testGoldenSwapContract = do
+@@ -40,6 +47,7 @@ testGoldenSwapContract = do
+         Left err              -> assertFailure err
+         Right (_ :: Contract) -> return ()
+
++
+ -- | Checks that we can decode the Golden JSON Module for Swap
+ -- | TODO: If we are more of these tests, add a helper function with a Proxy type
+ testGoldenSwapModule :: IO ()
+diff --git a/marlowe-test/src/Spec/Marlowe/Service.hs b/marlowe-test/src/Spec/Marlowe/Service.hs
+new file mode 100644
+index 000000000..e02ef56f8
+--- /dev/null
++++ b/marlowe-test/src/Spec/Marlowe/Service.hs
+@@ -0,0 +1,73 @@
++-----------------------------------------------------------------------------
++--
++-- Module      :  $Headers
++-- License     :  Apache 2.0
++--
++-- Stability   :  Experimental
++-- Portability :  Portable
++--
++-- | Client for Isabelle-based Marlowe testing service.
++--
++-----------------------------------------------------------------------------
++
++
++{-# LANGUAGE LambdaCase #-}
++{-# LANGUAGE OverloadedStrings #-}
++{-# LANGUAGE RecordWildCards #-}
++
++
++module Spec.Marlowe.Service
++  ( -- * Testing
++    handle
++  , handleValues
++  ) where
++
++
++import Spec.Marlowe.Service.Random (generateValue)
++import Spec.Marlowe.Service.Serialization (roundtripSerialization)
++import Spec.Marlowe.Service.Types (Request(..), Response(..))
++
++import qualified Data.Aeson as A (Result(..), Value, fromJSON, object, toJSON, (.=))
++import qualified Language.Marlowe.Core.V1.Semantics as Marlowe
++  (computeTransaction, evalObservation, evalValue, playTrace)
++
++
++-- | Respond to a request expressed as JSON.
++handleValues :: A.Value -> IO A.Value
++handleValues request =
++  case A.fromJSON request of
++    A.Success request' -> A.toJSON <$> handle request'
++    A.Error message -> error message
++
++
++-- | Respond to a request.
++handle :: Request -> IO Response
++handle TestRoundtripSerialization{..} =
++  pure
++    . RequestResponse . A.toJSON
++    $ roundtripSerialization typeSerialized valueSerialized
++handle GenerateRandomValue{..} =
++  generateValue size seed typeSerialized
++    >>= \case
++      Right value -> pure . RequestResponse . A.object . pure $ "value" A..= value
++      Left failureResponse -> pure $ ResponseFailure{..}
++handle ComputeTransaction{..} =
++  let
++    valueResponse = A.toJSON $ Marlowe.computeTransaction transactionInput state contract
++  in
++    pure RequestResponse{..}
++handle PlayTrace{..} =
++  let
++    valueResponse = A.toJSON $ Marlowe.playTrace initialTime contract transactionInputs
++  in
++    pure RequestResponse{..}
++handle EvalValue{..} =
++  let
++    valueResponse = A.toJSON $ Marlowe.evalValue environment state value
++  in
++    pure RequestResponse{..}
++handle EvalObservation{..} =
++  let
++    valueResponse = A.toJSON $ Marlowe.evalObservation environment state observation
++  in
++    pure RequestResponse{..}
+diff --git a/marlowe-test/src/Spec/Marlowe/Service/Isabelle.hs b/marlowe-test/src/Spec/Marlowe/Service/Isabelle.hs
+new file mode 100644
+index 000000000..f4c5fcf0e
+--- /dev/null
++++ b/marlowe-test/src/Spec/Marlowe/Service/Isabelle.hs
+@@ -0,0 +1,39 @@
++-----------------------------------------------------------------------------
++--
++-- Module      :  $Headers
++-- License     :  Apache 2.0
++--
++-- Stability   :  Experimental
++-- Portability :  Portable
++--
++-- | Test Marlowe's Cardano implementation against the Isabelle-derived test server.
++--
++-----------------------------------------------------------------------------
++
++
++module Spec.Marlowe.Service.Isabelle
++  ( -- * Testing
++    tests
++  ) where
++
++
++import Spec.Marlowe.Service (handleValues)
++import Test.Tasty (TestTree, testGroup)
++
++import qualified Data.Aeson as A
++import qualified Marlowe.Spec.Core (tests)
++
++
++-- | Run the tests.
++tests :: TestTree
++tests =
++  testGroup "Marlowe Client"
++    [
++      Marlowe.Spec.Core.tests
++        $ \request ->
++          do
++            response <- handleValues $ A.toJSON request
++            case A.fromJSON response of
++              A.Success response' -> pure response'
++              A.Error message -> error message
++    ]
+diff --git a/marlowe-test/src/Spec/Marlowe/Service/Random.hs b/marlowe-test/src/Spec/Marlowe/Service/Random.hs
+new file mode 100644
+index 000000000..5a312aa0f
+--- /dev/null
++++ b/marlowe-test/src/Spec/Marlowe/Service/Random.hs
+@@ -0,0 +1,44 @@
++-----------------------------------------------------------------------------
++--
++-- Module      :  $Headers
++-- License     :  Apache 2.0
++--
++-- Stability   :  Experimental
++-- Portability :  Portable
++--
++-- | Arbitrary instance for an existentially quantified JSON type.
++--
++-----------------------------------------------------------------------------
++
++
++module Spec.Marlowe.Service.Random
++  ( -- * Testing
++    generateValue
++  ) where
++
++
++import Data.Jsonable (generateJsonable)
++import Spec.Marlowe.Semantics.Arbitrary ()
++import Spec.Marlowe.Service.Serialization (knownJsonTypes)
++import Spec.Marlowe.Service.Types (Seed(..), Size(..))
++import Test.QuickCheck (generate)
++import Test.QuickCheck.Gen (Gen(..))
++import Test.QuickCheck.Random (mkQCGen, newQCGen)
++
++import qualified Data.Aeson as A (Value)
++
++
++-- | Generate an arbitrary value.
++generateValue
++  :: Maybe Size  -- ^ The optional size paramater.
++  -> Maybe Seed  -- ^ The optional seed for the generator.
++  -> String  -- ^ The key for the type.
++  -> IO (Either String A.Value)  -- ^ Either the value or an error message.
++generateValue size seed =
++  mapM (generate' size seed) . generateJsonable knownJsonTypes
++  where
++    generate' :: Maybe Size -> Maybe Seed -> Gen a -> IO a
++    generate' (Just (Size s)) (Just (Seed r)) (MkGen g) = return $ g (mkQCGen r) s
++    generate' Nothing (Just (Seed r)) (MkGen g) = return $ g (mkQCGen r) 30
++    generate' (Just (Size s)) Nothing (MkGen g) = newQCGen >>= \r -> return $ g r s
++    generate' _ _ g = generate g
+diff --git a/marlowe-test/src/Spec/Marlowe/Service/Serialization.hs b/marlowe-test/src/Spec/Marlowe/Service/Serialization.hs
+new file mode 100644
+index 000000000..08a373eb9
+--- /dev/null
++++ b/marlowe-test/src/Spec/Marlowe/Service/Serialization.hs
+@@ -0,0 +1,106 @@
++-----------------------------------------------------------------------------
++--
++-- Module      :  $Headers
++-- License     :  Apache 2.0
++--
++-- Stability   :  Experimental
++-- Portability :  Portable
++--
++-- | Test of Marlowe's Cardano JSON implementation against the reference implementation.
++--
++-----------------------------------------------------------------------------
++
++
++{-# LANGUAGE OverloadedStrings #-}
++{-# LANGUAGE RecordWildCards #-}
++
++
++module Spec.Marlowe.Service.Serialization
++  ( -- * Types
++    SerializationResponse(..)
++  , knownJsonTypes
++    -- * Testing
++  , roundtripSerialization
++  ) where
++
++
++import Control.Applicative ((<|>))
++import Data.Aeson (FromJSON(..), ToJSON(..))
++import Data.Jsonable (JsonableType(JsonableType), KnownJsonable, isKnownJson, roundTripJsonable)
++import Data.Proxy (Proxy(..))
++import Spec.Marlowe.Semantics.Arbitrary ()
++
++import qualified Data.Aeson as A (Value, object, withObject, (.:), (.=))
++import qualified Language.Marlowe.Core.V1.Semantics as Marlowe
++import qualified Language.Marlowe.Core.V1.Semantics.Types as Marlowe
++
++
++-- | Response to a round-trip serialization request.
++data SerializationResponse =
++    -- | Success.
++    SerializationSuccess
++    {
++      valueReserialized :: A.Value  -- ^ The reserialized value.
++    }
++    -- | The type was not known.
++  | UnknownType
++    {
++      unknownType :: String  -- ^ The type.
++    }
++    -- | The deserialization or serialization failed.
++  | SerializationError
++    {
++      serializationError :: String  -- ^ The error message.
++    }
++  deriving (Eq, Ord, Read, Show)
++
++instance ToJSON SerializationResponse where
++  toJSON SerializationSuccess{..} = A.object . pure $ "serialization-success" A..= valueReserialized
++  toJSON UnknownType{..} = A.object . pure $ "unknown-type" A..= unknownType
++  toJSON SerializationError{..} = A.object . pure $ "serialization-error" A..= serializationError
++
++instance FromJSON SerializationResponse where
++  parseJSON =
++    A.withObject "SerializationResponse"
++      $ \o ->
++            (SerializationSuccess <$> o A..: "serialization-success")
++        <|> (UnknownType <$> o A..: "unknown-type")
++        <|> (SerializationError <$> o A..: "serialization-error")
++
++
++-- | Deserialize and then serialize a value.
++roundtripSerialization
++  :: String  -- ^ The key to the type.
++  -> A.Value  -- ^ The value.
++  -> SerializationResponse  -- ^ The result.
++roundtripSerialization typeSerialized valueSerialized =
++  if isKnownJson knownJsonTypes typeSerialized
++    then case roundTripJsonable knownJsonTypes typeSerialized valueSerialized of
++             Right valueReserialized -> SerializationSuccess{..}
++             Left serializationError -> SerializationError{..}
++    else UnknownType typeSerialized
++
++
++-- | List of known types that can be serialized and deserialized as JSON.
++knownJsonTypes :: KnownJsonable
++knownJsonTypes =
++  [
++    JsonableType "Core.Action" (Proxy :: Proxy Marlowe.Action)
++  , JsonableType "Core.Bound" (Proxy :: Proxy Marlowe.Bound)
++  , JsonableType "Core.Case" (Proxy :: Proxy (Marlowe.Case Marlowe.Contract))
++  , JsonableType "Core.ChoiceId" (Proxy :: Proxy Marlowe.ChoiceId)
++  , JsonableType "Core.Contract" (Proxy :: Proxy Marlowe.Contract)
++  , JsonableType "Core.Token" (Proxy :: Proxy Marlowe.Token)
++  , JsonableType "Core.Payee" (Proxy :: Proxy Marlowe.Payee)
++  , JsonableType "Core.Input" (Proxy :: Proxy Marlowe.Input)
++  , JsonableType "Core.Observation" (Proxy :: Proxy Marlowe.Observation)
++  , JsonableType "Core.Value" (Proxy :: Proxy (Marlowe.Value Marlowe.Observation))
++  , JsonableType "Core.Party" (Proxy :: Proxy Marlowe.Party)
++  , JsonableType "Core.State" (Proxy :: Proxy Marlowe.State)
++  , JsonableType "Core.Payment" (Proxy :: Proxy Marlowe.Payment)
++  , JsonableType "Core.Transaction" (Proxy :: Proxy Marlowe.TransactionInput)
++  , JsonableType "Core.TransactionOutput" (Proxy :: Proxy Marlowe.TransactionOutput)
++  , JsonableType "Core.TransactionWarning" (Proxy :: Proxy Marlowe.TransactionWarning)
++  , JsonableType "Core.TransactionError" (Proxy :: Proxy Marlowe.TransactionError)
++  , JsonableType "Core.IntervalError" (Proxy :: Proxy Marlowe.IntervalError)
++  ]
+diff --git a/marlowe-test/src/Spec/Marlowe/Service/Types.hs b/marlowe-test/src/Spec/Marlowe/Service/Types.hs
+new file mode 100644
+index 000000000..6bcbb7579
+--- /dev/null
++++ b/marlowe-test/src/Spec/Marlowe/Service/Types.hs
+@@ -0,0 +1,172 @@
++-----------------------------------------------------------------------------
++--
++-- Module      :  $Headers
++-- License     :  Apache 2.0
++--
++-- Stability   :  Experimental
++-- Portability :  Portable
++--
++-- | Marlowe types for the test service client.
++--
++-----------------------------------------------------------------------------
++
++
++{-# LANGUAGE GeneralizedNewtypeDeriving #-}
++{-# LANGUAGE LambdaCase #-}
++{-# LANGUAGE OverloadedStrings #-}
++{-# LANGUAGE RecordWildCards #-}
++
++
++module Spec.Marlowe.Service.Types
++  ( -- * Types
++    Request(..)
++  , Response(..)
++  , Seed(..)
++  , Size(..)
++  ) where
++
++
++import Control.Applicative ((<|>))
++import Data.Aeson (FromJSON(..), ToJSON(..))
++import Plutus.V1.Ledger.Api (POSIXTime(..))
++
++import qualified Data.Aeson as A (Value(Object, String), object, withObject, (.:), (.:?), (.=))
++import qualified Data.Aeson.Types as A (Parser)
++import qualified Language.Marlowe.Core.V1.Semantics as Marlowe
++import qualified Language.Marlowe.Core.V1.Semantics.Types as Marlowe
++
++newtype Size = Size Int deriving (Eq, Show, ToJSON, FromJSON)
++newtype Seed = Seed Int deriving (Eq, Show, ToJSON, FromJSON)
++
++data Request =
++    TestRoundtripSerialization
++    {
++      typeSerialized :: String
++    , valueSerialized :: A.Value
++    }
++  | GenerateRandomValue
++    {
++      typeSerialized :: String
++    , size :: Maybe Size
++    , seed :: Maybe Seed
++    }
++  | ComputeTransaction
++    {
++      transactionInput :: Marlowe.TransactionInput
++    , contract :: Marlowe.Contract
++    , state :: Marlowe.State
++    }
++  | PlayTrace
++    {
++      transactionInputs :: [Marlowe.TransactionInput]
++    , contract :: Marlowe.Contract
++    , initialTime :: POSIXTime
++    }
++  | EvalValue
++    {
++      environment :: Marlowe.Environment
++    , state :: Marlowe.State
++    , value :: Marlowe.Value Marlowe.Observation
++    }
++  | EvalObservation
++    {
++      environment :: Marlowe.Environment
++    , state :: Marlowe.State
++    , observation :: Marlowe.Observation
++    }
++    deriving (Eq, Show)
++
++instance FromJSON Request where
++  parseJSON =
++    A.withObject "Request"
++      $ \o ->
++        (o A..: "request" :: A.Parser String)
++          >>= \case
++            "test-roundtrip-serialization" -> TestRoundtripSerialization <$> o A..: "typeId" <*> o A..: "json"
++            "generate-random-value"        -> GenerateRandomValue <$> o A..: "typeId" <*> o A..:? "size" <*> o A..:? "seed"
++            "compute-transaction"          -> ComputeTransaction <$> o A..: "transactionInput" <*> o A..: "coreContract" <*> o A..: "state"
++            "playtrace"                    -> PlayTrace <$> o A..: "transactionInputs" <*> o A..: "coreContract" <*> (POSIXTime <$> o A..: "initialTime")
++            "eval-value"                   -> EvalValue <$> o A..: "environment" <*> o A..: "state" <*> o A..: "value"
++            "eval-observation"             -> EvalObservation <$> o A..: "environment" <*> o A..: "state" <*> o A..: "observation"
++            request                        -> fail $ "Request not understood: " <> show request <> "."
++
++instance ToJSON Request where
++  toJSON TestRoundtripSerialization{..} =
++    A.object
++      [
++        "request" A..= ("test-roundtrip-serialization" :: String)
++      , "typeId" A..= typeSerialized
++      , "json"  A..= valueSerialized
++      ]
++  toJSON GenerateRandomValue{..} =
++    A.object
++      [
++        "request" A..= ("generate-random-value" :: String)
++      , "typeId" A..= typeSerialized
++      , "size" A..= size
++      , "seed" A..= seed
++      ]
++  toJSON ComputeTransaction{..} =
++    A.object
++      [
++        "request" A..= ("compute-transaction" :: String)
++      , "transactionInput" A..= transactionInput
++      , "coreContract" A..= contract
++      , "state" A..= state
++      ]
++  toJSON PlayTrace{..} =
++    A.object
++      [
++        "request" A..= ("playtrace" :: String)
++      , "transactionInputs" A..= transactionInputs
++      , "coreContract" A..= contract
++      , "initialTime" A..= getPOSIXTime initialTime
++      ]
++  toJSON EvalValue{..} =
++    A.object
++      [
++        "request" A..= ("eval-value" :: String)
++      , "environment" A..= environment
++      , "state" A..= state
++      , "value" A..= value
++      ]
++  toJSON EvalObservation{..} =
++    A.object
++      [
++        "request" A..= ("eval-value" :: String)
++      , "environment" A..= environment
++      , "state" A..= state
++      , "observation" A..= observation
++      ]
++
++data Response =
++    InvalidRequest
++    {
++      errorInvalid :: String
++    }
++  | UnknownRequest
++  | RequestResponse
++    {
++      valueResponse :: A.Value
++    }
++  | RequestNotImplemented
++  | RequestTimeOut
++  | ResponseFailure
++    {
++      failureResponse :: String
++    }
++    deriving (Eq, Ord, Read, Show)
++
++instance FromJSON Response where
++    parseJSON (A.String "UnknownRequest") = return UnknownRequest
++    parseJSON (A.String "RequestNotImplemented") = return RequestNotImplemented
++    parseJSON (A.Object v) = (InvalidRequest <$> v A..: "invalid-request") <|> (RequestResponse <$> v A..: "request-response")
++    parseJSON _ = fail "Response must be either a string or an A.object"
++
++instance ToJSON Response where
++  toJSON UnknownRequest = "UnknownRequest"
++  toJSON RequestNotImplemented = "RequestNotImplemented"
++  toJSON RequestTimeOut = "RequestTimeOut"
++  toJSON (InvalidRequest err) = A.object . pure $ "invalid-request" A..= err
++  toJSON (RequestResponse res) = A.object . pure $ "request-response" A..= res
++  toJSON (ResponseFailure err) = A.object . pure $ "invalid-request" A..= err
+```

### 2023-03-29 response-to-audit-report.md
PLT-3382 Removes diff appendices for spec and tests.

--- 2023-03-27 response-to-audit-report.md
@@
 
-# 2 Findings
+## Executive Summary
+
+TODO: @bwbush will add text here after everything else is written.
+
+## Acknowledgement
+
+TODO: @bwbush will add text thanking the audit team.
+
+## Overview of Isabelle Changes
+
+## Overview of Changes to Marlowe-Cardano Specification and Haskell (Plutus) Implementation
+
+The audit report's main and high-severity concerns for the Plutus validators for Marlowe on Cardano fell into to four clusters:
+
+1. Incorrect handling of negative deposits
+    - "2.1.1 Negative deposits allow stealing funds"
+2. Incomplete prevention of double satisfaction
+    - "2.1.2 Contracts vulnerable to double satisfaction attacks"
+3. Lax enforcement of invariants in Marlowe state
+    - "2.1.7 Positive balances are not checked for the output state"
+    - "2.1.8 Non-validated Marlowe states"
+    - "2.1.9 Total balance of ending state uncomputed"
+    - "2.1.10 Unchecked ending balance"
+4. Difference semantics of association lists in Isabelle and Plutus
+    - "2.1.12 Different insertion functions used in Isabelle and Haskell code"
+
+Incorrect handling of negative deposits (item 1 above) was mitigated by the addition of a guard against this in Marlowe's Plutus validator so that its behavior was consistent with Marlowe semantics.
+Property-based tests were also added to check the correctness of the mitigation.
+
+Although the Marlowe validator had prevented double-satisfaction among multiple copies of the Marlowe validator script running in the same transaction, it did not prevent double-satisfaction in cases where the Marlowe validator run alongside another Plutus validator in the same transaction (item 2 above).
+Double satisfaction has now been prevented by enforcing that the Marlowe validator is the only Plutus script that runs in transactions that make payments to parties.
+This allows Marlowe contracts to coordinate with other Plutus scripts, but only under conditions where double satisfaction is impossible.
+Once again, property-based tests were added to check the correctness of this mitigation.
+
+The Marlowe validator had made optimistic assumptions about its own correct operation and hence did not check certain invariants in order to save on Plutus execution costs (item 3 above).
+The Marlowe semantics validator has now been thoroughly hardened against corruption of initial or final state so that it ensures that the three state-invariants of positive accounts, non-duplication of state entries (accounts, choices, and bound values), and a total internal value matching the script UTxO's value.
+This hardening was completed without unduly increasing the size (in bytes) of the validator or its Plutus execution cost: such was verified with simulated and on-chain measurement of execution costs for a library of real-life Marlowe contracts.
+Notably, the enforcement of the final invariants protects against funds becoming permanently locked in a Marlowe contract due to a garbled state: even in cases where the implementation of Marlowe semantics or parts of the Plutus validator are flawed, at least some of the execution paths of a Marlowe contract remain viable, so funds could be released even under nominally impossible but catastrophic difficulties.
+The property-based test suite was significantly enhanced to
+
+Mitigation of the difference between association lists (item 4 above) in Isabelle (`MList`) and Plutus (`AssocMap`) was handled by the aforementioned enforcement of invariants, by line-by-line code inspection and annotation, and by thoroughly enhanced property-based testing.
+Work is in progress on a formal proof of the equivalence of `MList` and `AssocMap` under pre-conditions that hold within Marlowe semantics.
+Note that porting Isabelle's `MList` implementation to Plutus would have enlarged the Marlowe semantics validator by at least 2000 bytes and rendered it too costly to execute within the Cardano ledger's present execution-cost limits.
+
+The audit report for Marlowe on Cardano was based on the commit hash [523f3d56](https://github.com/input-output-hk/marlowe-cardano/commit/523f3d56f22bf992ddb0b0c8a52bb7a5a188f9e9).
+The revisions and mitigations discussed here apply to [69468d66](https://github.com/input-output-hk/marlowe-cardano/commit/69468d6623e24a4ccd6659e378ae012c38ca01ce) of that repository.
+The appendix to this report list the differences between pre- and post-audit validator code for Marlowe on Cardano.
+
+The medium-severity concerns in the audit report center around missing tests or name shadowing.
+With the exception of "2.7.1 More precise failure checks", these have all been remedied either by code changes (in the case of the shadowing) or by the addition of a significant augmentation of the property-based test suite.
+The audit's low-severity concerns generally relate to insufficiently detailed text in the Marlowe Cardano specification, the need for more elaborate comments in the code, naming of variable bindings, or typographical errors.
+Nearly all of these finds have been addressed, with only the exception of a few cosmetic recommendations that were not adopted.
+
+# 2 Responses to Specific Findings
 
@@
 > The test `checkFixInterval` is never instantiated with an invalid interval that is in the past, meaning the function `fixInterval` is never tested for that case.
+
+Commit [811f048b](https://github.com/input-output-hk/marlowe-cardano/commit/811f048ba93db10972641c0c235864db1d6014e7) adds a property-based test for an invalid in terval in the past.
+
+## Appendix: Post-Audit Changes in Marlowe Validator
+
+```bash
+git diff 523f3d56f22bf992ddb0b0c8a52bb7a5a188f9e9 69468d6623e24a4ccd6659e378ae012c38ca01ce marlowe/src/Language/Marlowe/{Core/V1/Semantics*,Scripts.hs}
+```
+
+```diff
+diff --git a/marlowe/src/Language/Marlowe/Core/V1/Semantics.hs b/marlowe/src/Language/Marlowe/Core/V1/Semantics.hs
+index e31b8c478..82b57b903 100644
+--- a/marlowe/src/Language/Marlowe/Core/V1/Semantics.hs
++++ b/marlowe/src/Language/Marlowe/Core/V1/Semantics.hs
+@@ -94,12 +94,12 @@ module Language.Marlowe.Core.V1.Semantics
+   , TransactionError(..)
+   , TransactionWarning(..)
+     -- * Utility Functions
++  , allBalancesArePositive
+   , contractLifespanUpperBound
+   , isClose
+   , notClose
+   , paymentMoney
+   , totalBalance
+-  , validateBalances
+     -- * Serialisation
+   , currencySymbolFromJSON
+   , currencySymbolToJSON
+@@ -217,6 +217,26 @@ import Text.PrettyPrint.Leijen (comma, hang, lbrace, line, rbrace, space, text,
+ data Payment = Payment AccountId Payee Token Integer
+   deriving stock (Haskell.Eq, Haskell.Show)
+
++instance ToJSON Payment where
++  toJSON (Payment accountId payee token amount) =
++    object
++      [
++        "payment_from" .= accountId
++      , "to" .= payee
++      , "token" .= token
++      , "amount" .= amount
++      ]
++
++instance FromJSON Payment where
++  parseJSON =
++    withObject "Payment"
++      $ \o ->
++        Payment
++          <$> o .: "payment_from"
++          <*> o .: "to"
++          <*> o .: "token"
++          <*> o .: "amount"
++
+
+ -- | Extract the money value from a payment.
+ paymentMoney :: Payment -> Money
+@@ -293,7 +313,28 @@ data TransactionError = TEAmbiguousTimeIntervalError
+                       | TEUselessTransaction
+                       | TEHashMismatch
+   deriving stock (Haskell.Show, Generic, Haskell.Eq)
+-  deriving anyclass (ToJSON, FromJSON)
++
++instance ToJSON TransactionError where
++  toJSON TEAmbiguousTimeIntervalError = JSON.String "TEAmbiguousTimeIntervalError"
++  toJSON TEApplyNoMatchError = JSON.String "TEApplyNoMatchError"
++  toJSON (TEIntervalError intervalError) = object ["error" .= JSON.String "TEIntervalError", "context" .= intervalError]
++  toJSON TEUselessTransaction = JSON.String "TEUselessTransaction"
++  toJSON TEHashMismatch = JSON.String "TEHashMismatch"
++
++instance FromJSON TransactionError where
++  parseJSON (JSON.String s) =
++    case s of
++      "TEAmbiguousTimeIntervalError" -> return TEAmbiguousTimeIntervalError
++      "TEApplyNoMatchError" -> return TEApplyNoMatchError
++      "TEUselessTransaction" -> return TEUselessTransaction
++      "TEHashMismatch" -> return TEHashMismatch
++      _ -> Haskell.fail "Failed parsing TransactionError"
++  parseJSON (JSON.Object o) = do
++                                err <- o .: "error"
++                                if err Haskell.== ("TEIntervalError" :: Haskell.String)
++                                  then TEIntervalError <$> o .: "context"
++                                  else Haskell.fail "Failed parsing TransactionError"
++  parseJSON _ = Haskell.fail "Failed parsing TransactionError"
+
+ -- | Marlowe transaction input.
+@@ -319,6 +360,31 @@ data TransactionOutput =
+     | Error TransactionError
+   deriving stock (Haskell.Show)
+
++instance ToJSON TransactionOutput where
++  toJSON TransactionOutput{..} =
++    object
++      [
++        "warnings" .= txOutWarnings
++      , "payments" .= txOutPayments
++      , "state" .= txOutState
++      , "contract" .= txOutContract
++      ]
++  toJSON (Error err) = object ["transaction_error" .= err]
++
++instance FromJSON TransactionOutput where
++  parseJSON =
++    withObject "TransactionOutput"
++      $ \o ->
++        let
++          asTransactionOutput =
++            TransactionOutput
++              <$> o .: "warnings"
++              <*> o .: "payments"
++              <*> o .: "state"
++              <*> o .: "contract"
++          asError = Error <$> o .: "transaction_error"
++        in
++          asTransactionOutput <|> asError
+
+ -- | Parse a validator hash from JSON.
+ validatorHashFromJSON :: JSON.Value -> Parser ValidatorHash
+@@ -404,12 +470,22 @@ evalValue env state value = let
+                                    then 0
+                                    else n `Builtins.quotientInteger` d
+         ChoiceValue choiceId ->
++            -- SCP-5126: Given the precondition that `choices` contains no
++            -- duplicate entries, this lookup behaves identically to
++            -- Marlowe's Isabelle semantics given the precondition that
++            -- the initial state's `choices` in Isabelle was sorted and
++            -- did not contain duplicate entries.
+             case Map.lookup choiceId (choices state) of
+                 Just x  -> x
+                 Nothing -> 0
+         TimeIntervalStart    -> getPOSIXTime (fst (timeInterval env))
+         TimeIntervalEnd      -> getPOSIXTime (snd (timeInterval env))
+         UseValue valId       ->
++            -- SCP-5126: Given the precondition that `boundValues` contains
++            -- no duplicate entries, this lookup behaves identically to
++            -- Marlowe's Isabelle semantics given the precondition that
++            -- the initial state's `boundValues` in Isabelle was sorted
++            -- and did not contain duplicate entries.
+             case Map.lookup valId (boundValues state) of
+                 Just x  -> x
+                 Nothing -> 0
+@@ -425,6 +501,11 @@ evalObservation env state obs = let
+         AndObs lhs rhs          -> evalObs lhs && evalObs rhs
+         OrObs lhs rhs           -> evalObs lhs || evalObs rhs
+         NotObs subObs           -> not (evalObs subObs)
++                                   -- SCP-5126: Given the precondition that `choices` contains no
++                                   -- duplicate entries, this membership test behaves identically
++                                   -- to Marlowe's Isabelle semantics given the precondition that
++                                   -- the initial state's `choices` in Isabelle was sorted and did
++                                   -- not contain duplicate entries.
+         ChoseSomething choiceId -> choiceId `Map.member` choices state
+         ValueGE lhs rhs         -> evalVal lhs >= evalVal rhs
+         ValueGT lhs rhs         -> evalVal lhs > evalVal rhs
+@@ -439,6 +520,12 @@ evalObservation env state obs = let
+ refundOne :: Accounts -> Maybe ((Party, Token, Integer), Accounts)
+ refundOne accounts = case Map.toList accounts of
+     [] -> Nothing
++    -- SCP-5126: The return value of this function differs from
++    -- Isabelle semantics in that it returns the least-recently
++    -- added account-token combination rather than the first
++    -- lexicographically ordered one. Also, the sequence
++    -- `Map.fromList . tail . Map.toList` preserves the
++    -- invariants of order and non-duplication.
+     ((accId, token), balance) : rest ->
+         if balance > 0
+         then Just ((accId, token, balance), Map.fromList rest)
+@@ -447,18 +534,30 @@ refundOne accounts = case Map.toList accounts of
+
+ -- | Obtains the amount of money available an account.
+ moneyInAccount :: AccountId -> Token -> Accounts -> Integer
+-moneyInAccount accId token accounts = case Map.lookup (accId, token) accounts of
+-    Just x  -> x
+-    Nothing -> 0
++moneyInAccount accId token accounts =
++    -- SCP-5126: Given the precondition that `accounts` contains
++    -- no duplicate entries, this lookup behaves identically to
++    -- Marlowe's Isabelle semantics given the precondition that
++    -- the initial state's `accounts` in Isabelle was sorted and
++    -- did not contain duplicate entries.
++    case Map.lookup (accId, token) accounts of
++      Just x  -> x
++      Nothing -> 0
+
+ -- | Sets the amount of money available in an account.
+ updateMoneyInAccount :: AccountId -> Token -> Integer -> Accounts -> Accounts
+ updateMoneyInAccount accId token amount =
++    -- SCP-5126: Given the precondition that `accounts` contains
++    -- no duplicate entries, this deletion or insertion behaves
++    -- identically (aside from internal ordering) to Marlowe's
++    -- Isabelle semantics given the precondition that the initial
++    -- state's `accounts` in Isabelle was sorted and did not
++    -- contain duplicate entries.
+     if amount <= 0 then Map.delete (accId, token) else Map.insert (accId, token) amount
+
+--- | Add the given amount of money to an accoun (only if it is positive).
++-- | Add the given amount of money to an account (only if it is positive).
+ --   Return the updated Map.
+ addMoneyToAccount :: AccountId -> Token -> Integer -> Accounts -> Accounts
+ addMoneyToAccount accId token amount accounts = let
+@@ -482,6 +581,14 @@ giveMoney accountId payee token amount accounts = let
+ reduceContractStep :: Environment -> State -> Contract -> ReduceStepResult
+ reduceContractStep env state contract = case contract of
+
++    -- SCP-5126: Although `refundOne` refunds accounts-token combinations
++    -- in least-recently-added order and Isabelle semantics requires that
++    -- they be refunded in lexicographic order, `reduceContractUntilQuiescent`
++    -- ensures that the `Close` pattern will be executed until `accounts`
++    -- is empty. Thus, the net difference between the behavior here and the
++    -- Isabelle semantics is that the `ContractQuiescent` resulting from
++    -- `reduceContractUntilQuiescent` will contain payments in a different
++    -- order.
+     Close -> case refundOne (accounts state) of
+         Just ((party, token, amount), newAccounts) -> let
+             newState = state { accounts = newAccounts }
+@@ -522,7 +629,17 @@ reduceContractStep env state contract = case contract of
+     Let valId val cont -> let
+         evaluatedValue = evalValue env state val
+         boundVals = boundValues state
++        -- SCP-5126: Given the precondition that `boundValues` contains
++        -- no duplicate entries, this insertion behaves identically
++        -- (aside from internal ordering) to Marlowe's Isabelle semantics
++        -- given the precondition that the initial state's `boundValues`
++        -- in Isabelle was sorted and did not contain duplicate entries.
+         newState = state { boundValues = Map.insert valId evaluatedValue boundVals }
++        -- SCP-5126: Given the precondition that `boundValues` contains
++        -- no duplicate entries, this lookup behaves identically to
++        -- Marlowe's Isabelle semantics given the precondition that the
++        -- initial state's `boundValues` in Isabelle was sorted and did
++        -- not contain duplicate entries.
+         warn = case Map.lookup valId boundVals of
+               Just oldVal -> ReduceShadowing valId oldVal evaluatedValue
+               Nothing     -> ReduceNoWarning
+@@ -575,6 +692,11 @@ applyAction env state (IDeposit accId1 party1 tok1 amount) (Deposit accId2 party
+     else NotAppliedAction
+ applyAction _ state (IChoice choId1 choice) (Choice choId2 bounds) =
+     if choId1 == choId2 && inBounds choice bounds
++    -- SCP-5126: Given the precondition that `choices` contains no
++    -- duplicate entries, this insertion behaves identically (aside
++    -- from internal ordering) to Marlowe's Isabelle semantics
++    -- given the precondition that the initial state's `choices`
++    -- in Isabelle was sorted and did not contain duplicate entries.
+     then let newState = state { choices = Map.insert choId1 choice (choices state) }
+          in AppliedAction ApplyNoWarning newState
+     else NotAppliedAction
+@@ -592,18 +714,20 @@ getContinuation (MerkleizedInput _ inputContinuationHash continuation) (Merkleiz
+     else Nothing
+ getContinuation _ _ = Nothing
+
+--- | Try to apply an input to a list of cases.
++-- | Try to apply an input to a list of cases, accepting the first match.
+ applyCases :: Environment -> State -> Input -> [Case Contract] -> ApplyResult
+-applyCases env state input (headCase : tailCase) =
++applyCases env state input (headCase : tailCases) =
+     let inputContent = getInputContent input :: InputContent
+         action = getAction headCase :: Action
+         maybeContinuation = getContinuation input headCase :: Maybe Contract
+     in case applyAction env state inputContent action of
+          AppliedAction warning newState ->
++           -- Note that this differs from Isabelle semantics because
++           -- the Cardano semantics includes merkleization.
+            case maybeContinuation of
+              Just continuation -> Applied warning newState continuation
+              Nothing           -> ApplyHashMismatch
+-         NotAppliedAction -> applyCases env state input tailCase
++         NotAppliedAction -> applyCases env state input tailCases
+ applyCases _ _ _ [] = ApplyNoMatchError
+
+@@ -615,7 +739,7 @@ applyInput _ _ _ _                          = ApplyNoMatchError
+
+ -- | Propagate 'ReduceWarning' to 'TransactionWarning'.
+ convertReduceWarnings :: [ReduceWarning] -> [TransactionWarning]
+-convertReduceWarnings = foldr (\warn acc -> case warn of
++convertReduceWarnings = foldr (\warn acc -> case warn of  -- Note that `foldr` is used here for efficiency, differing from Isabelle.
+     ReduceNoWarning -> acc
+     ReduceNonPositivePay accId payee tok amount ->
+         TransactionNonPositivePay accId payee tok amount : acc
+@@ -655,12 +779,12 @@ applyAllInputs env state contract inputs = let
+                     curState
+                     cont
+                 (input : rest) -> case applyInput env curState input cont of
+-                    Applied applyWarn newState cont ->
++                    Applied applyWarn newState cont' ->
+                         applyAllLoop
+                             True
+                             env
+                             newState
+-                            cont
++                            cont'
+                             rest
+                             (warnings' ++ convertApplyWarning applyWarn)
+                             payments'
+@@ -759,8 +883,8 @@ totalBalance accounts = foldMap
+
+ -- | Check that all accounts have positive balance.
+-validateBalances :: State -> Bool
+-validateBalances State{..} = all (\(_, balance) -> balance > 0) (Map.toList accounts)
++allBalancesArePositive :: State -> Bool
++allBalancesArePositive State{..} = all (\(_, balance) -> balance > 0) (Map.toList accounts)
+
+ instance FromJSON TransactionInput where
+@@ -845,20 +969,26 @@ instance Eq Payment where
+ instance Eq ReduceWarning where
+     {-# INLINABLE (==) #-}
+     ReduceNoWarning == ReduceNoWarning = True
+-    (ReduceNonPositivePay acc1 p1 tn1 a1) == (ReduceNonPositivePay acc2 p2 tn2 a2) =
++    ReduceNoWarning == _ = False
++    ReduceNonPositivePay acc1 p1 tn1 a1 == ReduceNonPositivePay acc2 p2 tn2 a2 =
+         acc1 == acc2 && p1 == p2 && tn1 == tn2 && a1 == a2
+-    (ReducePartialPay acc1 p1 tn1 a1 e1) == (ReducePartialPay acc2 p2 tn2 a2 e2) =
++    ReduceNonPositivePay{} == _ = False
++    ReducePartialPay acc1 p1 tn1 a1 e1 == ReducePartialPay acc2 p2 tn2 a2 e2 =
+         acc1 == acc2 && p1 == p2 && tn1 == tn2 && a1 == a2 && e1 == e2
+-    (ReduceShadowing v1 old1 new1) == (ReduceShadowing v2 old2 new2) =
++    ReducePartialPay{} == _ = False
++    ReduceShadowing v1 old1 new1 == ReduceShadowing v2 old2 new2 =
+         v1 == v2 && old1 == old2 && new1 == new2
+-    _ == _ = False
++    ReduceShadowing{} == _ = False
++    ReduceAssertionFailed == ReduceAssertionFailed = True
++    ReduceAssertionFailed == _ = False
+
+ instance Eq ReduceEffect where
+     {-# INLINABLE (==) #-}
+     ReduceNoPayment == ReduceNoPayment           = True
++    ReduceNoPayment == _                         = False
+     ReduceWithPayment p1 == ReduceWithPayment p2 = p1 == p2
+-    _ == _                                       = False
++    ReduceWithPayment _ == _                     = False
+
+ -- Lifting data types to Plutus Core
+diff --git a/marlowe/src/Language/Marlowe/Core/V1/Semantics/Types.hs b/marlowe/src/Language/Marlowe/Core/V1/Semantics/Types.hs
+index a333b8fbb..bf495773b 100644
+--- a/marlowe/src/Language/Marlowe/Core/V1/Semantics/Types.hs
++++ b/marlowe/src/Language/Marlowe/Core/V1/Semantics/Types.hs
+@@ -108,7 +108,7 @@ import PlutusTx.Lift (makeLift)
+ import PlutusTx.Prelude hiding (encodeUtf8, mapM, (<$>), (<*>), (<>))
+ import Prelude (mapM, (<$>))
+ import qualified Prelude as Haskell
+-import Text.PrettyPrint.Leijen (text)
++import Text.PrettyPrint.Leijen (parens, text)
+
+ -- Functions that used in Plutus Core must be inlinable,
+@@ -126,8 +126,8 @@ data Party =
+   deriving stock (Generic,Haskell.Eq,Haskell.Ord)
+
+ instance Pretty Party where
+-  prettyFragment (Address network address) = text $ "Address " ++ Haskell.show (serialiseAddressBech32 network address)
+-  prettyFragment (Role role)               = text $ "Role "    ++ Haskell.show role
++  prettyFragment (Address network address) = parens $ text "Address " Haskell.<> prettyFragment (serialiseAddressBech32 network address)
++  prettyFragment (Role role)               = parens $ text "Role "    Haskell.<> prettyFragment role
+
+ instance Haskell.Show Party where
+   showsPrec _ (Address network address) = Haskell.showsPrec 11 $ Haskell.show (serialiseAddressBech32 network address)
+@@ -307,6 +307,13 @@ data State = State { accounts    :: Accounts
+ newtype Environment = Environment { timeInterval :: TimeInterval }
+   deriving stock (Haskell.Show,Haskell.Eq,Haskell.Ord)
+
++instance FromJSON Environment where
++  parseJSON = withObject "Environment"
++    (\v -> Environment <$> (posixIntervalFromJSON =<< v .: "timeInterval"))
++
++instance ToJSON Environment where
++  toJSON Environment{..} = object
++    [ "timeInterval" .= posixIntervalToJSON timeInterval]
+
+ -- | Input for a Marlowe contract. Correspond to expected 'Action's.
+ data InputContent = IDeposit AccountId Party Token Integer
+@@ -389,23 +396,24 @@ data IntervalError = InvalidInterval TimeInterval
+                    | IntervalInPastError POSIXTime TimeInterval
+   deriving stock (Haskell.Show, Generic, Haskell.Eq)
+
++
+ instance ToJSON IntervalError where
+   toJSON (InvalidInterval (s, e)) = A.object
+-    [ ("invalidInterval", toJSON (posixTimeToJSON s, posixTimeToJSON e)) ]
++    [ ("invalidInterval", A.object [("from", posixTimeToJSON s), ("to", posixTimeToJSON e)]) ]
+   toJSON (IntervalInPastError t (s, e)) = A.object
+-    [ ("intervalInPastError", toJSON (posixTimeToJSON t, posixTimeToJSON s, posixTimeToJSON e)) ]
++    [ ("intervalInPastError", A.object [("minTime", posixTimeToJSON t), ("from", posixTimeToJSON s), ("to", posixTimeToJSON e)]) ]
+
+ instance FromJSON IntervalError where
+   parseJSON (JSON.Object v) =
+     let
+       parseInvalidInterval = do
+-        (s, e) <- v .: "invalidInterval"
+-        InvalidInterval <$> ((,) <$> posixTimeFromJSON s <*> posixTimeFromJSON e)
++        o <- v .: "invalidInterval"
++        InvalidInterval <$> posixIntervalFromJSON o
+       parseIntervalInPastError = do
+-        (t, s, e) <- v .: "intervalInPastError"
++        o <- v .: "intervalInPastError"
+         IntervalInPastError
+-          <$> posixTimeFromJSON t
+-          <*> ((,) <$> posixTimeFromJSON s <*> posixTimeFromJSON e)
++          <$> (posixTimeFromJSON =<< o .: "minTime")
++          <*> posixIntervalFromJSON o
+     in
+       parseIntervalInPastError <|> parseInvalidInterval
+   parseJSON invalid =
+@@ -422,11 +430,18 @@ posixTimeFromJSON = \case
+   invalid ->
+       JSON.prependFailure "parsing POSIXTime failed, " (JSON.typeMismatch "Number" invalid)
+
++posixIntervalFromJSON :: A.Object -> Parser TimeInterval
++posixIntervalFromJSON o = (,) <$> (posixTimeFromJSON =<< o .: "from") <*> (posixTimeFromJSON =<< o .: "to")
+
+ -- | Serialise time as a JSON value.
+ posixTimeToJSON :: POSIXTime -> JSON.Value
+ posixTimeToJSON (POSIXTime n) = JSON.Number $ scientific n 0
+
++posixIntervalToJSON :: TimeInterval -> JSON.Value
++posixIntervalToJSON (from, to) = object
++  [ "from" .= posixTimeToJSON from
++  , "to" .= posixTimeToJSON to
++  ]
+
+ -- | Result of 'fixInterval'
+ data IntervalResult = IntervalTrimmed Environment State
+@@ -778,23 +793,24 @@ instance ToJSON Contract where
+ instance Eq Party where
+     {-# INLINABLE (==) #-}
+     Address n1 a1 == Address n2 a2 = n1 == n2 && a1 == a2
++    Address _ _   == _             = False
+     Role r1       == Role r2       = r1 == r2
+-    _             == _             = False
++    Role _        == _             = False
+
+ instance Eq ChoiceId where
+     {-# INLINABLE (==) #-}
+-    (ChoiceId n1 p1) == (ChoiceId n2 p2) = n1 == n2 && p1 == p2
++    ChoiceId n1 p1 == ChoiceId n2 p2 = n1 == n2 && p1 == p2
+
+ instance Eq Token where
+     {-# INLINABLE (==) #-}
+-    (Token n1 p1) == (Token n2 p2) = n1 == n2 && p1 == p2
++    Token n1 p1 == Token n2 p2 = n1 == n2 && p1 == p2
+
+ instance Eq ValueId where
+     {-# INLINABLE (==) #-}
+-    (ValueId n1) == (ValueId n2) = n1 == n2
++    ValueId n1 == ValueId n2 = n1 == n2
+
+ instance Pretty ValueId where
+@@ -804,78 +820,107 @@ instance Pretty ValueId where
+ instance Eq Payee where
+     {-# INLINABLE (==) #-}
+     Account acc1 == Account acc2 = acc1 == acc2
++    Account{}    == _            = False
+     Party p1 == Party p2         = p1 == p2
+-    _ == _                       = False
++    Party{}  == _                = False
+
+ instance Eq a => Eq (Value a) where
+     {-# INLINABLE (==) #-}
+-    AvailableMoney acc1 tok1 == AvailableMoney acc2 tok2 =
+-        acc1 == acc2 && tok1 == tok2
++    AvailableMoney acc1 tok1 == AvailableMoney acc2 tok2 = acc1 == acc2 && tok1 == tok2
++    AvailableMoney _    _    == _                        = False
+     Constant i1 == Constant i2 = i1 == i2
++    Constant{}  == _           = False
+     NegValue val1 == NegValue val2 = val1 == val2
++    NegValue{}    == _             = False
+     AddValue val1 val2 == AddValue val3 val4 = val1 == val3 && val2 == val4
++    AddValue{}         == _                  = False
+     SubValue val1 val2 == SubValue val3 val4 = val1 == val3 && val2 == val4
++    SubValue{}         == _                  = False
+     MulValue val1 val2 == MulValue val3 val4 = val1 == val3 && val2 == val4
++    MulValue{}         == _                  = False
+     DivValue val1 val2 == DivValue val3 val4 = val1 == val3 && val2 == val4
++    DivValue{}         == _                  = False
+     ChoiceValue cid1 == ChoiceValue cid2 = cid1 == cid2
++    ChoiceValue{}    == _                = False
+     TimeIntervalStart == TimeIntervalStart = True
+-    TimeIntervalEnd   == TimeIntervalEnd   = True
++    TimeIntervalStart == _                 = False
++    TimeIntervalEnd == TimeIntervalEnd = True
++    TimeIntervalEnd == _               = False
+     UseValue val1 == UseValue val2 = val1 == val2
+-    Cond obs1 thn1 els1 == Cond obs2 thn2 els2 =  obs1 == obs2 && thn1 == thn2 && els1 == els2
+-    _ == _ = False
++    UseValue{}    == _             = False
++    Cond obs1 thn1 els1 == Cond obs2 thn2 els2 = obs1 == obs2 && thn1 == thn2 && els1 == els2
++    Cond{}              == _                   = False
+
+ instance Eq Observation where
+     {-# INLINABLE (==) #-}
+     AndObs o1l o2l == AndObs o1r o2r           = o1l == o1r && o2l == o2r
++    AndObs{}       == _                        = False
+     OrObs  o1l o2l == OrObs  o1r o2r           = o1l == o1r && o2l == o2r
++    OrObs{}        == _                        = False
+     NotObs ol == NotObs or                     = ol == or
++    NotObs{}  == _                             = False
+     ChoseSomething cid1 == ChoseSomething cid2 = cid1 == cid2
++    ChoseSomething _    == _                   = False
+     ValueGE v1l v2l == ValueGE v1r v2r         = v1l == v1r && v2l == v2r
++    ValueGE{}       == _                       = False
+     ValueGT v1l v2l == ValueGT v1r v2r         = v1l == v1r && v2l == v2r
++    ValueGT{}       == _                       = False
+     ValueLT v1l v2l == ValueLT v1r v2r         = v1l == v1r && v2l == v2r
++    ValueLT{}       == _                       = False
+     ValueLE v1l v2l == ValueLE v1r v2r         = v1l == v1r && v2l == v2r
++    ValueLE{}       == _                       = False
+     ValueEQ v1l v2l == ValueEQ v1r v2r         = v1l == v1r && v2l == v2r
++    ValueEQ{}       == _                       = False
+     TrueObs  == TrueObs                        = True
++    TrueObs  == _                              = False
+     FalseObs == FalseObs                       = True
+-    _ == _                                     = False
++    FalseObs == _                              = False
+
+ instance Eq Action where
+     {-# INLINABLE (==) #-}
+     Deposit acc1 party1 tok1 val1 == Deposit acc2 party2 tok2 val2 =
+         acc1 == acc2 && party1 == party2 && tok1 == tok2 && val1 == val2
++    Deposit{}       == _ = False
+     Choice cid1 bounds1 == Choice cid2 bounds2 =
+         cid1 == cid2 && length bounds1 == length bounds2 && let
+             bounds = zip bounds1 bounds2
+             checkBound (Bound low1 high1, Bound low2 high2) = low1 == low2 && high1 == high2
+             in all checkBound bounds
++    Choice{}   == _ = False
+     Notify obs1 == Notify obs2 = obs1 == obs2
+-    _ == _ = False
++    Notify{} == _ = False
+
+ instance Eq a => Eq (Case a) where
+     {-# INLINABLE (==) #-}
+     Case acl cl == Case acr cr                       = acl == acr && cl == cr
++    Case{}      == _                                 = False
+     MerkleizedCase acl bsl == MerkleizedCase acr bsr = acl == acr && bsl == bsr
+-    _ == _                                           = False
++    MerkleizedCase{}       == _                      = False
+
+ instance Eq Contract where
+     {-# INLINABLE (==) #-}
+     Close == Close = True
++    Close == _ = False
+     Pay acc1 payee1 tok1 value1 cont1 == Pay acc2 payee2 tok2 value2 cont2 =
+         acc1 == acc2 && payee1 == payee2 && tok1 == tok2 && value1 == value2 && cont1 == cont2
++    Pay{} == _ = False
+     If obs1 cont1 cont2 == If obs2 cont3 cont4 =
+         obs1 == obs2 && cont1 == cont3 && cont2 == cont4
+-    When cases1 timeout1 cont1 == When cases2 timeout2 cont2 =
++    If{} == _ = False
++    When cases1 timeout1 cont1 == When cases2 timeout2 cont2 =  -- The sequences of tests are ordered for efficiency.
+         timeout1 == timeout2 && cont1 == cont2
+         && length cases1 == length cases2
+         && and (zipWith (==) cases1 cases2)
++    When{} == _ = False
+     Let valId1 val1 cont1 == Let valId2 val2 cont2 =
+         valId1 == valId2 && val1 == val2 && cont1 == cont2
++    Let{} == _ = False
+     Assert obs1 cont1 == Assert obs2 cont2 = obs1 == obs2 && cont1 == cont2
+-    _ == _ = False
++    Assert{}  == _ = False
+
+ instance Eq State where
+diff --git a/marlowe/src/Language/Marlowe/Scripts.hs b/marlowe/src/Language/Marlowe/Scripts.hs
+index 6027202b5..aa397d5fa 100644
+--- a/marlowe/src/Language/Marlowe/Scripts.hs
++++ b/marlowe/src/Language/Marlowe/Scripts.hs
+@@ -47,7 +47,9 @@ module Language.Marlowe.Scripts
+   , alternateMarloweValidator
+   , alternateMarloweValidatorHash
+   , marloweValidator
++  , marloweValidatorCompiled
+   , marloweValidatorHash
++  , mkMarloweValidator
+     -- * Payout Validator
+   , TypedRolePayoutValidator
+   , rolePayoutValidator
+@@ -61,8 +63,9 @@ import GHC.Generics (Generic)
+ import Language.Marlowe.Core.V1.Semantics as Semantics
+ import Language.Marlowe.Core.V1.Semantics.Types as Semantics
+ import Language.Marlowe.Pretty (Pretty(..))
++import Ledger.Typed.Scripts (unsafeMkTypedValidator)
+ import qualified Plutus.Script.Utils.Typed as Scripts
+-import Plutus.Script.Utils.V2.Typed.Scripts (mkTypedValidator, mkUntypedValidator)
++import Plutus.Script.Utils.V2.Typed.Scripts (mkTypedValidator)
+ import qualified Plutus.Script.Utils.V2.Typed.Scripts as Scripts
+ import qualified Plutus.V1.Ledger.Address as Address (scriptHashAddress)
+ import qualified Plutus.V1.Ledger.Value as Val
+@@ -92,11 +95,22 @@ import PlutusTx (makeIsDataIndexed, makeLift)
+ import qualified PlutusTx
+ import qualified PlutusTx.AssocMap as AssocMap
+ import PlutusTx.Plugin ()
+-import PlutusTx.Prelude as PlutusTxPrelude
++import PlutusTx.Prelude as PlutusTxPrelude hiding (traceError, traceIfFalse)
+ import qualified Prelude as Haskell
+ import Unsafe.Coerce (unsafeCoerce)
+
++-- Suppress traces, in order to save bytes.
++
++{-# INLINABLE traceError #-}
++traceError :: BuiltinString -> a
++traceError _ = error ()
++
++{-# INLINABLE traceIfFalse #-}
++traceIfFalse :: BuiltinString -> a -> a
++traceIfFalse _ = id
++
++
+ -- | Input to a Marlowe transaction.
+ type MarloweInput = [MarloweTxInput]
+
+@@ -136,7 +150,7 @@ rolePayoutValidator = mkTypedValidator @TypedRolePayoutValidator
+   $$(PlutusTx.compile [|| mkRolePayoutValidator ||])
+   $$(PlutusTx.compile [|| wrap ||])
+   where
+-    wrap = Scripts.mkUntypedValidator @(CurrencySymbol, TokenName) @()
++    wrap = Scripts.mkUntypedValidator @_ @(CurrencySymbol, TokenName) @()
+
+ {-# INLINABLE rolePayoutValidator #-}
+@@ -147,7 +161,7 @@ mkRolePayoutValidator :: (CurrencySymbol, TokenName)  -- ^ The datum is the curr
+                       -> Bool                         -- ^ Whether the transaction validated.
+ mkRolePayoutValidator (currency, role) _ ctx =
+     -- The role token for the correct currency must be present.
+-    -- [Marlowe-Cardano Specification: "16. Payment authorized".]
++    -- [Marlowe-Cardano Specification: "17. Payment authorized".]
+     Val.singleton currency role 1 `Val.leq` valueSpent (scriptContextTxInfo ctx)
+
+@@ -188,12 +202,10 @@ mkMarloweValidator
+             case closeInterval $ txInfoValidRange scriptContextTxInfo of
+                 Just interval' -> interval'
+                 Nothing        -> traceError "a"
+-    -- The incoming balance of each account must be positive.
+-    -- [Marlowe-Cardano Specification: "Constraint 13. Positive balances".]
+-    let positiveBalances = traceIfFalse "b" $ validateBalances marloweState
+
+     -- Find Contract continuation in TxInfo datums by hash or fail with error.
+     let inputs = fmap marloweTxInputToInput marloweTxInputs
++
+     {-  We do not check that a transaction contains exact input payments.
+         We only require an evidence from a party, e.g. a signature for PubKey party,
+         or a spend of a 'party role' token.  This gives huge flexibility by allowing
+@@ -201,22 +213,19 @@ mkMarloweValidator
+         Then, we check scriptOutput to be correct.
+      -}
+     let inputContents = fmap getInputContent inputs
++
+     -- Check that the required signatures and role tokens are present.
+     -- [Marlowe-Cardano Specification: "Constraint 14. Inputs authorized".]
+-    let inputsOk = validateInputs inputContents
+-
+-    -- Since individual balances were validated to be positive,
+-    -- the total balance is also positive.
+-    let inputBalance = totalBalance (accounts marloweState)
++    let inputsOk = allInputsAreAuthorized inputContents
+
+     -- [Marlowe-Cardano Specification: "Constraint 5. Input value from script".]
+-    -- The total incoming balance must match the actual script value being spent.
+-    let balancesOk = traceIfFalse "v" $ inputBalance == scriptInValue
+-
+-    let preconditionsOk = positiveBalances && balancesOk
++    -- [Marlowe-Cardano Specification: "Constraint 13. Positive balances".]
++    -- [Marlowe-Cardano Specification: "Constraint 19. No duplicates".]
++    -- Check that the initial state obeys the Semantic's invariants.
++    let preconditionsOk = checkState "i" scriptInValue marloweState
+
+-    -- Package the inputs to be applied in the semantics.
+     -- [Marlowe-Cardano Specification: "Constraint 0. Input to semantics".]
++    -- Package the inputs to be applied in the semantics.
+     let txInput = TransactionInput {
+             txInterval = interval,
+             txInputs = inputs }
+@@ -246,16 +255,25 @@ mkMarloweValidator
+
+                 checkContinuation = case txOutContract of
+                     -- [Marlowe-Cardano Specification: "Constraint 4. No output to script on close".]
+-                    Close -> traceIfFalse "c" checkScriptOutputAny
++                    Close -> traceIfFalse "c" hasNoOutputToOwnScript
+                     _ -> let
+                         totalIncome = foldMap collectDeposits inputContents
+                         totalPayouts = foldMap snd payoutsByParty
+-                        finalBalance = inputBalance + totalIncome - totalPayouts
+-                        -- The total account balance must be paid to a single output to the script.
+-                        -- [Marlowe-Cardano Specification: "Constraint 3. Single Marlowe output".]
+-                        -- [Marlowe-Cardano Specification: "Constraint 6. Output value to script."]
+-                        in checkOwnOutputConstraint marloweData finalBalance
++                        finalBalance = scriptInValue + totalIncome - totalPayouts
++                        in
++                             -- [Marlowe-Cardano Specification: "Constraint 3. Single Marlowe output".]
++                             -- [Marlowe-Cardano Specification: "Constraint 6. Output value to script."]
++                             -- Check that the single Marlowe output has the correct datum and value.
++                             checkOwnOutputConstraint marloweData finalBalance
++                             -- [Marlowe-Cardano Specification: "Constraint 18. Final balance."]
++                             -- [Marlowe-Cardano Specification: "Constraint 13. Positive balances".]
++                             -- [Marlowe-Cardano Specification: "Constraint 19. No duplicates".]
++                             -- Check that the final state obeys the Semantic's invariants.
++                          && checkState "o" finalBalance txOutState
+             preconditionsOk && inputsOk && payoutsOk && checkContinuation
++              -- [Marlowe-Cardano Specification: "20. Single satsifaction".]
++              -- Either there must be no payouts, or there must be no other validators.
++              && traceIfFalse "z" (null payoutsByParty || noOthers)
+         Error TEAmbiguousTimeIntervalError -> traceError "i"
+         Error TEApplyNoMatchError -> traceError "n"
+         Error (TEIntervalError (InvalidInterval _)) -> traceError "j"
+@@ -274,24 +292,73 @@ mkMarloweValidator
+         find (\TxInInfo{txInInfoOutRef} -> txInInfoOutRef == txOutRef) txInfoInputs
+     findOwnInput _ = Nothing
+
+-    -- The inputs being spent by this script.
+     -- [Marlowe-Cardano Specification: "2. Single Marlowe script input".]
++    -- The inputs being spent by this script, and whether other validators are present.
+     ownInput :: TxInInfo
+-    ownInput@TxInInfo{txInInfoResolved=TxOut{txOutAddress=ownAddress}} =
++    noOthers :: Bool
++    (ownInput@TxInInfo{txInInfoResolved=TxOut{txOutAddress=ownAddress}}, noOthers) =
+         case findOwnInput ctx of
+-            Just ownTxInInfo ->
+-                case filter (sameValidatorHash ownTxInInfo) (txInfoInputs scriptContextTxInfo) of
+-                    [i] -> i
+-                    _   -> traceError "w" -- Multiple Marlowe contract inputs with the same address are forbidden.
++            Just ownTxInInfo -> examineScripts (sameValidatorHash ownTxInInfo) Nothing True (txInfoInputs scriptContextTxInfo)
+             _ -> traceError "x" -- Input to be validated was not found.
+
++    -- Check for the presence of multiple Marlowe validators or other Plutus validators.
++    examineScripts
++      :: (ValidatorHash -> Bool)  -- Test for this validator.
++      -> Maybe TxInInfo           -- The input for this validator, if found so far.
++      -> Bool                     -- Whether no other validator has been found so far.
++      -> [TxInInfo]               -- The inputs remaining to be examined.
++      -> (TxInInfo, Bool)         -- The input for this validator and whehter no other validators are present.
++    -- This validator has not been found.
++    examineScripts _ Nothing _ [] = traceError "x"
++    -- This validator has been found, and other validators may have been found.
++    examineScripts _ (Just self) noOthers [] = (self, noOthers)
++    -- Found both this validator and another script, so we short-cut.
++    examineScripts _ (Just self) False _ = (self, False)
++     -- Found one script.
++    examineScripts f mSelf noOthers (tx@TxInInfo{txInInfoResolved=TxOut{txOutAddress=Ledger.Address (ScriptCredential vh) _}} : txs)
++      -- The script is this validator.
++      | f vh = case mSelf of
++                 -- We hadn't found it before, so we save it in `mSelf`.
++                 Nothing -> examineScripts f (Just tx) noOthers txs
++                 -- We already had found this validator before
++                 Just _ -> traceError "w"
++      -- The script is something else, so we set `noOther` to `False`.
++      | otherwise = examineScripts f mSelf False txs
++    -- An input without a validator is encountered.
++    examineScripts f self others (_ : txs) = examineScripts f self others txs
++
+     -- Check if inputs are being spent from the same script.
+-    sameValidatorHash:: TxInInfo -> TxInInfo -> Bool
+-    sameValidatorHash
+-        TxInInfo{txInInfoResolved=TxOut{txOutAddress=Ledger.Address (ScriptCredential vh1) _}}
+-        TxInInfo{txInInfoResolved=TxOut{txOutAddress=Ledger.Address (ScriptCredential vh2) _}} = vh1 == vh2
++    sameValidatorHash:: TxInInfo -> ValidatorHash -> Bool
++    sameValidatorHash TxInInfo{txInInfoResolved=TxOut{txOutAddress=Ledger.Address (ScriptCredential vh1) _}} vh2 = vh1 == vh2
+     sameValidatorHash _ _ = False
+
++    -- Check a state for the correct value, positive accounts, and no duplicates.
++    checkState :: BuiltinString -> Val.Value -> State -> Bool
++    checkState tag expected State{..} =
++      let
++        positiveBalance :: (a, Integer) -> Bool
++        positiveBalance (_, balance) = balance > 0
++        noDuplicates :: Eq k => AssocMap.Map k v -> Bool
++        noDuplicates am =
++          let
++            test [] = True           -- An empty list has no duplicates.
++            test (x : xs)            -- Look for a duplicate of the head in the tail.
++              | elem x xs = False    -- A duplicate is present.
++              | otherwise = test xs  -- Continue searching for a duplicate.
++          in
++            test $ AssocMap.keys am
++      in
++           -- [Marlowe-Cardano Specification: "Constraint 5. Input value from script".]
++           -- and/or
++           -- [Marlowe-Cardano Specification: "Constraint 18. Final balance."]
++           traceIfFalse ("v"  <> tag) (totalBalance accounts == expected)
++           -- [Marlowe-Cardano Specification: "Constraint 13. Positive balances".]
++        && traceIfFalse ("b"  <> tag) (all positiveBalance $ AssocMap.toList accounts)
++           -- [Marlowe-Cardano Specification: "Constraint 19. No duplicates".]
++        && traceIfFalse ("ea" <> tag) (noDuplicates accounts)
++        && traceIfFalse ("ec" <> tag) (noDuplicates choices)
++        && traceIfFalse ("eb" <> tag) (noDuplicates boundValues)
++
+     -- Look up the Datum hash for specific data.
+     findDatumHash' :: PlutusTx.ToData o => o -> Maybe DatumHash
+     findDatumHash' datum = findDatumHash (Datum $ PlutusTx.toBuiltinData datum) scriptContextTxInfo
+@@ -301,7 +368,7 @@ mkMarloweValidator
+     checkOwnOutputConstraint ocDatum ocValue =
+         let hsh = findDatumHash' ocDatum
+         in traceIfFalse "d" -- "Output constraint"
+-        $ checkScriptOutput ownAddress hsh ocValue getContinuingOutput
++        $ checkScriptOutput (==) ownAddress hsh ocValue getContinuingOutput
+
+     getContinuingOutput :: TxOut
+     getContinuingOutput = case filter (\TxOut{txOutAddress} -> ownAddress == txOutAddress) allOutputs of
+@@ -309,20 +376,14 @@ mkMarloweValidator
+         _     -> traceError "o" -- No continuation or multiple Marlowe contract outputs is forbidden.
+
+     -- Check that address, value, and datum match the specified.
+-    checkScriptOutput :: Ledger.Address -> Maybe DatumHash -> Val.Value -> TxOut -> Bool
+-    checkScriptOutput addr hsh value TxOut{txOutAddress, txOutValue, txOutDatum=OutputDatumHash svh} =
+-                    txOutValue == value && hsh == Just svh && txOutAddress == addr
+-    checkScriptOutput _ _ _ _ = False
+-
+-    -- Check that address and datum match the specified, and that value is at least that required.
+-    checkScriptOutputRelaxed :: Ledger.Address -> Maybe DatumHash -> Val.Value -> TxOut -> Bool
+-    checkScriptOutputRelaxed addr hsh value TxOut{txOutAddress, txOutValue, txOutDatum=OutputDatumHash svh} =
+-                    txOutValue `Val.geq` value && hsh == Just svh && txOutAddress == addr
+-    checkScriptOutputRelaxed _ _ _ _ = False
++    checkScriptOutput :: (Val.Value -> Val.Value -> Bool) -> Ledger.Address -> Maybe DatumHash -> Val.Value -> TxOut -> Bool
++    checkScriptOutput comparison addr hsh value TxOut{txOutAddress, txOutValue, txOutDatum=OutputDatumHash svh} =
++                    txOutValue `comparison` value && hsh == Just svh && txOutAddress == addr
++    checkScriptOutput _ _ _ _ _ = False
+
+     -- Check for any output to the script address.
+-    checkScriptOutputAny :: Bool
+-    checkScriptOutputAny = all ((/= ownAddress) . txOutAddress) allOutputs
++    hasNoOutputToOwnScript :: Bool
++    hasNoOutputToOwnScript = all ((/= ownAddress) . txOutAddress) allOutputs
+
+     -- All of the script outputs.
+     allOutputs :: [TxOut]
+@@ -339,8 +400,8 @@ mkMarloweValidator
+     marloweTxInputToInput (Input input) = NormalInput input
+
+     -- Check that inputs are authorized.
+-    validateInputs :: [InputContent] -> Bool
+-    validateInputs = all validateInputWitness
++    allInputsAreAuthorized :: [InputContent] -> Bool
++    allInputsAreAuthorized = all validateInputWitness
+       where
+         validateInputWitness :: InputContent -> Bool
+         validateInputWitness input =
+@@ -356,8 +417,10 @@ mkMarloweValidator
+
+     -- Tally the deposits in the input.
+     collectDeposits :: InputContent -> Val.Value
+-    collectDeposits (IDeposit _ _ (Token cur tok) amount) = Val.singleton cur tok amount
+-    collectDeposits _                                     = zero
++    collectDeposits (IDeposit _ _ (Token cur tok) amount)
++      | amount > 0    = Val.singleton cur tok amount  -- SCP-5123: Semantically negative deposits
++      | otherwise     = zero                          -- do not remove funds from the script's UTxO.
++    collectDeposits _ = zero
+
+     -- Extract the payout to a party.
+     payoutByParty :: Payment -> AssocMap.Map Party Val.Value
+@@ -374,12 +437,17 @@ mkMarloweValidator
+       where
+         payoutToTxOut :: (Party, Val.Value) -> Bool
+         payoutToTxOut (party, value) = case party of
++            -- [Marlowe-Cardano Specification: "Constraint 15. Sufficient Payment".]
++            -- SCP-5128: Note that the payment to an address may be split into several outputs but the payment to a role must be
++            -- a single output. The flexibily of multiple outputs accommodates wallet-related practicalities such as the change and
++            -- the return of the role token being in separate UTxOs in situations where a contract is also paying to the address
++            -- where that change and that role token are sent.
+             Address _ address  -> traceIfFalse "p" $ value `Val.leq` valuePaidToAddress address  -- At least sufficient value paid.
+             Role role -> let
+                 hsh = findDatumHash' (rolesCurrency, role)
+                 addr = Address.scriptHashAddress rolePayoutValidatorHash
+                 -- Some output must have the correct value and datum to the role-payout address.
+-                in traceIfFalse "r" $ any (checkScriptOutputRelaxed addr hsh value) allOutputs
++                in traceIfFalse "r" $ any (checkScriptOutput Val.geq addr hsh value) allOutputs
+
+     -- The key for the address must have signed.
+     txSignedByAddress :: Ledger.Address -> Bool
+@@ -402,7 +470,7 @@ marloweValidator :: Scripts.TypedValidator TypedMarloweValidator
+ marloweValidator =
+   let
+     mkUntypedMarloweValidator :: ValidatorHash -> BuiltinData -> BuiltinData -> BuiltinData -> ()
+-    mkUntypedMarloweValidator rp = mkUntypedValidator (mkMarloweValidator rp)
++    mkUntypedMarloweValidator rp = Scripts.mkUntypedValidator (mkMarloweValidator rp)
+
+     untypedValidator :: Scripts.Validator
+     untypedValidator = mkValidatorScript $
+@@ -410,10 +478,20 @@ marloweValidator =
+         `PlutusTx.applyCode` PlutusTx.liftCode rolePayoutValidatorHash
+
+     typedValidator :: Scripts.TypedValidator Scripts.Any
+-    typedValidator = Scripts.unsafeMkTypedValidator untypedValidator
++    typedValidator = unsafeMkTypedValidator (Scripts.Versioned untypedValidator Scripts.PlutusV2)
+   in
+     unsafeCoerce typedValidator
+
++
++marloweValidatorCompiled :: PlutusTx.CompiledCode (ValidatorHash -> PlutusTx.BuiltinData -> PlutusTx.BuiltinData -> PlutusTx.BuiltinData -> ())
++marloweValidatorCompiled =
++  let
++    mkUntypedMarloweValidator :: ValidatorHash -> PlutusTx.BuiltinData -> PlutusTx.BuiltinData -> PlutusTx.BuiltinData -> ()
++    mkUntypedMarloweValidator rp = Scripts.mkUntypedValidator (mkMarloweValidator rp)
++  in
++    $$(PlutusTx.compile [|| mkUntypedMarloweValidator ||])
++
++
+ -- | The hash of the Marlowe semantics validator.
+ marloweValidatorHash :: ValidatorHash
+ marloweValidatorHash = Scripts.validatorHash marloweValidator
+@@ -432,7 +510,7 @@ alternateMarloweValidator = Scripts.mkTypedValidator
+           $$(PlutusTx.compile [|| mkMarloweValidator ||])
+             `PlutusTx.applyCode`
+               PlutusTx.liftCode rolePayoutValidatorHash
+-        mkArgsValidator = mkUntypedValidator @MarloweData @MarloweInput
++        mkArgsValidator = Scripts.mkUntypedValidator @_ @MarloweData @MarloweInput
+         compiledArgsValidator =
+           $$(PlutusTx.compile [|| mkArgsValidator ||])
+```

### 2023-03-29 response-to-audit-report.md
PLT-3382 Added TODOs.

@@
-# DRAFT Response to Audit Report
-
-TODO:
+# Response to Audit Report
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO:
 
@@
 
-TODO: @bwbush will add text here after everything else is written.
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @bwbush will add text here after everything else is written.
 
@@
 
-TODO: @bwbush will add text thanking the audit team.
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @bwbush will add text thanking the audit team.
 
 ## Overview of Isabelle Changes
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.1.5 Insufficient documentation of Money preservation *(Severity: Medium)*
@@
 It is necessary to explain the meaning of these functions in sufficient detail so readers can understand the property.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 > 3.  Merkleizing and unmerkleizing a contract gives back the original contract.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 It would be more reliable to have explicit and weaker lemmas that assert equalities only when there are no errors.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.2 Undefined reference *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.3 Lack of explanation for necessity of `Environment` type *(Severity: Low)*
@@
 > An `Environment` type is introduced, but it is unclear why it is needed as it is defined as a synonym for time intervals.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.5 Unexplained interval data types *(Severity: Low)*
@@
 > The meaning of the data types `IntervalError` and `IntervalResult` needs to be explained.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.7 Code snippets switch languages *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.8 Repeated definition of `IntervalResult` *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.9 Poorly named variable `newAccount` *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.10 Poorly named variable `acc` in specification *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.11 Inaccurate specification of `giveMoney` *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.12 Redundant evaluation in `addMoneyToAccount` *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.13 Redundant statement regarding addition *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.14 Missing implementation for negation case of `evalValue` *(Severity: Low)*
@@
 > Negation for `evalValue` does not show the implementation, just one lemma about `NegValue`, which is inconsistent with how other operations are presented.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.16 Unclear division explanation *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.17 Discrepancy with `evalValue` *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.18 Missing `evalValue` lemmas in specification *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.19 Typo in **Use Value** case of `evalValue` *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.20 Unexplained parameters of `playTrace` *(Severity: Low)*
@@
 > The parameters of the function `playTrace` need to be explained.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 Even though the latter is an alias for the former, it is beneficial to use the `POSIXTime` name both for consistency and readability.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.23 Complicated definition of `allAccountsPositive` *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.24 Discrepancy with Isabelle code for `allAccountsPositive` *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.25 Misleading or incorrect formula for contract not holding funds *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.26 Different format for lemma statement *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.27 Function `isClosedAndEmpty` is unexplained *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.2.28 Top-down definitions *(Severity: Low)*
@@
 > In Section 2, the order of definitions is reversed, and the reader is thus faced with functions which call other functions that have not been introduced yet, despite the claim in Section 1.3 that the definitions will be presented bottom-up.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 This way, it would be much easier to search for the actual Isabelle code and proofs matching the informal specification text, and compare the two.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 > -   in `SingleInputTransactions.thy`, lemmas `applyAllInputsPrefix_aux`,\ `computeTransactionIterative`, and `computeTransactionStepEquivalence_error`
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.3.3 Confusing auxiliary lemmas *(Severity: Low)*
@@
 > -   in `SingleInputTransactions.thy`, lemma `applyAllInputsPrefix_aux`
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 > -   in `TimeRange.thy`, lemmas `resultOfReduceIsCompatibleToo`, `resultOfReductionLoopIsCompatibleToo`, `resultOfReduceUntilQuiescentIsCompatibleToo`, `reduceLoop_ifCaseLtCt`, and\ `reduceContractUntilQuiescent_ifCaseLtCt`
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 > The proofs of the lemmas can be found in […].
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.3.7 Inconsistent variable name `valTrans` *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.3.8 Unused binding `interAccs` *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.3.9 Undescriptive variable name `acc` *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.3.10 Misleading indentation *(Severity: Low)*
@@
 > The indentation is misleading: the premises on these lines are indented as if they are a part of the previous functional premise.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 This trivially follows from `playTraceAux_preserves_validAndPositive_state` but no such theorem is present.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.3.13 Misleading lemma names *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.3.14 Misleading variable name `reduced` *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.3.15 Undescriptive name `beforeApplyAllLoopIsUseless` *(Severity: Low)*
@@
 > A more descriptive name for this lemma could be `reduceContractUntilQuiescent_hasNoEffect_before_applyAllLoop`
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.3.17 Redundant `reduceContractStep` lemmas *(Severity: Low)*
@@
 > This lemma is weaker than `transferMoneyBetweenAccounts_preserves`. If we replace its usage at line 351 with `transferMoneyBetweenAccounts_preserves`, the proof goes through.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.3.19 Duplicated lemmas *(Severity: Low)*
@@
 > This lemma is defined twice, once in each of these files. One of them should be removed.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 Thus, no auxiliary lemma is needed.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 > ```
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.3.23 Inconsistent style when applying constructor *(Severity: Low)*
@@
 > The lines mentioned in these lemmas display the resulting constructor before the function application, which differs from the general style in the rest of the codebase.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.3.25 Typo with "independet" in multiple lemmas *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.3.26 Poorly named `acc` lemmas *(Severity: Low)*
@@
 > It is unclear what `acc` refers to in these lemma names, as the lemmas are about the independence of warnings and payments, not accounts.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.3.28 `playTrace_only_accepts_maxTransactionsInitialState` not written as `theorem` *(Severity: Low)*
@@
 > This lemma seems like the main result of this file. Assuming it is an important result, we recommend writing it as a `theorem` rather than a `lemma`.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.3.30 Function `validTimeInterval` unnecessarily unfolded in lemma *(Severity: Low)*
@@
 > For consistency, $a \leq b$ should be replaced by $\texttt{validTimeInterval}$.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.3.32 `playTrace_preserves_valid_state` not written as `theorem` *(Severity: Low)*
@@
 > This lemma seems like the main result of this file. Assuming it is an important result, we recommend writing it as a `theorem` instead.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.4.1 Variable shadowing in `applyAllLoop` *(Severity: Medium)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.4.2 Undescriptive name `moneyInPayment` *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.4.3 Typo in section name *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.4.4 Typo in comment *(Severity: Low)*
@@
 > Typo in comment: "endpoits".
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.4.6 Variable name discrepancy in `reductionLoop` *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.4.7 Typo in constructor *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.4.8 Unclear function name `calculateNonAmbiguousInterval` *(Severity: Low)*
@@
 > The meaning of the function is not obvious. It needs a comment to explain it.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.4.10 Misleading function names *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.4.11 Unused parameter in `maxTransactionCaseList` *(Severity: Low)*
@@
 
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+
 ### 2.4.12 Duplicated `isValidInterval` function *(Severity: Low)*
@@
 > This function duplicates $\texttt{validTimeInterval}$ from $\texttt{OptBoundTimeInterval.thy}$, and the latter has certain additional properties proven about it specifically, so it makes sense to use the latter in both cases.
+
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
-TODO: @hrajchert will add text here.
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here.
 
@@
 
-TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
 
@@
 
-TODO: @bwbush will add text here.
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @bwbush will add text here.
 

### 2023-03-29 response-to-audit-report.md
PLT-3382 Added reponse to test-failure precision.

@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO:
+**TODO:**
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @bwbush will add text here.
+A future version of the Marlowe test suite for `checkSemanticsTransaction` and `checkPayoutTransaction` will instrumenting Plutus scripts for the Marlowe validators so that the precise cause of the test failure can be verified.
 

### 2023-03-30 isabelle/generated/Arith.hs
PLT-3313: Exporting `integer_of_int` from Isabelle

@@ -1,9 +1,10 @@
 {-# LANGUAGE EmptyDataDecls, RankNTypes, ScopedTypeVariables #-}
 
 module
-  Arith(Int(..), equal_int, less_eq_int, less_int, Nat, integer_of_nat, Num(..),
-         One, Zero, Zero_neq_one, suc, uminus_int, zero_int, abs_int, one_int,
-         plus_int, zero_nat, minus_int, times_int, of_bool, divide_int)
+  Arith(Int(..), integer_of_int, equal_int, less_eq_int, less_int, Nat,
+         integer_of_nat, Num(..), One, Zero, Zero_neq_one, suc, uminus_int,
+         zero_int, abs_int, one_int, plus_int, zero_nat, minus_int, times_int,
+         of_bool, divide_int)
   where {
 
 import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),

### 2023-03-30 marlowe-spec-test/src/Marlowe/Spec/Core/Semantics.hs

@@ -200,7 +200,7 @@ playTrace_only_accepts_maxTransactionsInitialStateTest interpret = reproducibleP
     transactions <- run $ generate $ arbitraryValidInputs state contract
     let
         req :: Request JSON.Value
-        req = PlayTrace (integer_of_int $ minTime state) contract transactions
+        req = PlayTrace (Arith.integer_of_int $ minTime state) contract transactions
     RequestResponse res <- run $ liftIO $ interpret req
 
     case JSON.fromJSON res of
@@ -226,16 +226,13 @@ traceToSingleInputIsEquivalentTest interpret = reproducibleProperty "Single inpu
         pure (c,s,t)) `suchThat` \(_,_,t) -> t /= traceListToSingleInput t
 
     let
-        multipleInputs = PlayTrace (integer_of_int startTime) contract transactions
-        singletonInput = PlayTrace (integer_of_int startTime) contract (traceListToSingleInput transactions)
+        multipleInputs = PlayTrace (Arith.integer_of_int startTime) contract transactions
+        singletonInput = PlayTrace (Arith.integer_of_int startTime) contract (traceListToSingleInput transactions)
 
     RequestResponse resMultipleInputs <- run $ liftIO $ interpret multipleInputs
     RequestResponse resSingletonInput <- run $ liftIO $ interpret singletonInput
 
     assert $ resMultipleInputs == resSingletonInput
-
-integer_of_int :: Arith.Int -> Integer
-integer_of_int (Arith.Int_of_integer k) = k
 
 -- lemma reduceContractUntilQuiescentIdempotent :
 --    "reduceContractUntilQuiescent env state contract = ContractQuiescent reducedAfter wa pa nsta ncont ⟹
@@ -301,7 +298,7 @@ playTraceIsQuiescentTest interpret = reproducibleProperty "Calling playTrace is 
     transactions <- run $ generate $ arbitraryValidInputs state contract `suchThat` (not . null)
     let
         req :: Request JSON.Value
-        req = PlayTrace (integer_of_int $ minTime state) contract transactions
+        req = PlayTrace (Arith.integer_of_int $ minTime state) contract transactions
     RequestResponse res <- run $ liftIO $ interpret req
 
     case JSON.fromJSON res of

### 2023-03-30 marlowe-spec-test/src/Marlowe/Spec/Core/Semantics.hs
PLT-3313: review feedback

@@ -208,9 +208,9 @@ playTrace_only_accepts_maxTransactionsInitialStateTest interpret = reproducibleP
         monitor
           ( counterexample $
               "Request: " ++ showAsJson req ++ "\n"
-                ++ "Expected reponse to be quiescent" )
+                ++ "Expected maxTransactionsInitialState (" ++ show (maxTransactionsInitialState contract) ++ ")\n"
+                ++ "to be an upper bound for the number of transacations (" ++ show (length transactions) ++ ")")
         assert $ toInteger (length transactions) <= Arith.integer_of_nat (maxTransactionsInitialState contract)
-      JSON.Success _ -> pre False
       _ -> fail "JSON parsing failed!"
 
 -- SingleInputTransactions.thy

### 2023-04-04 response-to-audit-report.md
PLT-3382 Fixed typos and remedied omission.

--- 2023-03-29 response-to-audit-report.md
@@
 
-Incorrect handling of negative deposits (item 1 above) was mitigated by the addition of a guard against this in Marlowe's Plutus validator so that its behavior was consistent with Marlowe semantics.
+Incorrect handling of negative deposits (item 1 above) was mitigated by the addition of a guard against this behavior in Marlowe's Plutus validator so that its behavior is consistent with Marlowe semantics.
 Property-based tests were also added to check the correctness of the mitigation.
 
-Although the Marlowe validator had prevented double-satisfaction among multiple copies of the Marlowe validator script running in the same transaction, it did not prevent double-satisfaction in cases where the Marlowe validator run alongside another Plutus validator in the same transaction (item 2 above).
-Double satisfaction has now been prevented by enforcing that the Marlowe validator is the only Plutus script that runs in transactions that make payments to parties.
+Although the Marlowe validator had prevented double-satisfaction among multiple copies of the Marlowe validator script running in the same transaction, it did not prevent double-satisfaction in cases where the Marlowe validator ran alongside another Plutus validator in the same transaction (item 2 above).
+Double satisfaction is now prevented by enforcing that the Marlowe validator is the only Plutus script that runs during transactions that make payments to parties.
 This allows Marlowe contracts to coordinate with other Plutus scripts, but only under conditions where double satisfaction is impossible.
@@
 This hardening was completed without unduly increasing the size (in bytes) of the validator or its Plutus execution cost: such was verified with simulated and on-chain measurement of execution costs for a library of real-life Marlowe contracts.
-Notably, the enforcement of the final invariants protects against funds becoming permanently locked in a Marlowe contract due to a garbled state: even in cases where the implementation of Marlowe semantics or parts of the Plutus validator are flawed, at least some of the execution paths of a Marlowe contract remain viable, so funds could be released even under nominally impossible but catastrophic difficulties.
-The property-based test suite was significantly enhanced to
+Notably, the enforcement of the final invariants protects against funds becoming permanently locked in a Marlowe contract due to a garbled state: even in cases where the implementation of Marlowe semantics or parts of the Plutus validator were flawed, at least some of the execution paths of a Marlowe contract remain viable, so funds could be released even under nominally impossible but catastrophic difficulties.
+The property-based test suite was significantly enhanced to check double-satisfaction situations.
 
@@
 The Marlowe semantics validator had prevented a limited form of double-satisfaction attack by preventing two Marlowe validators to execute in the same transaction.
-The audit report correctly highlights that such an attack could be feasible in cases where another Plutus validator (not the same version of the Marlowe semantics validator) executes in a transactions where a payment to an address or to the Marlowe role-payout validator satisfies both the Marlowe semantics validator and the other validator.
+The audit report correctly highlights that such an attack could be feasible in cases where another Plutus validator (not the selfsame version of the Marlowe semantics validator) executes in a transactions where a payment to an address or to the Marlowe role-payout validator satisfies both the Marlowe semantics validator and the other validator.
 
@@
 However, as the audit report points out, such a flaw would result in the contract being forever blocked from further execution.
-Adding a check on the final output's validity would prevent a contract from ever reaching a fully blocked state.
+Adding a check on the final output's validity now prevents a contract from ever reaching a fully blocked state.
 Even if the semantics or Plutus code were flawed, at least some execution paths in the contract would remain viable, so that the contract could eventually terminate and would not permanently lock funds.
@@
 
-Prompted on this finding in the audit report, the Marlowe semantics validator has been altered to enforce all invariants (positive accounts; non-duplication of accounts, choices, and bound values; a total internal account balance that exactly matches the value in the script's UTxO) on both the initial and final states.
+Prompted by this finding in the audit report, the Marlowe semantics validator has been altered to enforce all invariants (positive accounts; non-duplication of accounts, choices, and bound values; a total internal account balance that exactly matches the value in the script's UTxO) on both the initial and final states.
 This prevents execution of a Marlowe contract that was inadvertently or purposefully created by off-chain code with an invalid initial state.
-It also prevents transactions that corrupt the final state; such corruption could only happen if there were a flaw in the Plutus validator or Marlowe semantics implementation, in which case that execution-path of the contract would be blocked.
+It also prevents transactions that corrupt the final state; such corruption could only happen if there were a flaw in the Plutus validator or Marlowe semantics implementation, in which case that particular execution-path of the contract would be blocked.
 The [Marlowe Best Practices Guide](https://github.com/input-output-hk/marlowe-cardano/blob/main/marlowe/best-practices.md) and [Marlowe Security Guide](https://github.com/input-output-hk/marlowe-cardano/blob/main/marlowe/security.md) warn of the importance of off-chain code creating valid initial state for Marlowe contracts.
@@
 
-> **File `Semantics.hs`, Several functions ``** 
+> **File `Semantics.hs`, Several functions** 
 > 
 > Where `MList.insert` is used in the Isabelle semantics, [AssocMap.insert](https://github.com/input-output-hk/plutus/blob/v1.0.0/plutus-tx/src/PlutusTx/AssocMap.hs#L147-L148) is used in the Cardano implementation.
-However, the functions are not equivalent as demonstrated by the following examples:
-> 
-> ``` text
-> AssocMap.insert 'a' 1 [('b', 0)] == [('b', 0), ('a', 1)]
-> -- whereas
-> MList.insert 'a' 1 [('b', 0)] == [('a', 1), ('b', 0)]
-> ```
-> 
-> ``` text
-> AssocMap.insert 'b' 1 [('a', 0), (‘b’, 0), (‘b’, 0)] == [('a', 0), ('b', 1), (‘b’, 1)]
-> -- whereas
-> MList.insert 'b' 1 [(‘a’, 0), ('b', 0), ('b', 0)] == [('a', 0), ('b', 1), ('b', 0)]
-> ```
+However, the functions are not equivalent as demonstrated by the following examples: […]
 > 
@@
 > 
-> Similarly, functions `AssocMap.delete` and `MList.delete` differ in behavior when the input map is not sorted:
-> 
-> ``` text
-> AssocMap.delete 'a' [(‘b’, 0), (‘a’, 0)] == [('b', 0)]
-> -- whereas
-> MList.delete 'a' [(‘b’, 0), ('a', 0)] == [('b', 0), ('a', 0)]
-> ```
-> 
-> Functions `AssocMap.lookup` and `MList.lookup` also differ in behavior when the input map is not sorted:
-> 
-> ``` text
-> AssocMap.lookup 'a' [(‘b’, 0), (‘a’, 0)] == Just 0
-> -- whereas
-> MList.lookup 'a' [(‘b’, 0), ('a', 0)] == Nothing
-> ```
+> Similarly, functions `AssocMap.delete` and `MList.delete` differ in behavior when the input map is not sorted: […]
+> 
+> Functions `AssocMap.lookup` and `MList.lookup` also differ in behavior when the input map is not sorted: […]
 > 
@@
 Work is underway to modify the Isabelle semantics and proofs to use a data structure similar to Plutus's `AssocMap`.
-Instead of modifying the Marlowe semantics validator, all usages of `AssocMap` were manually reviewed to verify that no vulnerability was introduced by its use (as compared to `MList`) *provided that the precondition of no duplicate entries in the `AssocMap` held*: the source code was annotated with informal reasoning documenting the safety of `AssocMap`'s use.
+Instead of modifying the Marlowe semantics validator to address this finding, all usages of `AssocMap` were manually reviewed to verify that no vulnerability was introduced by its use (as compared to `MList`) *provided that the precondition of no duplicate entries in the `AssocMap` held*: the source code was annotated with informal reasoning documenting the safety of `AssocMap`'s use.
 That manual review was augmented by a comprehensive set of property-based tests to check the behavior of `MList` (under the precondition that it was ordered and contained no duplicates) against `AssocMap` (under the precondition that it contained not duplicates) for all functions used in Marlowe semantics or in the Marlowe validator.
@@
 
-> **File `Spec/Marlowe/Semantics/Compute.hs`, ``** 
+> **File `Spec/Marlowe/Semantics/Compute.hs`** 
 > 
@@
 The test oracle is now applied to the Marlowe semantics implementation in Plutus that the validator use.
-The test oracle uses the refference implementation to generates test cases to challenge the Plutus implementation and then it checks the Plutus results against its own.
-This oracle provides sufficient coverage to address the audit-reports concerns regarding the lack of property-based tests for Section 3 of the Marlowe specification.
+The test oracle uses the reference implementation to generate test cases to challenge the Plutus implementation and then it checks the Plutus result against its own result.
+This oracle provides sufficient coverage to address this audit-reports concern regarding the lack of property-based tests for Section 3 of the Marlowe specification.
 
@@
 It is mentioned in a commented out line in `StaticAnalysis.thy`.
-Furthermore, the lemma can be expressed more concisely as $$\llparenthesis \mathit{interval} = \mathit{inte}, \mathit{inputs} = \mathit{inp\_h}\ \#\ \mathit{inp\_t} \rrparenthesis\ \#\ t = \mathit{traceListToSingleInput}\ \mathit{t2} % \mathrel{% \mbox{\fontfamily{cmr}\fontencoding{OT1}\selectfont=}}% \joinrel\Rightarrow\mathit{inp\_t} = []$$
+Furthermore, the lemma can be expressed more concisely as `$$\llparenthesis \mathit{interval} = \mathit{inte}, \mathit{inputs} = \mathit{inp\_h}\ \#\ \mathit{inp\_t} \rrparenthesis\ \#\ t = \mathit{traceListToSingleInput}\ \mathit{t2} % \mathrel{% \mbox{\fontfamily{cmr}\fontencoding{OT1}\selectfont=}}% \joinrel\Rightarrow\mathit{inp\_t} = []$$`
 
@@
 > The statement of this lemma is very verbose.
-A more natural (and slightly stronger) formulation could be $$\begin{aligned} &\mathit{playTraceAux}\ \mathit{txOut}\ [\ \llparenthesis \mathit{interval} = \mathit{inte}, \mathit{inputs} = [h] \rrparenthesis, \llparenthesis \mathit{interval} = \mathit{inte}, \mathit{inputs} = t \rrparenthesis \ ] \\ =\ &\mathit{playTraceAux}\ \mathit{txOut}\ [\ \llparenthesis \mathit{interval} = \mathit{inte}, \mathit{inputs} = h\ \#\ t \rrparenthesis \ ]\end{aligned}$$
+A more natural (and slightly stronger) formulation could be `$$\begin{aligned} &\mathit{playTraceAux}\ \mathit{txOut}\ [\ \llparenthesis \mathit{interval} = \mathit{inte}, \mathit{inputs} = [h] \rrparenthesis, \llparenthesis \mathit{interval} = \mathit{inte}, \mathit{inputs} = t \rrparenthesis \ ] \\ =\ &\mathit{playTraceAux}\ \mathit{txOut}\ [\ \llparenthesis \mathit{interval} = \mathit{inte}, \mathit{inputs} = h\ \#\ t \rrparenthesis \ ]\end{aligned}$$`
 
@@
 > 
-> These lemmas use the hypothesis $\mathit{minTime}\ \mathit{sta} \leq \mathit{iniTime}$ and build a state $\mathit{sta}\ \llparenthesis \mathit{minTime} := \mathit{iniTime} \rrparenthesis)$ while other lemmas simply say $\mathit{minTime}\ \mathit{sta} = \mathit{iniTime}$.
+> These lemmas use the hypothesis `$\mathit{minTime}\ \mathit{sta} \leq \mathit{iniTime}$` and build a state `$\mathit{sta}\ \llparenthesis \mathit{minTime} := \mathit{iniTime} \rrparenthesis)$` while other lemmas simply say `$\mathit{minTime}\ \mathit{sta} = \mathit{iniTime}$`.
 Readability would be improved by presenting these lemmas in the same style as the others, or documenting the need for these distinct presentations via code comments.
@@
 > 
-> The assumptions $$\forall x\ \mathit{tok}.\ \mathit{positiveMoneyInAccountOrNoAccount}\ x\ \mathit{tok}\ \mathit{accs}$$ in `addMoneyToAccountPositive_match` and $$\mathit{money > 0}$$ in `addMoneyToAccountPositive_noMatch` are unnecessary.
+> The assumptions `$$\forall x\ \mathit{tok}.\ \mathit{positiveMoneyInAccountOrNoAccount}\ x\ \mathit{tok}\ \mathit{accs}$$` in `addMoneyToAccountPositive_match` and `$$\mathit{money > 0}$$` in `addMoneyToAccountPositive_noMatch` are unnecessary.
 > 
@@
 > 
-> The assumption $$\begin{aligned} \mathit{newState} = \llparenthesis &\mathit{accounts} = \mathit{newAccounts}, \mathit{choices} = \mathit{newChoices}, \\ &\mathit{boundValues} = \mathit{newBoundValues}, \mathit{minTime} = \mathit{newMinTime} \rrparenthesis\end{aligned}$$ is unnecessary.
+> The assumption `$$\begin{aligned} \mathit{newState} = \llparenthesis &\mathit{accounts} = \mathit{newAccounts}, \mathit{choices} = \mathit{newChoices}, \\ &\mathit{boundValues} = \mathit{newBoundValues}, \mathit{minTime} = \mathit{newMinTime} \rrparenthesis\end{aligned}$$` is unnecessary.
 
@@
 
-> **File `Semantics.hs`, Several functions ``** 
+> **File `Semantics.hs`, Several functions** 
 > 
@@
 
-> **File `Several files`, `several functions`** 
+> **File `Several files`, Several functions** 
 > 
@@
 > The comment describing the function is overly concise, as it does not mention the function removing all non-positive accounts before the first positive one, and effectively `uncons`-ing the list.
-> 
+
+Commit [47c9deb2](https://github.com/input-output-hk/marlowe-cardano/commit/47c9deb2047a240d78eaf62306278ecfbe624bea) elaborates the documentation for `refundOne`.
+
 > **File `Semantics.hs`, Function `addMoneyToAccount`, line *461***
@@
 The discrepancies of implementation between the constraints in the specification and the Plutus code result from the emphasis on readability in the specification and of efficiency in the implementation.
-The need to minimize of validator size and execution cost in the implementation has sometimes resulted in less consise or readible code.
-The Plutus compile can be quite sensitive to the precise expression of a constraint as Haskell code.
+The need to minimize of validator size and execution cost in the implementation has sometimes resulted in less concise or less readable code.
+Note that the Plutus compiler can be quite sensitive to the precise expression of a constraint as Haskell code.
 
@@
 
-Commit [00a0c240](https://github.com/input-output-hk/marlowe-cardano/commit/00a0c2406665f06753aecf6dbb63933edb275436) moves the comment about the non-existant fifth argument to a separate sentance following the list of arguments.
+Commit [00a0c240](https://github.com/input-output-hk/marlowe-cardano/commit/00a0c2406665f06753aecf6dbb63933edb275436) moves the comment about the non-existent fifth argument to a separate sentence following the list of arguments.
 
@@
 
-> **File `marlowe-cardano-specification.md`, Various sections ``** 
+> **File `marlowe-cardano-specification.md`, Various sections** 
 > 
@@
 
-This mistatement is remedied in commit [8691f335](https://github.com/input-output-hk/marlowe-cardano/commit/8691f3351e55758cc750eee0800b1458f0a8c554).
+This misstatement is remedied in commit [8691f335](https://github.com/input-output-hk/marlowe-cardano/commit/8691f3351e55758cc750eee0800b1458f0a8c554).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @bwbush will add text here.
+A future version of the Marlowe test suite for `checkSemanticsTransaction` and `checkPayoutTransaction` will instrument Plutus scripts for the Marlowe validators so that the precise cause of the test failure can be verified.
 
@@
 
-Commit [ee2e5222](https://github.com/input-output-hk/marlowe-cardano/commit/ee2e52220e25736af29a1f01144d5fafbe9eed76) adds a property-based test that choice and deposit continue as expected; commit [333eee29](https://github.com/input-output-hk/marlowe-cardano/commit/333eee293aa3afe95d0aa7f678c7e7b7bad660ea) tests that choice produces expected continuation; commit [477284c7](https://github.com/input-output-hk/marlowe-cardano/commit/477284c78095482eba63adc2e6090ea68553db9d) tests that that deposits add to an account; commit [d241669b](https://github.com/input-output-hk/marlowe-cardano/commit/d241669bfb7ec8dad1cc92a7ca6185d08229817e) tests that payout substracts from an account; and commit [f33fc720](https://github.com/input-output-hk/marlowe-cardano/commit/f33fc7206522031c93253323f3b8ca083d8ca8c7) tests that Merkleization continues as expected.
+Commit [ee2e5222](https://github.com/input-output-hk/marlowe-cardano/commit/ee2e52220e25736af29a1f01144d5fafbe9eed76) adds a property-based test that choice and deposit continue as expected; commit [333eee29](https://github.com/input-output-hk/marlowe-cardano/commit/333eee293aa3afe95d0aa7f678c7e7b7bad660ea) tests that choice produces expected continuation; commit [477284c7](https://github.com/input-output-hk/marlowe-cardano/commit/477284c78095482eba63adc2e6090ea68553db9d) tests that that deposits add to an account; commit [d241669b](https://github.com/input-output-hk/marlowe-cardano/commit/d241669bfb7ec8dad1cc92a7ca6185d08229817e) tests that payout subtracts from an account; and commit [f33fc720](https://github.com/input-output-hk/marlowe-cardano/commit/f33fc7206522031c93253323f3b8ca083d8ca8c7) tests that merkleization continues as expected.
 
@@
 
-Commit [811f048b](https://github.com/input-output-hk/marlowe-cardano/commit/811f048ba93db10972641c0c235864db1d6014e7) adds a property-based test for an invalid in terval in the past.
+Commit [811f048b](https://github.com/input-output-hk/marlowe-cardano/commit/811f048ba93db10972641c0c235864db1d6014e7) adds a property-based test for an invalid interval in the past.
 
@@
 ```bash
-git diff 523f3d56f22bf992ddb0b0c8a52bb7a5a188f9e9 69468d6623e24a4ccd6659e378ae012c38ca01ce marlowe/src/Language/Marlowe/{Core/V1/Semantics*,Scripts.hs}
+git diff 523f3d56f22bf992ddb0b0c8a52bb7a5a188f9e9 47c9deb2047a240d78eaf62306278ecfbe624bea marlowe/src/Language/Marlowe/{Core/V1/Semantics*,Scripts.hs}
 ```
@@
 diff --git a/marlowe/src/Language/Marlowe/Core/V1/Semantics.hs b/marlowe/src/Language/Marlowe/Core/V1/Semantics.hs
-index e31b8c478..82b57b903 100644
+index e31b8c478..0d7e94a20 100644
 --- a/marlowe/src/Language/Marlowe/Core/V1/Semantics.hs
@@
          ValueGT lhs rhs         -> evalVal lhs > evalVal rhs
-@@ -439,6 +520,12 @@ evalObservation env state obs = let
+@@ -435,10 +516,16 @@ evalObservation env state obs = let
+         FalseObs                -> False
+
+--- | Pick the first account with money in it.
++-- | Pick the first account with money in it, discarding any accounts prior to that if they have a non-positive balance.
  refundOne :: Accounts -> Maybe ((Party, Token, Integer), Accounts)

### 2023-04-04 response-to-audit-report.md
PLT-3382 Drafted executive summary.

@@
 
-- [ ] Add prefatory text and executive summary.
+- [x] Add prefatory text and executive summary.
 - [ ] Add responses and mitigation commits to Isabelle-related findings.
@@
 - [ ] Fix display of equations or replace ones that cannot be fixed with ellipses ("...").
-- [ ] Replace broken links and cross references, or use elipses ("...").
+- [ ] Replace broken links and cross references, or use ellipses ("...").
 - [ ] Verify the section numbering is correct.
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @bwbush will add text here after everything else is written.
+Marlowe consists of a domain-specific language (DSL) for the financial industry and for peer-to-peer financial contracts, with a reference semantics written in the Isabelle theorem-proving language and an implementation for the Cardano blockchain written in the Haskell/Plutus programming language.
+Early in 2023 the Marlowe specifications, on-chain implementation, and validator-test suite were audited by Tweag's High Assurance Software Group: their report is available online and it categorizes findings as either high, medium, or low severity.
+The present document summarizes the subsequent changes, made in response to the audit's findings, to the Marlowe specifications, its implementation as a Plutus smart-contract validator, and the property-based tests for that validator.
+All of the high-severity findings have been remedied or mitigated through changes to specifications, proofs, or implementation.
+Nearly all of the medium- and low-severity ones have also been addressed; justification for not addressing the remainder has been provided.
+
+The high-severity findings related to handling of negative deposits, prevention of "double satisfaction", enforcement of state invariants, an implementation difference between the formal specification and the Plutus implementation, and the Isabelle proof of money preservation.
+Marlowe's Plutus validator has been altered to resolve these first three findings so that it correctly prevents negative deposits from altering the balance of internal accounts, it only allows other Plutus scripts to run during transactions that do not make Marlowe payments, and it rigorously enforces invariants for initial and final states.
+The fourth finding regarding the Plutus implementation was mitigated to code analysis and property-based testing.
+The fifth finding about the money-preservation proof was remedied by revising the proof.
+
+The medium-severity findings typically suggested improvements to the Isabelle proofs, addition of more extensive property-based testing, correction of infelicities in the implementation, or safe-coding improvements.
+The low-severity findings generally centered on the usefulness of code comments, naming of variable bindings, omissions in specifications, and correction of typographical errors.
+All of the findings that affect the clarity of the specifications, the coverage of testing, or the robustness of implementation were made.
+Numerous additional property-based tests were added to the test suite as a result of audit findings.
+
+The changes outlined here have not been re-audited.
+The alterations to Marlowe's Plutus validator (detailed in the Appendix) have been kept to the minimum needed to address the audit findings, but avoiding extensive changes that would might have lessened the relevance of the audit.
+
+Finally, please note that the audit's scope included the Marlowe language and its Plutus interpreter, but not individual Marlowe contracts.
+Although the Marlowe language and its implementation on Cardano aim to make smart contracts safer, it is nevertheless still possible to design individual instances of Marlowe contracts that exhibit unwanted behavior.
+Such undesirable behavior may result either from poor contract design or lack of testing and analysis of the contract.
+[A best-practices guide](https://github.com/input-output-hk/marlowe-cardano/blob/main/marlowe/best-practices.md) and [a security guide](https://github.com/input-output-hk/marlowe-cardano/blob/main/marlowe/security.md) provide advice for the safe and secure design and testing of individual Marlowe contracts, which can themselves be subject to audit or certification.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @bwbush will add text thanking the audit team.
+We thank the High Assurance Software Group at Tweag for their diligent and expert analyses of the Marlowe specifications, proofs, implementations, and tests and their clear and thorough communication of their findings.
 
@@
 The audit report for Marlowe on Cardano was based on the commit hash [523f3d56](https://github.com/input-output-hk/marlowe-cardano/commit/523f3d56f22bf992ddb0b0c8a52bb7a5a188f9e9).
-The revisions and mitigations discussed here apply to [69468d66](https://github.com/input-output-hk/marlowe-cardano/commit/69468d6623e24a4ccd6659e378ae012c38ca01ce) of that repository.
+The revisions and mitigations discussed here apply to [ac65eba1](https://github.com/input-output-hk/marlowe-cardano/commit/) of that repository.
 The appendix to this report list the differences between pre- and post-audit validator code for Marlowe on Cardano.
@@
 
-Commit [47c9deb2](https://github.com/input-output-hk/marlowe-cardano/commit/47c9deb2047a240d78eaf62306278ecfbe624bea) elaborates the documentation for `refundOne`.
+Commit [ac65eba1](https://github.com/input-output-hk/marlowe-cardano/commit/ac65eba119b4eec8febc3723527edcd99914fdac) elaborates the documentation for `refundOne`.
 
@@
 ```bash
-git diff 523f3d56f22bf992ddb0b0c8a52bb7a5a188f9e9 47c9deb2047a240d78eaf62306278ecfbe624bea marlowe/src/Language/Marlowe/{Core/V1/Semantics*,Scripts.hs}
+git diff 523f3d56f22bf992ddb0b0c8a52bb7a5a188f9e9 ac65eba119b4eec8febc3723527edcd99914fdac marlowe/src/Language/Marlowe/{Core/V1/Semantics*,Scripts.hs}
 ```

### 2023-04-04 isabelle/generated/Arith.hs
PLT-3325: Property-based tests for assets perservation theorem

@@ -2,9 +2,9 @@
 
 module
   Arith(Int(..), integer_of_int, equal_int, less_eq_int, less_int, Nat,
-         integer_of_nat, Num(..), One, Zero, Zero_neq_one, suc, uminus_int,
-         zero_int, abs_int, one_int, plus_int, zero_nat, minus_int, times_int,
-         of_bool, divide_int)
+         integer_of_nat, Num(..), One, Zero, Zero_neq_one, nat, plus_nat, suc,
+         uminus_int, zero_int, abs_int, one_int, plus_int, zero_nat, minus_int,
+         times_int, of_bool, divide_int)
   where {
 
 import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
@@ -76,11 +76,19 @@ instance Zero Integer where {
   zero = (0 :: Integer);
 };
 
+instance Orderings.Ord Integer where {
+  less_eq = (\ a b -> a <= b);
+  less = (\ a b -> a < b);
+};
+
 class (One a, Zero a) => Zero_neq_one a where {
 };
 
 instance Zero_neq_one Integer where {
 };
+
+nat :: Int -> Nat;
+nat k = Nat (Orderings.max (0 :: Integer) (integer_of_int k));
 
 plus_nat :: Nat -> Nat -> Nat;
 plus_nat m n = Nat (integer_of_nat m + integer_of_nat n);

### 2023-04-04 isabelle/generated/AssetsPreservation.hs

@@ -1,0 +1,74 @@
+{-# LANGUAGE EmptyDataDecls, RankNTypes, ScopedTypeVariables #-}
+
+module
+  AssetsPreservation(assetsInState, assetsInTransactions,
+                      assetsInExternalPayments)
+  where {
+
+import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
+  (>>=), (>>), (=<<), (&&), (||), (^), (^^), (.), ($), ($!), (++), (!!), Eq,
+  error, id, return, not, fst, snd, map, filter, concat, concatMap, reverse,
+  zip, null, takeWhile, dropWhile, all, any, Integer, negate, abs, divMod,
+  String, Bool(True, False), Maybe(Nothing, Just));
+import qualified Prelude;
+import qualified Semantics;
+import qualified List;
+import qualified SemanticsTypes;
+import qualified MultiAssets;
+import qualified Arith;
+
+assetsInInput :: SemanticsTypes.Input -> MultiAssets.Assets;
+assetsInInput (SemanticsTypes.IDeposit uu uv tok money) =
+  MultiAssets.asset tok (Arith.nat money);
+assetsInInput (SemanticsTypes.IChoice uw ux) = MultiAssets.zero_Assets;
+assetsInInput SemanticsTypes.INotify = MultiAssets.zero_Assets;
+
+addAccountAssets ::
+  ((SemanticsTypes.Party, SemanticsTypes.Token), Arith.Int) ->
+    MultiAssets.Assets -> MultiAssets.Assets;
+addAccountAssets ((uu, t), v) b =
+  MultiAssets.plus_Assets b (MultiAssets.asset t (Arith.nat v));
+
+assetsInAccounts ::
+  [((SemanticsTypes.Party, SemanticsTypes.Token), Arith.Int)] ->
+    MultiAssets.Assets;
+assetsInAccounts accs =
+  List.foldr addAccountAssets accs MultiAssets.zero_Assets;
+
+assetsInState :: SemanticsTypes.State_ext () -> MultiAssets.Assets;
+assetsInState state = assetsInAccounts (SemanticsTypes.accounts state);
+
+addInputAssets ::
+  SemanticsTypes.Input -> MultiAssets.Assets -> MultiAssets.Assets;
+addInputAssets i a = MultiAssets.plus_Assets (assetsInInput i) a;
+
[…]

### 2023-04-04 isabelle/generated/List.hs

@@ -1,6 +1,6 @@
 {-# LANGUAGE EmptyDataDecls, RankNTypes, ScopedTypeVariables #-}
 
-module List(foldl, distinct, sorted_wrt) where {
+module List(foldl, foldr, distinct, sorted_wrt) where {
 
 import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
   (>>=), (>>), (=<<), (&&), (||), (^), (^^), (.), ($), ($!), (++), (!!), Eq,
@@ -12,6 +12,10 @@ import qualified Prelude;
 foldl :: forall a b. (a -> b -> a) -> a -> [b] -> a;
 foldl f a [] = a;
 foldl f a (x : xs) = foldl f (f a x) xs;
+
+foldr :: forall a b. (a -> b -> b) -> [a] -> b -> b;
+foldr f [] = id;
+foldr f (x : xs) = f x . foldr f xs;
 
 member :: forall a. (Eq a) => [a] -> a -> Bool;
 member [] y = False;

### 2023-04-04 isabelle/generated/MultiAssets.hs

@@ -1,0 +1,34 @@
+{-# LANGUAGE EmptyDataDecls, RankNTypes, ScopedTypeVariables #-}
+
+module MultiAssets(Assets, asset, assetValue, plus_Assets, zero_Assets) where {
+
+import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
+  (>>=), (>>), (=<<), (&&), (||), (^), (^^), (.), ($), ($!), (++), (!!), Eq,
+  error, id, return, not, fst, snd, map, filter, concat, concatMap, reverse,
+  zip, null, takeWhile, dropWhile, all, any, Integer, negate, abs, divMod,
+  String, Bool(True, False), Maybe(Nothing, Just));
+import qualified Prelude;
+import qualified SemanticsTypes;
+import qualified Arith;
+
+newtype Assets = Abs_Assets (SemanticsTypes.Token -> Arith.Nat);
+
+asset :: SemanticsTypes.Token -> Arith.Nat -> Assets;
+asset xb xc =
+  Abs_Assets
+    (\ t -> (if SemanticsTypes.equal_Token t xb then xc else Arith.zero_nat));
+
+rep_Assets :: Assets -> SemanticsTypes.Token -> Arith.Nat;
+rep_Assets (Abs_Assets x) = x;
+
+assetValue :: SemanticsTypes.Token -> Assets -> Arith.Nat;
+assetValue xa xb = rep_Assets xb xa;
+
+plus_Assets :: Assets -> Assets -> Assets;
+plus_Assets xb xc =
+  Abs_Assets (\ tok -> Arith.plus_nat (rep_Assets xb tok) (rep_Assets xc tok));
+
+zero_Assets :: Assets;
+zero_Assets = Abs_Assets (\ _ -> Arith.zero_nat);
+
+}

### 2023-04-04 marlowe-spec-test/src/Marlowe/Spec/Core/Semantics.hs

@@ -12,6 +12,7 @@ import Arith
     abs_int,
     integer_of_nat
   )
+import AssetsPreservation (assetsInTransactions, assetsInState, assetsInExternalPayments)
 import Control.Monad.IO.Class (MonadIO (..))
 import Data.Aeson (ToJSON (..))
 import qualified Data.Aeson as JSON
@@ -41,6 +42,7 @@ import Marlowe.Spec.Reproducible
     reproducibleProperty', ReproducibleTest,
   )
 import Marlowe.Utils (showAsJson)
+import MultiAssets (plus_Assets, Assets, assetValue)
 import Orderings (Ord (..))
 import PositiveAccounts (validAndPositive_state)
 import QuickCheck.GenT (frequency, suchThat, liftGen, GenT)
@@ -60,7 +62,7 @@ import Semantics
 import SemanticsTypes
   ( Contract (..),
     State_ext (..),
-    Value (..)
+    Value (..), Token, Case (..), Action (..)
   )
 import SingleInputTransactions (traceListToSingleInput)
 import Test.QuickCheck (cover, withMaxSuccess)
@@ -69,6 +71,7 @@ import Test.QuickCheck.Property (counterexample)
 import Test.Tasty (TestTree, testGroup)
 import Timeout (isClosedAndEmpty)
 import TransactionBound (maxTransactionsInitialState)
+import Control.Monad (join)
 
 tests :: InterpretJsonRequest -> TestTree
 tests i = testGroup "Semantics" [ testSemantics i , testGuarantees i ]
@@ -101,6 +104,8 @@ testGuarantees i = testGroup "Guarantees"
     , timedOutTransaction_closes_contractTest i
     -- CloseIsSafe.thy
     , closeIsSafeTest i
+    -- AssetsPreservation.thy
+    , playTrace_preserves_assets i
     ]
 
 -- The default maxSuccess is 100 and this tests modifies that to 500 as it was empirically found that 10 out of 10 times
@@ -405,3 +410,51 @@ closeIsSafeTest interpret = reproducibleProperty "theorem closeIsSafe" do
              :: PropertyM ReproducibleTest Bool
       JSON.Success (TransactionError err ) -> fail $ "Unexpected Transaction Error: " ++ show err
       _ -> fail "JSON parsing failed!"
+
+-- AssetsPreservation.thy
+--
+-- theorem playTrace_preserves_assets :
+--   assumes "playTrace slot contract txs = TransactionOutput out"
+--     shows "assetsInTransactions txs
[…]

### 2023-04-17 response-to-audit-report.md
Added clarification about each payment
Co-authored-by: Pablo Lamela  <pablo.lamela@iohk.io>

@@
 
-Although the audit report's suggestion of including a unique datum (such as that identifying the contract instance) in any output payment from the contract would likely prevent double satisfaction, implementing such a change in the Marlowe validators would have entailed such extensive changes to the specification and implementation that the overall applicability of the audit findings might subsequently be called into question.
+Although the audit report's suggestion of including a unique datum for each payment (such as that identifying the contract instance) in any output payment from the contract would likely prevent double satisfaction, implementing such a change in the Marlowe validators would have entailed such extensive changes to the specification and implementation that the overall applicability of the audit findings might subsequently be called into question.
 Instead, a minimal change to the specification and implementation mitigates this vulnerability: the specification was amended with a "Constraint 20.

### 2023-04-17 response-to-audit-report.md
Added clarification about abstraction over implementations
Co-authored-by: Pablo Lamela  <pablo.lamela@iohk.io>
Elaborated on existence of "test-spec" oracle.
Co-authored-by: Pablo Lamela  <pablo.lamela@iohk.io>

@@
 The porting of `MList` from Isabelle to Plutus would have increased the size and execution cost of the Marlowe semantics validator beyond the present limits of the Cardano ledger.
-Work is underway to modify the Isabelle semantics and proofs to use a data structure similar to Plutus's `AssocMap`.
+Work is underway to modify the Isabelle semantics and proofs to make obvious that no assumptions about the implementation of association maps are made.
 Instead of modifying the Marlowe semantics validator to address this finding, all usages of `AssocMap` were manually reviewed to verify that no vulnerability was introduced by its use (as compared to `MList`) *provided that the precondition of no duplicate entries in the `AssocMap` held*: the source code was annotated with informal reasoning documenting the safety of `AssocMap`'s use.
@@
 
-Subsequent to the commencement of the audit, a "test oracle" was developed that checks implementations of Marlowe semantics (such as the Plutus validator) against a reference implementation generated directly from the Isabelle specification.
+Subsequent to the commencement of the audit, a "test oracle" (called test spec) was developed that checks implementations of Marlowe semantics (such as the Plutus validator) against a reference implementation generated directly from the Isabelle specification.
 That generated Haskell implementation is backed by Isabelle proofs of the correctness of Haskell code generation by Isabelle.

### 2023-04-17 response-to-audit-report.md
Added cross-reference link
Co-authored-by: Pablo Lamela  <pablo.lamela@iohk.io>
Fixed spelling
Co-authored-by: Pablo Lamela  <pablo.lamela@iohk.io>

@@
 Commit [00010ea4](https://github.com/input-output-hk/marlowe-cardano/commit/00010ea48439892ff9eec54fb93929f56bcb6c7b) adds a paragraph of guidance to the Marlowe Cardano specification.
-The mitigation above for "2.1.2 Contracts vulnerable to double satisfaction attacks" implements validator changes to address this.
+The mitigation above for [2.1.2](#222-contracts-vulnerable-to-double-satisfaction-attacks-severity-high) implements validator changes to address this.
 
@@
 
-The monetary policy is not a concern of the Marlowe validator, but commit [fe00af7d](https://github.com/input-output-hk/marlowe-cardano/commit/fe00af7d622bbd17c3327363d4650ee544d1fff7) adds a sentance referencing the discussion of monetary policies for role tokens in the [Marlowe Best Practices Guide](https://github.com/input-output-hk/marlowe-cardano/blob/main/marlowe/best-practices.md).
+The monetary policy is not a concern of the Marlowe validator, but commit [fe00af7d](https://github.com/input-output-hk/marlowe-cardano/commit/fe00af7d622bbd17c3327363d4650ee544d1fff7) adds a sentence referencing the discussion of monetary policies for role tokens in the [Marlowe Best Practices Guide](https://github.com/input-output-hk/marlowe-cardano/blob/main/marlowe/best-practices.md).
 

### 2023-04-17 response-to-audit-report.md
Fixed word choice.
Co-authored-by: Pablo Lamela  <pablo.lamela@iohk.io>

@@
 
-The asymmetric is a convenience and practicality related to the manner in which coin selection and transaction balancing typically occurs nowadays in wallets.
+The asymmetry is a convenience and practicality related to the manner in which coin selection and transaction balancing typically occurs nowadays in wallets.
 Commit [4e81470d](https://github.com/input-output-hk/marlowe-cardano/commit/4e81470de90ddd6858c3ad8aace96e90df5a86ae) adds to the Marlowe Cardano specification a paragraph justifying this asymmetry in payout style and a comment to the Marlowe semantics validator along the same lines.

### 2023-04-17 response-to-audit-report.md
Fixed word choice.
Co-authored-by: Pablo Lamela  <pablo.lamela@iohk.io>

@@
 
-Although this behavior is required by the Isabelle specification and implemented in the Plutus validator, the comment on `applyCases` has been editing in commit [08d6f34a](https://github.com/input-output-hk/marlowe-cardano/commit/08d6f34ac91da7aa34e7414c5bae0e33e9235d74) to reinforce that the first applicable `Case` is taken for a `When`.
+Although this behavior is required by the Isabelle specification and implemented in the Plutus validator, the comment on `applyCases` has been edited in commit [08d6f34a](https://github.com/input-output-hk/marlowe-cardano/commit/08d6f34ac91da7aa34e7414c5bae0e33e9235d74) to reinforce that the first applicable `Case` is taken for a `When`.
 

### 2023-04-17 response-to-audit-report.md
Fixed word choice.
Co-authored-by: Pablo Lamela  <pablo.lamela@iohk.io>

@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here.
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: @hrajchert will add text here, since it relates to the Isabelle specification.
+![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
 
@@
 The discrepancies of implementation between the constraints in the specification and the Plutus code result from the emphasis on readability in the specification and of efficiency in the implementation.
-The need to minimize of validator size and execution cost in the implementation has sometimes resulted in less concise or less readable code.
+The need to minimize the validator size and execution cost in the implementation has sometimes resulted in less concise or less readable code.
 Note that the Plutus compiler can be quite sensitive to the precise expression of a constraint as Haskell code.

### 2023-04-18 marlowe-spec-test/src/Marlowe/Spec/Core/Arbitrary.hs
PLT-3313: Dropping `Arbitrary` subclass relationship in `SemiArbitrary`

@@ -1,124 +1,35 @@
 {-# LANGUAGE BlockArguments    #-}
-{-# LANGUAGE DeriveAnyClass    #-}
 {-# LANGUAGE FlexibleInstances #-}
 {-# LANGUAGE LambdaCase        #-}
 {-# LANGUAGE OverloadedStrings #-}
-{-# LANGUAGE RecordWildCards   #-}
-{-# LANGUAGE TupleSections     #-}
 
 {-# OPTIONS_GHC -fno-warn-orphans #-}
 
 module Marlowe.Spec.Core.Arbitrary
-  ( genValue
-  , genObservation
-  , genToken
-  , genParty
-  , genPayee
-  , genChoiceId
-  , genBound
-  , genState
-  , genStateWithSize
-  , genEnvironment
-  , genContract
-  , genContract'
-  , arbitraryContractWeighted
-  , genGoldenContract
-  , genInterval
-  , genTransaction
-  , genAction
-  , genInput
-  , genPayment
-  , genTransactionWarning
-  , genIntervalError
-  , genTransactionError
-  , genTransactionOutput
-  , arbitraryChoiceName
-  , arbitraryFibonacci
-  , arbitraryNonnegativeInteger
-  , arbitraryPositiveInteger
-  , arbitraryTransaction
-  , arbitraryTimeInterval
-  , arbitraryTimeIntervalAfter
-  , arbitraryTimeIntervalBefore
-  , arbitraryTimeIntervalAround
-  , arbitraryValidTransactions
-  , shrinkChoiceName
-  , shrinkTimeInterval
-  , RandomResponse (..)
-  , greater_eq
-  )
   where
 
-import Arith (divide_int)
 import qualified Arith
-import Control.Applicative ((<|>))
-import Control.Exception (Exception, throwIO)
-import Control.Monad (liftM2)
-import Control.Monad.IO.Class (liftIO)
-import Data.Aeson (FromJSON (..), ToJSON (..), object, withObject, (.:), (.=))
-import Data.Data (Proxy (..))
 import Data.List (nub)
-import Marlowe.Spec.Interpret (InterpretJsonRequest, Request (..), parseValidResponse)
 import Marlowe.Spec.TypeId (TypeId (..))
-import Orderings (Ord (..), max)
-import QuickCheck.GenT (GenT, MonadGen (..), frequency, resize, sized, suchThat, vectorOf)
-import Semantics
-  ( Payment (..),
-    TransactionError (..),
[…]

### 2023-04-18 marlowe-spec-test/src/Marlowe/Spec/Core/Generators.hs

@@ -1,0 +1,317 @@
+{-# LANGUAGE BlockArguments    #-}
+{-# LANGUAGE DeriveAnyClass    #-}
+{-# LANGUAGE FlexibleInstances #-}
+{-# LANGUAGE LambdaCase        #-}
+{-# LANGUAGE OverloadedStrings #-}
+
+{-# OPTIONS_GHC -fno-warn-orphans #-}
+
+module Marlowe.Spec.Core.Generators
+  ( genToken
+  , genParty
+  , genPayee
+  , genChoiceId
+  , genValue
+  , genObservation
+  , genPayment
+  , genTransaction
+  , genInput
+  , genAction
+  , genState
+  , genContract
+  , genTransactionOutput
+  , genTransactionWarning
+  , RandomResponse (..)
+  )
+  where
+
+import Control.Applicative ((<|>))
+import Control.Exception (Exception, throwIO)
+import Control.Monad (liftM2)
+import Control.Monad.IO.Class (liftIO)
+import Data.Aeson (FromJSON (..), ToJSON (..), object, withObject, (.:), (.=))
+import Data.Data (Proxy (..))
+import Marlowe.Spec.Core.Arbitrary
+  ( arbitraryChoiceName,
+    arbitraryInteger,
+    arbitraryNonnegativeInteger,
+    genBound,
+    genTransactionError,
+    genValueId,
+  )
+import Marlowe.Spec.Interpret (InterpretJsonRequest, Request (..), parseValidResponse)
+import Marlowe.Spec.TypeId (TypeId (..))
+import QuickCheck.GenT
+  ( GenT,
+    MonadGen (..),
+    frequency,
+    resize,
+    scale,
+    sized,
+    suchThat,
+    vectorOf,
+  )
+import Semantics
+  ( Payment (..),
+    TransactionOutput (..),
+    TransactionOutputRecord_ext (..),
+    TransactionWarning (..),
+    Transaction_ext (..),
+  )
+import SemanticsGuarantees (valid_state)
+import SemanticsTypes
+  ( Action (..),
+    Case (..),
+    ChoiceId (..),
+    Contract (..),
+    Input (..),
+    Observation (..),
+    Party (..),
+    Payee (..),
+    State_ext (..),
+    Token (..),
+    Value (..),
+  )
+import Test.QuickCheck (Gen, chooseInt, getSize)
+import Test.QuickCheck.Arbitrary (Arbitrary (..))
+
[…]

### 2023-04-18 marlowe-spec-test/src/Marlowe/Spec/Core/Guarantees.hs

@@ -1,0 +1,731 @@
+{-# LANGUAGE BlockArguments    #-}
+{-# LANGUAGE FlexibleInstances #-}
+{-# LANGUAGE LambdaCase        #-}
+{-# LANGUAGE OverloadedStrings #-}
+{-# LANGUAGE RecordWildCards   #-}
+{-# LANGUAGE TupleSections     #-}
+
+{-# OPTIONS_GHC -fno-warn-orphans #-}
+
+module Marlowe.Spec.Core.Guarantees
+  ( arbitraryContractWeighted
+  , arbitraryTransaction
+  , arbitraryValidTransactions
+  , genContext
+  , Context
+  , SemiArbitrary (..)
+  , tests
+  )
+  where
+
+import Arith ( integer_of_nat)
+import qualified Arith
+import Control.Monad.IO.Class (MonadIO (..))
+import qualified Data.Aeson as JSON
+import Data.Function (on)
+import Data.List (nubBy)
+import Data.Map (Map, fromList, toList)
+import qualified Examples.Escrow as Escrow
+import qualified Examples.Swap as Swap
+import Marlowe.Spec.Core.Arbitrary
+  ( arbitraryChoiceName,
+    arbitraryInteger,
+    arbitraryNonnegativeInteger,
+    arbitraryPositiveInteger,
+    arbitraryTimeIntervalAfter,
+    arbitraryTimeIntervalBefore,
+    chooseinteger,
+    genBound,
+    genTransactionError,
+    genValueId,
+    greater_eq
+  )
+import Marlowe.Spec.Core.Generators (genParty, genToken)
+import Marlowe.Spec.Interpret
+  ( InterpretJsonRequest,
+    Request (..),
+    Response (..)
+  )
+import Marlowe.Spec.Reproducible
+  ( ReproducibleTest,
+    generate,
+    generateT,
+    reproducibleProperty
+  )
+import Marlowe.Utils (showAsJson)
+import Orderings (Ord (..), max)
+import PositiveAccounts (validAndPositive_state)
+import QuickCheck.GenT
+  ( Gen,
+    GenT,
+    MonadGen (..),
+    frequency,
+    listOf,
+    resize,
+    sized,
+    suchThat,
+    vectorOf
+  )
+import Semantics
+  ( Payment (..),
+    TransactionOutput (..),
+    TransactionOutputRecord_ext (..),
+    TransactionWarning (..),
+    Transaction_ext (..),
[…]

### 2023-04-18 marlowe-spec-test/src/Marlowe/Spec/Core/Semantics.hs

@@ -1,6 +1,5 @@
 {-# LANGUAGE OverloadedStrings #-}
 {-# LANGUAGE BlockArguments #-}
-{-# LANGUAGE TupleSections     #-}
 
 module Marlowe.Spec.Core.Semantics
   ( tests
@@ -8,102 +7,59 @@ module Marlowe.Spec.Core.Semantics
   where
 
 import Arith
-  ( Int (..),
+  ( abs_int,
     less_int,
-    abs_int,
-    integer_of_nat
   )
 import Control.Monad.IO.Class (MonadIO (..))
 import Data.Aeson (ToJSON (..))
 import qualified Data.Aeson as JSON
-import Marlowe.Spec.Core.Arbitrary
-  ( arbitraryTransaction,
-    arbitraryTimeIntervalAfter,
+import Marlowe.Spec.Core.Arbitrary (genEnvironment)
+import Marlowe.Spec.Core.Guarantees
+  ( SemiArbitrary (..),
+    arbitraryTransaction,
     arbitraryValidTransactions,
-    genContract,
-    genEnvironment,
-    genState,
-    genStateWithSize,
-    genValue,
-    genObservation,
-    greater_eq,
-    arbitraryContractWeighted,
-    genGoldenContract
+    genContext
   )
 import Marlowe.Spec.Interpret
   ( InterpretJsonRequest,
     Request (..),
-    Response (..),
+    Response (..)
   )
 import Marlowe.Spec.Reproducible
   ( assertResponse,
     generate,
     generateT,
     reproducibleProperty,
-    reproducibleProperty', ReproducibleTest,
+    reproducibleProperty'
   )
 import Marlowe.Utils (showAsJson)
 import Orderings (Ord (..))
-import PositiveAccounts (validAndPositive_state)
-import QuickCheck.GenT (frequency, suchThat, liftGen, GenT)
+import QuickCheck.GenT (suchThat)
 import Semantics
   ( TransactionOutput (..),
-    TransactionOutputRecord_ext (TransactionOutputRecord_ext),
+    TransactionOutputRecord_ext (..),
     Transaction_ext (..),
     computeTransaction,
-    evalValue,
     evalObservation,
-    isQuiescent,
-    txOutWarnings,
-    txOutPayments,
-    maxTimeContract,
-    inputs,
+    evalValue
   )
 import SemanticsTypes
   ( Contract (..),
     State_ext (..),
     Value (..)
   )
[…]

### 2023-04-18 marlowe-spec-test/src/Marlowe/Spec/Core/Serialization/Json.hs

@@ -8,23 +8,39 @@
 module Marlowe.Spec.Core.Serialization.Json where
 
 import Control.Applicative ((<|>))
-import Data.Aeson.Types (Result(..), ToJSON(..), FromJSON(..))
-import Data.Aeson (object, (.=), (.:), withObject)
-import qualified Data.Aeson.Types as JSON
-import Data.Text as T
-import Data.Proxy (Proxy(..))
-import MarloweCoreJson
-import GHC.Stack (HasCallStack)
-import Test.Tasty (TestTree, testGroup)
-import Marlowe.Spec.Interpret (Response (..), InterpretJsonRequest, Request (..), testResponse)
-import Marlowe.Spec.TypeId (TypeId(..), HasTypeId (..))
-import Test.Tasty.HUnit (Assertion, assertBool, testCase, (@=?))
-import qualified SemanticsTypes as C
-import Marlowe.Spec.Core.Arbitrary (genToken, genParty, genPayee, genChoiceId, genBound, genValue, genObservation, genAction, genContract, genInput, genTransaction, genPayment, genState, genTransactionWarning, genIntervalError, genTransactionError, genTransactionOutput)
 import Control.Monad.IO.Class (MonadIO)
-import Marlowe.Spec.Reproducible (generate, generateT, reproducibleProperty, assertResponse)
+import Data.Aeson (object, withObject, (.:), (.=))
+import Data.Aeson.Types (FromJSON (..), Result (..), ToJSON (..))
+import qualified Data.Aeson.Types as JSON
+import Data.Proxy (Proxy (..))
+import Data.Text as T
+import GHC.Stack (HasCallStack)
+import Marlowe.Spec.Core.Arbitrary
+import Marlowe.Spec.Core.Generators
+  ( genAction,
+    genChoiceId,
+    genContract,
+    genInput,
+    genObservation,
+    genParty,
+    genPayee,
+    genPayment,
+    genState,
+    genToken,
+    genTransaction,
+    genTransactionOutput,
+    genTransactionWarning,
+    genValue,
+  )
+import Marlowe.Spec.Interpret (InterpretJsonRequest, Request (..), Response (..), testResponse)
+import Marlowe.Spec.Reproducible (assertResponse, generate, generateT, reproducibleProperty)
+import Marlowe.Spec.TypeId (HasTypeId (..), TypeId (..))
+import MarloweCoreJson
 import QuickCheck.GenT (MonadGen (resize))
[…]

### 2023-04-18 marlowe-spec-test/src/Marlowe/Spec/LocalInterpret.hs

@@ -1,6 +1,5 @@
 {-# LANGUAGE BlockArguments    #-}
 {-# LANGUAGE FlexibleInstances #-}
-{-# LANGUAGE LambdaCase        #-}
 {-# LANGUAGE OverloadedStrings #-}
 
 {-# OPTIONS_GHC -fno-warn-orphans #-}
@@ -13,33 +12,19 @@ module Marlowe.Spec.LocalInterpret
 import qualified Arith (Int (..))
 import Data.Aeson (FromJSON, Result (..), ToJSON)
 import qualified Data.Aeson as JSON
-import Marlowe.Spec.Interpret (Response(..), Request(..))
-import Marlowe.Spec.TypeId (TypeId (..), fromTypeName)
-import Marlowe.Spec.Core.Serialization.Json
 import Data.Data (Proxy)
-import Marlowe.Spec.Core.Arbitrary
-  ( arbitraryChoiceName,
-    arbitraryFibonacci,
-    arbitraryPositiveInteger,
-    shrinkChoiceName
-  )
-import qualified Marlowe.Spec.Core.Arbitrary as RandomResponse
+import qualified Marlowe.Spec.Core.Generators as Gen
+import Marlowe.Spec.Core.Serialization.Json (SerializationResponse (..))
+import Marlowe.Spec.Interpret (Request (..), Response (..))
+import Marlowe.Spec.TypeId (TypeId (..), fromTypeName)
 import Semantics
-  ( playTrace,
-    computeTransaction,
+  ( computeTransaction,
+    evalObservation,
     evalValue,
-    evalObservation
+    playTrace
   )
-import SemanticsTypes
-  ( Action (..),
-    ChoiceId (..),
-    Observation (..),
-    Party (..),
-    Payee (..),
-    Token (Token),
-    Value (..),
-  )
-import Test.QuickCheck (Arbitrary (..), frequency, resize, sized, suchThat)
+import SemanticsTypes (Party (..), Token (..))
+import Test.QuickCheck (Arbitrary (..))
 import Test.QuickCheck.Gen (Gen (..))
 import Test.QuickCheck.Random (mkQCGen)
 
@@ -74,173 +59,11 @@ interpretLocal (GenerateRandomValue t@(TypeId name _) size seed) =
     $ RequestResponse
     $ JSON.toJSON
     $ case name of
-      "Core.Token" -> RandomResponse.RandomValue $ JSON.toJSON (generate' arbitrary :: Token)
[…]

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/Arbitrary.hs
PLT-3313: Organized imports, minor cleanups

@@ -6,6 +6,25 @@
 {-# OPTIONS_GHC -fno-warn-orphans #-}
 
 module Marlowe.Spec.Core.Arbitrary
+  ( arbitraryChoiceName
+  , arbitraryFibonacci
+  , arbitraryInteger
+  , arbitraryNonnegativeInteger
+  , arbitraryPositiveInteger
+  , arbitraryTimeInterval
+  , arbitraryTimeIntervalAfter
+  , arbitraryTimeIntervalBefore
+  , arbitraryTimeIntervalAround
+  , chooseArithInt
+  , genBound
+  , genEnvironment
+  , genInterval
+  , genIntervalError
+  , genTransactionError
+  , genValueId
+  , shrinkChoiceName
+  , shrinkTimeInterval
+  )
   where
 
 import qualified Arith
@@ -60,8 +79,8 @@ arbitraryInteger = Arith.Int_of_integer <$>
     ]
 
 -- | An arbitrary integer within an interval.
-chooseinteger :: (Arith.Int, Arith.Int) -> Gen Arith.Int
-chooseinteger (Arith.Int_of_integer i, Arith.Int_of_integer j) =
+chooseArithInt :: (Arith.Int, Arith.Int) -> Gen Arith.Int
+chooseArithInt (Arith.Int_of_integer i, Arith.Int_of_integer j) =
   Arith.Int_of_integer <$> QC.chooseInteger (i, j)
 
 -- | An arbitrary non-negative integer, mostly small.
@@ -103,7 +122,7 @@ arbitraryTimeIntervalBefore :: Arith.Int -> Arith.Int -> Gen (Arith.Int, Arith.I
 arbitraryTimeIntervalBefore lower upper =
   do
     start <- arbitraryNonnegativeInteger `suchThat` greater_eq lower
-    duration <- chooseinteger (0, upper - start - 1)
+    duration <- chooseArithInt (0, upper - start - 1)
     pure (start, start + duration)
 
 -- | Generate a semi-random time interval after a given time.
@@ -113,6 +132,9 @@ arbitraryTimeIntervalAfter lower =
     start <- arbitraryNonnegativeInteger `suchThat` less_eq lower
     duration <- arbitraryNonnegativeInteger
     pure (start, start + duration)
+
+greater_eq :: Orderings.Ord a => a -> a -> Bool
+greater_eq = flip less_eq
 
 -- | Shrink a generated time interval.
[…]

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/Guarantees.hs

@@ -3,13 +3,7 @@
 {-# LANGUAGE OverloadedStrings #-}
 
 module Marlowe.Spec.Core.Guarantees
-  ( arbitraryContractWeighted
-  , arbitraryTransaction
-  , arbitraryValidTransactions
-  , genContext
-  , Context
-  , SemiArbitrary (..)
-  , tests
+  ( tests
   )
   where
 
@@ -17,10 +11,7 @@ import Arith (integer_of_nat)
 import qualified Arith
 import Control.Monad.IO.Class (MonadIO (..))
 import qualified Data.Aeson as JSON
-import Marlowe.Spec.Core.Arbitrary
-  ( arbitraryTimeIntervalAfter,
-    greater_eq,
-  )
+import Marlowe.Spec.Core.Arbitrary (arbitraryTimeIntervalAfter)
 import Marlowe.Spec.Core.SemiArbitrary
 import Marlowe.Spec.Interpret
   ( InterpretJsonRequest,
@@ -149,10 +140,6 @@ traceToSingleInputIsEquivalentTest interpret = reproducibleProperty "theorem tra
       (JSON.Success _ , JSON.Success (TransactionError _)) -> pre False
       _ -> fail "JSON parsing failed"
 
-isWhen :: Contract -> Bool
-isWhen When {} = True
-isWhen _ = False
-
 -- QuiescentResults.thy
 --
 -- theorem computeTransactionIsQuiescent:
@@ -187,6 +174,10 @@ computeTransactionIsQuiescentTest interpret = reproducibleProperty "theorem comp
              :: PropertyM ReproducibleTest Bool
       JSON.Success (TransactionError err ) -> fail $ "Unexpected Transaction Error: " ++ show err
       _ -> fail "JSON parsing failed!"
+  where
+    isWhen :: Contract -> Bool
+    isWhen When {} = True
+    isWhen _ = False
 
 -- QuiescentResults.thy
 --
@@ -259,6 +250,8 @@ timedOutTransaction_closes_contractTest interpret = reproducibleProperty "theore
              :: PropertyM ReproducibleTest Bool
       JSON.Success (TransactionError err ) -> fail $ "Unexpected Transaction Error: " ++ show err
       _ -> fail "JSON parsing failed!"
+  where
+    greater_eq = flip less_eq
 
 -- CloseIsSafe.thy
 --

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/Semantics.hs

@@ -14,23 +14,23 @@ import Control.Monad.IO.Class (MonadIO (..))
 import Data.Aeson (ToJSON (..))
 import qualified Data.Aeson as JSON
 import Marlowe.Spec.Core.Arbitrary (genEnvironment)
-import Marlowe.Spec.Core.Guarantees
+import Marlowe.Spec.Core.SemiArbitrary
   ( SemiArbitrary (..),
     arbitraryTransaction,
     arbitraryValidTransactions,
-    genContext
+    genContext,
   )
 import Marlowe.Spec.Interpret
   ( InterpretJsonRequest,
     Request (..),
-    Response (..)
+    Response (..),
   )
 import Marlowe.Spec.Reproducible
   ( assertResponse,
     generate,
     generateT,
     reproducibleProperty,
-    reproducibleProperty'
+    reproducibleProperty',
   )
 import Marlowe.Utils (showAsJson)
 import Orderings (Ord (..))
@@ -41,12 +41,12 @@ import Semantics
     Transaction_ext (..),
     computeTransaction,
     evalObservation,
-    evalValue
+    evalValue,
   )
 import SemanticsTypes
   ( Contract (..),
     State_ext (..),
-    Value (..)
+    Value (..),
   )
 import Test.QuickCheck (withMaxSuccess)
 import Test.QuickCheck.Monadic (PropertyM, assert, monitor, run)

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/SemiArbitrary.hs

@@ -31,7 +31,7 @@ import Marlowe.Spec.Core.Arbitrary
     arbitraryPositiveInteger,
     arbitraryTimeIntervalAfter,
     arbitraryTimeIntervalBefore,
-    chooseinteger,
+    chooseArithInt,
     genBound,
     genTransactionError,
     genValueId,
@@ -446,7 +446,7 @@ arbitraryTransaction state (When cases timeout _) =
                  Deposit a p t v -> pure . IDeposit a p t $ evalValue (Environment_ext times ()) state v
                  Choice n bs     -> do
                                       Bound lower upper <- QC.elements bs
-                                      IChoice n <$> chooseinteger (lower, upper)
+                                      IChoice n <$> chooseArithInt (lower, upper)
                  Notify _        -> pure INotify
           case cont of
             Close -> pure $ Transaction_ext times [i] ()

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/Arbitrary.hs
PLT-3482: shrinking

@@ -355,6 +355,18 @@ instance Arbitrary Value where
       genUseValue = UseValue <$> arbitrary
       genCond = Cond <$> arbitrary <*> arbitrary <*> arbitrary
 
+  shrink (AvailableMoney x y) = [AvailableMoney x' y | x' <- shrink x] ++ [AvailableMoney x y' | y' <- shrink y]
+  shrink (Constant x)         = [Constant x' | x' <- shrink x]
+  shrink (NegValue x)         = [NegValue x' | x' <- shrink x]
+  shrink (AddValue x y)       = [AddValue x' y | x' <- shrink x] ++ [AddValue x y' | y' <- shrink y]
+  shrink (SubValue x y)       = [SubValue x' y | x' <- shrink x] ++ [SubValue x y' | y' <- shrink y]
+  shrink (MulValue x y)       = [MulValue x' y | x' <- shrink x] ++ [MulValue x y' | y' <- shrink y]
+  shrink (DivValue x y)       = [DivValue x' y | x' <- shrink x] ++ [DivValue x y' | y' <- shrink y]
+  shrink (ChoiceValue x)      = [ChoiceValue x' | x' <- shrink x]
+  shrink (UseValue x)         = [UseValue x' | x' <- shrink x]
+  shrink (Cond o x y)         = [Cond o' x y | o' <- shrink o] ++ [Cond o x' y | x' <- shrink x] ++ [Cond o x y' | y' <- shrink y]
+  shrink _                    = []
+
 instance Arbitrary Observation where
   arbitrary =
     frequency

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/Arbitrary.hs
PLT-3482: Added shrink

@@ -15,40 +15,46 @@ module Marlowe.Spec.Core.Arbitrary
   , arbitraryTimeIntervalAfter
   , arbitraryTimeIntervalBefore
   , arbitraryTimeIntervalAround
+  , arbitraryTransaction
+  , arbitraryValidTransactions
   , chooseArithInt
-  , genBound
-  , genEnvironment
-  , genInterval
-  , genIntervalError
-  , genTransactionError
-  , genValueId
-  , shrinkChoiceName
-  , shrinkTimeInterval
   )
   where
 
 import qualified Arith
 import Data.List (nub)
-import Marlowe.Spec.TypeId (TypeId (..))
-import Orderings (Ord (..))
-import Semantics (TransactionError (..))
+import Marlowe.Spec.TypeId ()
+import Orderings (Ord (..), max)
+import QuickCheck.GenT (MonadGen (..), frequency, resize, sized, suchThat)
+import Semantics
+  ( TransactionError (..),
+    TransactionOutput (..),
+    TransactionOutputRecord_ext (..),
+    Transaction_ext (..),
+    computeTransaction,
+    evalValue,
+  )
 import SemanticsTypes
   ( Action (..),
     Bound (..),
+    Case (..),
     ChoiceId (..),
+    Contract (..),
     Environment_ext (..),
+    Input (..),
+    IntervalError (..),
     Observation (..),
     Party (..),
     Payee (..),
+    State_ext (..),
     Token (..),
     Value (..),
     ValueId (..),
-    IntervalError (..),
+    minTime,
   )
 import Test.QuickCheck (Gen)
 import Test.QuickCheck.Arbitrary (Arbitrary (..))
-import QuickCheck.GenT (MonadGen (..), frequency, resize, sized, suchThat)
-import qualified Test.QuickCheck.Gen as QC (chooseInteger)
+import qualified Test.QuickCheck.Gen as QC (chooseInteger, elements)
 
 -- | Part of the Fibonacci sequence.
 fibonaccis :: Num a => [a]
@@ -136,22 +142,6 @@ arbitraryTimeIntervalAfter lower =
 greater_eq :: Orderings.Ord a => a -> a -> Bool
 greater_eq = flip less_eq
 
--- | Shrink a generated time interval.
-shrinkTimeInterval :: (Arith.Int, Arith.Int) -> [(Arith.Int, Arith.Int)]
-shrinkTimeInterval (start, end) =
-  let
[…]

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/Generators.hs

@@ -35,9 +35,6 @@ import Marlowe.Spec.Core.Arbitrary
   ( arbitraryChoiceName,
     arbitraryInteger,
     arbitraryNonnegativeInteger,
-    genBound,
-    genTransactionError,
-    genValueId,
   )
 import Marlowe.Spec.Interpret (InterpretJsonRequest, Request (..), parseValidResponse)
 import Marlowe.Spec.TypeId (TypeId (..))
@@ -176,7 +173,7 @@ genValue i = sized (
   genChoiceValue = ChoiceValue <$> genChoiceId i
   genTimeIntervalStart = pure TimeIntervalStart
   genTimeIntervalEnd = pure TimeIntervalEnd
-  genUseValue = UseValue <$> liftGen genValueId
+  genUseValue = UseValue <$> liftGen arbitrary
   genCond = Cond <$> genObservation i <*> genValue i <*> genValue i
 
 genObservation :: InterpretJsonRequest -> GenT IO Observation
@@ -218,7 +215,7 @@ genAction i = frequency
       [ (7, Deposit <$> genParty i <*> genParty i <*> genToken i <*> genValue i)
       , (2, do
           lSize <- liftGen $ chooseInt (1, 4)
-          Choice <$> genChoiceId i <*> vectorOf lSize (liftGen genBound)
+          Choice <$> genChoiceId i <*> vectorOf lSize (liftGen arbitrary)
         )
       , (1, Notify <$> genObservation i)
       ]
@@ -254,7 +251,7 @@ genState i = rawGen `suchThat` valid_state
       boundSize <- liftGen $ chooseInt (0, min n 4)
       accounts <- vectorOf accountSize ((genParty i >*< genToken i) >*< liftGen arbitraryNonnegativeInteger)
       choices <- vectorOf choiceSize (genChoiceId i >*< liftGen arbitraryInteger)
-      bounds <- vectorOf boundSize (liftGen genValueId >*< liftGen arbitraryInteger)
+      bounds <- vectorOf boundSize (liftGen arbitrary >*< liftGen arbitraryInteger)
       minTime' <- liftGen arbitraryInteger
       pure $ State_ext accounts choices bounds minTime' ()
     )
@@ -263,14 +260,14 @@ genTransactionWarning :: InterpretJsonRequest -> GenT IO TransactionWarning
 genTransactionWarning i = frequency
[…]

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/Guarantees.hs

@@ -11,7 +11,7 @@ import Arith (integer_of_nat)
 import qualified Arith
 import Control.Monad.IO.Class (MonadIO (..))
 import qualified Data.Aeson as JSON
-import Marlowe.Spec.Core.Arbitrary (arbitraryTimeIntervalAfter)
+import Marlowe.Spec.Core.Arbitrary (arbitraryTimeIntervalAfter, arbitraryValidTransactions, arbitraryTransaction)
 import Marlowe.Spec.Core.SemiArbitrary
 import Marlowe.Spec.Interpret
   ( InterpretJsonRequest,

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/Semantics.hs

@@ -13,12 +13,13 @@ import Arith
 import Control.Monad.IO.Class (MonadIO (..))
 import Data.Aeson (ToJSON (..))
 import qualified Data.Aeson as JSON
-import Marlowe.Spec.Core.Arbitrary (genEnvironment)
+import Marlowe.Spec.Core.Arbitrary
+  ( arbitraryTransaction,
+    arbitraryValidTransactions,
+  )
 import Marlowe.Spec.Core.SemiArbitrary
   ( SemiArbitrary (..),
-    arbitraryTransaction,
-    arbitraryValidTransactions,
-    genContext,
+    genContext
   )
 import Marlowe.Spec.Interpret
   ( InterpretJsonRequest,
@@ -48,7 +49,7 @@ import SemanticsTypes
     State_ext (..),
     Value (..),
   )
-import Test.QuickCheck (withMaxSuccess)
+import Test.QuickCheck (Arbitrary (..), withMaxSuccess)
 import Test.QuickCheck.Monadic (PropertyM, assert, monitor, run)
 import Test.QuickCheck.Property (counterexample)
 import Test.Tasty (TestTree, testGroup)
@@ -70,7 +71,7 @@ tests i = testGroup "Semantics"
 evalValueTest :: InterpretJsonRequest -> TestTree
 evalValueTest interpret = reproducibleProperty' "Eval Value" (withMaxSuccess 500) do
     context <- run $ generateT $ genContext interpret
-    env <- run $ generate genEnvironment
+    env <- run $ generate arbitrary
     state <- run $ generate $ semiArbitrary context
     value <- run $ generate $ semiArbitrary context
     let
@@ -82,7 +83,7 @@ evalValueTest interpret = reproducibleProperty' "Eval Value" (withMaxSuccess 500
 evalObservationTest :: InterpretJsonRequest -> TestTree
 evalObservationTest interpret = reproducibleProperty "Eval Observation" do
     context <- run $ generateT $ genContext interpret
-    env <- run $ generate genEnvironment
+    env <- run $ generate arbitrary
     state <- run $ generate $ semiArbitrary context
     observation <- run $ generate $ semiArbitrary context
     let
@@ -94,7 +95,7 @@ evalObservationTest interpret = reproducibleProperty "Eval Observation" do
 divisionRoundsTowardsZeroTest :: InterpretJsonRequest -> TestTree
 divisionRoundsTowardsZeroTest interpret = reproducibleProperty "Division rounding"  do
[…]

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/SemiArbitrary.hs

@@ -9,8 +9,6 @@
 
 module Marlowe.Spec.Core.SemiArbitrary
   ( arbitraryContractWeighted
-  , arbitraryTransaction
-  , arbitraryValidTransactions
   , genContext
   , Context (..)
   , SemiArbitrary (..)
@@ -29,18 +27,11 @@ import Marlowe.Spec.Core.Arbitrary
     arbitraryInteger,
     arbitraryNonnegativeInteger,
     arbitraryPositiveInteger,
-    arbitraryTimeIntervalAfter,
-    arbitraryTimeIntervalBefore,
-    chooseArithInt,
-    genBound,
-    genTransactionError,
-    genValueId,
   )
 import Marlowe.Spec.Core.Generators (genParty, genToken)
 import Marlowe.Spec.Interpret
   ( InterpretJsonRequest,
   )
-import Orderings (Ord (..), max)
 import QuickCheck.GenT
   ( Gen,
     GenT,
@@ -56,8 +47,6 @@ import Semantics
     TransactionOutputRecord_ext (..),
     TransactionWarning (..),
     Transaction_ext (..),
-    computeTransaction,
-    evalValue,
   )
 import SemanticsTypes
   ( Action (..),
@@ -65,7 +54,6 @@ import SemanticsTypes
     Case (..),
     ChoiceId (..),
     Contract (..),
-    Environment_ext (..),
     Input (..),
     Observation (..),
     Party (..),
@@ -74,11 +62,9 @@ import SemanticsTypes
     Token (..),
     Value (..),
     ValueId (..),
-    minTime,
   )
 import Test.QuickCheck (chooseInt, elements)
 import Test.QuickCheck.Arbitrary (Arbitrary (..))
-import qualified Test.QuickCheck.Gen as QC (elements)
 
 -- | Class for arbitrary values with respect to a context.
 class SemiArbitrary a where
@@ -118,7 +104,7 @@ genContext interpret = sized \n ->
     chosenNums <- liftGen $ listOf' n arbitraryInteger
     choiceIds <- listOf' n $ ChoiceId
       <$> (liftGen $ perturb arbitraryChoiceName choiceNames) <*> perturbM (genParty interpret) parties
-    valueIds <- liftGen $ listOf' n genValueId
+    valueIds <- liftGen $ listOf' n arbitrary
     values <- liftGen $ listOf' n arbitraryInteger
[…]

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/Serialization/Json.hs

@@ -15,7 +15,6 @@ import qualified Data.Aeson.Types as JSON
 import Data.Proxy (Proxy (..))
 import Data.Text as T
 import GHC.Stack (HasCallStack)
-import Marlowe.Spec.Core.Arbitrary
 import Marlowe.Spec.Core.Generators
   ( genAction,
     genChoiceId,
@@ -37,10 +36,12 @@ import Marlowe.Spec.Reproducible (assertResponse, generate, generateT, reproduci
 import Marlowe.Spec.TypeId (HasTypeId (..), TypeId (..))
 import MarloweCoreJson
 import QuickCheck.GenT (MonadGen (resize))
+import qualified Semantics as C
 import qualified SemanticsTypes as C
 import Test.QuickCheck.Monadic (PropertyM, run)
 import Test.Tasty (TestTree, testGroup)
 import Test.Tasty.HUnit (Assertion, assertBool, testCase, (@=?))
+import Test.QuickCheck (Gen, Arbitrary (..))
 
 data SerializationResponse transport
   = SerializationSuccess transport
@@ -202,7 +203,7 @@ arbitraryChoiceIdTest i = reproducibleProperty "ChoiceId" do
 
 arbitraryBoundTest :: InterpretJsonRequest -> TestTree
 arbitraryBoundTest i = reproducibleProperty "Bound" do
-  bound <- run $ generate genBound
+  bound <- run $ generate (arbitrary :: Gen C.Bound)
   propertyRoundtripTest i bound
 
 arbitraryValueTest :: InterpretJsonRequest -> TestTree
@@ -252,12 +253,12 @@ arbitraryTransactionWarningTest i = reproducibleProperty "TransactionWarning" do
 
 arbitraryIntervalErrorTest :: InterpretJsonRequest -> TestTree
 arbitraryIntervalErrorTest i = reproducibleProperty "IntervalError" do
-  warning <- run $ generate genIntervalError
+  warning <- run $ generate (arbitrary :: Gen C.IntervalError)
   propertyRoundtripTest i warning
 
 arbitraryTransactionErrorTest :: InterpretJsonRequest -> TestTree
 arbitraryTransactionErrorTest i = reproducibleProperty "TransactionError" do
-  txError <- run $ generate genTransactionError
+  txError <- run $ generate (arbitrary :: Gen C.TransactionError)
   propertyRoundtripTest i txError
 
 arbitraryTransactionOutputTest :: InterpretJsonRequest -> TestTree

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/SemiArbitrary.hs
PLT-3313: formatting

@@ -8,9 +8,11 @@
 {-# OPTIONS_GHC -fno-warn-orphans #-}
 
 module Marlowe.Spec.Core.SemiArbitrary
-  ( arbitraryContractWeighted
+  ( arbitraryCaseWeighted
+  , arbitraryContractWeighted
+  , arbitraryContractSized
+  , Context
   , genContext
-  , Context (..)
   , SemiArbitrary (..)
   )
   where
@@ -96,32 +98,53 @@ data Context =
 
 genContext :: InterpretJsonRequest -> GenT IO Context
 genContext interpret = sized \n ->
-  do
-    parties <- listOf' n $ genParty interpret
-    tokens <- listOf' n $ genToken interpret
-    amounts <- liftGen $ listOf' n arbitraryPositiveInteger
-    choiceNames <- liftGen $ listOf' n arbitraryChoiceName
-    chosenNums <- liftGen $ listOf' n arbitraryInteger
-    choiceIds <- listOf' n $ ChoiceId
-      <$> (liftGen $ perturb arbitraryChoiceName choiceNames) <*> perturbM (genParty interpret) parties
-    valueIds <- liftGen $ listOf' n arbitrary
-    values <- liftGen $ listOf' n arbitraryInteger
-    times <- liftGen $ listOf' n arbitraryPositiveInteger
-    caccounts <- fromList . nubBy ((==) `on` fst)
-      <$> listOf' n ((,) <$> ((,) <$> perturbM (genParty interpret) parties <*> perturbM (genToken interpret) tokens) <*> liftGen (perturb arbitraryPositiveInteger amounts))
-    cchoices <- fromList . nubBy ((==) `on` fst)
-      <$> listOf' n ((,) <$> liftGen (elements choiceIds) <*> liftGen (perturb arbitraryInteger chosenNums))
-    cboundValues <- fromList . nubBy ((==) `on` fst)
-      <$> listOf' n ((,) <$> liftGen (perturb arbitrary valueIds) <*> liftGen (perturb arbitraryInteger values))
-    pure Context {..}
+  let listOf gen = choose (1, n) >>= flip replicateM gen
+      listOfGen = liftGen . listOf
+      genParty' = genParty interpret
+      genToken' = genToken interpret
+   in do
+        parties <- listOf genParty'
+        tokens <- listOf genToken'
+        amounts <- listOfGen arbitraryPositiveInteger
+        choiceNames <- listOfGen arbitraryChoiceName
[…]

### 2023-04-19 response-to-audit-report.md

@@
 The medium-severity concerns in the audit report center around missing tests or name shadowing.
-With the exception of "2.7.1 More precise failure checks", these have all been remedied either by code changes (in the case of the shadowing) or by the addition of a significant augmentation of the property-based test suite.
+These have all been remedied either by code changes (in the case of the shadowing) or by the addition of a significant augmentation of the property-based test suite.
 The audit's low-severity concerns generally relate to insufficiently detailed text in the Marlowe Cardano specification, the need for more elaborate comments in the code, naming of variable bindings, or typographical errors.
@@
 
-A future version of the Marlowe test suite for `checkSemanticsTransaction` and `checkPayoutTransaction` will instrument Plutus scripts for the Marlowe validators so that the precise cause of the test failure can be verified.
+Commit [8d18385b](https://github.com/input-output-hk/marlowe-cardano/commit/8d18385b107fbce2f7a4559e6017d9e228ab958b) to the Marlowe test suite for `checkSemanticsTransaction` instruments Plutus scripts for the Marlowe validator so that the precise cause of the test failure is verified.
 

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/Generators.hs
PLT-3313: Always use `SemiArbitrary`

@@ -1,75 +1,42 @@
 {-# LANGUAGE BlockArguments    #-}
 {-# LANGUAGE DeriveAnyClass    #-}
 {-# LANGUAGE FlexibleInstances #-}
-{-# LANGUAGE LambdaCase        #-}
 {-# LANGUAGE OverloadedStrings #-}
+{-# LANGUAGE RecordWildCards #-}
 
 {-# OPTIONS_GHC -fno-warn-orphans #-}
 
 module Marlowe.Spec.Core.Generators
-  ( genToken
-  , genParty
-  , genPayee
-  , genChoiceId
-  , genValue
-  , genObservation
-  , genPayment
-  , genTransaction
-  , genInput
-  , genAction
-  , genState
-  , genContract
-  , genTransactionOutput
-  , genTransactionWarning
+  ( gen
+  , genContext
   , RandomResponse (..)
   )
   where
 
 import Control.Applicative ((<|>))
 import Control.Exception (Exception, throwIO)
-import Control.Monad (liftM2)
+import Control.Monad (replicateM)
 import Control.Monad.IO.Class (liftIO)
 import Data.Aeson (FromJSON (..), ToJSON (..), object, withObject, (.:), (.=))
 import Data.Data (Proxy (..))
+import Data.Function (on)
+import Data.List (nubBy)
+import Data.Map (fromList)
 import Marlowe.Spec.Core.Arbitrary
   ( arbitraryChoiceName,
     arbitraryInteger,
-    arbitraryNonnegativeInteger,
+    arbitraryPositiveInteger,
   )
+import Marlowe.Spec.Core.SemiArbitrary (Context (..), SemiArbitrary (..))
 import Marlowe.Spec.Interpret (InterpretJsonRequest, Request (..), parseValidResponse)
 import Marlowe.Spec.TypeId (TypeId (..))
-import QuickCheck.GenT
-  ( GenT,
-    MonadGen (..),
-    frequency,
-    resize,
-    scale,
-    sized,
-    suchThat,
-    vectorOf,
-  )
-import Semantics
-  ( Payment (..),
-    TransactionOutput (..),
-    TransactionOutputRecord_ext (..),
-    TransactionWarning (..),
-    Transaction_ext (..),
-  )
-import SemanticsGuarantees (valid_state)
+import QuickCheck.GenT (GenT, MonadGen (..), elements, frequency, resize, sized)
 import SemanticsTypes
-  ( Action (..),
-    Case (..),
[…]

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/Guarantees.hs

@@ -11,7 +11,8 @@ import Arith (integer_of_nat)
 import qualified Arith
 import Control.Monad.IO.Class (MonadIO (..))
 import qualified Data.Aeson as JSON
-import Marlowe.Spec.Core.Arbitrary (arbitraryTimeIntervalAfter, arbitraryValidTransactions, arbitraryTransaction)
+import Marlowe.Spec.Core.Arbitrary (arbitraryTimeIntervalAfter, arbitraryTransaction, arbitraryValidTransactions)
+import Marlowe.Spec.Core.Generators (genContext)
 import Marlowe.Spec.Core.SemiArbitrary
 import Marlowe.Spec.Interpret
   ( InterpretJsonRequest,

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/Semantics.hs

@@ -17,10 +17,7 @@ import Marlowe.Spec.Core.Arbitrary
   ( arbitraryTransaction,
     arbitraryValidTransactions,
   )
-import Marlowe.Spec.Core.SemiArbitrary
-  ( SemiArbitrary (..),
-    genContext
-  )
+import Marlowe.Spec.Core.SemiArbitrary (SemiArbitrary (..))
 import Marlowe.Spec.Interpret
   ( InterpretJsonRequest,
     Request (..),
@@ -53,6 +50,7 @@ import Test.QuickCheck (Arbitrary (..), withMaxSuccess)
 import Test.QuickCheck.Monadic (PropertyM, assert, monitor, run)
 import Test.QuickCheck.Property (counterexample)
 import Test.Tasty (TestTree, testGroup)
+import Marlowe.Spec.Core.Generators (genContext)
 
 tests :: InterpretJsonRequest -> TestTree
 tests i = testGroup "Semantics"

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/SemiArbitrary.hs

@@ -2,7 +2,6 @@
 {-# LANGUAGE FlexibleInstances #-}
 {-# LANGUAGE LambdaCase        #-}
 {-# LANGUAGE OverloadedStrings #-}
-{-# LANGUAGE RecordWildCards   #-}
 {-# LANGUAGE TupleSections     #-}
 
 {-# OPTIONS_GHC -fno-warn-orphans #-}
@@ -11,33 +10,22 @@ module Marlowe.Spec.Core.SemiArbitrary
   ( arbitraryCaseWeighted
   , arbitraryContractWeighted
   , arbitraryContractSized
-  , Context
-  , genContext
+  , Context (..)
   , SemiArbitrary (..)
   )
   where
 
 import qualified Arith
-import Control.Monad (replicateM)
-import Data.Function (on)
-import Data.List (nubBy)
-import Data.Map (Map, fromList, toList)
+import Data.Map (Map, toList)
 import qualified Examples.Escrow as Escrow
 import qualified Examples.Swap as Swap
 import Marlowe.Spec.Core.Arbitrary
-  ( arbitraryChoiceName,
-    arbitraryInteger,
+  ( arbitraryInteger,
     arbitraryNonnegativeInteger,
     arbitraryPositiveInteger,
   )
-import Marlowe.Spec.Core.Generators (genParty, genToken)
-import Marlowe.Spec.Interpret
-  ( InterpretJsonRequest,
-  )
 import QuickCheck.GenT
   ( Gen,
-    GenT,
-    MonadGen (..),
     frequency,
     resize,
     sized,
@@ -95,62 +83,6 @@ data Context =
   , cchoices     :: Map ChoiceId Arith.Int       -- ^ Choices for state.
   , cboundValues :: Map ValueId Arith.Int        -- ^ Bound values for state.
   } deriving Show
-
-genContext :: InterpretJsonRequest -> GenT IO Context
-genContext interpret = sized \n ->
-  let listOf gen = choose (1, n) >>= flip replicateM gen
-      listOfGen = liftGen . listOf
-      genParty' = genParty interpret
-      genToken' = genToken interpret
-   in do
-        parties <- listOf genParty'
-        tokens <- listOf genToken'
-        amounts <- listOfGen arbitraryPositiveInteger
-        choiceNames <- listOfGen arbitraryChoiceName
-        chosenNums <- listOfGen arbitraryInteger
-        choiceIds <-
-          listOf $
-            ChoiceId
[…]

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/Serialization/Json.hs

@@ -15,33 +15,28 @@ import qualified Data.Aeson.Types as JSON
 import Data.Proxy (Proxy (..))
 import Data.Text as T
 import GHC.Stack (HasCallStack)
-import Marlowe.Spec.Core.Generators
-  ( genAction,
-    genChoiceId,
-    genContract,
-    genInput,
-    genObservation,
-    genParty,
-    genPayee,
-    genPayment,
-    genState,
-    genToken,
-    genTransaction,
-    genTransactionOutput,
-    genTransactionWarning,
-    genValue,
+import Marlowe.Spec.Core.Generators (gen)
+import Marlowe.Spec.Interpret
+  ( InterpretJsonRequest,
+    Request (..),
+    Response (..),
+    testResponse,
   )
-import Marlowe.Spec.Interpret (InterpretJsonRequest, Request (..), Response (..), testResponse)
-import Marlowe.Spec.Reproducible (assertResponse, generate, generateT, reproducibleProperty)
+import Marlowe.Spec.Reproducible
+  ( assertResponse,
+    generate,
+    generateT,
+    reproducibleProperty,
+  )
 import Marlowe.Spec.TypeId (HasTypeId (..), TypeId (..))
 import MarloweCoreJson
-import QuickCheck.GenT (MonadGen (resize))
+import QuickCheck.GenT (GenT, MonadGen (resize))
 import qualified Semantics as C
 import qualified SemanticsTypes as C
+import Test.QuickCheck (Arbitrary (..), Gen)
 import Test.QuickCheck.Monadic (PropertyM, run)
 import Test.Tasty (TestTree, testGroup)
 import Test.Tasty.HUnit (Assertion, assertBool, testCase, (@=?))
-import Test.QuickCheck (Gen, Arbitrary (..))
 
 data SerializationResponse transport
   = SerializationSuccess transport
@@ -182,23 +177,23 @@ isUnknownType _ = False
 arbitraryTokenTest :: InterpretJsonRequest -> TestTree
 arbitraryTokenTest i = reproducibleProperty "Token" do
   -- Any token that is randomly generated by the interpreter should also pass the roundtrip test
-  token <- run $ generateT $ genToken i
+  token <- run $ generateT (gen i :: GenT IO C.Token)
   propertyRoundtripTest i token
 
 arbitraryPartyTest :: InterpretJsonRequest -> TestTree
 arbitraryPartyTest i = reproducibleProperty "Party" do
[…]

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/Guarantees.hs
PLT-3313: Organized imports

@@ -11,9 +11,16 @@ import Arith (integer_of_nat)
 import qualified Arith
 import Control.Monad.IO.Class (MonadIO (..))
 import qualified Data.Aeson as JSON
-import Marlowe.Spec.Core.Arbitrary (arbitraryTimeIntervalAfter, arbitraryTransaction, arbitraryValidTransactions)
+import Marlowe.Spec.Core.Arbitrary
+  ( arbitraryTimeIntervalAfter,
+    arbitraryTransaction,
+    arbitraryValidTransactions,
+  )
 import Marlowe.Spec.Core.Generators (genContext)
 import Marlowe.Spec.Core.SemiArbitrary
+  ( Context (..),
+    SemiArbitrary (..),
+  )
 import Marlowe.Spec.Interpret
   ( InterpretJsonRequest,
     Request (..),

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/Arbitrary.hs
PLT-3313: Dropping not needed `Arbitrary` instances

@@ -6,26 +6,30 @@
 {-# OPTIONS_GHC -fno-warn-orphans #-}
 
 module Marlowe.Spec.Core.Arbitrary
-  ( arbitraryChoiceName
-  , arbitraryFibonacci
-  , arbitraryInteger
-  , arbitraryNonnegativeInteger
-  , arbitraryPositiveInteger
-  , arbitraryTimeInterval
-  , arbitraryTimeIntervalAfter
-  , arbitraryTimeIntervalBefore
-  , arbitraryTimeIntervalAround
-  , arbitraryTransaction
-  , arbitraryValidTransactions
-  , chooseArithInt
+  ( arbitraryChoiceName,
+    arbitraryFibonacci,
+    arbitraryInteger,
+    arbitraryNonnegativeInteger,
+    arbitraryPositiveInteger,
+    arbitraryTimeInterval,
+    arbitraryTimeIntervalAfter,
+    arbitraryTimeIntervalBefore,
+    arbitraryTimeIntervalAround,
+    arbitraryTransaction,
+    arbitraryValidTransactions,
+    chooseArithInt,
   )
-  where
+where
 
 import qualified Arith
 import Data.List (nub)
 import Marlowe.Spec.TypeId ()
 import Orderings (Ord (..), max)
-import QuickCheck.GenT (MonadGen (..), frequency, resize, sized, suchThat)
+import QuickCheck.GenT
+  ( MonadGen (..),
+    frequency,
+    suchThat,
+  )
 import Semantics
   ( TransactionError (..),
     TransactionOutput (..),
@@ -38,17 +42,13 @@ import SemanticsTypes
   ( Action (..),
     Bound (..),
     Case (..),
-    ChoiceId (..),
     Contract (..),
     Environment_ext (..),
     Input (..),
     IntervalError (..),
-    Observation (..),
     Party (..),
-    Payee (..),
     State_ext (..),
     Token (..),
-    Value (..),
     ValueId (..),
     minTime,
   )
@@ -142,6 +142,61 @@ arbitraryTimeIntervalAfter lower =
 greater_eq :: Orderings.Ord a => a -> a -> Bool
 greater_eq = flip less_eq
 
+-- | Generate a random step for a contract.
+arbitraryTransaction :: State_ext ()             -- ^ The state of the contract.
+                     -> Contract                 -- ^ The contract.
+                     -> Gen (Transaction_ext ()) -- ^ Generator for a transaction input for a single step.
[…]

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/SemiArbitrary.hs

@@ -101,6 +101,12 @@ instance Prelude.Ord ChoiceId where
 instance Prelude.Ord ValueId where
   compare (ValueId v1) (ValueId v2) = compare v1 v2
 
+instance Prelude.Ord Party where
+  compare (Address a) (Address b) = compare a b
+  compare (Address _) (Role _)    = LT
+  compare (Role _) (Address _)    = GT
+  compare (Role a) (Role b)       = compare a b
+
 instance SemiArbitrary Arith.Int where
   fromContext = times
 
@@ -425,9 +431,3 @@ arbitraryContractSized :: Int           -- ^ The maximum depth.
                        -> Context       -- ^ The Marlowe context.
                        -> Gen Contract  -- ^ Generator for a contract.
 arbitraryContractSized = arbitraryContractWeighted . (`replicate` defaultContractWeights)
-
-instance Prelude.Ord Party where
-  compare (Address a) (Address b) = compare a b
-  compare (Address _) (Role _)    = LT
-  compare (Role _) (Address _)    = GT
-  compare (Role a) (Role b)       = compare a b

### 2023-04-19 marlowe-spec-test/src/Marlowe/Spec/Core/Arbitrary.hs
PLT-3313: small refactoring

@@ -239,40 +239,12 @@ shrinkTimeInterval (start, end) =
       , (end  , end  )
       ]
 
-genValueId :: Gen ValueId
-genValueId = arbitraryFibonacci randomValueIds
-
-genBound :: Gen Bound
-genBound = do
-  lower <- arbitraryInteger
-  extent <- arbitraryNonnegativeInteger
-  pure $ Bound lower (lower + extent)
-
+-- | Generate a time interval.
 genInterval :: Gen (Arith.Int, Arith.Int)
 genInterval = do
   lower <- arbitraryNonnegativeInteger
   extent <- arbitraryNonnegativeInteger
   pure (lower, lower + extent)
-
-genIntervalError :: Gen IntervalError
-genIntervalError = do
-  lower <- arbitraryInteger
-  extent <- arbitraryNonnegativeInteger
-  frequency
-    [ (1, pure $ InvalidInterval (lower, lower + extent))
-    , (1, IntervalInPastError <$> liftGen arbitraryNonnegativeInteger <*> pure  (lower, lower + extent) )
-    ]
-
-genTransactionError :: Gen TransactionError
-genTransactionError = frequency
-    [ (1, pure TEAmbiguousTimeIntervalError)
-    , (1, pure TEApplyNoMatchError)
-    , (1, TEIntervalError <$> genIntervalError)
-    , (1, pure TEUselessTransaction)
-    ]
-
-genEnvironment :: Gen (Environment_ext ())
-genEnvironment = Environment_ext <$> genInterval <*> pure ()
 
 -- | Some value identifiers.
 randomValueIds :: [ValueId]
@@ -300,11 +272,15 @@ instance Arbitrary Arith.Int where
   arbitrary = arbitraryInteger
 
 instance Arbitrary ValueId where
-  arbitrary = genValueId
+  arbitrary = arbitraryFibonacci randomValueIds
   shrink = shrinkString (\(ValueId x) -> x) randomValueIds
 
 instance Arbitrary Bound where
-  arbitrary = genBound
+  arbitrary = do
+    lower <- arbitraryInteger
+    extent <- arbitraryNonnegativeInteger
+    pure $ Bound lower (lower + extent)
+
   shrink (Bound lower upper) =
     let
       mid = (lower + upper) `Arith.divide_int` 2
@@ -363,15 +339,28 @@ randomRoleNames =
   ]
 
 instance Arbitrary TransactionError where
[…]

### 2023-04-24 isabelle/generated/ListTools.hs
Remove old generated file

@@ -1,16 +1,0 @@
-{-# LANGUAGE EmptyDataDecls, RankNTypes, ScopedTypeVariables #-}
-
-module ListTools(anya) where {
-
-import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
-  (>>=), (>>), (=<<), (&&), (||), (^), (^^), (.), ($), ($!), (++), (!!), Eq,
-  error, id, return, not, fst, snd, map, filter, concat, concatMap, reverse,
-  zip, null, takeWhile, dropWhile, all, any, Integer, negate, abs, divMod,
-  String, Bool(True, False), Maybe(Nothing, Just));
-import qualified Prelude;
-
-anya :: forall a. (a -> Bool) -> [a] -> Bool;
-anya f [] = False;
-anya f (h : t) = f h || anya f t;
-
-}

### 2023-04-24 response-to-audit-report.md
Add Isabelle responses

@@
 - [x] Add prefatory text and executive summary.
-- [ ] Add responses and mitigation commits to Isabelle-related findings.
+- [x] Add responses and mitigation commits to Isabelle-related findings.
 - [x] Add responses and mitigation commits to Haskell-related findings.
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here.
+The audit report's main and high-severity concerns for the Isabelle Proofs were:
+
+- 2.1.6 Missing description of Merkleization
+- 2.1.4 Inaccurate formulation of Money preservation
+
+We didn't mention the merkleization in the specification document as we thought of it as an implementation detail.
+This does lessen the applicability of the Isabelle proofs to the different implementations (marlowe-cardano, purescript-marlowe, etc).
+To check that an implementation is compliant to the specification we use the `spec-test` tool, which uses property based tests to check that a custom implementation behaves equally to the Isabelle implementation.
+Without the proper formalization, the tool can only test for the non-merkleized cases.
+In practice this gives us confidence that the merkleized interpreter works the same for non-merkleized cases.
+To test the correctness of merkleized cases, the marlowe-cardano repository has specific tests.
+The Merkleization formalization will be introduced in a future version of the specification, the initial formulation started in [PR-171](https://github.com/input-output-hk/marlowe/pull/171).
+
+The Money preservation theory was only taking into account the amount of assets and not the type.
+This meant that the proof wasn't considering a case were a contract could receive `20 ada`, `15 djed` and pay back `20 djed` and `15 ada`.
+We added a new MultiAssets type and refactored the Isabelle code (without modifying the interpreter) to prove that the Assets are preserved.
+
+Other medium and low priorities findings regarding documentation and legibility of the specification were also addressed.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+To address this, a new `MultiAssets.thy` theory was created.
+This theory defines a new type that ties together the number of tokens and the type of token, and it implements some type classes to be able to work with normal addition and substraction.
+Also the MoneyPreservation was refactored into AssetsPreservation, addressing issue 2.1.4 among others.
+
+These changes can be found on [PR-161](https://github.com/input-output-hk/marlowe/pull/161)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The theory was completly refactored in [PR-161](https://github.com/input-output-hk/marlowe/pull/161), adding the necessary documentation.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We didn't mention the merkleization in the specification document as we thought of it as an implementation detail.
+This does lessen the applicability of the Isabelle proofs to the different implementations (marlowe-cardano, purescript-marlowe, etc).
+To check that an implementation is compliant to the specification we use the `spec-test` tool, which uses property based tests to check that a custom implementation behaves equally to the Isabelle implementation.
+Without the proper formalization, the tool can only test for the non-merkleized cases.
+In practice this gives us confidence that the merkleized interpreter works the same for non-merkleized cases.
+To test the correctness of merkleized cases, the marlowe-cardano repository has specific tests.
+
+The Merkleization formalization will be introduced in a future version of the specification, the initial formulation started in [PR-171](https://github.com/input-output-hk/marlowe/pull/171).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This module was refactored in [PR-161](https://github.com/input-output-hk/marlowe/pull/161) and these functions were no longer needed.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We improved documentation regarding choices in [PR-168](https://github.com/input-output-hk/marlowe/pull/168) and [PR-182](https://github.com/input-output-hk/marlowe/pull/182)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We fixed this in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We added more documentation regarding the environment in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We added more documentation regarding the execution environment in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+These types were properly documented in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This type was properly documented in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The cause of this problem was that initially we had the Isabelle code separated from the documentation generation and we coulnd't find a way to display the Isabelle code, so we showed the exported Haskell instead.
+We agree that it is confusing so we refactored to a literate programming style in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We removed the extra definition in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We renamed as suggested in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This function is confusing, we are likely to remove it in a future release.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We disagree on this observation. The function `updateMoneyInAccount` checks if its money parameter is non positive to always have accounts with positive balance. The
+`addMoneyToAccount` function calls `updateMoneyInAccount` with `balance + money` and checks that `money` is positive. This makes sure that `addMoneyToAccount` always increases the assets.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The statement was removed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The way we document evalValue was modified in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This lemma was removed from the PDF specification and replaced with lemmas that uses `evalValue` instead.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This was corrected to use `integer` instead of `natural` and further rephrased in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This section was rewritten an only the Division lemmas are now present as they are the most relevant ones.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This section was rewritten and the issue was fixed
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The function `playTrace` was documented in [PR-182](https://github.com/input-output-hk/marlowe/pull/182) in section 2.2.7.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This happened because we were using a type alias, which gets fully expanded when being referenced by the documentation.
+[PR-182](https://github.com/input-output-hk/marlowe/pull/182) address this by making the whole theory in literal programming mode.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This was taken into account in the new Asset preservation theory and a [paragraph](https://github.com/input-output-hk/marlowe/pull/161/files#diff-060b0c9f7bde4e9ebc4e7b9d6a2202af75213be8d44d7f4745e02054be5186deR52-R53) was added to explain the case.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The proposed definition is simpler and it is likely to be implemented in the future.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The definition in the specification is the exported haskell code of the Isabelle code.
+This is a limitation that Isabelle has when importing code from other theories.
+A possible solution is to change the Accounts positive theory to be literal programming.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The statement was indeed unconnected from the description.
+In [PR-186](https://github.com/input-output-hk/marlowe/pull/186) we refactored the guarantees section and now the section "3.2 Finite contracts"
+includes the theorem `timedOutTransaction_closes_contract` which shows that after closing a timeouted contract, the accounts are emptied.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This was addressed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+In [PR-186](https://github.com/input-output-hk/marlowe/pull/186) we add a small explanation, in the future we'll probably expand on this by refactoring the guarantees as literate programming.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This section was redone in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+In [PR-186](https://github.com/input-output-hk/marlowe/pull/186) we added the names of the theories in the PDF.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We agree with this observation.
+New lemmas and theories are proven with this recommendation in mind and some theories were refactored (like the Asset preservation).
+It will take some time until all theories are in this format.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+Same response as 2.3.1
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We agree that the proposed prove is easier, but the lemma will become obsolete once we replace the custom MList with
+`HOL-Library/finiteMap`, which addresses other problems found in the report.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The MoneyPreservation was refactored into Asset preservation and now there is a single reference to `giveMoney`. As mentioned in observation 2.2.11, giveMoney is likely to be removed in the future.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This theory was refactored in [PR-161](https://github.com/input-output-hk/marlowe/pull/161) and the lemma was no longer needed.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The valid and positive preservation is needed in lower-level functions so that higher-level functions can use that fact when proving other properties.
+This makes `playTrace` not needing that proof as it is the highest-level function.
+The theorem  could be useful on its own and might get added in the future.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We agree on this observation and it will addressed in a future refactor of the theory file.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We agree on these observations and they will likely be addressed when we refactor the corresponding files.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+* Lemma `transferMoneyBetweenAccounts_preserves_aux` was removed in [PR-161](https://github.com/input-output-hk/marlowe/pull/161)
+* Lemma `reduceOne_onlyIfNonEmptyState`, `valid_state_valid_choices` were removed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
+* Lemma `reduceContractUntilQuiescent_ifDifferentReduced` was promoted to a theorem with an explanation in  [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
+* Lemma `positiveMoneyInAccountOrNoAccount_gtZero_preservation` was removed in favour of `positiveMoneyInAccountOrNoAccount_gtZero_preservation` in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This lemma was removed in [PR-161](https://github.com/input-output-hk/marlowe/pull/161)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+Lemma `accountsArePositive` was removed in favour of `computeTransaction_gtZero` in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+Lemma `computeTransaction_preserves_valid_state_aux` was removed and `computeTransaction_preserve_valid_state` simplified in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+These issues were resolved in [PR-168](https://github.com/input-output-hk/marlowe/pull/168) and [PR-169](https://github.com/input-output-hk/marlowe/pull/169)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+These observation regarding readability will be addressed when we refactor the theory.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This issue was resolved in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This observation regarding readability will be addressed when we refactor the theory.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We agree on the observation, will likely promote to theorem in the future
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The lemma was promoted to a theorem in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
 
@@
 
+We removed the unnecessary assumptions in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
+
 ## 2.4 Isabelle implementation
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
-
 ### 2.4.1 Variable shadowing in `applyAllLoop` *(Severity: Medium)*
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The variable was renamed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The theory was refactored into AssetsPreservation and the function is now called `assetsInExternalPayment`
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The typo was fixed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We agree that it is confusing and will be refactored in the future.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This was an artifact of showing the exported haskell code. It is resolved in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This function was documented in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
+and moved to `TimeRange.thy` in [PR-182](https://github.com/input-output-hk/marlowe/pull/182).
+The function is not part of the Marlowe spec yet, it is a helper we hope will be useful in the future to construct the transactions.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We agree with the observation and it will be addressed in a future refactor.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We agree on the observations, for the moment some simple refactor and comments were added as part of
+[PR-180](https://github.com/input-output-hk/marlowe/pull/180/files#diff-638dc578b752590b5f2bace45f1e1ae9e522351f62056e6f92aa272424657b03).
+Further refactor and optional renaming might happen in the future.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The function `isValidInterval` was removed in favour of `validTimeInterval` in [PR-168](https://github.com/input-output-hk/marlowe/pull/168/files#diff-9c936a78b5d8eed5683713bb31c59366b2a0794c5f11e9807f7fa61d1b11b7c2L230)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here.
+This was addressed in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This misstatement is remedied in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)
 

### 2023-04-27 isabelle/generated/Examples/Swap.hs
Update to isabelle 2022

@@ -14,7 +14,7 @@ import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
   String, Bool(True, False), Maybe(Nothing, Just));
 import qualified Prelude;
 import qualified Semantics;
-import qualified Stringa;
+import qualified Str;
 import qualified SemanticsTypes;
 import qualified Arith;
 
@@ -59,62 +59,62 @@ swap p1 p2 =
       (makePayment p1 p2 (makePayment p2 p1 SemanticsTypes.Close)));
 
 adaToken :: SemanticsTypes.Token;
-adaToken = SemanticsTypes.Token (Stringa.implode []) (Stringa.implode []);
+adaToken = SemanticsTypes.Token (Str.implode []) (Str.implode []);
 
 adaProvider :: SemanticsTypes.Party;
 adaProvider =
   SemanticsTypes.Role
-    (Stringa.implode
-      [Stringa.Char True False False False False False True False,
-        Stringa.Char False False True False False True True False,
-        Stringa.Char True False False False False True True False,
-        Stringa.Char False False False False False True False False,
-        Stringa.Char False False False False True False True False,
-        Stringa.Char False True False False True True True False,
-        Stringa.Char True True True True False True True False,
-        Stringa.Char False True True False True True True False,
-        Stringa.Char True False False True False True True False,
-        Stringa.Char False False True False False True True False,
-        Stringa.Char True False True False False True True False,
-        Stringa.Char False True False False True True True False]);
+    (Str.implode
+      [Str.Char True False False False False False True False,
+        Str.Char False False True False False True True False,
+        Str.Char True False False False False True True False,
+        Str.Char False False False False False True False False,
+        Str.Char False False False False True False True False,
+        Str.Char False True False False True True True False,
+        Str.Char True True True True False True True False,
+        Str.Char False True True False True True True False,
+        Str.Char True False False True False True True False,
+        Str.Char False False True False False True True False,
+        Str.Char True False True False False True True False,
+        Str.Char False True False False True True True False]);
 
 dollarToken :: SemanticsTypes.Token;
 dollarToken =
   SemanticsTypes.Token
-    (Stringa.implode
-      [Stringa.Char False False False True True True False False,
-        Stringa.Char True False True False True True False False,
-        Stringa.Char False True False False False True True False,
-        Stringa.Char False True False False False True True False,
-        Stringa.Char False True True False True True False False,
[…]

### 2023-04-27 isabelle/generated/Str.hs

@@ -1,6 +1,6 @@
 {-# LANGUAGE EmptyDataDecls, RankNTypes, ScopedTypeVariables #-}
 
-module Stringa(Char(..), implode) where {
+module Str(Char(..), implode) where {
 
 import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
   (>>=), (>>), (=<<), (&&), (||), (^), (^^), (.), ($), ($!), (++), (!!), Eq,

### 2023-04-27 isabelle/generated/Examples/Swap.hs
Fix imports after merge

@@ -13,7 +13,6 @@ import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
   zip, null, takeWhile, dropWhile, all, any, Integer, negate, abs, divMod,
   String, Bool(True, False), Maybe(Nothing, Just));
 import qualified Prelude;
-import qualified Semantics;
 import qualified Stringa;
 import qualified SemanticsTypes;
 import qualified Arith;
@@ -125,40 +124,40 @@ swapExample =
       (SemanticsTypes.Constant (Arith.Int_of_integer (20 :: Integer)))
       dollarToken (Arith.Int_of_integer (1664816400000 :: Integer)) ());
 
-partialExecutionPathPayments :: [Semantics.Payment];
+partialExecutionPathPayments :: [SemanticsTypes.Payment];
 partialExecutionPathPayments =
-  [Semantics.Payment adaProvider (SemanticsTypes.Party adaProvider) adaToken
-     (Arith.Int_of_integer (10 :: Integer))];
+  [SemanticsTypes.Payment adaProvider (SemanticsTypes.Party adaProvider)
+     adaToken (Arith.Int_of_integer (10 :: Integer))];
 
-successfulExecutionPathPayments :: [Semantics.Payment];
+successfulExecutionPathPayments :: [SemanticsTypes.Payment];
 successfulExecutionPathPayments =
-  [Semantics.Payment adaProvider (SemanticsTypes.Party dollarProvider) adaToken
-     (Arith.Int_of_integer (10 :: Integer)),
-    Semantics.Payment dollarProvider (SemanticsTypes.Party adaProvider)
+  [SemanticsTypes.Payment adaProvider (SemanticsTypes.Party dollarProvider)
+     adaToken (Arith.Int_of_integer (10 :: Integer)),
+    SemanticsTypes.Payment dollarProvider (SemanticsTypes.Party adaProvider)
       dollarToken (Arith.Int_of_integer (20 :: Integer))];
 
-partialExecutionPathTransactions :: [Semantics.Transaction_ext ()];
+partialExecutionPathTransactions :: [SemanticsTypes.Transaction_ext ()];
 partialExecutionPathTransactions =
-  [Semantics.Transaction_ext
+  [SemanticsTypes.Transaction_ext
      (Arith.Int_of_integer (1664812600000 :: Integer),
        Arith.Int_of_integer (1664812700000 :: Integer))
[…]

### 2023-04-27 isabelle/generated/SingleInputTransactions.hs

@@ -8,23 +8,22 @@ import Prelude ((==), (/=), (<), (<=), (>=), (>), (+), (-), (*), (/), (**),
   zip, null, takeWhile, dropWhile, all, any, Integer, negate, abs, divMod,
   String, Bool(True, False), Maybe(Nothing, Just));
 import qualified Prelude;
-import qualified Semantics;
 import qualified SemanticsTypes;
 import qualified Arith;
 
 inputsToTransactions ::
   (Arith.Int, Arith.Int) ->
-    [SemanticsTypes.Input] -> [Semantics.Transaction_ext ()];
-inputsToTransactions ti [] = [Semantics.Transaction_ext ti [] ()];
-inputsToTransactions ti [inp1] = [Semantics.Transaction_ext ti [inp1] ()];
+    [SemanticsTypes.Input] -> [SemanticsTypes.Transaction_ext ()];
+inputsToTransactions ti [] = [SemanticsTypes.Transaction_ext ti [] ()];
+inputsToTransactions ti [inp1] = [SemanticsTypes.Transaction_ext ti [inp1] ()];
 inputsToTransactions ti (headInput : v : va) =
-  Semantics.Transaction_ext ti [headInput] () :
+  SemanticsTypes.Transaction_ext ti [headInput] () :
     inputsToTransactions ti (v : va);
 
 traceListToSingleInput ::
-  [Semantics.Transaction_ext ()] -> [Semantics.Transaction_ext ()];
+  [SemanticsTypes.Transaction_ext ()] -> [SemanticsTypes.Transaction_ext ()];
 traceListToSingleInput [] = [];
-traceListToSingleInput (Semantics.Transaction_ext si inps () : tailTx) =
+traceListToSingleInput (SemanticsTypes.Transaction_ext si inps () : tailTx) =
   inputsToTransactions si inps ++ traceListToSingleInput tailTx;
 
 }

### 2023-04-27 marlowe-spec-test/src/Marlowe/Spec/Core/Arbitrary.hs

@@ -25,11 +25,7 @@ import Data.List (nub)
 import Marlowe.Spec.TypeId ()
 import Orderings (Ord (..), max)
 import Semantics
-  ( TransactionError (..),
-    TransactionOutput (..),
-    TransactionOutputRecord_ext (..),
-    Transaction_ext (..),
-    computeTransaction,
+  ( computeTransaction,
     evalValue,
   )
 import SemanticsTypes
@@ -42,6 +38,10 @@ import SemanticsTypes
     IntervalError (..),
     State_ext (..),
     ValueId (..),
+    TransactionError (..),
+    TransactionOutput (..),
+    TransactionOutputRecord_ext (..),
+    Transaction_ext (..),
     minTime,
   )
 import Test.QuickCheck (Gen, frequency, suchThat, oneof)

### 2023-04-27 marlowe-spec-test/src/Marlowe/Spec/Core/Semantics.hs

@@ -34,10 +34,7 @@ import Marlowe.Utils (showAsJson)
 import Orderings (Ord (..))
 import QuickCheck.GenT (suchThat)
 import Semantics
-  ( TransactionOutput (..),
-    TransactionOutputRecord_ext (..),
-    Transaction_ext (..),
-    computeTransaction,
+  ( computeTransaction,
     evalObservation,
     evalValue,
   )
@@ -45,6 +42,9 @@ import SemanticsTypes
   ( Contract (..),
     State_ext (..),
     Value (..),
+    TransactionOutput (..),
+    TransactionOutputRecord_ext (..),
+    Transaction_ext (..),
   )
 import Test.QuickCheck (Arbitrary (..), withMaxSuccess)
 import Test.QuickCheck.Monadic (PropertyM, assert, monitor, run)

### 2023-04-27 marlowe-spec-test/src/Marlowe/Spec/Core/SemiArbitrary.hs

@@ -32,13 +32,6 @@ import QuickCheck.GenT
     sized,
     vectorOf,
   )
-import Semantics
-  ( Payment (..),
-    TransactionOutput (..),
-    TransactionOutputRecord_ext (..),
-    TransactionWarning (..),
-    Transaction_ext (..),
-  )
 import SemanticsTypes
   ( Action (..),
     Bound (..),
@@ -53,6 +46,11 @@ import SemanticsTypes
     Token (..),
     Value (..),
     ValueId (..),
+    Payment (..),
+    TransactionOutput (..),
+    TransactionOutputRecord_ext (..),
+    TransactionWarning (..),
+    Transaction_ext (..),
   )
 import Test.QuickCheck (chooseInt, elements)
 import Test.QuickCheck.Arbitrary (Arbitrary (..))

### 2023-05-08 response-to-audit-report.md
Apply suggestions from code review
Co-authored-by: Pablo Lamela  <pablo.lamela@iohk.io>

@@
 This does lessen the applicability of the Isabelle proofs to the different implementations (marlowe-cardano, purescript-marlowe, etc).
-To check that an implementation is compliant to the specification we use the `spec-test` tool, which uses property based tests to check that a custom implementation behaves equally to the Isabelle implementation.
+To check that an implementation is compliant with the specification we use the `spec-test` tool, which uses property based tests to check that a custom implementation behaves equally to the Isabelle implementation.
 Without the proper formalization, the tool can only test for the non-merkleized cases.
@@
 This theory defines a new type that ties together the number of tokens and the type of token, and it implements some type classes to be able to work with normal addition and substraction.
-Also the MoneyPreservation was refactored into AssetsPreservation, addressing issue 2.1.4 among others.
-
-These changes can be found on [PR-161](https://github.com/input-output-hk/marlowe/pull/161)
+Also, the MoneyPreservation was refactored into AssetsPreservation, addressing issue 2.1.4 among others.
+
+These changes can be found in [PR-161](https://github.com/input-output-hk/marlowe/pull/161)
 
@@
 
-The cause of this problem was that initially we had the Isabelle code separated from the documentation generation and we coulnd't find a way to display the Isabelle code, so we showed the exported Haskell instead.
+The cause of this problem was that initially we had the Isabelle code separated from the documentation generation and we couldn't find a way to display the Isabelle code, so we showed the exported Haskell instead.
 We agree that it is confusing so we refactored to a literate programming style in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)
@@
 
-This lemma was removed from the PDF specification and replaced with lemmas that uses `evalValue` instead.
+This lemma was removed from the PDF specification and replaced with lemmas that use `evalValue` instead.
 
@@
 
-This section was rewritten an only the Division lemmas are now present as they are the most relevant ones.
+This section was rewritten and only the Division lemmas are now present as they are the most relevant ones.
 
@@
 
-This section was rewritten and the issue was fixed
+This section was [rewritten](https://github.com/input-output-hk/marlowe/pull/182) and the issue was fixed
 
@@
 This happened because we were using a type alias, which gets fully expanded when being referenced by the documentation.
-[PR-182](https://github.com/input-output-hk/marlowe/pull/182) address this by making the whole theory in literal programming mode.
+[PR-182](https://github.com/input-output-hk/marlowe/pull/182) addresses this by making the whole theory in literal programming mode.
 
@@
 
-The definition in the specification is the exported haskell code of the Isabelle code.
+The definition in the specification is the exported Haskell code of the Isabelle code.
 This is a limitation that Isabelle has when importing code from other theories.
@@
 
-In [PR-186](https://github.com/input-output-hk/marlowe/pull/186) we add a small explanation, in the future we'll probably expand on this by refactoring the guarantees as literate programming.
+In [PR-186](https://github.com/input-output-hk/marlowe/pull/186) we added a small explanation; in the future, we'll probably expand on this by refactoring the guarantees as literate programming.
 
@@
 We agree with this observation.
-New lemmas and theories are proven with this recommendation in mind and some theories were refactored (like the Asset preservation).
+New lemmas and theories are proven with this recommendation in mind, and some theories were refactored (like the Asset preservation).
 It will take some time until all theories are in this format.
@@
 
-We agree that the proposed prove is easier, but the lemma will become obsolete once we replace the custom MList with
+We agree that the proposed proof is easier, but the lemma will become obsolete once we replace the custom MList with
 `HOL-Library/finiteMap`, which addresses other problems found in the report.
@@
 
-We agree on this observation and it will addressed in a future refactor of the theory file.
+We agree on this observation and it will be addressed in a future refactor of the theory file.
 
@@
 
-These observation regarding readability will be addressed when we refactor the theory.
+These observations regarding readability will be addressed when we refactor the theory.
 
@@
 
-We agree on the observation, will likely promote to theorem in the future
+We agree on the observation, and will likely promote it to theorem in the future.
 
@@
 
-We removed the unnecessary assumptions in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
+We removed the unnecessary assumptions in [PR-168](https://github.com/input-output-hk/marlowe/pull/168).
 
@@
 
-The variable was renamed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)
+The variable was renamed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168).
 
@@
 
-The theory was refactored into AssetsPreservation and the function is now called `assetsInExternalPayment`
+The theory was refactored into AssetsPreservation and the function is now called `assetsInExternalPayment`.
 
@@
 
-We agree that it is confusing and will be refactored in the future.
+We agree that it is confusing, and it will be refactored in the future.
 
@@
 
-We agree on the observations, for the moment some simple refactor and comments were added as part of
+We agree on the observations, for the moment some simple refactoring and comments were added as part of
 [PR-180](https://github.com/input-output-hk/marlowe/pull/180/files#diff-638dc578b752590b5f2bace45f1e1ae9e522351f62056e6f92aa272424657b03).
@@
 
-We agree on this observation and it will addressed in a future refactor of the theory file.
+We agree on this observation and it will addressed in a future refactoring of the theory file.
 

### 2023-05-09 response-to-audit-report.md
PLT-5502 Proofread and edited Isabelle-related findings.

--- 2023-04-19 response-to-audit-report.md
@@
-# Response to Audit Report
-
-**TODO:**
-
-- [x] Add prefatory text and executive summary.
-- [ ] Add responses and mitigation commits to Isabelle-related findings.
-- [x] Add responses and mitigation commits to Haskell-related findings.
-- [ ] Fix display of equations or replace ones that cannot be fixed with ellipses ("...").
-- [ ] Replace broken links and cross references, or use ellipses ("...").
-- [ ] Verify the section numbering is correct.
-- [ ] Add table of contents.
-- [ ] Proofread.
+# Response to Audit Report for the Marlowe Specification and for Marlowe on Cardano
 
@@
 
-The high-severity findings related to handling of negative deposits, prevention of "double satisfaction", enforcement of state invariants, an implementation difference between the formal specification and the Plutus implementation, and the Isabelle proof of money preservation.
+The high-severity findings related to handling of negative deposits, prevention of "double satisfaction", enforcement of state invariants, an implementation difference between the formal specification versus the Plutus implementation, and the Isabelle proof of money preservation.
 Marlowe's Plutus validator has been altered to resolve these first three findings so that it correctly prevents negative deposits from altering the balance of internal accounts, it only allows other Plutus scripts to run during transactions that do not make Marlowe payments, and it rigorously enforces invariants for initial and final states.
-The fourth finding regarding the Plutus implementation was mitigated to code analysis and property-based testing.
-The fifth finding about the money-preservation proof was remedied by revising the proof.
+The fourth finding regarding the Plutus implementation was mitigated via code analysis and property-based testing.
+The fifth finding about the money-preservation proof was remedied by revising the Isabelle proof.
 
@@
 
-Finally, please note that the audit's scope included the Marlowe language and its Plutus interpreter, but not individual Marlowe contracts.
+Finally, please note that the audit's scope included the Marlowe language and its Plutus interpreter, but not individual Marlowe contracts or off-chain code.
 Although the Marlowe language and its implementation on Cardano aim to make smart contracts safer, it is nevertheless still possible to design individual instances of Marlowe contracts that exhibit unwanted behavior.
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here.
+The audit report's main and high-severity concerns for the Isabelle Proofs were:
+
+- 2.1.6 Missing description of Merkleization
+- 2.1.4 Inaccurate formulation of Money preservation
+
+We did not mention the merkleization in the specification document as it was considered an implementation detail.
+This does lessen the applicability of the Isabelle proofs to the different implementations (Marlowe on Cardano, Marlowe in PureScript, etc).
+To check that an implementation is compliant with the specification we use the `spec-test` tool, which uses property based tests to check that a custom implementation behaves equally to the Isabelle implementation.
+Without the proper formalization, the tool can only test for the non-merkleized cases.
+In practice this gives us confidence that the merkleized interpreter works the same for non-merkleized cases.
+To test the correctness of merkleized cases, the marlowe-cardano repository has specific tests.
+The Merkleization formalization will be introduced in a future version of the specification, the initial formulation started in [PR-171](https://github.com/input-output-hk/marlowe/pull/171).
+
+The Money preservation theory had only taken into account the amount of assets and not the type.
+This meant that the proof was not considering a case where a contract could receive `20 ada`,  `15 djed`, and pay back `20 djed` and `15 ada`.
+We added a new `MultiAssets` type and refactored the Isabelle code (without modifying the interpreter) to prove that the Assets are preserved.
+
+Other medium and low priorities findings regarding documentation and legibility of the specification were also addressed.
 
@@
 
-The Marlowe validator had made optimistic assumptions about its own correct operation and hence did not check certain invariants in order to save on Plutus execution costs (item 3 above).
+The Marlowe validator had made optimistic assumptions about its own correct operation and hence did not check certain invariants in order to reduce Plutus execution costs (item 3 above).
 The Marlowe semantics validator has now been thoroughly hardened against corruption of initial or final state so that it ensures that the three state-invariants of positive accounts, non-duplication of state entries (accounts, choices, and bound values), and a total internal value matching the script UTxO's value.
 This hardening was completed without unduly increasing the size (in bytes) of the validator or its Plutus execution cost: such was verified with simulated and on-chain measurement of execution costs for a library of real-life Marlowe contracts.
-Notably, the enforcement of the final invariants protects against funds becoming permanently locked in a Marlowe contract due to a garbled state: even in cases where the implementation of Marlowe semantics or parts of the Plutus validator were flawed, at least some of the execution paths of a Marlowe contract remain viable, so funds could be released even under nominally impossible but catastrophic difficulties.
+Notably, the enforcement of the final invariants protects against funds becoming permanently locked in a Marlowe contract due to a garbled state: even in cases where the implementation of Marlowe semantics or parts of the Plutus validator were flawed, at least some of the execution paths of a Marlowe contract might remain viable, so funds could be released even under nominally impossible but catastrophic difficulties.
 The property-based test suite was significantly enhanced to check double-satisfaction situations.
@@
 The medium-severity concerns in the audit report center around missing tests or name shadowing.
-These have all been remedied either by code changes (in the case of the shadowing) or by the addition of a significant augmentation of the property-based test suite.
+These have all been remedied either by code changes (in the case of the shadowing) or by the implementation of a significant augmentation of the property-based test suite.
 The audit's low-severity concerns generally relate to insufficiently detailed text in the Marlowe Cardano specification, the need for more elaborate comments in the code, naming of variable bindings, or typographical errors.
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+To address this, a new `MultiAssets.thy` Isabelle theory was created.
+This theory defines a new type that ties together the number of tokens and the type of token, and it implements some type classes to be able to work with normal addition and subtraction.
+Also, the `MoneyPreservation` was refactored into `AssetsPreservation`, addressing issue 2.1.4 among others.
+
+These changes can be found in [PR-161](https://github.com/input-output-hk/marlowe/pull/161)
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The theory was completely refactored in [PR-161](https://github.com/input-output-hk/marlowe/pull/161), adding the necessary documentation.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We did not mention the merkleization in the specification document as it was considered an implementation detail.
+This does lessen the applicability of the Isabelle proofs to the different implementations (Marlowe on Cardano, Marlowe on PureScript, etc).
+To check that an implementation is compliant to the specification we use the `spec-test` tool, which uses property based tests to check that a custom implementation behaves equally to the Isabelle implementation.
+Without the proper formalization, the tool can only test for the non-merkleized cases.
+In practice this gives us confidence that the merkleized interpreter works the same for non-merkleized cases.
+To test the correctness of merkleized cases, the marlowe-cardano repository has specific tests.
+
+The Merkleization formalization will be introduced in a future version of the specification, the initial formulation started in [PR-171](https://github.com/input-output-hk/marlowe/pull/171).
 
@@
 Adding a check on the final output's validity now prevents a contract from ever reaching a fully blocked state.
-Even if the semantics or Plutus code were flawed, at least some execution paths in the contract would remain viable, so that the contract could eventually terminate and would not permanently lock funds.
+Even if the semantics or Plutus code were flawed, at least some execution paths in the contract might remain viable, so that the contract could eventually terminate and would not permanently lock funds.
 
@@
 As the audit report recognizes, this concern is closely related to the three previous concerns.
-The revised validator's enforcement that of the three invariants for the final state ensures that the ending balance of the internal accounts in the state matches the actual output paid to the script.
+The revised validator's enforcement of the three invariants for the final state ensures that the ending balance of the internal accounts in the state matches the actual output paid to the script.
 
@@
 
-Subsequent to the commencement of the audit, a "test oracle" (called test spec) was developed that checks implementations of Marlowe semantics (such as the Plutus validator) against a reference implementation generated directly from the Isabelle specification.
+Subsequent to the commencement of the audit, a "test oracle" (called `test-spec`) was developed that checks implementations of Marlowe semantics (such as the Plutus validator) against a reference implementation generated directly from the Isabelle specification.
 That generated Haskell implementation is backed by Isabelle proofs of the correctness of Haskell code generation by Isabelle.
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We improved documentation regarding choices in [PR-168](https://github.com/input-output-hk/marlowe/pull/168) and [PR-182](https://github.com/input-output-hk/marlowe/pull/182).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We fixed this in [PR-182](https://github.com/input-output-hk/marlowe/pull/182).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We added more documentation regarding the environment in [PR-182](https://github.com/input-output-hk/marlowe/pull/182).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We added more documentation regarding the execution environment in [PR-182](https://github.com/input-output-hk/marlowe/pull/182).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+These types were properly documented in [PR-182](https://github.com/input-output-hk/marlowe/pull/182).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This type was properly documented in [PR-182](https://github.com/input-output-hk/marlowe/pull/182).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The cause of this problem was that initially we had the Isabelle code separated from the documentation generation and we could not find a way to display the Isabelle code, so we showed the exported Haskell instead.
+We agree that it is confusing so we refactored to a literate programming style in [PR-182](https://github.com/input-output-hk/marlowe/pull/182).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We removed the extra definition in [PR-168](https://github.com/input-output-hk/marlowe/pull/168).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We renamed as suggested in [PR-182](https://github.com/input-output-hk/marlowe/pull/182).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This function is confusing, we will likely to remove it in a future release.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We disagree with this observation. The function `updateMoneyInAccount` checks if its money parameter is non-positive, so as to always have accounts with positive balance. The
+`addMoneyToAccount` function calls `updateMoneyInAccount` with `balance + money` and checks that `money` is positive. This ensures that `addMoneyToAccount` always increases the assets.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The statement was removed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The way we document `evalValue` was modified in [PR-182](https://github.com/input-output-hk/marlowe/pull/182).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This lemma was removed from the specification and replaced with lemmas that use `evalValue` instead.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This was corrected to use `integer` instead of `natural` and further rephrased in [PR-182](https://github.com/input-output-hk/marlowe/pull/182).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This section was rewritten and only the Division lemmas are now present, as they are the most relevant ones.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This section was [rewritten](https://github.com/input-output-hk/marlowe/pull/182) and the issue was fixed.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This happened because we were using a type alias, which gets fully expanded when being referenced by the documentation.
+[PR-182](https://github.com/input-output-hk/marlowe/pull/182) addresses this by making the whole theory in literate programming mode.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The definition in the specification is the exported Haskell code of the Isabelle code.
+This is a limitation that Isabelle has when importing code from other theories.
+A possible solution is to change the Accounts positive theory to be literal programming.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The statement was indeed unconnected from the description.
+In [PR-186](https://github.com/input-output-hk/marlowe/pull/186) we refactored the guarantees section and now the section "3.2 Finite contracts"
+includes the theorem `timedOutTransaction_closes_contract` which shows that after closing a timed-out contract, the accounts are emptied.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This was addressed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+In [PR-186](https://github.com/input-output-hk/marlowe/pull/186) we added a small explanation; in the future, we will probably expand on this by refactoring the guarantees as literate programming.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This section was redone in [PR-182](https://github.com/input-output-hk/marlowe/pull/182).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+In [PR-186](https://github.com/input-output-hk/marlowe/pull/186) we added the names of the theories in the specification.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We agree with this observation.
+New lemmas and theories are proven with this recommendation in mind, and some theories were refactored (like the Asset preservation).
+It will take some time until all theories are in this format.
 
@@
 >
-> ``` {.text escapeinside="||" mathescape="true"}
->   lemma insert_set:
->     "set (map fst (insert a b xs)) = set (map fst xs) |$\cup$| { a }"
-> 
->   lemma insert_sorted:
->     "List.sorted (map fst c) |$\Longrightarrow$| List.sorted (map fst (insert a b c))"
-> 
->   lemma insert_distinct :
->     "List.distinct (map fst c)
->      |$\Longrightarrow$|
->      List.sorted (map fst c)
->      |$\Longrightarrow$|
->      List.distinct (map fst (MList.insert a b c))"
+> [. . .]
+>
+> which then can be combined in the proof of `insert_valid` as follows:
+>
+> [. . .]
+>
+> The proofs of the lemmas can be found in […].
+
+We agree that the proposed proof is easier, but the lemma will become obsolete once we replace the custom `MList` with
+`HOL-Library/finiteMap`, which addresses other problems found in the report.
+
+### 2.3.6 Repeated verbose expression *(Severity: Low)*
+
+> **File `MoneyPreservation.thy`, lemma `removeMoneyFromAccount_preservation`, line *202***
+>
+> The expression
+>
 > ```
-> 
-> which then can be combined in the proof of `insert_valid` as follows:
-> 
-> ``` {.text escapeinside="||" mathescape="true"}
->   theorem insert_valid2 : "valid_map c |$\Longrightarrow$| valid_map (MList.insert a b c)"
->     using insert_sorted[of c a b] insert_distinct[of c a b] by fastforce
-> ```
-> 
-> The proofs of the lemmas can be found in […].
-
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
-
-### 2.3.6 Repeated verbose expression *(Severity: Low)*
-
-> **File `MoneyPreservation.thy`, lemma `removeMoneyFromAccount_preservation`, line *202***
-> 
-> The expression
-> 
-> ``` text
 >   giveMoney
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The `MoneyPreservation` was refactored into Asset preservation and now there is a single reference to `giveMoney`.
+As mentioned in observation 2.2.11, `giveMoney` is likely to be removed in the future.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The valid and positive preservation is needed in lower-level functions so that higher-level functions can use that fact when proving other properties.
+This makes `playTrace` not needing that proof as it is the highest-level function.
+The theorem could be useful on its own and might be added in the future.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We agree on this observation and it will be addressed in a future refactor of the theory file.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+* Lemma `transferMoneyBetweenAccounts_preserves_aux` was removed in [PR-161](https://github.com/input-output-hk/marlowe/pull/161).
+* Lemma `reduceOne_onlyIfNonEmptyState`, `valid_state_valid_choices` were removed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168).
+* Lemma `reduceContractUntilQuiescent_ifDifferentReduced` was promoted to a theorem with an explanation in [PR-168](https://github.com/input-output-hk/marlowe/pull/168).
+* Lemma `positiveMoneyInAccountOrNoAccount_gtZero_preservation` was removed in favor of `positiveMoneyInAccountOrNoAccount_gtZero_preservation` in [PR-168](https://github.com/input-output-hk/marlowe/pull/168).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This lemma was removed in [PR-161](https://github.com/input-output-hk/marlowe/pull/161).
 
@@
 >
-> ``` text
+> ```
 >   using
@@
 >
-> ``` text
+> ```
 >   using transferMoneyBetweenAccounts_preserves validAndPositive_state.simps by auto
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+Lemma `accountsArePositive` was removed in favor of `computeTransaction_gtZero` in [PR-168](https://github.com/input-output-hk/marlowe/pull/168).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+Lemma `computeTransaction_preserves_valid_state_aux` was removed and `computeTransaction_preserve_valid_state` simplified in [PR-168](https://github.com/input-output-hk/marlowe/pull/168).
 
@@
 >
-> ``` {.text escapeinside="||" mathescape="true"}
->   valid_map (((thisAccId, tok), money) # tail) |$\Longrightarrow$|
->   allAccountsPositive (((thisAccId, tok), money) # tail) |$\Longrightarrow$|
->   moneyInAccount thisAccId tok (((thisAccId, tok), money) # tail) > 0"
-> ```
+> [...]
 >
@@
 >
-> ``` {.text escapeinside="||" mathescape="true"}
->   cases "moneyInAccount accId tok ((thisAccIdTok, money) # tail) + moneyToPay |$\leq$| 0"
-> ```
+> [...]
 >
@@
 >
-> ``` {.text escapeinside="||" mathescape="true"}
->   cases "moneyInAccount accId tok ((thisAccIdTok, money) # tail) |$\leq$| 0"
-> ```
-
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+> [...]
+
+This lemma was removed in [PR-161](https://github.com/input-output-hk/marlowe/pull/161).
 
@@
 >
-> ``` text
->   lemma inIntervalIdempotency1 :
->     assumes "inInterval (x, y) (intersectInterval b c)"
->     shows "inInterval (x, y) b"
->   proof (cases b)
->     case [simp]:(Pair b1 b2)
->     thus ?thesis proof (cases c)
->       case (Pair c1 c2)
->       thus ?thesis using assms by (cases c1; cases c2; cases b1;cases b2; simp)
->     qed
->   qed
-> ```
+> [...]
 >
@@
 >
-> ``` {.text escapeinside="||" mathescape="true"}
->   have "(x < y) = (x |$\le$| y |$\land$| |$\neg$| y |$\le$| (x :: Token))"
->     by (meson less_Tok.simps less_Token_def less_eq_Token_def linearToken)
->   thus "(x < y) = (x |$\le$| y |$\land$| |$\neg$| y |$\le$| x)" by simp
-> ```
+> [...]
 >
@@
 >
-> ``` {.text escapeinside="||" mathescape="true"}
->   show "(x < y) = (x |$\le$| y |$\land$| |$\neg$| y |$\le$| (x :: Token))"
->     by (meson less_Tok.simps less_Token_def less_eq_Token_def linearToken)
-> ```
+> [...]
 >
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+These observations regarding readability will be addressed when we refactor the theory.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This issue was resolved in [PR-168](https://github.com/input-output-hk/marlowe/pull/168).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We agree on the observation, and will likely promote it to a theorem in the future.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We agree on this observation and it will be addressed in a future refactoring of the theory file.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The lemma was promoted to a theorem in [PR-168](https://github.com/input-output-hk/marlowe/pull/168).
 
@@
 >
-> ``` {.text escapeinside="||" mathescape="true"}
->   Reduced
->     ReduceNoWarning
->     (ReduceWithPayment (Payment party (Party party) tok2 money))
->     (state|$\llparenthesis$|accounts := newAccount|$\rrparenthesis$|) Close
->     =
->   Reduced wa eff newState newCont
-> ```
+> [...]
 >
@@
 >
-> The assumption\ $\mathit{env} = \llparenthesis \mathit{timeInterval} = (a, b) \rrparenthesis$ is unnecessary.
+> The assumption [...] is unnecessary.
 >
@@
 >
-> The assumption `$$\begin{aligned} \mathit{newState} = \llparenthesis &\mathit{accounts} = \mathit{newAccounts}, \mathit{choices} = \mathit{newChoices}, \\ &\mathit{boundValues} = \mathit{newBoundValues}, \mathit{minTime} = \mathit{newMinTime} \rrparenthesis\end{aligned}$$` is unnecessary.
+> The assumption [...] is unnecessary.
+
+We removed the unnecessary assumptions in [PR-168](https://github.com/input-output-hk/marlowe/pull/168).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
-
 ### 2.4.1 Variable shadowing in `applyAllLoop` *(Severity: Medium)*
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The variable was renamed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The theory was refactored into `AssetsPreservation` and the function is now called `assetsInExternalPayment`.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The typo was fixed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We agree that it is confusing, and it will be refactored in the future.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This was an artifact of showing the exported haskell code. It is resolved in [PR-182](https://github.com/input-output-hk/marlowe/pull/182).
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+This function was documented in [PR-168](https://github.com/input-output-hk/marlowe/pull/168) and moved to `TimeRange.thy` in [PR-182](https://github.com/input-output-hk/marlowe/pull/182).
+The function is not part of the Marlowe spec yet: it is a helper we hope will be useful in the future to construct the transactions.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We agree with the observation and it will be addressed in a future refactoring.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We agree on the observations, for the moment some simple refactoring and comments were added as part of [PR-180](https://github.com/input-output-hk/marlowe/pull/180/files#diff-638dc578b752590b5f2bace45f1e1ae9e522351f62056e6f92aa272424657b03).
+Further refactoring and optional renaming might happen in the future.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+We agree on this observation and it will addressed in a future refactoring of the theory file.
 
@@
 
-![#f03c15](https://placehold.co/15x15/f03c15/f03c15.png) TODO: Isabelle-related text to be added here, since it relates to the Isabelle specification.
+The function `isValidInterval` was removed in favor of `validTimeInterval` in [PR-168](https://github.com/input-output-hk/marlowe/pull/168/files#diff-9c936a78b5d8eed5683713bb31c59366b2a0794c5f11e9807f7fa61d1b11b7c2L230)
 
@@
 For the time being, the `contractLifespanUpperBound` and `isClose` have been retained in `Language.Marlowe.Core.V1.Semantics`.
-They will be relocated in the future if validator and non-validator code is separated into different modules.
+They will be relocated in the future if validator and non-validator code are separated into different modules.
 

### 2023-05-11 response-to-audit-report.md
PLT-5900: small adjustments

@@
 
-Marlowe consists of a domain-specific language (DSL) for the financial industry and for peer-to-peer financial contracts, with a reference semantics written in the Isabelle theorem-proving language and an implementation for the Cardano blockchain written in the Haskell/Plutus programming language.
-Early in 2023 the Marlowe specifications, on-chain implementation, and validator-test suite were audited by Tweag's High Assurance Software Group: their report is available online and it categorizes findings as either high, medium, or low severity.
-The present document summarizes the subsequent changes, made in response to the audit's findings, to the Marlowe specifications, its implementation as a Plutus smart-contract validator, and the property-based tests for that validator.
-All of the high-severity findings have been remedied or mitigated through changes to specifications, proofs, or implementation.
-Nearly all of the medium- and low-severity ones have also been addressed; justification for not addressing the remainder has been provided.
-
-The high-severity findings related to handling of negative deposits, prevention of "double satisfaction", enforcement of state invariants, an implementation difference between the formal specification versus the Plutus implementation, and the Isabelle proof of money preservation.
-Marlowe's Plutus validator has been altered to resolve these first three findings so that it correctly prevents negative deposits from altering the balance of internal accounts, it only allows other Plutus scripts to run during transactions that do not make Marlowe payments, and it rigorously enforces invariants for initial and final states.
+Marlowe is a domain-specific language (DSL) for multi-party financial contracts.
+Operational semantics are formalized in the Isabelle proof-assistant and an implementation of Marlowe for the Cardano blockchain is written in the Haskell/Plutus programming language.
+Early in 2023 the Marlowe specification, the on-chain implementation and the validator test-suite were audited by Tweag's High Assurance Software Group.
+The report is available online.
+It categorizes findings of either high, medium or low severity.
+The present document summarizes the subsequent changes made to the Marlowe specification, the implementation of the validator and the property-based tests for the validator.
+All of the high-severity findings have been remedied or mitigated through changes to the specification, the proofs in Isabelle or the implementation.
+Nearly all of the medium-severity and low-severity findings have also been addressed; justification for not addressing the remainder has been provided.
+
+The high-severity findings comprise handling of negative deposits, prevention of "double satisfaction", enforcement of state invariants, an implementation difference between the formal specification versus the Plutus implementation and the proof of money preservation in Isabelle.
+The Marlowe validator has been altered to resolve these first three findings so that it correctly prevents negative deposits from altering the balance of internal accounts, it only allows other Plutus scripts to run during transactions that do not make Marlowe payments, and it rigorously enforces invariants for initial and final states.
 The fourth finding regarding the Plutus implementation was mitigated via code analysis and property-based testing.
@@
 The medium-severity findings typically suggested improvements to the Isabelle proofs, addition of more extensive property-based testing, correction of infelicities in the implementation, or safe-coding improvements.
-The low-severity findings generally centered on the usefulness of code comments, naming of variable bindings, omissions in specifications, and correction of typographical errors.
+The low-severity findings generally centered on the usefulness of code comments, naming of variable bindings, omissions in specifications and correction of typographical errors.
 All of the findings that affect the clarity of the specifications, the coverage of testing, or the robustness of implementation were made.
@@
 
-The changes outlined here have not been re-audited.
-The alterations to Marlowe's Plutus validator (detailed in the Appendix) have been kept to the minimum needed to address the audit findings, but avoiding extensive changes that would might have lessened the relevance of the audit.
-
-Finally, please note that the audit's scope included the Marlowe language and its Plutus interpreter, but not individual Marlowe contracts or off-chain code.
+The changes outlined here have not been re-audited. The alterations to the Marlowe validator (detailed in the Appendix) have been kept to the minimum needed to address exclusively the audit findings.
+
+Finally, please note that the scope of the audit included the Marlowe language and its Plutus interpreter, but not individual Marlowe contracts or off-chain code.
 Although the Marlowe language and its implementation on Cardano aim to make smart contracts safer, it is nevertheless still possible to design individual instances of Marlowe contracts that exhibit unwanted behavior.
@@
 
-We did not mention the merkleization in the specification document as it was considered an implementation detail.
+We did not mention merkleization in the specification document as it was considered an implementation detail.
 This does lessen the applicability of the Isabelle proofs to the different implementations (Marlowe on Cardano, Marlowe in PureScript, etc).
@@
 To test the correctness of merkleized cases, the marlowe-cardano repository has specific tests.
-The Merkleization formalization will be introduced in a future version of the specification, the initial formulation started in [PR-171](https://github.com/input-output-hk/marlowe/pull/171).
+The formalization of merkleization will be introduced in a future version of the specification (initial work started in [PR-171](https://github.com/input-output-hk/marlowe/pull/171).
 
@@
 
-The audit report's main and high-severity concerns for the Plutus validators for Marlowe on Cardano fell into to four clusters:
+The audit report's main and high-severity concerns for the Plutus validators for Marlowe on Cardano fell into four clusters:
 
@@
 
-This function is confusing, we will likely to remove it in a future release.
+This function is confusing, we will likely remove it in a future release.
 
@@
 
-> **File `specification-v3-rc1.pdf`, Section `3.3 Possitive Accounts`, page *30***
+> **File `specification-v3-rc1.pdf`, Section `3.3 Positive Accounts`, page *30***
 >

### 2023-05-11 response-to-audit-report.md
PLT-5900: small adjustments

@@
 Marlowe is a domain-specific language (DSL) for multi-party financial contracts.
-Operational semantics are formalized in the Isabelle proof-assistant and an implementation of Marlowe for the Cardano blockchain is written in the Haskell/Plutus programming language.
+Operational semantics are formalized in the Isabelle proof-assistant and an interpreter for Marlowe for the Cardano blockchain is provided as Plutus validator script.
 Early in 2023 the Marlowe specification, the on-chain implementation and the validator test-suite were audited by Tweag's High Assurance Software Group.
@@
 The fourth finding regarding the Plutus implementation was mitigated via code analysis and property-based testing.
-The fifth finding about the money-preservation proof was remedied by revising the Isabelle proof.
+The fifth finding about the proof of the money preservation theorem was remedied by revising the Isabelle code.
 

### 2023-05-11 response-to-audit-report.md
PLT-5900: small adjustments

@@
 The high-severity findings comprise handling of negative deposits, prevention of "double satisfaction", enforcement of state invariants, an implementation difference between the formal specification versus the Plutus implementation and the proof of money preservation in Isabelle.
-The Marlowe validator has been altered to resolve these first three findings so that it correctly prevents negative deposits from altering the balance of internal accounts, it only allows other Plutus scripts to run during transactions that do not make Marlowe payments, and it rigorously enforces invariants for initial and final states.
+The Marlowe validator has been altered to resolve these first three findings so that it correctly prevents negative deposits from altering the balance of internal accounts, it only allows other Plutus scripts to run during transactions that do not make Marlowe payments and it rigorously enforces invariants for initial and final states.
 The fourth finding regarding the Plutus implementation was mitigated via code analysis and property-based testing.
@@
 All of the findings that affect the clarity of the specifications, the coverage of testing, or the robustness of implementation were made.
-Numerous additional property-based tests were added to the test suite as a result of audit findings.
+Numerous additional property-based tests were added to the test suite as a result of the audit findings.
 
@@
 
-Finally, please note that the scope of the audit included the Marlowe language and its Plutus interpreter, but not individual Marlowe contracts or off-chain code.
+Finally, please note that the scope of the audit included the Marlowe language and interpreter, but not individual Marlowe contracts or off-chain code.
 Although the Marlowe language and its implementation on Cardano aim to make smart contracts safer, it is nevertheless still possible to design individual instances of Marlowe contracts that exhibit unwanted behavior.
@@
 
-We thank the High Assurance Software Group at Tweag for their diligent and expert analyses of the Marlowe specifications, proofs, implementations, and tests and their clear and thorough communication of their findings.
+We thank the High Assurance Software Group at Tweag for their diligent and expert analyses of the Marlowe specifications, proofs, implementations and tests and their clear and thorough communication of their findings.
 
@@
 The Money preservation theory had only taken into account the amount of assets and not the type.
-This meant that the proof was not considering a case where a contract could receive `20 ada`,  `15 djed`, and pay back `20 djed` and `15 ada`.
+This meant that the proof was not considering a case where a contract could receive `20 ada`, `15 djed` and pay back `20 djed` and `15 ada`.
 We added a new `MultiAssets` type and refactored the Isabelle code (without modifying the interpreter) to prove that the Assets are preserved.

### 2023-05-11 response-to-audit-report.md
PLT-5900: small adjustments

@@
 All of the high-severity findings have been remedied or mitigated through changes to the specification, the proofs in Isabelle or the implementation.
-Nearly all of the medium-severity and low-severity findings have also been addressed; justification for not addressing the remainder has been provided.
+Nearly all of the medium-severity and low-severity findings have also been addressed, otherwise justification for not addressing a finding has been provided.
 
@@
 To test the correctness of merkleized cases, the marlowe-cardano repository has specific tests.
-The formalization of merkleization will be introduced in a future version of the specification (initial work started in [PR-171](https://github.com/input-output-hk/marlowe/pull/171).
+The formalization of merkleization will be introduced in a future version of the specification (initial work started in [PR-171](https://github.com/input-output-hk/marlowe/pull/171)).
 

### 2023-05-11 response-to-audit-report.md
Update response-to-audit-report.md

@@
 
-The medium-severity findings typically suggested improvements to the Isabelle proofs, addition of more extensive property-based testing, correction of infelicities in the implementation, or safe-coding improvements.
+The medium-severity findings typically suggested improvements to the Isabelle proofs, addition of more extensive property-based testing, correction of infelicities in the implementation or safe-coding improvements.
 The low-severity findings generally centered on the usefulness of code comments, naming of variable bindings, omissions in specifications and correction of typographical errors.
-All of the findings that affect the clarity of the specifications, the coverage of testing, or the robustness of implementation were made.
+All of the findings that affect the clarity of the specifications, the coverage of testing or the robustness of implementation were made.
 Numerous additional property-based tests were added to the test suite as a result of the audit findings.
@@
 Such undesirable behavior may result either from poor contract design or lack of testing and analysis of the contract.
-[A best-practices guide](https://github.com/input-output-hk/marlowe-cardano/blob/main/marlowe/best-practices.md) and [a security guide](https://github.com/input-output-hk/marlowe-cardano/blob/main/marlowe/security.md) provide advice for the safe and secure design and testing of individual Marlowe contracts, which can themselves be subject to audit or certification.
+[A best-practices guide](https://github.com/input-output-hk/marlowe-cardano/blob/main/marlowe/best-practices.md) and [a security guide](https://github.com/input-output-hk/marlowe-cardano/blob/main/marlowe/security.md) provide advice for the safe and secure design and testing of individual Marlowe contracts which can themselves be subject to audit or certification.
 
@@
 
-We did not mention the merkleization in the specification document as it was considered an implementation detail.
+We did not mention merkleization in the specification document as it was considered an implementation detail.
 This does lessen the applicability of the Isabelle proofs to the different implementations (Marlowe on Cardano, Marlowe on PureScript, etc).
